### 2011-03-04 report/report.tex
minor changes + new directory for the report

\documentclass[a4paper,english,titlepage]{article}

\usepackage{mathcomp,amssymb,mathptm,verimagstudent}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}

\newtheorem{definition}{Definition}[section]
\newtheorem{theorem}{Theorem}[section]

\title{Analyse de programme par découverte de chemin}
\author{Julien Henry}
\date{2011}
\shorttitle{Analyse de programme par découverte de chemin}
\institute{Grenoble-INP}
\abstract{an abstract}
\keywords{keywords}
\tutor{David Monniaux - Matthieu Moy}
\notes{some notes}
\webaddress{Julien.Henry@imag.fr}
\reporttype{Rapport de M2R}

\begin{document}

  \maketitlepage

  \tableofcontents

  \newpage

\section{Introduction}

\part{Interprétation abstraite}

	Pour calculer les variables live en SSA, des algorithmes plus optimisés sont
	expliqués dans la thèse de B. Boissinot\cite{Boi10}

	Interpretation abstraite
	\cite{CC77}
	\cite{CH78}

	\cite{GopanR06}

	Utilisation de la bibliothèque APRON \cite{JM09}

\part{Contribution}

  \section{Conclusion}

  \appendix

\bibliographystyle{plain}
  \bibliography{report}
\end{document}

### 2011-03-04 src/AI.cc

@@ -141,7 +141,7 @@ void AI::computeEnv(Node * n) {
 		visit(*i);
 	}
 	
-	// for each predecessor, we iterate on the their variables, and we insert
+	// for each predecessor, we iterate on their variables, and we insert
 	// them if they are associated to a value which is still live in our Block
 	// We do this for both int and real variables
 	for (pred_iterator p = pred_begin(b), E = pred_end(b); p != E; ++p) {
@@ -202,12 +202,9 @@ void AI::computeHull(
 			}
 			// we still need to add phi variables into our domain 
 			// and assign them the the right value
-			fouts() << "AAAAAAAAAAAAAAAAAAAAAAAAAAAAAAAAAAAAAAAAAAAAAAAAA\n";
-			X->print();
-
-			for (int i = 0; i < n->phi_vars[pred].name.size()>0; i++) {
-				ap_environment_fdump(stdout,n->phi_vars[pred].expr[i].env);
-			}
+			//for (int i = 0; i < n->phi_vars[pred].name.size()>0; i++) {
+			//	ap_environment_fdump(stdout,n->phi_vars[pred].expr[i].env);
+			//}
 			X->assign_texpr_array(
 					&n->phi_vars[pred].name[0],
 					&n->phi_vars[pred].expr[0],
@@ -513,7 +510,6 @@ void AI::visitSwitchInst (SwitchInst &I){
 	
 	for (int i = 0; i < num; i++) {
 		ConstantInt * CaseValue = I.getCaseValue(i);
-		BasicBlock * Successor = I.getSuccessor(i);
 		ap_texpr1_t * CaseExp = get_ap_expr(n,CaseValue);
 
 		common_environment(&ConditionExp,&CaseExp);

### 2011-03-16 src/AI.cc
beginning of the presentation

@@ -985,6 +985,8 @@ void AI::visitInstAndAddVarIfNecessary(Instruction &I) {
 	
 	if (get_ap_type((Value*)&I, ap_type)) return;
 
+	if (!LV->isLiveThroughBlock(&I,n->bb) && !LV->isUsedInBlock(&I,n->bb)) return;
+	
 	if (ap_type == AP_RTYPE_INT) { 
 		env = ap_environment_alloc(&var,1,NULL,0);
 	} else {

### 2011-03-21 ex/boustrophedon_expansed.c
presentation

@@ -4,7 +4,7 @@ int main() {
 	int x;
 	int d;
 	x = 0;
-	d = 0;
+	d = 1;
 
 	while(1) {
 		if (x > 0) {

### 2011-04-08 article/gpl2011.tex
update in the construction of rho : lots of variables didn't need to be primed

\documentclass{llncs}
\usepackage{a4wide}
\usepackage[french]{babel}
\usepackage[utf8]{inputenc}

\title{Analyse statique par Interprétation Abstraite}
\titlerunning{}
\subtitle{Raffinement de la méthode en utilisant le SMT-solving}
\author{Julien Henry}
\institute{CNRS - Verimag}

\begin{document}
\maketitle

\begin{abstract}
	L'analyse statique consiste à trouver à la compilation des propriétés sur un
	programme, en particulier l'ensemble des valeurs possibles pour chacunes des
	variables durant l'exécution. Ceci permet par exemple de prouver que le
	programme ne va pas faire de divisions par 0, ne va pas avoir de
	débordements arithmétiques, ou encore à calculer des invariants de boucle,
	etc. Trouver l'ensemble exact des valeurs possibles pour les variables est
	impossible, c'est pourquoi on a recours à des surapproximations : on calcule
	un ensemble plus grand, qui contient toutes les valeurs possibles des
	variables pendant l'exécution, mais aussi d'autres valeurs pour simplifier
	l'analyse : c'est le cas de l'Interprétation Abstraite, qui est une méthode
	bien connue de vérification statique.  
	On peut alors essayer de limiter cette surapproximation, grâce à une
	technique par SMT-solving, pour gagner en précision.
\end{abstract}

\section{Introduction}

\section{Interprétation abstraite}

\end{document}

### 2011-04-08 ex/compile_llvm.sh

@@ -52,7 +52,8 @@ if [ -z $OUTPUT ] ; then
 fi
 
 clang -DNDEBUG -fno-exceptions -emit-llvm -c $FILENAME -o $OUTPUT
-opt -mem2reg -loopsimplify -lowerswitch $OUTPUT -o $OUTPUT
+#opt -mem2reg -loopsimplify -lowerswitch $OUTPUT -o $OUTPUT
+opt -mem2reg -lowerswitch $OUTPUT -o $OUTPUT
 
 if [ $GRAPH -eq 1 ] ; then
 	opt -dot-cfg $OUTPUT -o $OUTPUT

### 2011-04-08 ex/nested_loop3.c

@@ -2,8 +2,6 @@ void mat_product(unsigned n) {
 if (n>0) {
   for(unsigned i=0; i<n; i++) {
     for(unsigned j=0; j<n; j++) {
-      for(unsigned k=0; k<n; k++) {
-      }
     }
   }
 }

### 2011-04-08 src/AI.cc

@@ -316,9 +316,9 @@ void AI::computeTransform (Node * n, std::list<BasicBlock*> path, Abstract &Xtem
 	focuspath.assign(path.begin(), path.end());
 	focusblock = 0;
 	
-	fouts() << "POLYHEDRA BEGINNING\n";
-	fouts().flush();
-	Xtemp.print();
+	//fouts() << "POLYHEDRA BEGINNING\n";
+	//fouts().flush();
+	//Xtemp.print();
 
 	std::list<BasicBlock*>::iterator B = path.begin(), E = path.end();
 	for (; B != E; ++B) {
@@ -348,9 +348,9 @@ void AI::computeTransform (Node * n, std::list<BasicBlock*> path, Abstract &Xtem
 	//Xtemp.set_top(env);
 	Xtemp.change_environment(env);
 
-	fouts() << "POLYHEDRA AFTER CHANGE OF ENVIRONMENT\n";
-	fouts().flush();
-	Xtemp.print();
+	//fouts() << "POLYHEDRA AFTER CHANGE OF ENVIRONMENT\n";
+	//fouts().flush();
+	//Xtemp.print();
 
 	std::list<std::vector<ap_tcons1_array_t*>*>::iterator i, e;
 	for (i = constraints.begin(), e = constraints.end(); i!=e; ++i) {
@@ -370,14 +370,14 @@ void AI::computeTransform (Node * n, std::list<BasicBlock*> path, Abstract &Xtem
 			Xtemp.join_array(env,A);
 		}
 	}
-	fouts() << "POLYHEDRA AFTER MEETING CONSTRAINTS\n";
-	fouts().flush();
-	Xtemp.print();
+	//fouts() << "POLYHEDRA AFTER MEETING CONSTRAINTS\n";
+	//fouts().flush();
+	//Xtemp.print();
 
 	Xtemp.assign_texpr_array(&PHIvars.name[0],&PHIvars.expr[0],PHIvars.name.size(),NULL);
-	fouts() << "POLYHEDRA AFTER PHI ASSIGNATIONS\n";
-	fouts().flush();
-	Xtemp.print();
+	//fouts() << "POLYHEDRA AFTER PHI ASSIGNATIONS\n";
+	//fouts().flush();
+	//Xtemp.print();
 }
 
 
@@ -436,9 +436,9 @@ void AI::computeNode(Node * n) {
 		Join.push_back(new Abstract(Xtemp));
 		Xtemp->join_array(Xtemp->main->env,Join);
 
[…]

### 2011-04-08 src/SMT.cc

@@ -379,10 +379,11 @@ void SMT::computeRhoRec(Function &F,
 		BasicBlock * b,
 		BasicBlock * dest,
 		bool newPr,
-		std::set<BasicBlock*> visited) {
+		std::set<BasicBlock*> * visited) {
 
-	if (visited.count(b)) return;
-	visited.insert(b);
+	bool first = (visited->count(b) > 0);
+	visited->insert(b);
+	
 
 	if (!Pr[&F]->count(b) || newPr) {
 		// we recursively construct Rho, starting from the predecessors
@@ -398,6 +399,11 @@ void SMT::computeRhoRec(Function &F,
 			}
 		}
 	}
+
+	if (first) return;
+	fouts() << "constructing formula for node " << b << "\n";
+	fouts().flush();
+
 
 	// we create a boolean reachability predicate for the basicblock
 	SMT_var bvar = man->SMT_mk_bool_var(getNodeName(b,false));
@@ -442,16 +448,16 @@ void SMT::computeRhoRec(Function &F,
 	}
 	
 	// x' = x if we don't reach this block
-	std::vector<SMT_expr> affect;
-	std::set<Value*> empty;
-	std::set<Value*>::iterator it = exist_prime.begin(), et = exist_prime.end();
-	for (; it != et; ++it) {
-		affect.push_back(man->SMT_mk_eq(getValueExpr(*it,empty), getValueExpr(*it,exist_prime)));
-	}
-	std::vector<SMT_expr> implies;
-	implies.push_back(man->SMT_mk_expr_from_bool_var(bvar));
-	implies.push_back(man->SMT_mk_and(affect));
-	rho_components.push_back(man->SMT_mk_or(implies));
+//	std::vector<SMT_expr> affect;
+//	std::set<Value*> empty;
+//	std::set<Value*>::iterator it = exist_prime.begin(), et = exist_prime.end();
+//	for (; it != et; ++it) {
+//		affect.push_back(man->SMT_mk_eq(getValueExpr(*it,empty), getValueExpr(*it,exist_prime)));
+//	}
+//	std::vector<SMT_expr> implies;
+//	implies.push_back(man->SMT_mk_expr_from_bool_var(bvar));
+//	implies.push_back(man->SMT_mk_and(affect));
+//	rho_components.push_back(man->SMT_mk_or(implies));
[…]

### 2011-04-08 src/SMT.h

@@ -56,7 +56,7 @@ class SMT : public ModulePass, public InstVisitor<SMT> {
 							BasicBlock * b,
 							BasicBlock * dest,
 							bool newPr,
-							std::set<BasicBlock*> visited);
+							std::set<BasicBlock*> * visited);
 		void computeRho(Function &F);
 	public:
 		static char ID;

### 2011-04-12 article/gpl2011.tex
first version of the article gpl2011

@@
 \usepackage[utf8]{inputenc}
+\usepackage{listings}
+\usepackage{placeins}
+\usepackage{pgf}
+\usepackage{tikz}
+\usetikzlibrary{arrows,automata}
 
-\title{Analyse statique par Interprétation Abstraite}
+\title{Analyse statique par découverte de chemin}
 \titlerunning{}
-\subtitle{Raffinement de la méthode en utilisant le SMT-solving}
+\subtitle{}
 \author{Julien Henry}
-\institute{CNRS - Verimag}
+\institute{CNRS - Verimag \\ \email{Julien.Henry@imag.fr}}
 
@@
 
-\begin{abstract}
+\section{Introduction}
+
 L'analyse statique consiste à trouver à la compilation des propriétés sur un
-	programme, en particulier l'ensemble des valeurs possibles pour chacunes des
+programme, comme l'ensemble des valeurs possibles pour chacunes des
 variables durant l'exécution. Ceci permet par exemple de prouver que le
-	programme ne va pas faire de divisions par 0, ne va pas avoir de
-	débordements arithmétiques, ou encore à calculer des invariants de boucle,
-	etc. Trouver l'ensemble exact des valeurs possibles pour les variables est
-	impossible, c'est pourquoi on a recours à des surapproximations : on calcule
-	un ensemble plus grand, qui contient toutes les valeurs possibles des
-	variables pendant l'exécution, mais aussi d'autres valeurs pour simplifier
-	l'analyse : c'est le cas de l'Interprétation Abstraite, qui est une méthode
-	bien connue de vérification statique.  
-	On peut alors essayer de limiter cette surapproximation, grâce à une
-	technique par SMT-solving, pour gagner en précision.
-\end{abstract}
+programme ne va pas avoir de débordements arithmétiques, à calculer des
+invariants de boucle, à faire des optimisations de code lors de la compilation, etc.
 
-\section{Introduction}
+L'Interprétation Abstraite est une méthode classique d'analyse statique, qui consiste à calculer une surapproximation de
+l'ensemble des états possibles d'un programme numérique. 
+L'état d'un programme est caractérisé par la valeur de chacunes de ses variables
+numériques, et par la position à laquelle on se trouve dans le code.
+Un cas particulier
+d'Interprétation Abstraite est l'analyse par relations linéaires: On surapproxime
+l'ensemble des états possibles du programme par un polyhèdre convexe, pour découvrir des propriétés linéaires entre les différentes variables
+numériques du programme.
+Cette méthode implique des surapproximations, nécessaires pour rendre
+l'analyse moins coûteuse : il faut alors essayer de
+trouver le bon compromis entre la précision et le coût. De nombreux travaux
+consistent à trouver de nouvelles techniques pour améliorer cette précision à
+moindre coût. La méthode proposée ici consiste à s'aider des
+progrès dans la recherche sur le SMT-solving
+pour raffiner la méthode
+d'interprétation abstraite.
 
@@
 
+L'application de l'Interprétation Abstraite à l'analyse de
+relations linéaires consiste à attacher à chaque point de contrôle du programme un polyhèdre convexe dont les dimensions sont les variables numériques. On travaille sur
+le graphe de flot de contrôle du programme (voir figure) : le programme peut
+être vu comme un graphe où chaque transition est étiquetée par une ou plusieurs
+instructions, et peut être soumise à condition(s). On calcule itérativement les
+différents polyhèdres $A_i$ à chaque noeud $s_i$ du graphe de la façon suivante:
+
+Au départ, le polyhèdre de l'état initial est l'espace entier (aucune
+contrainte sur les variables), et celui des autres états est le polyhèdre vide
+(l'état est considéré comme étant inaccessible).
+
+Puis, on itère ce calcul pour tous les états du graphe, tant que la suite des
+polyhèdres $A_i$ est croissante pour la relation d'inclusion :
+\begin{itemize}
+\item On part d'un état $i$, et on met à jour chaque état $j$ où $j$ est un
+successeur de $i$ dans le graphe, en calculant l'image $Y$ du polyhèdre $A_i$
+par la transition $i \rightarrow j$. 
+\item Le nouveau polyhèdre $A_j$ est
+l'enveloppe convexe de $Y$ avec l'ancienne valeur de $A_j$, notée $Y \sqcup
+A_j$, qui est le plus petit polyhèdre convexe incluant $A_j$ et $Y$. Les valeurs
+successives de $A_j$ au cours de l'analyse définissent donc bien une suite croissante.
+\end{itemize}
+Après un certain nombre d'itérations, chaque polyhèdre a atteint une limite, et
+l'analyse s'arrête \footnote{En fait, pour garantir la terminaison, on a recours
+à un opérateur dit d'élargissement, qui fait une extrapolation lorsqu'un
+polyhèdre risque de croître indéfiniment.}. Les polyhèdres résultats permettent de déduire des
+invariants sur les variables du programme.
+
+\begin{figure}[!h]
+\begin{minipage}[b]{0.5\linewidth}
+\begin{lstlisting}
+// position s1
+x = 0;
+i = 0;
+//position s2
+while (i < n) {
+	// position s3
+	i++;
+	x = x+2;
+}
+// position s4
+\end{lstlisting}
+\end{minipage}
+\begin{minipage}[b]{0.5\linewidth}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1.5cm,
+semithick]
+  \tikzstyle{every state}=[]
+
+	\node[initial,state] (A)                    {$s_1$};
+	\node[state]         (B) [below of=A] {$s_2$};
+	\node[state]         (C) [below of=B] {$s_3$};
+	\node[state]         (D) [left of=B]       {$s_4$};
+
+	\path (A) edge               node {$x=0;\ i=0;$} (B);
+	\path (B) edge               node {$i < n$} (C);
+	\path (B) edge               node {$i\geq n$} (D);
+	\path (C) edge [out=0, in=0, distance=2cm] node [right] {$i++;\ x=x+2;$} (B);
+
+\end{tikzpicture}
+\end{minipage}
+\caption{Exemple de programme, et son graphe de flot de contrôle associé}
+\end{figure}
+%\FloatBarrier
+
+Dans l'exemple de la figure 1, en supposant que $n>0$, l'analyse donne:
+\begin{itemize}
+\item $A_1$ est le polyhèdre $\{n>0\}$: au point $s_1$, on sait
+seulement que $n>0$.
+\item $A_2$ est le polyhèdre $\{0 \leq x \leq 2n,\ 0 \leq i \leq n\}$.
+\item $A_3$ est le polyhèdre $\{0 \leq x < 2n,\ 0 \leq i < n\}$.
+\item $A_4$ est le polyhèdre $\{i=n,\ x=2n\}$. On connaît donc directement la
+valeur des variables à la fin du programme, ce qui permet des optimisations de
+compilation.
+\end{itemize}
+
+\section{Découverte de chemin}
+
+Nous proposons un raffinement de la méthode d'analyse par Interprétation
+Abstraite, en guidant plus astucieusement les itérations de point fixe. En
+effet, la vitesse et la précision de l'analyse dépend de l'ordre dans lequel on
+traite les noeuds du graphe.
+
+Au lieu de travailler sur le graphe de flot de contrôle et de calculer un
+polyhèdre à chacun des noeuds, on choisit un sous-ensemble $P$ des noeuds et on fait
+le calcul de point fixe sur le graphe ne contenant que les noeuds de $P$, et où
+une transition correspond à un chemin possible dans le graphe de départ.
+
+\begin{figure}
+\begin{minipage}[b]{0.5\linewidth}
+\begin{center}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
+semithick]
+  \tikzstyle{every state}=[minimum size=0.4cm]
+  \tikzstyle{PRstate}=[fill=black]
+
+	\node[initial,state,PRstate] (A)                    {};
+	\node[state,PRstate]         (B) [below of=A] {};
+	\node[state,PRstate]         (C) [left of=B] {};
+	\node[state]         (D) [below left of=B]       {};
+	\node[state]         (E) [below right of=B]       {};
+	\node[state]         (F) [below left of=E]       {};
+
+	\path (A) edge               node {} (B);
+	\path (B) edge               node {} (C);
+	\path (B) edge               node {} (D);
+	\path (B) edge               node {} (E);
+	\path (D) edge               node {} (F);
+	\path (E) edge               node {} (F);
+	\path (F) edge [out=0, in=0, distance=2cm] node [right] {} (B);
+\end{tikzpicture}
+\end{center}
+\end{minipage}
+\begin{minipage}[b]{0.5\linewidth}
+\begin{center}
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1.3cm,
+semithick]
+  \tikzstyle{every state}=[minimum size=0.4cm]
+  \tikzstyle{PRstate}=[fill=black]
+
+	\node[initial,state,PRstate] (A)                    {};
+	\node[state,PRstate]         (B) [below of=A] {};
+	\node[state,PRstate]         (C) [left of=B] {};
+
+	\path (A) edge               node {} (B);
+	\path (B) edge               node {} (C);
+	\path (B) edge     [loop right]          node {} (B);
+	\path (B) edge     [loop below]          node {} (B);
+\end{tikzpicture}
+\end{center}
+\end{minipage}
+\caption{En noir, les noeuds de $P$. À gauche, un graphe de
+flot de contrôle, et à droite son équivalent avec uniquement les noeuds de $P$.
+L'un des noeuds contient deux boucles, chacune correspondant à un chemin
+possible dans le graphe de gauche.}
+\end{figure}
+Ce nouveau graphe contient moins d'états, mais un nombre potentiellement
+exponentiellement plus grand que le graphe d'origine. C'est pourquoi la
+technique consiste à ne pas construire explicitement ce graphe, et à procéder
+comme suit:
+
+Lors de l'analyse d'un noeud $i \in P$ du graphe, on encode la question suivante
+sous forme de formule logique: ``Existe-t-il un chemin entre $i$ et $j \in P$,
+tel que les variables numériques sont dans $A_i$ au noeud $i$, et tel que les
+variables après le passage par ce chemin ne sont pas dans $A_j$ ?'', et on la
+résout avec un SMT-solver. Si la réponse est ``oui'', celà veut dire qu'on a
+trouvé un chemin qui fait avancer l'analyse, et on peut alors calculer la
+transformation associée à ce chemin, et mettre à jour $A_j$. L'analyse d'un
+noeud $i$ s'arrête lorsque plus aucun chemin n'est trouvé.
+
+\section{Conclusion}
+
+Nous proposons une méthode pour améliorer la précision de l'analyse par
+Interprétation Abstraite, en guidant l'analyse vers les chemins du graphe à
+explorer. Le fait de considérer des chemins complets et non pas de simples
+transitions permet de gagner en précision, et l'utilisation du SMT-solving
+permet de limiter le coût des calculs et la consommation mémoire.
+
 \end{document}

### 2011-04-12 article/gpl2011.tex
implementation of narrowing phase OK

@@
 Un cas particulier
-d'Interprétation Abstraite est l'analyse par relations linéaires: On surapproxime
-l'ensemble des états possibles du programme par un polyhèdre convexe, pour découvrir des propriétés linéaires entre les différentes variables
+d'Interprétation Abstraite est l'analyse par relations linéaires: on surapproxime
+l'ensemble des états possibles du programme à un certain point par un polyhèdre convexe, pour découvrir des propriétés linéaires entre les différentes variables
 numériques du programme.
@@
 pour raffiner la méthode
-d'interprétation abstraite.
+d'Interprétation Abstraite.
 
@@
 relations linéaires consiste à attacher à chaque point de contrôle du programme un polyhèdre convexe dont les dimensions sont les variables numériques. On travaille sur
-le graphe de flot de contrôle du programme (voir figure) : le programme peut
+le graphe de flot de contrôle du programme (Fig. 1) : le programme peut
 être vu comme un graphe où chaque transition est étiquetée par une ou plusieurs
@@
 Abstraite, en guidant plus astucieusement les itérations de point fixe. En
-effet, la vitesse et la précision de l'analyse dépend de l'ordre dans lequel on
+effet, la vitesse et la précision de l'analyse dépendent de l'ordre dans lequel on
 traite les noeuds du graphe.
@@
 \end{figure}
-Ce nouveau graphe contient moins d'états, mais un nombre potentiellement
-exponentiellement plus grand que le graphe d'origine. C'est pourquoi la
+Ce nouveau graphe contient moins d'états, mais un nombre de transitions potentiellement
+exponentiellement plus grand que dans le graphe d'origine. C'est pourquoi la
 technique consiste à ne pas construire explicitement ce graphe, et à procéder

### 2011-04-12 src/AI.cc

@@ -63,7 +63,7 @@ bool AI::runOnModule(Module &M) {
 		n = Nodes[b];
 		fouts() << "\n\nRESULT FOR BASICBLOCK: -------------------" << *b << "-----\n";
 		fouts().flush();
-		n->X->print(true);
+		n->Y->print(true);
 
 		fflush(stdout);
 		delete it->second;
@@ -117,6 +117,7 @@ void AI::printPath(std::list<BasicBlock*> path) {
 void AI::computeFunction(Function * F) {
 	BasicBlock * b;
 	Node * n;
+	Node * current;
 
 	// A = {first basicblock}
 	b = F->begin();
@@ -142,13 +143,28 @@ void AI::computeFunction(Function * F) {
 	computeEnv(n);
 	n->create_env(&env);
 	n->X->set_top(env);
+	n->Y = new Abstract(man,env);
+	n->Y->set_top(env);
 	A.push(n);
 
 	// Simple Abstract Interpretation algorithm
 	while (!A.empty()) {
-		n = A.top();
+		current = A.top();
 		A.pop();
-		computeNode(n);
+		computeNode(current);
+	}
+	
+	is_computed.clear();
+	A.push(n);
+
+	fouts() << "NARROWING ITERATIONS\n";
+	fouts().flush();
+
+	// narrowing phase
+	while (!A.empty()) {
+		current = A.top();
+		A.pop();
+		narrowNode(current);
 	}
 }
 
@@ -384,7 +400,6 @@ void AI::computeTransform (Node * n, std::list<BasicBlock*> path, Abstract &Xtem
 void AI::computeNode(Node * n) {
 	BasicBlock * b = n->bb;
 	Abstract * Xtemp;
-	bool update = false;
 	Node * Succ;
 
 	if (is_computed.count(n) && is_computed[n]) {
@@ -404,7 +419,7 @@ void AI::computeNode(Node * n) {
 		fouts().flush();
 		LSMT->push_context();
 		// creating the SMT formula we want to check
-		SMT_expr smtexpr = LSMT->createSMTformula(*n->bb->getParent(), n->bb);
+		SMT_expr smtexpr = LSMT->createSMTformula(n->bb,false);
 		std::list<BasicBlock*> path;
 		DEBUG(
 			LSMT->man->SMT_print(smtexpr);
@@ -445,18 +460,6 @@ void AI::computeNode(Node * n) {
 			if (Succ->widening == 0) {
[…]

### 2011-04-12 src/AI.h

@@ -99,6 +99,9 @@ class AI : public ModulePass, public InstVisitor<AI> {
 
 		/// computeNode - compute and update the Abstract value of the Node n
 		void computeNode(Node * n);
+		
+		/// narrowNode - apply narrowing at node n
+		void narrowNode(Node * n);
 
 		/// computeCondition - creates the constraint arrays resulting from a
 		/// comparison instruction.

### 2011-04-12 src/Abstract.cc

@@ -70,6 +70,10 @@ bool Abstract::is_leq (Abstract *d) {
 	return false;
 }
 
+bool Abstract::is_bottom() {
+	return ap_abstract1_is_bottom(man,main);
+}
+
 /// widening - Compute the widening operation according to the Gopan & Reps
 /// approach
 void Abstract::widening(Node * n) {

### 2011-04-12 src/Abstract.h

@@ -39,6 +39,9 @@ class Abstract {
 		/// is_leq - return true iff this <= d
 		bool is_leq(Abstract * d);
 
+		/// is_bottom - return true iff the abstract value is at bottom
+		bool is_bottom();
+
 		/// widening - applies the widening operator, according to its
 		/// definition in the domain.
 		void widening(Node * n);

### 2011-04-12 src/Node.cc

@@ -28,11 +28,13 @@ Node::Node(ap_manager_t * _man, BasicBlock * _bb) {
 	man = _man;
 	env = ap_environment_alloc_empty();
 	X = new Abstract(_man,env);
+	Y = new Abstract(_man,env);
 	widening = 0;
 }
 
 Node::~Node() {
 	delete X;
+	delete Y;
 }
 
 /// computeSCC - compute the strongly connected components and the loop 

### 2011-04-12 src/Node.h

@@ -36,6 +36,8 @@ class Node {
 
 		/// X - Abstract domain 
 		Abstract * X;
+		/// Y - Abstract domain after narrowing operations
+		Abstract * Y;
 		ap_environment_t * env;
 
 		/// count the number of computation of X since the last widening

### 2011-04-12 src/SMT.cc

@@ -446,20 +446,6 @@ void SMT::computeRhoRec(Function &F,
 		bvar_exp = man->SMT_mk_or(implies);
 		rho_components.push_back(bvar_exp);
 	}
-	
-	// x' = x if we don't reach this block
-//	std::vector<SMT_expr> affect;
-//	std::set<Value*> empty;
-//	std::set<Value*>::iterator it = exist_prime.begin(), et = exist_prime.end();
-//	for (; it != et; ++it) {
-//		affect.push_back(man->SMT_mk_eq(getValueExpr(*it,empty), getValueExpr(*it,exist_prime)));
-//	}
-//	std::vector<SMT_expr> implies;
-//	implies.push_back(man->SMT_mk_expr_from_bool_var(bvar));
-//	implies.push_back(man->SMT_mk_and(affect));
-//	rho_components.push_back(man->SMT_mk_or(implies));
-	// end of x' = x
-
 }
 
 void SMT::computeRho(Function &F) {
@@ -488,7 +474,8 @@ void SMT::pop_context() {
 
 /// createSMTformula - create the smt formula that is described in the paper
 ///
-SMT_expr SMT::createSMTformula(Function &F, BasicBlock * source) {
+SMT_expr SMT::createSMTformula(BasicBlock * source, bool narrow) {
+	Function &F = *source->getParent();
 	std::vector<SMT_expr> formula;
 	formula.push_back(getRho(F));
 
@@ -513,8 +500,11 @@ SMT_expr SMT::createSMTformula(Function &F, BasicBlock * source) {
 		std::vector<SMT_expr> SuccExp;
 		SMT_var succvar = man->SMT_mk_bool_var(getNodeName(*i,false));
 		SuccExp.push_back(man->SMT_mk_expr_from_bool_var(succvar));
-
-		SuccExp.push_back(man->SMT_mk_not(AbstractToSmt(*i,Nodes[*i]->X)));
+		
+		if (narrow)
+			SuccExp.push_back(man->SMT_mk_not(AbstractToSmt(*i,Nodes[*i]->Y)));
+		else
+			SuccExp.push_back(man->SMT_mk_not(AbstractToSmt(*i,Nodes[*i]->X)));
 		
 		Or.push_back(man->SMT_mk_and(SuccExp));
 	}

### 2011-04-12 src/SMT.h

@@ -82,7 +82,7 @@ class SMT : public ModulePass, public InstVisitor<SMT> {
 		void push_context();
 		void pop_context();
 
-		SMT_expr createSMTformula(Function &F, BasicBlock * source);
+		SMT_expr createSMTformula(BasicBlock * source, bool narrow);
 		bool SMTsolve(SMT_expr expr, std::list<BasicBlock*> * path);
 
 		SMT_expr texpr1ToSmt(ap_texpr1_t texpr);

### 2011-04-18 report/report.tex
beginning of the report

@@
-\documentclass[a4paper,english,titlepage]{article}
+\documentclass[a4paper,english,titlepage,11pt]{article}
 
@@
 \usepackage[T1]{fontenc}
+\usepackage{amsmath}
+\usepackage{mathenv}
+\usepackage{tikz}
+\usepackage{listings}
+\usepackage{placeins}
 
@@
 \newtheorem{theorem}{Theorem}[section]
+\newtheorem{proposition}[theorem]{Proposition}
 
-\title{Analyse de programme par découverte de chemin}
+\def\R{\mathbb{R}}
+\def\N{\mathbb{N}}
+\def\Q{\mathbb{Q}}
+\def\P{\mathcal{P}}
+
+\definecolor{trefle}{rgb}{0,0.5,0}
+\definecolor{moka}{rgb}{0.5,0.25,0}
+\definecolor{minuit}{rgb}{0,0,0.5}
+
+\tikzstyle{arrow}=[->,line width=.05cm,draw=red!90!blue!60!black]
+
+\usetikzlibrary{snakes,arrows,shapes,backgrounds,shadows,automata,patterns}
+\usepgflibrary{snakes}
+
+\tikzstyle{state}=[circle,fill=black!25,minimum size=13pt,inner sep=0pt]
+\tikzstyle{transition}=[rectangle,thick,draw=black!75,
+  			  fill=black!20,minimum size=4mm]
+\tikzstyle{transition2}=[rectangle,thick,draw=blue!75,
+  			  fill=blue,minimum size=4mm,blue]
+\tikzstyle{PRstate}=[circle,fill=red!90,minimum size=13pt,inner sep=0pt]
+\tikzstyle{polyhedra}=[blue!25,opacity=0.5,pattern=north west lines,pattern
+color=blue]
+\tikzstyle{line}=[black,thick]
+
+\lstnewenvironment{C}
+{\lstset{language=C,
+		basicstyle=\ttfamily,
+		commentstyle=\color{moka}\textit,
+		keywordstyle=\color{minuit},
+		identifierstyle=\color{trefle},
+		showstringspaces=false}}
+{}
+
+\title{Static Analysis by Path Focusing}
 \author{Julien Henry}
 \date{2011}
-\shorttitle{Analyse de programme par découverte de chemin}
+\shorttitle{Static Analysis by Path Focusing}
 \institute{Grenoble-INP}
@@
 \webaddress{Julien.Henry@imag.fr}
-\reporttype{Rapport de M2R}
+\reporttype{M2R internship's report}
 
@@
 
-\section{Introduction}
+\section*{Introduction}
 
-\part{Interprétation abstraite}
+\part{Abstract Interpretation: state of the art}
+
+\section{Abstract Interpretation}
+
+Abstract Interpretation \cite{CC77,CousotCousot92-1} is a general method to find approximate solutions of
+fixpoint equations. This method is used for program analysis, since analyzing a
+program often comes down to solving fixpoint equations $\Phi(x) = x$.
+
+However, most of the time, the solution of this fixpoint equation must be
+computed in a complex domain: in program analysis, this could be the state space
+of the program, i.e the set of all possible states of the program. This
+computation quickly becomes too costly.
+
+\subsection{Abstraction of the domain}
+
+Abstract Interpretation method proposes to represent more efficiently the
+elements of this complex domain $C$ of concrete values, by choosing an simpler
+domain  $A$ called \emph{abstract domain}. The relation between these two
+domains is characterized by two functions, usually called 
+$\alpha\ :\ C \mapsto A$, $\gamma \ : \ A \mapsto C$, such as:
+
+$$\forall x \in C, \forall y \in A, \alpha(x) \leq_{A} y \Leftrightarrow x
+\leq_{C} \gamma(y) $$
+
+where $\leq_C$ and $\leq_A$ are the order relations on $C$ and $A$. Then, the
+fixpoint computation is done in the abstract domain, using the approximation of
+the function $\Phi$, i.e $\alpha(\Phi) = \alpha \circ \Phi \circ \gamma$, also
+noted $\Phi^\#$.
+We have the following result:
+
+\begin{proposition}
+If $C$ is a complete lattice, and if $\Phi$ is increasing from $C$ to $C$, then
+$$\alpha(lfp(\Phi)) \leq_A lfp(\alpha(\Phi))$$
+where $lfp(\Phi)$ denotes the least fixpoint of $\Phi$.
+\end{proposition}
+
+In fact, one compute the least fixpoint of the function in the abstract
+domain, and the result of the computation also gives an upper approximation of
+the fixpoint in the concrete domain.
+
+\subsection{Termination}
+
+Termination of the fixpoint computation has to be guaranteed. This termination
+depends on the properties of the abstract domain:
+this one should be finite, or of finite depth, meaning that there doesn't exist
+any sequence $(y_i)_{i \geq 0}$ of elements of the abstract domain $A$ such as
+$\forall i \geq 0,\ y_i \leq_A y_{i+1}$. Indeed, in case of a domain of infinite
+depth, the least fixpoint
+computation could run indefinitely, since the ascending sequence of elements of
+$A$ is infinite, and the fixpoint in never reached.
+
+Most of the domains that are used for program analysis doesn't satisfy this
+finiteness property. To ensure convergence in this case, another approximation
+is performed: a new operator is defined, called \emph{widening operator}, that
+extrapolates the limit of a sequence of abstract values
+\cite{CC77,CousotCousot92-4}. This \emph{widening
+operator} is usually noted $\nabla: A \times A \rightarrow A$, and satisfies
+the following properties:
+
+\begin{itemize}
+\item $\forall y_1, y_2 \in A,\ y_1 \leq_A y_1 \nabla y_2$ and $y_2 \leq_A y_1
+\nabla y_2$. This guarantees the correctness of the result.
+\item For any increasing sequence $y_0 \leq_A y_1 \leq_A \dots$, the sequence
+defined by 
+$$\left\{ \begin{array}{rll}
+y'_0 & = &  y_0 \\
+y'_{i+1} & = &  y'_i\ \nabla\ y_{i+1},\ \  \forall i > 0
+\end{array} \right.$$
+is not strictly increasing. Then, applying the widening operator when the
+sequence may increase indefinitely makes the computation converge to a fixpoint
+in finite time. 
+\end{itemize}
+
+The least fixpoint of a function $\Phi^\#$ is noted $\overline{y}$ and is equal
+to $\displaystyle \lim_{i \geq 0}\ y_i$, where $y_0 = \perp$ (the least element
+of $A$) and $y_{i+1} = \Phi^\# (y_i)$. Instead of computing this least fixpoint,
+one compute an upper approximation of it, by computing the following ascending
+sequence:
+
+$$\left\{ \begin{array}{rll}
+y'_0 &=& \perp \\
+y'_{i+1} &=& y'_i\ \nabla\ \Phi^\#(y'_i),\ \ \forall i > 0
+\end{array} \right.$$
+
+which converges towards $\tilde{y}$, where $\tilde{y} \geq_A \overline{y}$.
+
+$\tilde{y}$ is a correct upper approximation of the least fixpoint of $\Phi^\#$,
+and the classical technique is to regain some precision lost by the widening
+operator by computing a descending sequence:
+
+$$\left\{ \begin{array}{rll}
+y''_0 &=& \tilde{y} \\
+y''_{i+1} &=& \Phi^\#(y''_i), \ \ \forall i > 0
+\end{array} \right.$$
+
+Each element of this descending sequence is still an upper approximation of the
+least fixpoint $\overline{y}$. So, this sequence often allows to find
+approximate least fixpoints that are more precise than the one obtained after
+the ascending sequence.
+
+\section{Linear Relation Analysis}
+
+Linear Relation Analysis \cite{CH78} (also noted LRA) is a direct application of
+Abstract Interpretation. It aims at computing an upper approximation of the
+reachable states of a program containing numerical variables. The set of
+possible assignments for the numerical variables is abstracted by convex
+polyhedron. This technique discovers invariant linear relations between
+the numerical variables at each control point of a program.
+
+\subsection{Convex polyhedra}
+
+Let $x_1, x_2, \dots ,x_n$ be the numerical variables of a program (assume they
+all are in $\Q$). A state of
+the program is then a point $\overrightarrow{x} \in \Q^n$.
+
+In $\Q^n$, the set of polyhedra is a lattice. The least element is $\perp$
+(the empty polyhedron), and the greatest element is $\top$ (the whole $Q^n$).
+
+The classical union of two convex polyhedra may not be a convex polyhedra.
+Therefore, we use the convex hull operation, noted $\sqcup$, for joining
+polyhedra. 
+If $X_1, X_2$ are two convex polyhedra, $X_1 \sqcup X_2$ is the
+smallest convex polyhedron containing both $X_1$ and $X_2$.
+
+Since the domain of convex polyhedra is of infinite depth, a widening operator
+is defined to ensure convergence of the technique. There have been different
+work proposing a definition of the widening operator in LRA
+\cite{CH78,Hal79,HPR97,BlanchetCousotEtAl_PLDI03}, in order to fight
+against the loss of precision induced by widening.
+
+\subsection{Program Analysis}
+
+Throughout the report, a program will be represented by a control flow graph:
+\begin{itemize}
+\item $P$ is a finite set of control points.
+\item $I_p$ is the (possibly empty)
+set of initial values for each control point $p \in P$.
+\item $E \subseteq  P \times P$ is a set of directed edges. Each edge $e \in E$
+has a semantic $\tau_e: \P(\Q^n) \rightarrow \P(\Q^n)$, $\P(\Q^n)$ being the set
+of possible values of $\overrightarrow{x}$. $\tau_e$ maps a set of states before
+the transition to the set of states after the transition.
+\end{itemize}
+
+\begin{figure}[!h]
+   \begin{minipage}[c]{.46\linewidth}
+\begin{C}
+x = 0;
+y = 0;
+while (true) {
+	if (x <= 50) 
+		y++;
+	else 
+		y--;
+	if (y < 0) break;
+	x++;
+}
+\end{C}
+   \end{minipage} \hfill
+   \begin{minipage}[c]{.46\linewidth}
+\begin{tikzpicture}[->,>=stealth',auto,node distance=1.5cm,
+                    semithick,font=\footnotesize]
+
+	\node[state] (n0) {$p_0$};
+	\node[state] (n1) [below of=n0] {$p_1$};
+	\node[state] (n2) [below left of=n1] {$p_2$};
+	\node[state] (n3) [below right of=n1] {$p_3$};
+	\node[state] (n4) [below left of=n3] {$p_4$};
+	\node[state] (n5) [below of=n4] {$p_5$};
+	\node[state] (n6) [below of=n5] {$p_6$};
+	\node[state] (n7) [left of=n6] {$p_7$};
+
+	\node (n8) [right of=n6] {};
+	\node (n9) [right of=n1] {};
+
+  \path [transition] 
+		(n0) edge              node {$x,y \leftarrow 0$} (n1);
+  \path [transition] 
+        (n1) edge			   node [left] {$x \leq 50$} (n2);
+  \path [transition] 
+        (n1)  edge              node [right] {$x \geq 51$} (n3);
+  \path [transition] 
+        (n2) edge              node [left] {$y \leftarrow y+1$} (n4);
+  \path [transition] 
+        (n3) edge			   node [right] {$y \leftarrow y-1$} (n4);
+  \path [transition] 
+        (n4) edge			   node {$y \geq 0$} (n5);
+  \path [transition] 
+		(n4) edge  [out = 180, in=90] node [left] {$y \leq -1$} (n7);
+  \path [transition] 
+        (n5) edge              node {$x \leftarrow x+1$} (n6);
+  \path [transition] 
+        (n6) edge [out=0, in=0, distance=3.5cm] node {} (n1);
+
+\end{tikzpicture}
+   \end{minipage}
+   \caption{Example of program, and its associate control flow graph. This
+   program comes from \cite{GopanR06}.}
+\end{figure}
+\FloatBarrier
+
+To each state $p \in P$ of the control flow graph, we associate an abstract
+value $X_p \in D$, $D$ being in our case the domain of convex polyhedra over
+$\Q^n$. Since the exact operation $\tau$ may not be expressed in the abstract
+domain, we abstract it into $\tau^\#$ such as $\forall X \in D, \tau(X)
+\subseteq \tau^\#(X)$.
+
+The computation aims to find a solution for this system of abstract semantic
+inequalities:
+
+$$\left\{ \begin{array}{l}
+\forall p \in P, \ \ I_p \subseteq X_p \\
+\forall (p',p) \in E,\ \  \tau^\#_{(p',p)}(X_{p'}) \subseteq X_p
+\end{array} \right.$$
+
+The function $\Phi^\#$ previously described is the following:
+
+$$\Phi^\#\left[ \begin{array}{c}
+\vdots \\
+X_p \\
+\vdots
+\end{array} \right] = \left[ \begin{array}{c} 
+\vdots \\
+I_p \sqcup \displaystyle \bigsqcup_{(p',p) \in E} \tau^\# (X_{p'}) \\
+\vdots
+\end{array} \right]$$
+
+The fixpoint computation algorithm consist in replacing iteratively each $X_p$ by its value
+on the right hand side, until the convergence.
+
+In the case of abstract domain with infinite ascending sequence, we use the
+previously defined widening operator:
+
+$$\left[ \begin{array}{c}
+\vdots \\
+X_p \\
+\vdots
+\end{array} \right] \longleftarrow \left[ \begin{array}{c} 
+\vdots \\
+X_p \nabla \left( I_p \sqcup \displaystyle \bigsqcup_{(p',p) \in E} \tau^\#
+(X_{p'}) \right) \\
+\vdots
+\end{array} \right]$$
+
+After a finite number of steps, the computation has reached the fixpoint and the
+value $X_p$ at the end give the different linear relations between the
+different numerical variables of the program at the control point $p$.
+
+\subsection{Precision of the analysis}
+
+In order to guarantee the termination of the computation, there is no need to
+apply the widening operator at each control point $p \in P$. It is sufficient to
+apply it on a subset of $P$, called $P_w$, such as removing the nodes of $P_w$
+disconnect every cycles. For instance, one could choose $P_w$ as the set of the
+heads of loops.
+
+There exist several iteration strategies for computing the fixpoint: whatever
+the order we choose for updating the different $X_p, p\in P$, the result at the
+end of the analysis is correct. Yet, the precision of the result and the time
+before convergence could be very different. 
+There have been some work to find efficient iteration strategies
+\cite{Bou92}, such as first stabilizing innermost loops, or stabilizing strongly
+connected components of the graph.
+
+\part{Path Focusing technique}
 
@@
 	expliqués dans la thèse de B. Boissinot\cite{Boi10}
-
-	Interpretation abstraite
-	\cite{CC77}
-	\cite{CH78}
-
-	\cite{GopanR06}
 

### 2011-04-20 report/report.tex
report in progress

@@
 
-\usepackage{mathcomp,amssymb,mathptm,verimagstudent}
+\usepackage{mathcomp,amssymb,verimagstudent}
 \usepackage[utf8]{inputenc}
@@
 
+\newcommand*\system[1]{\left[ \begin{array}{lllll}#1 \end{array}\right.}
+
 \def\R{\mathbb{R}}
+\def\Z{\mathbb{Z}}
 \def\N{\mathbb{N}}
@@
 \tikzstyle{state}=[circle,fill=black!25,minimum size=13pt,inner sep=0pt]
-\tikzstyle{transition}=[rectangle,thick,draw=black!75,
-  			  fill=black!20,minimum size=4mm]
-\tikzstyle{transition2}=[rectangle,thick,draw=blue!75,
-  			  fill=blue,minimum size=4mm,blue]
-\tikzstyle{PRstate}=[circle,fill=red!90,minimum size=13pt,inner sep=0pt]
+\tikzstyle{transition}=[rectangle,semithick,draw=black!75,
+  			  minimum size=4mm]
+\tikzstyle{transition2}=[transition,rectangle,thick,dashed,
+  			  minimum size=4mm]
+\tikzstyle{PRstate}=[circle,double,draw,fill=blue!15,minimum size=13pt,inner sep=0pt]
 \tikzstyle{polyhedra}=[blue!25,opacity=0.5,pattern=north west lines,pattern
@@
 \tikzstyle{line}=[black,thick]
+
+\lstnewenvironment{LLVM}
+{\lstset{language=C,
+		basicstyle=\ttfamily\footnotesize,
+		commentstyle=\color{moka}\textit,
+		keywordstyle=\color{minuit},
+		identifierstyle=\color{trefle},
+		showstringspaces=false}}
+{}
 
@@
 defined by 
-$$\left\{ \begin{array}{rll}
+$$\system{
 y'_0 & = &  y_0 \\
 y'_{i+1} & = &  y'_i\ \nabla\ y_{i+1},\ \  \forall i > 0
-\end{array} \right.$$
+}$$
 is not strictly increasing. Then, applying the widening operator when the
@@
 
-$$\left\{ \begin{array}{rll}
+$$\system{
 y'_0 &=& \perp \\
 y'_{i+1} &=& y'_i\ \nabla\ \Phi^\#(y'_i),\ \ \forall i > 0
-\end{array} \right.$$
+}$$
 
@@
 
-$$\left\{ \begin{array}{rll}
+$$\system{
 y''_0 &=& \tilde{y} \\
 y''_{i+1} &=& \Phi^\#(y''_i), \ \ \forall i > 0
-\end{array} \right.$$
+}$$
 
@@
 the numerical variables at each control point of a program.
+
+This technique is:
+\begin{itemize}
+\item \emph{sound}: Each possible assignment for the variables in the real
+program is included in the abstract value.
+\item \emph{incomplete}: The abstract value also contains assignments for the
+variables that are not possible in the real program.
+\end{itemize}
 
@@
 \cite{CH78,Hal79,HPR97,BlanchetCousotEtAl_PLDI03}, in order to fight
-against the loss of precision induced by widening.
+against the loss of precision induced by widening:
 
@@
 
-$$\left\{ \begin{array}{l}
+$$\system{
 \forall p \in P, \ \ I_p \subseteq X_p \\
 \forall (p',p) \in E,\ \  \tau^\#_{(p',p)}(X_{p'}) \subseteq X_p
-\end{array} \right.$$
+}$$
 
@@
 \subsection{Precision of the analysis}
-
 In order to guarantee the termination of the computation, there is no need to
@@
 
+\subsubsection{Iteration strategies}
 There exist several iteration strategies for computing the fixpoint: whatever
@@
 
+\subsubsection{Sources of imprecision}
+ \label{multigraph}
+Linear Relation Analysis is sound but incomplete: there is loss of precision
+during the computation, due to different reasons:
+\begin{enumerate}
+\item Since an abstract domain is used to represent the set of possible states,
+such as the domain of convex polyhedra, there is an upper approximation of the
+set of possible states to the smallest convex polyhedra including this set. This
+loss of precision is unavoidable. Still, there are techniques proposing to
+compute disjunctive invariants \cite{GulwaniZ10} to limit this upper approximation: at each
+program point, one compute a union of convex polyhedra instead of a single
+one.
+\item The widening operator, used to ensure convergence of the technique,
+also induces a loss of precision. The classical technique is then to compute
+narrowing iterations, which often recover some precision. 
+\item There is also a loss of precision each time a point in the control flow
+graph has several incoming edges. This is the case for instance for
+\emph{if} statements, loops, etc. In this case, the abstract value of this point
+is a convex hull of several convex polyhedra, inducing an upper approximation.
+Indeed, if $X_1, X_2$ are convex polyhedra,
+$$X_1 \cup X_2 \subseteq X_1 \sqcup X_2$$
+
+To limit the number of such points, a method could be
+to expand the control flow graph. For instance, instead of considering a graph
+with a sequence of $n$ \emph{if-then-else} between point $p_1$ and $p_n$, one could remove the $n$ merge node
+of these \emph{if-then-else}, each of them having two incoming edges, and adding 
+$2^n$ edges from $p_1$ to $p_n$, corresponding to the different paths through
+the \emph{if-then-else} statements. At the end, $p_n$ is the only point with
+several incoming edges. This graph transformation results in an exponential
+blowup, because of the exponential growth of the number of edges in the graph.
+\end{enumerate}
+
+\subsubsection{Acceleration}
+
+Typically, to analyze a program with loops, the classical approach is to use the
+widening operator, for example at the head of the loops, to ensure convergence.
+Since this induces imprecision, another technique called \emph{acceleration}
+aims to compute the exact effect of the loop when possible \cite{Gon07,GH06}. For instance, for
+the following program, one could directly compute an invariant for the loop:
+
+\begin{minipage}[c]{.39\linewidth}
+\begin{C}
+x = 0;
+y = 0;
+while (x < N) {
+	y = y+2;
+	x = x+5;
+}
+\end{C}
+\end{minipage} 
+\begin{minipage}[c]{.59\linewidth}
+$$\begin{array}{llll}
+\exists k \geq 0,& & x  = k \\
+		&\wedge & y = 2k \\
+		&\wedge & \forall k', (0 \leq k' \leq k)
+				\Rightarrow (5k' \leq N)
+\end{array}$$
+\end{minipage}
+
+Acceleration is possible when the loop is simple enough, and come to compute the
+transitive closure $\tau_e^+$ of $\tau_e$, the transition function associated to
+the loop.
+
 \part{Path Focusing technique}
 
-	Pour calculer les variables live en SSA, des algorithmes plus optimisés sont
-	expliqués dans la thèse de B. Boissinot\cite{Boi10}
+	In this part, we present two techniques that aim to isolate the different
+	paths of the program to compute more precise invariants.
 
-	Utilisation de la bibliothèque APRON \cite{JM09}
+ \section{Lookahead Widening}
 
-\part{Contribution}
+ \section{Path Focusing}
 
-  \section{Conclusion}
+	\subsection{Multigraph}
+
+	\subsubsection{Expanding the control flow graph}
+	The main idea of the technique is to compute the fixpoint iterations on an
+	expanded multigraph instead of the classical control flow graph. A
+	multigraph is a graph that can have several edges from a point $p$ to a
+	point $q$.
+
+	Let $(P,E)$ be the control flow graph of the program.
+	First, one choose a set of points $P_W \subseteq P$, such that the graph
+	obtained after removing these points has no cycle. The widening operator
+	will be applied at these points. The choice of $P_W$ can be guided by
+	\cite{Bou92}.
+	Then, one choose another set $P_R \subseteq P$ of nodes, satisfying the
+	following properties:
+	\begin{itemize}
+	\item $I_p = \emptyset$ for each $p \in P \ P_R$, meaning that the initial
+	points of the program are included in $P_R$.
+	\item $P_W \subseteq P_R$: all the widening points are in $P_R$.
+	\end{itemize}
+
+	The multigraph is then the graph such as:
+	\begin{itemize}
+	\item each element of $P_R$ is a point of the graph.
+	\item for each path $p_1 \rightarrow p_2 \rightarrow \dots \rightarrow p_k$
+	in the control flow graph, such as $p_1 \in P_R, p_k \in P_R$,
+	there is a transition in the multigraph from $p_1$ to $p_k$ with the
+	semantic $$\tau_{p_1 \rightarrow \dots \rightarrow p_k} = \tau_{p_{k-1}
+	\rightarrow p_k} \circ
+	\dots \circ \tau_{p_1 \rightarrow p_2}$$ where $\tau_{p_i \rightarrow
+	p_{i+1}}$ is the transformation associated to the semantic of $p_i
+	\rightarrow p_{i+1}$.
+	\end{itemize}
+
+	As explained in \ref{multigraph}, separating the different
+	possible paths between $p$ and $q$ reduce the number of points with a join
+	operation, and results in a better precision.
+	This can result in an exponential blowup in the size of the graph. To avoid
+	it, the technique is to never construct the multigraph, and only compute
+	parts of it when needed.
+
+	Intuitively, choosing a small set $P_R$ will result in a better precision,
+	but a higher cost. 
+	Indeed, the number of paths between two points of $P_R$ will grow if $P_R$
+	contains few elements.
+	A good choice for $P_R$ requires to find a good
+	compromise between precision and cost of the technique.
+
+\begin{figure}[!h]
+\label{CFG}
+\centering
+\begin{minipage}[c]{.39\linewidth}
+\begin{tikzpicture}[->,>=stealth',auto,node distance=1.5cm,
+                    semithick,font=\footnotesize]
+
+	\node[PRstate] (n0) {$p_0$};
+	\node[PRstate] (n1) [below of=n0] {$p_1$};
+	\node[state] (n2) [below left of=n1] {$p_2$};
+	\node[state] (n3) [below right of=n1] {$p_3$};
+	\node[state] (n4) [below left of=n3] {$p_4$};
+	\node[state] (n5) [below of=n4] {$p_5$};
+	\node[state] (n6) [below of=n5] {$p_6$};
+	\node[PRstate] (n7) [left of=n6] {$p_7$};
+
+  \path [transition] 
+		(n0) edge              node {$x,y \leftarrow 0$} (n1);
+  \path [transition2] 
+        (n1) edge			   node [left] {$x \leq 50$} (n2);
+  \path [transition] 
+        (n1)  edge              node [right] {$x \geq 51$} (n3);
+  \path [transition2] 
+        (n2) edge              node [left] {$y \leftarrow y+1$} (n4);
+  \path [transition] 
+        (n3) edge			   node [right] {$y \leftarrow y-1$} (n4);
+  \path [transition2] 
+        (n4) edge			   node {$y \geq 0$} (n5);
+  \path [transition] 
+		(n4) edge  [out = 180, in=90] node [left] {$y \leq -1$} (n7);
+  \path [transition2] 
+        (n5) edge              node {$x \leftarrow x+1$} (n6);
+  \path [transition2] 
+        (n6) edge [out=0, in=0, distance=3.5cm] node {} (n1);
+\end{tikzpicture}
+\end{minipage} 
+$\Longrightarrow$
+\begin{minipage}[c]{.49\linewidth}
+\begin{tikzpicture}[->,>=stealth',auto,node distance=3.2cm,
+                    semithick,font=\footnotesize]
+
+	\node[PRstate] (n0) {$p_0$};
+	\node[PRstate] (n1) [below of=n0,yshift=1cm] {$p_1$};
+	\node[PRstate] (n7) [below of=n1] {$p_7$};
+
+	\node (n8) at (-3,0) {};
+	\node (n8) at (3,0) {};
+
+  \path [transition] 
+		(n0) edge              node [yshift=5mm] {$x,y \leftarrow 0$} (n1);
+  \path [transition] 
+		(n1) edge    [bend left]  node [right,yshift=-0.8cm]  {$
+		\begin{array}{l}
+			x \geq 51 \\
+			y \leq 0 \\
+			y \leftarrow y-1
+		\end{array}
+		$} (n7);
+  \path [transition2] 
+		(n1) edge    [loop left,distance=1.2cm]  node [left,xshift=3mm] {$
+		\begin{array}{l}
+			x \leq 50 \\
+			y \geq -1 \\
+			y \leftarrow y+1 \\
+			x \leftarrow x+1 
+		\end{array}
+		$} (n1);
+  \path [transition] 
+		(n1) edge    [loop right,distance=1.2cm]  node [right,xshift=-2mm] {$
+		\begin{array}{l}
+			x \geq 51 \\
+			y \geq 1 \\
+			y \leftarrow y-1 \\
+			x \leftarrow x+1 
+		\end{array}
+		$} (n1);
+  \path [transition] 
+		(n1) edge    [bend right]  node [left,yshift=-0.8cm,xshift=4mm] {$
+		\begin{array}{l}
+			x \leq 50 \\
+			y \leq -2 \\
+			y \leftarrow y-1
+		\end{array}
+		$} (n7);
+\end{tikzpicture}
+\end{minipage}
+\caption{Example of classical control flow graph, and its associate expanded
+multigraph, where $P_R = \{p_0, p_1, p_7\}$. The path dashed from $p_1$ to $p_1$
+in the control flow graph corresponds to a simple dashed transition in the multigraph,
+having the same semantic.}
+\end{figure}
+\FloatBarrier
+
+%\begin{tikzpicture}[->,>=stealth',auto,node distance=1.5cm,
+%                    semithick,font=\footnotesize]
+%
+%	\node[state] (n0) {$p_0$};
+%	\node[state] (n1) [below of=n0] {$p_1$};
+%	\node[state] (n2) [below left of=n1] {$p_2$};
+%	\node[state] (n3) [below right of=n1] {$p_3$};
+%	\node[state] (n4) [below left of=n3] {$p_4$};
+%	\node[state] (n5) [below of=n4] {$p_5$};
+%	\node[state] (n6) [below of=n5] {$p_6$};
+%	\node[state] (n7) [left of=n6] {$p_7$};
+%
+%	\node (n8) [right of=n6] {};
+%	\node (n9) [right of=n1] {};
+%	\node (n88) [above right of=n8] {};
+%	\node (n99) [below right of=n9] {};
+%
+%	\path[transition] plot[->,smooth,tension=0.7] coordinates {
+%		(n1)
+%		(n2)
+%		(-0.4,-3.8)
+%		(n7)
+%	};
+%	\path[transition] plot[->,smooth,tension=0.7] coordinates {
+%		(n1)
+%		(0.7,-2.6)
+%		(0,-3.8)
+%		(n7)
+%	};
+%	\path[transition] plot[->,smooth,tension=0.7] coordinates {
+%		(n1)
+%		(-0.8,-2.4)
+%		(-0.1,-3.8)
+%		(0,-6.3)
+%		(2.5,-5.5)
+%		(2.5,-2.5)
+%		(n1)
+%	};
+%	\path[transition] plot[->,smooth,tension=0.7] coordinates {
+%		(n1)
+%		(n3)
+%		(0.3,-3.8)
+%		(0.3,-6)
+%		(2.1,-5.5)
+%		(2.1,-2.5)
+%		(n1)
+%	};
+%
+%  \path [transition] 
+%		(n0) edge              node {$x,y \leftarrow 0$} (n1);
+%
+%	\node[state] (n0) {$p_0$};
+%	\node[state] (n1) [below of=n0] {$p_1$};
+%	\node[state] (n2) [below left of=n1] {$p_2$};
+%	\node[state] (n3) [below right of=n1] {$p_3$};
+%	\node[state] (n4) [below left of=n3] {$p_4$};
+%	\node[state] (n5) [below of=n4] {$p_5$};
+%	\node[state] (n6) [below of=n5] {$p_6$};
+%	\node[state] (n7) [left of=n6] {$p_7$};
+%
+%\end{tikzpicture}
+
+	\subsubsection{Static Single Assignment form}
+
+	Path focusing technique uses control flow graphs in static single assignment
+	(SSA) form. The main concept behind the SSA form is that every variable is
+	only being assigned once.
+
+\begin{figure}[!h]
+\centering
+\begin{minipage}[c]{.39\linewidth}
+\begin{C}
+x = 0;
+y = 0;
+x = x + 2;
+y = x + 1;
+\end{C}
+\end{minipage} 
+$\Longrightarrow$ \hfill
+\begin{minipage}[c]{.49\linewidth}
+\begin{C}
+x.0 = 0;
+y.0 = 0;
+x.1 = x.0 + 2;
+y.1 = x.1 + 1;
+\end{C}
+\end{minipage}
+\caption{On the left side, a simple program. On the right side, the same
+program in SSA form.}
+\end{figure}
+
+This single definition property cannot be achieved with only the renaming of
+variables assigned most than once. In some cases, two distinct definition reach
+the same use, depending on the actual execution. To solve this issue, SSA form
+uses $\Phi$-functions.
+
+A $\Phi$-function is usually inserted at points of the control flow graph that
+have several incoming transitions, and has the same number of arguments as the point
+has incoming transitions. If we suppose these transitions are ordered, then the
+$\Phi$-function returns the value of its $i$-th argument if the point is reached
+from the $i$-th transition.
+
+\begin{figure}[!h]
+\centering
+\begin{minipage}[c]{.39\linewidth}
+\begin{C}
+if (c) x = 0;
+else x = 1;
+
+y = x+1;
+\end{C}
+\end{minipage} 
+$\Longrightarrow$ \hfill
+\begin{minipage}[c]{.49\linewidth}
+\begin{C}
+if (c) x.0 = 0;
+else x.1 = 0;
+
+x.2 = Phi (x.0, x.1)
+y.0 = x.2 + 1;
+\end{C}
+\end{minipage}
+\end{figure}
+\FloatBarrier
+
+This static single assignment form is required for the computation of the focus
+path.
+
+	\subsection{Choice of the focus paths}
+
+	\subsubsection{Reachability problem}
+	Path focusing technique aims to discover which paths to focus on for the
+	computation of the fixpoint iterations. 
+	Since the multigraph is not constructed explicitly, the different paths
+	starting and finishing on node in $P_R$ are not known. 
+	The technique proposes to express the focus path as the solution of a
+	satisfiability modulo theory (SMT) problem:
+
+	\begin{center} \emph{
+		``Is there a path starting on control point $p$, with numerical
+		variables in $X_p$, that ends on point $q$, with variables that are not
+		in $X_q$ ?''}
+	\end{center}
+
+	This problem could be rephrased as:
+	``Is there a path that starts on $p$, that can make the fixpoint computation
+	progress ?''
+	Indeed, the idea is to only focus on paths that make the abstract values
+	grow.
+
+	This is a reachability problem, that can be expressed as an SMT-formula,
+	which is a logic formula with boolean variables, and elements of a certain
+	theory $T$. Depending on the program, one can choose different theories:
+	\begin{itemize}
+	\item For programs with rational variables, whose operations (instructions,
+	transition conditions\dots) are all linear arithmetic, $T$ can be the theory
+	of linear real arithmetic (LRA).
+	\item For programs with integer variables (with a numerical state space in
+	$\Z^n$) $T$ can be the theory of linear integer arithmetic (LIA).
+	\end{itemize}
+
+	Despite deciding the satisfiability of such formula is NP-complete, there
+	has been lots of research on decision procedures \cite{Kroening08} and there exist nowadays
+	efficient programs, known as SMT-solvers, that can decide the satisfiability
+	of LRA or LIA formulas. Well-known efficient SMT-solvers are
+	Z3 \cite{MouraB08} and Yices \cite{DutertreM06}.
+
+	If the problem has a solution,  the SMT-solver will be in position to show
+	which path in the control flow graph is selected, by giving a model: an
+	assignments of the boolean variables such as the formula is true.
+	Among these boolean variables, some of them are associated to a transition
+	in the control flow graph:
+	\begin{itemize}
+	\item if the model has set to \emph{true} the boolean variable $b_e$ associated
+	to the transition $e$, then the focus path goes through this transition.
+	\item in the other case, the focus path doesn't go through this transition.
+	\end{itemize}
+
+	Some others are associated to the points of the control flow graph. These
+	are called \emph{reachability predicates}, and are set to \emph{true} in the
+	model if and only if the focus path goes through these points.
+
+	\subsubsection{Construction of the formula}
+	The first point of the Path Focusing technique is to compute the SMT formula
+	$\rho$ expressing the semantic of the program. For doing so, all the
+	operations in the program have to be expressible within the theory we
+	choose. Otherwise, the SMT-solver would not be able to decide if the formula
+	is \emph{sat} or not.
+
+	Since we want to find a path starting in a control point in $P_R$ and ending
+	in a point in $P_R$, we need to disconnect in the formula the points $p_i$
+	in $P_R$ into a source point $p_i^s$, with only outgoing transitions, and a
+	destination point $p_i^d$, with only incoming transitions.
+
+	We construct the $\rho$ formula as follow:
+
+\begin{enumerate}
+\item Each numerical SSA variable of the control flow graph has its definition
+in the $\rho$ formula.
+\item For each operation in the program, we encode its semantic, expressed in
+terms of the boolean and numerical SSA variables. For instance, if the CFG
+contains the affectation $x.2 = x.1 + 1$, then we simply add $x.2 = x.1 + 1$ in
+the formula.
+\item For each transition $(p_i,p_j) \in E$, we set $t_{i,j} = c(i,j)$, where
+$c(i,j)$ express the condition that needs to be true to go through the
+transition. For instance, $c(i,j)$ could be of the form $x < y$, \dots
+If the transition is undeterministic, then $c(i,j) = c_{i,j}$, where
+$c_{i,j}$ is a boolean variable left undeterministic.
+\item For each point $p_i \notin P_R$, we set the reachability predicate 
+$b_i = \displaystyle \bigvee_{(p_j,p_i)\in E} t_{i,j}$.
+\item For each point $p_i^d \in P_R$, we set the reachability predicate 
+$b_i^d = \displaystyle \bigvee_{(p_j,p_i)\in E} t_{i,j}$. 
+\item $\Phi$-variables are treated differently, depending on the control point
+$p$ where they are defined. To be completed !!
+
+\end{enumerate}
+
+\begin{figure}[!h]
+\begin{minipage}[c]{.49\linewidth}
+\begin{tikzpicture}[->,>=stealth',auto,node distance=1.5cm,
+                    semithick,font=\footnotesize]
+
+\tikzstyle{s}=[rectangle,fill=black!25,minimum size=13pt,inner sep=0pt]
+\tikzstyle{t}=[rectangle,semithick,draw=black!75,
+  			  minimum size=4mm]
+\tikzstyle{PRs}=[rectangle,draw,fill=blue!15,minimum size=13pt,inner sep=0pt]
+\tikzstyle{polyhedra}=[blue!25,opacity=0.5,pattern=north west lines,pattern
+color=blue]
+\tikzstyle{line}=[black,thick]
+
+	\node[PRs] (n0) {$p_0$};
+	\node[PRs] (n1) [below of=n0] {$\begin{array}{l}p_1 \\ x.1 =
+	\Phi(x.0,x.1) \\ y.1 = \Phi(y.0,y.4)\end{array}$};
+	\node[s] (n2) [below left of=n1,yshift=-0.5cm] {$p_2$};
+	\node[s] (n3) [below right of=n1,yshift=-0.5cm] {$p_3$};
+	\node[s] (n4) [below left of=n3,yshift=-0.2cm] {$\begin{array}{l}p_4\\ y.4 =
+	\Phi(y.2,y.3)\end{array}$};
+	\node[s] (n5) [below of=n4] {$p_5$};
+	\node[s] (n6) [below of=n5] {$p_6$};
+	\node[PRs] (n7) [left of=n6] {$p_7$};
+
+  \path [transition] 
+		(n0) edge              node {$\begin{array}{l} x.0 = 0 \\ y.0 =
+		0\end{array}$} (n1);
+  \path [transition] 
+        (n1) edge			   node [left] {$x.1 \leq 50$} (n2);
+  \path [transition] 
+        (n1)  edge              node [right] {$x.1 \geq 51$} (n3);
+  \path [transition] 
+        (n2) edge              node [left] {$y.2 = y.1+1$} (n4);
+  \path [transition] 
+        (n3) edge			   node [right] {$y.3 = y.1-1$} (n4);
+  \path [transition] 
+        (n4) edge			   node {$y.4 \geq 0$} (n5);
+  \path [transition] 
+		(n4) edge  [out = 180, in=90] node [left] {$y.4 \leq -1$} (n7);
+  \path [transition] 
+        (n5) edge              node {$x.2 \leftarrow x.1+1$} (n6);
+  \path [transition] 
+        (n6) edge [out=0, in=0, distance=3.5cm] node {} (n1);
+\end{tikzpicture}
+(a) SSA form of the control flow graph seen in figure \ref{CFG}.
+\end{minipage} 
+$\Longrightarrow$ \hfill
+\begin{minipage}[c]{.49\linewidth}
+\begin{tikzpicture}[->,>=stealth',auto,node distance=1.5cm,
+                    semithick,font=\footnotesize]
+
+\tikzstyle{s}=[rectangle,fill=black!25,minimum size=13pt,inner sep=0pt]
+\tikzstyle{t}=[rectangle,semithick,draw=black!75,
+  			  minimum size=4mm]
+\tikzstyle{PRs}=[rectangle,draw,fill=blue!15,minimum size=13pt,inner sep=0pt]
+\tikzstyle{polyhedra}=[blue!25,opacity=0.5,pattern=north west lines,pattern
+color=blue]
+\tikzstyle{line}=[black,thick]
+
+	\node[PRs] (n0s) {$p_0^s$};
+	\node[PRs] (n1s) [right of=n0s,xshift=1.3cm] {$p_1^s$};
+	\node[PRs] (n7s) [right of=n1s,xshift=1.3cm] {$p_7^s$};
+	\node[s] (n2) [below left of=n1s] {$p_2$};
+	\node[s] (n3) [below right of=n1s] {$p_3$};
+	\node[s] (n4) [below left of=n3,yshift=-0.2cm] {$\begin{array}{l}p_4\\ y.4 =
+	\Phi(y.2,y.3)\end{array}$};
+	\node[s] (n5) [below of=n4] {$p_5$};
+	\node[s] (n6) [below of=n5,yshift=0.3cm] {$p_6$};
+	\node[PRs] (n1d) [below of=n6] {$\begin{array}{l}p_1^d \\ x.1' =
+	\Phi(x.0,x.1) \\ y.1' = \Phi(y.0,y.4)\end{array}$};
+	\node[PRs] (n7) [right of=n1d,xshift=0.6cm] {$p_7^d$};
+	\node[PRs] (n0d) [left of=n1d,xshift=-0.6cm] {$p_0^d$};
+
+  \path [transition] 
+		(n0s) edge   [out=-90]           node [left] {$\begin{array}{l} x.0 = 0 \\ y.0 =
+		0\end{array}$} (n1d);
+  \path [transition] 
+        (n1s) edge			   node [left] {$x.1 \leq 50$} (n2);
+  \path [transition] 
+        (n1s)  edge              node [right] {$x.1 \geq 51$} (n3);
+  \path [transition] 
+        (n2) edge              node [left] {$y.2 = y.1+1$} (n4);
+  \path [transition] 
+        (n3) edge			   node [right] {$y.3 = y.1-1$} (n4);
+  \path [transition] 
+        (n4) edge			   node {$y.4 \geq 0$} (n5);
+  \path [transition] 
+		(n4) edge  [out = 0, in=90] node [left] {$y.4 \leq -1$} (n7);
+  \path [transition] 
+        (n5) edge              node {$x.2 \leftarrow x.1+1$} (n6);
+  \path [transition] 
+        (n6) edge				node {} (n1d);
+\end{tikzpicture}
+
+(b) control flow graph (a), such as it is represented by the SMT formula $\rho$.
+Points in $P_R$ have been split into a ``source'' point and a ``destination''
+point, and $\Phi$-variables associated to points in $P_R$ have been primed.
+\end{minipage} 
+\caption{}
+\end{figure}
+
+	\subsection{Algorithm}
+
+	\subsection {Choice of $P_R$}
+
+	\subsection{Disjunctive invariants}
+
+\part{Implementation}
+
+\emph{Lookahead Widening} and \emph{Path Focusing}
+techniques have been implemented into a small analyzer, so that we can compare
+the precision of their results. In this part, we explain some details about
+these implementations.
+
+ \section{Infrastructure}
+The analyzer is based on the \emph{Low Level Virtual Machine} (LLVM)
+\cite{LLVM:CGO04}, which is a compilation infrastructure used for instance by
+the \emph{clang} compiler.
+
+% We need layers to draw the block diagram
+\pgfdeclarelayer{background}
+\pgfdeclarelayer{foreground}
+\pgfsetlayers{background,main,foreground}
+
+% Define a few styles and constants
+\tikzstyle{sensor}=[draw, fill=blue!20, text width=5em, 
+    text centered, minimum height=2.5em]
+\tikzstyle{ann} = [above, text width=5em]
+\tikzstyle{IR} = [sensor, text width=6em, fill=red!20, 
+    minimum height=12em, rounded corners]
+\def\leftdist{2.3}
+\def\rightdist{5.0}
+\def\edgedist{2.5}
+
+\begin{figure}[!h]
+\centering
+\begin{tikzpicture}
+    \node (IR) [IR] {Internal Representation\\ (IR)};
+    % Note the use of \path instead of \node at ... below. 
+    \path (IR.130)+(-\leftdist,0) node (c) [sensor] {C};
+    %\path (IR.-150)+(-\leftdist,0) node (cpp) [sensor] {C++};
+	\node [sensor] (cpp) [below of=c, yshift=-0.3cm] {C++};
+	\node  (lang) [below of=cpp] {$\vdots$};
+
+    \path (IR.130)+(+\rightdist,0) node (x8664) [sensor] {Binary\\ x86\_64};
+	\node [sensor] (i686) [below of=x8664, yshift=-0.3cm] {Binary\\ i686};
+	\node  (Out) [below of=i686] {$\vdots$};
+
+	\path [draw, ->] (cpp) -- (IR);
+	\path [draw, ->] (c) -- (IR);
+	\path [draw, ->] (lang) -- (IR);
+
+	\path [draw, ->] (IR) -- (x8664);
+	\path [draw, ->] (IR) -- (i686);
+	\path [draw, ->] (IR) -- (Out);
+
+	% We could simply have written (gyros) .. (naveq.140). However, it's
+    % best to avoid hard coding coordinates
+    \node (Input) [below of=lang] {Input};
+    \node (Output) [below of=Out] {Output};
+    \path (IR.south west)+(-0.6,-0.4) node (Frontend) {Frontend};
+    \path (IR.south east)+(+0.6,-0.4) node (Backend) {Backend};
+
+    % Now it's time to draw the colored IMU and INS rectangles.
+    % To draw them behind the blocks we use pgf layers. This way we  
+    % can use the above block coordinates to place the backgrounds   
+    \begin{pgfonlayer}{background}
+        % Compute a few helper coordinates
+        \path (c.west |- IR.north)+(-0.5,0.3) node (a) {};
+        \path (Frontend.south -| x8664.east)+(+0.5,-0.2) node (b) {};
+        \path[fill=yellow!20,rounded corners, draw=black!50, dashed]
+            (a) rectangle (b);
+        \path (c.north west)+(-0.2,0.2) node (a) {};
+        \path (Input.south -| c.east)+(+0.2,-0.2) node (b) {};
+        \path[fill=blue!10,rounded corners, draw=black!50, dashed]
+            (a) rectangle (b);
+        \path (x8664.north west)+(-0.2,0.2) node (a) {};
+        \path (Output.south -| x8664.east)+(+0.2,-0.2) node (b) {};
+        \path[fill=blue!10,rounded corners, draw=black!50, dashed]
+            (a) rectangle (b);
+    \end{pgfonlayer}
+\end{tikzpicture}
+\end{figure}
+\FloatBarrier
+
+The advantage of using such an infrastructure is that the analyzer can check
+programs written in different languages: C, C++, \dots, without needing to write
+a specific frontend for each of them. Here, the analyzer works directly on the
+internal representation (IR), which is a graph representation of
+the program. The control flow graph of the program can easily be extracted from
+this representation.
+
+\begin{figure}[!h]
+\begin{minipage}[c]{.46\linewidth}
+\begin{C}
+int f(int n) {
+	int x;
+	int y = n;
+	if (n > 0) {
+		x = 50;
+		y++;
+	} else {
+		x = 10;
+		y--;
+	}
+	return x+y;
+}
+\end{C}
+\end{minipage}
+\begin{minipage}[c]{.46\linewidth}
+\begin{LLVM}
+define i32 @f(i32 %n) nounwind {
+  %1 = icmp sgt i32 %n, 0
+  br i1 %1, label %2, label %4
+
+; <label>:2                                       ; preds = %0
+  %3 = add nsw i32 %n, 1
+  br label %6
+
+; <label>:4                                       ; preds = %0
+  %5 = add nsw i32 %n, -1
+  br label %6
+
+; <label>:6                                       ; preds = %4, %2
+  %x.0 = phi i32 [ 50, %2 ], [ 10, %4 ]
+  %y.0 = phi i32 [ %3, %2 ], [ %5, %4 ]
+  %7 = add nsw i32 %x.0, %y.0
+  ret i32 %7
+}
+\end{LLVM}
+\end{minipage}
+\caption{Example of C program and its internal representation in SSA form}
+\end{figure}
+\FloatBarrier
+
+ \section{Abstract domain representation}
+
+		The analyzer is implemented with the APRON library \cite{JM09}.
+		% live variables
+		To compute the set of live variables in SSA: \cite{Boi10}
+
+ \section{Experiments}
+
+  \section*{Conclusion}
 

### 2011-04-21 report/report.tex
fix compilation problem in the analyzer + update of the report

@@
 \def\P{\mathcal{P}}
+\def\ite{\textnormal{ite\ }}
 
@@
  \section{Lookahead Widening}
+
+To Be Completed
 
@@
 contains the affectation $x.2 = x.1 + 1$, then we simply add $x.2 = x.1 + 1$ in
-the formula.
-\item For each transition $(p_i,p_j) \in E$, we set $t_{i,j} = c(i,j)$, where
+the formula. This equality expression is noted $assign(x.2)$. 
+$assign(x)$ is defined for each numerical SSA-variable of the program.
+\item For each transition $(p_i,p_j) \in E$, we set $t_{i,j} = B_i\ \wedge \
+c(i,j)$, where $B_i = b_i^s$ if $p_i \in P_R$, or $b_i$ if not.
 $c(i,j)$ express the condition that needs to be true to go through the
@@
 \item For each point $p_i \notin P_R$, we set the reachability predicate 
-$b_i = \displaystyle \bigvee_{(p_j,p_i)\in E} t_{i,j}$.
+$b_i = \displaystyle \bigvee_{(p_j,p_i)\in E} t_{j,i}$.
 \item For each point $p_i^d \in P_R$, we set the reachability predicate 
-$b_i^d = \displaystyle \bigvee_{(p_j,p_i)\in E} t_{i,j}$. 
-\item $\Phi$-variables are treated differently, depending on the control point
-$p$ where they are defined. To be completed !!
-
+$b_i^d = \displaystyle \bigvee_{(p_j,p_i)\in E} t_{j,i}$. 
+\item Numerical $\Phi$-variables assignations have to be expressed in the $\rho$
+formula. We simply use $\ite$ (\emph{if-then-else}) statements, standardly defined in the
+SMT-Lib format \cite{BarST-SMTLIB}, for giving the right value to the variable,
+depending on the incoming transition we come from. Generally, for a
+$Phi$-variable $x = \Phi(x_1,x_2,\dots,x_n)$ in control point $p_i$ with the
+ordered incoming transitions $t_{j_1,i},t_{j_2,i},\dots,t_{j_n,i}$, we set:
+
+$$x = \ite (t_{j_1,i})\  (x_1)\  (\ite (t_{j_2,i})\  (x_2)\  (\ite\  \dots))$$
+
+where $\ite (c)\ (x)\ (y)$ is equal to $x$ if $c = true$, else $y$.
+
+If the point where the $\Phi$-variable $x$ is defined is in
+$P_R$, one need to rename $x$ into $x'$:
+\begin{itemize}
+\item $x$ will be the value of $x$ at the beginning of the path.
+\item $x'$ will be the value of $x$ at the end of the path.
+\end{itemize}
+Indeed, there can be uses of $x$ along the path, and these uses refer to the
+old value of $x$. 
 \end{enumerate}
+
+Finally, the $\rho$ formula is the following:
+
+\begin{eqnarray*}
+\rho = & \displaystyle\bigwedge_{p_i \notin P_R}& \left[ (b_i = \bigvee_{(p_h,p_i) \in E}
+t_{h,i}) \wedge ( \bigwedge_{(p_i,p_j)\in E} t_{i,j} = b_i \wedge c(i,j))
+\right] \wedge \\
+ & \displaystyle\bigwedge_{p_i \in P_R}& \left[ (b_i^d = \bigvee_{(p_h,p_i) \in E}
+ t_{h,i}) \wedge ( \bigwedge_{(p_i,p_j)\in E} t_{i,j} = b_i^s \wedge
+ c(i,j)) \right] \wedge \\
+ & \displaystyle\bigwedge_{x \in \Sigma}& assign(x)
+\end{eqnarray*}
+
+where $\Sigma$ is the set of the numerical SSA-variables of the program.
 
@@
 \end{minipage} 
-\caption{}
+\caption{Example (cont'd)}
 \end{figure}

### 2011-04-21 src/Analyzer.cc

@@ -12,12 +12,12 @@ SMT_Solver manager;
 llvm::raw_ostream *Out;
 
 void show_help() {
-        puts("Usage :\n \
+        std::cout << "Usage :\n \
 \tanalyzer -h OR analyzer [-d] [-y]  -i <filename> \n \
 -h : help\n \
 -i : input file\n \
 -y : use Yices instead of Z3 SMT-solver\
--d : debug\n");
+-d : debug\n";
 }
 
 SMT_Solver getSMTSolver() {
@@ -54,18 +54,18 @@ int main(int argc, char* argv[]) {
             manager = YICES_MANAGER;
             break;
         case '?':
-            printf("Error : Unknown option %c\n", optopt);
+            std::cout << "Error : Unknown option" << optopt << "\n";
             bad_use = true;
         }   
     } 
     if (!help) {
         if (!filename) {
-            printf("No input file specified.\n");
+            std::cout << "No input file specified\n";
             bad_use = true;
         }
 
         if (bad_use) {
-            printf("Bad use !\n");
+            std::cout << "Bad use\n";
             show_help();
             exit(EXIT_FAILURE);
         }

### 2011-04-21 src/yices.cc

@@ -1,6 +1,7 @@
 #include <cstddef>
 #include <vector>
 #include <sstream>
+#include <stdio.h>
 
 #include "llvm/Module.h"
 #include "llvm/Instructions.h"

### 2011-04-21 report/report.tex
update of the report

@@
 \usepackage{placeins}
+\usepackage{algorithm}
+\usepackage{algpseudocode}
 
@@
 	\subsubsection{Reachability problem}
+	\label{reachability}
 	Path focusing technique aims to discover which paths to focus on for the
@@
 
+	\subsubsection{Expression of the reachability problem}
+
+	$\rho$ express the semantic of the program. We still need to create the
+	formula expressing the reachability problem defined in \ref{reachability}.
+
+	The abstract domain $D$ we use is those of convex polyhedra, so an abstract
+	value $X \in D$ is simply a system of linear inequalities between the
+	different SSA-variables of the program. Then, it is easy to encode the
+	property $x \in X$ into an SMT-formula: one simply write the conjunction of
+	these inequalities, with the name of the variables in the vector $x$.
+
+	The formula given to the SMT-solver is:
+
+	$$\rho \wedge b_i^s \wedge \bigwedge_{j\neq i,\ j\in P_R} (\neg b_j^s)
+	\wedge x_i \in X_{p_i} \wedge \bigvee_{j/(p_i,p_j)\in E} (b_2^d \wedge \neg
+	(x_j \in X_{p_j}))$$
+
+	\begin{itemize}
+	\item we search a path starting in $p_i$: $b_i^s$ has to be \emph{true}, and
+	all other source points are \emph{false}.
+	\item the vector $x_i$ of the numerical variables at point $p_i$ is in our
+	abstract value $X_{p_i}$
+	\item we search a path arriving in a successor $p_j$ of $p_i$, such as the
+	vector of the numerical variables $x_j$ is not yet included in the current
+	abstract value $X_{p_j}$ of $p_j$. $b_j^d$ is \emph{true}, and the
+	conjunction of the inequalities given by $X_{p_j}$ is \emph{false}.
+	\end{itemize}
+
 	\subsection{Algorithm}
 
-	\subsection {Choice of $P_R$}
+	Path Focusing algorithm is described in Algorithm \ref{pathfocusingalgo}.
+\begin{algorithm}
+\caption{Path Focusing}\label{pathfocusingalgo}
+\begin{algorithmic}[1] 
+\Procedure{pathfocusing}{$P$,$E$,$P_R$}
+\ForAll {$p \in P_R$}
+	\State Compute $Succ(p)$, the set of the successors of $p$ in the multigraph
+\EndFor
+\State $\rho \gets$ computeRho$(P,E)$
+\State $A \gets \emptyset$
+\ForAll {$p \in P_R / I_p \neq \emptyset$}
+	\State $A \gets A \cup p$
+\EndFor
+\While{$A \neq \emptyset$}
+	\State Select $p_i \in A$
+	\State $A \gets A \setminus \{p_i\}$
+	\While{true}
+		\State $res \gets SmtSolve\left[\rho \wedge b_i^s \wedge
+		\displaystyle\bigwedge_{\substack{j\neq i \\
+		j\in P_R}} (\neg b_j^s) \wedge x_i \in X_{p_i} \wedge
+		\bigvee_{\substack{j \\ p_j\in Succ(p_i)}} \left(b_2^d \wedge \neg (x'_j \in
+		X_{p_j})\right)\right]$
+		\If {$res = unsat$}
+			\State \textbf{break}
+		\EndIf
+		\State Compute the focus path $e$ from $p_i$ to $p_j$
+		\State $Y \gets \tau_e^\#(X_{p_i})$
+		\If {$p_j \in P_W$}
+			\State $X_{p_j} \gets X_{p_j} \nabla (X_{p_j} \sqcup Y)$
+		\Else
+			\State $X_{p_j} \gets X_{p_j} \sqcup Y$
+		\EndIf
+		\State $A \gets A \cup \{p_j\}$
+	\EndWhile
+\EndWhile
+\State Possibly do some narrowing steps
+\State Compute $\{X_{p_i},\ i \notin P_R\}$
+\State \textbf{return} $\{X_{p_i},\ i \in P\}$
+\EndProcedure
+\end{algorithmic}
+\end{algorithm}
+
+	\subsection {Precision of the analysis}
 
 	\subsection{Disjunctive invariants}
+
+	In this subsection, we propose an extension of the path focusing technique
+	to compute disjunctive invariants, in order to improve precision of the
+	analysis.
+
+	\cite{GulwaniZ10} proposes a technique to compute transitive closure of a
+	transition system, i.e compute the invariants for a loop having its semantic
+	expressed by the transition system.
+
+	A transition system for a control point $p$ is a DNF formula where each
+	disjunct is the semantic of a path from $p$ to itself. It is of the form
+	$$\bigvee_{1 \leq i \leq n} \tau_{p,i}$$ where $n$ is the number of
+	paths, and $\tau_{p,i}$ is the semantic of the $i$-th path from $p$ to itself.
+
+	The technique aims to compute a disjunctive invariant for this control
+	point: 
+	$$\displaystyle\bigvee_{1 \leq j \leq m} X_{p,j}$$
+	where $X_{p,j}$ is a conjunction of linear inequalities, i.e a convex
+	polyhedra.
+
+	The essence of the method is to
+	choose an integer $\delta \in \{1,..,m\}$, and a mapping function
+	$\sigma: \{1,..,m\} \times \{1,..,n\} \mapsto \{1,..,m\}$.
+	For each polyhedra of the disjunctive invariant, and for each path in the
+	graph, the image of the polyhedra $X_{p,j}$ by the transition
+	$\tau_{p,i}$ is joined with $X_{p,\sigma(j,i)}$.
+
+\begin{algorithm}
+\caption{Transitive closure}\label{gulwani}
+\begin{algorithmic}[1] 
+\Procedure{TransitiveClosure}{$\bigvee_{i=1}^n X_{p,i}$}
+\ForAll {$j \in \{1,..,m\} \setminus \{\delta\}$}
+	\State $X_{p,j} \gets \perp$
+\EndFor
+\State $X_{p,\delta} \gets Id$
+\Repeat
+	\For {$i \in \{1,..,n\}$ and $j \in \{1,..,m\}$}
+		\State $X_{p,\sigma(j,i)} \gets X_{p,\sigma(j,i)} \sqcup
+		\tau_{p,i}(X_{p,j})$
+	\EndFor
+\Until {no change in $\bigvee_{j=1}^m X_{p,j}$}
+\EndProcedure
+\end{algorithmic}
+\end{algorithm}
+
+There are heuristics to choose $m,\delta$ and $\sigma$ \cite[Section
+5]{GulwaniZ10}.
+
+This algorithm requires to enumerate all the paths and to compute their
+semantics, which may result in an exponential blowup. In addition, the
+\emph{for} loop at line $7$ iterates on all the $n \times m$ values of $(i,j)$:
+some of the join operations may be useless, in the sense that the value of 
+$X_{p,\sigma(j,i)}$ may not change. Here, one could use the same technique as
+Path Focusing to avoid these drawbacks: one can use SMT-solving to find the
+paths $\tau_{p,i}$.
+
+\begin{algorithm}
+\caption{Transitive closure with implicit transition system}\label{gulwani}
+\begin{algorithmic}[1] 
+\Procedure{TransitiveClosureImplicit}{$p$}
+\ForAll {$j \in \{1,..,m\} \setminus \{\delta\}$}
+	\State $X_{j} \gets \perp$
+\EndFor
+\State $X_{\delta} \gets Id$
+\While {true}
+
+		\State $res \gets SmtSolve\left[\rho \wedge b_k^s \wedge
+		\bigvee_{j=1}^m (B_j \wedge x \in X_{j})
+		\wedge b_k^d
+		\wedge \neg \left(\bigvee_{j=1}^m x' \in X_{j}\right)\right]$
+
+	\If {$res = unsat$}
+		\State \textbf{break}
+	\EndIf
+	\State Compute $\tau_{p,i}$ from $res$ 
+	\State $j$ is the number such as $B_j=true$
+	\State $X_{p,\sigma(j,i)} \gets X_{p,\sigma(j,i)} \sqcup
+	\tau_{p,i}(X_{p,j})$
+\EndWhile
+\EndProcedure
+\end{algorithmic}
+\end{algorithm}
+
+A difference is that here, we don't know the number of paths. Then, the mapping
+function $\sigma$ has to be defined on $\{1,..,m\} \times \N$.
 

### 2011-04-22 report/report.tex
update in the report

@@
 
+Static analysis aims at automatically computing properties on programs, such as
+the possible values of their variables during execution. This allows to show for
+instance that a program will not overflow, will not divide by zero, to compute
+loop invariants\dots
+The main applications of static analysis are compilation-time optimisation, and
+satefy critical systems, such as avionics.
+
+Abstract Interpretation is a general framework used for static analysis.
+Linear Relation Analysis (LRA) is a direct application of this framework, that
+computes an upper-approximation of the set of the possible states of a numerical
+program. The state of a program is defined by the position where we are in the
+code, and the actual values of the different numerical variables.
+The set of possible states is expressed as a least fixpoint of a set of
+equations defining the semantic of the program.
+
+A fundamental fact in static analysis is that it cannot be perfectly precise:
+either the set of possible states we compute is unsound (it doesn't contain all
+the possible states during execution), either it is incomplete (it contains
+states that can never be reached during execution). 
+Abstract Interpretation is incomplete, since it always computes an
+upper-approximation of the set of possible states.
+
+Linear Relation Analysis approximates the set of possible states as a convex
+polyhedron where the dimensions are the numerical variables of the program.
+Instead of convex polyhedra, one could choose intervals, octagons, etc.
+The fixpoint equation is computed iteratively; that is, successive
+approximations of the set of reachable states are computed until they converge
+to a fixpoint.
+
+Satisfiability Modulo Theory (SMT) solving is a field in computer science that
+aims to decide the satisfiability of a logic formula containing boolean
+predicates and elements of a theory, such as linear arithmetic relations between
+integers.
+In this report, we present a novel refinment of the Abstract Interpretation
+technique, that uses SMT-solving to guide the fixpoint iterations such as it
+temporarily focus on a selected path in the control flow graph (the graph
+expressing the semantic of the program), in order to
+obtain more precise results at reasonable cost.
+
+\subsection*{Contribution and Organization}
+
+In this report we make the following contributions:
+\begin{itemize}
+\item We present a new technique, refered as \emph{Path Focusing}, that takes
+benefits of the recent advances in the field of SMT-solving to improve precision
+of static analysis by abstract interpretation. We also present an adaptation of
+this technique to compute disjunctive invariants.
+\item We present an implementation of this technique into a small analyzer, as
+well as the implementation of another technique proposed in \cite{GopanR06},
+such as we can compare precision and cost of the these two methods.
+\end{itemize}
+
+The report is organized as follows:
+Part \ref{stateoftheartpart} introduces the state of the art in Abstract
+Interpretation and more specifically in Linear Relation Analysis.
+Part \ref{pathfocusingpart} describes the two techniques
+\emph{Lookahead Widening} and \emph{Path Focusing}.
+Part \ref{implementationpart} presents the implementation of these two
+techniques into an analyzer, and the experimental results we obtained.
+
 \part{Abstract Interpretation: state of the art}
-
+\label{stateoftheartpart}
 \section{Abstract Interpretation}
@@
 
-\part{Path Focusing technique}
+\part{Path Focusing technique} \label{pathfocusingpart}
 
@@
 
-	We construct the $\rho$ formula as follow:
+	We construct the $\rho$ formula as follows:
 
@@
 	\wedge x_i \in X_{p_i} \wedge \bigvee_{j/(p_i,p_j)\in E} (b_2^d \wedge \neg
-	(x_j \in X_{p_j}))$$
+	(x_j' \in X_{p_j}))$$
+
+	where $x_i$ is the vector of the SSA-variables at the beginning of the path,
+	and $x_j'$ is the vector of the SSA-variable at the end: in this second
+	vector, $\Phi$-variables defined in $p_j$ are primed.
 
@@
 	\item we search a path arriving in a successor $p_j$ of $p_i$, such as the
-	vector of the numerical variables $x_j$ is not yet included in the current
+	vector of the numerical variables $x_j'$ is not yet included in the current
 	abstract value $X_{p_j}$ of $p_j$. $b_j^d$ is \emph{true}, and the
@@
 \EndWhile
-\State Possibly do some narrowing steps
+\State Possibly perform some narrowing steps
 \State Compute $\{X_{p_i},\ i \notin P_R\}$
@@
 	\subsection {Precision of the analysis}
+
+	To Be Completed\dots
 
@@
 
-\begin{algorithm}
-\caption{Transitive closure with implicit transition system}\label{gulwani}
+\begin{algorithm}[!h]
+\caption{Transitive closure with implicit transition system}\label{gulwani2}
 \begin{algorithmic}[1] 
@@
 	\State Compute $\tau_{p,i}$ from $res$ 
-	\State $j$ is the number such as $B_j=true$
+	\State Take $j \in \{ k | B_k = true\}$ 
 	\State $X_{p,\sigma(j,i)} \gets X_{p,\sigma(j,i)} \sqcup
@@
 
+The new algorithm we propose is described in Algorithm \ref{gulwani2}.
 A difference is that here, we don't know the number of paths. Then, the mapping
-function $\sigma$ has to be defined on $\{1,..,m\} \times \N$.
-
-\part{Implementation}
+function $\sigma$ has to be defined on $\{1,..,m\} \times \N$. We could also
+compute this number of path at the beginning of the algorithm.
+
+\part{Implementation}\label{implementationpart}
 
@@
 
+\subsection{LLVM internal representation}
 % We need layers to draw the block diagram
@@
 \begin{tikzpicture}
-    \node (IR) [IR] {Internal Representation\\ (IR)};
+    \node (IR) [IR] {\small Internal\\Representation\\ (IR)};
     % Note the use of \path instead of \node at ... below. 
@@
 internal representation (IR), which is a graph representation of
-the program. The control flow graph of the program can easily be extracted from
+the program. The control flow graph can easily be extracted from
 this representation.
@@
 \begin{figure}[!h]
-\begin{minipage}[c]{.46\linewidth}
+\begin{minipage}[c]{.45\linewidth}
 \begin{C}
-int f(int n) {
-	int x;
-	int y = n;
-	if (n > 0) {
-		x = 50;
-		y++;
-	} else {
-		x = 10;
-		y--;
+int main() {
+   int x = 0;
+   int y = 0;
+
+   while (1) {
+      if (x <= 50) y++;
+      else y--;
+
+      if (y < 0) break;
+      x++;
    }
-	return x+y;
 }
@@
 \end{minipage}
-\begin{minipage}[c]{.46\linewidth}
-\begin{LLVM}
-define i32 @f(i32 %n) nounwind {
-  %1 = icmp sgt i32 %n, 0
-  br i1 %1, label %2, label %4
-
-; <label>:2                                       ; preds = %0
-  %3 = add nsw i32 %n, 1
-  br label %6
-
-; <label>:4                                       ; preds = %0
-  %5 = add nsw i32 %n, -1
-  br label %6
-
-; <label>:6                                       ; preds = %4, %2
-  %x.0 = phi i32 [ 50, %2 ], [ 10, %4 ]
-  %y.0 = phi i32 [ %3, %2 ], [ %5, %4 ]
-  %7 = add nsw i32 %x.0, %y.0
-  ret i32 %7
-}
-\end{LLVM}
+\begin{minipage}[c]{.45\linewidth}
+\includegraphics[width=8.5cm]{"images/cfg_gopan"}
 \end{minipage}
-\caption{Example of C program and its internal representation in SSA form}
+\caption{Our running example and its internal representation in SSA form}
 \end{figure}
@@
 
+\subsection{Drawbacks}
+
+During the implementation, we encountered problems due to the use of the LLVM
+internal representation. Indeed, this format is not provided for static
+analysis, but for code generation. Then, some informations we need to do precise
+static analysis are lacking.
+
+This is the case for the distinction between \emph{signed} and \emph{unsigned}
+integers, which is inexistant in the representation, since the compiler doesn't
+need this information to generate code. Yet, for static analysis, one need to
+have this distinction to improve precision:
+for instance, if $n$ is unsigned, one could add the constraint $n \geq 0$ in our
+	abstract value at the beginning of the analysis.
+	This lack of precision may be recovered by using disjunctive invariants: one
+	could attach two polyhedra to the control point: the first one with the
+	constraint $n \geq 0$, and the second one with $n < 0$.
+
  \section{Abstract domain representation}
 
-		The analyzer is implemented with the APRON library \cite{JM09}.
-		% live variables
+		The analyzer is implemented with the APRON library \cite{JM09}, which
+		is a common interface to libraries implementing all the features
+		for several abstract domains, such as convex polyhedra (NewPolka), 
+		octagons (OCT), intervals (BOX).
+		In our case, we used the convex polyhedra library, but we could switch
+		to octagons or intervals easily.
+
+		\subsection{Dimensions of the Abstract values}
 		To compute the set of live variables in SSA: \cite{Boi10}
@@
 
+	To Be Completed
+
   \section*{Conclusion}
+
+	To Be Completed
 

### 2011-05-02 report/report.tex
minor changes in the report

@@
 The main applications of static analysis are compilation-time optimisation, and
-satefy critical systems, such as avionics.
+the proof of satefy properties in critical systems, such as avionics.
 
@@
 computes an upper-approximation of the set of the possible states of a numerical
-program. The state of a program is defined by the position where we are in the
-code, and the actual values of the different numerical variables.
+program. The state of a program is defined by the position of the program
+counter in the code, and the current values of the different numerical variables.
 The set of possible states is expressed as a least fixpoint of a set of
-equations defining the semantic of the program.
+equations defining the semantics of the program.
 
@@
 either the set of possible states we compute is unsound (it doesn't contain all
-the possible states during execution), either it is incomplete (it contains
+the possible states during execution), or it is incomplete (it contains
 states that can never be reached during execution). 
-Abstract Interpretation is incomplete, since it always computes an
+Abstract Interpretation is sound but incomplete, since it always computes an
 upper-approximation of the set of possible states.
@@
 Satisfiability Modulo Theory (SMT) solving is a field in computer science that
-aims to decide the satisfiability of a logic formula containing boolean
+aims to decide the satisfiability of a logic formula containing Boolean
 predicates and elements of a theory, such as linear arithmetic relations between
 integers.
-In this report, we present a novel refinment of the Abstract Interpretation
-technique, that uses SMT-solving to guide the fixpoint iterations such as it
-temporarily focus on a selected path in the control flow graph (the graph
-expressing the semantic of the program), in order to
+In this report, we present a novel refinement of the Abstract Interpretation
+technique, that uses SMT-solving to guide the fixpoint iterations, so that it
+temporarily focuses on a selected path in the control flow graph (the graph
+expressing the semantics of the program), in order to
 obtain more precise results at reasonable cost.

### 2011-05-02 report/report.tex
report

@@
 \def\ite{\textnormal{ite\ }}
+\def\lfp{\textnormal{\it lfp}}
 
@@
 of the program, i.e the set of all possible states of the program. This
-computation quickly becomes too costly.
+computation quickly becomes too costly or doesn't terminate.
 
@@
 If $C$ is a complete lattice, and if $\Phi$ is increasing from $C$ to $C$, then
-$$\alpha(lfp(\Phi)) \leq_A lfp(\alpha(\Phi))$$
-where $lfp(\Phi)$ denotes the least fixpoint of $\Phi$.
+$$\alpha(\lfp(\Phi)) \leq_A \lfp(\alpha(\Phi))$$
+where $\lfp(\Phi)$ denotes the least fixpoint of $\Phi$.
 \end{proposition}
@@
 any sequence $(y_i)_{i \geq 0}$ of elements of the abstract domain $A$ such as
-$\forall i \geq 0,\ y_i \leq_A y_{i+1}$. Indeed, in case of a domain of infinite
+$\forall i \geq 0,\ y_i <_A y_{i+1}$. Indeed, in case of a domain of infinite
 depth, the least fixpoint
@@
 
+\cite{Monniaux_HOSC09} gives a more general definition of the widening operator.
+
 The least fixpoint of a function $\Phi^\#$ is noted $\overline{y}$ and is equal
@@
 
-In $\Q^n$, the set of polyhedra is a lattice. The least element is $\perp$
+In $\Q^n$, the set of polyhedra ordered by inclusion is a lattice. The least element is $\perp$
 (the empty polyhedron), and the greatest element is $\top$ (the whole $Q^n$).
@@
 \cite{CH78,Hal79,HPR97,BlanchetCousotEtAl_PLDI03}, in order to fight
-against the loss of precision induced by widening:
+against the loss of precision induced by widening.
 
@@
 \item $I_p$ is the (possibly empty)
-set of initial values for each control point $p \in P$.
+set of initial values for each control point $p \in P$. For an initial control
+point $p$, i.e a starting point of the program, $I_p$ is not empty. It is empty
+in the other case.
 \item $E \subseteq  P \times P$ is a set of directed edges. Each edge $e \in E$
@@
 
+\begin{figure}[h!]
+\centering
+\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1.5cm,
+                    semithick]
+
+  \node[state] (p)             [label=left:$X_{p}$]       {$p$};
+  \node[state]         (p2) [above  of=p] {};
+  \node[state]         (p1) [left of=p2,label=left:$X_{p'}$] {$p'$};
+  \node[state]         (p3) [right of=p2] {};
+  \node         (p22) [below  of=p] {};
+  \node         (p12) [left of=p22] {};
+  \node         (p32) [right of=p22] {};
+
+  \path [transition] (p1) edge              node {} (p);
+  \path [transition] (p2) edge              node {} (p);
+  \path [transition] (p3) edge              node {} (p);
+  \path [transition2] (p) edge              node {} (p12);
+  \path [transition2] (p) edge              node {} (p22);
+  \path [transition2] (p) edge              node {} (p32);
+\end{tikzpicture}
+\end{figure}
+
 The function $\Phi^\#$ previously described is the following:
@@
 
-In the case of abstract domain with infinite ascending sequence, we use the
+In the case of an abstract domain with infinite ascending sequence, we use the
 previously defined widening operator:
@@
 apply it on a subset of $P$, called $P_w$, such as removing the nodes of $P_w$
-disconnect every cycles. For instance, one could choose $P_w$ as the set of the
+disconnects every cycles. For instance, one could choose $P_w$ as the set of the
 heads of loops.
@@
 There exist several iteration strategies for computing the fixpoint: whatever
-the order we choose for updating the different $X_p, p\in P$, the result at the
+order we choose for updating the different $X_p, p\in P$, the result at the
 end of the analysis is correct. Yet, the precision of the result and the time
@@
 narrowing iterations, which often recover some precision.
+
+For instance, considering the program:
+\begin{C}
+for (int i=0; i < 100; i++) {
+}
+\end{C}
+The analysis starts with $i=0$. After one iteration, one obtain $i \in
+[0,1]$, \dots. After a widening operation, the result becomes $i \in [0,
++\infty]$, which is an invariant, but very imprecise. We can do a new iteration,
+with $i\in [0, +\infty]$ at the beginning of the loop, and we find that the
+values for $i$ at the next step are in $[0, 100]$. This is still an invariant,
+but much better.
+
 \item There is also a loss of precision each time a point in the control flow
@@
 
-Typically, to analyze a program with loops, the classical approach is to use the
+Typically, to analyze a program with loops, the approach is to use the
 widening operator, for example at the head of the loops, to ensure convergence.
@@
 
+	Intuitively, expanding the graph comes back to consider independently the
+	different paths in the control flow graph, and thus to be more precise,
+	because paths that are not feasible will be ignored. Additionally, the
+	number of control points with several incoming transitions will be reduced,
+	so the loss of precision due to the convex hull of polyhedra may be
+	minimized.
+
+\begin{figure}[!h]
+\label{multigraph}
+\centering
+\begin{minipage}[c]{.19\linewidth}
+\begin{tikzpicture}[->,>=stealth',auto,node distance=1.1cm,
+                    semithick,font=\footnotesize]
+
+	\node[PRstate] (n0) {$p$};
+	\node[state] (n1) [below left of=n0] {};
+	\node[state] (n2) [below right of=n0] {};
+	\node[state] (n3) [below right of=n1] {};
+	\node[state] (n4) [below left of=n3] {};
+	\node[state] (n5) [below right of=n3] {};
+	\node[state] (n6) [below right of=n4] {};
+	\node[state] (n7) [below left of=n6] {};
+	\node[state] (n8) [below right of=n6] {};
+	\node[PRstate] (n9) [below right of=n7] {$q$};
+
+  \path [transition] 
+		(n0) edge              node {} (n1);
+  \path [transition] 
+		(n0) edge              node {} (n2);
+  \path [transition] 
+		(n1) edge              node {} (n3);
+  \path [transition] 
+		(n2) edge              node {} (n3);
+  \path [transition] 
+		(n3) edge              node {} (n4);
+  \path [transition] 
+		(n3) edge              node {} (n5);
+  \path [transition] 
+		(n4) edge              node {} (n6);
+  \path [transition] 
+		(n5) edge              node {} (n6);
+  \path [transition] 
+		(n6) edge              node {} (n7);
+  \path [transition] 
+		(n6) edge              node {} (n8);
+  \path [transition] 
+		(n7) edge              node {} (n9);
+  \path [transition] 
+		(n8) edge              node {} (n9);
+\end{tikzpicture}
+\end{minipage} 
+$\Longrightarrow$
+\begin{minipage}[c]{.19\linewidth}
+\begin{tikzpicture}[->,>=stealth',auto,node distance=1.1cm,
+                    semithick,font=\footnotesize]
+
+	\node[PRstate] (n0) {$p$};
+	\node (n1) [below left of=n0] {};
+	\node (n2) [below right of=n0] {};
+	\node (n3) [below right of=n1] {};
+	\node (n4) [below left of=n3] {};
+	\node (n5) [below right of=n3] {};
+	\node (n6) [below right of=n4] {};
+	\node (n7) [below left of=n6] {};
+	\node (n8) [below right of=n6] {};
+	\node[PRstate] (n9) [below right of=n7] {$q$};
+
+  \path [transition] 
+		(n0) edge  [out=0, in=0]            node {} (n9);
+  \path [transition] 
+		(n0) edge  [out=180, in=180]        node {} (n9);
+  \path [transition] 
+		(n0) edge  [out=205, in=155]            node {} (n9);
+  \path [transition] 
+		(n0) edge  [out=230, in=130]            node {} (n9);
+  \path [transition] 
+		(n0) edge  [out=255, in=105]            node {} (n9);
+  \path [transition] 
+		(n0) edge  [out=-25, in=25]            node {} (n9);
+  \path [transition] 
+		(n0) edge  [out=-50, in=50]            node {} (n9);
+  \path [transition] 
+		(n0) edge  [out=-75, in=75]            node {} (n9);
+\end{tikzpicture}
+\end{minipage}
+\caption{One can expand the graph in the left and obtain the associate graph in
+the right.}
+\end{figure}
+\FloatBarrier
+
 	Let $(P,E)$ be the control flow graph of the program.
@@
 	\begin{itemize}
-	\item $I_p = \emptyset$ for each $p \in P \ P_R$, meaning that the initial
+	\item $I_p = \emptyset$ for each $p \in P \setminus P_R$, meaning that the initial
 	points of the program are included in $P_R$.
@@
 	which path in the control flow graph is selected, by giving a model: an
-	assignments of the boolean variables such as the formula is true.
+	assignments of the boolean variables such that the formula is true.
 	Among these boolean variables, some of them are associated to a transition
@@
 terms of the boolean and numerical SSA variables. For instance, if the CFG
-contains the affectation $x.2 = x.1 + 1$, then we simply add $x.2 = x.1 + 1$ in
+contains the assignment $x.2 = x.1 + 1$, then we simply add $x.2 = x.1 + 1$ in
 the formula. This equality expression is noted $assign(x.2)$. 
@@
 transition. For instance, $c(i,j)$ could be of the form $x < y$, \dots
-If the transition is undeterministic, then $c(i,j) = c_{i,j}$, where
-$c_{i,j}$ is a boolean variable left undeterministic.
+If the transition is non-deterministic, then $c(i,j) = c_{i,j}$, where
+$c_{i,j}$ is a boolean variable left non-deterministic.
 \item For each point $p_i \notin P_R$, we set the reachability predicate 
@@
 depending on the incoming transition we come from. Generally, for a
-$Phi$-variable $x = \Phi(x_1,x_2,\dots,x_n)$ in control point $p_i$ with the
+$\Phi$-variable $x = \Phi(x_1,x_2,\dots,x_n)$ in control point $p_i$ with the
 ordered incoming transitions $t_{j_1,i},t_{j_2,i},\dots,t_{j_n,i}$, we set:
@@
 
-	$\rho$ express the semantic of the program. We still need to create the
+	The SMT-formula $\rho$ express the semantic of the program. We still need to create the
 	formula expressing the reachability problem defined in \ref{reachability}.
@@
 	different SSA-variables of the program. Then, it is easy to encode the
-	property $x \in X$ into an SMT-formula: one simply write the conjunction of
+	property $x \in X$ into an SMT-formula: one simply writes the conjunction of
 	these inequalities, with the name of the variables in the vector $x$.
 
-	The formula given to the SMT-solver is:
+	To compute the reachability strating from point $p_i$, the formula given to
+	the SMT-solver is:
 
@@
 The advantage of using such an infrastructure is that the analyzer can check
-programs written in different languages: C, C++, \dots, without needing to write
+programs written in various languages: C, C++, \dots, without needing to write
 a specific frontend for each of them. Here, the analyzer works directly on the

### 2011-05-03 report/report.tex
report

@@
 \label{stateoftheartpart}
+
+\section{Introductive example}
+
+We consider the following program:
+
+\begin{figure}[!h]
+   \begin{minipage}[c]{.46\linewidth}
+\begin{C}
+x = 0;
+while (x < 100) {
+	x++;
+}
+\end{C}
+   \end{minipage} \hfill
+   \begin{minipage}[c]{.46\linewidth}
+\begin{tikzpicture}[->,>=stealth',auto,node distance=1.7cm,
+                    semithick,font=\footnotesize]
+
+	\node[state] (n0) {$p_0$};
+	\node[state] (n1) [below of=n0] {$p_1$};
+	\node[state] (n2) [below of=n1] {$p_2$};
+	\node[state] (n3) [left of=n1] {$p_3$};
+
+  \path [transition] 
+		(n0) edge              node {$x = 0$} (n1);
+  \path [transition] 
+        (n1) edge			   node [above] {$x \geq 100?$} (n3);
+  \path [transition] 
+        (n1) edge			   node [left] {$x < 100?$} (n2);
+  \path [transition] 
+        (n2) edge [out=0, in=0, distance=2cm] node [right] {$x = x+1$} (n1);
+
+\end{tikzpicture}
+   \end{minipage}
+\end{figure}
+\FloatBarrier
+
+Static analysis aims to discover some properties about programs. In this
+example, one would to compute the set of possible values for the variable $x$
+during execution.
+The graph at the right side represents the program. The nodes of the graph are
+called \emph{program points}. We can compute for instance the set $Y_1$ of
+values for $x$ at point $p_1$. This set depends on the sets $Y_0$ and $Y_2$,
+since there are two edges arriving at $p_1$: the first one comes from $p_0$, and
+the second one from $p_2$.
+
+There is a relation between $Y_1$ and $Y_2$ :
+\begin{eqnarray*}
+Y_1 &=& \{x\ |\ x=0 \vee \exists x' \in Y_2, x=x'+1 \} \\
+Y_2 &=& \{x\ |\ x \in Y_1 \wedge x < 100 \}
+\end{eqnarray*}
+
+We see that $Y_1$ and $Y_2$ can be computed step by step:
+we start with $Y_1 = Y_2 = \emptyset$.
+Then, we add iteratively new elements in the sets.
+\begin{itemize}
+\item  We see that $0 \in Y_1$: $Y_1 = \{0\}$.
+\item $Y_1$ contains $0$. So, $Y_2$ also contains $0$ by definition: $Y_2 =
+\{0\}$.
+\item $Y_2$ contains $0$, so $0+1=1$ is in $Y_1$: $Y_1 = \{0,1\}$.
+\item Again, we find $Y_2 = \{0,1\}$.
+\item $Y_2$ now contains $1$, so $1+1=2 \in Y_1$: $Y_1 = \{0,1,2\}$.
+\item We continue the iteration until we find no more elements in the sets\dots
+\end{itemize}
+
+This computation is a fixpoint computation: one updates the different sets until
+we find no more elements. These sets are strictly increasing (we always add new
+elements in the sets), and if we find no new elements to add, that means we have
+reached a limit.
+
 \section{Abstract Interpretation}
@@
 Since the domain of convex polyhedra is of infinite depth, a widening operator
-is defined to ensure convergence of the technique. There have been different
+is defined to ensure convergence of the technique. There have been various
 work proposing a definition of the widening operator in LRA
@@
    program comes from \cite{GopanR06}.}
+\label{runningexample}
 \end{figure}
@@
 Linear Relation Analysis is sound but incomplete: there is loss of precision
-during the computation, due to different reasons:
+during the computation, due to various reasons:
 \begin{enumerate}
@@
 
-To Be Completed
+In some cases, a loop can have a non-regular behaviour, especially when it has
+several paths. Some of these paths can stay impossible for a while, and become
+possible after a certain number of iterations. However, the widening operator
+extrapolates without taking care of these new possible paths. This can induce a
+loss of precision, that Lookahead widening aims to limit.
+
+Lookahead widening \cite{GopanR06} is a technique that aims to isolate the
+different loop phases during the analysis, in order to obtain more precise
+results.
+
+In the program already depicted in figure \ref{runningexample}, there is a loop
+with two perfectly distinct phases:
+\begin{itemize}
+\item During the 51 first iterations, both $x$ and $y$ are incremented.
+\item During the 51 last iterations, $x$ is incremented and $y$ is decremented.
+\end{itemize}
+
+When computing the fixpoint iterations, we only consider a subset of the graph
+for a while, until the convergence. In our example:
+\begin{itemize}
+\item Step 1:
+	we only consider the path of
+	the loop that is possible at the first iteration. The second path is
+	ignored, and will be treated later. we compute the iterations, and we obtain
+	an first invariant for this part of the graph. Then, we can do some
+	narrowing steps in order to recover some precision, before adding new paths
+	in the graph.
+\item Step 2:
+	Now, we add the new paths into our graph if they have become possible. In
+	our case, the second path of the loop is now possible, so we add the path in
+	the graph and compute new iterations until convergence\dots
+\item Step 3: finally, the path from $p_4$ to $p_7$, so we add it into the graph and
+compute the invariant for $p_7$.
+\end{itemize}
+
+\begin{figure}[!h]
+   \begin{minipage}[c]{.46\linewidth}
+\begin{tikzpicture}[->,>=stealth',auto,node distance=1.5cm,
+                    semithick,font=\footnotesize]
+
+	\node[state] (n0) {$p_0$};
+	\node[state] (n1) [below of=n0] {$p_1$};
+	\node[state] (n2) [below left of=n1] {$p_2$};
+	\node[state] (n3) [below right of=n1] {$p_3$};
+	\node[state] (n4) [below left of=n3] {$p_4$};
+	\node[state] (n5) [below of=n4] {$p_5$};
+	\node[state] (n6) [below of=n5] {$p_6$};
+	\node[state] (n7) [left of=n6] {$p_7$};
+
+	\node (n8) [right of=n6] {};
+	\node (n9) [right of=n1] {};
+
+  \path [transition] 
+		(n0) edge              node {$x,y \leftarrow 0$} (n1);
+  \path [transition] 
+        (n1) edge			   node [left] {$x \leq 50$} (n2);
+  \path [transition2] 
+        (n1)  edge              node [right] {$x \geq 51$} (n3);
+  \path [transition] 
+        (n2) edge              node [left] {$y \leftarrow y+1$} (n4);
+  \path [transition2] 
+        (n3) edge			   node [right] {$y \leftarrow y-1$} (n4);
+  \path [transition] 
+        (n4) edge			   node {$y \geq 0$} (n5);
+  \path [transition2] 
+		(n4) edge  [out = 180, in=90] node [left] {$y \leq -1$} (n7);
+  \path [transition] 
+        (n5) edge              node {$x \leftarrow x+1$} (n6);
+  \path [transition] 
+        (n6) edge [out=0, in=0, distance=3.5cm] node {} (n1);
+
+\end{tikzpicture}
+\centering Step 1
+   \end{minipage} \hfill
+   \begin{minipage}[c]{.46\linewidth}
+\begin{tikzpicture}[->,>=stealth',auto,node distance=1.5cm,
+                    semithick,font=\footnotesize]
+
+	\node[state] (n0) {$p_0$};
+	\node[state] (n1) [below of=n0] {$p_1$};
+	\node[state] (n2) [below left of=n1] {$p_2$};
+	\node[state] (n3) [below right of=n1] {$p_3$};
+	\node[state] (n4) [below left of=n3] {$p_4$};
+	\node[state] (n5) [below of=n4] {$p_5$};
+	\node[state] (n6) [below of=n5] {$p_6$};
+	\node[state] (n7) [left of=n6] {$p_7$};
+
+	\node (n8) [right of=n6] {};
+	\node (n9) [right of=n1] {};
+
+  \path [transition] 
+		(n0) edge              node {$x,y \leftarrow 0$} (n1);
+  \path [transition] 
+        (n1) edge			   node [left] {$x \leq 50$} (n2);
+  \path [transition] 
+        (n1)  edge              node [right] {$x \geq 51$} (n3);
+  \path [transition] 
+        (n2) edge              node [left] {$y \leftarrow y+1$} (n4);
+  \path [transition] 
+        (n3) edge			   node [right] {$y \leftarrow y-1$} (n4);
+  \path [transition] 
+        (n4) edge			   node {$y \geq 0$} (n5);
+  \path [transition2] 
+		(n4) edge  [out = 180, in=90] node [left] {$y \leq -1$} (n7);
+  \path [transition] 
+        (n5) edge              node {$x \leftarrow x+1$} (n6);
+  \path [transition] 
+        (n6) edge [out=0, in=0, distance=3.5cm] node {} (n1);
+
+\end{tikzpicture}
+\centering Step 2
+   \end{minipage}
+   \caption{Dotted arrows are ignored by the analysis.}
+\label{gopanreps}
+\end{figure}
+\FloatBarrier
 
@@
 	Path focusing technique uses control flow graphs in static single assignment
-	(SSA) form. The main concept behind the SSA form is that every variable is
+	(SSA) form. The main concept behind the SSA form is that, syntactically,
+	every variable is	
 	only being assigned once.
@@
 	This is a reachability problem, that can be expressed as an SMT-formula,
-	which is a logic formula with boolean variables, and elements of a certain
+	which is a logic formula with Boolean variables, and elements of a certain
 	theory $T$. Depending on the program, one can choose different theories:
@@
 
-	Despite deciding the satisfiability of such formula is NP-complete, there
-	has been lots of research on decision procedures \cite{Kroening08} and there exist nowadays
+	Although deciding the satisfiability of such formula is NP-complete, there
+	has been much research on decision procedures \cite{Kroening08} and there exist nowadays
 	efficient programs, known as SMT-solvers, that can decide the satisfiability
@@
 	which path in the control flow graph is selected, by giving a model: an
-	assignments of the boolean variables such that the formula is true.
-	Among these boolean variables, some of them are associated to a transition
+	assignments of the Boolean variables such that the formula is true.
+	Among these Boolean variables, some of them are associated to a transition
 	in the control flow graph:
 	\begin{itemize}
-	\item if the model has set to \emph{true} the boolean variable $b_e$ associated
+	\item if the model has set to \emph{true} the Boolean variable $b_e$ associated
 	to the transition $e$, then the focus path goes through this transition.
@@
 	Path Focusing algorithm is described in Algorithm \ref{pathfocusingalgo}.
+	We maintain a set $A$ of point in $P_R$ that need to be treated.
+	We select iteratively one of these control points, and search for new focus
+	paths starting from this point, until there is no more path to be treated.
+	At the end of the analysis, one could perform some narrowing iterations in
+	order to recover some precision.
 \begin{algorithm}
@@
 
-	To Be Completed\dots
+	One can refine the \emph{Path-focusing} algorithm by treating specially the
+	self loops of the multigraph. Indeed, instead of simply apply the widening
+	operator, one could compute some narrowing steps to regain precision before
+	focusing on another path. This is the same principle as the one we find in
+	\emph{Lookahead widening} \cite{GopanR06}.
+
+	So, the algorithm has a special case if the path we focus on goes from $p_i$
+	to itself: instead of computing $\tau_e^\#(X_{p_i})$, one compute
+	$SelfLoop(\tau_e^\#,X_{p_i})$, defined in algorithm \ref{loopiteralgo}.
+
+	The $SelfLoop$ procedure actually computes a widening operation, and
+	computes a narrowing step in order to recover precision. In some cases,
+	instead of performing widening/narrowing iterations, one could use
+	acceleration techniques to find even more precise invariants \cite{GH06}.
+
+	In order for this refinement to be precise, one needs to delay the widening
+	operator when we consider a self-loop for the first time. Then, our
+	algorithm uses a set $U$ of paths that have already been treated. If the
+	self-loop we focus on is not in $U$, we apply a simple union instead of
+	widening. Widening will be applied next time to guarantee the convergence.
+
+\begin{algorithm}
+\caption{SelfLoop}
+\label{loopiteralgo}
+\begin{algorithmic}[1] 
+\Procedure{SelfLoop}{$\tau_e^\#$,$X_{p_i}$}
+	\State $Y \gets \tau_e^\#(X_{p_i})$
+	\State $X' \gets X_{p_i} \sqcup Y$
+	\State $Y \gets \tau_e^\#(X')$
+	\State \textbf{return} $Y$
+\EndProcedure
+\end{algorithmic}
+\end{algorithm}
+
+\begin{algorithm}
+\caption{Path Focusing with special treatment for self loops}
+\label{pathfocusingoptalgo}
+\begin{algorithmic}[1] 
+\Procedure{pathfocusing}{$P$,$E$,$P_R$}
+\ForAll {$p \in P_R$}
+	\State Compute $Succ(p)$, the set of the successors of $p$ in the multigraph
+\EndFor
+\State $\rho \gets$ computeRho$(P,E)$
+\State $A \gets \emptyset$
+\ForAll {$p \in P_R / I_p \neq \emptyset$}
+	\State $A \gets A \cup p$
+\EndFor
+\While{$A \neq \emptyset$}
+	\State Select $p_i \in A$
+	\State $A \gets A \setminus \{p_i\}$
+	\State $U \gets \emptyset$
+	\While{true}
+		\State $res \gets SmtSolve\left[\rho \wedge b_i^s \wedge
+		\displaystyle\bigwedge_{\substack{j\neq i \\
+		j\in P_R}} (\neg b_j^s) \wedge x_i \in X_{p_i} \wedge
+		\bigvee_{\substack{j \\ p_j\in Succ(p_i)}} \left(b_2^d \wedge \neg (x'_j \in
+		X_{p_j})\right)\right]$
+		\If {$res = unsat$}
+			\State \textbf{break}
+		\EndIf
+		\State Compute the focus path $e$ from $p_i$ to $p_j$
+		\If{$p_i = p_j$}
+			\State $Y \gets SelfLoop(\tau_e^\#,X_{p_i})$
+		\Else
+			\State $Y \gets \tau_e^\#(X_{p_i})$
+		\EndIf
+		\If {$(p_j \in P_W) \wedge (p_i \neq p_j \vee e \in U)$}
+			\State $X_{p_j} \gets X_{p_j} \nabla (X_{p_j} \sqcup Y)$
+		\Else
+			\State $X_{p_j} \gets X_{p_j} \sqcup Y$
+			\State $U \gets U \cup \{e\}$
+		\EndIf
+		\State $A \gets A \cup \{p_j\}$
+	\EndWhile
+\EndWhile
+\State Possibly perform some narrowing steps
+\State Compute $\{X_{p_i},\ i \notin P_R\}$
+\State \textbf{return} $\{X_{p_i},\ i \in P\}$
+\EndProcedure
+\end{algorithmic}
+\end{algorithm}
 
@@
 function $\sigma$ has to be defined on $\{1,..,m\} \times \N$. We could also
-compute this number of path at the beginning of the algorithm.
-
+compute this number of paths at the beginning of the algorithm.
+
+\FloatBarrier
 \part{Implementation}\label{implementationpart}
@@
 programs written in various languages: C, C++, \dots, without needing to write
-a specific frontend for each of them. Here, the analyzer works directly on the
+a specific frontend for each of them. Here, the analyzer directly works on the
 internal representation (IR), which is a graph representation of
@@
 this representation.
+
+Our analyzer takes as parameter the LLVM internal representation of a program,
+and returns linear relations between the numerical variables at each control
+point of this program. It first applies some optimisation passes over the IR,
+and then computes the linear relations by abstract interpretation.
 
@@
 internal representation. Indeed, this format is not provided for static
-analysis, but for code generation. Then, some informations we need to do precise
+analysis, but for code generation. Then, some informations we need for a precise
 static analysis are lacking.
@@
 
+Another example of information that was missing is the set of live variables.
+Indeed, in LLVM, live variables computation is processed after the register
+allocation. Then, the data structure we work on doesn't have live variables
+informations yet. To solve this problem, we coded our own pass, that aims to
+compute these live variables after the transformation in SSA form.
+
  \section{Abstract domain representation}
 
-		The analyzer is implemented with the APRON library \cite{JM09}, which
+	The analyzer is implemented using the APRON library \cite{JM09}, which
 	is a common interface to libraries implementing all the features
@@
 	\subsection{Dimensions of the Abstract values}
-		To compute the set of live variables in SSA: \cite{Boi10}
+
+	In our implementation, we tried to reduce as much as possible the dimensions
+	of the abstract values at each program point. Indeed, there is no need to
+	consider the whole set of the program numerical variables at each point $p$,
+	since lots of these variables are not \emph{live} in $p$. 
+	The definition of a \emph{live} variable is the following:
+
+	\begin{definition}[Liveness]
+	A variable is \emph{live} at the control point $p$ iff:
+	\begin{itemize}
+	\item its value is available at $p$, meaning that there is a path from the
+	point defining the variable and $p$.
+	\item its value might be used in the future, meaning that there exist a path
+	starting in $p$, that arrives in a point using the variable.
+	\end{itemize}
+	\end{definition}
+
+	LLVM internal representation allows to find the control point where
+	a variable is defined and used, thanks to def-use chains: we can
+	easily get the set of points where a variable is used, and we directly have
+	a pointer to its definition.
+	With all these informations, we implemented an LLVM pass that computes 
+	the set of live variables in SSA, using the liveness check algorithm
+	described in \cite[section 3.3]{Boi10}.
+
+	At a program point $p$, the dimensions of the associate abstract value could
+	only be the numerical variables that are live at $p$. This is
+	not exactly the case in our implementation, because of another 
+	optimisation: we only consider
+	variables that are not a linear combination of other variables.
+
+	For instance, assume that $x,y,z$ are numerical variables of a program,
+	$x$ is defined as $x = y+z$, and $x,y,z$ are live at point $p$. Instead of having
+	$x$ as a dimension for $X_p$, we only have $y$ and $z$. All the properties
+	for $x$ can be directly extracted from $X_p$ and the information $x=y+z$.
+	This is an optimisation in the sense that there is redundant information in
+	the abstract value if both $x,y$ and $z$ are dimensions of $X_p$.
+
+	Then, liveness definition can be adapted in our case:
+
+	\begin{definition}[Liveness\dots]
+	A variable $v$ is \emph{live\dots} at the control point $p$ iff:
+	\begin{itemize}
+	\item 
+	its value is available at $p$, meaning that there is a path from the
+	point defining the variable to $p$,
+	\item one of these conditions holds:
+		\begin{itemize}
+		\item $v$ is live in $p$.
+		\item There is a variable $v'$, defined as a linear combination of other
+		variables $v_1, v_2, \dots, v_n$, such that $\exists i \in \{1,\dots,n\}, v = v_i$,
+		and $v'$ is live\dots in $p$.
+		\end{itemize}
+	\end{itemize}
+	\end{definition}
+
+	Finally, a variable is a dimension of $X_p$ iff it is live\dots and it is
+	not defined as a linear combination of program variables.
+
+	\subsection{SMT-solving}
+
+	\emph{Path focusing} technique requires to decide the satisfiability of some
+	SMT formula, and to extract a model when the formula is satisfiable, i.e an
+	assignement of the variables such that the formula is \emph{true}. For doing
+	this, our implementation has an interface with Microsoft Z3 \cite{MouraB08},
+	and Yices \cite{DutertreM06}. One can easily switch from one to another with
+	an argument we give as parameter to the analyzer. In this way, we can
+	compare the effectiveness of both SMT-solvers.
 

### 2011-05-04 report/report.tex
update of the report

@@
 elements in the sets), and if we find no new elements to add, that means we have
-reached a limit.
+reached a limit. Abstract Interpretation gives a general framework to do such
+kind of fixpoint computation.
 
@@
 
-The fixpoint computation algorithm consist in replacing iteratively each $X_p$ by its value
+The fixpoint computation algorithm consists in replacing iteratively each $X_p$ by its value
 on the right hand side, until the convergence.
@@
 
+\section{Example}
+
+We apply the linear relation analysis technique over the program in figure
+\ref{runningexample}. The set of state where we apply widening is $\{p_1\}$: it
+is sufficient, since removing $p_1$ cuts all cycles in the graph.
+
+To each point of the control flow graph $\{p_0, p_1,\dots, p_7\}$, we attach a
+convex polyhedron, whose dimensions are the numerical variables $x,y$.
+For instance, at point $p_1$, the effective set of possible assignments for
+$(x,y)$ is the union of the two segments $(0,0) - (51,51)$ and $(51,51) -
+(102,0)$, depicted in figure \ref{result}.a. Since this union of segments is not
+a convex polyhedra, the best result we can wait for after abstraction is the
+polyhedra depicted in figure \ref{result}.b, which is the smallest convex
+polyhedra containing the two segments.
+
+\begin{figure}[!h]
+\begin{minipage}[c]{.39\linewidth}
+\centering
+\begin{tikzpicture}[y=.2cm, x=.2cm,font=\footnotesize]
+
+	\node (t1) at (-5,-4) {};
+	\node (t2) at (13,11) {};
+	\draw[line] (0,0) -- (8,8) -- cycle;
+	\draw[line] (16,0) -- (8,8) -- cycle;
+
+ 	%axis
+	\draw (0,0) -- coordinate (x axis mid) (20,0);
+    \draw (0,0) -- coordinate (y axis mid) (0,10);
+
+	%ticks and labels      
+     		\draw [dotted](8,8) -- (-3pt,8) 
+     			node[anchor=east] {$51$}; 
+     		\draw [dotted] (8,8) -- (8,-3pt) 
+     			node[anchor=north] {$51$}; 
+     		\draw [dotted] (16,1pt) -- (16,-3pt) 
+     			node[anchor=north] {$102$};
+
+	\node[right=1.9cm] at (x axis mid) {$x$};
+	\node[above=1.2cm] at (y axis mid) {$y$};
+
+	\node[right=3cm] at (y axis mid) {$p_1$};
+\end{tikzpicture}
+
+a. During execution, $(x,y)$ can only be in one of these two segments.
+\end{minipage} 
+\begin{minipage}[c]{.59\linewidth}
+\centering
+\begin{tikzpicture}[y=.2cm, x=.2cm,font=\footnotesize]
+
+	\node (t1) at (-5,-4) {};
+	\node (t2) at (13,11) {};
+	\draw[line] (0,0) -- (8,8) -- cycle;
+	\draw[line] (16,0) -- (8,8) -- cycle;
+
+	\fill[polyhedra] (0,0) -- (8,8) -- (16,0) -- cycle;
+
+ 	%axis
+	\draw (0,0) -- coordinate (x axis mid) (20,0);
+    \draw (0,0) -- coordinate (y axis mid) (0,10);
+
+	%ticks and labels      
+     		\draw [dotted](8,8) -- (-3pt,8) 
+     			node[anchor=east] {$51$}; 
+     		\draw [dotted] (8,8) -- (8,-3pt) 
+     			node[anchor=north] {$51$}; 
+     		\draw [dotted] (16,1pt) -- (16,-3pt) 
+     			node[anchor=north] {$102$};
+
+	\node[right=1.9cm] at (x axis mid) {$x$};
+	\node[above=1.2cm] at (y axis mid) {$y$};
+
+	\node[right=3cm] at (y axis mid) {$p_1$};
+\end{tikzpicture}
+
+b. The best result we can have is the polyhedra in hashed lines.
+\end{minipage}
+\caption{Best results we can get from the analysis}
+\label{result}
+\end{figure}
+
+We apply abstract interpretation to this program. Let $X_i$ be the polyhedra
+attached to point $p_i$. Initially, $X_i = \perp$ for each $i \in \{1,7\}$, and
+$X_0 = \top$. The tabular in figure \ref{iterations} shows the evolution of the
+polyhedron during the iterations.
+
+After 3 iterations, we have reached an invariant. We can then apply a narrowing
+iteration, that recovers some precision. Finally, at point $p_1$, we see that
+the obtained result is not as precise as we could wish (Figure \ref{result}.b).
+This is due to the widening operator applied at point $p_1$, and the convex hull
+applied at point $p_4$.
+
+\begin{figure}[!h]
+\centering
+\caption{Polyhedron attached to the different points during iterations.}
+\label{iterations}
+\begin{tabular}{|c||c|c|c|c|} \hline
+Iteration & $X_0 \sqcup X_6$ & $X_1$ (widening) & $X_4$ & $X_6$ \\ \hline
+1& 
+\begin{tikzpicture}[y=.1cm, x=.1cm,font=\footnotesize]
+	\node (t1) at (-5,-4) {};
+	\node (t2) at (13,11) {};
+	\fill[line] (0,0) circle (0.8);
+	%\draw[line] (0,0) -- (8,8) -- cycle;
+	%\draw[line] (16,0) -- (8,8) -- cycle;
+	%\fill[polyhedra] (0,0) -- (8,8) -- (16,0) -- cycle;
+ 	%axis
+	\draw (0,0) -- coordinate (x axis mid) (20,0);
+    \draw (0,0) -- coordinate (y axis mid) (0,10);
+	%ticks and labels      
+	\node[right=0.9cm] at (x axis mid) {$x$};
+	\node[above=0.6cm] at (y axis mid) {$y$};
+\end{tikzpicture} 
+&
+\begin{tikzpicture}[y=.1cm, x=.1cm,font=\footnotesize]
+	\node (t1) at (-5,-4) {};
+	\node (t2) at (13,11) {};
+	\fill[line] (0,0) circle (0.8);
+ 	%axis
+	\draw (0,0) -- coordinate (x axis mid) (20,0);
+    \draw (0,0) -- coordinate (y axis mid) (0,10);
+	%ticks and labels      
+	\node[right=0.9cm] at (x axis mid) {$x$};
+	\node[above=0.6cm] at (y axis mid) {$y$};
+\end{tikzpicture} 
+& 
+\begin{tikzpicture}[y=.1cm, x=.1cm,font=\footnotesize]
+	\node (t1) at (-5,-4) {};
+	\node (t2) at (13,11) {};
+	\fill[line] (0,2) circle (0.8);
+ 	%axis
+	\draw (0,0) -- coordinate (x axis mid) (20,0);
+    \draw (0,0) -- coordinate (y axis mid) (0,10);
+	%ticks and labels      
+     		\draw [dotted](1pt,2) -- (-3pt,2) 
+     			node[anchor=east] {$1$}; 
+	\node[right=0.9cm] at (x axis mid) {$x$};
+	\node[above=0.6cm] at (y axis mid) {$y$};
+\end{tikzpicture} 
+& 
+\begin{tikzpicture}[y=.1cm, x=.1cm,font=\footnotesize]
+	\node (t1) at (-5,-4) {};
+	\node (t2) at (13,11) {};
+	\fill[line] (2,2) circle (0.8);
+ 	%axis
+	\draw (0,0) -- coordinate (x axis mid) (20,0);
+    \draw (0,0) -- coordinate (y axis mid) (0,10);
+	%ticks and labels      
+     		\draw [dotted](2,2) -- (-3pt,2) 
+     			node[anchor=east] { {\tiny $1$}}; 
+     		\draw [dotted](2,2) -- (2,-3pt) 
+     			node[anchor=north] { {\tiny $1$}}; 
+	\node[right=0.9cm] at (x axis mid) {$x$};
+	\node[above=0.6cm] at (y axis mid) {$y$};
+\end{tikzpicture} 
+\\ \hline
+2&
+\begin{tikzpicture}[y=.1cm, x=.1cm,font=\footnotesize]
+	\node (t1) at (-5,-4) {};
+	\node (t2) at (13,11) {};
+	\draw[line] (0,0) -- (2,2) -- cycle;
+ 	%axis
+	\draw (0,0) -- coordinate (x axis mid) (20,0);
+    \draw (0,0) -- coordinate (y axis mid) (0,10);
+	%ticks and labels      
+     		\draw [dotted](2,2) -- (-3pt,2) 
+     			node[anchor=east] { {\tiny $1$}}; 
+     		\draw [dotted](2,2) -- (2,-3pt) 
+     			node[anchor=north] { {\tiny $1$}}; 
+	\node[right=0.9cm] at (x axis mid) {$x$};
+	\node[above=0.6cm] at (y axis mid) {$y$};
+\end{tikzpicture} 
+& 
+\begin{tikzpicture}[y=.1cm, x=.1cm,font=\footnotesize]
+	\node (t1) at (-5,-4) {};
+	\node (t2) at (13,11) {};
+	\draw[line] (0,0) -- (12,12) -- cycle;
+ 	%axis
+	\draw (0,0) -- coordinate (x axis mid) (20,0);
+    \draw (0,0) -- coordinate (y axis mid) (0,10);
+	%ticks and labels      
+	\node[right=0.9cm] at (x axis mid) {$x$};
+	\node[above=0.6cm] at (y axis mid) {$y$};
+\end{tikzpicture} 
+&
+\begin{tikzpicture}[y=.1cm, x=.1cm,font=\footnotesize]
+	\node (t1) at (-5,-4) {};
+	\node (t2) at (13,11) {};
+	\draw[line] (0,2) -- (6,8) -- cycle;
+	\draw[line] (8,6) -- (14,12) -- cycle;
+	\fill[polyhedra] (0,2) -- (10,12) -- (14,12) -- (8,6) -- cycle;
+ 	%axis
+	\draw (0,0) -- coordinate (x axis mid) (20,0);
+    \draw (0,0) -- coordinate (y axis mid) (0,10);
+	%ticks and labels      
+	     	\draw [dotted](6,8) -- (-3pt,8) 
+     			node[anchor=east,yshift=2pt] { {\tiny $51$}}; 
+     		\draw [dotted] (8,6) -- (-3pt,6) 
+     			node[anchor=east, yshift=-2pt] { {\tiny $50$}}; 
+     		\draw [dotted](8,6) -- (8,-3pt) 
+     			node[anchor=north,xshift=2pt] { {\tiny $51$}}; 
+     		\draw [dotted] (6,8) -- (6,-3pt) 
+     			node[anchor=north,xshift=-2pt] { {\tiny $50$}};
+     		\draw [dotted](1pt,2) -- (-3pt,2) 
+     			node[anchor=east] { {\tiny $1$}}; 
+	\node[right=0.9cm] at (x axis mid) {$x$};
+	\node[above=0.6cm] at (y axis mid) {$y$};
+\end{tikzpicture} 
+&
+\begin{tikzpicture}[y=.1cm, x=.1cm,font=\footnotesize]
+	\node (t1) at (-5,-4) {};
+	\node (t2) at (13,11) {};
+	%\draw[line] (2,2) -- (6,8) -- cycle;
+	%\draw[line] (8,6) -- (14,12) -- cycle;
+	\fill[polyhedra] (2,2) -- (12,12) -- (16,12) -- (10,6) -- cycle;
+ 	%axis
+	\draw (0,0) -- coordinate (x axis mid) (20,0);
+    \draw (0,0) -- coordinate (y axis mid) (0,10);
+	%ticks and labels      
+     		\draw [dotted] (10,6) -- (-3pt,6) 
+     			node[anchor=east, yshift=-2pt] { {\tiny $50$}}; 
+     		\draw [dotted](10,6) -- (10,-3pt) 
+     			node[anchor=north,xshift=2pt] { {\tiny $52$}}; 
+     		\draw [dotted](2,2) -- (-3pt,2) 
+     			node[anchor=east] { {\tiny $1$}}; 
+     		\draw [dotted](2,2) -- (2,-3pt) 
+     			node[anchor=north] { {\tiny $1$}}; 
+	\node[right=0.9cm] at (x axis mid) {$x$};
+	\node[above=0.6cm] at (y axis mid) {$y$};
+\end{tikzpicture} 
+\\ \hline
+3& 
+\begin{tikzpicture}[y=.1cm, x=.1cm,font=\footnotesize]
+	\node (t1) at (-5,-4) {};
+	\node (t2) at (13,11) {};
+	%\draw[line] (2,2) -- (6,8) -- cycle;
+	%\draw[line] (8,6) -- (14,12) -- cycle;
+	\fill[polyhedra] (0,0) -- (12,12) -- (16,12) -- (10,6) -- cycle;
+ 	%axis
+	\draw (0,0) -- coordinate (x axis mid) (20,0);
+    \draw (0,0) -- coordinate (y axis mid) (0,10);
+	%ticks and labels      
+     		\draw [dotted] (10,6) -- (-3pt,6) 
+     			node[anchor=east, yshift=-2pt] { {\tiny $50$}}; 
+     		\draw [dotted](10,6) -- (10,-3pt) 
+     			node[anchor=north,xshift=2pt] { {\tiny $52$}}; 
+	\node[right=0.9cm] at (x axis mid) {$x$};
+	\node[above=0.6cm] at (y axis mid) {$y$};
+\end{tikzpicture} 
+& 
+\begin{tikzpicture}[y=.1cm, x=.1cm,font=\footnotesize]
+	\node (t1) at (-5,-4) {};
+	\node (t2) at (13,11) {};
+	\draw[line] (-4,-4) -- (12,12) -- cycle;
+	\fill[polyhedra] (-4,-4) -- (12,12) -- (20,12) -- (20,-4) -- cycle;
+ 	%axis
+	\draw (0,0) -- coordinate (x axis mid) (20,0);
+    \draw (0,0) -- coordinate (y axis mid) (0,10);
+	%ticks and labels      
+	\node[right=0.9cm] at (x axis mid) {$x$};
+	\node[above=0.6cm] at (y axis mid) {$y$};
+\end{tikzpicture} 
+& 
+\begin{tikzpicture}[y=.1cm, x=.1cm,font=\footnotesize]
+	\node (t1) at (-5,-4) {};
+	\node (t2) at (13,11) {};
+	\draw[line] (-4,-2) -- (10,12) -- cycle;
+	\fill[polyhedra] (-4,-4) -- (-4,-2) -- (10,12) -- (20,12) -- (20,-4) -- cycle;
+ 	%axis
+	\draw (0,0) -- coordinate (x axis mid) (20,0);
+    \draw (0,0) -- coordinate (y axis mid) (0,10);
+	%ticks and labels      
+     		\draw [dotted](1pt,2) -- (-3pt,2) 
+     			node[anchor=east] { {\tiny $1$}}; 
+	\node[right=0.9cm] at (x axis mid) {$x$};
+	\node[above=0.6cm] at (y axis mid) {$y$};
+\end{tikzpicture} 
+&
+\begin{tikzpicture}[y=.1cm, x=.1cm,font=\footnotesize]
+	\node (t1) at (-5,-4) {};
+	\node (t2) at (13,11) {};
+	\draw[line] (0,0) -- (12,12) -- cycle;
+	\fill[polyhedra] (0,0) -- (12,12) -- (20,12) -- (20,0) -- cycle;
+ 	%axis
+	\draw (0,0) -- coordinate (x axis mid) (20,0);
+    \draw (0,0) -- coordinate (y axis mid) (0,10);
+	%ticks and labels      
+	\node[right=0.9cm] at (x axis mid) {$x$};
+	\node[above=0.6cm] at (y axis mid) {$y$};
+\end{tikzpicture} 
+\\ \hline
+narrowing& 
+\begin{tikzpicture}[y=.1cm, x=.1cm,font=\footnotesize]
+	\node (t1) at (-5,-4) {};
+	\node (t2) at (13,11) {};
+	\draw[line] (0,0) -- (12,12) -- cycle;
+	\fill[polyhedra] (0,0) -- (12,12) -- (20,12) -- (20,0) -- cycle;
+ 	%axis
+	\draw (0,0) -- coordinate (x axis mid) (20,0);
+    \draw (0,0) -- coordinate (y axis mid) (0,10);
+	%ticks and labels      
+	\node[right=0.9cm] at (x axis mid) {$x$};
+	\node[above=0.6cm] at (y axis mid) {$y$};
+\end{tikzpicture} 
+& 
+\begin{tikzpicture}[y=.1cm, x=.1cm,font=\footnotesize]
+	\node (t1) at (-5,-4) {};
+	\node (t2) at (13,11) {};
+	\draw[line] (0,0) -- (12,12) -- cycle;
+	\fill[polyhedra] (0,0) -- (12,12) -- (20,12) -- (20,0) -- cycle;
+ 	%axis
+	\draw (0,0) -- coordinate (x axis mid) (20,0);
+    \draw (0,0) -- coordinate (y axis mid) (0,10);
+	%ticks and labels      
+	\node[right=0.9cm] at (x axis mid) {$x$};
+	\node[above=0.6cm] at (y axis mid) {$y$};
+\end{tikzpicture} 
+& 
+\begin{tikzpicture}[y=.1cm, x=.1cm,font=\footnotesize]
+	\node (t1) at (-5,-4) {};
+	\node (t2) at (13,11) {};
+	\draw[line] (0,2) -- (10,12) -- cycle;
+	\draw[line] (0,2) -- (10,-2) -- cycle;
+	\draw[line] (20,-2) -- (10,-2) -- cycle;
+	\fill[polyhedra] (0,2) -- (10,12) -- (20,12) -- (20,-2) --
+	(10,-2) -- cycle;
+ 	%axis
+	\draw (0,0) -- coordinate (x axis mid) (20,0);
+    \draw (0,0) -- coordinate (y axis mid) (0,10);
+	%ticks and labels      
+     		\draw [dotted](1pt,2) -- (-3pt,2) 
+     			node[anchor=east] { {\tiny $1$}}; 
+     		\draw [dotted](10,-2) -- (-3pt,-2) 
+     			node[anchor=east] { {\tiny $-1$}}; 
+     		\draw [dotted](10,-2) -- (10,1pt) 
+     			node[anchor=south] { {\tiny $51$}}; 
+	\node[right=0.9cm] at (x axis mid) {$x$};
+	\node[above=0.6cm] at (y axis mid) {$y$};
+\end{tikzpicture} 
+& 
+\begin{tikzpicture}[y=.1cm, x=.1cm,font=\footnotesize]
+	\node (t1) at (-5,-4) {};
+	\node (t2) at (13,11) {};
+	\draw[line] (2,2) -- (12,12) -- cycle;
+	\draw[line] (2,2) -- (7,0) -- cycle;
+	\fill[polyhedra] (2,2) -- (12,12) -- (20,12) -- (20,0) --
+	(7,0) -- cycle;
+ 	%axis
+	\draw (0,0) -- coordinate (x axis mid) (20,0);
+    \draw (0,0) -- coordinate (y axis mid) (0,10);
+	%ticks and labels      
+     		\draw [dotted](2,2) -- (-3pt,2) 
+     			node[anchor=east] { {\tiny $1$}}; 
+     		\draw [dotted](2,2) -- (2,-3pt) 
+     			node[anchor=north] { {\tiny $1$}}; 
+	\node[right=0.9cm] at (x axis mid) {$x$};
+	\node[above=0.6cm] at (y axis mid) {$y$};
+\end{tikzpicture} 
+\\ \hline
+\end{tabular}
+\end{figure}
+
 \part{Path Focusing technique} \label{pathfocusingpart}
@@
 	ignored, and will be treated later. we compute the iterations, and we obtain
-	an first invariant for this part of the graph. Then, we can do some
+	an first invariant for this part of the graph: at point $p_1$, $x =
+	y \wedge x \geq 0$.
+	Then, we can do some
 	narrowing steps in order to recover some precision, before adding new paths
-	in the graph.
+	in the graph. At point $p_1$ after narrowing, $x=y \wedge 0 \leq x \leq 51$.
 \item Step 2:
 	Now, we add the new paths into our graph if they have become possible. In
-	our case, the second path of the loop is now possible, so we add the path in
-	the graph and compute new iterations until convergence\dots
-\item Step 3: finally, the path from $p_4$ to $p_7$, so we add it into the graph and
-compute the invariant for $p_7$.
+	our case, the second path of the loop is now possible, because $x$ can be
+	equal to $51$, so we add the path in
+	the graph and compute new iterations until convergence\dots. After
+	convergence, at point $p_1$, $x + y - 102 \leq 0 \wedge x
+	\leq y$. We then apply a narrowing step and at $p_1$, 
+$x + y - 102 \leq 0 \wedge x \leq y \wedge y \geq 0$.
+\item Step 3: finally, the path from $p_4$ to $p_7$ becomes possible, so we add
+it into the graph and compute the invariant for $p_7$.
 \end{itemize}
+
+Finally, at $p_1$, the result we obtain is the best we can have using convex
+polyhedra (see Figure \ref{result}.b). This method allows to find the
+constraint $x + y - 102 \leq 0$, which is not the case in the classical abstract
+interpretation technique.
 
@@
 \caption{Our running example and its internal representation in SSA form}
+\label{IR}
 \end{figure}
 \FloatBarrier
+
+As we see if Figure \ref{IR}, LLVM IR is slightly different from
+our definition of control flow graph: it uses the notion of \emph{BasicBlock}.
+A \emph{BasicBlock} is a series of instructions that are necessarily executed
+one after the others, meaning that if the first intruction of a BasicBlock is
+executed, then the rest of the instructions are necessarily executed exactly
+once, in order.
+
+In our case, a control point will be a BasicBlock of the program.
 
@@
 
+	\subsection{Attachment to LLVM Internal Representation}
+
+	We attach only one abstract value to each BasicBlock of the IR. Indeed,
+	there would be a huge memory consumption if we would attach a polyhedra
+	after each instruction. A BasicBlock begins with all its $\Phi$-variables,
+	and ends with the other instructions. We choose to attach our abstract value
+	just after the $\Phi$-variables assignments. In this way, the
+	$\Phi$-variables of the associate BasicBlock will be part of the dimensions
+	of the polyhedron.
+
+\begin{figure}[!h]
+\centering
+\begin{tikzpicture}
+	\tikzstyle{bb} = [text width=6em, minimum height=12em]
+	\tikzstyle{part}=[draw, fill=blue!10, text width=5em, 
+    text centered, minimum height=2.5em, rounded corners, dashed]
+	\tikzstyle{abstract}=[draw, circle, fill=red!20, text width=5em, 
+    text centered, minimum height=2em,
+	decorate,
+    decoration={random steps,
+		segment length=6pt,
+		amplitude=4pt}
+	]
+    \node (bb) [bb]  {};
+    \node (Phi) [part] [above of=bb] {\small $\Phi$-variables};
+    \node (middle)  [below of=Phi] {};
+    \node (inst) [part] [below of=bb] {\small Instructions};
+	\node [abstract] (A) [left of=bb, xshift=-2cm] {Abstract \\ value};
+
+	\path [draw, ->] (A) -- (middle);
+	\path [draw, ->] (Phi) -- (inst);
+
+    \path (inst.south)+(0,-1) node (label) {BasicBlock};
+
+    \begin{pgfonlayer}{background}
+        % Compute a few helper coordinates
+        \path (Phi.north east)+(0.2,0.2) node (a) {};
+        \path (inst.south west)+(-0.2,-0.2) node (b) {};
+        \path[fill=yellow!20,rounded corners, draw=black!50, dashed]
+            (a) rectangle (b);
+
+        %\path (Phi.north west)+(-0.2,0.2) node (a) {};
+        %\path (Phi.south east)+(+0.2,-0.2) node (b) {};
+        %\path[fill=blue!10,rounded corners, draw=black!50, dashed]
+        %    (a) rectangle (b);
+        %\path (inst.north west)+(-0.2,0.2) node (a) {};
+        %\path (inst.south east)+(+0.2,-0.2) node (b) {};
+        %\path[fill=blue!10,rounded corners, draw=black!50, dashed]
+        %    (a) rectangle (b);
+       % \path (IR.north west)+(-0.2,0.2) node (a) {};
+       % \path (IR.south -| IR.east)+(+0.2,-0.2) node (b) {};
+       % \path[fill=blue!10,rounded corners, draw=black!50, dashed]
+       %     (a) rectangle (b);
+    \end{pgfonlayer}
+\end{tikzpicture}
+\end{figure}
+\FloatBarrier
+
 	\subsection{Dimensions of the Abstract values}
@@
 
-	\begin{definition}[Liveness\dots]
-	A variable $v$ is \emph{live\dots} at the control point $p$ iff:
+	\begin{definition}[Liveness by linearity]
+	A variable $v$ is \emph{live by linearity} at the control point $p$ iff:
 	\begin{itemize}
@@
 		variables $v_1, v_2, \dots, v_n$, such that $\exists i \in \{1,\dots,n\}, v = v_i$,
-		and $v'$ is live\dots in $p$.
+		and $v'$ is live by linearity in $p$.
 		\end{itemize}
@@
 
-	Finally, a variable is a dimension of $X_p$ iff it is live\dots and it is
-	not defined as a linear combination of program variables.
+	Finally, a variable is a dimension of $X_p$ iff it is live by linearity and
+	it is not defined as a linear combination of program variables.
+
+	There is a special case for $\Phi$-variables: in some cases, a
+	$\Phi$-variable can be considered as a linear combination of program
+	variables, when only a single incoming value is possible. This is the case
+	when all but one polyhedra associated to a predessor are at bottom.
+	For instance, let say point $p$ has $p_1, p_2, \dots, p_n$ as ordered 
+	predecessors. At point $p$, we have $x = \Phi(x_1, x_2,\dots,x_n)$.
+	If $\forall i \in \{2,\dots,n\}, X_i = \perp$, and $X_1 \neq \perp$, then we
+	can use the linear relation $x = x_1$ instead of the $\Phi$ function. In
+	practical cases, this is a way to reduce the size of the abstract values.
 

### 2011-05-06 report/report.tex
report

@@
 \institute{Grenoble-INP}
-\abstract{an abstract}
-\keywords{keywords}
+\abstract{
+	Program verification aims at statically discover properties on programs,
+	such as the values that can take the different variables during execution.
+	Abstract Interpretation is a technique that allows to compute an
+	approximation of the set of these values, since it is impossible to compute
+	the real set in general.
+	This report takes place in the many attempts to improve the precision of
+	static Analysis by Abstract Interpretation. It proposes a technique that
+	takes benefit of SMT-solving to obtain more precise results at reasonable
+	cost.
+	\\ \\  
+	La vérification de programme consiste à découvrir statiquement des
+	propriétés sur des programmes, comme l'ensemble des valeurs que peuvent
+	prendre les variables durant l'exécution. L'Interprétation Abstraite est une
+	technique permettant de calculer une approximation de cet ensemble, le
+	véritable ensemble étant impossible à calculer en général.
+	Ce report s'inscrit dans la lignée des travaux visant à améliorer la
+	précision de l'analyse par interprétation abstraite, et propose une
+	technique tirant parti du SMT-solving pour obtenir de meilleurs résultats
+	à un coût raisonnable.
+}
+\keywords{Static Program Analysis, Verification, Abstract Interpretation, Invariant
+Generation}
 \tutor{David Monniaux - Matthieu Moy}
@@
 
-\section*{Introduction}
+\part{Introduction}
 
@@
 Abstract Interpretation is sound but incomplete, since it always computes an
-upper-approximation of the set of possible states.
-
-Linear Relation Analysis approximates the set of possible states as a convex
-polyhedron where the dimensions are the numerical variables of the program.
+upper-approximation of the set of possible states. The challenge is then to be
+as precise as possible, so that we can prove properties about the program.
+
+Linear Relation Analysis over-approximates the set of possible states as a
+convex polyhedron where the dimensions are the numerical variables of the
+program.
 Instead of convex polyhedra, one could choose intervals, octagons, etc.
@@
 well as the implementation of another technique proposed in \cite{GopanR06},
-such as we can compare precision and cost of the these two methods.
+such that we can compare precision and cost of the these two methods.
 \end{itemize}
@@
 techniques into an analyzer, and the experimental results we obtained.
+Finally, Part \ref{future} introduces some directions to explore.
 
@@
 
-Abstract Interpretation \cite{CC77,CousotCousot92-1} is a general method to find approximate solutions of
+Abstract Interpretation \cite{CC77,CousotCousot92-1} is a general method for
+finding approximate solutions of
 fixpoint equations. This method is used for program analysis, since analyzing a
-program often comes down to solving fixpoint equations $\Phi(x) = x$.
+program often comes down to solving fixpoint equations, because of the loops and
+recursive procedures.
 
@@
 Abstract Interpretation method proposes to represent more efficiently the
-elements of this complex domain $C$ of concrete values, by choosing an simpler
-domain  $A$ called \emph{abstract domain}. The relation between these two
-domains is characterized by two functions, usually called 
-$\alpha\ :\ C \mapsto A$, $\gamma \ : \ A \mapsto C$, such as:
-
-$$\forall x \in C, \forall y \in A, \alpha(x) \leq_{A} y \Leftrightarrow x
-\leq_{C} \gamma(y) $$
-
-where $\leq_C$ and $\leq_A$ are the order relations on $C$ and $A$. Then, the
-fixpoint computation is done in the abstract domain, using the approximation of
-the function $\Phi$, i.e $\alpha(\Phi) = \alpha \circ \Phi \circ \gamma$, also
-noted $\Phi^\#$.
-We have the following result:
+elements of this complex domain $C$ of concrete values,
+by choosing an simpler domain  $A$ called \emph{abstract domain}.
+
+A concretization function $\gamma$ maps the elements of $A$ to elements of $C$:
+\begin{eqnarray*}
+\gamma: & A &\longrightarrow C \\
+		& x &\longmapsto \gamma(x)
+\end{eqnarray*}
+
+The concrete semantics is a monotonic function $\Phi: C \rightarrow C$.
+One can choose an abstract semantics $\Phi^\#: A \rightarrow A$ satisfying this
+condition:
+
+$$\forall x \in A, \Phi \circ \gamma (x) \subseteq_C \gamma \circ \Phi^\#(x)$$
+
+$\Phi^\#$ is then an abstraction of $\Phi$, that gives an upper
+approximation of $\gamma(x)$ when composed with $\gamma$.
 
 \begin{proposition}
-If $C$ is a complete lattice, and if $\Phi$ is increasing from $C$ to $C$, then
-$$\alpha(\lfp(\Phi)) \leq_A \lfp(\alpha(\Phi))$$
-where $\lfp(\Phi)$ denotes the least fixpoint of $\Phi$.
+Suppose $C$ is a complete lattice, and $\Phi$ is increasing from $C$ to $C$.
+Each $x \in A$, satisfying the condition $\Phi^\#(x) = x$, is an abstraction of
+the least fixpoint of $\Phi$.
 \end{proposition}
 
-In fact, one compute the least fixpoint of the function in the abstract
-domain, and the result of the computation also gives an upper approximation of
-the fixpoint in the concrete domain.
+Indeed, if $\Phi^\#(x) = x$, then $\Phi \circ \gamma (x) \subseteq \gamma(x)
+= \gamma \circ \Phi^\#(x)$. Abstract interpretation technique aims at computing
+one of these $x$. It computes the stationary limit of an ascending sequence
+$(x_i)_{i \geq 0}$ defined by the induction:
+
+$$\left[
+\begin{array}{lll}
+x_0 &=& \perp \\
+x_{n+1} &=& \Phi^\#(x_n)
+\end{array}
+\right.
+$$
+%The relation between these two
+%domains is characterized by two functions, usually called 
+%$\alpha\ :\ C \mapsto A$, $\gamma \ : \ A \mapsto C$, such as:
+%
+%$$\forall x \in C, \forall y \in A, \alpha(x) \leq_{A} y \Leftrightarrow x
+%\leq_{C} \gamma(y) $$
+%
+%where $\leq_C$ and $\leq_A$ are the order relations on $C$ and $A$. Then, the
+%fixpoint computation is done in the abstract domain, using the approximation of
+%the function $\Phi$, i.e $\alpha(\Phi) = \alpha \circ \Phi \circ \gamma$, also
+%noted $\Phi^\#$.
+%We have the following result:
+%
+%\begin{proposition}
+%If $C$ is a complete lattice, and if $\Phi$ is increasing from $C$ to $C$, then
+%$$\alpha(\lfp(\Phi)) \leq_A \lfp(\alpha(\Phi))$$
+%where $\lfp(\Phi)$ denotes the least fixpoint of $\Phi$.
+%\end{proposition}
+%
+%In fact, one compute the least fixpoint of the function in the abstract
+%domain, and the result of the computation also gives an upper approximation of
+%the fixpoint in the concrete domain.
 
@@
 this one should be finite, or of finite depth, meaning that there doesn't exist
-any sequence $(y_i)_{i \geq 0}$ of elements of the abstract domain $A$ such as
-$\forall i \geq 0,\ y_i <_A y_{i+1}$. Indeed, in case of a domain of infinite
+any sequence $(x_i)_{i \geq 0}$ of elements of the abstract domain $A$ such that
+$\forall i \geq 0,\ x_i <_A x_{i+1}$. Indeed, in case of a domain of infinite
 depth, the least fixpoint
@@
 \begin{itemize}
-\item $\forall y_1, y_2 \in A,\ y_1 \leq_A y_1 \nabla y_2$ and $y_2 \leq_A y_1
-\nabla y_2$. This guarantees the correctness of the result.
-\item For any increasing sequence $y_0 \leq_A y_1 \leq_A \dots$, the sequence
+\item $\forall x_1, x_2 \in A,\ x_1 \leq_A x_1 \nabla x_2$ and $x_2 \leq_A x_1
+\nabla x_2$. This guarantees the correctness of the result.
+\item For any increasing sequence $x_0 \leq_A x_1 \leq_A \dots$, the sequence
 defined by 
 $$\system{
-y'_0 & = &  y_0 \\
-y'_{i+1} & = &  y'_i\ \nabla\ y_{i+1},\ \  \forall i > 0
+x'_0 & = &  x_0 \\
+x'_{i+1} & = &  x'_i\ \nabla\ x_{i+1},\ \  \forall i > 0
 }$$
@@
 
-The least fixpoint of a function $\Phi^\#$ is noted $\overline{y}$ and is equal
-to $\displaystyle \lim_{i \geq 0}\ y_i$, where $y_0 = \perp$ (the least element
-of $A$) and $y_{i+1} = \Phi^\# (y_i)$. Instead of computing this least fixpoint,
+The least fixpoint of a function $\Phi^\#$ is noted $\overline{x}$ and is equal
+to $\displaystyle \lim_{i \geq 0}\ x_i$, where $x_0 = \perp$ (the least element
+of $A$) and $x_{i+1} = \Phi^\# (x_i)$. Instead of computing this least fixpoint,
 one compute an upper approximation of it, by computing the following ascending
@@
 $$\system{
-y'_0 &=& \perp \\
-y'_{i+1} &=& y'_i\ \nabla\ \Phi^\#(y'_i),\ \ \forall i > 0
+x'_0 &=& \perp \\
+x'_{i+1} &=& x'_i\ \nabla\ \Phi^\#(x'_i),\ \ \forall i > 0
 }$$
 
-which converges towards $\tilde{y}$, where $\tilde{y} \geq_A \overline{y}$.
-
-$\tilde{y}$ is a correct upper approximation of the least fixpoint of $\Phi^\#$,
+which converges towards $\tilde{x}$, where $\tilde{x} \geq_A \overline{x}$.
+
+$\tilde{x}$ is a correct upper approximation of the least fixpoint of $\Phi^\#$,
 and the classical technique is to regain some precision lost by the widening
@@
 $$\system{
-y''_0 &=& \tilde{y} \\
-y''_{i+1} &=& \Phi^\#(y''_i), \ \ \forall i > 0
+x''_0 &=& \tilde{x} \\
+x''_{i+1} &=& \Phi^\#(x''_i), \ \ \forall i > 0
 }$$
@@
 Each element of this descending sequence is still an upper approximation of the
-least fixpoint $\overline{y}$. So, this sequence often allows to find
+least fixpoint $\overline{x}$. So, this sequence often allows to find
 approximate least fixpoints that are more precise than the one obtained after
@@
 $\Q^n$. Since the exact operation $\tau$ may not be expressed in the abstract
-domain, we abstract it into $\tau^\#$ such as $\forall X \in D, \tau(X)
+domain, we abstract it into $\tau^\#$ such that $\forall X \in D, \tau(X)
 \subseteq \tau^\#(X)$.
@@
 apply the widening operator at each control point $p \in P$. It is sufficient to
-apply it on a subset of $P$, called $P_w$, such as removing the nodes of $P_w$
+apply it on a subset of $P$, called $P_w$, such that removing the nodes of $P_w$
 disconnects every cycles. For instance, one could choose $P_w$ as the set of the
@@
 
-Lookahead widening \cite{GopanR06} is a technique that aims to isolate the
-different loop phases during the analysis, in order to obtain more precise
-results.
+Lookahead widening \cite{GopanR06} is a technique that isolates the
+different loop phases during the analysis, and stabilizes each of these loop
+phases before proceeding to the next. One only consider paths that are feasible
+at the first iteration, and forget the others for a while. The iterations end
+when the analysis of these paths has converged. After that, one can compute some
+narrowing iterations. Then, paths that
+become feasible are reinserted into the control flow graph, and we redo
+iterations.
 
@@
 
+In the classical abstract interpretation technique, narrowing iterations are
+computed at the end of the analysis, when all the fixpoint computation has
+converged.
+The interest of this method is to apply narrowing iterations right after each
+path computation. This allows to recover precision before analyzing new paths,
+and then to be more precise.
+
  \section{Path Focusing}
@@
 	Let $(P,E)$ be the control flow graph of the program.
-	First, one choose a set of points $P_W \subseteq P$, such that the graph
+	First, one choose a set of points $P_W \subseteq P$, so that the graph
 	obtained after removing these points has no cycle. The widening operator
@@
 
-	The multigraph is then the graph such as:
+	The multigraph is then the graph such that:
 	\begin{itemize}
@@
 	\item for each path $p_1 \rightarrow p_2 \rightarrow \dots \rightarrow p_k$
-	in the control flow graph, such as $p_1 \in P_R, p_k \in P_R$,
+	in the control flow graph, such that $p_1 \in P_R, p_k \in P_R$,
 	there is a transition in the multigraph from $p_1$ to $p_k$ with the
@@
 having the same semantic.}
+\label{multigraphtransformGopan}
 \end{figure}
@@
 	which path in the control flow graph is selected, by giving a model: an
-	assignments of the Boolean variables such that the formula is true.
+	assignments of the Boolean variables so that the formula is true.
 	Among these Boolean variables, some of them are associated to a transition
@@
 \caption{Example (cont'd)}
+\label{multigraphGopan}
 \end{figure}
@@
 	abstract value $X_{p_i}$
-	\item we search a path arriving in a successor $p_j$ of $p_i$, such as the
+	\item we search a path arriving in a successor $p_j$ of $p_i$, such that the
 	vector of the numerical variables $x_j'$ is not yet included in the current
@@
 	widening. Widening will be applied next time to guarantee the convergence.
+
+	When the self loop is accelerable \cite{Gon07}, instead of doing a
+	widening/narrowing sequence, one can directly accelerate the loop and find
+	its transitive closure.
 
@@
 
+	\subsection{Example}
+
+	We apply the \emph{Path Focusing} technique to our running example from
+	Figures \ref{runningexample}, \ref{multigraphtransformGopan} and
+	\ref{multigraphGopan}.
+	The control points in $P_R$ are $p_0$, $p_1$ and $p_7$. One computes
+	iteratively the associate abstract values $X_0$, $X_1$ and $X_7$.
+
+\begin{itemize}
+\item Step 0: Initially, $X_0 = \top, X_1 = \perp, X_7 = \perp$, and $A =
+\{p_0\}$.
+\item Step 1: We start in $p_0$. Is there a path starting in $p_0$, that arrives
+in a state in $P_R$ and makes its abstract value grow? The SMT-solver answers
+\emph{yes}, and gives the path $p_0 \rightarrow p_1$. We update $X_1$:
+it is the polyhedron $x=y=0$. We update $A = \{p_1\}$. 
+We do another SMT query: is there another path
+starting in $p_0$ that makes an abstract value grow? Answer is no.
+\item Step 2: We now work on $p_1$. Is there a path starting in $p_1$ that
+arrives in a successor of $p_1$, and makes its abstract value grow? Answer is
+yes, and the model given by the SMT-solver is the path $p_1 \rightarrow p_2
+\rightarrow p_4 \rightarrow p_5 \rightarrow p_6 \rightarrow p_1$. Actually, this
+is the first phase of the loop. 
+The image of $x=y=0$ by this transition is $x=y=1$. The new $X_1$ is then $x=y
+\wedge 0\leq x \leq 1$. The path is a self loop, but is seen for the first time,
+so we do not apply widening yet. So, we redo an SMT-query, that gives us exactly
+the same path. Then, we can do a widening/narrowing sequence. 
+After widening, we have $x=y \wedge x \geq 0$, and narrowing gives $X_1 = (x=y)
+\wedge (0 \leq x \leq 51)$.
+\item Step 3: We still are in $p_1$: we ask for another focus path, and the
+SMT-solver finds the path $p_1 \rightarrow p_3
+\rightarrow p_4 \rightarrow p_5 \rightarrow p_6 \rightarrow p_1$. This is the
+second phase of the loop, which is possible since $(x,y)=(51,51) \in X_1$.
+Again, this is a self-loop, but we do not widen yet. The
+image of $X_1$ by the transition is the single point $(52,50)$. $X_1$ is then
+the convex hull of this point with the old $X_1$. A new SMT query gives the same
+path, and we can now apply widening/narrowing. We obtain
+$X_1 = (x \leq y) \wedge (102 - x - y \geq 0)$ after widening, and narrowing
+gives $X_1 = (x \leq y) \wedge (102 - x - y \geq 0) \wedge y \geq 0$.
+We can see that $X_1$ is the smallest polyhedron containing the possible values
+for $(x,y)$ (see Figure \ref{result}.b).
+\item Step 4: Again, we ask the SMT-solver for a new focus path, and it returns
+$p_1 \rightarrow p_3 \rightarrow p_4 \rightarrow p_7$. The image of the
+polyhedra $X_1$ after this path transformation is the single point $y=-1 \wedge
+x = 102$. We can update $X_7 = {(102,-1)}$. Path focusing technique gives us the
+precise result of this program. After that, the SMT-solver can't find any new
+path, so the analysis terminates.
+\end{itemize}
+
 	\subsection{Disjunctive invariants}
@@
 		\item There is a variable $v'$, defined as a linear combination of other
-		variables $v_1, v_2, \dots, v_n$, such that $\exists i \in \{1,\dots,n\}, v = v_i$,
+		variables $v_1, v_2, \dots, v_n$, so that $\exists i \in \{1,\dots,n\}, v = v_i$,
 		and $v'$ is live by linearity in $p$.
@@
 
-	\subsection{SMT-solving}
+	\section{SMT-solving}
 
@@
 	SMT formula, and to extract a model when the formula is satisfiable, i.e an
-	assignement of the variables such that the formula is \emph{true}. For doing
+	assignement of the variables so that the formula is \emph{true}. For doing
 	this, our implementation has an interface with Microsoft Z3 \cite{MouraB08},
@@
 
+	To be completed\dots
+
+\part{Future Work}
+\label{future}
+\emph{Path Focusing} technique allows to work on control flow graphs with a huge
+number of paths, since they are computed only when needed. With this in mind,
+one could think of several improvements using path focusing, by encoding into
+the control flow graph new informations as paths, such as arithmetic overflows,
+alias analysis, \dots
+In this part, we explain how to encode such kind of informations into implicit
+paths of the CFG.
+
+	\section{Arithmetic Overflows}
+
+Some operations such as additions, soustractions\dots may overflow if their
+result is not in the type's definition interval. For instance, an
+\emph{int} variable can only take values in $[-2^{31}, 2^{31} -1]$. Then, there
+are two kind of overflows. Suppose $N=2^{31}$:
+\begin{itemize}
+\item The result $r$ of the operation may be greater than $N-1$. In this case,
+the result we obtain is actually $r-2N$.
+\item $r$ may be lesser than $-N$. Then, the result is $r+2N$.
+\end{itemize}
+
+The idea would be to distinguish the correct behaviour and the two kind of
+overflow in the control flow graph, each behaviour corresponding to a specific
+path (see Figure \ref{overflow}).
+
+\begin{figure}[!h]
+\centering
+\begin{tikzpicture}[->,>=stealth',auto,node distance=3cm,
+                    semithick,font=\footnotesize]
+	\node[state] (n0) {};
+	\node[state] (n1) [below of=n0] {};
+
+	\path [transition] 
+		(n0) edge [out=-140,in=140]  node [xshift=-0.2mm] 
+		{$\begin{array}{l} x+y < -N\ /\\ r=x+y+2N\end{array}$} (n1)
+		(n0) edge [in=170,out=190,distance=2cm]  node [left]
+		{$\begin{array}{l} x+y > N-1\ / \\ r=x+y-2N\end{array}$} (n1)
+		(n0) edge [in=10,out=-10,distance=2.5cm]  node [right] 
+		{$\begin{array}{l} -N \leq x+y \leq N-1\ /\\ r=x+y\end{array}$} (n1)
+		;
+\end{tikzpicture}
+\caption{How to treat overflows of the addition $x+y$}
+\label{overflow}
+\end{figure}
+
+This results in a huge blowup in the number of paths, but hopefully a lot of
+them will never be possible, so they will never be computed by our algorithm.
+
+	\section{Alias analysis}
+
+One could use the same idea as for overflows, in order to treat aliases.
+Assume a pointer $p$ that possibly points to variables $x,y,z$ or $t$. For each
+operation involving $*p$, one distinguishes these different cases by
+creating a specific path for each of them. 
+If the pointers are manipulated in a tricky way, one considers they may
+point to each variable of the program. Again, there is an exponential blowup in
+the number of paths, but these are treated implicitly.
+
+\begin{figure}[!h]
+\centering
+\begin{tikzpicture}[->,>=stealth',auto,node distance=3cm,
+                    semithick,font=\footnotesize]
+	\node[state] (n0) {};
+	\node[state] (n1) [below of=n0] {};
+
+	\path [transition] 
+		(n0) edge [in=170,out=190,distance=2.5cm]  node [left] 
+		{$\begin{array}{l}p=\&x/ \\ x++\end{array}$} (n1);
+	\path [transition] 
+		(n0) edge  [in=120, out=-120] node [left,xshift=0.2cm] 
+		{$\begin{array}{l}p=\&y/ \\ y++\end{array}$} (n1);
+	\path [transition] 
+		(n0) edge [in=60,out=-60]  node  [right,xshift=-0.2cm]
+		{$\begin{array}{l}p=\&z/ \\ z++\end{array}$} (n1);
+	\path [transition] 
+		(n0) edge [in=10,out=-10,distance=2.5cm]  node [right] 
+		{$\begin{array}{l}p=\&t/ \\ t++\end{array}$} (n1);
+\end{tikzpicture}
+\caption{How to treat pointers: $(*p)++$}
+\label{pointers}
+\end{figure}
+
+\part{Conclusion}
+
 	To Be Completed
 
-  \section*{Conclusion}
-
-	To Be Completed
-
   \appendix
 
-\bibliographystyle{plain}
+\bibliographystyle{alpha}
   \bibliography{report}

### 2011-05-06 report/report.tex
report

@@
-\documentclass[a4paper,english,titlepage,11pt]{article}
+\documentclass[a4paper,english,titlepage,11pt]{report}
 
@@
 
-\part{Introduction}
+\chapter{Introduction}
 
@@
 
-\part{Abstract Interpretation: state of the art}
+\chapter{Abstract Interpretation: state of the art}
 \label{stateoftheartpart}
@@
 values for $i$ at the next step are in $[0, 100]$. This is still an invariant,
-but much better.
+but much better. We can repeat such iterations as long as the set narrows, and
+we always obtain a correct over-approximation.
+In practice, a single narrowing iteration is enough to recover the precision.
 
@@
 
-\part{Path Focusing technique} \label{pathfocusingpart}
+\chapter{Path Focusing technique} \label{pathfocusingpart}
 
@@
 
-	The essence of the method is to
+	The principle of the method is to
 	choose an integer $\delta \in \{1,..,m\}$, and a mapping function
@@
 \FloatBarrier
-\part{Implementation}\label{implementationpart}
+\chapter{Implementation}\label{implementationpart}
 
@@
 \begin{figure}[!h]
-\begin{minipage}[c]{.45\linewidth}
+\begin{minipage}[c]{.35\linewidth}
+{\small
 \begin{C}
@@
 \end{C}
+}
 \end{minipage}
-\begin{minipage}[c]{.45\linewidth}
-\includegraphics[width=8.5cm]{"images/cfg_gopan"}
+\begin{minipage}[c]{.69\linewidth}
+%\includegraphics[width=8.5cm]{"images/cfg_gopan"}
+
+{\tiny
+% \begin{tikzpicture}[anchor=mid,>=latex',line join=bevel,]
+\begin{tikzpicture}[>=latex',line join=bevel,scale=0.5]
+  \pgfsetlinewidth{1bp}
+%%
+\begin{scope}
+  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
+  \pgfsetstrokecolor{strokecol}
+  \draw (273bp,13bp) node {CFG for 'main' function};
+\end{scope}
+  \pgfsetcolor{black}
+  % Edge: Node0x9828da8 -> Node0x9828dd8
+  \draw [->] (356.55bp,400.82bp) .. controls (333.72bp,390.36bp) and (308.41bp,378.75bp)  .. (274.21bp,363.07bp);
+  % Edge: Node0x9828b18 -> Node0x9828d48
+  \draw [->] (310bp,694.72bp) .. controls (310bp,686.41bp) and (310bp,677bp)  .. (310bp,657.09bp);
+  % Edge: Node0x9828dd8 -> Node0x9828e08
+  \draw [->] (70bp,241bp) .. controls (70bp,229.11bp) and (70bp,216.19bp)  .. (70bp,194.35bp);
+  % Edge: Node0x9828d48 -> Node0x9828da8
+  \draw [->] (456bp,530bp) .. controls (475.15bp,530bp) and (473.05bp,509.87bp)  .. (461.87bp,479.21bp);
+  % Edge: Node0x9828e08 -> Node0x9828e68
+  \draw [->] (70bp,133.91bp) .. controls (70bp,122.63bp) and (70bp,109.61bp)  .. (70bp,87.235bp);
+  % Edge: Node0x9828d78 -> Node0x9828dd8
+  \draw [->] (155.32bp,400.82bp) .. controls (154.06bp,392.13bp) and (152.69bp,382.65bp)  .. (149.87bp,363.2bp);
+  % Edge: Node0x9828d48 -> Node0x9828d78
+  \draw [->] (164bp,530bp) .. controls (145.24bp,530bp) and (143.41bp,510.23bp)  .. (148.69bp,479.21bp);
+  % Edge: Node0x9828e38 -> Node0x9828d48
+  \draw [->] (285.99bp,203.34bp) .. controls (287.87bp,215.1bp) and (289.73bp,228.04bp)  .. (291bp,240bp) .. controls (300.66bp,331.07bp) and (305.54bp,436.13bp)  .. (308.25bp,516.72bp);
+  % Edge: Node0x9828dd8 -> Node0x9828e38
+  %\draw [->] (284bp,254bp) .. controls (293.39bp,254bp) and (292.88bp,234.23bp)  .. (287.99bp,203.21bp);
+  \draw [->] (210bp,240bp) -- (240.99bp,203.21bp);
+  % Node: Node0x9828d48
+\begin{scope}
+  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
+  \pgfsetstrokecolor{strokecol}
+  \draw (165bp,517bp) -- (165bp,657bp) -- (455bp,657bp) -- (455bp,517bp) -- cycle;
+  \draw (165bp,543bp) -- (455bp,543bp);
+  \draw (310bp,517bp) -- (310bp,543bp);
+  \draw (173bp,639bp) node[right] {\%1:};
+  \draw (173bp,605bp) node[right] { \%x.0 = phi i32 [ 0, \%0 ], [ \%11, \%10 ]};
+  \draw (173bp,587bp) node[right] { \%y.1 = phi i32 [ 0, \%0 ], [ \%y.0, \%10 ]};
+  \draw (173bp,569bp) node[right] { \%2 = icmp sle i32 \%x.0, 50};
+  \draw (173bp,551bp) node[right] { br i1 \%2, label \%3, label \%5};
+  \draw (238bp,525bp) node {T};
+  \draw (383bp,525bp) node {F};
+\end{scope}
+  % Node: Node0x9828b18
+\begin{scope}
+  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
+  \pgfsetstrokecolor{strokecol}
+  \draw (261bp,695bp) -- (261bp,755bp) -- (360bp,755bp) -- (360bp,695bp) -- cycle;
+  \draw (269bp,737bp) node[right] {\%0:};
+  \draw (269bp,703bp) node[right] { br label \%1};
+\end{scope}
+  % Node: Node0x9828e68
+\begin{scope}
+  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
+  \pgfsetstrokecolor{strokecol}
+  \draw (30bp,27bp) -- (30bp,87bp) -- (111bp,87bp) -- (111bp,27bp) -- cycle;
+  \draw (38bp,69bp) node[right] {\%12:};
+  \draw (38bp,35bp) node[right] { ret i32 0};
+\end{scope}
+  % Node: Node0x9828e38
+\begin{scope}
+  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
+  \pgfsetstrokecolor{strokecol}
+  \draw (172bp,125bp) -- (172bp,203bp) -- (387bp,203bp) -- (387bp,125bp) -- cycle;
+  \draw (180bp,185bp) node[right] {\%10:};
+  \draw (180bp,151bp) node[right] { \%11 = add nsw i32 \%x.0, 1};
+  \draw (180bp,133bp) node[right] { br label \%1};
+\end{scope}
+  % Node: Node0x9828da8
+\begin{scope}
+  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
+  \pgfsetstrokecolor{strokecol}
+  \draw (338bp,401bp) -- (338bp,479bp) -- (547bp,479bp) -- (547bp,401bp) -- cycle;
+  \draw (346bp,461bp) node[right] {\%5:};
+  \draw (346bp,427bp) node[right] { \%6 = add nsw i32 \%y.1, -1};
+  \draw (346bp,409bp) node[right] { br label \%7};
+\end{scope}
+  % Node: Node0x9828d78
+\begin{scope}
+  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
+  \pgfsetstrokecolor{strokecol}
+  \draw (59bp,401bp) -- (59bp,479bp) -- (263bp,479bp) -- (263bp,401bp) -- cycle;
+  \draw (67bp,461bp) node[right] {\%3:};
+  \draw (67bp,427bp) node[right] { \%4 = add nsw i32 \%y.1, 1};
+  \draw (67bp,409bp) node[right] { br label \%7};
+\end{scope}
+  % Node: Node0x9828dd8
+\begin{scope}
+  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
+  \pgfsetstrokecolor{strokecol}
+  \draw (-1bp,241bp) -- (-1bp,363bp) -- (283bp,363bp) -- (283bp,241bp) -- cycle;
+  \draw (-1bp,267bp) -- (283bp,267bp);
+  \draw (142bp,241bp) -- (142bp,267bp);
+  \draw (8bp,345bp) node[right] {\%7:};
+  \draw (8bp,311bp) node[right] { \%y.0 = phi i32 [ \%4, \%3 ], [ \%6, \%5 ]};
+  \draw (8bp,293bp) node[right] { \%8 = icmp slt i32 \%y.0, 0};
+  \draw (8bp,275bp) node[right] { br i1 \%8, label \%9, label \%10};
+  \draw (71bp,249bp) node {T};
+  \draw (212bp,249bp) node {F};
+\end{scope}
+  % Node: Node0x9828e08
+\begin{scope}
+  \definecolor{strokecol}{rgb}{0.0,0.0,0.0};
+  \pgfsetstrokecolor{strokecol}
+  \draw (16bp,134bp) -- (16bp,194bp) -- (124bp,194bp) -- (124bp,134bp) -- cycle;
+  \draw (24bp,176bp) node[right] {\%9:};
+  \draw (24bp,142bp) node[right] { br label \%12};
+\end{scope}
+%
+\end{tikzpicture}
+}
+
 \end{minipage}
@@
 our definition of control flow graph: it uses the notion of \emph{BasicBlock}.
-A \emph{BasicBlock} is a series of instructions that are necessarily executed
+A \emph{BasicBlock} is a sequence of instructions that are necessarily executed
 one after the others, meaning that if the first intruction of a BasicBlock is
@@
 	abstract value at the beginning of the analysis.
-	This lack of precision may be recovered by using disjunctive invariants: one
+	This lack of precision may be limited by using disjunctive invariants: one
 	could attach two polyhedra to the control point: the first one with the
@@
 	We attach only one abstract value to each BasicBlock of the IR. Indeed,
-	there would be a huge memory consumption if we would attach a polyhedra
+	there would be a huge memory consumption if we would attach a polyhedron
 	after each instruction. A BasicBlock begins with all its $\Phi$-variables,
@@
 
-\part{Future Work}
+\chapter{Future Work}
 \label{future}
@@
 
-\part{Conclusion}
+	\section{Combining Lookahead Widening and Path Focusing}
+
+	Path Focusing technique works on the expanded multigraph.
+	Its limitation is that it does narrowing
+	sequences during the iterations only for self-loops of this multigraph.
+
+	Lookahead Widening works on the classical control flow graph, and stabilizes
+	temporarily the fixpoint analysis on a part of this graph, forgetting paths
+	that are not feasible at the first iteration.
+
+	One could try to combine these two techniques, and apply Lookahead Widening
+	on the expanded multigraph, where the different paths are maintained
+	implicit thanks to Path Focusing technique. The advantage of this approach
+	compared to the classical Path Focusing is that narrowing iterations could
+	be computed on loops that are not self-loops.
+
+\chapter{Conclusion}
 

### 2011-05-09 report/report.tex
minor changes in the report

@@
 
-The least fixpoint of a function $\Phi^\#$ is noted $\overline{x}$ and is equal
-to $\displaystyle \lim_{i \geq 0}\ x_i$, where $x_0 = \perp$ (the least element
-of $A$) and $x_{i+1} = \Phi^\# (x_i)$. Instead of computing this least fixpoint,
+The least fixpoint of a function $\Phi^\#$ is noted $\overline{x}$.
+Instead of computing this least fixpoint,
 one compute an upper approximation of it, by computing the following ascending
@@
 	Let $(P,E)$ be the control flow graph of the program.
-	First, one choose a set of points $P_W \subseteq P$, so that the graph
+	First, one chooses a set of widening points $P_W \subseteq P$, so that the graph
 	obtained after removing these points has no cycle. The widening operator
@@
 	\cite{Bou92}.
-	Then, one choose another set $P_R \subseteq P$ of nodes, satisfying the
+	Then, one chooses another set $P_R \subseteq P$ of nodes, satisfying the
 	following properties:

### 2011-05-11 report/report.tex
removed all debug printing for performance evaluation + report update

@@
 \usepackage{algpseudocode}
+\usepackage{multirow}
 
@@
 
+\subsection{Transformation passes}
+
+LLVM internal representation can be applied transformation passes that modifies
+parts of it. These passes are mostly used for optimisation purpose. In our
+analyzer, we apply some already existing passes before running our 
+abstract interpretation pass. Among the passes we run, one could cite:
+\begin{itemize}
+\item the \emph{PromoteMemoryToRegister} Pass, that transforms the internal
+representation by promoting the memory variables to Static Single Assignment
+(SSA) registers. This pass transform the control flow graph such that all the
+variables are in SSA form, and removes most of the \emph{Load/Store} operations
+into memory. It is useful since our analyzer doesn't have a memory model, and
+then loses lots of precision in the case of \emph{Load/Store} instructions.
+\item the \emph{LoopSimplify} and \emph{LoopInfo} Passes, that transform all loops into a canonical
+format, and compute some information related to loops (for instance, the set of
+BasicBlocks that are parts of the loop, etc.).
+\item the \emph{LowerSwitch} Pass, that transform a switch instruction into its
+corresponding \emph{if-then-else} instructions. After running this pass, there
+is no more \emph{Switch} instruction in the control flow graph, so the
+implementation of the static analysis is easier.
+\end{itemize}
+
 \subsection{Drawbacks}
@@
 	\item its value is available at $p$, meaning that there is a path from the
-	point defining the variable and $p$.
+	point defining the variable to $p$.
 	\item its value might be used in the future, meaning that there exist a path
@@
 
+	\section{Example}
+
+	In our running example (Figures \ref{runningexample},
+	\ref{multigraphtransformGopan} and \ref{multigraphGopan}), the analyzer
+	outputs the following result for the basicblock $1$:
+
+\begin{C}
+RESULT FOR BASICBLOCK: -------------------
+; <label>:1                                 ; preds = %10, %0
+  %x.0 = phi i32 [ 0, %0 ], [ %11, %10 ]
+  %y.0 = phi i32 [ 0, %0 ], [ %y.1, %10 ]
+  %2 = icmp sle i32 %x.0, 50
+  br i1 %2, label %3, label %5
+-----
+environment: dim = (2,0), count = 77
+ 0: x.0
+ 1: y.0
+polyhedron of dim (2,0)
+array of constraints of size 3
+ 0: -x.0 - y.0 + 102 >= 0
+ 1: y.0 >= 0
+ 2: x.0 - y.0 >= 0
+\end{C}
+
+First, it shows the BasicBlock we work on, and its associated abstract value:
+\begin{itemize}
+\item The environment of the abstract value is its dimensions. Here, $x.0$ and
+$y.0$ are the dimensions of the polyhedron.
+\item The array of constraints defining the polyhedron. The three constraints
+here define the polyhedra seen in Figure \ref{result}.b.
+\end{itemize}
+
  \section{Experiments}
@@
 	To be completed\dots
+
+\begin{tabular}{|c|c|c|c||c|c|} \hline
+
+			& \multicolumn{3}{c||}{Path Focusing} & \multicolumn{2}{c|}{Lookahead
+			Widening} \\ \hline
+Benchmark	& \multicolumn{2}{c|}{time} & result	& time	& result	\\
+\cline{2-3}	& z3 & Yices				& & & \\
+\hline
+
+\end{tabular}
 
@@
 
-	To Be Completed
-
-  \appendix
+	In this report, we proposed a new approach to guide the iterations of the
+	analyzer, in order to distinguish the different paths and to treat them
+	independently. The loss of precision due to widening and control-flow merges
+	is then reduced. This technique allows us, for instance, to do narrowing
+	steps right after widening, or to use acceleration techniques when a path
+	actually is a simple loop.
+
+	We implemented this technique into a small analyzer, 
+	and it appears that in lots of cases, 
+	the precision of the result is improved 
+	compared to the classical abstract interpretation technique.
+	We compared it with \emph{Lookahead Widening}, and our technique seems to
+	perform better in the case of loops containing control-flow merges.
 
@@
   \bibliography{report}
+
+  \appendix
+
+\begin{figure}
+\begin{C}
+(and (= bd_0 false)
+        (= t_0_1 bs_0)
+        (or (not bd_0) true)
+        (= b_5 t_1_5)
+        (= x_%6_ (+ x_y.0_ -1))
+        (= t_5_7 b_5)
+        (or (not b_5) true)
+        (= b_3 t_1_3)
+        (= x_%4_ (+ x_y.0_ 1))
+        (= t_3_7 b_3)
+        (or (not b_3) true)
+        (= b_7
+           (or t_5_7 t_3_7))
+        (= t_7_9 (and b_7 (< x_y.1_ 0)))
+        (= t_7_10
+           (and b_7 (not (< x_y.1_ 0))))
+        (or (not b_7)
+            (= x_y.1_ (ite t_3_7 x_%4_ x_%6_)))
+        (= b_10 t_7_10)
+        (= x_%11_ (+ x_x.0_ 1))
+        (= t_10_1 b_10)
+        (or (not b_10) true)
+        (= bd_1
+           (or t_10_1 t_0_1))
+        (= t_1_3 (and bs_1 (<= x_x.0_ 50)))
+        (= t_1_5
+           (and bs_1 (not (<= x_x.0_ 50))))
+        (or (not bd_1)
+            (and (= x'_x.0_ (ite t_0_1 0 x_%11_))
+                 (= x'_y.0_ (ite t_0_1 0 x_y.1_))))
+        (= b_9 t_7_9)
+        (= t_9_12 b_9)
+        (or (not b_9) true)
+        (= bd_12 t_9_12)
+        (or (not bd_12) true))
+\end{C}
+\caption{Expression of $\rho$ associated to the program in Figures
+\ref{runningexample}, \ref{multigraphtransformGopan} and
+\ref{multigraphGopan}.}
+\end{figure}
+
 \end{document}

### 2011-05-11 src/AI.cc

@@ -297,7 +297,9 @@ void AI::computeTransform (Node * n, std::list<BasicBlock*> path, Abstract &Xtem
 	std::list<std::vector<ap_tcons1_array_t*>*>::iterator i, e;
 	for (i = constraints.begin(), e = constraints.end(); i!=e; ++i) {
 		if ((*i)->size() == 1) {
-				tcons1_array_print((*i)->front());
+				DEBUG(
+					tcons1_array_print((*i)->front());
+				);
 				Xtemp.meet_tcons_array((*i)->front());
 				// creating the associated lincons for a future widening with
 				// threshold
@@ -368,10 +370,12 @@ void AI::computeNode(Node * n) {
 
 	while (true) {
 		is_computed[n] = true;
-		Out->changeColor(raw_ostream::RED,true);
-		*Out << "--------------- NEW SMT SOLVE -------------------------\n";
-		Out->resetColor();
-		Out->flush();
+		DEBUG(
+			Out->changeColor(raw_ostream::RED,true);
+			*Out << "--------------- NEW SMT SOLVE -------------------------\n";
+			Out->resetColor();
+			Out->flush();
+		);
 		LSMT->push_context();
 		// creating the SMT formula we want to check
 		SMT_expr smtexpr = LSMT->createSMTformula(n->bb,false);
@@ -471,10 +475,12 @@ void AI::narrowNode(Node * n) {
 	while (true) {
 		is_computed[n] = true;
 
-		Out->changeColor(raw_ostream::RED,true);
-		*Out << "--------------- NEW SMT SOLVE -------------------------\n";
-		Out->resetColor();
-		Out->flush();
+		DEBUG(
+			Out->changeColor(raw_ostream::RED,true);
+			*Out << "--------------- NEW SMT SOLVE -------------------------\n";
+			Out->resetColor();
+			Out->flush();
+		);
 		LSMT->push_context();
 		// creating the SMT formula we want to check
 		SMT_expr smtexpr = LSMT->createSMTformula(n->bb,true);
@@ -825,8 +831,9 @@ void AI::visitPHINode (PHINode &I){
 
 	ap_texpr_rtype_t ap_type;
 	if (get_ap_type((Value*)&I, ap_type)) return;
-	*Out << I << "\n";
-
+	DEBUG(
+		*Out << I << "\n";
+	);
 	for (unsigned i = 0; i < I.getNumIncomingValues(); i++) {
[…]

### 2011-05-11 src/yices.cc

@@ -15,6 +15,7 @@
 
 #include "yices.h"
 #include "Analyzer.h"
+#include "Debug.h"
 
 using namespace llvm;
 
@@ -207,11 +208,13 @@ bool yices::SMT_check(SMT_expr a, std::set<std::string> * true_booleans) {
 	yices_assert(ctx,(yices_expr)a);
 	//*Out << "\n";
 	if (yices_check(ctx) == l_true) {
+		DEBUG(
 		*Out << "sat\n";
+		);
 		yices_var_decl_iterator it = yices_create_var_decl_iterator(ctx);
 		yices_model m              = yices_get_model(ctx);
 		
-
+		DEBUG(
 		//Save position of current standard output
 		fpos_t pos;
 		fgetpos(stdout, &pos);
@@ -233,6 +236,7 @@ bool yices::SMT_check(SMT_expr a, std::set<std::string> * true_booleans) {
 			*Out << c;
 
 		*Out << "\n";
+		);
 		
 		while (yices_iterator_has_next(it)) {
 			yices_var_decl d         = yices_iterator_next(it);
@@ -253,7 +257,9 @@ bool yices::SMT_check(SMT_expr a, std::set<std::string> * true_booleans) {
 		}
 		yices_del_iterator(it);
 	} else {
+		DEBUG(
 		*Out << "unsat\n";
+		);
 		return false;
 	}
 	return true;

### 2011-05-11 src/z3_manager.cc

@@ -7,6 +7,7 @@
 
 #include "z3_manager.h"
 #include "Analyzer.h"
+#include "Debug.h"
 
 z3_manager::z3_manager() {
 	Z3_config config = Z3_mk_config();
@@ -218,13 +219,19 @@ bool z3_manager::SMT_check(SMT_expr a, std::set<std::string> * true_booleans){
 
 	switch (result) {
 		case Z3_L_FALSE:
-			*Out << "unsat\n";
+			DEBUG(
+				*Out << "unsat\n";
+			);
 			return false;
 		case Z3_L_UNDEF:
+			DEBUG(
 			*Out << "unknown\n";
+			);
 			return false;
 		case Z3_L_TRUE:
+			DEBUG(
 			*Out << "sat\n";
+			);
 			unsigned n = Z3_get_model_num_constants(ctx,m);
 			for (unsigned i = 0; i < n; i++) {
 				Z3_func_decl decl = Z3_get_model_constant(ctx,m,i);
@@ -232,41 +239,58 @@ bool z3_manager::SMT_check(SMT_expr a, std::set<std::string> * true_booleans){
 				Z3_eval_func_decl (ctx,m,decl,&v);
 				Z3_symbol symbol = Z3_get_decl_name(ctx,decl);
 				std::string name (Z3_get_symbol_string (ctx,symbol));
+				DEBUG(
 				*Out << name << " = ";
-
+				);
 				Z3_sort_kind sort = Z3_get_sort_kind(ctx,Z3_get_sort(ctx,v));
 
 				switch (sort) {
 					case Z3_BOOL_SORT: 
 						switch (Z3_get_bool_value(ctx,v)) {
 							case Z3_L_FALSE:
-								*Out << "false\n";
+								DEBUG(
+									*Out << "false\n";
+								);
 								break;
 							case Z3_L_UNDEF:
+								DEBUG(
 								*Out << "undef\n";
+								);
 								break;
 							case Z3_L_TRUE:
+								DEBUG(
 								*Out << "true\n";
+								);
 								true_booleans->insert(name);
 								break;
 						}
 						break;
 					case Z3_INT_SORT:
+						DEBUG(
 						int i;
 						Z3_get_numeral_int (ctx,v,&i);
 						*Out << i << "\n";
+						);
 						break;
 					case Z3_REAL_SORT:
+						DEBUG(
 						*Out << "real value\n";
+						);
 						break;
 					case Z3_BV_SORT:
+						DEBUG(
[…]

### 2011-05-11 report/report.tex
report update + added some benchmarks

@@
 
-	To be completed\dots
-
-\begin{tabular}{|c|c|c|c||c|c|} \hline
+	In this section, we evaluate our \emph{Lookahead Widening} and \emph{Path
+	Focusing} implementations. We can compare their execution time in some
+	practical cases, and the precision of their results. Table \ref{experiments}
+	gives the results of the analyzer in various examples, some of them taken
+	from the recent literature \cite{CGGMP05,GopanR06}.
+
+	In all experiments, the set $P_R$ is chosen equal to $P_W \cup P_{ret} \cup
+	P_{init}$,
+	where $P_W$ is the set of widening points. $P_{init}$ is the set of starting
+	points of the program, and $P_{ret}$ is the set of points that have no
+	successors.
+
+\begin{table}[!h]
+\centering \scriptsize
+\begin{tabular}{|c|c|c|l||c|l|} \hline
 
 			& \multicolumn{3}{c||}{Path Focusing} & \multicolumn{2}{c|}{Lookahead
-			Widening} \\ \hline
+			Widening} \\ \cline{2-6}
 Benchmark	& \multicolumn{2}{c|}{time} & result	& time	& result	\\
-\cline{2-3}	& z3 & Yices				& & & \\
-\hline
+\cline{2-3}	& Z3 & Yices				& & & \\
+\hline \hline
+
+Gopan.c & 0.088s & 0.080s & \begin{tabular}{l}
+-x.0 - y.0 + 102 >= 0 \\
+y.0 >= 0              \\
+x.0 - y.0 >= 0        
+\end{tabular} & 0.032 & \begin{tabular}{l}
+-x.0 - y.0 + 102 >= 0 \\
+y.0 >= 0              \\
+x.0 - y.0 >= 0        
+\end{tabular} \\ \hline
+
+Beyer.c & 0.084s & 0.076s & \begin{tabular}{l}
+ -x.0  + 100 >= 0 \\
+ y.0 - 50 >= 0    \\
+ -x.0 + y.0 >= 0  
+\end{tabular} & 0.032 & \begin{tabular}{l}
+-x.0  + 100 >= 0 \\
+ y.0 - 50 >= 0 \\
+-x.0 + y.0 >= 0
+\end{tabular} \\ \hline
+
+Boustr.c & 0.100s & 0.088s & \begin{tabular}{l}
+ -2x.0 + d.0 + 1999 >= 0 \\
+ -2x.0 + 3d.0 + 2001 >= 0 \\
+ -d.0 + 1 >= 0
+\end{tabular} & 0.064 & \begin{tabular}{l}
+ -1996x.0 + 4991d.0 + 1998995 >= 0\\
+ -2x.0 + d.0 + 1999 >= 0 \\
+ -d.0 + 1 >= 0 \\
+\end{tabular} \\ \hline
+
+ex8.cpp & 0.516s & 0.408s & \begin{tabular}{l}
+ -y.0 - x.0 + 1 >= 0\\
+ -y.0 + x.0 + 1 >= 0\\
+ y.0 - x.0 + 1 >= 0\\
+ y.0 + x.0 + 1 >= 0
+\end{tabular} & 0.208 & \begin{tabular}{l}
+-x.0 + 1 >= 0 \\
+x.0 + 1 >= 0 
+\end{tabular} \\ \hline
+
+ex9.cpp & 0.084s & 0.082s & \begin{tabular}{l}
+ r.0 = 0 \\
+-i.0 + 100 >= 0 \\
+i.0 >= 0
+\end{tabular} & 0.052 & \begin{tabular}{l}
+-i.0 + 100 >= 0\\
+r.0 >= 0 \\
+i.0 - r.0 >= 0 \\
+24i.0 - 25r.0 + 75 >= 0
+\end{tabular} \\ \hline
+
+test1.c & 0.064s & 0.064s & \begin{tabular}{l}
+2j.0 + i.0 - 21 = 0 \\
+-j.0 + 10 >= 0 \\
+j.0 - 6 >= 0
+\end{tabular} & 0.016 & \begin{tabular}{l}
+2j.0 + i.0 - 21 = 0 \\
+-j.0 + 10 >= 0 \\
+j.0 - 6 >= 0
+\end{tabular} \\ \hline
 
 \end{tabular}
+\caption{Results of \emph{Path Focusing} and \emph{Lookahead Widening}
+techniques applied to various benchmarks}
+\label{experiments}
+\end{table}
+
+\emph{Path Focusing} appears to always perform similarly or better than 
+\emph{Lookahead Widening}. Indeed, it is more precise in the case of programs
+containing control-flow merges inside of loops, because of the distinction
+between all the paths of the loop.
 

### 2011-05-11 ex/ex8.cc

@@ -1,0 +1,82 @@
+#include<iostream>
+
+int main() {
+	int x;
+	int y;
+	int d;
+	x = 1;
+	y = 0;
+
+	while(1) {
+		std::cout << x << " " << y << "\n";
+		if (x > 0) {
+		} else if (x < 0) {
+		} else {
+			if (y > -1) {
+			} else if (y < -1) {
+			} else {
+				x=1;
+			}
+			if (y > 1) {
+			} else if (y < 1) {
+			} else {
+				x=-1;
+			}
+		}
+
+		if (y > 0) {
+		} else if (y < 0) {
+		} else {
+			if (x > -1) {
+			} else if (x < -1) {
+			} else {
+				y=-1;
+			}
+			if (x > 1) {
+			} else if (x < 1) {
+			} else {
+				y=1;
+			}
+		}
+
+
+		if (y > 1) {
+		} else if (y < 1) {
+		} else {
+			if (x > 1) {
+			} else if (x < 1) {
+			} else {
+				x=0;
+			}
+		}
+
+		if (y > -1) {
+		} else if (y < -1) {
+		} else {
+			if (x > -1) {
+			} else if (x < -1) {
+			} else {
+				x=0;
+			}
+		}
+		if (x > 1) {
+		} else if (x < 1) {
+		} else {
+			if (y > -1) {
+			} else if (y < -1) {
+			} else {
+				y=0;
+			}
+		}
+		if (x > -1) {
+		} else if (x < -1) {
+		} else {
+			if (y > 1) {
+			} else if (y < 1) {
+			} else {
+				y=0;
+			}
+		}
+	}
+	return 1;
+}

### 2011-05-13 report/report.tex
report + added counters in the code

@@
 	cost.
-	\\ \\  
+}
+\resume{  
 	La vérification de programme consiste à découvrir statiquement des
@@
 }
-\keywords{Static Program Analysis, Verification, Abstract Interpretation, Invariant
-Generation}
+\keywords{Static Program Analysis, Verification, Abstract Interpretation,
+Invariant Generation, Path Focusing, SMT-Solving}
+\motscles{Analyse Statique de Programmes, Vérification, Interprétation
+Abstraite, Génération d'invariants, Découverte de Chemins, SMT-Solving}
 \tutor{David Monniaux - Matthieu Moy}
@@
 	\cite{GulwaniZ10} proposes a technique to compute transitive closure of a
-	transition system, i.e compute the invariants for a loop having its semantic
-	expressed by the transition system.
-
-	A transition system for a control point $p$ is a DNF formula where each
+	loop, i.e compute the invariants for a loop having its semantic
+	expressed by a transition system. The definition of a transition system is
+	the following:
+
+	\begin{definition}
+	A transition system for a control point $p$ is a DNF formula (i.e a
+	disjunction of conjunctions), where each
 	disjunct is the semantic of a path from $p$ to itself. It is of the form
 	$$\bigvee_{1 \leq i \leq n} \tau_{p,i}$$ where $n$ is the number of
-	paths, and $\tau_{p,i}$ is the semantic of the $i$-th path from $p$ to itself.
+	paths, and $\tau_{p,i}$ is the semantic of the $i$-th path from $p$ to
+	itself.
+	\end{definition}
 
@@
 function $\sigma$ has to be defined on $\{1,..,m\} \times \N$. We could also
-compute this number of paths at the beginning of the algorithm.
+easily compute this number of paths when computing the $\rho$ formula.
 
@@
 
+\subsection{Diseq comparisons}
+
+The set of elements $x$ satisfying $x \neq N$ is not convex: it is the union of
+two intervals: $]-\infty, N[\  \cup\  ]N,+\infty[$. Since we use convex polyhedra to
+represent the set of possible states of our program, we loose information when
+the analysis comes across a \emph{diseq} operation.
+One classical approach would be to compute a disjunctive invariant, by attaching
+two different polyhedron to the BasicBlock. However, we prefered the following
+technique: we transform all the operations of the form $x \neq y$ into
+$x < y \vee x > y$, meaning that we separate the cases \emph{lesser than} and
+\emph{greater than} into two distinct paths.
+
+\begin{minipage}[c]{.29\linewidth}
+\begin{C}
+if (x != y) {
+	...
+}
+\end{C}
+\end{minipage} $\Rightarrow$
+\begin{minipage}[c]{.69\linewidth}
+\centering
+\begin{C}
+	if (x > y) {
+	} else if (x < y) {
+	} else {
+		...
+	}
+\end{C}
+\end{minipage}
+
+To do so, one could implement an LLVM pass that transform the internal
+representation according to this principle. However, this pass has not been
+implemented in our analyzer, and this transformation has to be done manually.
+
 	\section{SMT-solving}
@@
 	an argument we give as parameter to the analyzer. In this way, we can
-	compare the effectiveness of both SMT-solvers.
+	compare the efficiency of both SMT-solvers.
+
+	\section{Limitations}
+
+	Our analyzer could be extended in many ways. Indeed, it is intra-procedural,
+	meaning that it does not consider that a variable may change after a
+	function call. In addition, the return value of a function call is
+	considered unconstrained. This limitation can be addressed by simply
+	inlining the function call, provided the function is not recursive.
+	There have been research to find efficient interprocedural analysis
+	techniques \cite{Bourdoncle90,Bou92}.
+
+	Additionnaly, our analyzer has no memory model:
+\begin{itemize}
+\item \emph{Store} instructions, that store a value in the memory, has no effect
+in the analysis.
+\item \emph{Load} instructions, that get a value $v$ in the memory and assign it
+to a variable $x$, doesn't provide properties about the value $v$. Then, our
+analyzer considers $x$ can take all values of its type.
+\end{itemize}
+In the case of local variables, the transformation of the internal
+representation into SSA form removes the \emph{Load/Store} instructions.
+The use of global variables instead of local ones will induce huge
+loss of precision, since there will still be \emph{Load/Store} instructions in
+the IA.
 
@@
 
-	In all experiments, the set $P_R$ is chosen equal to $P_W \cup P_{ret} \cup
+	In all experiments, the set $P_R$ is chosen equals to $P_W \cup P_{ret} \cup
 	P_{init}$,
@@
 
+ex6.c & 0.104s & $\infty$ & \begin{tabular}{l}
+-y.0 + 51 >= 0 \\
+y.0 >= 0 \\
+x.0 - 51y.0 + 2550 >= 0 \\
+x.0 >= 0
+\end{tabular} & 0.032 & \begin{tabular}{l}
+y.0 >= 0
+\end{tabular} \\ \hline
+
+ex8c.c & 191m50.935s & $\infty$ & \begin{tabular}{l}
+c.0 + a.0 = 0 \\
+b.0 + d.0 = 0\\
+-b.0 - c.0 + 1 >= 0\\
+-b.0 + c.0 + 1 >= 0\\
+b.0 - c.0 + 1 >= 0\\
+b.0 + c.0 + 1 >= 0\\
+\end{tabular} & 0.928 & \begin{tabular}{l}
+-b.0 + 1 >= 0\\
+-b.0 + a.0 + 1 >= 0\\
+-c.0 + 1 >= 0\\
+-c.0 + d.0 + 1 >= 0\\
+-a.0 + 1 >= 0\\
+-d.0 + 1 >= 0\\
+a.0 + 1 >= 0\\
+c.0 - d.0 + 1 >= 0\\
+c.0 + d.0 + 1 >= 0\\
+b.0 - a.0 + 1 >= 0\\
+b.0 + 1 >= 0
+\end{tabular} \\ \hline
+
 \end{tabular}
@@
 between all the paths of the loop.
+
+To be completed\dots
+
+\begin{table}[!h]
+\begin{tabular}{|l||c|c|c|c|c|c|c|} \hline
+Benchmark  & Gopan.c & Beyer.c & Boustr.c & ex8.cpp & ex9.cpp & test1.c & ex6.c\\ \hline \hline
+paths (total) & 5 & 4 & 10 & 75626 & 8 & 3 & 5 \\ \hline
+paths (computed) & 5 & 4 & 4 & 4 & 4 & 3 & 4\\ \hline
+iterations  & 6 & 5 & 5 & 4 & 4 & 3 & 4\\ \hline
+\end{tabular}
+\caption{Number of path in the expanded multigraph, number of path the
+algorithm actually computes, and total number of iterations}
+\end{table}
+
+To be completed\dots
 
@@
 
+	Validation of softwares has become an essential domain in industry,
+	especially those dealing with embedded systems, where a bug in the program
+	can endanger the safety of people, or induce huge costs to fix. These
+	domains constitute a privileged application field for formal verification
+	techniques.
+	Abstract Interpretation is a common used technique to do static analysis of
+	programs. It aims to compute iteratively an over-approximation of the set of
+	possible states of the program, and allows to prove properties such as
+	``In this program, $x$ is always between 1 and 10'', ``The program never
+	divides by zero'', etc.
+
 	In this report, we proposed a new approach to guide the iterations of the
-	analyzer, in order to distinguish the different paths and to treat them
+	analyzer, in order to distinguish the different paths of the program
+	and to treat them
 	independently. The loss of precision due to widening and control-flow merges
@@
 
-	We implemented this technique into a small analyzer, 
-	and it appears that in lots of cases, 
+	Since distinguishing all the paths leads to an exponential blowup in time
+	and memory, we do compute a path only when needed, otherwise the path is
+	kept implicit. In this way, memory consumption is preserved, and the
+	computation time is reduced, but still potentially exponential.
+	In order to find which path we should compute, we use SMT-solving to find
+	paths that makes the abstract interpretation analysis progress.
+
+	We implemented this technique into a small analyzer dealing with C and C++
+	programs, and run it on several
+	benchmarks.
+	It appears that in lots of cases, 
 	the precision of the result is improved 
@@
 	perform better in the case of loops containing control-flow merges.
+
+	Our technique performs better because of the distinction between
+	paths. Then, encoding some properties, such as arithmetic overflows or
+	aliases, into distinct paths in the control flow graph is one direction to
+	explore.
+
+	Our analyzer is able to deal with small- and middle-sized programs, but lose
+	precision in the case of programs with pointers or several functions.
+	It could be improved in many ways, in order to support
+	for instance interprocedural analysis. A memory model could be added to
+	handle programs with pointers and \emph{load/store} instructions.
+	One could also try to compute disjunctive invariants using the technique
+	described in this report.
+	To be completed\dots
 

### 2011-05-13 ex/ex8.cc

@@ -1,4 +1,4 @@
-#include<iostream>
+//#include<iostream>
 
 int main() {
 	int x;
@@ -8,7 +8,7 @@ int main() {
 	y = 0;
 
 	while(1) {
-		std::cout << x << " " << y << "\n";
+	//	std::cout << x << " " << y << "\n";
 		if (x > 0) {
 		} else if (x < 0) {
 		} else {

### 2011-05-13 src/AI.cc

@@ -29,6 +29,7 @@ using namespace llvm;
 
 char AI::ID = 0;
 
+
 static RegisterPass<AI> X("AI", "Abstract Interpretation Pass", false, true);
 
 const char * AI::getPassName() const {
@@ -81,6 +82,9 @@ bool AI::runOnModule(Module &M) {
 		}
 		delete it->second;
 	}
+	*Out << "Number of iterations: " << n_iterations << "\n";
+	*Out << "Number of paths: " << n_totalpaths << "\n";
+	*Out << "Number of paths computed: " << n_paths << "\n";
 	return 0;
 }
 
@@ -143,9 +147,11 @@ void AI::computeFunction(Function * F) {
 	// get the information about live variables from the LiveValues pass
 	LV = &(getAnalysis<Live>(*F));
 	LI = &(getAnalysis<LoopInfo>(*F));
-	
+
+	*Out << "Computing Pr and Rho...";
 	LSMT->getPr(*F);
 	LSMT->getRho(*F);
+	*Out << "OK\n";
 	// add all function's arguments into the environment of the first bb
 	for (Function::arg_iterator a = F->arg_begin(), e = F->arg_end(); a != e; ++a) {
 		Argument * arg = a;
@@ -359,23 +365,23 @@ void AI::computeNode(Node * n) {
 		return;
 	}
 	
-	DEBUG (
+	//DEBUG (
 		Out->changeColor(raw_ostream::GREEN,true);
 		*Out << "#######################################################\n";
 		*Out << "Computing node: " << b << "\n";
 		Out->resetColor();
 		*Out << *b << "\n";
-	);
+	//);
 
 
 	while (true) {
 		is_computed[n] = true;
-		DEBUG(
+		//DEBUG(
 			Out->changeColor(raw_ostream::RED,true);
 			*Out << "--------------- NEW SMT SOLVE -------------------------\n";
 			Out->resetColor();
 			Out->flush();
-		);
+		//);
 		LSMT->push_context();
 		// creating the SMT formula we want to check
 		SMT_expr smtexpr = LSMT->createSMTformula(n->bb,false);
@@ -395,6 +401,9 @@ void AI::computeNode(Node * n) {
 		);
 	
 		Succ = Nodes[path.back()];
+
+		n_iterations++;
+		if (!pathtree->exist(path)) n_paths++;
 
 		// computing the image of the abstract value by the path's tranformation
 		Xtemp = new Abstract(n->X);

### 2011-05-13 src/Analyzer.cc

@@ -7,6 +7,7 @@
 
 #include "Execute.h" 
 #include "Analyzer.h" 
+#include "Debug.h" 
 
 SMT_Solver manager;
 llvm::raw_ostream *Out;
@@ -35,6 +36,9 @@ int main(int argc, char* argv[]) {
 	bool debug = false;
 
 	manager = Z3_MANAGER;
+	n_totalpaths = 0;
+	n_paths = 0;
+	n_iterations = 0;
 
 	 while ((o = getopt(argc, argv, "hdi:o:y")) != -1) {
         switch (o) {

### 2011-05-13 src/Debug.h

@@ -7,4 +7,8 @@
 #define DEBUG(X) do { } while (0);
 #endif
 
+extern int n_paths;
+extern int n_iterations;
+extern int n_totalpaths;
+
 #endif

### 2011-05-13 src/SMT.cc

@@ -19,6 +19,7 @@
 #include "yices.h"
 #include "Expr.h"
 #include "apron.h"
+#include "Debug.h"
 
 using namespace std;
 
@@ -388,21 +389,24 @@ void SMT::computeRhoRec(Function &F,
 
 	bool first = (visited->count(b) > 0);
 	visited->insert(b);
-	
+
 
 	if (!Pr[&F]->count(b) || newPr) {
 		// we recursively construct Rho, starting from the predecessors
 		BasicBlock * pred;
 		for (pred_iterator p = pred_begin(b), E = pred_end(b); p != E; ++p) {
 			pred = *p;
-			computeRhoRec(F,pred,dest,false,visited);
-			if (!Pr[&F]->count(pred))
+			//computeRhoRec(F,pred,dest,false,visited);
+			if (!Pr[&F]->count(pred)) {
+				computeRhoRec(F,pred,dest,false,visited);
 				primed[b].insert(primed[pred].begin(),primed[pred].end());
-			else {
+			} else {
 				Pr_succ[pred].insert(dest);
 				Pr_pred[dest].insert(pred);
 			}
 		}
+	} else {
+		n_totalpaths++;
 	}
 
 	if (first) return;

### 2011-05-16 report/report.tex
Bug fix: there was a bug in the case of OR, AND,... instructions for integer variables

@@
 \usepackage{multirow}
+\usepackage[toc,page]{appendix}
 
@@
 well as the implementation of another technique proposed in \cite{GopanR06},
-such that we can compare precision and cost of the these two methods.
+such that we can compare precision and cost of these two methods.
 \end{itemize}
@@
 Static analysis aims to discover some properties about programs. In this
-example, one would to compute the set of possible values for the variable $x$
+example, one would compute the set of possible values for the variable $x$
 during execution.
@@
 
-This computation is a fixpoint computation: one updates the different sets until
+This computation is a fixpoint computation: the different sets are updated until
 we find no more elements. These sets are strictly increasing (we always add new
@@
 $$
+
+where $\perp$ is the least element of $A$.
 %The relation between these two
@@
 Since the domain of convex polyhedra is of infinite depth, a widening operator
-is defined to ensure convergence of the technique. There have been various
-work proposing a definition of the widening operator in LRA
+is defined to ensure convergence of the technique. 
+Intuitively, the polyhedron $P \nabla Q$ is obtained after 
+removing from the system of linear equations defining $P$ all the inequalities
+that are not satisfied by $Q$.
+For instance, if $P = \{(x,y)\ |\ 0 \leq x \leq y \leq 1 \}$ and $Q = 
+\{(x,y)\ |\ 0 \leq x \leq y \leq 2 \}$, then the result of the widening operator
+will be $P \nabla Q = \{(x,y)\ |\ 0 \leq x \leq y \}$.
+
+There have been various
+work proposing a definition or a refinment of the widening operator in LRA
 \cite{CH78,Hal79,HPR97,BlanchetCousotEtAl_PLDI03}, in order to fight
-against the loss of precision induced by widening.
+against loss of precision:
+\begin{itemize}
+\item \emph{Delayed widening}: instead of applying windening at the first
+iteration of the analysis, one apply it after a number $n$ of iterations. During
+the $n-1$ first iterations, only a convex hull is computed. In some cases,
+delaying the widening operation is more precise. The algorithm still terminates,
+since widening is applied within a finite time after $n$ iterations. Most of the
+cases, $n$ can be choosen equals to $2$.
+\item \emph{Widening with threshold (or ``up to'')}: In some cases, instead of
+brutally applying the widening operator, a fixed set of linear inequalities $T$
+can be chosen, and the widening operator $\nabla_T$ is defined as follows:
+$P \nabla_T Q$ is the intersection of $P \nabla Q$ with all the inequalities in
+$T$ satisfied by both $P$ and $Q$.
+
+For instance, consider the program:
+\begin{C}
+while (x <= 50) {
+	x++;
+	y++;
+}
+\end{C}
+If $P = \{(x,y)\ |\ 0 \leq x \leq y \leq 1 \}$, after one iteration of the loop,
+we have $Q = \{(x,y)\ |\ 0 \leq x \leq y \leq 2 \}$. Instead of a classic
+widening (which gives the result $P \nabla Q = \{(x,y)\ |\ 0 \leq x \leq y \}$),
+a widening with the threshold $x \leq 50$ gives 
+$P \nabla_T Q = \{(x,y)\ |\ 0 \leq x \leq y \leq 50\}$. In such cases, widening
+with threshold helps to find better invariants.
+\end{itemize}
 
@@
 apply it on a subset of $P$, called $P_w$, such that removing the nodes of $P_w$
-disconnects every cycles. For instance, one could choose $P_w$ as the set of the
+disconnects every cycles. For instance, $P_w$ can be chosen as the set of the
 heads of loops.
@@
 compute disjunctive invariants \cite{GulwaniZ10} to limit this upper approximation: at each
-program point, one compute a union of convex polyhedra instead of a single
+program point, a union of convex polyhedra is computed instead of a single
 one.
@@
 \end{C}
-The analysis starts with $i=0$. After one iteration, one obtain $i \in
-[0,1]$, \dots. After a widening operation, the result becomes $i \in [0,
+The analysis starts with $i=0$. After one iteration, $i \in
+[0,1]$, then $i\in [0,2]$ after the second iteration\dots. 
+After a widening operation, the result becomes $i \in [0,
 +\infty]$, which is an invariant, but very imprecise. We can do a new iteration,
@@
 to expand the control flow graph. For instance, instead of considering a graph
-with a sequence of $n$ \emph{if-then-else} between point $p_1$ and $p_n$, one could remove the $n$ merge node
-of these \emph{if-then-else}, each of them having two incoming edges, and adding 
-$2^n$ edges from $p_1$ to $p_n$, corresponding to the different paths through
+with a sequence of $n$ \emph{if-then-else} between point $p_1$ and $p_n$, 
+the $n$ merge node
+of these \emph{if-then-else} could be removed, each of them having two incoming
+edges, and 
+$2^n$ edges from $p_1$ to $p_n$ are added, corresponding to the different paths through
 the \emph{if-then-else} statements. At the end, $p_n$ is the only point with
@@
 aims to compute the exact effect of the loop when possible \cite{Gon07,GH06}. For instance, for
-the following program, one could directly compute an invariant for the loop:
+the following program, an invariant could be directly computed for the loop:
 
@@
 at the first iteration, and forget the others for a while. The iterations end
-when the analysis of these paths has converged. After that, one can compute some
-narrowing iterations. Then, paths that
+when the analysis of these paths has converged. After that, some
+narrowing iterations can be computed. Then, paths that
 become feasible are reinserted into the control flow graph, and we redo
@@
 	\cite{Bou92}.
-	Then, one chooses another set $P_R \subseteq P$ of nodes, satisfying the
+	Then, another set $P_R \subseteq P$ of nodes is chosen, satisfying the
 	following properties:
@@
 	which is a logic formula with Boolean variables, and elements of a certain
-	theory $T$. Depending on the program, one can choose different theories:
+	theory $T$. Depending on the program, different theories may be used:
 	\begin{itemize}
@@
 If the point where the $\Phi$-variable $x$ is defined is in
-$P_R$, one need to rename $x$ into $x'$:
+$P_R$, $x$ needs to be renamed into $x'$:
 \begin{itemize}
@@
 	paths starting from this point, until there is no more path to be treated.
-	At the end of the analysis, one could perform some narrowing iterations in
+	At the end of the analysis, some narrowing iterations can be performed in
 	order to recover some precision.
@@
 	self loops of the multigraph. Indeed, instead of simply apply the widening
-	operator, one could compute some narrowing steps to regain precision before
-	focusing on another path. This is the same principle as the one we find in
-	\emph{Lookahead widening} \cite{GopanR06}.
+	operator,  some narrowing steps could be performed to regain precision
+	before focusing on another path. This is the same principle as the one we
+	find in \emph{Lookahead widening} \cite{GopanR06}.
 
@@
 	computes a narrowing step in order to recover precision. In some cases,
-	instead of performing widening/narrowing iterations, one could use
-	acceleration techniques to find even more precise invariants \cite{GH06}.
-
-	In order for this refinement to be precise, one needs to delay the widening
-	operator when we consider a self-loop for the first time. Then, our
+	instead of performing widening/narrowing iterations, 
+	acceleration techniques could be used 
+	to find even more precise invariants \cite{GH06}.
+
+	In order for this refinement to be precise, the widening
+	operator needs to be delayed
+	when we consider a self-loop for the first time. Then, our
 	algorithm uses a set $U$ of paths that have already been treated. If the
@@
 	When the self loop is accelerable \cite{Gon07}, instead of doing a
-	widening/narrowing sequence, one can directly accelerate the loop and find
+	widening/narrowing sequence,  acceleration directly finds
 	its transitive closure.
@@
 	\ref{multigraphGopan}.
-	The control points in $P_R$ are $p_0$, $p_1$ and $p_7$. One computes
+	The control points in $P_R$ are $p_0$, $p_1$ and $p_7$. 
+	The $Ŕho$ formula associated to this program is depicted in
+	Figure \ref{rhoformula}.
+	One computes
 	iteratively the associate abstract values $X_0$, $X_1$ and $X_7$.
@@
 some of the join operations may be useless, in the sense that the value of 
-$X_{p,\sigma(j,i)}$ may not change. Here, one could use the same technique as
-Path Focusing to avoid these drawbacks: one can use SMT-solving to find the
-paths $\tau_{p,i}$.
+$X_{p,\sigma(j,i)}$ may not change. Here, the same technique as
+Path Focusing is achievable to avoid these drawbacks:  SMT-solving is used
+to find the paths $\tau_{p,i}$.
 
@@
 easily compute this number of paths when computing the $\rho$ formula.
+
+	\section{Efficiency comparison: example}
+
+In this subsection, we propose an example showing that \emph{Path Focusing}
+technique performs better than \emph{Lookahead Widening} in the case of loops
+with intermediate control-flow merges.
+
+We consider the program depicted in figure \ref{ex9}.
+Here, we would like to show that the returned value $r$ is zero.
+
+\begin{figure}[!h]
+\begin{minipage}[c]{.49\linewidth}
+\begin{C}
+x = 0;
+i = 0;
+r = 0;
+while (i < 100) {
+	if (i < 50)
+		x = -1;
+	else
+		x = 1;
+
+	if (x == 0) r++;
+	i++;
+}
+return r;
+\end{C}
+\end{minipage}
+\begin{minipage}[c]{.49\linewidth}
+\begin{tikzpicture}[->,>=stealth',auto,node distance=1.5cm,
+                    semithick,font=\footnotesize]
+
+	\node[state] (n0) {$p_0$};
+	\node[state] (n1) [below of=n0] {$p_1$};
+	\node[state] (n2) [below of=n1] {$p_2$};
+	\node[state] (n3) [below left of=n2] {$p_3$};
+	\node[state] (n4) [below right of=n2] {$p_4$};
+	\node[state] (n5) [below right of=n3] {$p_5$};
+	\node[state] (n6) [below left of=n5] {$p_6$};
+	\node[state] (n7) [below right of=n5] {$p_7$};
+	\node[state] (n8) [below left of=n7] {$p_8$};
+	\node[state] (n9) [left of=n1] {$p_9$};
+
+  \path [transition] 
+		(n0) edge              node {$x,i,r \leftarrow 0$} (n1);
+  \path [transition] 
+		(n1) edge              node {$i < 100$} (n2);
+  \path [transition] 
+		(n1) edge              node {$i \geq 100$} (n9);
+  \path [transition] 
+        (n2) edge			   node [left] {$i < 50$} (n3);
+  \path [transition] 
+        (n2)  edge              node [right] {$i \geq 50$} (n4);
+  \path [transition] 
+        (n3) edge              node [left] {$x \leftarrow -1$} (n5);
+  \path [transition] 
+        (n4) edge			   node [right] {$x \leftarrow 1$} (n5);
+  \path [transition] 
+        (n5) edge			   node [left] {$x = 0$} (n6);
+  \path [transition] 
+        (n5) edge			   node [right] {$x \neq 0$} (n7);
+  \path [transition] 
+        (n6) edge              node [left] {$r \gets r+1$} (n8);
+  \path [transition] 
+        (n7) edge              node {} (n8);
+  \path [transition] 
+        (n8) edge [out=0, in=0, distance=3.5cm] node [right] {$i \gets i+1$} (n1);
+
+\end{tikzpicture}
+\end{minipage}
+\caption{Example of program. $p_5$ is a control-flow merge point, which is in
+the middle of the containing loop.}
+\label{ex9}
+\end{figure}
+
+\paragraph{Lookahead Widening} 
+The analysis begins with $p_1: x=i=r=0$. Then, the path $p_1 \rightarrow p_2
+\rightarrow p_3 \rightarrow p_5 \rightarrow p_7 \rightarrow p_1$ is the only one
+possible. The technique is then to stabilize the iterations over this subgraph,
+removing $p_4$ and $p_6$ for a while. After stabilization and narrowing phases,
+we obtain $ 0 \leq i \leq 50 \wedge r=0 \wedge -1 \leq x \leq 0$ at
+control point $p_1$, and 
+$ 0 \leq i \leq 49 \wedge r=0 \wedge x=-1$ at point $p_5$.
+
+Then, point $p_4$ becomes reachable. We continue the iterations and have to do
+a convex hull of polyhedra at point $p_5$, because of the two possible incoming
+transitions.
+We obtain the polyhedron
+$ 0 \leq i \leq 50 \wedge r=0 \wedge -1 \leq x \leq 1$.
+Point $p_6$ now becomes reachable, since $x=0$ is inside our polyhedron.
+Finally, the analysis terminates after some steps and finds $0 \leq r \leq i$.
+\emph{Lookahead Widening} is not able to prove that $r = 0$ at the end of the
+funtion.
+
+\paragraph{Path Focusing} 
+The analysis of the loop starts with 
+$p_1: x=i=r=0$. Then, the SMT-solver discovers the path $p_1 \rightarrow p_2
+\rightarrow p_3 \rightarrow p_5 \rightarrow p_7 \rightarrow p_1$. We do a
+widening/narrowing sequence and we find 
+$ 0 \leq i \leq 50 \wedge r=0 \wedge -1 \leq x \leq 0$ at $p_1$. The smt gives
+another path: 
+$p_1 \rightarrow p_2
+\rightarrow p_4 \rightarrow p_5 \rightarrow p_7 \rightarrow p_1$. After
+analyzing this path, we obtain 
+$ 0 \leq i \leq 100 \wedge r=0 \wedge x=0$. Analysis terminates since the
+SMT-solver gives $unsat$ answers. Here, We never focused on paths going through
+point $p_6$, so our analysis gives us the invariant $x=0$ at the end of the
+program. Indeed, there is no feasible paths going through $p_6$, but
+\emph{Lookahead Widening} don't know it because of the convex hull at point
+$p_5$.
 
@@
 analyzer, we apply some already existing passes before running our 
-abstract interpretation pass. Among the passes we run, one could cite:
+abstract interpretation pass. Among the passes we run, we could cite:
 \begin{itemize}
@@
 integers, which is inexistant in the representation, since the compiler doesn't
-need this information to generate code. Yet, for static analysis, one need to
-have this distinction to improve precision:
-for instance, if $n$ is unsigned, one could add the constraint $n \geq 0$ in our
-	abstract value at the beginning of the analysis.
+need this information to generate code. Yet, for static analysis, 
+this distinction is required to improve precision:
+for instance, if $n$ is unsigned, the constraint $n \geq 0$ could be added in
+	our abstract value at the beginning of the analysis.
 	This lack of precision may be limited by using disjunctive invariants: one
@@
 
-To do so, one could implement an LLVM pass that transform the internal
+To do so, we could implement an LLVM pass that transform the internal
 representation according to this principle. However, this pass has not been
@@
 the IA.
+
+Additionally, some optimization in the code could be proposed in order to
+reduce time and/or memory complexity. For instance, paths that have already been
+computed are until now stored in a tree. An optimization could be to store them
+in a Binary Decision Diagram, in order to reduce significantly the memory
+consumption when there is a huge number of paths.
 
@@
 	In this section, we evaluate our \emph{Lookahead Widening} and \emph{Path
-	Focusing} implementations. We can compare their execution time in some
+	Focusing} implementations. We compare their execution time in some
 	practical cases, and the precision of their results. Table \ref{experiments}
@@
 
+\emph{Path Focusing} appears to always perform similarly or better than 
+\emph{Lookahead Widening}. Indeed, it is more precise in the case of programs
+containing control-flow merges inside of loops, because of the distinction
+between all the paths of the loop. In simpler programs, it performs as well as
+\emph{Lookahead Widening}
+
+\emph{ex6.c} is a program containing non-linear arithmetic: there is an integer
+multiplication inside the loop. The Apron library handles multiplication using
+a linearization technique. However, multiplications is not allowed in LIA
+(Linear Integer Arithmetic) or LRA (Linear Real Arithmetic). Then, Yices
+SMT-solver errs with a message ``not implemented'' (which is the reason of the
+$\infty$ symbol in the table), when Z3 still works correctly.
+
+We tried our technique on example with a huge number of paths: ex8, ex8b and
+ex8c (see Table \ref{npaths}). These benchmarks are constructed such that the
+loop contains only a few
+possible paths, but a lot of unfeasible paths. The idea is to show that our
+technique doesn't blow up, even with a huge expanded multigraph. Results show
+that in such benchmarks, our technique finds a much better invariant and is even
+faster than \emph{Lookahead Widening}.
+
+For small programs, our execution time may not be significant. Indeed, the cost
+for reading the program, initializing the SMT-solver, \dots, may have a
+	non-negligible cost compared to the abstract interpretation analysis itself.
+	The only interest of these times is to show our algorithm is pretty fast.
+
+\begin{table}[!h]
+\begin{tabular}{|l||c|c|c|c|c|c|c|c|c|} \hline
+Benchmark  & Gopan & Beyer & Boustr & ex9 & test1 & ex6 & ex8 &
+ex8c & ex8b\\ \hline \hline
+paths (total) & 5 & 4 & 10 & 8 & 3 & 5 & 75626 & 5.7$.10^9$ & $4.10^{14}$\\
+\hline
+paths (computed) & 5 & 4 & 4 & 4 & 3 & 4 & 4 & 4 & 4\\ \hline
+iterations  & 6 & 5 & 5 & 4 & 3 & 4 & 4 & 4 & 4\\ \hline
+\end{tabular}
+\caption{Number of path in the expanded multigraph, number of path the
+algorithm actually computes, and total number of iterations}
+\label{npaths}
+\end{table}
+
+The last remark is about the efficiency of Yices and Z3. Yices seems to always
+be slightly faster than Z3, but in some cases, Yices appeared to blow up, or to
+err with a message ``not implemented'', for instance when adding non-linear
+relations into the formula. Then, it was a good choice to link our analyzer with
+both SMT-solvers.
+
 \begin{table}[!h]
@@
 
-Gopan.c & 0.088s & 0.080s & \begin{tabular}{l}
+Gopan.c & 0.076s & 0.076s & \begin{tabular}{l}
 -x.0 - y.0 + 102 >= 0 \\
@@
  -d.0 + 1 >= 0 \\
-\end{tabular} \\ \hline
-
-ex8.cpp & 0.516s & 0.408s & \begin{tabular}{l}
- -y.0 - x.0 + 1 >= 0\\
- -y.0 + x.0 + 1 >= 0\\
- y.0 - x.0 + 1 >= 0\\
- y.0 + x.0 + 1 >= 0
-\end{tabular} & 0.208 & \begin{tabular}{l}
--x.0 + 1 >= 0 \\
-x.0 + 1 >= 0 
 \end{tabular} \\ \hline
@@
 
-ex8c.c & 191m50.935s & $\infty$ & \begin{tabular}{l}
-c.0 + a.0 = 0 \\
+ex8.c & 0.516s & 0.408s & \begin{tabular}{l}
+ -y.0 - x.0 + 1 >= 0\\
+ -y.0 + x.0 + 1 >= 0\\
+ y.0 - x.0 + 1 >= 0\\
+ y.0 + x.0 + 1 >= 0
+\end{tabular} & 0.208 & \begin{tabular}{l}
+-x.0 + 1 >= 0 \\
+x.0 + 1 >= 0 
+\end{tabular} \\ \hline
+
+ex8c.c & 0.656s & 0.548s & \begin{tabular}{l}
+c.0 + a.0 = 0\\
 b.0 + d.0 = 0\\
@@
 b.0 + c.0 + 1 >= 0\\
-\end{tabular} & 0.928 & \begin{tabular}{l}
+\end{tabular} & 0.916s & \begin{tabular}{l}
 -b.0 + 1 >= 0\\
@@
 b.0 - a.0 + 1 >= 0\\
-b.0 + 1 >= 0
+b.0 + 1 >= 0\\
+\end{tabular} \\ \hline
+
+ex8b.c & 1.016s & 0.816s & \begin{tabular}{l}
+c.0 + y.0 = 0\\
+c.0 + a.0 = 0\\
+b.0 + d.0 = 0\\
+b.0 + x.0 = 0\\
+-b.0 - c.0 + 1 >= 0\\
+-b.0 + c.0 + 1 >= 0\\
+b.0 - c.0 + 1 >= 0\\
+b.0 + c.0 + 1 >= 0\\
+\end{tabular} & 36.300s & \begin{tabular}{l}
+-b.0 + 1 >= 0 \\
+-b.0 + a.0 + 1 >= 0\\
+-c.0 + 1 >= 0\\
+-c.0 + d.0 + 1 >= 0\\
+-a.0 + 1 >= 0\\
+-y.0 - x.0 + 1 >= 0\\
+-y.0 + x.0 + 1 >= 0\\
+-d.0 + 1 >= 0\\
+x.0 + 1 >= 0\\
+y.0 - x.0 + 1 >= 0\\
+y.0 + 1 >= 0\\
+a.0 + 1 >= 0\\
+c.0 - d.0 + 1 >= 0\\
+c.0 + d.0 + 1 >= 0\\
+b.0 - a.0 + 1 >= 0\\
+b.0 + 1 >= 0\\
 \end{tabular} \\ \hline
@@
 \caption{Results of \emph{Path Focusing} and \emph{Lookahead Widening}
-techniques applied to various benchmarks}
+techniques applied to various benchmarks. The invariant we give is the one at
+the head of the loop.}
 \label{experiments}
 \end{table}
-
-\emph{Path Focusing} appears to always perform similarly or better than 
-\emph{Lookahead Widening}. Indeed, it is more precise in the case of programs
-containing control-flow merges inside of loops, because of the distinction
-between all the paths of the loop.
-
-To be completed\dots
-
-\begin{table}[!h]
-\begin{tabular}{|l||c|c|c|c|c|c|c|} \hline
-Benchmark  & Gopan.c & Beyer.c & Boustr.c & ex8.cpp & ex9.cpp & test1.c & ex6.c\\ \hline \hline
-paths (total) & 5 & 4 & 10 & 75626 & 8 & 3 & 5 \\ \hline
-paths (computed) & 5 & 4 & 4 & 4 & 4 & 3 & 4\\ \hline
-iterations  & 6 & 5 & 5 & 4 & 4 & 3 & 4\\ \hline
-\end{tabular}
-\caption{Number of path in the expanded multigraph, number of path the
-algorithm actually computes, and total number of iterations}
-\end{table}
-
-To be completed\dots
 
@@
 	``In this program, $x$ is always between 1 and 10'', ``The program never
-	divides by zero'', etc.
+	divides by zero'', etc. The objective is to obtain a result as precise as
+	possible, by limiting the loss of precision during the analysis, induced for
+	instance by the widening operator, or control-flow merges.
 
@@
 
-	Our technique performs better because of the distinction between
-	paths. Then, encoding some properties, such as arithmetic overflows or
-	aliases, into distinct paths in the control flow graph is one direction to
-	explore.
-
-	Our analyzer is able to deal with small- and middle-sized programs, but lose
+	Our analyzer is able to deal with small- and middle-sized programs, but
+	looses
 	precision in the case of programs with pointers or several functions.
+	It has been tested with program having loops containing several hundreds
+	lines of code, and gave results after a time in the order of the second.
 	It could be improved in many ways, in order to support
@@
 	described in this report.
-	To be completed\dots
+	Since our technique combines well with acceleration techniques, another
+	direction is to implement this last technique to improve our results
+	obtained with classical widening/narrowing.
+
+	\emph{Path Focusing} technique performs better because of the distinction
+	between
+	paths. Then, encoding some properties in the control flow graph, such as
+	arithmetic overflows or aliases, is one direction to explore. One could also
+	try to combine \emph{Path Focusing} with other techniques, such as
+	\emph{Lookahead Widening}.
 
@@
 
+\addcontentsline{toc}{chapter}{Bibliography}
   \appendix
-
 \begin{figure}
@@
 \ref{runningexample}, \ref{multigraphtransformGopan} and
-\ref{multigraphGopan}.}
+\ref{multigraphGopan}, in the SMT-lib format.}
+\label{rhoformula}
 \end{figure}

### 2011-05-16 src/AI.cc

@@ -148,10 +148,12 @@ void AI::computeFunction(Function * F) {
 	LV = &(getAnalysis<Live>(*F));
 	LI = &(getAnalysis<LoopInfo>(*F));
 
-	*Out << "Computing Pr and Rho...";
+	*Out << "Computing Pr...";
 	LSMT->getPr(*F);
+	*Out << "OK\nComputing Rho...";
 	LSMT->getRho(*F);
 	*Out << "OK\n";
+	LSMT->man->SMT_print(LSMT->getRho(*F));
 	// add all function's arguments into the environment of the first bb
 	for (Function::arg_iterator a = F->arg_begin(), e = F->arg_end(); a != e; ++a) {
 		Argument * arg = a;
@@ -258,6 +260,7 @@ void AI::computeEnv(Node * n) {
 
 void AI::computeTransform (Node * n, std::list<BasicBlock*> path, Abstract &Xtemp) {
 	
+	*Out << "computeTransform...\n";
 	// setting the focus path, such that the instructions can be correctly
 	// handled
 	focuspath.clear();
@@ -289,7 +292,6 @@ void AI::computeTransform (Node * n, std::list<BasicBlock*> path, Abstract &Xtem
 		}
 		focusblock++;
 	}
-
 	ap_environment_t * env = NULL;
 	succ->create_env(&env);
 
@@ -324,11 +326,11 @@ void AI::computeTransform (Node * n, std::list<BasicBlock*> path, Abstract &Xtem
 				A.push_back(X2);
 				// creating the associated lincons for a future widening with
 				// threshold
-				ap_abstract1_t AX = ap_abstract1_of_tcons_array(man,Xtemp.main->env,(*it));
-				ap_lincons1_array_t lincons = ap_abstract1_to_lincons_array(man,&AX);
-				linconssize += ap_lincons1_array_size(&lincons);
-				linconsts.push_back(lincons);
-				ap_abstract1_clear(man,&AX);
+				//ap_abstract1_t AX = ap_abstract1_of_tcons_array(man,Xtemp.main->env,(*it));
+				//ap_lincons1_array_t lincons = ap_abstract1_to_lincons_array(man,&AX);
+				//linconssize += ap_lincons1_array_size(&lincons);
+				//linconsts.push_back(lincons);
+				//ap_abstract1_clear(man,&AX);
 			}
 			Xtemp.join_array(env,A);
[…]

### 2011-05-16 src/SMT.cc

@@ -398,15 +398,14 @@ void SMT::computeRhoRec(Function &F,
 			pred = *p;
 			//computeRhoRec(F,pred,dest,false,visited);
 			if (!Pr[&F]->count(pred)) {
-				computeRhoRec(F,pred,dest,false,visited);
-				primed[b].insert(primed[pred].begin(),primed[pred].end());
+				if (!visited->count(pred)) 
+					computeRhoRec(F,pred,dest,false,visited);
+				//primed[b].insert(primed[pred].begin(),primed[pred].end());
 			} else {
 				Pr_succ[pred].insert(dest);
 				Pr_pred[dest].insert(pred);
 			}
 		}
-	} else {
-		n_totalpaths++;
 	}
 
 	if (first) return;

### 2011-05-18 report/report.tex
added some missing code for a more correct construction of rho (when condition branchments depend on values of boolean Phi nodes)

@@
 \abstract{
-	Program verification aims at statically discover properties on programs,
+	Program verification is aimed at statically discover properties on programs,
 	such as the values that can take the different variables during execution.
@@
 
-Static analysis aims at automatically computing properties on programs, such as
+Static analysis is aimed at automatically computing properties on programs, such as
 the possible values of their variables during execution. This allows to show for
@@
 Satisfiability Modulo Theory (SMT) solving is a field in computer science that
-aims to decide the satisfiability of a logic formula containing Boolean
+is aimed to decide the satisfiability of a logic formula containing Boolean
 predicates and elements of a theory, such as linear arithmetic relations between
@@
 
-Static analysis aims to discover some properties about programs. In this
+Static analysis is aimed at discovering some properties about programs. In this
 example, one would compute the set of possible values for the variable $x$
@@
 Indeed, if $\Phi^\#(x) = x$, then $\Phi \circ \gamma (x) \subseteq \gamma(x)
-= \gamma \circ \Phi^\#(x)$. Abstract interpretation technique aims at computing
+= \gamma \circ \Phi^\#(x)$. Abstract interpretation technique is aimed at computing
 one of these $x$. It computes the stationary limit of an ascending sequence
@@
 Linear Relation Analysis \cite{CH78} (also noted LRA) is a direct application of
-Abstract Interpretation. It aims at computing an upper approximation of the
+Abstract Interpretation. It is aimed at computing an upper approximation of the
 reachable states of a program containing numerical variables. The set of
@@
 
-The computation aims to find a solution for this system of abstract semantic
+The computation is aimed at finding a solution for this system of abstract semantic
 inequalities:
@@
 Since this induces imprecision, another technique called \emph{acceleration}
-aims to compute the exact effect of the loop when possible \cite{Gon07,GH06}. For instance, for
+is aimed at computing the exact effect of the loop when possible \cite{Gon07,GH06}. For instance, for
 the following program, an invariant could be directly computed for the loop:
@@
 extrapolates without taking care of these new possible paths. This can induce a
-loss of precision, that Lookahead widening aims to limit.
+loss of precision, that Lookahead widening tries to limit.
 
@@
 	\label{reachability}
-	Path focusing technique aims to discover which paths to focus on for the
+	Path focusing technique tries to discover which paths to focus on for the
 	computation of the fixpoint iterations. 
@@
 	The control points in $P_R$ are $p_0$, $p_1$ and $p_7$. 
-	The $Ŕho$ formula associated to this program is depicted in
+	The $\rho$ formula associated to this program is depicted in
 	Figure \ref{rhoformula}.
@@
 
-	The technique aims to compute a disjunctive invariant for this control
+	The technique is aimed at computing a disjunctive invariant for this control
 	point: 
@@
 easily compute this number of paths when computing the $\rho$ formula.
+
+	\subsection{Removing Identity transitions}
+
+	When computing iterations, some of the loop paths may not change the state
+	of the variables, i.e the transition function associated to this path is the
+	identity function. This happens in the case of expansive transition
+	relations $\Phi$, defined by the property $X \subseteq \Phi(X)$.
+
+	Considering the set of initial states $I$, and the transition function 
+	$\tau$ mapping a state to its successor, we can define:
+
+	$$\Phi(X) = \{ x'\ |\ \exists x \in X, x' = \tau(x)\} \cup I$$
+
+	$\Phi$ is the function we classically use to find an inductive invariant
+	$X$,satisfying the property $\Phi(X) \subseteq X$. To improve the analysis,
+	we propose to compute this invariant using another function $\Psi$, by
+	removing the transitions that are actually the identity. Indeed, these
+	transitions make no change to the set $X$.
+	The $\Psi$ function could be expressed as follows:
+
+	$$\Psi(X) = \{ x'\ |\ \exists x \in X, x' = \tau(x) \wedge x' \neq x\}
+	\cup I$$
+
+	It is easy to prove that $\Psi(X) \subseteq X \Leftrightarrow \Phi(X)
+	\subseteq(X)$, so we can use $\Psi$ instead of $\Phi$ to compute our
+	invariant.
+
+	In the case of path focusing algorithm, we simply conjunct our formula with
+	$(i \neq j) \vee (x'_j \neq x_i)$, such that the SMT-solving only proposes
+	paths that are not the identity.
 
@@
 allocation. Then, the data structure we work on doesn't have live variables
-informations yet. To solve this problem, we coded our own pass, that aims to
-compute these live variables after the transformation in SSA form.
+informations yet. To solve this problem, we coded our own pass, that is aimed at
+computing these live variables after the transformation in SSA form.
 
@@
 	Abstract Interpretation is a common used technique to do static analysis of
-	programs. It aims to compute iteratively an over-approximation of the set of
+	programs. It is aimed at computing iteratively an over-approximation of the set of
 	possible states of the program, and allows to prove properties such as

### 2011-05-18 src/SMT.cc

@@ -264,7 +264,33 @@ SMT_expr SMT::getValueExpr(Value * v, std::set<Value*> ssa_defs) {
 	SMT_var var;
 
 	ap_texpr_rtype_t ap_type;
-	if (get_ap_type(v, ap_type)) return NULL;
+	if (get_ap_type(v, ap_type)) {
+		*Out << "getValueExpr returns NULL\n";
+		// this may be a boolean
+		if (ap_type == AP_RTYPE_INT) {
+			// this is a boolean
+			// so we can create a boolean variable
+			SMT_expr cond = NULL;
+			if (isa<CmpInst>(v)) {
+				cond = computeCondition(dyn_cast<CmpInst>(v));
+			} else if (isa<PHINode>(v)) {
+				cond = computeCondition(dyn_cast<PHINode>(v));
+			} else if (ConstantInt * vint = dyn_cast<ConstantInt>(v)) {
+				if (vint->isZero()) {
+					cond = man->SMT_mk_false();
+				} else if (vint->isOne()) {
+					cond = man->SMT_mk_true();
+				}
+			} 
+			if (cond == NULL) {
+				SMT_var cvar = man->SMT_mk_bool_var(getUndeterministicChoiceName(v));
+				cond = man->SMT_mk_expr_from_bool_var(cvar);
+			}
+			return cond;
+		} else {
+			return NULL;
+		}
+	}
 
 	if (isa<Constant>(v)) {
 		if (isa<ConstantInt>(v)) { 
@@ -556,10 +582,7 @@ void SMT::visitReturnInst (ReturnInst &I) {
 }
 
 SMT_expr SMT::computeCondition(PHINode * inst) {
-
-	//SMT_expr expr = getValueExpr(inst);	
 	return construct_phi_ite(*inst,0,inst->getNumIncomingValues());
-
 }
 
 SMT_expr SMT::computeCondition(CmpInst * inst) {
@@ -648,8 +671,6 @@ void SMT::visitBranchInst (BranchInst &I) {
 		if (isa<CmpInst>(I.getOperand(0)))
 			cond = computeCondition(dyn_cast<CmpInst>(I.getOperand(0)));
 		else if (isa<PHINode>(I.getOperand(0))) {
-			return;
-			// NOT IMPLEMENTED !!
 			cond = computeCondition(dyn_cast<PHINode>(I.getOperand(0)));
 		} else {
[…]

### 2011-05-18 report/report.tex
update of the report - Real experiments still need to be done

@@
 	$(i \neq j) \vee (x'_j \neq x_i)$, such that the SMT-solving only proposes
-	paths that are not the identity.
+	paths that are not the identity. Actually, it never focuses on such paths
+	during the acending sequence, but during narrowing iterations that occur 
+	at the end of the algorithm.
+
+	Ignoring these identity transitions makes the narrowing sequence recover
+	more precision.
+	Indeed, narrowing sequence updates the invariant $X$ with $\Phi(X)$, which
+	is still an invariant, while precision is recovered, meaning that $\Phi(X)
+	\subset X$. In the case of an expansive function, where $X \subseteq
+	\Phi(X)$, we never have $\Phi(X) \subset X$, so narrowing iterations are
+	useless.
 
@@
 
+	\subsection{Benchmarks}
+
 \emph{Path Focusing} appears to always perform similarly or better than 
@@
 \end{table}
+
+	\subsection{Analysis of real code}
+
+	In this section, we show that \emph{Path Focusing} technique performs better
+	that \emph{Lookahead} widening on some code taken from real open-source
+	programs.
+
+	To Be Completed\dots
 

### 2011-05-18 ex/ex2c.c

@@ -1,0 +1,34 @@
+
+int main(int argc, char ** argv) {
+	int x;
+	int y;
+	int i;
+	int toto;
+	
+	while (argc > 100000 || x >= 42) {
+	//x = 2;
+	//y = x + 50;
+	//y = y + 6;
+	//y = x + y + 5;
+
+	//toto = x == y || x == 24;
+	//if (toto) {
+	//	x++;
+	//}
+
+
+	for (i=0; i < y; i++) {
+//		x = x + 2;
+//		x = f(x);
+//		x = x - 2;
+	}
+
+	//x = x+1;
+
+//	for (i=0; i < y; i++) {
+//		x = x + 2;
+//	}
+
+	}
+	return x;
+}

### 2011-05-20 report/report.tex
loops are now unrolled once before doing the analysis

@@
 reachable states of a program containing numerical variables. The set of
-possible assignments for the numerical variables is abstracted by convex
+possible assignments for the numerical variables is abstracted by a convex
 polyhedron. This technique discovers invariant linear relations between
@@
 
-The classical union of two convex polyhedra may not be a convex polyhedra.
+The classical union of two convex polyhedra may not be a convex polyhedron.
 Therefore, we use the convex hull operation, noted $\sqcup$, for joining
@@
 such as the domain of convex polyhedra, there is an upper approximation of the
-set of possible states to the smallest convex polyhedra including this set. This
+set of possible states to the smallest convex polyhedron including this set. This
 loss of precision is unavoidable. Still, there are techniques proposing to
@@
 (102,0)$, depicted in figure \ref{result}.a. Since this union of segments is not
-a convex polyhedra, the best result we can wait for after abstraction is the
-polyhedra depicted in figure \ref{result}.b, which is the smallest convex
-polyhedra containing the two segments.
+a convex polyhedron, the best result we can wait for after abstraction is the
+polyhedron depicted in figure \ref{result}.b, which is the smallest convex
+polyhedron containing the two segments.
 
@@
 
-b. The best result we can have is the polyhedra in hashed lines.
+b. The best result we can have is the polyhedron in hashed lines.
 \end{minipage}
@@
 
-We apply abstract interpretation to this program. Let $X_i$ be the polyhedra
+We apply abstract interpretation to this program. Let $X_i$ be the polyhedron
 attached to point $p_i$. Initially, $X_i = \perp$ for each $i \in \{1,7\}$, and
 $X_0 = \top$. The tabular in figure \ref{iterations} shows the evolution of the
-polyhedron during the iterations.
+polyhedra during the iterations.
 
@@
 	\subsubsection{Expression of the reachability problem}
-
+	\label{smtformula}
 	The SMT-formula $\rho$ express the semantic of the program. We still need to create the
@@
 
-	In the case of path focusing algorithm, we simply conjunct our formula with
+	In the case of path focusing algorithm, we simply conjunct our formula
+	(see \ref{smtformula}) with
 	$(i \neq j) \vee (x'_j \neq x_i)$, such that the SMT-solving only proposes
@@
 
+\section{Unrolling loops}
+
+In some cases, the number of iterations inside a loop during execution 
+may be equals to zero. This happens if the condition of the \emph{while}
+statement is false at the first time the head of loop is reached.
+
+In our analyzer, this kind of loops gives imprecise results. In this section, we
+propose to unroll loops once in order to avoid it.
+This behaviour can happen in very simple programs, such as the one
+depicted in Figure \ref{loop}. Indeed, in the case of $n \leq 0$, the condition
+$i < n$ is always false.
+
+\begin{figure}[!h]
+\begin{minipage}[c]{.29\linewidth}
+\begin{C}
+i=0;
+while (i < n)
+	i++;
+\end{C}
+\end{minipage}
+\begin{minipage}[c]{.69\linewidth}
+\centering
+\begin{tikzpicture}[->,>=stealth',auto,node distance=1.3cm,
+                    semithick,font=\footnotesize]
+
+	\node[state] (n0) {$p_0$};
+	\node[state] (n1) [below of=n0] {$p_1$};
+	\node[state] (n3) [below right of=n1] {$p_3$};
+	\node[state] (n4) [below of=n3] {$p_4$};
+	\node[state] (n2) [left of=n4,xshift=-1cm] {$p_f$};
+
+  \path [transition] 
+		(n0) edge              node {$i \gets 0$} (n1);
+  \path [transition] 
+        (n1) edge [out=180,in=90]	node [left] {$i \geq n$} (n2);
+  \path [transition] 
+        (n1) edge			   node [right] {$i < n$} (n3);
+  \path [transition] 
+		(n3) edge              node {$i \gets i+1$} (n4);
+  \path [transition] 
+        (n4) edge [out=0, in=0, distance=2.5cm] node {} (n1);
+
+\end{tikzpicture}
+\end{minipage}
+\caption{If $n \leq 0$, there is no loop iterations.}
+\label{loop}
+\end{figure}
+
+In this example, our program behaves as follows:
+\begin{itemize}
+\item At point $p_1$, we first have the polyhedra $i=0$, where $n$ in
+unconstrained.
+\item The image of this polyhedra after the loop transition is $i \leq n \wedge
+i=1$. After a convex hull, we obtain for $p_1$ $0 \leq i \leq 1$.
+\item Again, we
+compute the image of this polyhedron by the transition, $1 \leq i \leq 2 \wedge
+i \leq n$, and after widening we
+obtain at $p_1$ the polyhedron $0 \leq i$.
+\item Finally, at point $p_f$, we have the polyhedron $0 \leq i \wedge i \geq
+n$.
+\end{itemize}
+
+Now, after unrolling the loop once, the program is the one in Figure
+\ref{loopunroll}. The analysis gives:
+\begin{itemize}
+\item At point $p_1$, we have $i=0$, then we directly have at point $p_f$ the
+polyhedron $i=0 \wedge n \leq 0$.
+\item At point $p_4$, we first have $i=1 \wedge n \geq 1$.
+\item the image of the polyhedron after the loop transition is $2 \leq n \wedge
+i=2$, which gives after a convex hull $ 1 \leq i \leq 2 \wedge i \leq n$.
+\item After one iteration and an application of widening, we find at $p_4$ the
+polyhedron $1 \leq i \leq n$.
+\item Then, at the exit point $p_f'$ of the loop, we have the polyhedron
+$i = n \wedge i \geq 1$.
+\end{itemize}
+
+\begin{figure}[!h]
+\begin{minipage}[c]{.29\linewidth}
+\begin{C}
+i=0;
+if (i < n) {
+	i++;
+	while (i < n)
+		i++;
+}
+\end{C}
+\end{minipage}
+\begin{minipage}[c]{.69\linewidth}
+\centering
+\begin{tikzpicture}[->,>=stealth',auto,node distance=1.3cm,
+                    semithick,font=\footnotesize]
+
+	\node[state] (n0) {$p_0$};
+	\node[state] (n1) [below of=n0] {$p_1$};
+	\node[state] (n3) [below right of=n1] {$p_3$};
+	\node[state] (n4) [below of=n3] {$p_4$};
+	\node[state] (n2) [left of=n4,xshift=-1cm] {$p_f$};
+	\node[state] (n22) [below of=n2,yshift=0.5cm] {$p_f'$};
+	\node[state] (n5) [below of=n4] {$p_5$};
+	\node[state] (n6) [below of=n5] {$p_6$};
+
+  \path [transition] 
+		(n0) edge              node {$i \gets 0$} (n1);
+  \path [transition] 
+        (n1) edge [out=180,in=90]	node [left] {$i \geq n$} (n2);
+  \path [transition] 
+        (n1) edge			   node [right] {$i < n$} (n3);
+  \path [transition] 
+		(n3) edge              node {$i \gets i+1$} (n4);
+  \path [transition] 
+		(n4) edge              node {$i \geq n$} (n22);
+  \path [transition] 
+		(n4) edge              node {$i < n$} (n5);
+  \path [transition] 
+		(n5) edge              node [right] {$i \gets i+1$} (n6);
+  \path [transition] 
+        (n6) edge [out=0, in=0, distance=2.5cm] node {} (n4);
+
+\end{tikzpicture}
+\end{minipage}
+\caption{Program of Figure \ref{loop} after unrolling the loop once.}
+\label{loopunroll}
+\end{figure}
+
+The convex hull of the polyhedron from $p_f$ with the one from $p_f'$ also gives
+$0 \leq i \wedge i \geq n$. Then, unrolling the loop does not give a better
+solution, since
+$p_f$ and $p_f'$ may be the same control point.
+
+However, our technique does a convex hull only if this exit point is in $P_R$. 
+If not, we consider independently the paths starting from $p_4$ and the ones
+starting from $p_1$, hence the better precision. Indeed, this comes to computing
+a disjunctive invariant for the loop, by distinguishing the case where the
+number of iterations is zero.
+
+In our analyzer, we unroll every loops once, by using LLVM optimisation
+passes such as \emph{loop-unroll}, \emph{loop-rotate} or \emph{loop-simplify}.
+
 	\section{SMT-solving}
@@
 	In this section, we show that \emph{Path Focusing} technique performs better
-	that \emph{Lookahead} widening on some code taken from real open-source
+	that \emph{Lookahead widening} on some code taken from real open-source
 	programs.

### 2011-05-20 ex/analyzer.sh

@@ -10,6 +10,7 @@ OPTIONS :
 
 	-o [FILE ]: name of the IR generated file
 	-p        : print the optimized llvm bytecode
+	-u        : unroll loops once
 	-G        : generate the .dot CFG
 	-g        : use Lookahead Widening instead of Path Focusing
 	-y        : use Yices instead of Microsoft Z3
@@ -21,6 +22,7 @@ PRINT=0
 GRAPH=0
 GOPAN=0
 YICES=0
+UNROLL=0
 COMPILE_OPTIONS=
 
 while getopts “hpygGi:o:c:” opt ; do
@@ -37,6 +39,9 @@ while getopts “hpygGi:o:c:” opt ; do
 			;;
 		i)
 			FILENAME=$OPTARG
+			;;
+		u)
+			UNROLL=1
 			;;
 		G)
 			GRAPH=1
@@ -71,7 +76,11 @@ clang -DNDEBUG -fno-exceptions $COMPILE_OPTIONS -emit-llvm -c $FILENAME -o $OUTP
 "
 clang -DNDEBUG -fno-exceptions $COMPILE_OPTIONS -emit-llvm -c $FILENAME -o $OUTPUT
 #opt -mem2reg -loopsimplify -lowerswitch $OUTPUT -o $OUTPUT
-opt -mem2reg -lowerswitch $OUTPUT -o $OUTPUT
+if [ $UNROLL -eq 1 ] ; then
+	opt -mem2reg -lowerswitch -loops  -loop-simplify -loop-rotate -lcssa -loop-unroll -unroll-count=1 $OUTPUT -o $OUTPUT
+else
+	opt -mem2reg -lowerswitch $OUTPUT -o $OUTPUT
+fi
 
 if [ $GRAPH -eq 1 ] ; then
 	opt -dot-cfg $OUTPUT -o $OUTPUT

### 2011-05-20 ex/compile_llvm.sh

@@ -53,7 +53,7 @@ fi
 
 clang -DNDEBUG -fno-exceptions -emit-llvm -c $FILENAME -o $OUTPUT
 #opt -mem2reg -loopsimplify -lowerswitch $OUTPUT -o $OUTPUT
-opt -mem2reg -lowerswitch $OUTPUT -o $OUTPUT
+opt -mem2reg -lowerswitch -loops  -loop-simplify -loop-rotate -lcssa -loop-unroll -unroll-count=1 $OUTPUT -o $OUTPUT
 
 if [ $GRAPH -eq 1 ] ; then
 	opt -dot-cfg $OUTPUT -o $OUTPUT

### 2011-05-20 src/AI.cc

@@ -294,9 +294,6 @@ void AI::computeTransform (Node * n, std::list<BasicBlock*> path, Abstract &Xtem
 	//Xtemp.set_top(env);
 	Xtemp.change_environment(env);
 
-	//std::vector<ap_lincons1_array_t> linconsts;
-	//size_t linconssize = 0;
-
 	std::list<std::vector<ap_tcons1_array_t*>*>::iterator i, e;
 	for (i = constraints.begin(), e = constraints.end(); i!=e; ++i) {
 		if ((*i)->size() == 1) {
@@ -304,13 +301,6 @@ void AI::computeTransform (Node * n, std::list<BasicBlock*> path, Abstract &Xtem
 					tcons1_array_print((*i)->front());
 				);
 				Xtemp.meet_tcons_array((*i)->front());
-				// creating the associated lincons for a future widening with
-				// threshold
-				//ap_abstract1_t AX = ap_abstract1_of_tcons_array(man,Xtemp.main->env,(*i)->front());
-				//ap_lincons1_array_t lincons = ap_abstract1_to_lincons_array(man,&AX);
-				//linconssize += ap_lincons1_array_size(&lincons);
-				//linconsts.push_back(lincons);
-				//ap_abstract1_clear(man,&AX);
 		} else {
 			std::vector<Abstract*> A;
 			std::vector<ap_tcons1_array_t*>::iterator it, et;
@@ -319,44 +309,16 @@ void AI::computeTransform (Node * n, std::list<BasicBlock*> path, Abstract &Xtem
 				X2 = new Abstract(&Xtemp);
 				X2->meet_tcons_array(*it);
 				A.push_back(X2);
-				// creating the associated lincons for a future widening with
-				// threshold
-				//ap_abstract1_t AX = ap_abstract1_of_tcons_array(man,Xtemp.main->env,(*it));
-				//ap_lincons1_array_t lincons = ap_abstract1_to_lincons_array(man,&AX);
-				//linconssize += ap_lincons1_array_size(&lincons);
-				//linconsts.push_back(lincons);
-				//ap_abstract1_clear(man,&AX);
 			}
-			//if (A.size() > 0)
-				Xtemp.join_array(env,A);
+			Xtemp.join_array(env,A);
 		}
 	}
 
-	//
-	//linconstraints = ap_lincons1_array_make(Xtemp.main->env,linconssize);
[…]

### 2011-05-20 src/SMT.cc

@@ -265,7 +265,6 @@ SMT_expr SMT::getValueExpr(Value * v, std::set<Value*> ssa_defs) {
 
 	ap_texpr_rtype_t ap_type;
 	if (get_ap_type(v, ap_type)) {
-		*Out << "getValueExpr returns NULL\n";
 		// this may be a boolean
 		if (ap_type == AP_RTYPE_INT) {
 			// this is a boolean
@@ -288,6 +287,7 @@ SMT_expr SMT::getValueExpr(Value * v, std::set<Value*> ssa_defs) {
 			}
 			return cond;
 		} else {
+			*Out << "ERROR: getValueExpr returns NULL\n";
 			return NULL;
 		}
 	}
@@ -757,7 +757,6 @@ void SMT::visitGetElementPtrInst (GetElementPtrInst &I) {
 SMT_expr SMT::construct_phi_ite(PHINode &I, unsigned i, unsigned n) {
 	if (i == n-1) {
 		// this is the last possible value of the PHI-variable
-		*Out << "Hello1 !\n";
 		return getValueExpr(I.getIncomingValue(i), primed[I.getParent()]);
 	}
 	SMT_expr incomingVal = 	getValueExpr(I.getIncomingValue(i), primed[I.getParent()]);
@@ -766,11 +765,6 @@ SMT_expr SMT::construct_phi_ite(PHINode &I, unsigned i, unsigned n) {
 	SMT_expr incomingBlock = man->SMT_mk_expr_from_bool_var(evar);
 
 	SMT_expr tail = construct_phi_ite(I,i+1,n);
-	*Out << "Hello !\n";
-	man->SMT_print(tail);
-	*Out << "\n";
-	man->SMT_print(incomingVal);
-	*Out << "\n";
 	return man->SMT_mk_ite(incomingBlock,incomingVal,tail);
 }
 

### 2011-05-23 report/report.tex
applied David's patches

@@
 
-\usepackage{mathcomp,amssymb,verimagstudent}
+\usepackage{verimagstudent}
+\usepackage{mathcomp,amssymb}
 \usepackage[utf8]{inputenc}
@@
 \usepackage[toc,page]{appendix} 
-
+% DM: verimagstudent charge hyperref, et hyperref doit habituellement
+% DM: passer en dernier
 \newtheorem{definition}{Definition}[section]
@@
 \def\lfp{\textnormal{\it lfp}}
+\newcommand{\widening}{\mathop{\triangledown}}
 
@@
 \abstract{
-	Program verification is aimed at statically discover properties on programs,
+	Program verification is aims at statically discovering properties on programs,
 	such as the values that can take the different variables during execution.
@@
 
-Static analysis is aimed at automatically computing properties on programs, such as
+Static analysis aims at automatically computing properties on programs, such as
 the possible values of their variables during execution. This allows to show for
-instance that a program will not overflow, will not divide by zero, to compute
+instance that a program will not overflow, will not divide by zero, and to compute
 loop invariants\dots
 The main applications of static analysis are compilation-time optimisation, and
-the proof of satefy properties in critical systems, such as avionics.
+the proof of safety properties in critical systems, such as avionics.
 
@@
 
-Satisfiability Modulo Theory (SMT) solving is a field in computer science that
-is aimed to decide the satisfiability of a logic formula containing Boolean
+Satisfiability Modulo Theory (SMT) solving is a technique
+for deciding the satisfiability of a logic formula containing Boolean
 predicates and elements of a theory, such as linear arithmetic relations between
@@
 
-\subsection*{Contribution and Organization}
+\subsection*{Contribution and Organisation}
 
@@
 \begin{itemize}
-\item We present a new technique, refered as \emph{Path Focusing}, that takes
+\item We present a new technique, referred as \emph{Path Focusing}, that takes
 benefits of the recent advances in the field of SMT-solving to improve precision
@@
 this technique to compute disjunctive invariants.
-\item We present an implementation of this technique into a small analyzer, as
+\item We present an implementation of this technique into a small analyser, as
 well as the implementation of another technique proposed in \cite{GopanR06},
@@
 
-The report is organized as follows:
+The report is organised as follows:
 Part \ref{stateoftheartpart} introduces the state of the art in Abstract
@@
 Part \ref{implementationpart} presents the implementation of these two
-techniques into an analyzer, and the experimental results we obtained.
+techniques into an analyser, and the experimental results we obtained.
 Finally, Part \ref{future} introduces some directions to explore.
@@
 
-\section{Introductive example}
+\section{Introductory example}
 
@@
 \end{eqnarray*}
-
+% DM: d'habitude on met des ' sur les variables de l'etat d'arrivee
+% DM: et pas de depart
 We see that $Y_1$ and $Y_2$ can be computed step by step:
@@
 finding approximate solutions of
-fixpoint equations. This method is used for program analysis, since analyzing a
+fixpoint equations. This method is used for program analysis, since analysing a
 program often comes down to solving fixpoint equations, because of the loops and
@@
 
-Abstract Interpretation method proposes to represent more efficiently the
+Abstract Interpretation proposes to represent more efficiently the
 elements of this complex domain $C$ of concrete values,
 by choosing an simpler domain  $A$ called \emph{abstract domain}. 
+% DM: un peu lourd et peu clair
 
@@
 Indeed, if $\Phi^\#(x) = x$, then $\Phi \circ \gamma (x) \subseteq \gamma(x)
-= \gamma \circ \Phi^\#(x)$. Abstract interpretation technique is aimed at computing
-one of these $x$. It computes the stationary limit of an ascending sequence
+= \gamma \circ \Phi^\#(x)$. Abstract interpretation aims at computing
+such an~$x$. It computes the stationary limit of an ascending sequence
 $(x_i)_{i \geq 0}$ defined by the induction:
@@
 $$
-
 where $\perp$ is the least element of $A$.
@@
 depends on the properties of the abstract domain:
-this one should be finite, or of finite depth, meaning that there doesn't exist
+this one should be finite, of finite height,
+or more generally satisfy the \emph{ascending chain condition},
+meaning that there doesn't exist
 any sequence $(x_i)_{i \geq 0}$ of elements of the abstract domain $A$ such that
-$\forall i \geq 0,\ x_i <_A x_{i+1}$. Indeed, in case of a domain of infinite
-depth, the least fixpoint
+$\forall i \geq 0,\ x_i <_A x_{i+1}$. Indeed, otherwise, the least fixpoint
 computation could run indefinitely, since the ascending sequence of elements of
@@
 
-Most of the domains that are used for program analysis doesn't satisfy this
-finiteness property. To ensure convergence in this case, another approximation
+Many of the domains that are used for program analysis, including those of intervals and convex polyhedra, don't satisfy this ascending chain condition.
+To ensure convergence in this case, another approximation
 is performed: a new operator is defined, called \emph{widening operator}, that
@@
 \cite{CC77,CousotCousot92-4}. This \emph{widening
-operator} is usually noted $\nabla: A \times A \rightarrow A$, and satisfies
+operator} is usually noted $\widening: A \times A \rightarrow A$, and satisfies
 the following properties:
@@
 \begin{itemize}
-\item $\forall x_1, x_2 \in A,\ x_1 \leq_A x_1 \nabla x_2$ and $x_2 \leq_A x_1
-\nabla x_2$. This guarantees the correctness of the result.
+\item $\forall x_1, x_2 \in A,\ x_1 \leq_A x_1 \widening x_2$ and $x_2 \leq_A x_1
+\widening x_2$. This guarantees the correctness of the result.
+% DM: souvent n'est defini que pour x1<=x2 ce qui oblige a utiliser
+% DM: une union abstraite juste avant
 \item For any increasing sequence $x_0 \leq_A x_1 \leq_A \dots$, the sequence
@@
 x'_0 & = &  x_0 \\
-x'_{i+1} & = &  x'_i\ \nabla\ x_{i+1},\ \  \forall i > 0
+x'_{i+1} & = &  x'_i\ \widening\ x_{i+1},\ \  \forall i > 0
 }$$
@@
 \cite{Monniaux_HOSC09} gives a more general definition of the widening operator.
+% DM: c'est gentil de me citer!
 
@@
 x'_0 &=& \perp \\
-x'_{i+1} &=& x'_i\ \nabla\ \Phi^\#(x'_i),\ \ \forall i > 0
+x'_{i+1} &=& x'_i\ \widening\ \Phi^\#(x'_i),\ \ \forall i > 0
 }$$
@@
 
-Since the domain of convex polyhedra is of infinite depth, a widening operator
+Since the domain of convex polyhedra is of infinite height, a widening operator
 is defined to ensure convergence of the technique. 
-Intuitively, the polyhedron $P \nabla Q$ is obtained after 
+Intuitively, the polyhedron $P \widening Q$ is obtained after 
 removing from the system of linear equations defining $P$ all the inequalities
-that are not satisfied by $Q$.
+that are not satisfied by $Q$ (the actual definition is somewhat more involved,
+see e.g. \cite{BagnaraHRZ05SCP}).
 For instance, if $P = \{(x,y)\ |\ 0 \leq x \leq y \leq 1 \}$ and $Q = 
 \{(x,y)\ |\ 0 \leq x \leq y \leq 2 \}$, then the result of the widening operator
-will be $P \nabla Q = \{(x,y)\ |\ 0 \leq x \leq y \}$.
+will be $P \widening Q = \{(x,y)\ |\ 0 \leq x \leq y \}$.
 
 There have been various
-work proposing a definition or a refinment of the widening operator in LRA
+work proposing a definition or a refinement of the widening operator in LRA
 \cite{CH78,Hal79,HPR97,BlanchetCousotEtAl_PLDI03}, in order to fight
@@
 \begin{itemize}
-\item \emph{Delayed widening}: instead of applying windening at the first
+\item \emph{Delayed widening}: instead of applying widening at the first
 iteration of the analysis, one apply it after a number $n$ of iterations. During
@@
 since widening is applied within a finite time after $n$ iterations. Most of the
-cases, $n$ can be choosen equals to $2$.
+cases, $n$ can be chosen equals to $2$.
 \item \emph{Widening with threshold (or ``up to'')}: In some cases, instead of
 brutally applying the widening operator, a fixed set of linear inequalities $T$
-can be chosen, and the widening operator $\nabla_T$ is defined as follows:
-$P \nabla_T Q$ is the intersection of $P \nabla Q$ with all the inequalities in
+can be chosen, and the widening operator $\widening_T$ is defined as follows:
+$P \widening_T Q$ is the intersection of $P \widening Q$ with all the inequalities in
 $T$ satisfied by both $P$ and $Q$.
@@
 we have $Q = \{(x,y)\ |\ 0 \leq x \leq y \leq 2 \}$. Instead of a classic
-widening (which gives the result $P \nabla Q = \{(x,y)\ |\ 0 \leq x \leq y \}$),
+widening (which gives the result $P \widening Q = \{(x,y)\ |\ 0 \leq x \leq y \}$),
 a widening with the threshold $x \leq 50$ gives 
-$P \nabla_T Q = \{(x,y)\ |\ 0 \leq x \leq y \leq 50\}$. In such cases, widening
+$P \widening_T Q = \{(x,y)\ |\ 0 \leq x \leq y \leq 50\}$. In such cases, widening
 with threshold helps to find better invariants.
@@
 \vdots \\
-X_p \nabla \left( I_p \sqcup \displaystyle \bigsqcup_{(p',p) \in E} \tau^\#
+X_p \widening \left( I_p \sqcup \displaystyle \bigsqcup_{(p',p) \in E} \tau^\#
 (X_{p'}) \right) \\
@@
 There have been some work to find efficient iteration strategies
-\cite{Bou92}, such as first stabilizing innermost loops, or stabilizing strongly
+\cite{Bou92}, such as first stabilising innermost loops, or stabilising strongly
 connected components of the graph.
@@
 we always obtain a correct over-approximation.
-In practice, a single narrowing iteration is enough to recover the precision.
+In practise, a single narrowing iteration is enough to recover the precision.
 
@@
 to expand the control flow graph. For instance, instead of considering a graph
-with a sequence of $n$ \emph{if-then-else} between point $p_1$ and $p_n$, 
+with a sequence of $n$ \emph{if-then-else} between points $p_1$ and $p_n$, 
 the $n$ merge node
@@
 
-Typically, to analyze a program with loops, the approach is to use the
+Typically, to analyse a program with loops, the approach is to use the
 widening operator, for example at the head of the loops, to ensure convergence.
@@
 Lookahead widening \cite{GopanR06} is a technique that isolates the
-different loop phases during the analysis, and stabilizes each of these loop
+different loop phases during the analysis, and stabilises each of these loop
 phases before proceeding to the next. One only consider paths that are feasible
@@
 The interest of this method is to apply narrowing iterations right after each
-path computation. This allows to recover precision before analyzing new paths,
+path computation. This allows to recover precision before analysing new paths,
 and then to be more precise.
@@
 	so the loss of precision due to the convex hull of polyhedra may be
-	minimized.
+	minimised.
 
@@
 have several incoming transitions, and has the same number of arguments as the point
-has incoming transitions. If we suppose these transitions are ordered, then the
+has incoming transitions. Assuming these transitions are ordered, the
 $\Phi$-function returns the value of its $i$-th argument if the point is reached
@@
 		variables in $X_p$, that ends on point $q$, with variables that are not
-		in $X_q$ ?''}
+		in $X_q$?''}
 	\end{center}
@@
 	efficient programs, known as SMT-solvers, that can decide the satisfiability
-	of LRA or LIA formulas. Well-known efficient SMT-solvers are
+	of LRA or LIA formulae. Well-known efficient SMT-solvers are
 	Z3 \cite{MouraB08} and Yices \cite{DutertreM06}.
@@
 \item For each operation in the program, we encode its semantic, expressed in
-terms of the boolean and numerical SSA variables. For instance, if the CFG
+terms of the Boolean and numerical SSA variables. For instance, if the CFG
 contains the assignment $x.2 = x.1 + 1$, then we simply add $x.2 = x.1 + 1$ in
@@
 c(i,j)$, where $B_i = b_i^s$ if $p_i \in P_R$, or $b_i$ if not.
-$c(i,j)$ express the condition that needs to be true to go through the
+$c(i,j)$ expresses the condition that needs to be true to go through the
 transition. For instance, $c(i,j)$ could be of the form $x < y$, \dots
 If the transition is non-deterministic, then $c(i,j) = c_{i,j}$, where
-$c_{i,j}$ is a boolean variable left non-deterministic.
+$c_{i,j}$ is a Boolean variable left non-deterministic.
 \item For each point $p_i \notin P_R$, we set the reachability predicate 
@@
 \item Numerical $\Phi$-variables assignations have to be expressed in the $\rho$
-formula. We simply use $\ite$ (\emph{if-then-else}) statements, standardly defined in the
+formula. We simply use $\ite$ (\emph{if-then-else}) statements, a standard feature of the
 SMT-Lib format \cite{BarST-SMTLIB}, for giving the right value to the variable,
@@
 
-	To compute the reachability strating from point $p_i$, the formula given to
+	To compute the reachability starting from point $p_i$, the formula given to
 	the SMT-solver is:
@@
 
-	Path Focusing algorithm is described in Algorithm \ref{pathfocusingalgo}.
-	We maintain a set $A$ of point in $P_R$ that need to be treated.
+	The Path Focusing algorithm is described in Algorithm \ref{pathfocusingalgo}.
+	We maintain a set $A$ of points in $P_R$ that need to be treated.
 	We select iteratively one of these control points, and search for new focus
@@
 		\If {$p_j \in P_W$}
-			\State $X_{p_j} \gets X_{p_j} \nabla (X_{p_j} \sqcup Y)$
+			\State $X_{p_j} \gets X_{p_j} \widening (X_{p_j} \sqcup Y)$
 		\Else
@@
 	operator,  some narrowing steps could be performed to regain precision
-	before focusing on another path. This is the same principle as the one we
-	find in \emph{Lookahead widening} \cite{GopanR06}.
+	before focusing on another path. This relates to the ideas in
+        \emph{Lookahead widening} \cite{GopanR06}.
 
@@
 
-	When the self loop is accelerable \cite{Gon07}, instead of doing a
+	When the self loop can be accelerated \cite{Gon07}, instead of doing a
 	widening/narrowing sequence,  acceleration directly finds
@@
 		\If {$(p_j \in P_W) \wedge (p_i \neq p_j \vee e \in U)$}
-			\State $X_{p_j} \gets X_{p_j} \nabla (X_{p_j} \sqcup Y)$
+			\State $X_{p_j} \gets X_{p_j} \widening (X_{p_j} \sqcup Y)$
 		\Else
@@
 
+%DM: en quoi tout ceci est de la transitive closure?
+%DM: transitive closure c'est sur des relations i/o
+%DM: note que tu pourrais faire des relations (summaries de boucles, etc.)
+%DM: en gardant dans tes polyedres une copie des variables en tete
+%DM: du bloc/fonction/truc
+
 	\subsection{Disjunctive invariants}
@@
 	paths that are not the identity. Actually, it never focuses on such paths
-	during the acending sequence, but during narrowing iterations that occur 
+	during the ascending sequence, but during narrowing iterations that occur 
 	at the end of the algorithm.
@@
 \rightarrow p_3 \rightarrow p_5 \rightarrow p_7 \rightarrow p_1$ is the only one
-possible. The technique is then to stabilize the iterations over this subgraph,
-removing $p_4$ and $p_6$ for a while. After stabilization and narrowing phases,
+possible. The technique is then to stabilise the iterations over this subgraph,
+removing $p_4$ and $p_6$ for a while. After stabilisation and narrowing phases,
 we obtain $ 0 \leq i \leq 50 \wedge r=0 \wedge -1 \leq x \leq 0$ at
@@
 \emph{Lookahead Widening} is not able to prove that $r = 0$ at the end of the
-funtion.
+function.
 
@@
 widening/narrowing sequence and we find 
-$ 0 \leq i \leq 50 \wedge r=0 \wedge -1 \leq x \leq 0$ at $p_1$. The smt gives
+$ 0 \leq i \leq 50 \wedge r=0 \wedge -1 \leq x \leq 0$ at $p_1$. The SMT gives
 another path: 
@@
 \rightarrow p_4 \rightarrow p_5 \rightarrow p_7 \rightarrow p_1$. After
-analyzing this path, we obtain 
+analysing this path, we obtain 
 $ 0 \leq i \leq 100 \wedge r=0 \wedge x=0$. Analysis terminates since the
@@
 \emph{Lookahead Widening} and \emph{Path Focusing}
-techniques have been implemented into a small analyzer, so that we can compare
+techniques have been implemented into a small analyser, so that we can compare
 the precision of their results. In this part, we explain some details about
@@
  \section{Infrastructure}
-The analyzer is based on the \emph{Low Level Virtual Machine} (LLVM)
+The analyser is based on the \emph{Low Level Virtual Machine} (LLVM)
 \cite{LLVM:CGO04}, which is a compilation infrastructure used for instance by
@@
 
-The advantage of using such an infrastructure is that the analyzer can check
+The advantage of using such an infrastructure is that the analyser can check
 programs written in various languages: C, C++, \dots, without needing to write
-a specific frontend for each of them. Here, the analyzer directly works on the
+a specific frontend for each of them. Here, the analyser directly works on the
 internal representation (IR), which is a graph representation of
@@
 
-Our analyzer takes as parameter the LLVM internal representation of a program,
+Our analyser takes as parameter the LLVM internal representation of a program,
 and returns linear relations between the numerical variables at each control
@@
 parts of it. These passes are mostly used for optimisation purpose. In our
-analyzer, we apply some already existing passes before running our 
+analyser, we apply some already existing passes before running our 
 abstract interpretation pass. Among the passes we run, we could cite:
@@
 variables are in SSA form, and removes most of the \emph{Load/Store} operations
-into memory. It is useful since our analyzer doesn't have a memory model, and
+into memory. It is useful since our analyser doesn't have a memory model, and
 then loses lots of precision in the case of \emph{Load/Store} instructions.
@@
 This is the case for the distinction between \emph{signed} and \emph{unsigned}
-integers, which is inexistant in the representation, since the compiler doesn't
+integers, which is nonexistent in the representation, since the compiler doesn't
 need this information to generate code. Yet, for static analysis, 
@@
 
-	The analyzer is implemented using the APRON library \cite{JM09}, which
+	The analyser is implemented using the APRON library \cite{JM09}, which
 	is a common interface to libraries implementing all the features
@@
 	variables, when only a single incoming value is possible. This is the case
-	when all but one polyhedra associated to a predessor are at bottom.
+	when all but one polyhedra associated to a predecessor are at bottom.
 	For instance, let say point $p$ has $p_1, p_2, \dots, p_n$ as ordered 
@@
 One classical approach would be to compute a disjunctive invariant, by attaching
-two different polyhedron to the BasicBlock. However, we prefered the following
+two different polyhedron to the BasicBlock. However, we preferred the following
 technique: we transform all the operations of the form $x \neq y$ into
@@
 representation according to this principle. However, this pass has not been
-implemented in our analyzer, and this transformation has to be done manually.
+implemented in our analyser, and this transformation has to be done manually.
 
@@
 
-In our analyzer, this kind of loops gives imprecise results. In this section, we
+In our analyser, this kind of loops gives imprecise results. In this section, we
 propose to unroll loops once in order to avoid it.
@@
 
-In our analyzer, we unroll every loops once, by using LLVM optimisation
+In our analyser, we unroll every loops once, by using LLVM optimisation
 passes such as \emph{loop-unroll}, \emph{loop-rotate} or \emph{loop-simplify}.
@@
 	SMT formula, and to extract a model when the formula is satisfiable, i.e an
-	assignement of the variables so that the formula is \emph{true}. For doing
+	assignment of the variables so that the formula is \emph{true}. For doing
 	this, our implementation has an interface with Microsoft Z3 \cite{MouraB08},
 	and Yices \cite{DutertreM06}. One can easily switch from one to another with
-	an argument we give as parameter to the analyzer. In this way, we can
+	an argument we give as parameter to the analyser. In this way, we can
 	compare the efficiency of both SMT-solvers.
@@
 
-	Our analyzer could be extended in many ways. Indeed, it is intra-procedural,
+	Our analyser could be extended in many ways. Indeed, it is intra-procedural,
 	meaning that it does not consider that a variable may change after a
@@
 
-	Additionnaly, our analyzer has no memory model:
+	Additionally, our analyser has no memory model:
 \begin{itemize}
@@
 to a variable $x$, doesn't provide properties about the value $v$. Then, our
-analyzer considers $x$ can take all values of its type.
+analyser considers $x$ can take all values of its type.
 \end{itemize}
@@
 
-Additionally, some optimization in the code could be proposed in order to
+Additionally, some optimisation in the code could be proposed in order to
 reduce time and/or memory complexity. For instance, paths that have already been
-computed are until now stored in a tree. An optimization could be to store them
+computed are until now stored in a tree. An optimisation could be to store them
 in a Binary Decision Diagram, in order to reduce significantly the memory
@@
 	In our running example (Figures \ref{runningexample},
-	\ref{multigraphtransformGopan} and \ref{multigraphGopan}), the analyzer
-	outputs the following result for the basicblock $1$:
+	\ref{multigraphtransformGopan} and \ref{multigraphGopan}), the analyser
+	outputs the following result for the basic block $1$:
 
@@
 	practical cases, and the precision of their results. Table \ref{experiments}
-	gives the results of the analyzer in various examples, some of them taken
+	gives the results of the analyser in various examples, some of them taken
 	from the recent literature \cite{CGGMP05,GopanR06}.
@@
 For small programs, our execution time may not be significant. Indeed, the cost
-for reading the program, initializing the SMT-solver, \dots, may have a
+for reading the program, initialising the SMT-solver, \dots, may have a
 	non-negligible cost compared to the abstract interpretation analysis itself.
@@
 err with a message ``not implemented'', for instance when adding non-linear
-relations into the formula. Then, it was a good choice to link our analyzer with
+relations into the formula. Then, it was a good choice to link our analyser with
 both SMT-solvers.
@@
 
-Some operations such as additions, soustractions\dots may overflow if their
+Some operations such as additions, subtractions\dots may overflow if their
 result is not in the type's definition interval. For instance, an
@@
 
-	Lookahead Widening works on the classical control flow graph, and stabilizes
+	Lookahead Widening works on the classical control flow graph, and stabilises
 	temporarily the fixpoint analysis on a part of this graph, forgetting paths
@@
 
-	Validation of softwares has become an essential domain in industry,
+	Validation of software has become an essential domain in industry,
 	especially those dealing with embedded systems, where a bug in the program
@@
 	In this report, we proposed a new approach to guide the iterations of the
-	analyzer, in order to distinguish the different paths of the program
+	analyser, in order to distinguish the different paths of the program
 	and to treat them
@@
 
-	We implemented this technique into a small analyzer dealing with C and C++
+	We implemented this technique into a small analyser dealing with C and C++
 	programs, and run it on several
@@
 
-	Our analyzer is able to deal with small- and middle-sized programs, but
+	Our analyser is able to deal with small- and middle-sized programs, but
 	looses

### 2011-05-23 src/AI.cc

@@ -75,7 +75,6 @@ bool AI::runOnModule(Module &M) {
 	}
 
 	*Out << "Number of iterations: " << n_iterations << "\n";
-	*Out << "Number of paths: " << n_totalpaths << "\n";
 	*Out << "Number of paths computed: " << n_paths << "\n";
 	return 0;
 }
@@ -196,7 +195,6 @@ void AI::computeFunction(Function * F) {
 }
 
 void AI::computeEnv(Node * n) {
-	*Out << "computeEnv\n";
 	BasicBlock * b = n->bb;
 	Node * pred = NULL;
 	std::map<Value*,std::set<ap_var_t> >::iterator i, e;
@@ -229,7 +227,6 @@ void AI::computeEnv(Node * n) {
 			//*Out << "Hello\n";
 			for (i = pred->intVar.begin(), e = pred->intVar.end(); i != e; ++i) {
 				if (LV->isLiveThroughBlock((*i).first,b)) {
-					*Out << *((*i).first) << " is live through " << b << "\n";
 					intVars[(*i).first].insert((*i).second.begin(),(*i).second.end());
 				}
 			}

### 2011-05-23 src/AIGopan.cc

@@ -8,7 +8,7 @@
 #include "llvm/PassManager.h"
 #include "llvm/Support/CFG.h"
 #include "llvm/Support/FormattedStream.h"
-#include "llvm/Analysis/LiveValues.h"
+//#include "llvm/Analysis/LiveValues.h"
 #include "llvm/Analysis/LoopInfo.h"
 
 #include "llvm/Analysis/Passes.h"

### 2011-05-23 src/AIGopan.h

@@ -10,7 +10,7 @@
 #include "llvm/Support/ErrorHandling.h"
 #include "llvm/Support/FormattedStream.h"
 #include "llvm/Support/InstVisitor.h"
-#include "llvm/Analysis/LiveValues.h"
+//#include "llvm/Analysis/LiveValues.h"
 #include "llvm/Analysis/LoopInfo.h"
 
 #include "ap_global1.h"

### 2011-05-23 src/Analyzer.cc

@@ -19,7 +19,7 @@ void show_help() {
 -h : help\n \
 -i : input file\n \
 -o : output file\n \
--g : use Lookahead Widening technique\
+-g : use Lookahead Widening technique\n \
 -y : use Yices instead of the Z3 SMT-solver (unused when -g)\n";
 }
 

### 2011-05-23 src/z3_manager.cc

@@ -12,9 +12,9 @@
 z3_manager::z3_manager() {
 	Z3_config config = Z3_mk_config();
 	Z3_set_param_value(config, "MODEL", "true");
-	Z3_set_param_value(config, "MODEL_V2", "true");
+	//DM Z3_set_param_value(config, "MODEL_V2", "true");
 	ctx = Z3_mk_context(config);
-	Z3_set_logic(ctx,"QF_LIA");
+	//DM Z3_set_logic(ctx,"QF_LIA");
 
 	int_type = Z3_mk_int_sort(ctx);
 	float_type = Z3_mk_real_sort(ctx);

### 2011-05-25 report/report.tex
new implementation of Lookahead Widening.

@@
 \item $Y_2$ now contains $1$, so $1+1=2 \in Y_1$: $Y_1 = \{0,1,2\}$.
-\item We continue the iteration until we find no more elements in the sets\dots
+\item We continue the iteration until we find no more new elements in the
+sets\dots
 \end{itemize}
@@
 elements of this complex domain $C$ of concrete values,
-by choosing an simpler domain  $A$ called \emph{abstract domain}. 
+by choosing a simpler domain  $A$ called \emph{abstract domain}. 
 % DM: un peu lourd et peu clair
@@
 x'_0 & = &  x_0 \\
-x'_{i+1} & = &  x'_i\ \widening\ x_{i+1},\ \  \forall i > 0
+x'_{i+1} & = &  x'_i\ \widening\ x_{i+1},\ \  \forall i \geq 0
 }$$
@@
 x'_0 &=& \perp \\
-x'_{i+1} &=& x'_i\ \widening\ \Phi^\#(x'_i),\ \ \forall i > 0
+x'_{i+1} &=& x'_i\ \widening\ \Phi^\#(x'_i),\ \ \forall i \geq 0
 }$$
@@
 x''_0 &=& \tilde{x} \\
-x''_{i+1} &=& \Phi^\#(x''_i), \ \ \forall i > 0
+x''_{i+1} &=& \Phi^\#(x''_i), \ \ \forall i \geq 0
 }$$
@@
 
-Linear Relation Analysis \cite{CH78} (also noted LRA) is a direct application of
+Linear Relation Analysis \cite{CH78} (also denoted LRA) is a direct application
+of
 Abstract Interpretation. It is aimed at computing an upper approximation of the
@@
 delaying the widening operation is more precise. The algorithm still terminates,
-since widening is applied within a finite time after $n$ iterations. Most of the
-cases, $n$ can be chosen equals to $2$.
+since widening is applied within a finite time after $n$ iterations. In most of
+the cases, $n$ can be chosen equals to $2$.
 \item \emph{Widening with threshold (or ``up to'')}: In some cases, instead of
@@
 \item $E \subseteq  P \times P$ is a set of directed edges. Each edge $e \in E$
-has a semantic $\tau_e: \P(\Q^n) \rightarrow \P(\Q^n)$, $\P(\Q^n)$ being the set
+has a semantics $\tau_e: \P(\Q^n) \rightarrow \P(\Q^n)$, $\P(\Q^n)$ being the
+set
 of possible values of $\overrightarrow{x}$. $\tau_e$ maps a set of states before
@@
    \end{minipage}
-   \caption{Example of program, and its associate control flow graph. This
+   \caption{Example of program, and its associated control flow graph. This
    program comes from \cite{GopanR06}.}
@@
 
-Acceleration is possible when the loop is simple enough, and come to compute the
-transitive closure $\tau_e^+$ of $\tau_e$, the transition function associated to
-the loop.
+Acceleration is possible when the loop is simple enough, and comes to compute
+the transitive closure $\tau_e^+$ of $\tau_e$, the transition function
+associated to the loop.
 
@@
 extrapolates without taking care of these new possible paths. This can induce a
-loss of precision, that Lookahead widening tries to limit.
+loss of precision, that \emph{Lookahead widening} tries to limit.
 
@@
 	the loop that is possible at the first iteration. The second path is
-	ignored, and will be treated later. we compute the iterations, and we obtain
+	ignored, and will be treated later. We compute the iterations, and we obtain
 	an first invariant for this part of the graph: at point $p_1$, $x =
@@
 \end{minipage}
-\caption{One can expand the graph in the left and obtain the associate graph in
+\caption{One can expand the graph in the left and obtain the associated graph in
 the right.}
@@
 \end{minipage}
-\caption{Example of classical control flow graph, and its associate expanded
+\caption{Example of classical control flow graph, and its associated expanded
 multigraph, where $P_R = \{p_0, p_1, p_7\}$. The path dashed from $p_1$ to $p_1$
@@
 
-This static single assignment form is required for the computation of the focus
+This static single assignment form is used for the computation of the focus
 path.
@@
 \begin{eqnarray*}
-\rho = & \displaystyle\bigwedge_{p_i \notin P_R}& \left[ (b_i = \bigvee_{(p_h,p_i) \in E}
-t_{h,i}) \wedge ( \bigwedge_{(p_i,p_j)\in E} t_{i,j} = b_i \wedge c(i,j))
+\rho = & \displaystyle\bigwedge_{p_i \notin P_R}& \left[ \left(b_i = \bigvee_{(p_h,p_i) \in E}
+t_{h,i}\right) \wedge \left( \bigwedge_{(p_i,p_j)\in E} t_{i,j} = b_i \wedge
+c(i,j)\right)
 \right] \wedge \\
- & \displaystyle\bigwedge_{p_i \in P_R}& \left[ (b_i^d = \bigvee_{(p_h,p_i) \in E}
- t_{h,i}) \wedge ( \bigwedge_{(p_i,p_j)\in E} t_{i,j} = b_i^s \wedge
- c(i,j)) \right] \wedge \\
+ & \displaystyle\bigwedge_{p_i \in P_R}& \left[ \left(b_i^d = \bigvee_{(p_h,p_i) \in E}
+ t_{h,i}\right) \wedge \left( \bigwedge_{(p_i,p_j)\in E} t_{i,j} = b_i^s \wedge
+ c(i,j)\right) \right] \wedge \\
  & \displaystyle\bigwedge_{x \in \Sigma}& assign(x)
@@
 	\label{smtformula}
-	The SMT-formula $\rho$ express the semantic of the program. We still need to create the
+	The SMT-formula $\rho$ express the semantics of the program. We still need
+	to create the
 	formula expressing the reachability problem defined in \ref{reachability}.
@@
 	One computes
-	iteratively the associate abstract values $X_0$, $X_1$ and $X_7$.
+	iteratively the associated abstract values $X_0$, $X_1$ and $X_7$.
 
@@
 We do another SMT query: is there another path
-starting in $p_0$ that makes an abstract value grow? Answer is no.
+starting in $p_0$ that makes an abstract value grow? The answer is no.
 \item Step 2: We now work on $p_1$. Is there a path starting in $p_1$ that
@@
 
-	\subsection{Removing Identity transitions}
+	\subsection{Removing identity transitions}
 
@@
 	$\Phi$ is the function we classically use to find an inductive invariant
-	$X$,satisfying the property $\Phi(X) \subseteq X$. To improve the analysis,
+	$X$, satisfying the property $\Phi(X) \subseteq X$. To improve the analysis,
 	we propose to compute this invariant using another function $\Psi$, by
@@
 
-We consider the program depicted in figure \ref{ex9}.
+We consider the program depicted in Figure \ref{ex9}, page \pageref{ex9}.
 Here, we would like to show that the returned value $r$ is zero.
@@
 widening/narrowing sequence and we find 
-$ 0 \leq i \leq 50 \wedge r=0 \wedge -1 \leq x \leq 0$ at $p_1$. The SMT gives
-another path: 
+$ 0 \leq i \leq 50 \wedge r=0 \wedge -1 \leq x \leq 0$ at $p_1$. The SMT-solver
+gives another path: 
 $p_1 \rightarrow p_2
@@
 Another example of information that was missing is the set of live variables.
-Indeed, in LLVM, live variables computation is processed after the register
+Indeed, in LLVM, live variables computation is processed during the register
 allocation. Then, the data structure we work on doesn't have live variables
@@
 	just after the $\Phi$-variables assignments. In this way, the
-	$\Phi$-variables of the associate BasicBlock will be part of the dimensions
+	$\Phi$-variables of the associated BasicBlock will be part of the dimensions
 	of the polyhedron.
@@
 
-	At a program point $p$, the dimensions of the associate abstract value could
+	At a program point $p$, the dimensions of the associated abstract value could
 	only be the numerical variables that are live at $p$. This is
@@
 	when all but one polyhedra associated to a predecessor are at bottom.
-	For instance, let say point $p$ has $p_1, p_2, \dots, p_n$ as ordered 
+	For instance, let's say point $p$ has $p_1, p_2, \dots, p_n$ as ordered 
 	predecessors. At point $p$, we have $x = \Phi(x_1, x_2,\dots,x_n)$.
@@
 technique: we transform all the operations of the form $x \neq y$ into
-$x < y \vee x > y$, meaning that we separate the cases \emph{lesser than} and
+$x < y \vee x > y$, meaning that we separate the cases \emph{less than} and
 \emph{greater than} into two distinct paths.
@@
 In some cases, the number of iterations inside a loop during execution 
-may be equals to zero. This happens if the condition of the \emph{while}
+may be equal to zero. This happens if the condition of the \emph{while}
 statement is false at the first time the head of loop is reached.
@@
 \begin{itemize}
-\item At point $p_1$, we first have the polyhedra $i=0$, where $n$ in
+\item At point $p_1$, we first have the polyhedron $i=0$, where $n$ in
 unconstrained.
-\item The image of this polyhedra after the loop transition is $i \leq n \wedge
+\item The image of this polyhedron after the loop transition is $i \leq n \wedge
 i=1$. After a convex hull, we obtain for $p_1$ $0 \leq i \leq 1$.
@@
 	There have been research to find efficient interprocedural analysis
-	techniques \cite{Bourdoncle90,Bou92}.
+	techniques \cite{Bourdoncle90,Bou92,jeannet09b}.
 
@@
 
-	In all experiments, the set $P_R$ is chosen equals to $P_W \cup P_{ret} \cup
+	In all experiments, the set $P_R$ is chosen equal to $P_W \cup P_{ret} \cup
 	P_{init}$,
@@
 \emph{Lookahead Widening}. Indeed, it is more precise in the case of programs
-containing control-flow merges inside of loops, because of the distinction
+containing control-flow merges inside of loops, thanks to the distinction
 between all the paths of the loop. In simpler programs, it performs as well as
@@
 
-	\emph{Path Focusing} technique performs better because of the distinction
+	\emph{Path Focusing} technique performs better thanks to the distinction
 	between

### 2011-05-25 src/AIGopan2.cc

@@ -15,7 +15,7 @@
 #include "ap_global1.h"
 #include "pk.h"
 
-#include "AI.h"
+#include "AIGopan2.h"
 #include "Expr.h"
 #include "Node.h"
 #include "apron.h"
@@ -27,23 +27,23 @@
 
 using namespace llvm;
 
-char AI::ID = 0;
+char AIGopan2::ID = 0;
 
 
-static RegisterPass<AI> X("AI", "Abstract Interpretation Pass", false, true);
+static RegisterPass<AIGopan2> X("AIGopan2", "AbstractGopan Interpretation Pass", false, true);
 
-const char * AI::getPassName() const {
-	return "AI";
+const char * AIGopan2::getPassName() const {
+	return "AIGopan2";
 }
 
-void AI::getAnalysisUsage(AnalysisUsage &AU) const {
+void AIGopan2::getAnalysisUsage(AnalysisUsage &AU) const {
 	AU.setPreservesAll();
 	AU.addRequired<LoopInfo>();
 	AU.addRequired<Live>();
 	AU.addRequired<SMT>();
 }
 
-bool AI::runOnModule(Module &M) {
+bool AIGopan2::runOnModule(Module &M) {
 	Function * F;
 	BasicBlock * b;
 	Node * n;
@@ -68,7 +68,7 @@ bool AI::runOnModule(Module &M) {
 				Out->changeColor(raw_ostream::MAGENTA,true);
 				*Out << "\n\nRESULT FOR BASICBLOCK: -------------------" << *b << "-----\n";
 				Out->resetColor();
-				n->Y->print(true);
+				n->Xgopan->print(true);
 			}
 			delete Nodes[b];
 		}
@@ -79,7 +79,7 @@ bool AI::runOnModule(Module &M) {
 	return 0;
 }
 
-void AI::initFunction(Function * F) {
+void AIGopan2::initFunction(Function * F) {
 	Node * n;
 	// we create the Node objects associated to each basicblock
 	for (Function::iterator i = F->begin(), e = F->end(); i != e; ++i) {
@@ -96,7 +96,7 @@ void AI::initFunction(Function * F) {
 	*Out << *F;
 }
 
-void AI::printBasicBlock(BasicBlock* b) {
+void AIGopan2::printBasicBlock(BasicBlock* b) {
 	Node * n = Nodes[b];
 	LoopInfo &LI = getAnalysis<LoopInfo>(*(b->getParent()));
 	if (LI.isLoopHeader(b)) {
[…]

### 2011-05-25 src/AIGopan2.h

@@ -1,0 +1,182 @@
+#ifndef _AIGOPAN2_H
+#define _AIGOPAN2_H
+
+#include <queue>
+#include <vector>
+
+#include "llvm/Module.h"
+#include "llvm/Pass.h"
+#include "llvm/PassManager.h"
+#include "llvm/Support/ErrorHandling.h"
+#include "llvm/Support/FormattedStream.h"
+#include "llvm/Support/InstVisitor.h"
+#include "llvm/Analysis/LoopInfo.h"
+
+#include "ap_global1.h"
+#include "pk.h"
+
+#include "apron.h"
+#include "Node.h"
+#include "Live.h"
+#include "SMT.h"
+#include "PathTree.h"
+
+using namespace llvm;
+
+class AIGopan2 : public ModulePass, public InstVisitor<AIGopan2> {
+
+	private:
+		/// LV - result of the Live pass
+		Live * LV;
+		/// LI - result of the LoopInfo pass
+		LoopInfo * LI;
+		/// LSMT - result of the SMT pass
+		SMT * LSMT;
+		/// A - list of active Nodes, that still have to be computed
+		std::priority_queue<Node*,std::vector<Node*>,NodeCompare> A;
+		/// is_computed - remember the Nodes that don't need to be recomputed.
+		/// This is used to remove duplicates in the A list.
+		std::map<Node*,bool> is_computed;
+	
+		/// man - apron manager we use along the pass
+		ap_manager_t* man;
+
+		/// paths - remembers all the paths that have already been
+		/// visited
+		PathTree* pathtree;
+
+		std::map<BasicBlock*,std::list<BasicBlock*> > lastpath;
+
+	public:
+		static char ID;	
+	
+	public:
+
+		AIGopan2 () : ModulePass(ID), LV(NULL), LI(NULL), LSMT(NULL) {
+				man = pk_manager_alloc(true);
+				pathtree = new PathTree();
+				init_apron();
+				//linconstraints = ap_lincons1_array_make(ap_environment_alloc_empty(),0);
+			}
+
+		~AIGopan2 () {
+				ap_manager_free(man);
+				delete pathtree;
+			}
+
+		const char *getPassName() const;
+		void getAnalysisUsage(AnalysisUsage &AU) const;
+		bool runOnModule(Module &M);
+
+		void computeFunction(Function * F);
+
+		/// initFunction - initialize the function by creating the Node
+		/// objects, and computing the strongly connected components.
+		void initFunction(Function * F);
+		
[…]

### 2011-05-25 src/AbstractGopan.cc

@@ -53,10 +53,15 @@ void AbstractGopan::set_bottom(ap_environment_t * env) {
 }
 
 void AbstractGopan::change_environment(ap_environment_t * env) {
+
 	if (!ap_environment_is_eq(env,main->env))
 		*main = ap_abstract1_change_environment(man,true,main,env,false);
 	if (pilot != main && !ap_environment_is_eq(env,pilot->env))
 		*pilot = ap_abstract1_change_environment(man,true,pilot,env,false);
+}
+
+bool AbstractGopan::is_bottom() {
+	return ap_abstract1_is_bottom(man,main);
 }
 
 bool AbstractGopan::is_leq (AbstractGopan *d) {
@@ -76,25 +81,25 @@ bool AbstractGopan::is_leq (AbstractGopan *d) {
 void AbstractGopan::widening(Node * n) {
 	ap_abstract1_t Xmain_widening;
 	ap_abstract1_t Xpilot_widening;
-	ap_abstract1_t Xpilot;
 
 	if (is_leq(n->Xgopan)) {
 		Xmain_widening = ap_abstract1_copy(man,n->Xgopan->main);
 		Xpilot_widening = ap_abstract1_copy(man,n->Xgopan->pilot);
-	} else if (ap_abstract1_is_leq(man,pilot,n->Xgopan->pilot)) {
-		Xmain_widening = ap_abstract1_copy(man,pilot);
-		Xpilot_widening = ap_abstract1_copy(man,pilot);
 	} else {
-		// we apply the widening operator only on the pilot value,
-		// and we only join the main values.
-		Xmain_widening = ap_abstract1_join(man,false,n->Xgopan->main,main);
-		// before widening, n->Xgopan->pilot has to be included in pilot
-		Xpilot = ap_abstract1_join(man,false,n->Xgopan->pilot,pilot);
-		Xpilot_widening = ap_abstract1_widening(man,n->Xgopan->pilot,&Xpilot);
-		ap_abstract1_clear(man,&Xpilot);
-	}
-	if (pilot != main)
+		ap_abstract1_t dpUcm = ap_abstract1_join(man,false,pilot,n->Xgopan->main);
+		if (ap_abstract1_is_leq(man,&dpUcm,n->Xgopan->pilot)) {
+			Xmain_widening = ap_abstract1_copy(man,&dpUcm);
+			Xpilot_widening = ap_abstract1_copy(man,&dpUcm);
[…]

### 2011-05-25 src/AbstractGopan.h

@@ -36,6 +36,9 @@ class AbstractGopan {
 		/// change_environment - change the environment of the abstract value
 		void change_environment(ap_environment_t * env);
 
+		/// is_bottom - return true iff this is bottom
+		bool is_bottom();
+
 		/// is_leq - return true iff this <= d
 		bool is_leq(AbstractGopan * d);
 

### 2011-05-25 src/Execute.cc

@@ -18,6 +18,7 @@
 
 #include "AI.h"
 #include "AIGopan.h"
+#include "AIGopan2.h"
 #include "Node.h"
 #include "Execute.h"
 #include "Live.h"
@@ -93,7 +94,7 @@ void execute::exec(std::string InputFilename, std::string OutputFilename) {
 
 	ModulePass *AIPass;
 	if (useLookaheadWidening())
-		AIPass = new AIGopan();
+		AIPass = new AIGopan2();
 	else
 		AIPass = new AI();
 	FunctionPass *LoopInfoPass = new LoopInfo();

### 2011-05-25 src/Node.cc

@@ -112,12 +112,16 @@ void Node::create_env(ap_environment_t ** env, Live * LV) {
 
 	for (std::map<Value*,std::set<ap_var_t> >::iterator i = intVar.begin(),
 			e = intVar.end(); i != e; ++i) {
-		if (LV->isLiveThroughBlock((*i).first,bb) || LV->isUsedInBlock((*i).first,bb) || isa<UndefValue>((*i).first))
+		if (LV->isLiveThroughBlock((*i).first,bb)
+			|| LV->isUsedInBlock((*i).first,bb)
+			|| isa<UndefValue>((*i).first))
 			Sintvars.insert((*i).second.begin(), (*i).second.end());
 	}
 	for (std::map<Value*,std::set<ap_var_t> >::iterator i = realVar.begin(),
 			e = realVar.end(); i != e; ++i) {
-		if (LV->isLiveThroughBlock((*i).first,bb) || LV->isUsedInBlock((*i).first,bb) || isa<UndefValue>((*i).first))
+		if (LV->isLiveThroughBlock((*i).first,bb)
+			|| LV->isUsedInBlock((*i).first,bb)
+			|| isa<UndefValue>((*i).first))
 			Srealvars.insert((*i).second.begin(), (*i).second.end());
 	}
 

### 2011-06-01 report/report.tex
update of the report + fixed a bug in the comparison of abstract values

@@
 
-\emph{Path Focusing} appears to always perform similarly or better than 
+\emph{Path Focusing} appears to perform similarly or better than 
 \emph{Lookahead Widening}. Indeed, it is more precise in the case of programs
@@
 \centering \scriptsize
-\begin{tabular}{|c|c|c|l||c|l|} \hline
-
-			& \multicolumn{3}{c||}{Path Focusing} & \multicolumn{2}{c|}{Lookahead
-			Widening} \\ \cline{2-6}
-Benchmark	& \multicolumn{2}{c|}{time} & result	& time	& result	\\
-\cline{2-3}	& Z3 & Yices				& & & \\
+\begin{tabular}{|c|c|c|l||c||c|l|} \hline
+
+ & \multicolumn{3}{c||}{Path Focusing} & & \multicolumn{2}{c|}{Lookahead
+			Widening}   \\ \cline{2-4} \cline{6-7}
+Benchmark	& \multicolumn{2}{c|}{time} & result & cmp.	& time	& result	\\
+\cline{2-3}	& Z3 & Yices				& & & & \\
 \hline \hline
@@
 x.0 - y.0 >= 0        
-\end{tabular} & 0.032 & \begin{tabular}{l}
+\end{tabular} & $=$ & 0.032 & \begin{tabular}{l}
 -x.0 - y.0 + 102 >= 0 \\
@@
  -x.0 + y.0 >= 0  
-\end{tabular} & 0.032 & \begin{tabular}{l}
+\end{tabular}& $=$ & 0.032 & \begin{tabular}{l}
 -x.0  + 100 >= 0 \\
@@
  -d.0 + 1 >= 0
-\end{tabular} & 0.064 & \begin{tabular}{l}
+\end{tabular}& $\subset$ & 0.064 & \begin{tabular}{l}
  -1996x.0 + 4991d.0 + 1998995 >= 0\\
@@
 i.0 >= 0
-\end{tabular} & 0.052 & \begin{tabular}{l}
+\end{tabular}& $\subset$ & 0.052 & \begin{tabular}{l}
 -i.0 + 100 >= 0\\
@@
 j.0 - 6 >= 0
-\end{tabular} & 0.016 & \begin{tabular}{l}
+\end{tabular}& $=$ & 0.016 & \begin{tabular}{l}
 2j.0 + i.0 - 21 = 0 \\
@@
 x.0 >= 0
-\end{tabular} & 0.032 & \begin{tabular}{l}
+\end{tabular}& $\subset$ & 0.032 & \begin{tabular}{l}
 y.0 >= 0
@@
  y.0 + x.0 + 1 >= 0
-\end{tabular} & 0.208 & \begin{tabular}{l}
+\end{tabular}& $\subset$ & 0.208 & \begin{tabular}{l}
 -x.0 + 1 >= 0 \\
@@
 b.0 + c.0 + 1 >= 0\\
-\end{tabular} & 0.916s & \begin{tabular}{l}
+\end{tabular}& $\subset$ & 0.916s & \begin{tabular}{l}
 -b.0 + 1 >= 0\\
@@
 b.0 + c.0 + 1 >= 0\\
-\end{tabular} & 36.300s & \begin{tabular}{l}
+\end{tabular}& $\subset$ & 36.30s & \begin{tabular}{l}
 -b.0 + 1 >= 0 \\
@@
 b.0 + 1 >= 0\\
-\end{tabular} \\ \hline
+\end{tabular}\\ \hline
 
@@
 
-	To Be Completed\dots
+	\subsubsection{Analysis of cBench programs}
+	Collective Benchmark (cBench) \cite{cbench} 
+	is a collection of open-source sequential
+	programs assembled by the community to enable realistic
+	benchmarking and research on program.
+
+	We compare \emph{Lookahead Widening} and \emph{Path Focusing} techniques on
+	these benchmarks, and we respectively count the number of BasicBlocks where
+	the invariant is more precise with the first technique, or with the second
+	one, or if it the invariant is the same (see Table \ref{cbench}).
+
+	\emph{Path Focusing} technique finds more precise invariants in some cases,
+	whereas \emph{Lookahead Widening} performs better in other cases.
+	This is partly due to the widening operator: even if the analysis is more
+	subtle, the result at the end may be worse.
+
+\begin{table}[!h]
+\centering
+\begin{tabular}{|c||c|c|c|c|} \hline 
+Benchmark & Path Focusing & Lookahead Widening & Same invariant & Total \\
+\hline \hline
+automotive\_bitcount & 1 & 0 & 39 & 40 \\ \hline
+automotive\_qsort1 & 7 & 3 & 6 & 16 \\ \hline
+automotive\_susan\_c & 0 & 0 & 3 & 3 \\ \hline
+automotive\_susan\_e & 0 & 0 & 3 & 3 \\ \hline
+automotive\_susan\_s & 0 & 0 & 3 & 3 \\ \hline
+bzip2d & 8 & 0 & 22 & 30 \\ \hline
+bzip2e & 8 & 0 & 22 & 30 \\ \hline
+consumer\_jpeg\_c & 25 & 9 & 694 & 728 \\ \hline
+consumer\_jpeg\_d & 22 & 9 & 661 & 692 \\ \hline
+consumer\_lame & 24 & 30 & 361  & 415 \\ \hline
+consumer\_mad & 9 & 20 & 567 & 596 \\ \hline
+consumer\_tiff2bw & 10 & 24 & 561 & 595 \\ \hline
+consumer\_tiff2rgba & 8 & 25 & 457 & 490\\ \hline
+consumer\_tiffdither & 9 & 23 & 560 & 592 \\ \hline
+consumer\_tiffmedian & 10 & 24 & 609 & 643 \\ \hline
+network\_patricia & 2 & 0 & 21 & 23 \\ \hline
+office\_rsynth & 8 & 1 & 157 & 166\\ \hline
+office\_stringsearch1 & 2 & 2 & 46 & 50\\ \hline
+security\_blowfish\_d & 0 & 1 & 26 & 27\\ \hline
+security\_blowfish\_e & 0 & 1 & 26 & 27\\ \hline
+security\_rijndael\_d & 2 & 0 & 30 & 32\\ \hline
+security\_rijndael\_e & 2 & 0 & 30 & 32\\ \hline
+security\_sha  & 1 & 0 & 26 & 27\\ \hline
+telecom\_adpcm\_c & 0 & 0 & 10 & 10\\ \hline
+telecom\_adpcm\_d & 0 & 0 & 10 & 10\\ \hline
+telecom\_CRC32 & 0 & 0 & 13 & 13\\ \hline
+telecom\_gsm & 4 & 1 & 88 & 93\\ \hline
+\end{tabular}
+\caption{Number of basicblocks where the computed invariant is more precise with
+\emph{Path Focusing}, \emph{Lookahead Widening}, or the same with both
+techniques.}
+\label{cbench}
+\end{table}
+
+	\subsubsection{Analysis of Open Source projects}
 

### 2011-06-01 src/AI.cc

@@ -912,8 +912,10 @@ void AI::visitPHINode (PHINode &I){
 					*Out << "\n";
 				);
 				set_ap_expr(&I,expr);
-				ap_environment_t * env = expr->env;
-				insert_env_vars_into_node_vars(env,n,(Value*)&I);
+				if (LV->isLiveThroughBlock(&I,n->bb) || LV->isUsedInBlock(&I,n->bb)) {
+					ap_environment_t * env = expr->env;
+					insert_env_vars_into_node_vars(env,n,(Value*)&I);
+				}
 			}
 		}
 	}

### 2011-06-01 src/Compare.cc

@@ -27,11 +27,18 @@ void Compare::getAnalysisUsage(AnalysisUsage &AU) const {
 }
 
 int Compare::compareAbstract(Abstract * A, AbstractGopan * B) {
-	ap_environment_t * lcenv = common_environment(
+	ap_environment_t * cenv = intersect_environment(
 			A->main->env,
 			B->main->env);
-	A->change_environment(lcenv);
-	B->change_environment(lcenv);
+	
+	*Out << "ABSTRACT VALUES:\n";
+	A->print();
+	B->print(true);
+	A->change_environment(cenv);
+	B->change_environment(cenv);
+	*Out << "NEW ABSTRACT VALUES:\n";
+	A->print();
+	B->print(true);
 
 	if (A->is_eq(B)) {
 		return 0;
@@ -76,6 +83,9 @@ bool Compare::runOnModule(Module &M) {
 						PF++;
 						break;
 					case -1:
+						*Out << "LW is better in " << *b << "\n";
+						n->Y->print();
+						n->Xgopan->print(true);
 						LW++;
 						break;
 					default:

### 2011-06-01 src/Expr.cc

@@ -103,6 +103,32 @@ ap_environment_t * common_environment(
 }
 
 
+ap_environment_t * intersect_environment(
+	ap_environment_t * env1, 
+	ap_environment_t * env2) {
+	ap_environment_t * lcenv = common_environment(env1,env2);
+	ap_environment_t * intersect = ap_environment_copy(lcenv);	
+
+	for (size_t i = 0; i < lcenv->intdim + lcenv->realdim; i++) {
+		*Out << "Hello\n";
+		ap_var_t var = ap_environment_var_of_dim(lcenv,(ap_dim_t)i);
+		*Out << "Hello2\n";
+		if (!ap_environment_mem_var(env1,var) || !ap_environment_mem_var(env2,var)) {
+		*Out << "Hello3 " << ap_var_to_string(var) << "\n";
+			size_t size = intersect->intdim + intersect->realdim;
+			*Out << "Removing variable " << ap_var_to_string(var) << " from the environment\n";
+			intersect = ap_environment_remove(intersect,&var,1);
+			*Out << "Removing variable " << ap_var_to_string(var) << " ok : intersect=" << intersect << "\n";
+		}	
+	}	
+
+	*Out << "RESULT OF THE INTERSECTION\n";
+	environment_print(intersect);	
+	return intersect;
+}
+
+
+
 /// common_environment = modifies exp1 and exp2 such that 
 /// they have the same common environment
 ///

### 2011-06-01 src/Expr.h

@@ -29,6 +29,11 @@ ap_environment_t * common_environment(
 		ap_environment_t * env1,
 		ap_environment_t * env2);
 
+/// intersect_environment - compute the intersection of the two environments.
+ap_environment_t * intersect_environment(
+		ap_environment_t * env1,
+		ap_environment_t * env2);
+
 /// common_environment - modifies the two expression by giving them the same
 /// least common environment.
 void common_environment(ap_texpr1_t ** exp1, ap_texpr1_t ** exp2);

### 2011-06-03 report/report.tex
update of the report + minor changes in the code

@@
 
+	These benchmarks often contain non-linear arithmetic, especially
+	multiplications. This is not a problem to the Apron library since it uses
+	linearization to handle them, but this sometimes implies an execution time
+	blowup for the SMT-solver:
+\begin{itemize}
+\item \emph{Yices} does not work at all when the formula contains non-linear
+arithmetic, so we could not use it. Indeed, Linear Real Arithmetic or Linear
+Integer Arithmetic theories do not treat multiplications.
+\item \emph{Z3} implements some linearization techniques 
+in order to deal with non-linear
+arithmetic, even if the used theory is LRA or LIA. In lots of cases, it is
+sufficient for deciding the satisfiability of our formulae, but it seems
+sometimes to have a huge execution time blowup. 
+When running our benchmarks, we fixed a time limit (5 minutes) for the analysis,
+and ignore in our results the functions that make the SMT-solver blow.
+\end{itemize}
+	This drawback could be solved in the future, by linearizing parts of the
+	SMT-formula that make the computation too costly, or make it fail (in some
+	cases, the SMT-solver is unable to decide the satisfiability of the formula,
+	and answers ``unknown'').
+
 	\emph{Path Focusing} technique finds more precise invariants in some cases,
@@
 automotive\_bitcount	& 1 & 0 & 39 & 40 \\ \hline
-automotive\_qsort1 & 7 & 3 & 6 & 16 \\ \hline
+automotive\_qsort1		& 7 & 0 & 9 & 16 \\ \hline
 automotive\_susan\_c	& 0 & 0 & 3 & 3 \\ \hline
@@
 bzip2e					& 8 & 0 & 22 & 30 \\ \hline
-consumer\_jpeg\_c & 25 & 9 & 694 & 728 \\ \hline
-consumer\_jpeg\_d & 22 & 9 & 661 & 692 \\ \hline
-consumer\_lame & 24 & 30 & 361  & 415 \\ \hline
-consumer\_mad & 9 & 20 & 567 & 596 \\ \hline
-consumer\_tiff2bw & 10 & 24 & 561 & 595 \\ \hline
-consumer\_tiff2rgba & 8 & 25 & 457 & 490\\ \hline
-consumer\_tiffdither & 9 & 23 & 560 & 592 \\ \hline
-consumer\_tiffmedian & 10 & 24 & 609 & 643 \\ \hline
+consumer\_jpeg\_c		& 35 & 0 & 693 & 728 \\ \hline
+consumer\_jpeg\_d		& 30 & 0 & 662 & 692 \\ \hline
+consumer\_lame			& 31 & 12 & 372  & 415 \\ \hline
+consumer\_mad			& 9 & 8 & 579 & 596 \\ \hline
+consumer\_tiff2bw		& 18 & 1 & 576 & 595 \\ \hline
+consumer\_tiff2rgba		& 19 & 1 & 574 & 594\\ \hline
+consumer\_tiffdither	& 17 & 1 & 574 & 592 \\ \hline
+consumer\_tiffmedian	& 21 & 2 & 620 & 643 \\ \hline
 network\_patricia		& 2 & 0 & 21 & 23 \\ \hline
-office\_rsynth & 8 & 1 & 157 & 166\\ \hline
-office\_stringsearch1 & 2 & 2 & 46 & 50\\ \hline
-security\_blowfish\_d & 0 & 1 & 26 & 27\\ \hline
-security\_blowfish\_e & 0 & 1 & 26 & 27\\ \hline
+office\_rsynth			& 8 & 0 & 158 & 166\\ \hline
+office\_stringsearch1	& 5 & 1 & 44 & 50\\ \hline
+security\_blowfish\_d	& 4 & 0 & 23 & 27\\ \hline
+security\_blowfish\_e	& 4 & 0 & 23 & 27\\ \hline
 security\_rijndael\_d	& 2 & 0 & 30 & 32\\ \hline
@@
 telecom\_CRC32			& 0 & 0 & 13 & 13\\ \hline
-telecom\_gsm & 4 & 1 & 88 & 93\\ \hline
+telecom\_gsm			& 4 & 0 & 89 & 93\\ \hline
 \end{tabular}
-\caption{Number of basicblocks where the computed invariant is more precise with
+\caption{Number of basicblocks where the computed invariant is more precise by
 \emph{Path Focusing}, \emph{Lookahead Widening}, or the same with both
@@
 	\subsubsection{Analysis of Open Source projects}
+
+We compare \emph{Lookahead Widening} and \emph{Path Focusing} on various
+GNU projects ( see Table \ref{opensource}), so that we can see their precision
+on really used code.
+
+\begin{table}
+\centering
+\begin{tabular}{|c||c|c|c|c|} \hline 
+Benchmark & Path Focusing & Lookahead Widening & Same invariant & Total \\
+\hline \hline
+Bc			& 0 & 0 & 177 & 177 \\ \hline
+Gawk		& 4 & 0 & 287 & 291 \\ \hline
+Gnuchess	& 16 & 16 & 1447 & 1479 \\ \hline
+Gnugo		& 120 & 10 & 1286 & 1416 \\ \hline
+Grep		& 4 & 1 & 326 & 331 \\ \hline
+Gzip		& 16 & 8 & 207 & 231 \\ \hline
+Make		&  5 & 5 & 304 & 314 \\ \hline
+Tar			& 13 & 2 & 437 & 452 \\ \hline
+Wget		& 20 & 3 & 751 & 774 \\ \hline
+\end{tabular}
+\caption{Result of the analysis of various open-source projects.}
+\label{opensource}
+\end{table}
 

### 2011-06-03 ex/analyzer-bc.sh

@@ -11,6 +11,7 @@ OPTIONS :
 	-o [FILE ]: name of the IR generated file
 	-p        : print the optimized llvm bytecode
 	-u        : unroll loops once
+	-c        : compare both techniques
 	-G        : generate the .dot CFG
 	-g        : use Lookahead Widening instead of Path Focusing
 	-y        : use Yices instead of Microsoft Z3
@@ -22,10 +23,10 @@ GRAPH=0
 GOPAN=0
 YICES=0
 UNROLL=0
-COMPILE_OPTIONS=
+COMPARE=0
 RESULT=
 
-while getopts “hpygruGi:o:c:” opt ; do
+while getopts “hpygruGi:o:c” opt ; do
 	case $opt in
 		h)
 			usage
@@ -39,6 +40,9 @@ while getopts “hpygruGi:o:c:” opt ; do
 			;;
 		u)
 			UNROLL=1
+			;;
+		c)
+			COMPARE=1
 			;;
 		G)
 			GRAPH=1
@@ -93,14 +97,16 @@ fi
 NAME=`basename $OUTPUT`
 RESULT=/tmp/${NAME%%.*}.result
 echo "running analyzer on $NAME"
-
-if [ $GOPAN -eq 1 ] ; then
-	/home/jhenry/m2r/src/analyzer -g -i $OUTPUT
+if [ $COMPARE -eq 1 ] ; then
+	/home/jhenry/m2r/src/analyzer -c -i $OUTPUT
 else
-	if [ $YICES -eq 1 ] ; then
-		/home/jhenry/m2r/src/analyzer -y -i $OUTPUT
+	if [ $GOPAN -eq 1 ] ; then
+		/home/jhenry/m2r/src/analyzer -g -i $OUTPUT
 	else
-		/home/jhenry/m2r/src/analyzer -i $OUTPUT
+		if [ $YICES -eq 1 ] ; then
+			/home/jhenry/m2r/src/analyzer -y -i $OUTPUT
+		else
+			/home/jhenry/m2r/src/analyzer -i $OUTPUT
+		fi
 	fi
 fi
-

### 2011-06-03 src/AI.cc

@@ -448,6 +448,8 @@ void AI::computeNode(Node * n) {
 		if (LI->isLoopHeader(Succ->bb) && (Succ != n) && !only_join) {
 				//if (Succ->widening == 1) {
 				Xtemp->widening(Succ);
+				lastpath[Succ->bb].clear();
+				lastpath[n->bb].clear();
 				//Xtemp->widening_threshold(Succ, &linconstraints);
 				//Succ->widening = 0;
 			//	} else {

### 2011-06-03 src/Compare.cc

@@ -31,14 +31,8 @@ int Compare::compareAbstract(Abstract * A, AbstractGopan * B) {
 			A->main->env,
 			B->main->env);
 	
-	*Out << "ABSTRACT VALUES:\n";
-	A->print();
-	B->print(true);
 	A->change_environment(cenv);
 	B->change_environment(cenv);
-	*Out << "NEW ABSTRACT VALUES:\n";
-	A->print();
-	B->print(true);
 
 	if (A->is_eq(B)) {
 		return 0;

### 2011-06-03 src/Expr.cc

@@ -110,20 +110,12 @@ ap_environment_t * intersect_environment(
 	ap_environment_t * intersect = ap_environment_copy(lcenv);	
 
 	for (size_t i = 0; i < lcenv->intdim + lcenv->realdim; i++) {
-		*Out << "Hello\n";
 		ap_var_t var = ap_environment_var_of_dim(lcenv,(ap_dim_t)i);
-		*Out << "Hello2\n";
 		if (!ap_environment_mem_var(env1,var) || !ap_environment_mem_var(env2,var)) {
-		*Out << "Hello3 " << ap_var_to_string(var) << "\n";
 			size_t size = intersect->intdim + intersect->realdim;
-			*Out << "Removing variable " << ap_var_to_string(var) << " from the environment\n";
 			intersect = ap_environment_remove(intersect,&var,1);
-			*Out << "Removing variable " << ap_var_to_string(var) << " ok : intersect=" << intersect << "\n";
 		}	
 	}	
-
-	*Out << "RESULT OF THE INTERSECTION\n";
-	environment_print(intersect);	
 	return intersect;
 }
 

### 2011-06-03 src/z3_manager.cc

@@ -202,14 +202,14 @@ SMT_expr z3_manager::SMT_mk_ge (SMT_expr a1, SMT_expr a2){
 }
 
 void z3_manager::SMT_print(SMT_expr a){
-	*Out << Z3_benchmark_to_smtlib_string(ctx,
-				"name",
-				"logic",
-				"unknown",
-				"",
-				0,
-				NULL,
-				(Z3_ast)a);
+//	*Out << Z3_benchmark_to_smtlib_string(ctx,
+//				"name",
+//				"logic",
+//				"unknown",
+//				"",
+//				0,
+//				NULL,
+//				(Z3_ast)a);
 }
 
 int z3_manager::SMT_check(SMT_expr a, std::set<std::string> * true_booleans){
@@ -239,58 +239,58 @@ int z3_manager::SMT_check(SMT_expr a, std::set<std::string> * true_booleans){
 				Z3_eval_func_decl (ctx,m,decl,&v);
 				Z3_symbol symbol = Z3_get_decl_name(ctx,decl);
 				std::string name (Z3_get_symbol_string (ctx,symbol));
-				DEBUG(
-				*Out << name << " = ";
-				);
+				//DEBUG(
+				//*Out << name << " = ";
+				//);
 				Z3_sort_kind sort = Z3_get_sort_kind(ctx,Z3_get_sort(ctx,v));
 
 				switch (sort) {
 					case Z3_BOOL_SORT: 
 						switch (Z3_get_bool_value(ctx,v)) {
 							case Z3_L_FALSE:
-								DEBUG(
-									*Out << "false\n";
-								);
+								//DEBUG(
+								//	*Out << "false\n";
+								//);
 								break;
 							case Z3_L_UNDEF:
-								DEBUG(
-								*Out << "undef\n";
-								);
+								//DEBUG(
+								//*Out << "undef\n";
+								//);
 								break;
 							case Z3_L_TRUE:
-								DEBUG(
-								*Out << "true\n";
-								);
+								//DEBUG(
+								//*Out << "true\n";
+								//);
 								true_booleans->insert(name);
 								break;
 						}
 						break;
 					case Z3_INT_SORT:
-						DEBUG(
-						int i;
-						Z3_get_numeral_int (ctx,v,&i);
-						*Out << i << "\n";
-						);
+						//DEBUG(
+						//int i;
+						//Z3_get_numeral_int (ctx,v,&i);
+						//*Out << i << "\n";
+						//);
 						break;
 					case Z3_REAL_SORT:
[…]

### 2011-06-06 src/AI.cc
- Path Focusing: now, the set of already seen paths is cleared at the beginning   of computeNode.   => Better results !

@@ -365,7 +365,8 @@ void AI::computeNode(Node * n) {
 		Out->resetColor();
 		*Out << *b << "\n";
 	);
-
+		//if (!isequal(path,lastpath[n->bb])) {
+	pathtree->clear();
 
 	while (true) {
 		is_computed[n] = true;
@@ -397,15 +398,15 @@ void AI::computeNode(Node * n) {
 		Succ = Nodes[path.back()];
 
 		n_iterations++;
-		if (!pathtree->exist(path)) n_paths++;
-
-		if (!isequal(path,lastpath[n->bb])) {
+		//if (!isequal(path,lastpath[n->bb])) {
+		if (!pathtree->exist(path)) {
+			n_paths++;
 			only_join = true;
 		} else {
 			only_join = false;
 		}
-		lastpath[n->bb].clear();
-		lastpath[n->bb].assign(path.begin(),path.end());
+		//lastpath[n->bb].clear();
+		//lastpath[n->bb].assign(path.begin(),path.end());
 
 		// computing the image of the abstract value by the path's tranformation
 		Xtemp = new Abstract(n->X);

### 2011-06-06 src/PathTree.cc

@@ -38,6 +38,13 @@ void PathTree::insert(std::list<BasicBlock*> path) {
 	}
 }
 
+void PathTree::clear() {
+	while (!start.name.empty()) 
+		start.name.pop_back();
+	while (!start.next.empty()) 
+		start.next.pop_back();
+}
+
 bool PathTree::exist(std::list<BasicBlock*> path) {
 	std::list<BasicBlock*> workingpath;
 	pathnode * v;

### 2011-06-06 src/PathTree.h

@@ -25,6 +25,8 @@ class PathTree {
 
 		void insert(std::list<BasicBlock*> path);
 
+		void clear();
+
 		bool exist(std::list<BasicBlock*> path);
 };
 

### 2011-06-06 src/z3_manager.cc
new version of the slides for the GDR GPL

@@ -215,14 +215,14 @@ SMT_expr z3_manager::SMT_mk_ge (SMT_expr a1, SMT_expr a2){
 }
 
 void z3_manager::SMT_print(SMT_expr a){
-	*Out << Z3_benchmark_to_smtlib_string(ctx,
-				"name",
-				"logic",
-				"unknown",
-				"",
-				0,
-				NULL,
-				(Z3_ast)a);
+	//*Out << Z3_benchmark_to_smtlib_string(ctx,
+	//			"name",
+	//			"logic",
+	//			"unknown",
+	//			"",
+	//			0,
+	//			NULL,
+	//			(Z3_ast)a);
 }
 
 int z3_manager::SMT_check(SMT_expr a, std::set<std::string> * true_booleans){
@@ -252,58 +252,58 @@ int z3_manager::SMT_check(SMT_expr a, std::set<std::string> * true_booleans){
 				Z3_eval_func_decl (ctx,m,decl,&v);
 				Z3_symbol symbol = Z3_get_decl_name(ctx,decl);
 				std::string name (Z3_get_symbol_string (ctx,symbol));
-				DEBUG(
-				*Out << name << " = ";
-				);
+				//DEBUG(
+				//*Out << name << " = ";
+				//);
 				Z3_sort_kind sort = Z3_get_sort_kind(ctx,Z3_get_sort(ctx,v));
 
 				switch (sort) {
 					case Z3_BOOL_SORT: 
 						switch (Z3_get_bool_value(ctx,v)) {
 							case Z3_L_FALSE:
-								DEBUG(
-									*Out << "false\n";
-								);
+								//DEBUG(
+								//	*Out << "false\n";
+								//);
 								break;
 							case Z3_L_UNDEF:
-								DEBUG(
-								*Out << "undef\n";
-								);
+								//DEBUG(
+								//*Out << "undef\n";
+								//);
 								break;
 							case Z3_L_TRUE:
-								DEBUG(
-								*Out << "true\n";
-								);
+								//DEBUG(
+								//*Out << "true\n";
+								//);
 								true_booleans->insert(name);
 								break;
 						}
 						break;
 					case Z3_INT_SORT:
-						DEBUG(
-						int i;
-						Z3_get_numeral_int (ctx,v,&i);
-						*Out << i << "\n";
-						);
+						//DEBUG(
+						//int i;
+						//Z3_get_numeral_int (ctx,v,&i);
+						//*Out << i << "\n";
+						//);
 						break;
[…]

### 2011-06-09 report/report.tex
report update

@@
 \tikzstyle{state}=[circle,fill=black!25,minimum size=13pt,inner sep=0pt]
+\tikzstyle{rstate}=[rectangle,fill=black!25,minimum size=13pt,inner sep=0pt]
 \tikzstyle{transition}=[rectangle,semithick,draw=black!75,
@@
 
+\lstnewenvironment{Csmall}
+{\lstset{language=C,
+		basicstyle=\ttfamily\small,
+		commentstyle=\color{moka}\textit,
+		keywordstyle=\color{minuit},
+		identifierstyle=\color{trefle},
+		showstringspaces=false}}
+{}
 \title{Static Analysis by Path Focusing}
@@
 \tutor{David Monniaux - Matthieu Moy}
-\notes{some notes}
+%\notes{some notes}
 \webaddress{Julien.Henry@imag.fr}
 \reporttype{M2R internship's report}
+
+\acknowledgement{
+	I would like to thank my two tutors, David Monniaux and Matthieu Moy, for
+	their direction of my master thesis, and for their help during all my
+	internship. I thank them for all their explanations that clearified lots of
+	my questions, and for their advices for the redaction of this report.
+}
 
@@
 	\State $Y \gets \tau_e^\#(X_{p_i})$
-	\State $X' \gets X_{p_i} \sqcup Y$
+	\State $X' \gets X_{p_i}\widening (X_{p_i} \sqcup Y)$
 	\State $Y \gets \tau_e^\#(X')$
@@
 	The $\rho$ formula associated to this program is depicted in
-	Figure \ref{rhoformula}.
+	Figure \ref{rhoformula} (see appendices).
 	One computes
@@
 \hline \hline
+%automotive\_bitcount	& 1 & 0 & 39 & 40 \\ \hline
+%automotive\_qsort1		& 7 & 0 & 9 & 16 \\ \hline
+%automotive\_susan\_c	& 0 & 0 & 3 & 3 \\ \hline
+%automotive\_susan\_e	& 0 & 0 & 3 & 3 \\ \hline
+%automotive\_susan\_s	& 0 & 0 & 3 & 3 \\ \hline
+%bzip2d					& 8 & 0 & 22 & 30 \\ \hline
+%bzip2e					& 8 & 0 & 22 & 30 \\ \hline
+%consumer\_jpeg\_c		& 35 & 0 & 693 & 728 \\ \hline
+%consumer\_jpeg\_d		& 30 & 0 & 662 & 692 \\ \hline
+%consumer\_lame			& 31 & 12 & 372  & 415 \\ \hline
+%consumer\_mad			& 9 & 8 & 579 & 596 \\ \hline
+%consumer\_tiff2bw		& 18 & 1 & 576 & 595 \\ \hline
+%consumer\_tiff2rgba		& 19 & 1 & 574 & 594\\ \hline
+%consumer\_tiffdither	& 17 & 1 & 574 & 592 \\ \hline
+%consumer\_tiffmedian	& 21 & 2 & 620 & 643 \\ \hline
+%network\_patricia		& 2 & 0 & 21 & 23 \\ \hline
+%office\_rsynth			& 8 & 0 & 158 & 166\\ \hline
+%office\_stringsearch1	& 5 & 1 & 44 & 50\\ \hline
+%security\_blowfish\_d	& 4 & 0 & 23 & 27\\ \hline
+%security\_blowfish\_e	& 4 & 0 & 23 & 27\\ \hline
+%security\_rijndael\_d	& 2 & 0 & 30 & 32\\ \hline
+%security\_rijndael\_e	& 2 & 0 & 30 & 32\\ \hline
+%security\_sha			& 1 & 0 & 26 & 27\\ \hline
+%telecom\_adpcm\_c		& 0 & 0 & 10 & 10\\ \hline
+%telecom\_adpcm\_d		& 0 & 0 & 10 & 10\\ \hline
+%telecom\_CRC32			& 0 & 0 & 13 & 13\\ \hline
+%telecom\_gsm			& 4 & 0 & 89 & 93\\ \hline
+%%%%%%%%%%%%
 automotive\_bitcount & 1 & 0 & 39 & 40 \\ \hline
-automotive\_qsort1		& 7 & 0 & 9 & 16 \\ \hline
-automotive\_susan\_c	& 0 & 0 & 3 & 3 \\ \hline
-automotive\_susan\_e	& 0 & 0 & 3 & 3 \\ \hline
-automotive\_susan\_s	& 0 & 0 & 3 & 3 \\ \hline
-bzip2d					& 8 & 0 & 22 & 30 \\ \hline
-bzip2e					& 8 & 0 & 22 & 30 \\ \hline
-consumer\_jpeg\_c		& 35 & 0 & 693 & 728 \\ \hline
-consumer\_jpeg\_d		& 30 & 0 & 662 & 692 \\ \hline
-consumer\_lame			& 31 & 12 & 372  & 415 \\ \hline
-consumer\_mad			& 9 & 8 & 579 & 596 \\ \hline
-consumer\_tiff2bw		& 18 & 1 & 576 & 595 \\ \hline
-consumer\_tiff2rgba		& 19 & 1 & 574 & 594\\ \hline
-consumer\_tiffdither	& 17 & 1 & 574 & 592 \\ \hline
-consumer\_tiffmedian	& 21 & 2 & 620 & 643 \\ \hline
+automotive\_qsort1 & 9 & 0 & 13 & 22 \\ \hline
+automotive\_susan\_c & 0 & 2 & 59 & 61 \\ \hline
+automotive\_susan\_e & 0 & 2 & 59 & 61 \\ \hline
+automotive\_susan\_s & 0 & 2 & 59 & 61 \\ \hline
+bzip2d & 70 & 3 & 81 & 154 \\ \hline
+bzip2e & 70 & 3 & 81 & 154 \\ \hline
+consumer\_jpeg\_c & 22 & 8 & 711 & 741 \\ \hline
+consumer\_jpeg\_d & 13 & 8 & 667 & 688 \\ \hline
+consumer\_lame & 13 & 15 & 366 & 394 \\ \hline
+consumer\_mad & 12 & 11 & 609 & 632 \\ \hline
+consumer\_tiff2bw & 19 & 3 & 593 & 615 \\ \hline
+consumer\_tiff2rgba & 21 & 6 & 661 & 688 \\ \hline
+consumer\_tiffdither & 21 & 3 & 662 & 686 \\ \hline
+consumer\_tiffmedian & 25 & 5 & 707 & 737 \\ \hline
+network\_dijkstra & 0 & 0 & 19 & 19 \\ \hline
 network\_patricia & 2 & 0 & 21 & 23 \\ \hline
-office\_rsynth			& 8 & 0 & 158 & 166\\ \hline
-office\_stringsearch1	& 5 & 1 & 44 & 50\\ \hline
-security\_blowfish\_d	& 4 & 0 & 23 & 27\\ \hline
-security\_blowfish\_e	& 4 & 0 & 23 & 27\\ \hline
-security\_rijndael\_d	& 2 & 0 & 30 & 32\\ \hline
-security\_rijndael\_e	& 2 & 0 & 30 & 32\\ \hline
-security\_sha			& 1 & 0 & 26 & 27\\ \hline
-telecom\_adpcm\_c		& 0 & 0 & 10 & 10\\ \hline
-telecom\_adpcm\_d		& 0 & 0 & 10 & 10\\ \hline
-telecom\_CRC32			& 0 & 0 & 13 & 13\\ \hline
-telecom\_gsm			& 4 & 0 & 89 & 93\\ \hline
+office\_rsynth & 6 & 0 & 158 & 164 \\ \hline
+office\_stringsearch1 & 5 & 1 & 44 & 50 \\ \hline
+security\_blowfish\_d & 4 & 0 & 23 & 27 \\ \hline
+security\_blowfish\_e & 4 & 0 & 23 & 27 \\ \hline
+security\_rijndael\_d & 1 & 0 & 31 & 32 \\ \hline
+security\_rijndael\_e & 1 & 0 & 31 & 32 \\ \hline
+security\_sha & 1 & 0 & 26 & 27 \\ \hline
+telecom\_CRC32 & 0 & 0 & 13 & 13 \\ \hline
+telecom\_adpcm\_c & 0 & 0 & 10 & 10 \\ \hline
+telecom\_adpcm\_d & 0 & 0 & 10 & 10 \\ \hline
+telecom\_gsm & 5 & 0 & 153 & 158 \\ \hline
 \end{tabular}
@@
 
+The results we obtain are very promising. Most of the cases, our technique
+performs as well as \emph{Lookahead Widening}, and discovers regularly more
+precise invariants.
+
+However, \emph{Lookahead Widening} seems to have better results in some cases.
+We already have some ideas for explaining this lack of precision in our
+technique:
+
+\begin{itemize}
+\item We use the domain of convex polyhedra as the abstract domain. Since this
+domain requires a widening operator, a more subtle analysis technique may
+sometimes discover invariants that are less precise than a more simplified
+technique.
+\item Our technique performs well in the case of self loops in the multigraph,
+thanks to the widening/narrowing phases (or acceleration). Yet, in the case of
+paths that do not constitute a self loop in the multigraph, one perform a
+widening operation, but no narrowing iteration. Apparently, lots of the cases
+where \emph{Lookahead Widening} performs better than \emph{Path Focusing} are
+due to this drawback. An example of such loss of precision is depicted in
+appendix \ref{lostprecision}.
+An good idea that could solve this problem is to try to combine
+the two methods.
+\end{itemize}
+
+Although these project are not safety-critical and do not necessarily need
+static analysis, they give a good idea of the quality of our technique and the
+scalability of our implementation.
+
 \begin{table}
@@
 Bc			& 0 & 0 & 177 & 177 \\ \hline
-Gawk		& 4 & 0 & 287 & 291 \\ \hline
-Gnuchess	& 16 & 16 & 1447 & 1479 \\ \hline
-Gnugo		& 120 & 10 & 1286 & 1416 \\ \hline
-Grep		& 4 & 1 & 326 & 331 \\ \hline
-Gzip		& 16 & 8 & 207 & 231 \\ \hline
-Make		&  5 & 5 & 304 & 314 \\ \hline
-Tar			& 13 & 2 & 437 & 452 \\ \hline
-Wget		& 20 & 3 & 751 & 774 \\ \hline
+Gawk		& 4 & 3 & 284 & 291 \\ \hline
+Gnuchess	& 22 & 33 & 1506 & 1561 \\ \hline
+Gnugo		& 105 & 35 & 1303 & 1443 \\ \hline
+Grep		& 4 & 3 & 323 & 330 \\ \hline
+Gzip		& 9 & 3 & 189 & 201 \\ \hline
+Make		&  7 & 11 & 457 & 475 \\ \hline
+Tar			& 16 & 5 & 555 & 576 \\ \hline
+Wget		& 8 & 12 & 715 & 735 \\ \hline
 \end{tabular}
@@
   \appendix
-\begin{figure}
+\appendixpage
+\renewcommand{\thesection}{\Alph{section}}
+\section{Generated $\rho$ formula}
+\label{rhoformula}
+\begin{figure}[!h]
 \begin{C}
@@
 \ref{multigraphGopan}, in the SMT-lib format.}
-\label{rhoformula}
 \end{figure}
+\FloatBarrier
+
+\newpage
+\section{Example of loss of precision in Path Focusing}
+\label{lostprecision}
+The following graph is a simplified example taken from one of our benchmarks,
+where \emph{Path Focusing} technique obtains bad results.
+\begin{figure}[!h]
+\centering
+\begin{tikzpicture}[->,>=stealth',auto,node distance=2.5cm,
+                    semithick,font=\footnotesize]
+
+	\node[rstate] (n0) {$p_0$};
+	\node[rstate] (n1) [below of=n0] {\begin{tabular}{l}
+	$p_1$: \\
+	$i \gets \Phi(0,i')$
+	\end{tabular}};
+	\node[rstate] (n2) [below left of=n1] {$p_2$};
+	\node[rstate] (n3) [below right of=n2] {$p_3$};
+
+  \path [transition] 
+		(n0) edge              node {} (n1);
+  \path [transition] 
+        (n1) edge			   node [left] {$i \leq 50$} (n2);
+  \path [transition] 
+        (n2) edge	[loop left, distance=2cm]	   node  {} (n2);
+  \path [transition] 
+        (n2) edge			   node  {} (n3);
+  \path [transition] 
+        (n3) edge			   node [right] {$i' \gets i+1$} (n1);
+\end{tikzpicture}
+\end{figure}
+
+$p_1$ and $p_2$ are both loop headers. We have $P_R = \{p_1, p_2\}$.
+
+\begin{itemize}
+\item We start with $p_1$ : $i=0$. The SMT-solver finds the path $p_1 \rightarrow p_2$.
+\item At point $p_2$ : $i=0$. Then, we focus on the path $p_2 \rightarrow p_3 \rightarrow p_1$.
+The obtained invariant in $p_1$ becomes $i \in [0,1]$. 
+Since this is a loop header, widening is applied: $i \geq 0$.
+\item The new focus path is $p_1 \rightarrow p_2$. The image of the polyhedron $i
+\geq 0$ by this transition is $0 \leq i \leq 50$. Again, $p_2$  is a loop
+header, and widening gives $i \geq 0$.
+\end{itemize}
+In this example, we do not have narrowing iterations since we never focus on self
+loops.
 

### 2011-06-12 report/report.tex
report ~final

@@
 	I would like to thank my two tutors, David Monniaux and Matthieu Moy, for
-	their direction of my master thesis, and for their help during all my
-	internship. I thank them for all their explanations that clearified lots of
-	my questions, and for their advices for the redaction of this report.
+	their direction of my master thesis, and for their help everytime I needed
+	it. I thank them for all their explanations that clearified lots of
+	my questions, and for their advices for the redaction of this report.\\
+	I also thank Laure Gonnord for inviting me to present my work at the GDR GPL
+	2011.\\
+	Finally, I thank all the Synchrone team and the Verimag laboratory for their
+	cheerful welcome.
 }

### 2011-06-13 report/report.tex
report

@@
 \usepackage[toc,page]{appendix} 
-% DM: verimagstudent charge hyperref, et hyperref doit habituellement
-% DM: passer en dernier
 \newtheorem{definition}{Definition}[section]
@@
 \abstract{
-	Program verification is aims at statically discovering properties on programs,
+	Program verification is aimed at statically discovering properties on programs,
 	such as the values that can take the different variables during execution.
@@
 A fundamental fact in static analysis is that it cannot be perfectly precise:
-either the set of possible states we compute is unsound (it doesn't contain all
-the possible states during execution), or it is incomplete (it contains
-states that can never be reached during execution). 
+either the analysis is unsound (the set of possible states we compute does not contain all
+the possible states during execution), or is incomplete (the computed set
+contains states that can never be reached during execution). 
 Abstract Interpretation is sound but incomplete, since it always computes an
@@
 
-\section{Introductory example}
+\section{Introductive example}
 
@@
 \end{eqnarray*}
-% DM: d'habitude on met des ' sur les variables de l'etat d'arrivee
-% DM: et pas de depart
 We see that $Y_1$ and $Y_2$ can be computed step by step:
@@
 
-The concrete semantics is a monotonic function $\Phi: C \rightarrow C$.
+The concrete semantics is a function $\Phi: C \rightarrow C$.
+In the introductive example, this function $\Phi$ is the following:
+
+$$\Phi\left[ \begin{array}{c}
+Y_1 \\
+Y_2 \\
+Y_3 \\
+Y_4
+\end{array} \right] = \left[ \begin{array}{cc} 
+x\ |& x \in \R \\
+x\ |& x = 0 \vee \exists\  x'\in Y_2, x = x'+1 \\
+x\ |& x \in Y_1 \wedge x < 100 \\
+x\ |& x \in Y_1 \wedge x \geq 100 
+\end{array} \right]$$
+
 One can choose an abstract semantics $\Phi^\#: A \rightarrow A$ satisfying this
@@
 Suppose $C$ is a complete lattice, and $\Phi$ is increasing from $C$ to $C$.
-Each $x \in A$, satisfying the condition $\Phi^\#(x) = x$, is an abstraction of
+Each $x \in A$, satisfying the condition $\Phi^\#(x) \subseteq x$, is an abstraction of
 the least fixpoint of $\Phi$.
@@
 
-Indeed, if $\Phi^\#(x) = x$, then $\Phi \circ \gamma (x) \subseteq \gamma(x)
-= \gamma \circ \Phi^\#(x)$. Abstract interpretation aims at computing
+Indeed, if $\Phi^\#(x) \subseteq x$, then $\Phi \circ \gamma (x) 
+\subseteq \gamma \circ \Phi^\#(x) \subseteq \gamma(x)$.
+Abstract interpretation aims at computing
 such an~$x$. It computes the stationary limit of an ascending sequence
@@
 x_0 &=& \perp \\
-x_{n+1} &=& \Phi^\#(x_n)
+x_{n+1} &=& x_n \sqcup \Phi^\#(x_n)
 \end{array}
@@
 $$
-where $\perp$ is the least element of $A$.
+where $\perp$ is the least element of $A$, and $\sqcup$ is an operator verifying
+$x_1 \cup x_2 \subseteq x_1 \sqcup x_2$ and $x_1 \sqcup x_2 \in A$.
 %The relation between these two
@@
 \item $\forall x_1, x_2 \in A,\ x_1 \leq_A x_1 \widening x_2$ and $x_2 \leq_A x_1
-\widening x_2$. This guarantees the correctness of the result.
-% DM: souvent n'est defini que pour x1<=x2 ce qui oblige a utiliser
-% DM: une union abstraite juste avant
+\widening x_2$. This guarantees the correctness of the result. Most of the time,
+this operator is only defined for $x_1 \leq x_2$. In this case, we use $x_1
+\widening (x_1 \sqcup x_2)$ instead.
 \item For any increasing sequence $x_0 \leq_A x_1 \leq_A \dots$, the sequence
@@
 \cite{Monniaux_HOSC09} gives a more general definition of the widening operator.
-% DM: c'est gentil de me citer!
 
@@
 x'_0 &=& \perp \\
-x'_{i+1} &=& x'_i\ \widening\ \Phi^\#(x'_i),\ \ \forall i \geq 0
+x'_{i+1} &=& x'_i\ \widening\ ( x'_i \sqcup \Phi^\#(x'_i)),\ \ \forall i \geq 0
 }$$
@@
 If $P = \{(x,y)\ |\ 0 \leq x \leq y \leq 1 \}$, after one iteration of the loop,
-we have $Q = \{(x,y)\ |\ 0 \leq x \leq y \leq 2 \}$. Instead of a classic
+we have 
+\\$Q = \{(x,y)\ |\ 0 \leq x \leq y \leq 2 \}$. Instead of a classic
 widening (which gives the result $P \widening Q = \{(x,y)\ |\ 0 \leq x \leq y \}$),

### 2011-06-13 report/report.tex
report

@@
 loop invariants\dots
-The main applications of static analysis are compilation-time optimisation, and
+The main applications of static analysis are compile-time optimisations, and
 the proof of safety properties in critical systems, such as avionics.
@@
 \begin{itemize}
-\item We present a new technique, referred as \emph{Path Focusing}, that takes
+\item We present a new technique, refered as \emph{Path Focusing}, that takes
 benefits of the recent advances in the field of SMT-solving to improve precision
@@
 Static analysis is aimed at discovering some properties about programs. In this
-example, one would compute the set of possible values for the variable $x$
-during execution.
+example, we would like to compute the set of possible values for the variable $x$ during execution.
 The graph at the right side represents the program. The nodes of the graph are
@@
 of the program, i.e the set of all possible states of the program. This
-computation quickly becomes too costly or doesn't terminate.
+computation quickly becomes too costly or does not terminate.
 
@@
 or more generally satisfy the \emph{ascending chain condition},
-meaning that there doesn't exist
+meaning that there does not exist
 any sequence $(x_i)_{i \geq 0}$ of elements of the abstract domain $A$ such that
@@
 
-Many of the domains that are used for program analysis, including those of intervals and convex polyhedra, don't satisfy this ascending chain condition.
-To ensure convergence in this case, another approximation
+Many of the domains that are used for program analysis, including those of
+intervals and convex polyhedra, do not satisfy this ascending chain condition. To ensure convergence in this case, another approximation
 is performed: a new operator is defined, called \emph{widening operator}, that
@@
 Let $x_1, x_2, \dots ,x_n$ be the numerical variables of a program (assume they
-all are in $\Q$). A state of
+ are all in $\Q$). A state of
 the program is then a point $\overrightarrow{x} \in \Q^n$.
@@
 since widening is applied within a finite time after $n$ iterations. In most of
-the cases, $n$ can be chosen equals to $2$.
+the cases, $n$ can be chosen equal to $2$.
 \item \emph{Widening with threshold (or ``up to'')}: In some cases, instead of
@@
 value $X_p \in D$, $D$ being in our case the domain of convex polyhedra over
-$\Q^n$. Since the exact operation $\tau$ may not be expressed in the abstract
+$\Q^n$. Since the exact operation $\tau$ may not be expressible in the abstract
 domain, we abstract it into $\tau^\#$ such that $\forall X \in D, \tau(X)
@@
 After a finite number of steps, the computation has reached the fixpoint and the
-value $X_p$ at the end give the different linear relations between the
+value $X_p$ at the end gives various linear relations between the
 different numerical variables of the program at the control point $p$.
@@
 disconnects every cycles. For instance, $P_w$ can be chosen as the set of the
-heads of loops.
+loop headers.
 
@@
 end of the analysis is correct. Yet, the precision of the result and the time
-before convergence could be very different. 
+before convergence can be very different. 
 There have been some work to find efficient iteration strategies
@@
 	to the transition $e$, then the focus path goes through this transition.
-	\item in the other case, the focus path doesn't go through this transition.
+	\item in the other case, the focus path does not go through this transition.
 	\end{itemize}
@@
 x = 102$. We can update $X_7 = {(102,-1)}$. Path focusing technique gives us the
-precise result of this program. After that, the SMT-solver can't find any new
+precise result of this program. After that, the SMT-solver cannot find any new
 path, so the analysis terminates.
@@
 The new algorithm we propose is described in Algorithm \ref{gulwani2}.
-A difference is that here, we don't know the number of paths. Then, the mapping
+A difference is that here, we do not know the number of paths. Then, the mapping
 function $\sigma$ has to be defined on $\{1,..,m\} \times \N$. We could also
@@
 program. Indeed, there is no feasible paths going through $p_6$, but
-\emph{Lookahead Widening} don't know it because of the convex hull at point
+\emph{Lookahead Widening} do not know it because of the convex hull at point
 $p_5$.
@@
 variables are in SSA form, and removes most of the \emph{Load/Store} operations
-into memory. It is useful since our analyser doesn't have a memory model, and
+into memory. It is useful since our analyser does not have a memory model, and
 then loses lots of precision in the case of \emph{Load/Store} instructions.
@@
 This is the case for the distinction between \emph{signed} and \emph{unsigned}
-integers, which is nonexistent in the representation, since the compiler doesn't
+integers, which is nonexistent in the representation, since the compiler does
+not
 need this information to generate code. Yet, for static analysis, 
@@
 Indeed, in LLVM, live variables computation is processed during the register
-allocation. Then, the data structure we work on doesn't have live variables
+allocation. Then, the data structure we work on does not have live variables
 informations yet. To solve this problem, we coded our own pass, that is aimed at
@@
 \item \emph{Load} instructions, that get a value $v$ in the memory and assign it
-to a variable $x$, doesn't provide properties about the value $v$. Then, our
+to a variable $x$, does not provide properties about the value $v$. Then, our
 analyser considers $x$ can take all values of its type.
@@
 possible paths, but a lot of unfeasible paths. The idea is to show that our
-technique doesn't blow up, even with a huge expanded multigraph. Results show
+technique does not blow up, even with a huge expanded multigraph. Results show
 that in such benchmarks, our technique finds a much better invariant and is even

### 2011-06-13 report/report.tex
report += suggestions de matthieu

@@
 	véritable ensemble étant impossible à calculer en général.
-	Ce report s'inscrit dans la lignée des travaux visant à améliorer la
+	Ce rapport s'inscrit dans la lignée des travaux visant à améliorer la
 	précision de l'analyse par interprétation abstraite, et propose une
@@
 well as the implementation of another technique proposed in \cite{GopanR06},
-such that we can compare precision and cost of these two methods.
+such that we can compare precision and cost of these two methods. This
+implementation has been tested on a wide range of programs, including small
+test-cases as well as real-life programs.
 \end{itemize}
@@
   \path [transition] 
-		(n0) edge              node {$x = 0$} (n1);
-  \path [transition] 
-        (n1) edge			   node [above] {$x \geq 100?$} (n3);
-  \path [transition] 
-        (n1) edge			   node [left] {$x < 100?$} (n2);
-  \path [transition] 
-        (n2) edge [out=0, in=0, distance=2cm] node [right] {$x = x+1$} (n1);
+		(n0) edge              node {$x \gets 0$} (n1);
+  \path [transition] 
+        (n1) edge			   node [above] {$x \geq 100$} (n3);
+  \path [transition] 
+        (n1) edge			   node [left] {$x < 100$} (n2);
+  \path [transition] 
+        (n2) edge [out=0, in=0, distance=2cm] node [right] {$x \gets x+1$} (n1);
 
@@
 \emph{Lookahead Widening}. Indeed, it is more precise in the case of programs
-containing control-flow merges inside of loops, thanks to the distinction
+containing control-flow merges inside  loops, thanks to the distinction
 between all the paths of the loop. In simpler programs, it performs as well as
@@
  -d.0 + 1 >= 0
-\end{tabular}& $\subset$ & 0.064 & \begin{tabular}{l}
+\end{tabular}& $\varsubsetneq$ & 0.064 & \begin{tabular}{l}
  -1996x.0 + 4991d.0 + 1998995 >= 0\\
@@
 i.0 >= 0
-\end{tabular}& $\subset$ & 0.052 & \begin{tabular}{l}
+\end{tabular}& $\varsubsetneq$ & 0.052 & \begin{tabular}{l}
 -i.0 + 100 >= 0\\
@@
 x.0 >= 0
-\end{tabular}& $\subset$ & 0.032 & \begin{tabular}{l}
+\end{tabular}& $\varsubsetneq$ & 0.032 & \begin{tabular}{l}
 y.0 >= 0
@@
  y.0 + x.0 + 1 >= 0
-\end{tabular}& $\subset$ & 0.208 & \begin{tabular}{l}
+\end{tabular}& $\varsubsetneq$ & 0.208 & \begin{tabular}{l}
 -x.0 + 1 >= 0 \\
@@
 b.0 + c.0 + 1 >= 0\\
-\end{tabular}& $\subset$ & 0.916s & \begin{tabular}{l}
+\end{tabular}& $\varsubsetneq$ & 0.916s & \begin{tabular}{l}
 -b.0 + 1 >= 0\\
@@
 b.0 + c.0 + 1 >= 0\\
-\end{tabular}& $\subset$ & 36.30s & \begin{tabular}{l}
+\end{tabular}& $\varsubsetneq$ & 36.30s & \begin{tabular}{l}
 -b.0 + 1 >= 0 \\
@@
 thanks to the widening/narrowing phases (or acceleration). Yet, in the case of
-paths that do not constitute a self loop in the multigraph, one perform a
+paths that do not constitute a self loop in the multigraph, one performs a
 widening operation, but no narrowing iteration. Apparently, lots of the cases
@@
 appendix \ref{lostprecision}.
-An good idea that could solve this problem is to try to combine
-the two methods.
+An idea that could solve this problem is to try to combine
+the two methods (see \ref{combining}).
 \end{itemize}
@@
 	\section{Combining Lookahead Widening and Path Focusing}
-
+	\label{combining}
 	Path Focusing technique works on the expanded multigraph.
@@
 
-	Our analyser is able to deal with small- and middle-sized programs, but
-	looses
+	Our analyser is able to deal with small- and middle-sized test-cases, as
+	well as real-life programs. It looses
 	precision in the case of programs with pointers or several functions.

### 2011-06-13 report/report.tex
bug fix in the report

@@
 \usetikzlibrary{snakes,arrows,shapes,backgrounds,shadows,automata,patterns}
-\usepgflibrary{snakes}
+%\usepgflibrary{snakes}
 
@@
 \subsubsection{Sources of imprecision}
- \label{multigraph}
+ \label{imprecision}
 Linear Relation Analysis is sound but incomplete: there is loss of precision
@@
 
-	As explained in \ref{multigraph}, separating the different
+	As explained in \ref{imprecision}, separating the different
 	possible paths between $p$ and $q$ reduce the number of points with a join

### 2011-06-13 report/report.tex
minor changes in the report
bug fix in the report
report (final version)

@@
-\documentclass[a4paper,english,titlepage,11pt]{report}
-
-\usepackage{verimagstudent}
+\documentclass[a4paper,english,titlepage,11pt,twoside]{report}
+
+\usepackage[pdftex]{graphicx}
 \usepackage{mathcomp,amssymb}
@@
 \usepackage{multirow}
+\usepackage{verimagstudent}
 \usepackage[toc,page]{appendix}
+
 \newtheorem{definition}{Definition}[section]
@@
 \abstract{
-	Program verification is aimed at statically discovering properties on programs,
+	Program verification aims at statically discovering properties on programs,
 	such as the values that can take the different variables during execution.
-	Abstract Interpretation is a technique that allows to compute an
-	approximation of the set of these values, since it is impossible to compute
-	the real set in general.
+	Abstract Interpretation is a technique that  computes an
+	over-approximation of the set of these values, since it is impossible to
+	compute the real set in general.
 	This report takes place in the many attempts to improve the precision of
@@
 \webaddress{Julien.Henry@imag.fr}
-\reporttype{M2R internship's report}
+\reporttype{M2R Internship Report}
 
@@
 program.
-Instead of convex polyhedra, one could choose intervals, octagons, etc.
+Instead of convex polyhedra, one could choose intervals, octagons, etc, that are
+subclasses of the class of convex polyhedra.
 The fixpoint equation is computed iteratively; that is, successive
@@
 The fixpoint computation algorithm consists in replacing iteratively each $X_p$ by its value
-on the right hand side, until the convergence.
+on the right hand side, until convergence.
 
@@
 apply it on a subset of $P$, called $P_w$, such that removing the nodes of $P_w$
-disconnects every cycles. For instance, $P_w$ can be chosen as the set of the
+disconnects every cycles. For instance, in structured programs, $P_w$ can be chosen as the set of the
 loop headers.
@@
 
-Acceleration is possible when the loop is simple enough, and comes to compute
+Acceleration is possible when the loop is simple enough, and comes to computing
 the transitive closure $\tau_e^+$ of $\tau_e$, the transition function
@@
 We apply the linear relation analysis technique over the program in figure
-\ref{runningexample}. The set of state where we apply widening is $\{p_1\}$: it
+\ref{runningexample}. The set of control state where we apply widening is $\{p_1\}$: it
 is sufficient, since removing $p_1$ cuts all cycles in the graph.
@@
 	our case, the second path of the loop is now possible, because $x$ can be
-	equal to $51$, so we add the path in
+	equal to $51$, so we add the path to
 	the graph and compute new iterations until convergence\dots. After
@@
  \section{Path Focusing}
+
+	In this section, we present a refinement of Abstract Interpretation, refered
+	as \emph{Path Focusing} technique \cite{MG11}.
 
@@
 	there is a transition in the multigraph from $p_1$ to $p_k$ with the
-	semantic $$\tau_{p_1 \rightarrow \dots \rightarrow p_k} = \tau_{p_{k-1}
+	semantics $$\tau_{p_1 \rightarrow \dots \rightarrow p_k} = \tau_{p_{k-1}
 	\rightarrow p_k} \circ
 	\dots \circ \tau_{p_1 \rightarrow p_2}$$ where $\tau_{p_i \rightarrow
-	p_{i+1}}$ is the transformation associated to the semantic of $p_i
+	p_{i+1}}$ is the transformation associated to the semantics of $p_i
 	\rightarrow p_{i+1}$.
@@
 in the control flow graph corresponds to a simple dashed transition in the multigraph,
-having the same semantic.}
+having the same semantics.}
 \label{multigraphtransformGopan}
@@
 	The first point of the Path Focusing technique is to compute the SMT formula
-	$\rho$ expressing the semantic of the program. For doing so, all the
+	$\rho$ expressing the semantics of the program. For doing so, all the
 	operations in the program have to be expressible within the theory we
@@
 in the $\rho$ formula.
-\item For each operation in the program, we encode its semantic, expressed in
+\item For each operation in the program, we encode its semantics, expressed in
 terms of the Boolean and numerical SSA variables. For instance, if the CFG
-contains the assignment $x.2 = x.1 + 1$, then we simply add $x.2 = x.1 + 1$ in
+contains the assignment $x.2 = x.1 + 1$, then we simply add $x.2 = x.1 + 1$ to
 the formula. This equality expression is noted $assign(x.2)$. 
@@
 	One can refine the \emph{Path-focusing} algorithm by treating specially the
-	self loops of the multigraph. Indeed, instead of simply apply the widening
+	self loops of the multigraph. Indeed, instead of simply applying the widening
 	operator,  some narrowing steps could be performed to regain precision
@@
 	to itself: instead of computing $\tau_e^\#(X_{p_i})$, one compute
-	$SelfLoop(\tau_e^\#,X_{p_i})$, defined in algorithm \ref{loopiteralgo}.
-
-	The $SelfLoop$ procedure actually computes a widening operation, and
+	\textit{SelfLoop}$(\tau_e^\#,X_{p_i})$, defined in algorithm \ref{loopiteralgo}.
+
+	The \textit{SelfLoop} procedure actually computes a widening operation, and
 	computes a narrowing step in order to recover precision. In some cases,
@@
 	\cite{GulwaniZ10} proposes a technique to compute transitive closure of a
-	loop, i.e compute the invariants for a loop having its semantic
+	loop, i.e compute the invariants for a loop having its semantics
 	expressed by a transition system. The definition of a transition system is
@@
 	disjunction of conjunctions), where each
-	disjunct is the semantic of a path from $p$ to itself. It is of the form
+	disjunct is the semantics of a path from $p$ to itself. It is of the form
 	$$\bigvee_{1 \leq i \leq n} \tau_{p,i}$$ where $n$ is the number of
-	paths, and $\tau_{p,i}$ is the semantic of the $i$-th path from $p$ to
+	paths, and $\tau_{p,i}$ is the semantics of the $i$-th path from $p$ to
 	itself.
@@
 easily compute this number of paths when computing the $\rho$ formula.
+
+Computing disjunctive invariants is a field we have to work on. The work
+described here is only preliminary, and shows it is possible to improve existing
+techniques with our ideas, so that we avoid explicit enumeration of
+exponential-sized sets of paths.
 
@@
 \begin{itemize}
-\item the \emph{PromoteMemoryToRegister} Pass, that transforms the internal
+\item the \emph{PromoteMemoryToRegister} Pass (-mem2reg), that transforms the internal
 representation by promoting the memory variables to Static Single Assignment
@@
 then loses lots of precision in the case of \emph{Load/Store} instructions.
-\item the \emph{LoopSimplify} and \emph{LoopInfo} Passes, that transform all loops into a canonical
+\item the \emph{LoopSimplify} (-loopsimplify) and \emph{LoopInfo} (-loopinfo) Passes, that transform all loops into a canonical
 format, and compute some information related to loops (for instance, the set of
 BasicBlocks that are parts of the loop, etc.).
-\item the \emph{LowerSwitch} Pass, that transform a switch instruction into its
+\item the \emph{LowerSwitch} Pass (-lowerswitch), that transform a switch instruction into its
 corresponding \emph{if-then-else} instructions. After running this pass, there
@@
 	of the polyhedron.
+
+	Morally, this means the $\Phi$-variables are on incoming transitions,
+	whereas the other instructions are on outgoing transitions.
 
@@
 
-	\subsection{Dimensions of the Abstract values}
+	\subsection{Dimensions of the abstract values}
 
@@
 two intervals: $]-\infty, N[\  \cup\  ]N,+\infty[$. Since we use convex polyhedra to
-represent the set of possible states of our program, we loose information when
+represent the set of possible states of our program, we lose information when
 the analysis comes across a \emph{diseq} operation.
@@
 multiplication inside the loop. The Apron library handles multiplication using
-a linearization technique. However, multiplications is not allowed in LIA
+a linearization technique. However, multiplication is not allowed in LIA
 (Linear Integer Arithmetic) or LRA (Linear Real Arithmetic). Then, Yices
@@
 	This is partly due to the widening operator: even if the analysis is more
-	subtle, the result at the end may be worse.
+	subtle, the result at the end may be worse. This is a well-known result in
+	abstract interpretation: when using widening operations, the final result is
+	not monotonic with the quality of the abstraction.
 
@@
 An idea that could solve this problem is to try to combine
-the two methods (see \ref{combining}).
+the two methods, and get the best of both worlds (see \ref{combining}).
 \end{itemize}
@@
 	techniques.
-	Abstract Interpretation is a common used technique to do static analysis of
+	Abstract Interpretation is a commonly used technique to do static analysis of
 	programs. It is aimed at computing iteratively an over-approximation of the set of
@@
 	Our analyser is able to deal with small- and middle-sized test-cases, as
-	well as real-life programs. It looses
+	well as real-life programs. It loses
 	precision in the case of programs with pointers or several functions.

### 2011-06-17 src/Compare.cc
soutenance.tex updated

@@ -41,7 +41,7 @@ int Compare::compareAbstract(Abstract * A, AbstractGopan * B) {
 	} else if (B->is_leq(A)) {
 		return -1;
 	} else {
-		return 0;
+		return -2;
 	}
 }
 
@@ -54,6 +54,7 @@ bool Compare::runOnModule(Module &M) {
 	int equal = 0;
 	int LW = 0;
 	int PF = 0;
+	int UN = 0;
 
 	Out->changeColor(raw_ostream::BLUE,true);
 	*Out << "\n\n\n"
@@ -76,6 +77,9 @@ bool Compare::runOnModule(Module &M) {
 					case 1:
 						PF++;
 						break;
+					case -2:
+						UN++;
+						break;
 					case -1:
 						*Out << "LW is better in " << *b << "\n";
 						n->Y->print();
@@ -93,6 +97,7 @@ bool Compare::runOnModule(Module &M) {
 	*Out << PF << " : Path focusing is better\n";
 	*Out << LW << " : Lookahead Widening is better\n";
 	*Out << equal << " : Same result\n";
+	*Out << UN << " : uncomparable\n";
 	return true;
 }
 

### 2011-08-03 src/Analyzer.cc
algo.tex added: algorithm combining path focusing + lookahead widening

@@ -53,7 +53,7 @@ Apron_Manager_Type getApronManager() {
 }
 
 
-void setApronManager(char * domain) {
+bool setApronManager(char * domain) {
 	std::string d;
 	d.assign(domain);
 	
@@ -73,7 +73,9 @@ void setApronManager(char * domain) {
 		ap_manager = PKGRID;
 	} else {
 		std::cout << "Wrong parameter defining the abstract domain\n";
+		return 1;
 	}
+	return 0;
 }
 
 std::set<llvm::Function*> ignoreFunction;
@@ -116,7 +118,9 @@ int main(int argc, char* argv[]) {
             break;
         case 'd':
             domain = optarg;
-			setApronManager(domain);
+			if (setApronManager(domain)) {
+				bad_use = true;
+			}
             break;
         case 'i':
             filename = optarg;

### 2011-08-30 papers/PLDI2012/implicitization.tex
PLDI 2012

\documentclass[preprint]{sigplanconf}

% The following \documentclass options may be useful:
%
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% authoryear    To obtain author/year citation style instead of numeric.

\usepackage{amsmath}
\usepackage{dmnatbib}
\usepackage{listings}
\lstset{language=C,basicstyle={\rm\fontfamily{ppl}\selectfont},mathescape=true,numbers=left}
\usepackage{hyperref}

\begin{document}

\conferenceinfo{PLDI '12}{11-16 June 2012, Beijing} 
\copyrightyear{2012} 
\copyrightdata{[to be supplied]}

\titlebanner{banner above paper title}        % These are ignored unless
\preprintfooter{short description of paper}   % 'preprint' option specified.

\title{Implicit yaddayadda}
\subtitle{Subtitle Text, if any}

\authorinfo{Anonymous}
           {anonymous}
           {anonymous.gal@gmail.com}

\maketitle

\begin{abstract}
This is the text of the abstract.
\end{abstract}

% \category{CR-number}{subcategory}{third-level}
\category{D.2.4}{Software/Program Verification}{Formal methods}
\category{F.3.1}{Specifying and Verifying and Reasoning about Programs}{Assertions, Invariants, Mechanical verification, Pre- and post-conditions}
\category{F.3.2}{Semantics of Programming Languages}{Program analysis}
\category{F.4.1}{Mathematical Logic}{Logic and constraint programming, Mechanical theorem proving}
\category{I.2.3}{Deduction and Theorem Proving}{Inference engines, Logic programming, Resolution}

\terms
theory, verification, algorithms

\keywords
SMT-solving, static analysis, abstract interpretation, disjunctions

\section{Introduction}
% TODO expliquer 'implicite'
Static analysis by abstract interpretation is a fully automatic program analysis method. When applied to imperative programs, it computes an inductive invariant mapping each program location (or a subset thereof) to a set of states represented symbolically~\cite{CousotCousot_JLC92}.
For instance, if we are only interested in scalar numerical program variables, such a set may be a convex polyhedron (the set of solutions of a system of linear inequalities)~\cite{CousotHalbwachs78,PPL,BagnaraHZ08SCP}.

In such an analysis, information may flow forward (one computes the polyhedron after a program statement as the image, or a super-set thereof, of the polyhedron before by the semantics of the statement) or backward;
forward program analysis computes super-sets of the states reachable from the initialization of the program, backward program analysis computes super-sets of the states co-reachable from some property of interest (for instance, the violation of an assertion).
In forward analysis, control-flow joins correspond to computation of convex hulls if using convex polyhedra (more generally, they correspond to least upper bounds in a lattice); in backward analysis, it is control-flow splits.

It is a known limitation of program analysis by abstract interpretation that this convex hull, or more generally, least upper bound operation, may introduce states that cannot occur in the real program: for instance, the convex hull of the intervals $[-2,-1]$ and $[1,2]$ is $[-2,2]$, strictly larger than the union of the two.
Such introduction may prevent proving desired program properties, for instance $\neq 0$. The alternative is to keep the union symbolic (e.g. compute using $[-2,-1] \cup [1,2]$) and thus compute in the \emph{disjunctive completion}
of the lattice, but the number of terms in the union may grow exponentially with the number of successive tests in the program to analyze, not to mention difficulties for designing suitable widening operators for enforcing the convergence of fixpoint iterations~\cite{PPL,BagnaraHZ08SCP,DBLP:journals/sttt/BagnaraHZ07}.
The exponential growth of the number of terms in the union may be controlled by heuristics that judiciously apply least upper bound operations, as in the \emph{trace partitioning domain} \cite{Rival_Mauborgne_TOPLAS07} implemented in the Astr\'ee analyzer~\cite{ASTREE_TASE07,ASTREE_ESOP05,ASTREE_PLDI03}. % TODO ne garder que les papiers les plus significatifs par rapport au sujet évoqué

Assuming we are  interested in a loop-free program fragment, the above approach of keeping symbolic unions gives the same results as performing the analysis separately over every path in the fragment.
A recent method for finding inductive loop invariants \cite{DBLP:conf/pldi/GulwaniZ10} is based on this idea: each path inside the loop body is considered separately.
Two recent proposals use SMT-solving \cite{Kroening_Strichman_08} as a decision procedure for the satisfiability of first-order arithmetic formulas in order to enumerate only paths that are needed for the progress of the analysis \cite{Gawlitza_Monniaux_ESOP11,Monniaux_Gonnord_SAS11}.

An additional claim of the methods that distinguish paths inside the loop body \cite{DBLP:conf/pldi/GulwaniZ10,Monniaux_Gonnord_SAS11} is that they tend to generate better invariants that methods that do not, by behaving better with respect to the \emph{widening operators} \cite{CousotCousot_JLC92}
used for enforcing convergence when searching for loop invariants by Kleene iterations. A related technique, \emph{guided static analysis} \cite{DBLP:conf/sas/GopanR07}, computes successive loop invariants for increasing subsets of the transitions taken into account, until all transitions are considered; again, the claim is that this approach avoids some gross over-approximation introduced by widenings.

All these methods improve the precision of the analysis by keeping the same abstract domain (say, convex polyhedra) but changing the operations applied and their ordering. An alternative is to change the abstract domain, for instance by moving from intervals to octagons or convex polyhedra \cite{DBLP:journals/lisp/Mine06,Mine_PhD}, or the widening operator: for instance one may replace the classical widening on convex polyhedra \cite{CousotHalbwachs78,Halbwachs_PhD} by an improved version \cite{BagnaraHRZ05SCP}, or use widening ``up to'', that is, use linear inequalities syntactically present in the program source code as possible widening steps~\cite{Polka:FMSD:97}.

There are many possible combinations of the above techniques, and it is not evident which ones perform more or less precisely or more or less efficiency on real-life examples. One needs to experiment. Unfortunately, the published literature on the subject lacks experimental comparative assessments. One purpose of this article is therefore to propose such experimental results.
\medskip

This article makes the following contributions:
\begin{itemize}
\item We improve the technique for obtaining disjunctive invariants from \citet{DBLP:conf/pldi/GulwaniZ10} by replacing the explicit exhaustive enumeration of paths by a sequence of SMT queries.
\item We recast the guided static analysis technique from \citet{DBLP:conf/sas/GopanR07} on the expanded multigraph from \citet{Monniaux_Gonnord_SAS11}, considering entire paths instead of individual transitions, using SMT queries and binary decision diagrams.
\item We implemented these techniques, in addition to ``classical'' iterations and the original guided static analysis, inside a prototype static analyzer.
This tool uses the LLVM bitcode format \cite{Lattner:2004:LCF:977395.977673,LLVM_langref} as input, which can be produced by compilation from C, C++ and Fortran, enabling it to be run on many real-life programs.
With respect to abstract domains on numerical variables, it uses the APRON library \cite{DBLP:conf/cav/JeannetM09}, which supports a variety of abstract domains from which we can choose with minimal changes to our analyzer.
\item We conducted extensive experiments with this tool, on real-life programs, thus obtaining experimental comparisons between the various combinations of techniques.
\end{itemize}

\section{A simple, motivating example}
Consider the following program, adapted from \cite{Monniaux_Gonnord_SAS11}:
\lstinputlisting{rlim_int.c}
This program implements a construct commonly found in control programs (in e.g. automotive or avionics): a rate limiter. For the sake of simplicity, we chose it to be fed a nondeterministic input clamped between $[-100000,100000]$, but in a real system it would be integrated in a reactive control loop and connected to other signals.

The expected inductive invariant is $\verb|x_old| \in [-100000,100000]$, but classical abstract interpretation using intervals (or octagons or polyhedra) finds $\verb|x_old| \in (-\infty,+\infty)$. % TODO chercher ref Astree

Let us briefly why. Widening iterations converge to $\lstinline|x_old| \in (-\infty,+\infty)$; let us now see why decreasing iterations fail to recover the desired invariant. % TODO expliquer widening suivi de narrowing
The \lstinline|x >= x_old+10| test at line~7, if taken, yields $\lstinline|x_old| \in (-\infty,99990)$; followed by \lstinline|x = x_old+10|, we obtain $\lstinline|x| \in (-\infty,100000]$, and the same after union with the no-operation ``else'' branch. Line 8 yields $\lstinline|x| \in (-\infty,+\infty)$.

We could use ``guided widening'' or ``widening with thresholds'', propagating the ``magic values'' $\pm 100000$ associated to \lstinline|x| into~\lstinline|x_old|, but these syntactic approaches cannot directly cope with programs for which $\lstinline|x|  \in [-100000,+100000]$ is itself obtained by analysis.
The guided static analysis of \citet{DBLP:conf/sas/GopanR07} performs slightly better, and obtains $\verb|x_old| \in (-\infty,100000]$.

In contrast, let us distinguish all four possible execution paths through the tests at lines 7 and~8. The path through both ``else'' branches is infeasible; the program is thus equivalent to:
\lstinputlisting{rlim_int2.c}

Classical interval analysis on this program yields $\lstinline|x_old| \in [-100000,100000]$.

% Cas a probleme:
% Initialisation (x,y) à {(1,0), (0,1)}
% puis boucle: si y = 0 alors x++
% fait un polyèdre qui finit par perdre y <= 1 et x >= 0

%\appendix
%\section{Appendix Title}

%This is the text of the appendix, if you need one.

%\acks

% Acknowledgments, if needed.

% We recommend abbrvnat bibliography style.
\bibliographystyle{dmabbrvnat}
\bibliography{implicitization}
\end{document}

% Pour Emacs:
% Local Variables:
% reftex-cite-format: natbib
% End:

### 2011-08-30 papers/PLDI2012/rlim_int.c

@@ -1,0 +1,11 @@
+extern int input();
+void rate_limiter() {
+  int x_old = 0;
+  while (1) {
+    int x = input();
+    if (x < -100000 || x > 100000) x=0;
+    if (x >= x_old+10) x = x_old+10;
+    if (x <= x_old-10) x = x_old-10;
+    x_old = x;
+  }
+}

### 2011-08-30 papers/PLDI2012/rlim_int2.c

@@ -1,0 +1,11 @@
+extern int input();
+void rate_limiter() {
+  int x_old = 0;
+  while (1) {
+    int x = input();
+    if (x < -100000 || x > 100000) x=0;
+    if (x >= x_old+10) x = x_old+10;
+    else if (x <= x_old-10) x = x_old-10;
+    else x_old = x;
+  }
+}

### 2011-08-30 papers/PLDI2012/implicitization.tex
implicitization...

@@
 This article makes the following contributions:
-\begin{itemize}
-\item We improve the technique for obtaining disjunctive invariants from \citet{DBLP:conf/pldi/GulwaniZ10} by replacing the explicit exhaustive enumeration of paths by a sequence of SMT queries.
-\item We recast the guided static analysis technique from \citet{DBLP:conf/sas/GopanR07} on the expanded multigraph from \citet{Monniaux_Gonnord_SAS11}, considering entire paths instead of individual transitions, using SMT queries and binary decision diagrams.
+\begin{enumerate}
+\item We improve the technique for obtaining disjunctive invariants from \citet{DBLP:conf/pldi/GulwaniZ10} by replacing the explicit exhaustive enumeration of paths by a sequence of SMT queries.\label{contr:disjunctive}
+\item We recast the guided static analysis technique from \citet{DBLP:conf/sas/GopanR07} on the expanded multigraph from \citet{Monniaux_Gonnord_SAS11}, considering entire paths instead of individual transitions, using SMT queries and binary decision diagrams.\label{contr:guided_multigraph}
 \item We implemented these techniques, in addition to ``classical'' iterations and the original guided static analysis, inside a prototype static analyzer.
@@
 \item We conducted extensive experiments with this tool, on real-life programs, thus obtaining experimental comparisons between the various combinations of techniques.
-\end{itemize}
+\end{enumerate}
 
@@
 Classical interval analysis on this program yields $\lstinline|x_old| \in [-100000,100000]$.
+We have transformed the first program into the second, manually pruning out infeasible paths; yet in general the resulting program could be exponentially larger than the first, even though not all feasible paths are needed to compute the invariant.
+Following recent suggestions \cite{Gawlitza_Monniaux_ESOP11,Monniaux_Gonnord_SAS11}, we avoid this space explosion by keeping the second program implicit and working from the first program, yet simulating the analysis of the second program.
+Contribution \ref{contr:disjunctive} (\S\ref{sec:disjunctive}) replaces the exponential expansion of disjunctive invariant generation from \citet{DBLP:conf/pldi/GulwaniZ10} by an implicit representation; contribution \ref{contr:guided_multigraph} (\S\ref{sec:guided_multigraph}) recasts the ``guided analysis'' from \citet{DBLP:conf/sas/GopanR07} on an implicit representation of the paths in lieu of the individual transitions.
+
+\section{Guided widening}
+\label{sec:guided_multigraph}
+
+\section{Disjunctive invariants}
+\label{sec:disjunctive}
+
+\section{Experimental comparisons}
+
+\section{Conclusion}
 

### 2011-08-30 papers/PLDI2012/implicitization.tex
figure pompee du rapport de Julien

@@
 \usepackage{dmnatbib}
+
 \usepackage{listings}
 \lstset{language=C,basicstyle={\rm\fontfamily{ppl}\selectfont},mathescape=true,numbers=left}
+
+\usepackage{tikz}
+\usetikzlibrary{arrows,automata}
+
+\tikzstyle{state}=[circle,fill=black!25,minimum size=13pt,inner sep=0pt]
+\tikzstyle{rstate}=[rectangle,fill=black!25,minimum size=13pt,inner sep=0pt]
+\tikzstyle{transition}=[rectangle,semithick,draw=black!75,
+  			  minimum size=4mm]
+\tikzstyle{transition2}=[transition,rectangle,thick,dashed,
+  			  minimum size=4mm]
+\tikzstyle{PRstate}=[circle,double,draw,fill=blue!15,minimum size=13pt,inner sep=0pt]
+\tikzstyle{polyhedra}=[blue!25,opacity=0.5,pattern=north west lines,pattern
+color=blue]
+\tikzstyle{line}=[black,thick]
+
 \usepackage{hyperref}
@@
 
+\begin{figure}
+\label{fig:multigraph}
+\centering
+\begin{minipage}[c]{.19\textwidth}
+\begin{tikzpicture}[->,>=stealth',auto,node distance=1.1cm,
+                    semithick,font=\footnotesize]
+
+	\node[PRstate] (n0) {$p$};
+	\node[state] (n1) [below left of=n0] {};
+	\node[state] (n2) [below right of=n0] {};
+	\node[state] (n3) [below right of=n1] {};
+	\node[state] (n4) [below left of=n3] {};
+	\node[state] (n5) [below right of=n3] {};
+	\node[state] (n6) [below right of=n4] {};
+	\node[state] (n7) [below left of=n6] {};
+	\node[state] (n8) [below right of=n6] {};
+	\node[PRstate] (n9) [below right of=n7] {$q$};
+
+  \path [transition] 
+		(n0) edge              node {} (n1);
+  \path [transition] 
+		(n0) edge              node {} (n2);
+  \path [transition] 
+		(n1) edge              node {} (n3);
+  \path [transition] 
+		(n2) edge              node {} (n3);
+  \path [transition] 
+		(n3) edge              node {} (n4);
+  \path [transition] 
+		(n3) edge              node {} (n5);
+  \path [transition] 
+		(n4) edge              node {} (n6);
+  \path [transition] 
+		(n5) edge              node {} (n6);
+  \path [transition] 
+		(n6) edge              node {} (n7);
+  \path [transition] 
+		(n6) edge              node {} (n8);
+  \path [transition] 
+		(n7) edge              node {} (n9);
+  \path [transition] 
+		(n8) edge              node {} (n9);
+\end{tikzpicture}
+\end{minipage} 
+$\Longrightarrow$
+\begin{minipage}[c]{.19\textwidth}
+\begin{tikzpicture}[->,>=stealth',auto,node distance=1.1cm,
+                    semithick,font=\footnotesize]
+
+	\node[PRstate] (n0) {$p$};
+	\node (n1) [below left of=n0] {};
+	\node (n2) [below right of=n0] {};
+	\node (n3) [below right of=n1] {};
+	\node (n4) [below left of=n3] {};
+	\node (n5) [below right of=n3] {};
+	\node (n6) [below right of=n4] {};
+	\node (n7) [below left of=n6] {};
+	\node (n8) [below right of=n6] {};
+	\node[PRstate] (n9) [below right of=n7] {$q$};
+
+  \path [transition] 
+		(n0) edge  [out=0, in=0]            node {} (n9);
+  \path [transition] 
+		(n0) edge  [out=180, in=180]        node {} (n9);
+  \path [transition] 
+		(n0) edge  [out=205, in=155]            node {} (n9);
+  \path [transition] 
+		(n0) edge  [out=230, in=130]            node {} (n9);
+  \path [transition] 
+		(n0) edge  [out=255, in=105]            node {} (n9);
+  \path [transition] 
+		(n0) edge  [out=-25, in=25]            node {} (n9);
+  \path [transition] 
+		(n0) edge  [out=-50, in=50]            node {} (n9);
+  \path [transition] 
+		(n0) edge  [out=-75, in=75]            node {} (n9);
+\end{tikzpicture}
+\end{minipage}
+\caption{Expansion of the transition graph into a multigraph.}
+\end{figure}
+
 Classical interval analysis on this program yields $\lstinline|x_old| \in [-100000,100000]$.
-We have transformed the first program into the second, manually pruning out infeasible paths; yet in general the resulting program could be exponentially larger than the first, even though not all feasible paths are needed to compute the invariant.
-Following recent suggestions \cite{Gawlitza_Monniaux_ESOP11,Monniaux_Gonnord_SAS11}, we avoid this space explosion by keeping the second program implicit and working from the first program, yet simulating the analysis of the second program.
+We have transformed the first program into the second, manually pruning out infeasible paths; yet in general the resulting program could be exponentially larger than the first (as in Fig.~\ref{fig:multigraph}), even though not all feasible paths are needed to compute the invariant.
+Following recent suggestions \cite{Gawlitza_Monniaux_ESOP11,Monniaux_Gonnord_SAS11}, we avoid this space explosion by keeping the second program implicit and working from the first program, yet simulating the analysis of the second program, that is, on a transition multigraph (Fig.~\ref{fig:multigraph}).
 Contribution \ref{contr:disjunctive} (\S\ref{sec:disjunctive}) replaces the exponential expansion of disjunctive invariant generation from \citet{DBLP:conf/pldi/GulwaniZ10} by an implicit representation; contribution \ref{contr:guided_multigraph} (\S\ref{sec:guided_multigraph}) recasts the ``guided analysis'' from \citet{DBLP:conf/sas/GopanR07} on an implicit representation of the paths in lieu of the individual transitions.

### 2011-08-30 papers/PLDI2012/implicitization.tex
refs

@@
 \begin{enumerate}
-\item We improve the technique for obtaining disjunctive invariants from \citet{DBLP:conf/pldi/GulwaniZ10} by replacing the explicit exhaustive enumeration of paths by a sequence of SMT queries.\label{contr:disjunctive}
-\item We recast the guided static analysis technique from \citet{DBLP:conf/sas/GopanR07} on the expanded multigraph from \citet{Monniaux_Gonnord_SAS11}, considering entire paths instead of individual transitions, using SMT queries and binary decision diagrams.\label{contr:guided_multigraph}
+\item We recast the guided static analysis technique from \citet{DBLP:conf/sas/GopanR07} on the expanded multigraph from \citet{Monniaux_Gonnord_SAS11}, considering entire paths instead of individual transitions, using SMT queries and binary decision diagrams.\label{contr:guided_multigraph} (See \S\ref{contr:guided_multigraph})
+\item We improve the technique for obtaining disjunctive invariants from \citet{DBLP:conf/pldi/GulwaniZ10} by replacing the explicit exhaustive enumeration of paths by a sequence of SMT queries.\label{contr:disjunctive} (See \S\ref{contr:disjunctive})
 \item We implemented these techniques, in addition to ``classical'' iterations and the original guided static analysis, inside a prototype static analyzer.
@@
 With respect to abstract domains on numerical variables, it uses the APRON library \cite{DBLP:conf/cav/JeannetM09}, which supports a variety of abstract domains from which we can choose with minimal changes to our analyzer.
-\item We conducted extensive experiments with this tool, on real-life programs, thus obtaining experimental comparisons between the various combinations of techniques.
+\item We conducted extensive experiments with this tool, on real-life programs, thus obtaining experimental comparisons between the various combinations of techniques. (\S\ref{sec:experiments})
 \end{enumerate}
@@
 \section{Experimental comparisons}
+\label{sec:experiments}
 

### 2011-08-31 papers/PLDI2012/implicitization.tex
pouf pouf

@@
 \usepackage{listings}
-\lstset{language=C,basicstyle={\rm\fontfamily{ppl}\selectfont},mathescape=true,numbers=left}
+\lstset{language=C,mathescape=true,numbers=left} % basicstyle={\rm\fontfamily{ppl}\selectfont} trop reconnaissable
 
@@
 Following recent suggestions \cite{Gawlitza_Monniaux_ESOP11,Monniaux_Gonnord_SAS11}, we avoid this space explosion by keeping the second program implicit and working from the first program, yet simulating the analysis of the second program, that is, on a transition multigraph (Fig.~\ref{fig:multigraph}).
-Contribution \ref{contr:disjunctive} (\S\ref{sec:disjunctive}) replaces the exponential expansion of disjunctive invariant generation from \citet{DBLP:conf/pldi/GulwaniZ10} by an implicit representation; contribution \ref{contr:guided_multigraph} (\S\ref{sec:guided_multigraph}) recasts the ``guided analysis'' from \citet{DBLP:conf/sas/GopanR07} on an implicit representation of the paths in lieu of the individual transitions.
+Contribution \ref{contr:disjunctive} (\S\ref{sec:disjunctive}) replaces the exponential expansion of disjunctive invariant generation from \citet{DBLP:conf/pldi/GulwaniZ10} by an implicit representation;
+contribution \ref{contr:guided_multigraph} (\S\ref{sec:guided_multigraph}) recasts the ``guided analysis'' from \citet{DBLP:conf/sas/GopanR07} on an implicit representation of the paths in lieu of the individual transitions.
 
-\section{Guided widening}
+\section{Guided static analysis}
 \label{sec:guided_multigraph}
+\emph{Guided static analysis} was proposed by \citet{DBLP:conf/sas/GopanR07} as an improved over classical upward Kleene iterations with widening.
+Consider the following program:
+\lstinputlisting{gopan_reps.c}
+
+Classical iterations on the domain of convex polyhedra \cite{Halbwachs_PhD,CousotHalbwachs78} or octagons \cite{Mine_PhD,DBLP:journals/lisp/Mine06} start with $x = 0 \land x = 0$, then continue with $x = y \land 0 \leq x \leq 1$.
+The widening operator extrapolates from these two iterations and yields $x = y \land x \geq 0$.
+From there, the ``else'' branch at line 5 may be taken; with further widening, $0 \leq y \leq x$ is obtained as a loop invariant, and thus the postcondition computed at line 9 is $x \geq 0 \land y = 0$.
+Yet the strongest invariant is $(0 \leq x \leq 51 \land y = x) \lor (51 \leq x \leq 102 \land x+y=102)$, and its convex hull, a convex polyhedron, is
+\begin{equation}
+y \leq x \land y \leq 102-x \land y \geq 0.\label{eqn:triangle}
+\end{equation} % TODO fig
+Intuitively, this disappointing result is obtained because widening extrapolates from the first iterations of the loop, but the loop has two different phases ($x \leq 50$ and $x > 50$) with different behaviors, thus the extrapolation from the first phase is not valid for the second.
+
+\citeauthor{DBLP:conf/sas/GopanR07}' idea is to analyze the first phase of the loop with a widening and narrowing sequence, and thus obtain $0 \leq x \leq 50 \land y = x$, and then analyze the second phase, finally obtaining invariant~\ref{eqn:triangle}; each phase is identified by the tests taken or not taken.
+
+The analysis starts by identifying the test taken and not taken during the first iteration of the loop, starting in the loop initialization.
+The branches not taken are pruned from the loop body, yielding:
+\begin{lstlisting}[numbers=none]
+  while(1) {
+    if(x <= 50) y++;
+    else break; /* not taken in phase 1 */
+    if(y < 0) break;
+    x++;
+  }
+\end{lstlisting}
+
+Analyzing this loop using widening and narrowing on convex polyhedra or octagons yields the loop invariant $0 \leq x \leq 51 \land y = x$.
+Now, the transition at line~5 becomes feasible; and we analyze the full loop, starting iterations from $0 \leq x \leq 51 \land y = x$, and obtain invariant~\ref{eqn:triangle}.
+
+More generally, this analysis method considers an ascending sequence of subsets of the transitions in the loop body (left side of Fig.~\ref{fig:multigraph});
+for each subset, an inductive invariant is computed for the program restricted to it.
+The starting subset are the transitions reachable in one step from the loop initialization.
+If for a given subset $S$ in the sequence, no transitions outside of $S$ are reachable from the inductive invariant attached to $S$, then iterations stop;
+otherwise, add these transitions to $S$ and iterate more.
+Termination ensues from the finiteness of the control-flow graph.
 

### 2011-09-09 papers/PLDI2012/implicitization.tex
typo
suite du blabla introductif

@@
 
-\usepackage{amsmath}
+\usepackage{amsmath,amsfonts}
 \usepackage{dmnatbib}
+\usepackage[mathscr]{euscript}
 
@@
 \usepackage{hyperref}
+
+\newcommand{\abstr}[1]{#1^\sharp}
+\newcommand{\parts}[1]{\mathscr{P}(#1)}
+\newcommand{\ZZ}{\mathbb{Z}}
+\newcommand{\QQ}{\mathbb{Q}}
 
@@
 
-\section{A simple, motivating example}
+\section{Bases}
+\subsection{Static analysis by abstract interpretation}
+Let $X$ be the set of possible states of the program variables; for instance, if the program has 3 unbounded integer variables, then $X = \ZZ^3$.
+The set $\parts{X}$ of subsets of $X$, partially ordered by inclusion, is often referred to as the \emph{concrete domain}.
+An \emph{abstract domain} is a set $\abstr{X}$ equipped with a partial order $\sqsubseteq$; for instance, it can be the domain of convex polyhedra in $\QQ^3$ ordered by geometric inclusion.
+In this article, all abstract domains are supposed to contain machine-representable objects, and all $\sqsubseteq$ order relations are supposed to be decidable.
+The concrete and abstract domains are connected by a monotone \emph{concretization} function $\gamma: (\abstr{X},\sqsubseteq) \rightarrow (\parts{X},\subseteq)$: an element $\abstr{x} \in \abstr{X}$ represents a set $\gamma(\abstr{x})$.%
+%
+\footnote{%
+Some presentations of abstract interpretation identify an element $\abstr{x}$ with the set of states $\gamma(\abstr{x})$ that it represents.
+This leads to simpler notations, but may also confuse if, as in some of our constructions, there exist several $\abstr{x}$ with identical $\gamma(\abstr{x})$, but distinct algorithmic behaviors.}
+
+We also assume a join operator $\sqcap: \abstr{X} \times \abstr{X} \rightarrow \abstr{X}$, with infix notation; in practice, it is generally a least upper bound operation, but we only need it to satisfy $\gamma(\abstr{x}) \cup \gamma(\abstr{y}) \subseteq \gamma(\abstr{x} \sqcap \abstr{y})$ for all $\abstr{x},\abstr{y}$.
+
+\subsection{SMT-solving}
+TODO
+
+\subsection{A simple, motivating example}
 Consider the following program, adapted from \cite{Monniaux_Gonnord_SAS11}:
@@
 automotive or avionics): a rate limiter.
-For the sake of simplicity, we chose it to be fed a nondeterministic input clamped between $[-100000,100000]$, but in a real system it would be integrated in a reactive control loop and connected to other signals.
+For the sake of simplicity, we chose it to be fed a nondeterministic input clamped between $[-100000,100000]$, but in a real system it would be integrated in a reactive control loop and its input connected to a complex system with unknown output range.
 
@@
 % TODO chercher ref Astree
+Let us briefly see why.
 
-Let us briefly why.
-Widening iterations converge to $\lstinline|x_old| \in (-\infty,+\infty)$; let us now see why decreasing iterations fail to recover the desired invariant.
-% TODO expliquer widening suivi de narrowing
+Widening iterations converge to $\lstinline|x_old| \in (-\infty,+\infty)$; let us now see why decreasing iterations fail to recover the desired invariant. % TODO expliquer widening suivi de narrowing
 The \lstinline|x >= x_old+10| test at line~7, if taken, yields $\lstinline|x_old| \in (-\infty,99990)$; followed by \lstinline|x = x_old+10|, we obtain $\lstinline|x| \in (-\infty,100000]$, and the same after union with the no-operation ``else'' branch.
@@
 We have transformed the first program into the second, manually pruning out infeasible paths; yet in general the resulting program could be exponentially larger than the first (as in Fig.~\ref{fig:multigraph}), even though not all feasible paths are needed to compute the invariant.
-Following recent suggestions \cite{Gawlitza_Monniaux_ESOP11,Monniaux_Gonnord_SAS11}, we avoid this space explosion by keeping the second program implicit and working from the first program, yet simulating the analysis of the second program, that is, on a transition multigraph (Fig.~\ref{fig:multigraph}).
+Following recent suggestions \cite{Gawlitza_Monniaux_ESOP11,Monniaux_Gonnord_SAS11}, we avoid this space explosion by keeping the second program implicit while simulating its analysis.
+This means we work on an implicitly represented transition multigraph (Fig.~\ref{fig:multigraph}); its compact representation is given by the transition graph of the first program.
+
 Contribution \ref{contr:disjunctive} (\S\ref{sec:disjunctive}) replaces the exponential expansion of disjunctive invariant generation from \citet{DBLP:conf/pldi/GulwaniZ10} by an implicit representation;
@@
 % We recommend abbrvnat bibliography style.
+
 \bibliographystyle{dmabbrvnat}

### 2011-09-20 papers/PLDI2012/implicitization.tex
Explication de la technique combinée en cours...

@@
 \usepackage[mathscr]{euscript}
+\usepackage{algorithmic,algorithm}
 
@@
 
-\section{Guided static analysis}
+\subsection{Guided static analysis}
 \label{sec:guided_multigraph}
@@
 
+\section{Guided analysis over the paths}
+\label{sec:combined}
+
+Guided static analysis \cite{DBLP:conf/sas/GopanR07} applies to the transition
+graph of the program. One could instead apply the same technique to the implicit
+multigraph from \cite{Monniaux_Gonnord_SAS11}, thus avoiding control flow merges with
+unfeasible paths.
+
+Let $P$ be the set of control points in the transition graph, 
+$P_W \subseteq P$ the set of widening points
+such that removing the points in $P_W$ gives an acyclic graph.
+One can choose a set $P_R$ such that $P_W \subseteq P_R \subseteq P$.
+The combination of these two techniques considers an ascending sequence of
+subsets of the paths between two points in $P_R$.
+
+The set of paths is kept implicit by an SMT formula $\rho$ expressing
+the semantics of the program, assuming that the transition semantics can be
+expressed within a decidable theory.
+
+This formula contains Boolean \emph{reachability predicates} $b_i$ for each
+control points $p_i \notin P_R$, $b_i^s$ and $b_i^d$ for each $p_i \in P_R$, so
+that a path 
+$p_{i_1} \rightarrow p_{i_2} \rightarrow \dots \rightarrow p_{i_n}$ 
+between two points $p_{i_1}, p_{i_n} \in P_R$ 
+can easily be expressed as the
+cunjunction $b_{i_1}^s \wedge \bigwedge_{2 \leq k < n} b_{i_k} \wedge b_{i_n}^d$.
+
+The current working subset of paths is stored using a compact representation,
+such as Binary Decision Diagrams.
+
+\subsection{Ascending iterations}
+\label{subsec:ascending}
+
+The algorithm for computing an inductive invariant over a subgraph is the Path
+Focused algorithm from \citet{Monniaux_Gonnord_SAS11}.
+
+In order to find which path to focus on, we construct an SMT formula $f(p_i)$, whose
+model when satisfiable is a path that starts in $p_i$, goes to a successor $p_j
+\in P_R$ of $p_i$, such that the image of $X_{p_i}^s$ by the path transformation
+is not included in the actual $X_{p_j}^s$.
+Intuitively, such a path makes the abstract value $X_{p_j}^s$ grow, and thus is
+an interesting path to focus on. When the formula becomes unsatisfiable, it
+means the analysis of $p_i$ is finished.
+
+If we note $Succ(i)$ the set of indices $j$ such that $p_j \in P_R$ is a
+successor of $p_i$ in the expanded multigraph, and $X_i^s$ the abstract value
+associated to $p_i$ :
+$$f(p_i) = \rho \wedge b_i^s \wedge 
+\displaystyle\bigwedge_{j \in P_R \atop j \neq i} \neg
+b_j^s \wedge x \in X_i^s \wedge \displaystyle\bigvee_{j \in Succ(i)} (b_j^d \wedge
+x \notin X_j^s)$$
+
+Since we do not work on the entire transition graph but on a subset of it, we
+have to cunjunct the formula $f(p_i)$ with the actual set of working paths,
+noted $P$, expressed as a boolean formula, where the boolean variables are the
+\emph{reachability predicates} of the control points. We can easily construct
+this formula from the binary decision diagram using dynamic programming, and
+avoiding an exponentially sized formula. In other words, we force the SMT solver
+to give us a path which is in $P$.
+
+\subsection{Adding new paths}
+\label{subsec:addingpaths}
+
+Our technique computes the fixpoint iterations on a ascending sequence of
+subgraphs, until the complete graph is reached.
+
+When the analysis of a subgraph is finished, meaning that the abstract values
+for each control points has converged to an inductive invariant for this subgraph,
+the next subgraph to work on has to be computed.
+
+This new subgraph is the union of the paths of the previous one with a set
+$P'$ of new paths that become feasible regarding the current abstract values.
+$P'$ is progressively constructed during the analysis of the subgraph.
+
+We distinguish two phases in the analysis of a point $p_i \in P_R$:
+\begin{enumerate}
+\item We search for paths starting at point $p_i$, such that these paths are in
+$P$, i.e are included in the working subgraph. Each time we find a path, we
+update the abstract value of the destination point of the path. This is the
+phase explained in \ref{subsec:ascending}.
+\item  When the precedent phase terminates, there is no more paths in
+$P$, starting at $p_i$, 
+that make the abstract values of the successors grow (otherwise, the SMT
+solver would not have answered ``\emph{unsat}''). However, there may
+exist such paths in the entire multigraph, that are not in $P$. The second phase
+computes these paths and adds them to $P'$.
+\end{enumerate}
+
+It is not necessary to add into $P'$ all the new feasible paths. We just have
+to add a new path into $P'$ until no more path can make the invariant grow,
+as proposed by Algorithm \ref{algo:computepaths}. For that, we use another
+abstract value associated to a control point $p_j$, noted $X_j^d$, which is
+different from $X_j^s$.
+
+\begin{algorithm}
+	\caption{ComputeNewPaths}
+	\label{algo:computepaths}
+	\begin{algorithmic}[1] 
+	\input{algo/compute-paths}
+	\end{algorithmic}
+\end{algorithm}
+
+\subsection{Algorithm}
+
+The algorithm for Guided static analysis on the implicit multigraph is described
+in Algorithm \ref{algo:combined}.
+
+We maintain a set of active control points $A$, containing the points in $P_R$
+that still need to be processed. We apply the two previously described phases to
+the elements of $A$ until $A$ is empty. When $A$ is empty, it means the
+invariant computation of the subgraph is finished. As proposed by
+\citet{DBLP:conf/sas/GopanR07}, we can do some narrowing iterations, and we
+restart the analysis with a new subgraph if needed.
+\begin{algorithm}
+	\caption{Guided static analysis on implicit multigraph}
+	\label{algo:combined}
+	\begin{algorithmic}[1] 
+	\input{algo/algo-combined}
+	\end{algorithmic}
+\end{algorithm}
+
 \section{Disjunctive invariants}

### 2011-09-20 papers/PLDI2012/implicitization.tex
typo

@@
 
+We also assume that the program in expressed in SSA form, meaning that each
+variable is only assigned once in the transition graph. This is not a
+restriction, since there exist standard algorithm that transform a program into
+an SSA form.
+
 The set of paths is kept implicit by an SMT formula $\rho$ expressing
@@
 The current working subset of paths is stored using a compact representation,
-such as Binary Decision Diagrams.
+such as binary decision diagrams.
 
@@
 have to cunjunct the formula $f(p_i)$ with the actual set of working paths,
-noted $P$, expressed as a boolean formula, where the boolean variables are the
+noted $P$, expressed as a Boolean formula, where the Boolean variables are the
 \emph{reachability predicates} of the control points. We can easily construct
@@
 
-We distinguish two phases in the analysis of a point $p_i \in P_R$:
+We distinguish two phases when analysing a point $p_i \in P_R$:
 \begin{enumerate}

### 2011-09-20 papers/PLDI2012/implicitization.tex
algorithme

@@
 phase explained in \ref{subsec:ascending}.
-\item  When the precedent phase terminates, there is no more paths in
+\item  When the previous phase terminates, there is no more paths in
 $P$, starting at $p_i$, 
@@
 exist such paths in the entire multigraph, that are not in $P$. The second phase
-computes these paths and adds them to $P'$.
+computes these paths and adds them to $P'$. We also use SMT solving to discover
+these new paths, but we slightly change the SMT formula given to the SMT solver.
+We now simply check for the satisfiability of $f(p_i)$ instead of $f(p_i) \wedge
+P$. Since we already know that $f(p_i) \wedge P$ is \emph{unsatisfiable}, all
+the paths given by the SMT solver won't be in $P$.
 \end{enumerate}
@@
 
+Termination of this algorihm is guaranteed, because:
+\begin{itemize}
+\item 
+the subset of paths $P$ is
+strictly increasing at each loop iteration, and bounded by the finite set of
+paths in the entire graph. 
+\item the set $P'$ always verifies $P \wedge P' = false$ by construction, 
+which guarantees that $P$
+will eventually be empty after a finite number of loop iterations.
+\end{itemize}
 \section{Disjunctive invariants}

### 2011-09-21 papers/PLDI2012/implicitization.tex
parties 2 et 3

@@
 
-We also assume that the program in expressed in SSA form, meaning that each
-variable is only assigned once in the transition graph. This is not a
-restriction, since there exist standard algorithm that transform a program into
-an SSA form.
-
 The set of paths is kept implicit by an SMT formula $\rho$ expressing
 the semantics of the program, assuming that the transition semantics can be
-expressed within a decidable theory.
+expressed within a decidable theory. For an easy construction of $\rho$, 
+we also assume that the program in expressed in SSA form, meaning that each
+variable is only assigned once in the transition graph. This is not a
+restriction, since there exist standard algorithms that transform a program into
+an SSA format.
 
@@
 cunjunction $b_{i_1}^s \wedge \bigwedge_{2 \leq k < n} b_{i_k} \wedge b_{i_n}^d$.
+The Boolean $b_{i}^s$ is $true$ when the path starts at point $p_i$, whereas
+$b_i^d$ is $true$ when the path arrives at $p_i$.
 
@@
 model when satisfiable is a path that starts in $p_i$, goes to a successor $p_j
-\in P_R$ of $p_i$, such that the image of $X_{p_i}^s$ by the path transformation
-is not included in the actual $X_{p_j}^s$.
-Intuitively, such a path makes the abstract value $X_{p_j}^s$ grow, and thus is
+\in P_R$ of $p_i$, such that the image of $X_{i}^s$ by the path transformation
+is not included in the actual $X_{j}^s$.
+Intuitively, such a path makes the abstract value $X_{j}^s$ grow, and thus is
 an interesting path to focus on. When the formula becomes unsatisfiable, it
@@
 avoiding an exponentially sized formula. In other words, we force the SMT solver
-to give us a path which is in $P$.
+to give us a path included in $P$.
 
@@
 It is not necessary to add into $P'$ all the new feasible paths. We just have
-to add a new path into $P'$ until no more path can make the invariant grow,
+to add them one after another, until no more path can make the invariant grow,
 as proposed by Algorithm \ref{algo:computepaths}. For that, we use another
@@
 paths in the entire graph. 
-\item the set $P'$ always verifies $P \wedge P' = false$ by construction, 
+\item the set $P'$ always verifies $P \cap P' = \emptyset$ by construction, 
 which guarantees that $P$
@@
 \end{itemize}
+
 \section{Disjunctive invariants}
 \label{sec:disjunctive}
+
+\citet{DBLP:conf/pldi/GulwaniZ10} proposes a technique for computing disjunctive invariants, by
+distinguishing all the paths inside a loop. In
+this section, we propose to improve this technique by using SMT queries to find
+interesting paths, the objective being to avoid an explicit exhaustive
+enumeration of an exponential number of paths.
+
+This technique is aimed at computing for a loop header $p_i$ a disjunctive invariant
+$\bigvee_{1\leq j \leq m} X_{i,j}$. 
+To do so, one chooses an integer $\delta \in [1,m]$, and
+a mapping function $\sigma: [1,m] \times [1,n] \mapsto [1,m]$, assuming there
+are $n$ distinct paths in the loop.  For each
+abstract value of the disjunctive invariant, and for each path $\tau_{i,k}$ in the loop, the
+image of $X_{i,j}$ by the path is joined with
+$X_{i,\sigma(j,k)}$.
+Initially, the $\delta$-th abstract value is assigned to the initial states of
+$p_i$, and each other abstract value is assigned to bottom.
+
+Our method improves this technique in two ways :
+\begin{itemize}
+\item Instead of enumerating the whole set of paths, we keep them implicit and
+compute them only when needed.
+\item At each loop iteration of the algorithm explained in
+\cite{DBLP:conf/pldi/GulwaniZ10}, an image by a path and a union of abstract
+values are computed for each path and each disjunct. 
+Lots of these operations may be useless, in the sense that the resulting
+abstract value may not change.
+In our approach, we do this only for paths and disjuncts that makes the
+resulting abstract value grow.
+\end{itemize}
+
+Our improvement consists in a slight modification of the SMT formula we solve.
+We introduce in this formula Boolean variables $\{d_k, 1 \leq k \leq m\}$, so
+that we can easily find in the model which abstract value of the disjunction we
+have to choose. The resulting formula that is given to the SMT solver is defined
+by $g(p_i)$.
+
+\begin{eqnarray*}
+g(p_i) = & \rho \wedge b_i^s \wedge 
+\displaystyle\bigwedge_{j \in P_R \atop j \neq i} \neg b_j^s  \\
+ & \wedge 
+\displaystyle\bigvee_{1 \leq k \leq m} (d_k \wedge x \in X_{i,k} \wedge \bigwedge_{l \neq k}
+\neg d_l) \\
+ & \wedge
+\displaystyle\bigvee_{j \in Succ(i)} 
+(b_j^d \wedge \bigwedge_{1 \leq k \leq m} (x \notin X_{j,k}))
+\end{eqnarray*}
+
+Our algorithm for computing disjunctive invariants is described in Algorithm
+\ref{algo:disjunctive}
+
+\begin{algorithm}[!h]
+\caption{Disjunctive invariant computation with implicit paths}\label{gulwani2}
+\label{algo:disjunctive}
+\begin{algorithmic}[1] 
+\FORALL {$k \in \{1,..,m\} \setminus \{\delta\}$}
+	\STATE $X_{i,k} \gets \perp$
+\ENDFOR
+\STATE $X_{i,\delta} \gets I_{p_i}$
+\WHILE {true}
+		\STATE $res \gets SmtSolve\left[g(p_i)\right]$
+	\IF {$res = unsat$}
+		\STATE \textbf{break}
+	\ENDIF
+	\STATE Compute the path $\tau_{i,k}$ from $res$ 
+	\STATE Take $j \in \{ l | d_l = true\}$ 
+	\STATE $X_{i,\sigma(j,k)} \gets X_{i,\sigma(j,k)} \sqcup
+	\tau_{i,k}(X_{i,j})$
+\ENDWHILE
+\end{algorithmic}
+\end{algorithm}
 

### 2011-09-23 papers/PLDI2012/implicitization.tex
pldi

@@
 \subsection{Guided static analysis}
-\label{sec:guided_multigraph}
+\label{sec:guided}
 \emph{Guided static analysis} was proposed by \citet{DBLP:conf/sas/GopanR07} as an improved over classical upward Kleene iterations with widening.
@@
 
-\section{Guided analysis over the paths}
-\label{sec:combined}
+\subsection{Path-focused technique}
+\label{sec:path_focusing}
 
-Guided static analysis \cite{DBLP:conf/sas/GopanR07} applies to the transition
-graph of the program. One could instead apply the same technique to the implicit
-multigraph from \cite{Monniaux_Gonnord_SAS11}, thus avoiding control flow merges with
-unfeasible paths.
+\emph{Path-focused} technique from \citet{Monniaux_Gonnord_SAS11} proposes to
+distinguish the different paths in the program in order to avoid loss of
+precision due to merge operations. Since the number of paths may be exponential,
+the technique keeps them implicit and computes them when needed using
+SMT-solving. Then, the iterations are computed over a reduced multigraph instead
+of the classical transition graph.
 
@@
 One can choose a set $P_R$ such that $P_W \subseteq P_R \subseteq P$.
-The combination of these two techniques considers an ascending sequence of
-subsets of the paths between two points in $P_R$.
 
@@
 The Boolean $b_{i}^s$ is $true$ when the path starts at point $p_i$, whereas
-$b_i^d$ is $true$ when the path arrives at $p_i$.
+$b_i^d$ is $true$ when the path arrives at $p_i$. In other words, we split the
+points in $P_R$ into a \emph{source} point, with only outgoing transitions, and
+a \emph{destination} point, with only incoming transitions, so that the
+resulting graph is acyclic and there is no paths going through control
+points in $P_R$.
 
+In order to find focus paths, we solve an SMT formula which is satisfiable when
+there exist a path starting at a point $p_i \in P_R$ in a state included in the
+actual invariant candidate $X_i$, and arriving at a point $p_j \in P_R$ in a
+state outside of $X_j$. In this case, we construct this path using the model and
+update $X_j$. When $p_i = p_j$, meaning that the path is actually a self-loop,
+we can apply a widening/narrowing sequence, or even compute the transitive
+closure of the loop using abstract acceleration. % TODO citation laure gonnord
+
+\section{Guided analysis over the paths}
+\label{sec:guided_multigraph}
+
+Guided static analysis \cite{DBLP:conf/sas/GopanR07} applies to the transition
+graph of the program. One could instead apply the same technique to the implicit
+multigraph from \cite{Monniaux_Gonnord_SAS11}, thus avoiding control flow merges with
+unfeasible paths.
+In this section, we use the same notations as \ref{sec:path_focusing}.
+
+The combination of these two techniques considers an ascending sequence of
+subsets of the paths between two points in $P_R$.
 The current working subset of paths is stored using a compact representation,
@@
 
-The algorithm for computing an inductive invariant over a subgraph is the Path
-Focused algorithm from \citet{Monniaux_Gonnord_SAS11}.
+The algorithm for computing an inductive invariant over a subgraph is the Path-focused algorithm from \citet{Monniaux_Gonnord_SAS11}.
 
@@
 Since we do not work on the entire transition graph but on a subset of it, we
-have to cunjunct the formula $f(p_i)$ with the actual set of working paths,
+cunjunct the formula $f(p_i)$ with the actual set of working paths,
 noted $P$, expressed as a Boolean formula, where the Boolean variables are the
@@
 abstract value associated to a control point $p_j$, noted $X_j^d$, which is
-different from $X_j^s$.
+distinct from $X_j^s$, and initialized to $X_j^s$.
 
@@
 invariant computation of the subgraph is finished. As proposed by
-\citet{DBLP:conf/sas/GopanR07}, we can do some narrowing iterations, and we
-restart the analysis with a new subgraph if needed.
+\citet{DBLP:conf/sas/GopanR07}, we can do some narrowing iterations.
+These narrowing iterations allow to recover precision lost by
+widening, before taking into account new feasible paths.
+Thus, our technique combines both the advantages of \emph{Guided Static
+Analysis} and \emph{Path-focusing}.
+We then restart the analysis with a new subgraph if needed.
+
 \begin{algorithm}
@@
 
-\citet{DBLP:conf/pldi/GulwaniZ10} proposes a technique for computing disjunctive invariants, by
+\citet{DBLP:conf/pldi/GulwaniZ10} propose a technique for computing disjunctive invariants, by
 distinguishing all the paths inside a loop. In
@@
 
+$m$, $\delta$ and $\sigma$ can be defined heuristically.
+For instance, one could defined $\sigma$ so that $\sigma(j,k)$ only depends on the
+last transition of the path, or else construct it dynamically during the
+analysis.
+
 Our method improves this technique in two ways :
@@
 
-Our improvement consists in a slight modification of the SMT formula we solve.
+Our improvement consists in a slight modification of the SMT formula we solve in
+\ref{sec:guided_multigraph}.
 We introduce in this formula Boolean variables $\{d_k, 1 \leq k \leq m\}$, so

### 2011-09-23 papers/PLDI2012/implicitization.tex
exemple pour la technique combinée

@@
 \usepackage{amsmath,amsfonts}
+\usepackage{mathcomp,amssymb}
 \usepackage{dmnatbib}
@@
 \newcommand{\QQ}{\mathbb{Q}}
+\newcommand{\widening}{\mathop{\triangledown}}
 
@@
 \subsection{A simple, motivating example}
+\label{subsec:rate_lim}
 Consider the following program, adapted from \cite{Monniaux_Gonnord_SAS11}:
@@
 
+\subsection{Example}
+
+We revise the rate limiter described in \ref{subsec:rate_lim}. In this example,
+\emph{Path-focused} technique works well because all the paths starting at the
+loop header are actually self loops. In such a case, the technique performs a
+widening/narrowing sequence or accelerates the loop, thus leading to a precise
+invariant. However, in real cases, there also exist paths that are not
+self loops, and \emph{Path-focusing} only performs a widening. 
+This widening may induce unrecoverable loss of precision.
+
+Suppose the main loop of the rate limiter contains a nested loop:
+\lstinputlisting{rlim_int_loop.c}
+
+We choose $P_R$ as the set of loop headers of the function, plus the initial
+state. In this case, we have three elements in $P_R$.
+We also unroll the loop once, in order to distinguish the paths that don't go
+through the loop from the others:
+\begin{lstlisting}[numbers=none]
+if (wait) {
+   while (wait()) {}
+}
+\end{lstlisting}
+
+The main loop in the expanded multigraph has then 24 distinct paths, half of
+them being self loops, and the other half going to the header of the nested loop.
+
+Guided static analysis from \citet{DBLP:conf/sas/GopanR07} yields at line 5 
+$\lstinline|x_old| \in (-\infty,+\infty)$.
+Path-focused technique from \citet{Monniaux_Gonnord_SAS11} performs slightly
+better, and finds $\lstinline|x_old| \in (-\infty,10000)$.
+Our combined technique gives us the expected invariant 
+$\lstinline|x_old| \in (-10000,10000)$.
+
 \section{Disjunctive invariants}
@@
 \ref{sec:guided_multigraph}.
-We introduce in this formula Boolean variables $\{d_k, 1 \leq k \leq m\}$, so
+We introduce in this formula Boolean variables $\{d_j, 1 \leq j \leq m\}$, so
 that we can easily find in the model which abstract value of the disjunction we
@@
 by $g(p_i)$.
+When the formula is satisfiable, we know that the index $j$ of the starting
+disjunct that has to be chosen is the one for which the associate Boolean value
+$d_j$ is \emph{true} in the model. Then, we can easily compute the value of 
+$\sigma(j,k)$, thus know the index of the disjunct to join with.
 
@@
 Our algorithm for computing disjunctive invariants is described in Algorithm
-\ref{algo:disjunctive}
+\ref{algo:disjunctive}. In this algorithm, instead of simply joining the two
+abstract values, we use the widening operator to ensure the algorithm terminates
+when using abstract domains with infinite height.
 
@@
 	\STATE Take $j \in \{ l | d_l = true\}$ 
-	\STATE $X_{i,\sigma(j,k)} \gets X_{i,\sigma(j,k)} \sqcup
-	\tau_{i,k}(X_{i,j})$
+	\STATE $X_{i,\sigma(j,k)} \gets X_{i,\sigma(j,k)} \widening (X_{i,\sigma(j,k)} \sqcup \tau_{i,k}(X_{i,j}))$
 \ENDWHILE

### 2011-09-23 papers/PLDI2012/rlim_int_loop.c
rate limiter with nested loop

@@ -1,0 +1,13 @@
+extern int input();
+extern int wait();
+void rate_limiter() {
+  int x_old = 0;
+  while (1) {
+    int x = input();
+    if (x < -100000 || x > 100000) x=0;
+    if (x >= x_old+10) x = x_old+10;
+    if (x <= x_old-10) x = x_old-10;
+    x_old = x;
+    while (wait()) {}
+  }
+}

### 2011-09-27 papers/PLDI2012/implicitization.tex
petites modifs
Make sure math formula isn't cut before a )

@@
 In this article, all abstract domains are supposed to contain machine-representable objects, and all $\sqsubseteq$ order relations are supposed to be decidable.
-The concrete and abstract domains are connected by a monotone \emph{concretization} function $\gamma: (\abstr{X},\sqsubseteq) \rightarrow (\parts{X},\subseteq)$: an element $\abstr{x} \in \abstr{X}$ represents a set $\gamma(\abstr{x})$.%
+The concrete and abstract domains are connected by a monotone \emph{concretization} function $\gamma: \left(\abstr{X},\sqsubseteq\right) \rightarrow (\parts{X},\subseteq)$: an element $\abstr{x} \in \abstr{X}$ represents a set $\gamma(\abstr{x})$.%
 %
@@
 invariant. However, in real cases, there also exist paths that are not
-self loops, and \emph{Path-focusing} only performs a widening. 
+self loops, where \emph{Path-focusing} only performs a widening. 
 This widening may induce unrecoverable loss of precision.

### 2011-09-27 papers/PLDI2012/implicitization.tex
typos

@@
 \label{sec:guided}
-\emph{Guided static analysis} was proposed by \citet{DBLP:conf/sas/GopanR07} as an improved over classical upward Kleene iterations with widening.
+\emph{Guided static analysis} was proposed by \citet{DBLP:conf/sas/GopanR07} as an improvement over classical upward Kleene iterations with widening.
 Consider the following program:
@@
 expressed within a decidable theory. For an easy construction of $\rho$, 
-we also assume that the program in expressed in SSA form, meaning that each
+we also assume that the program is expressed in SSA form, meaning that each
 variable is only assigned once in the transition graph. This is not a

### 2011-09-27 papers/PLDI2012/implicitization.tex
Full path in \input{}.
The Makefile doesn't find the dependancy otherwise
typos and rewording

@@
 Guided static analysis \cite{DBLP:conf/sas/GopanR07} applies to the transition
-graph of the program. One could instead apply the same technique to the implicit
+graph of the program. We now present a new technique applying this analysis on the implicit
 multigraph from \cite{Monniaux_Gonnord_SAS11}, thus avoiding control flow merges with
@@
 computes these paths and adds them to $P'$. We also use SMT solving to discover
-these new paths, but we slightly change the SMT formula given to the SMT solver.
-We now simply check for the satisfiability of $f(p_i)$ instead of $f(p_i) \wedge
+these new paths, but we slightly change the SMT formula given to the SMT solver:
+we now simply check for the satisfiability of $f(p_i)$ instead of $f(p_i) \wedge
 P$. Since we already know that $f(p_i) \wedge P$ is \emph{unsatisfiable}, all
@@
 	\begin{algorithmic}[1] 
-	\input{algo/compute-paths}
+	\input{algo/compute-paths.tex}
 	\end{algorithmic}
@@
 	\begin{algorithmic}[1] 
-	\input{algo/algo-combined}
+	\input{algo/algo-combined.tex}
 	\end{algorithmic}
@@
 \item the set $P'$ always verifies $P \cap P' = \emptyset$ by construction, 
-which guarantees that $P$
+which guarantees that $P'$
 will eventually be empty after a finite number of loop iterations.
@@
 state. In this case, we have three elements in $P_R$.
-We also unroll the loop once, in order to distinguish the paths that don't go
+We also unroll the loop once, in order to distinguish the paths that do not go
 through the loop from the others:

### 2011-09-27 papers/PLDI2012/implicitization.tex
add remarks in section on combined techniques

@@
 % authoryear    To obtain author/year citation style instead of numeric.
+
+\newcommand{\MM}[1]{{\color{blue} TODO(MM): #1}}
 
@@
 
+\MM{$X^s$ n'est pas defini.
+A priori, il suffit d'ajouter un truc du style ...
+``except that we call the abstract values $X_i^s$ instead of $X_i$ to leave room for another abstract value $X^d$ later in the algorithm'' ici.}
+
 The combination of these two techniques considers an ascending sequence of
@@
 such as binary decision diagrams.
+
+\MM{Je trouve qu'il manque une vue de haut niveau quelque part par ici. Dire qu'en gros, l'algo, c'est :
+
+\begin{algorithm}
+  \begin{algorithmic}
+    \STATE \textit{Subset} $\gets \emptyset$
+    \WHILE{true}
+    \STATE Ascending iteration on a \textit{Subset}
+    \STATE Narrow
+    \STATE Add new paths to \textit{Subset}
+    \ENDWHILE
+  \end{algorithmic}
+\end{algorithm}
+
+Sans ca, enchainer tout de suite sur ``ascending iterations'' est un peu brutal.
+}
 

### 2011-10-10 papers/PLDI2012/implicitization.tex
labels and references to algo

@@
 
-\MM{Je trouve qu'il manque une vue de haut niveau quelque part par ici. Dire qu'en gros, l'algo, c'est :
+\MM{Je trouve qu'il manque une vue de haut niveau quelque part par
+  ici. Dire qu'en gros, l'algo, c'est (en pseudo-code ou en langue naturelle) :
 
@@
     \WHILE{true}
-    \STATE Ascending iteration on a \textit{Subset}
-    \STATE Narrow
-    \STATE Add new paths to \textit{Subset}
+    \STATE Ascending iteration on a \textit{Subset} (lines~\ref{alg=start-ascending}-\ref{alg=end-ascending} in algorithm~\ref{alg=combined})
+    \STATE Narrow (line~\ref{alg=narrowing} in algorithm~\ref{alg=combined})
+    \STATE Add new paths to \textit{Subset} (lines~\ref{alg=start-add-paths}-\ref{alg=end-add-paths} in algorithm~\ref{alg=combined})
     \ENDWHILE
@@
 Sans ca, enchainer tout de suite sur ``ascending iterations'' est un peu brutal.
+
+Ensuite, on peut rafiner la définition de ``Subset'' ($A$ et $P$), et
+enchainer sur l'algo 2.
 }

### 2011-10-10 papers/PLDI2012/implicitization.tex
Fix references

--- 2011-09-27 papers/PLDI2012/implicitization.tex
@@
     \WHILE{true}
-    \STATE Ascending iteration on a \textit{Subset}
-    \STATE Narrow
-    \STATE Add new paths to \textit{Subset}
+    \STATE Ascending iteration on a \textit{Subset} (lines~\ref{alg=start-ascending}-\ref{alg=end-ascending} in algorithm~\ref{algo:combined})
+    \STATE Narrow (line~\ref{alg=narrowing} in algorithm~\ref{algo:combined})
+    \STATE Add new paths to \textit{Subset} (lines~\ref{alg=start-add-paths}-\ref{alg=end-add-paths} in algorithm~\ref{algo:combined})
     \ENDWHILE

### 2011-10-10 papers/PLDI2012/implicitization.tex
Typos and remarks

@@
 $m$, $\delta$ and $\sigma$ can be defined heuristically.
-For instance, one could defined $\sigma$ so that $\sigma(j,k)$ only depends on the
+For instance, one could define $\sigma$ so that $\sigma(j,k)$ only depends on the
 last transition of the path, or else construct it dynamically during the
@@
 that we can easily find in the model which abstract value of the disjunction we
-have to choose. The resulting formula that is given to the SMT solver is defined
+have to choose.
+\MM{Le ``we have to choose'' est un peu vague. Tu veux dire ``which
+  abstract value of the disjunction has to be chosen to make the
+  invariant grow'' ?}
+The resulting formula that is given to the SMT solver is defined
 by $g(p_i)$.
@@
 
+\MM{Comment cet algo se place par rapport à l'algo
+  \ref{algo:combined} ? Si je comprends bien, la boucle while
+  remplace l'algo \ref{algo:computepaths}, et l'initialisation au
+  début remplace le début de \ref{algo:combined}. Faudrait rendre ça
+  explicite.}
 \begin{algorithm}[!h]

### 2011-10-11 papers/PLDI2012/implicitization.tex
reorganisation du papier, suite aux remarques de Matthieu

@@
 unfeasible paths.
-In this section, we use the same notations as \ref{sec:path_focusing}.
+In this section, we use the same notations as \ref{sec:path_focusing}, except
+that we call the abstract values $X_i^s$ instead of $X_i$ to leave room for
+another abstract value $X_i^d$ later in the algorithm.
 
-\MM{$X^s$ n'est pas defini.
-A priori, il suffit d'ajouter un truc du style ...
-``except that we call the abstract values $X_i^s$ instead of $X_i$ to leave room for another abstract value $X^d$ later in the algorithm'' ici.}
-
-The combination of these two techniques considers an ascending sequence of
+The combination of these two techniques aims at first discovering a precise
+inductive invariant for a subset of paths between two points in $P_R$, 
+by the mean of ascending and narrowing iterations. When an
+inductive invariant has been found, we add new feasible paths to the subset and
+compute an inductive invariant for this new subset, starting with the results
+from the previous analysis.
+In other words, our technique considers an ascending sequence of
 subsets of the paths between two points in $P_R$.
-The current working subset of paths is stored using a compact representation,
-such as binary decision diagrams.
-
-\MM{Je trouve qu'il manque une vue de haut niveau quelque part par
-  ici. Dire qu'en gros, l'algo, c'est (en pseudo-code ou en langue naturelle) :
+We iterate the operations until we consider the whole program, i.e all the
+feasible paths. The result will then be an inductive invariant of the entire program.
 
@@
     \WHILE{true}
-    \STATE Ascending iteration on a \textit{Subset} (lines~\ref{alg=start-ascending}-\ref{alg=end-ascending} in algorithm~\ref{algo:combined})
+    \STATE Ascending iteration on \textit{Subset} (lines~\ref{alg=start-ascending}-\ref{alg=end-ascending} in algorithm~\ref{algo:combined})
     \STATE Narrow (line~\ref{alg=narrowing} in algorithm~\ref{algo:combined})
@@
 
-Sans ca, enchainer tout de suite sur ``ascending iterations'' est un peu brutal.
+%\MM{Je trouve qu'il manque une vue de haut niveau quelque part par
+%  ici. Dire qu'en gros, l'algo, c'est (en pseudo-code ou en langue naturelle) :
+%
+%Sans ca, enchainer tout de suite sur ``ascending iterations'' est un peu brutal.
+%
+%Ensuite, on peut rafiner la définition de ``Subset'' ($A$ et $P$), et
+%enchainer sur l'algo 2.
+%}
 
-Ensuite, on peut rafiner la définition de ``Subset'' ($A$ et $P$), et
-enchainer sur l'algo 2.
-}
+\subsection{Algorithm}
 
-\subsection{Ascending iterations}
+The algorithm for Guided static analysis on the implicit multigraph is described
+in Algorithm \ref{algo:combined}.
+
+The current working subset of paths, noted $P$, is stored using a
+compact representation, such as binary decision diagrams.
+We also maintain a set of active control points $A$, containing the points in $P_R$
+that still need to be processed. 
+We apply the Path-focused technique\cite{Monniaux_Gonnord_SAS11} to
+the elements of $A$ until $A$ is empty. 
+We also check which paths become feasible, so that we can include them into the
+next subset of paths at the next iteration. These new feasible paths are
+temporarily stored in a set $P'$. 
+When $A$ is empty, it means the
+invariant computation of the subgraph is finished. As proposed by
+\citet{DBLP:conf/sas/GopanR07}, we can do some narrowing iterations.
+These narrowing iterations allow to recover precision lost by
+widening, before taking into account new feasible paths.
+Thus, our technique combines both the advantages of \emph{Guided Static
+Analysis} and \emph{Path-focusing}.
+We then restart the analysis with a new subgraph if needed, by adding into $P$
+the paths in $P'$. The set of active points $A$ is assigned to $A'$, which is
+the set of control points that are starting points of a new feasible path in
+$P'$.
+
+\begin{algorithm}
+	\caption{Guided static analysis on implicit multigraph}
+	\label{algo:combined}
+	\begin{algorithmic}[1] 
+	\input{algo/algo-combined.tex}
+	\end{algorithmic}
+\end{algorithm}
+
+\subsection{Ascending iterations by Path-focusing}
 \label{subsec:ascending}
 
-The algorithm for computing an inductive invariant over a subgraph is the Path-focused algorithm from \citet{Monniaux_Gonnord_SAS11}.
+The algorithm for computing an inductive invariant over a subgraph is the
+Path-focused algorithm from \citet{Monniaux_Gonnord_SAS11} with special
+treatment for self loops (line~\ref{alg=pf} in algorithm~\ref{algo:combined}).
 
@@
 Intuitively, such a path makes the abstract value $X_{j}^s$ grow, and thus is
-an interesting path to focus on. When the formula becomes unsatisfiable, it
-means the analysis of $p_i$ is finished.
+an interesting path to focus on. We loop until the formula becomes unsatisfiable,
+meaning that the analysis of $p_i$ is finished.
 
@@
 exist such paths in the entire multigraph, that are not in $P$. The second phase
-computes these paths and adds them to $P'$. We also use SMT solving to discover
+computes these paths and adds them to $P'$ (line~\ref{alg=computeNewPaths} in
+algorithm~\ref{algo:combined}). We also use SMT solving to discover
 these new paths, but we slightly change the SMT formula given to the SMT solver:
@@
 
-\subsection{Algorithm}
-
-The algorithm for Guided static analysis on the implicit multigraph is described
-in Algorithm \ref{algo:combined}.
-
-We maintain a set of active control points $A$, containing the points in $P_R$
-that still need to be processed. We apply the two previously described phases to
-the elements of $A$ until $A$ is empty. When $A$ is empty, it means the
-invariant computation of the subgraph is finished. As proposed by
-\citet{DBLP:conf/sas/GopanR07}, we can do some narrowing iterations.
-These narrowing iterations allow to recover precision lost by
-widening, before taking into account new feasible paths.
-Thus, our technique combines both the advantages of \emph{Guided Static
-Analysis} and \emph{Path-focusing}.
-We then restart the analysis with a new subgraph if needed.
-
-\begin{algorithm}
-	\caption{Guided static analysis on implicit multigraph}
-	\label{algo:combined}
-	\begin{algorithmic}[1] 
-	\input{algo/algo-combined.tex}
-	\end{algorithmic}
-\end{algorithm}
-
+\subsection{Termination}
 Termination of this algorihm is guaranteed, because:
@@
 We introduce in this formula Boolean variables $\{d_j, 1 \leq j \leq m\}$, so
-that we can easily find in the model which abstract value of the disjunction we
-have to choose.
-\MM{Le ``we have to choose'' est un peu vague. Tu veux dire ``which
-  abstract value of the disjunction has to be chosen to make the
-  invariant grow'' ?}
+that we can easily find in the model which abstract value of the disjunction has
+to be chosen to make the invariant grow.
 The resulting formula that is given to the SMT solver is defined

### 2011-10-11 papers/PLDI2012/implicitization.tex
partie disjunctive invariant

@@
 
-Our improvement consists in a slight modification of the SMT formula we solve in
+Our improvement consists in a modification of the SMT formula we solve in
 \ref{sec:guided_multigraph}.
@@
 
-Our algorithm for computing disjunctive invariants is described in Algorithm
-\ref{algo:disjunctive}. In this algorithm, instead of simply joining the two
-abstract values, we use the widening operator to ensure the algorithm terminates
-when using abstract domains with infinite height.
-
-\MM{Comment cet algo se place par rapport à l'algo
-  \ref{algo:combined} ? Si je comprends bien, la boucle while
-  remplace l'algo \ref{algo:computepaths}, et l'initialisation au
-  début remplace le début de \ref{algo:combined}. Faudrait rendre ça
-  explicite.}
+In our algorithm, the initialization of the abstract values slightly differs from
+algorithm \ref{algo:combined} line~\ref{alg=X-init}, since we now have to
+initialize each disjunct. Line~\ref{alg=X-init} is then replaced by:
 \begin{algorithm}[!h]
-\caption{Disjunctive invariant computation with implicit paths}\label{gulwani2}
-\label{algo:disjunctive}
 \begin{algorithmic}[1] 
@@
 \STATE $X_{i,\delta} \gets I_{p_i}$
+\end{algorithmic}
+\end{algorithm}
+
+Furthermore, the Path-focused algorithm (line~\ref{alg=pf} from algorithm
+\ref{algo:combined}) is enhanced to deal with disjunctive invariants, and
+is detailed in algorithm \ref{algo:disjunctive}.
+
+The \emph{Update} function can classically assign to $X_{i,\sigma(j,k)}$ the
+value $X_{i,\sigma(j,k)} \widening (X_{i,\sigma(j,k)} \sqcup
+\tau_{i,k}(X_{i,j}))$, or can integrate the special treatment for self loops
+proposed by \citet{Monniaux_Gonnord_SAS11}, with widening/narrowing sequence or
+acceleration.
+
+\begin{algorithm}[!h]
+\caption{Disjunctive invariant computation with implicit paths}\label{gulwani2}
+\label{algo:disjunctive}
+\begin{algorithmic}[1] 
 \WHILE {true}
@@
 	\STATE Take $j \in \{ l | d_l = true\}$ 
-	\STATE $X_{i,\sigma(j,k)} \gets X_{i,\sigma(j,k)} \widening (X_{i,\sigma(j,k)} \sqcup \tau_{i,k}(X_{i,j}))$
+	\STATE Update($X_{i,\sigma(j,k)})$
 \ENDWHILE
@@
 \end{algorithm}
+
+The $\sigma$ function takes as parameters the index of the starting abstract
+value, and the path we focus on. If we dynamically construct this function
+during the analysis, we should store the results of the sigma function into a
+compact representation, such as Algebraic Decision Diagrams.
 

### 2011-10-11 papers/PLDI2012/implicitization.tex
ajout d'une phrase qui explique ce qu'est A
refs to Z3 and Yices

@@
 We also maintain a set of active control points $A$, containing the points in $P_R$
-that still need to be processed. 
+that still need to be processed. Each time the abstract value of a control point
+$p$ is updated, $p$ is inserted in $A$.
+
 We apply the Path-focused technique\cite{Monniaux_Gonnord_SAS11} to
@@
 Guided static analysis from \citet{DBLP:conf/sas/GopanR07} yields at line 5
-$\lstinline|x_old| \in (-\infty,+\infty)$.
+$\lstinline|x_old| \in~]-\infty,+\infty[$.
 Path-focused technique from \citet{Monniaux_Gonnord_SAS11} performs slightly
-better, and finds $\lstinline|x_old| \in (-\infty,10000)$.
+better, and finds $\lstinline|x_old| \in~]-\infty,10000]$.
 Our combined technique gives us the expected invariant 
-$\lstinline|x_old| \in (-10000,10000)$.
+$\lstinline|x_old| \in~[-10000,10000]$.
+
+In this example, we see that our technique actually combines best of
+\emph{Guided Static Analysis} and \emph{Path Focusing}. Still, we need to unroll
+the loop once in order to get precise results. Section \ref{sec:disjunctive}
+proposes a new technique that gives precise invariants without needing such
+graph transformations.
 
@@
 
+Our analyzer uses Yices \cite{DBLP:conf/cav/DutertreM06}
+ or Microsoft Z3\cite{DBLP:conf/tacas/MouraB08} as SMT-solver.
 \section{Conclusion}

### 2011-10-11 papers/PLDI2012/implicitization.tex
introduction de la partie experiments

@@
 
-Our analyzer uses Yices \cite{DBLP:conf/cav/DutertreM06}
- or Microsoft Z3\cite{DBLP:conf/tacas/MouraB08} as SMT-solver.
+We have implemented our proposed solutions inside a prototype intraprocedural
+static analyzer,
+as well as the classical abstract interpretation algorithm, and the state-of-the-art
+techniques \emph{Path Focusing} \cite{Monniaux_Gonnord_SAS11} and \emph{Guided
+Static Analysis} \cite{DBLP:conf/sas/GopanR07}.
+We also can choose among different abstract domains: convex polyhedra, octagons,
+intervals, grids (convex polyhedra and linear congruences).
+
+For the techniques involving SMT-solving, our analyzer uses Yices
+\cite{DBLP:conf/cav/DutertreM06} or Microsoft Z3\cite{DBLP:conf/tacas/MouraB08}
+as SMT-solver.
+
+We conducted extensive experiments on real-life programs in order to compare the
+different techniques. We also compared the precision of the analysis depending
+on the choice of the abstract domain.
+
 \section{Conclusion}

### 2011-10-11 papers/PLDI2012/implicitization.tex
+blurb SMT

--- 2011-10-11 papers/PLDI2012/implicitization.tex
@@
 \newcommand{\QQ}{\mathbb{Q}}
+\newcommand{\RR}{\mathbb{R}}
 \newcommand{\widening}{\mathop{\triangledown}}
@@
 \subsection{SMT-solving}
-TODO
+Boolean satisfiability (SAT) is the canonical NP-complete problem: given a propositional formula (e.g.
+$(a \lor \neg b) \land (\neg a \lor b \lor \neg c)$), decide whether it is satisfiable --- and, if so, output a satisfying assignment.
+Despite an exponential worst-case complexity, the DPLL algorithm \cite{Kroening_Strichman_08,Handbook_SAT} solves many useful SAT problems in practice.
+
+SAT was extended to \emph{satisfiability modulo theory} (SMT): in addition to propositional literals, SMT formulas admit atoms from a theory.
+For instance, the theories of linear integer arithmetic (LIA) and linear real arithmetic (LRA) have atoms of the form $a_1 x_1 + \dots + a_n x_n \bowtie C$ where $a_1,\dots,a_n,C$ are integer constants, $x_1\,dots,x_n$ are variables (interpreted over $\ZZ$ for LIA and $\RR$ or $\QQ$ for LRA), and $\bowtie$ is a comparison operator $=,\neq,<,\leq,>,\geq$.
+Satisfiability for LIA and LRA is NP-complete, yet tools based on DPLL(T) approach \cite{Kroening_Strichman_08,Handbook_SAT} solve many useful SMT problems in practice.
+All these tools provide a \emph{satisfying assignment} if the problem is satisfiable.
+
+Most SMT solvers, including Z3%
+\footnote{\url{http://research.microsoft.com/en-us/um/redmond/projects/z3/}}
+Yices,%
+\footnote{\url{http://yices.csl.sri.com/}}
+and all those supporting the full SMTLIB2 standard \cite{BarST-SMTLIB},
+offer an \emph{incremental} interface: the client program specifies the formula as an initially empty conjunction, to which additional constraints are added, and calls a ``check'' function answering whether it is satisfiable;
+it may then backtrack some of the constraints and add other ones without restarting from scratch.
 

### 2011-10-11 papers/PLDI2012/implicitization.tex
ajout d'une remarque dans disjunctive invariants
def Kleene

@@
 \subsection{Static analysis by abstract interpretation}
+\label{sec:static_analysis}
 Let $X$ be the set of possible states of the program variables; for instance, if the program has 3 unbounded integer variables, then $X = \ZZ^3$.
@@
 We also assume a join operator $\sqcap: \abstr{X} \times \abstr{X} \rightarrow \abstr{X}$, with infix notation; in practice, it is generally a least upper bound operation, but we only need it to satisfy $\gamma(\abstr{x}) \cup \gamma(\abstr{y}) \subseteq \gamma(\abstr{x} \sqcap \abstr{y})$ for all $\abstr{x},\abstr{y}$.
+
+Classically, one considers the control-flow graph of the program, with edges labeled with concrete transition relations (e.g.
+$x' = x+1$ for an instruction \lstinline|x = x+1;|), and attaches an abstract element to each control point.
+A concrete transition relation $\tau \subseteq X \times X$ is replaced by an abstract \emph{forward abstract transformer} $\abstr{\tau}: \abstr{X} \rightarrow \abstr{X}$, such that
+\begin{equation}
+\forall \abstr{x} \in \abstr{X}, x,x' \in X,~
+x \in \gamma(\abstr{x}) \land (x,x') \in \tau \implies
+x' \in \gamma \circ \abstr{\tau}(\abstr{x})
+\label{eqn:transformer_correct}
+\end{equation}
+It is easy to see that if to any control point $p \in P$ we attach an abstract element $\abstr{x}_p$ such that
+(i) for any $p$, $\gamma(\abstr{x}_p)$ includes all initial states possible at control node $p$
+(ii) for any $p,p'$, $\abstr{\tau}_{p,p'} (\abstr{x}_p) \sqsubseteq \abstr{x}_{p'}$, noting $\tau_{p,p'}$ the transition from $p$ to~$p'$, then $(\gamma(\abstr{x}_p))_{p \in P}$ form an \emph{inductive invariant}: by induction, the current program states always lie in $\gamma(\abstr{x}_p)$ at program location~$p$.
 
@@
 
-\emph{Path-focused} technique from \citet{Monniaux_Gonnord_SAS11} proposes to
-distinguish the different paths in the program in order to avoid loss of
+\citet{Monniaux_Gonnord_SAS11}'s \emph{path-focused} technique
+distinguishes the different paths in the program in order to avoid loss of
 precision due to merge operations. Since the number of paths may be exponential,
 the technique keeps them implicit and computes them when needed using
-SMT-solving. Then, the iterations are computed over a reduced multigraph instead
-of the classical transition graph.
+SMT-solving.
+The (accelerated) Kleene iterations (\S\ref{sec:static_analysis}) are computed over a reduced multigraph instead of the classical transition graph.
 

### 2011-10-11 papers/PLDI2012/implicitization.tex
abstract interpretation, suite

@@
 
-\usepackage{amsmath,amsfonts}
-\usepackage{mathcomp,amssymb}
+\usepackage{amsmath}
+\usepackage{mathcomp,amsfonts,amssymb,MnSymbol}
 \usepackage{dmnatbib}
@@
 (i) for any $p$, $\gamma(\abstr{x}_p)$ includes all initial states possible at control node $p$
-(ii) for any $p,p'$, $\abstr{\tau}_{p,p'} (\abstr{x}_p) \sqsubseteq \abstr{x}_{p'}$, noting $\tau_{p,p'}$ the transition from $p$ to~$p'$, then $(\gamma(\abstr{x}_p))_{p \in P}$ form an \emph{inductive invariant}: by induction, the current program states always lie in $\gamma(\abstr{x}_p)$ at program location~$p$.
+(ii) for any $p,p'$, $\abstr{\tau}_{p,p'} (\abstr{x}_p) \sqsubseteq \abstr{x}_{p'}$, noting $\tau_{p,p'}$ the transition from $p$ to~$p'$, then $(\gamma(\abstr{x}_p))_{p \in P}$ form an \emph{inductive invariant}: by induction, the program states always lie in $\gamma(\abstr{x}_p)$ at program location~$p$.
+
+\emph{Kleene iterations} compute such an inductive invariant as the stationary limit, if it exists, of the following system: for each $p$, initialize $\abstr{x}_p$ such that $\gamma(\abstr{x}_p)$ is a superset of the initial states at point $p$; then iterate the following: if $\abstr{\tau}_{p,p'} (\abstr{x}_p) \nsqsubseteq \abstr{x}_{p'}$, replace $\abstr{x}_{p'}$ by $\abstr{x}_{p'} \sqcup \abstr{\tau}_{p,p'} (\abstr{x}_p)$.
+Such a stationary limit is bound to exist if $\abstr{X}$ has no infinite ascending chain $a_1 \sqsubsetneq a_2 \sqsubsetneq \dots$; this condition is however not met by domains such as intervals or convex polyhedra.
 
@@
 during the analysis, we should store the results of the sigma function into a
-compact representation, such as Algebraic Decision Diagrams.
+compact representation, such as Algebraic Decision Diagrams. Notice that the
+image of a couple $(j,k)$ by the function $\sigma$ will be computed only if
+needed.
 

### 2011-10-11 papers/PLDI2012/implicitization.tex
static analysis

@@
 
-\emph{Kleene iterations} compute such an inductive invariant as the stationary limit, if it exists, of the following system: for each $p$, initialize $\abstr{x}_p$ such that $\gamma(\abstr{x}_p)$ is a superset of the initial states at point $p$; then iterate the following: if $\abstr{\tau}_{p,p'} (\abstr{x}_p) \nsqsubseteq \abstr{x}_{p'}$, replace $\abstr{x}_{p'}$ by $\abstr{x}_{p'} \sqcup \abstr{\tau}_{p,p'} (\abstr{x}_p)$.
+\emph{Kleene iterations} compute such an inductive invariant as the stationary limit, if it exists, of the following system: for each $p$, initialize $\abstr{x}_p$ such that $\gamma(\abstr{x}_p)$ is a superset of the initial states at point $p$; then iterate the following:
+if $\abstr{\tau}_{p,p'} (\abstr{x}_p) \nsqsubseteq \abstr{x}_{p'}$, replace $\abstr{x}_{p'}$ by $\abstr{x}_{p'} \sqcup \abstr{\tau}_{p,p'} (\abstr{x}_p)$.
 Such a stationary limit is bound to exist if $\abstr{X}$ has no infinite ascending chain $a_1 \sqsubsetneq a_2 \sqsubsetneq \dots$; this condition is however not met by domains such as intervals or convex polyhedra.
+
+\emph{Widening-accelerated Kleene iterations} proceed by replacing $\abstr{x}_{p'} \sqcup \abstr{\tau}_{p,p'} (\abstr{x}_p)$ by $\abstr{x}_{p'} \widening (\abstr{x}_{p'} \sqcup \abstr{\tau}_{p,p'} (\abstr{x}_p))$ where $\widening$ is a \emph{widening operator}: for all $\abstr{x},\abstr{y}$, $\gamma(\abstr{y}) \subseteq \gamma(\abstr{x} \widening \abstr{y})$, and there exists no sequence $\abstr{u}_1,\abstr{u}_2,\dots$ of the form $\abstr{u}_{n+1} = \abstr{u}_n \widening \abstr{v}_n$ where $\abstr{v}_n$ is another sequence.
+Such iterations necessarily lead to a stationary $(\abstr{x}_p)_{p \in P}$, which define an inductive invariant $(\gamma(\abstr{x}_p))_{p \in P}$.
+Note that this invariant is not, in general, the least one expressible in the abstract domain, and may depend on the iteration ordering (the successive choices~$p,p'$).
 

### 2011-10-11 papers/PLDI2012/implicitization.tex
ajout d'un figure dans l'exemple. Il reste à commenter

--- 2011-10-11 papers/PLDI2012/implicitization.tex
@@
 \newcommand{\MM}[1]{{\color{blue} TODO(MM): #1}}
+\newcommand{\JH}[1]{{\color{blue} TODO(JH): #1}}
 
@@
 \subsection{Static analysis by abstract interpretation}
-\label{sec:static_analysis}
 Let $X$ be the set of possible states of the program variables; for instance, if the program has 3 unbounded integer variables, then $X = \ZZ^3$.
@@
 We also assume a join operator $\sqcap: \abstr{X} \times \abstr{X} \rightarrow \abstr{X}$, with infix notation; in practice, it is generally a least upper bound operation, but we only need it to satisfy $\gamma(\abstr{x}) \cup \gamma(\abstr{y}) \subseteq \gamma(\abstr{x} \sqcap \abstr{y})$ for all $\abstr{x},\abstr{y}$.
-
-Classically, one considers the control-flow graph of the program, with edges labeled with concrete transition relations (e.g.
-$x' = x+1$ for an instruction \lstinline|x = x+1;|), and attaches an abstract element to each control point.
-A concrete transition relation $\tau \subseteq X \times X$ is replaced by an abstract \emph{forward abstract transformer} $\abstr{\tau}: \abstr{X} \rightarrow \abstr{X}$, such that
-\begin{equation}
-\forall \abstr{x} \in \abstr{X}, x,x' \in X,~
-x \in \gamma(\abstr{x}) \land (x,x') \in \tau \implies
-x' \in \gamma \circ \abstr{\tau}(\abstr{x})
-\label{eqn:transformer_correct}
-\end{equation}
-It is easy to see that if to any control point $p \in P$ we attach an abstract element $\abstr{x}_p$ such that
-(i) for any $p$, $\gamma(\abstr{x}_p)$ includes all initial states possible at control node $p$
-(ii) for any $p,p'$, $\abstr{\tau}_{p,p'} (\abstr{x}_p) \sqsubseteq \abstr{x}_{p'}$, noting $\tau_{p,p'}$ the transition from $p$ to~$p'$, then $(\gamma(\abstr{x}_p))_{p \in P}$ form an \emph{inductive invariant}: by induction, the current program states always lie in $\gamma(\abstr{x}_p)$ at program location~$p$.
 
@@
 
-\citet{Monniaux_Gonnord_SAS11}'s \emph{path-focused} technique
-distinguishes the different paths in the program in order to avoid loss of
+\emph{Path-focused} technique from \citet{Monniaux_Gonnord_SAS11} proposes to
+distinguish the different paths in the program in order to avoid loss of
 precision due to merge operations. Since the number of paths may be exponential,
 the technique keeps them implicit and computes them when needed using
-SMT-solving.
-The (accelerated) Kleene iterations (\S\ref{sec:static_analysis}) are computed over a reduced multigraph instead of the classical transition graph.
+SMT-solving. Then, the iterations are computed over a reduced multigraph instead
+of the classical transition graph.
 
@@
 \subsection{Termination}
-Termination of this algorihm is guaranteed, because:
+Termination of this algorithm is guaranteed, because:
 \begin{itemize}
@@
 better, and finds $\lstinline|x_old| \in~]-\infty,10000]$.
+
+Now, see how our technique performs on this example.
 Our combined technique gives us the expected invariant 
 $\lstinline|x_old| \in~[-10000,10000]$.
+
+\JH{Expliquer la figure 2}
+
+\begin{figure}
+\label{fig:multigraph}
+\centering
+\begin{minipage}[c]{.10\textwidth}
+\begin{tikzpicture}[->,>=stealth',auto,node distance=2.1cm,
+                    semithick,font=\footnotesize]
+
+	\node[PRstate] (n00) {$p_0$};
+	\node[PRstate] (n0) [below of=n00, yshift=0.8cm] {$p_1$};
+	\node[PRstate] (n1) [below of=n0] {$p_2$};
+	\node (label) [below of=n1, yshift=1.2cm] {Step 1};
+
+  \path [transition] 
+		(n00) edge              node {} (n0);
+\end{tikzpicture}
+\end{minipage} 
+\begin{minipage}[c]{.21\textwidth}
+\begin{tikzpicture}[->,>=stealth',auto,node distance=2.1cm,
+                    semithick,font=\footnotesize]
+
+	\node[PRstate] (n00) {$p_0$};
+	\node[PRstate] (n0) [below of=n00, yshift=0.8cm] {$p_1$};
+	\node[PRstate] (n1) [below of=n0] {$p_2$};
+	\node (label) [below of=n1, yshift=1.2cm] {Step 2};
+
+  \path [transition] 
+		(n00) edge  node {} (n0);
+  \path [transition] 
+		(n0) edge  [loop right] node [right] {$\neg c_7 \wedge c_8 \wedge \neg c_9$} (n0);
+  \path [transition] 
+		(n0) edge  [loop left] node [left] {$\neg c_7 \wedge c_8 \wedge \neg c_9$} (n0);
+  \path [transition] 
+		(n0) edge  [out=-45, in=45] node [right] {$\neg c_7 \wedge \neg c_8 \wedge c_9$} (n1);
+  \path [transition] 
+		(n0) edge  [out=-135, in=135] node [left] {$\neg c_7 \wedge \neg c_8 \wedge c_9$} (n1);
+\end{tikzpicture}
+\end{minipage}
+\caption{Ascending sequence of subgraphs}
+\end{figure}
 

### 2011-10-11 papers/PLDI2012/implicitization.tex
exemple

@@
 Now, see how our technique performs on this example.
-Our combined technique gives us the expected invariant 
-$\lstinline|x_old| \in~[-10000,10000]$.
 
-\JH{Expliquer la figure 2}
+Figure \ref{fig:example-graph} shows the sequence of subset of paths during the
+analysis. The points in $P_R$ are noted $p_i$, where $i$ is the corresponding
+line in the code: for instance, $p_5$ corresponds to the header of the main
+loop.
+
+\begin{enumerate}
+	\item The first non-empty subset of paths is depicted on Figure
+		\ref{fig:example-graph} Step 1. We apply Path-Focusing on this graph,
+		and discover for $p_5$ the inductive invariant $\lstinline|x_old| = 0$.
+		We also compute the set $P'$ of paths that have to be added at the next
+		iteration. 
+		For describing the different paths in the loop, we introduce the Boolean
+		values $c_7, c_8, c_9$, where $c_i$ is \emph{true} when the path goes
+		through the \emph{then} branch of the \emph{if-then-else} at line $i$, and
+		\emph{false} when the path goes through the \emph{else} branch.
+		We then add into $P'$ the following paths:
+\JH{A compl\'eter}
+\end{enumerate}
+
+%Our combined technique gives us the expected invariant 
+%$\lstinline|x_old| \in~[-10000,10000]$.
 
 \begin{figure}
-\label{fig:multigraph}
+\label{fig:example-graph}
 \centering
@@
 
-	\node[PRstate] (n00) {$p_0$};
-	\node[PRstate] (n0) [below of=n00, yshift=0.8cm] {$p_1$};
-	\node[PRstate] (n1) [below of=n0] {$p_2$};
+	\node[PRstate] (n00) {$p_4$};
+	\node[PRstate] (n0) [below of=n00, yshift=0.8cm] {$p_5$};
+	\node[PRstate] (n1) [below of=n0] {$p_{11}$};
 	\node (label) [below of=n1, yshift=1.2cm] {Step 1};
@@
 
-	\node[PRstate] (n00) {$p_0$};
-	\node[PRstate] (n0) [below of=n00, yshift=0.8cm] {$p_1$};
-	\node[PRstate] (n1) [below of=n0] {$p_2$};
+	\node[PRstate] (n00) {$p_4$};
+	\node[PRstate] (n0) [below of=n00, yshift=0.8cm] {$p_5$};
+	\node[PRstate] (n1) [below of=n0] {$p_{11}$};
 	\node (label) [below of=n1, yshift=1.2cm] {Step 2};

### 2011-10-13 papers/PLDI2012/implicitization.tex
exemple

@@
 \label{subsec:rate_lim}
+\JH{les tests des lignes 7 et 8 ont \'et\'e remplac\'es par des in\'egalit\'es
+strictes}
 Consider the following program, adapted from \cite{Monniaux_Gonnord_SAS11}:
@@
 \begin{enumerate}
-	\item The first non-empty subset of paths is depicted on Figure
+	\item The first subset of paths is depicted on Figure
 		\ref{fig:example-graph} Step 1. We apply Path-Focusing on this graph,
 		and discover for $p_5$ the inductive invariant $\lstinline|x_old| = 0$.
-		We also compute the set $P'$ of paths that have to be added at the next
-		iteration. 
-		For describing the different paths in the loop, we introduce the Boolean
-		values $c_7, c_8, c_9$, where $c_i$ is \emph{true} when the path goes
-		through the \emph{then} branch of the \emph{if-then-else} at line $i$, and
-		\emph{false} when the path goes through the \emph{else} branch.
-		We then add into $P'$ the following paths:
-\JH{A compl\'eter}
+	\item
+		We then compute the set $P'$ of paths that have to be added into the
+		next subgraph. 
+		The image of $\lstinline|x_old| = 0$ by the path that goes from $p_5$ to
+		itself, and that goes through the \emph{else} branch of each 
+		\emph{if-then-else}, is $-10 \leq \lstinline|x_old| \leq 10$. This path
+		is then added to our subgraph. Moreover, there is no other path whose
+		image is not in $-10 \leq \lstinline|x_old| \leq 10$.
+		Additionnally, the path that goes from $p_5$ to
+		$p_11$, that goes through the \emph{else} branch of each 
+		\emph{if-then-else} at lines 7,8 and 9, is also 
+		$-10 \leq \lstinline|x_old| \leq 10$, and thus is also added into the
+		new subgraph. Again, there is no new path that goes to $p_11$ with an
+		image outside of $-10 \leq \lstinline|x_old| \leq 10$.
+	\item
+		We apply Path Focusing on this new subgraph (\ref{fig:example-graph}
+		Step 2). When we have reached an inductive invariant, we narrow, and the
+		result for both $p_5$ and $p_11$ is the polyhedron 
+		$-10000 \leq \lstinline|x_old| \leq 10000$.
+	\item Finally, we compute the new subgraph. The SMT-solver does not find any
+		new path that makes the abstract values grow, and the algorithm
+		terminates.
 \end{enumerate}
 
-%Our combined technique gives us the expected invariant 
-%$\lstinline|x_old| \in~[-10000,10000]$.
+Our technique gives us the expected invariant 
+$\lstinline|x_old| \in~[-10000,10000]$. 
+Here, only 3 paths over the 25 have been computed during the analysis. In
+practice, depending on the order the SMT-solver returns the paths, other
+feasible paths could have been added during the analysis.
 
@@
 \centering
-\begin{minipage}[c]{.10\textwidth}
+\begin{minipage}[c]{.15\textwidth}
 \begin{tikzpicture}[->,>=stealth',auto,node distance=2.1cm,
@@
   \path [transition] 
-		(n00) edge              node {} (n0);
+		(n00) edge  node {$\lstinline|x_old| \gets 0$} (n0);
 \end{tikzpicture}
 \end{minipage} 
-\begin{minipage}[c]{.21\textwidth}
+\begin{minipage}[c]{.28\textwidth}
 \begin{tikzpicture}[->,>=stealth',auto,node distance=2.1cm,
@@
   \path [transition] 
-		(n00) edge  node {} (n0);
+		(n00) edge  node {$\lstinline|x_old| \gets 0$} (n0);
   \path [transition] 
-		(n0) edge  [loop right] node [right] {$\neg c_7 \wedge c_8 \wedge \neg c_9$} (n0);
+		(n0) edge  [loop left] node [left, xshift=0.2cm] {
+		$\begin{array}{r}
+			-10000 \leq \lstinline|x| \leq 10000 \\
+			\lstinline|x_old|-10 \leq \lstinline|x| \leq \lstinline|x_old|+10 /\\
+			\lstinline|x_old| \gets \lstinline|x|
+		\end{array}$
+		} (n0);
   \path [transition] 
-		(n0) edge  [loop left] node [left] {$\neg c_7 \wedge c_8 \wedge \neg c_9$} (n0);
-  \path [transition] 
-		(n0) edge  [out=-45, in=45] node [right] {$\neg c_7 \wedge \neg c_8 \wedge c_9$} (n1);
-  \path [transition] 
-		(n0) edge  [out=-135, in=135] node [left] {$\neg c_7 \wedge \neg c_8 \wedge c_9$} (n1);
+		(n0) edge node [left, yshift=-0.2cm, xshift=0.2cm] {
+		$\begin{array}{r}
+			-10000 \leq \lstinline|x| \leq 10000 \\
+			\lstinline|x_old|-10 \leq \lstinline|x| \leq \lstinline|x_old|+10 /\\
+			\lstinline|x_old| \gets \lstinline|x|
+		\end{array}$
+		} (n1);
 \end{tikzpicture}

### 2011-10-13 papers/PLDI2012/rlim_int.c

@@ -4,8 +4,8 @@ void rate_limiter() {
   while (1) {
     int x = input();
     if (x < -100000 || x > 100000) x=0;
-    if (x >= x_old+10) x = x_old+10;
-    if (x <= x_old-10) x = x_old-10;
+    if (x > x_old+10) x = x_old+10;
+    if (x < x_old-10) x = x_old-10;
     x_old = x;
   }
 }

### 2011-10-13 papers/PLDI2012/rlim_int2.c

@@ -4,8 +4,8 @@ void rate_limiter() {
   while (1) {
     int x = input();
     if (x < -100000 || x > 100000) x=0;
-    if (x >= x_old+10) x = x_old+10;
-    else if (x <= x_old-10) x = x_old-10;
+    if (x > x_old+10) x = x_old+10;
+    else if (x < x_old-10) x = x_old-10;
     else x_old = x;
   }
 }

### 2011-10-13 papers/PLDI2012/rlim_int_loop.c

@@ -5,8 +5,8 @@ void rate_limiter() {
   while (1) {
     int x = input();
     if (x < -100000 || x > 100000) x=0;
-    if (x >= x_old+10) x = x_old+10;
-    if (x <= x_old-10) x = x_old-10;
+    if (x > x_old+10) x = x_old+10;
+    if (x < x_old-10) x = x_old-10;
     x_old = x;
     while (wait()) {}
   }

### 2011-10-18 papers/PLDI2012/implicitization.tex
modification des algorithmes dans le papier

@@
 \newcommand{\widening}{\mathop{\triangledown}}
+
+% commentaires dans les algorithmes crees avec le package algorithmic
+\renewcommand{\algorithmiccomment}[1]{ {\color{gray} // #1}}
 
@@
     \WHILE{true}
+    \STATE Add new paths to \textit{Subset} (lines~\ref{alg=start-add-paths}-\ref{alg=end-add-paths} in algorithm~\ref{algo:combined})
     \STATE Ascending iteration on \textit{Subset} (lines~\ref{alg=start-ascending}-\ref{alg=end-ascending} in algorithm~\ref{algo:combined})
     \STATE Narrow (line~\ref{alg=narrowing} in algorithm~\ref{algo:combined})
-    \STATE Add new paths to \textit{Subset} (lines~\ref{alg=start-add-paths}-\ref{alg=end-add-paths} in algorithm~\ref{algo:combined})
     \ENDWHILE
@@
 compact representation, such as binary decision diagrams.
-We also maintain a set of active control points $A$, containing the points in $P_R$
-that still need to be processed. Each time the abstract value of a control point
-$p$ is updated, $p$ is inserted in $A$.
+We also maintain two sets of control points:
+\begin{itemize}
+	\item $A'$ contains the points in $P_R$ that may be the starting points of new
+		feasible paths.
+	\item $A$ contains the points in $P_R$ on which we apply the ascending iterations.
+	Each time the abstract value of a control point $p$ is updated, $p$ is
+	inserted in both $A$ and $A'$.
+\end{itemize}
 
+We first check which paths become feasible, so that we can include them into the
+next subset of paths. The set $P'$ of new feasible paths is added to the
+previous set $P$. Each starting points of paths in $P'$ is added into $A$.
 We apply the Path-focused technique\cite{Monniaux_Gonnord_SAS11} to
 the elements of $A$ until $A$ is empty. 
-We also check which paths become feasible, so that we can include them into the
-next subset of paths at the next iteration. These new feasible paths are
-temporarily stored in a set $P'$. 
-When $A$ is empty, it means the
+When $A$ becomes empty, it means the
 invariant computation of the subgraph is finished. As proposed by
@@
 These narrowing iterations allow to recover precision lost by
-widening, before taking into account new feasible paths.
+widening, before computing and taking into account new feasible paths.
 Thus, our technique combines both the advantages of \emph{Guided Static
 Analysis} and \emph{Path-focusing}.
-We then restart the analysis with a new subgraph if needed, by adding into $P$
-the paths in $P'$. The set of active points $A$ is assigned to $A'$, which is
-the set of control points that are starting points of a new feasible path in
-$P'$.
+
+After narrowing, we restart the analysis, starting from the computation the new
+subgraph.
 
@@
 $P'$ of new paths that become feasible regarding the current abstract values.
-$P'$ is progressively constructed during the analysis of the subgraph.
 
-We distinguish two phases when analysing a point $p_i \in P_R$:
+We distinguish two phases when analysing a subgraph $P$:
 \begin{enumerate}
-\item We search for paths starting at point $p_i$, such that these paths are in
-$P$, i.e are included in the working subgraph. Each time we find a path, we
-update the abstract value of the destination point of the path. This is the
-phase explained in \ref{subsec:ascending}.
+\item We search for paths starting at point $p_i \in P_R$, such that these paths
+	are in $P$, i.e are included in the working subgraph. Each time we find a
+	path, we update the abstract value of the destination point of the path.
+	This is the phase explained in \ref{subsec:ascending}.
 \item  When the previous phase terminates, there is no more paths in
@@
 that make the abstract values of the successors grow (otherwise, the SMT
-solver would not have answered ``\emph{unsat}''). However, there may
-exist such paths in the entire multigraph, that are not in $P$. The second phase
+solver would not have answered ``\emph{unsat}''). Narrowing iterations preserves
+this property. 
+However, there may exist such paths in the entire multigraph, that are not in
+$P$. The second phase
 computes these paths and adds them to $P'$ (line~\ref{alg=computeNewPaths} in

### 2011-10-19 papers/PLDI2012/implicitization.tex
Different color for JH's TODO
Fix a reference (contr: -> sec:)

@@
 \newcommand{\MM}[1]{{\color{blue} TODO(MM): #1}}
-\newcommand{\JH}[1]{{\color{blue} TODO(JH): #1}}
+\newcommand{\JH}[1]{{\color{teal} TODO(JH): #1}}
+\newcommand{\DM}[1]{{\color{violet} TODO(DM): #1}}
 
@@
 \begin{enumerate}
-\item We recast the guided static analysis technique from \citet{DBLP:conf/sas/GopanR07} on the expanded multigraph from \citet{Monniaux_Gonnord_SAS11}, considering entire paths instead of individual transitions, using SMT queries and binary decision diagrams.\label{contr:guided_multigraph} (See \S\ref{contr:guided_multigraph})
-\item We improve the technique for obtaining disjunctive invariants from \citet{DBLP:conf/pldi/GulwaniZ10} by replacing the explicit exhaustive enumeration of paths by a sequence of SMT queries.\label{contr:disjunctive} (See \S\ref{contr:disjunctive})
+\item We recast the guided static analysis technique from \citet{DBLP:conf/sas/GopanR07} on the expanded multigraph from \citet{Monniaux_Gonnord_SAS11}, considering entire paths instead of individual transitions, using SMT queries and binary decision diagrams.\label{contr:guided_multigraph} (See \S\ref{sec:guided_multigraph})
+\item We improve the technique for obtaining disjunctive invariants from \citet{DBLP:conf/pldi/GulwaniZ10} by replacing the explicit exhaustive enumeration of paths by a sequence of SMT queries.\label{contr:disjunctive} (See \S\ref{sec:disjunctive})
 \item We implemented these techniques, in addition to ``classical'' iterations and the original guided static analysis, inside a prototype static analyzer.

### 2011-10-21 papers/PLDI2012/implicitization.tex
fixes in intro

@@
 \usepackage{listings}
-\lstset{language=C,mathescape=true,numbers=left} % basicstyle={\rm\fontfamily{ppl}\selectfont} trop reconnaissable
+\lstset{language=C,mathescape=true,numbers=left,numbersep=0.8ex} % basicstyle={\rm\fontfamily{ppl}\selectfont} trop reconnaissable
 
@@
 \item We improve the technique for obtaining disjunctive invariants from \citet{DBLP:conf/pldi/GulwaniZ10} by replacing the explicit exhaustive enumeration of paths by a sequence of SMT queries.\label{contr:disjunctive} (See \S\ref{sec:disjunctive})
+% TODO: 'We implemented' pas une contribution; rephrase.
 \item We implemented these techniques, in addition to ``classical'' iterations and the original guided static analysis, inside a prototype static analyzer.
@@
 Let $X$ be the set of possible states of the program variables; for instance, if the program has 3 unbounded integer variables, then $X = \ZZ^3$.
-The set $\parts{X}$ of subsets of $X$, partially ordered by inclusion, is often referred to as the \emph{concrete domain}.
+The set $\parts{X}$ of subsets of $X$, partially ordered by inclusion, is the \emph{concrete domain}.
 An \emph{abstract domain} is a set $\abstr{X}$ equipped with a partial order $\sqsubseteq$; for instance, it can be the domain of convex polyhedra in $\QQ^3$ ordered by geometric inclusion.
@@
 (i) for any $p$, $\gamma(\abstr{x}_p)$ includes all initial states possible at control node $p$
-(ii) for any $p,p'$, $\abstr{\tau}_{p,p'} (\abstr{x}_p) \sqsubseteq \abstr{x}_{p'}$, noting $\tau_{p,p'}$ the transition from $p$ to~$p'$, then $(\gamma(\abstr{x}_p))_{p \in P}$ form an \emph{inductive invariant}: by induction, the program states always lie in $\gamma(\abstr{x}_p)$ at program location~$p$.
+(ii) for any $p,p'$, $\abstr{\tau}_{p,p'} (\abstr{x}_p) \sqsubseteq \abstr{x}_{p'}$, noting $\tau_{p,p'}$ the transition from $p$ to~$p'$, then $(\gamma(\abstr{x}_p))_{p \in P}$ form an \emph{inductive invariant}: by induction, when the control point is $p$, the program state always lies in $\gamma(\abstr{x}_p)$.
 
@@
 \emph{Widening-accelerated Kleene iterations} proceed by replacing $\abstr{x}_{p'} \sqcup \abstr{\tau}_{p,p'} (\abstr{x}_p)$ by $\abstr{x}_{p'} \widening (\abstr{x}_{p'} \sqcup \abstr{\tau}_{p,p'} (\abstr{x}_p))$ where $\widening$ is a \emph{widening operator}: for all $\abstr{x},\abstr{y}$, $\gamma(\abstr{y}) \subseteq \gamma(\abstr{x} \widening \abstr{y})$, and there exists no sequence $\abstr{u}_1,\abstr{u}_2,\dots$ of the form $\abstr{u}_{n+1} = \abstr{u}_n \widening \abstr{v}_n$ where $\abstr{v}_n$ is another sequence.
-Such iterations necessarily lead to a stationary $(\abstr{x}_p)_{p \in P}$, which define an inductive invariant $(\gamma(\abstr{x}_p))_{p \in P}$.
+Such iterations necessarily lead to a stationary limit $(\abstr{x}_p)_{p \in P}$, which defines an inductive invariant $(\gamma(\abstr{x}_p))_{p \in P}$.
 Note that this invariant is not, in general, the least one expressible in the abstract domain, and may depend on the iteration ordering (the successive choices~$p,p'$).
@@
 SAT was extended to \emph{satisfiability modulo theory} (SMT): in addition to propositional literals, SMT formulas admit atoms from a theory.
-For instance, the theories of linear integer arithmetic (LIA) and linear real arithmetic (LRA) have atoms of the form $a_1 x_1 + \dots + a_n x_n \bowtie C$ where $a_1,\dots,a_n,C$ are integer constants, $x_1\,dots,x_n$ are variables (interpreted over $\ZZ$ for LIA and $\RR$ or $\QQ$ for LRA), and $\bowtie$ is a comparison operator $=,\neq,<,\leq,>,\geq$.
+For instance, the theories of linear integer arithmetic (LIA) and linear real arithmetic (LRA) have atoms of the form $a_1 x_1 + \dots + a_n x_n \bowtie C$ where $a_1,\dots,a_n,C$ are integer constants, $x_1,\dots,x_n$ are variables (interpreted over $\ZZ$ for LIA and $\RR$ or $\QQ$ for LRA), and $\bowtie$ is a comparison operator $=,\neq,<,\leq,>,\geq$.
 Satisfiability for LIA and LRA is NP-complete, yet tools based on DPLL(T) approach \cite{Kroening_Strichman_08,Handbook_SAT} solve many useful SMT problems in practice.

### 2011-10-21 papers/PLDI2012/implicitization.tex
input(a, b) mieux que input() suivi de tests pour exclure out-of-range

@@
 \label{subsec:rate_lim}
-\JH{les tests des lignes 7 et 8 ont \'et\'e remplac\'es par des in\'egalit\'es
-strictes}
-Consider the following program, adapted from \cite{Monniaux_Gonnord_SAS11}:
+Consider the following program, adapted from \cite{Monniaux_Gonnord_SAS11}, where \lstinline|input($a$, $b$)| stands for a nondeterministic input in $[a,b]$:
 \lstinputlisting{rlim_int.c}
@@
 Widening iterations converge to $\lstinline|x_old| \in (-\infty,+\infty)$; let us now see why decreasing iterations fail to recover the desired invariant. % TODO expliquer widening suivi de narrowing
-The \lstinline|x >= x_old+10| test at line~7, if taken, yields $\lstinline|x_old| \in (-\infty,99990)$; followed by \lstinline|x = x_old+10|, we obtain $\lstinline|x| \in (-\infty,100000]$, and the same after union with the no-operation ``else'' branch.
-Line 8 yields $\lstinline|x| \in (-\infty,+\infty)$.
+The \lstinline|x >= x_old+10| test at line~6, if taken, yields $\lstinline|x_old| \in (-\infty,99990)$; followed by \lstinline|x = x_old+10|, we obtain $\lstinline|x| \in (-\infty,\allowbreak 100000]$, and the same after union with the no-operation ``else'' branch.
+Line~7 yields $\lstinline|x| \in (-\infty,+\infty)$.
 
-We could use ``guided widening'' or ``widening with thresholds'', propagating the ``magic values'' $\pm 100000$ associated to \lstinline|x| into~\lstinline|x_old|, but these syntactic approaches cannot directly cope with programs for which $\lstinline|x|  \in [-100000,+100000]$ is itself obtained by analysis.
+We could use ``widening up to'' or ``widening with thresholds'', propagating the ``magic values'' $\pm 100000$ associated to \lstinline|x| into~\lstinline|x_old|, but these syntactic approaches cannot directly cope with programs for which $\lstinline|x|  \in [-100000,+100000]$ is itself obtained by analysis.
 The guided static analysis of \citet{DBLP:conf/sas/GopanR07} performs slightly better, and obtains $\verb|x_old| \in (-\infty,100000]$.
 
-In contrast, let us distinguish all four possible execution paths through the tests at lines 7 and~8. The path through both ``else'' branches is infeasible; the program is thus equivalent to:
+In contrast, let us distinguish all four possible execution paths through the tests at lines 6 and~7. The path through both ``else'' branches is infeasible; the program is thus equivalent to:
 \lstinputlisting{rlim_int2.c}

### 2011-10-21 papers/PLDI2012/rlim_int.c

@@ -2,8 +2,7 @@ extern int input();
 void rate_limiter() {
   int x_old = 0;
   while (1) {
-    int x = input();
-    if (x < -100000 || x > 100000) x=0;
+    int x = input(-100000, 100000);
     if (x > x_old+10) x = x_old+10;
     if (x < x_old-10) x = x_old-10;
     x_old = x;

### 2011-10-21 papers/PLDI2012/rlim_int2.c

@@ -2,8 +2,7 @@ extern int input();
 void rate_limiter() {
   int x_old = 0;
   while (1) {
-    int x = input();
-    if (x < -100000 || x > 100000) x=0;
+    int x = input(-100000, 100000);
     if (x > x_old+10) x = x_old+10;
     else if (x < x_old-10) x = x_old-10;
     else x_old = x;

### 2011-10-21 papers/PLDI2012/rlim_int_loop.c

@@ -3,8 +3,7 @@ extern int wait();
 void rate_limiter() {
   int x_old = 0;
   while (1) {
-    int x = input();
-    if (x < -100000 || x > 100000) x=0;
+    int x = input(-100000, 100000);
     if (x > x_old+10) x = x_old+10;
     if (x < x_old-10) x = x_old-10;
     x_old = x;

### 2011-10-21 papers/PLDI2012/implicitization.tex
typo join operator

--- 2011-10-19 papers/PLDI2012/implicitization.tex
@@
 
-We also assume a join operator $\sqcap: \abstr{X} \times \abstr{X} \rightarrow \abstr{X}$, with infix notation; in practice, it is generally a least upper bound operation, but we only need it to satisfy $\gamma(\abstr{x}) \cup \gamma(\abstr{y}) \subseteq \gamma(\abstr{x} \sqcap \abstr{y})$ for all $\abstr{x},\abstr{y}$.
+We also assume a join operator $\sqcup: \abstr{X} \times \abstr{X} \rightarrow \abstr{X}$, with infix notation; in practice, it is generally a least upper bound operation, but we only need it to satisfy $\gamma(\abstr{x}) \cup \gamma(\abstr{y}) \subseteq \gamma(\abstr{x} \sqcup \abstr{y})$ for all $\abstr{x},\abstr{y}$.
 
@@
 
+In order to compare the various techniques, we compare by inclusion the obtained
+abstract values at each control points in $P_R$.
+
+\subsection{Precision of the various techniques}
+
+\subsection{Precision of Abstract Domains}
+
 \section{Conclusion}

### 2011-10-21 papers/PLDI2012/implicitization.tex
experiments

@@
 
+In this section, our experimental results come from the analysis of various
+open-source projects listed in Figure \ref{fig:projects}, written in C, C++ and
+Fortran.
+\begin{figure}[!h]
+	\centering
+\begin{tabular}{|l|l|l|} \hline
+	Name & LOC & $|P_R|$ \\ \hline
+	a2ps & 55k & 2012\\
+	gawk & 59k & 902\\ 
+	gnuchess & 38k & 1222\\ 
+	gnugo & 83k & 2801\\
+	grep & 35k & 820\\
+	gzip & 27k & 494\\
+	lapack/blas & 954k & 16422\\
+	make & 34k & 993\\ 
+	sed & 23k & 292\\
+	tar & 73k & 1712\\
+	\hline
+\end{tabular}
+\caption{List of analyzed open-source projects, with their respective number of
+lines of code, and their number of control points in $P_R$}
+\label{fig:projects}
+\end{figure}
+
 \subsection{Precision of the various techniques}

### 2011-10-21 papers/PLDI2012/implicitization.tex
graphique resultats
experiments

@@
 \usepackage{algorithmic,algorithm}
+%\usepackage{graphicx}
+%\usepackage{color}
 
@@
 
+\begin{figure}[h]
+  \begin{center}
+    \input{./techniques.tex}
+  \end{center} 
+  \vspace{-20px}
+  \caption{Comparison of the abstract values obtained by classical abstract
+  interpretation (S), \emph{Guided Static
+  Analysis} (G), \emph{Path-focused} technique (PF), our combined technique
+  (G+PF), and its version with disjunctive invariants (DIS).}
+  \label{fig:techniques}
+\end {figure}
+
+\JH{La comparaison DIS/G+PF n'est pas encore correcte (DIS pas compl\`etement
+impl\'ement\'e)\\} 
+Figures \ref{fig:techniques} shows the results of various comparisons between
+the different analysis techniques.
+
 \subsection{Precision of Abstract Domains}

### 2011-10-24 papers/PLDI2012/implicitization.tex
petites modifs

--- 2011-10-21 papers/PLDI2012/implicitization.tex
@@
 \usepackage{listings}
-\lstset{language=C,mathescape=true,numbers=left} % basicstyle={\rm\fontfamily{ppl}\selectfont} trop reconnaissable
+\lstset{language=C,mathescape=true,numbers=left,numbersep=0.8ex,tabsize=2} % basicstyle={\rm\fontfamily{ppl}\selectfont} trop reconnaissable
 
@@
 
-We also assume a join operator $\sqcup: \abstr{X} \times \abstr{X} \rightarrow \abstr{X}$, with infix notation; in practice, it is generally a least upper bound operation, but we only need it to satisfy $\gamma(\abstr{x}) \cup \gamma(\abstr{y}) \subseteq \gamma(\abstr{x} \sqcup \abstr{y})$ for all $\abstr{x},\abstr{y}$.
+We also assume a join operator $\sqcap: \abstr{X} \times \abstr{X} \rightarrow \abstr{X}$, with infix notation; in practice, it is generally a least upper bound operation, but we only need it to satisfy $\gamma(\abstr{x}) \cup \gamma(\abstr{y}) \subseteq \gamma(\abstr{x} \sqcap \abstr{y})$ for all $\abstr{x},\abstr{y}$.
 
@@
 
-In order to compare the various techniques, we compare by inclusion the obtained
-abstract values at each control points in $P_R$.
-
-\subsection{Precision of the various techniques}
-
-\subsection{Precision of Abstract Domains}
-
 \section{Conclusion}

### 2011-10-24 papers/PLDI2012/implicitization.tex

@@
 
+In order to compare the various techniques, we compare by inclusion the obtained
+abstract values at each control points in $P_R$.
+
+In this section, our experimental results come from the analysis of various
+open-source projects listed in Figure \ref{fig:projects}, written in C, C++ and
+Fortran.
+\begin{figure}[!h]
+	\centering
+\begin{tabular}{|l|l|l|} \hline
+	Name & LOC & $|P_R|$ \\ \hline
+	a2ps & 55k & 2012\\
+	gawk & 59k & 902\\ 
+	gnuchess & 38k & 1222\\ 
+	gnugo & 83k & 2801\\
+	grep & 35k & 820\\
+	gzip & 27k & 494\\
+	lapack/blas & 954k & 16422\\
+	make & 34k & 993\\ 
+	sed & 23k & 292\\
+	tar & 73k & 1712\\
+	\hline
+\end{tabular}
+\caption{List of analyzed open-source projects, with their respective number of
+lines of code, and their number of control points in $P_R$}
+\label{fig:projects}
+\end{figure}
+
+\subsection{Precision of the various techniques}
+
+\begin{figure}[h]
+  \begin{center}
+    \input{./techniques.tex}
+  \end{center} 
+  \vspace{-20px}
+  \caption{Comparison of the abstract values obtained by classical abstract
+  interpretation (S), \emph{Guided Static
+  Analysis} (G), \emph{Path-focused} technique (PF), our combined technique
+  (G+PF), and its version with disjunctive invariants (DIS).}
+  \label{fig:techniques}
+\end {figure}
+
+\JH{La comparaison DIS/G+PF n'est pas encore correcte (DIS pas compl\`etement
+impl\'ement\'e)\\} 
+Figures \ref{fig:techniques} shows the results of various comparisons between
+the different analysis techniques.
+
+\subsection{Precision of Abstract Domains}
+
 \section{Conclusion}

### 2011-10-25 papers/PLDI2012/implicitization.tex
tableau de résultats

@@
 
-\JH{La comparaison DIS/G+PF n'est pas encore correcte (DIS pas compl\`etement
-impl\'ement\'e)\\} 
 Figures \ref{fig:techniques} shows the results of various comparisons between
 the different analysis techniques.
+
+\begin{table*}
+\begin{center}
+\begin{tabular}{|l||c|c|c||c|c|c||c|c|c||c|c|c||c|c|c||} \hline
+\textbf{Benchmark} & \multicolumn{3}{c||}{\textbf{G/S}}
+& \multicolumn{3}{c||}{\textbf{PF/S}}
+& \multicolumn{3}{c||}{\textbf{PF/G}}
+& \multicolumn{3}{c||}{\textbf{G+PF/PF}}
+& \multicolumn{3}{c||}{\textbf{G+PF/G}} \\ \cline{2-16}
+& \textbf{lt} & \textbf{gt} & \textbf{un} 
+& \textbf{lt} & \textbf{gt} & \textbf{un} 
+& \textbf{lt} & \textbf{gt} & \textbf{un} 
+& \textbf{lt} & \textbf{gt} & \textbf{un} 
+& \textbf{lt} & \textbf{gt} & \textbf{un} \\
+ \hline
+ \input{table1.tex}
+\end{tabular}
+\end{center}
+\caption{Result of the comparison of the various techniques described in this
+paper: classic Abstract Interpretation (S), \emph{Guided Static Analysis} (G),
+\emph{Path-focused} technique (PF), and our combined technique (G+PF). For
+instance, \textbf{G/S} compares the benefits of \emph{Guided Static Analysis}
+over the classic Abstract interpretation algorithm. \textbf{lt} column gives the
+percentage of invariants more precise with the left-side technique,
+\textbf{gt} the percentage of invariants better with the right-side technique,
+and \textbf{un} gives the percentage of invariants that are uncomparable, i.e
+neither greater nor smaller.}
+\end{table*}
 

### 2011-10-25 papers/PLDI2012/implicitization.tex
experiments

@@
 	\centering
-\begin{tabular}{|l|l|l|} \hline
+\begin{tabular}{|l|r|r|} \hline
 	Name & LOC & $|P_R|$ \\ \hline
@@
 
-Figures \ref{fig:techniques} shows the results of various comparisons between
-the different analysis techniques.
-
 \begin{table*}
 \begin{center}
-\begin{tabular}{|l||c|c|c||c|c|c||c|c|c||c|c|c||c|c|c||} \hline
+\begin{tabular}{|l||r|r|r||r|r|r||r|r|r||r|r|r||r|r|r||} \hline
 \textbf{Benchmark} & \multicolumn{3}{c||}{\textbf{G/S}}
@@
 neither greater nor smaller.}
+\label{tab:techniques}
 \end{table*}
+
+Table \ref{tab:techniques} gives the results of the comparison between the
+different analysis techniques, applied to the benchmarks described in Figure
+\ref{fig:projects}. These results are also summarized by Figure \ref{fig:techniques}.
+
+\emph{Guided Static Analysis} from \citet{DBLP:conf/sas/GopanR07} improves the
+result of the classical Abstract Interpretation in $2.37\%$ of the control points
+in $P_R$.
+\emph{Path-focused} technique from \citet{Monniaux_Gonnord_SAS11} gives
+statistically better results, and finds better invariants than \emph{Guided
+Static Analysis} in $7.28\%$ of the cases. However, it also looses precision in an
+important number ($2.93\%$) of control points.
+Our combined technique gives the most promising results, since it is
+statistically more precise than the other techniques.
+
+The analysis using disjunctive invariants seems not to be as efficient as
+expected, and often finds greater invariants than the combined technique
+($11.64\%$). It probably means that our heuristic for the construction of
+$\sigma$ is not effective.
+
+While experiencing with techniques that use SMT-solving, we encountered some
+limitations due to non-linear arithmetic in the analyzed programs. Indeed, 
+the SMT-solver is not able to decide the satisfiability of some SMT-formulae
+expressing the semantics of non-linear programs. 
+In this case, we skipped the function for which the SMT-solver returned the
+``unknown'' result.
+This limitation occured very rarely in our experiments, except for the analysis
+of \emph{Lapack/Blas}, where 798 over the 1602 functions have been skipped.
+In such cases, such a technique should make the SMT-formula simpler to solve by using
+for instance \emph{linearization}. However, such transformations are not
+implemented in our analyzer.
 

### 2011-10-25 papers/PLDI2012/implicitization.tex
petites rectifs

@@
 \footnote{\url{http://research.microsoft.com/en-us/um/redmond/projects/z3/}}
-Yices,%
+and Yices,%
 \footnote{\url{http://yices.csl.sri.com/}}
@@
 
-\subsection{Path-focused technique}
+\subsection{Path-focusing}
 \label{sec:path_focusing}
@@
 Our technique gives us the expected invariant 
-$\lstinline|x_old| \in~[-10000,10000]$. 
+$\lstinline|x_old| \in~[-10000,\allowbreak 10000]$. 
 Here, only 3 paths over the 25 have been computed during the analysis. In
@@
 intervals, grids (convex polyhedra and linear congruences).
-
-For the techniques involving SMT-solving, our analyzer uses Yices
+For SMT-solving, our analyzer uses Yices
 \cite{DBLP:conf/cav/DutertreM06} or Microsoft Z3\cite{DBLP:conf/tacas/MouraB08}

### 2011-10-25 papers/PLDI2012/implicitization.tex
presentation du tableau

@@
 \begin{tabular}{|l|r|r|} \hline
-	Name & LOC & $|P_R|$ \\ \hline
-	a2ps & 55k & 2012\\
-	gawk & 59k & 902\\ 
-	gnuchess & 38k & 1222\\ 
-	gnugo & 83k & 2801\\
-	grep & 35k & 820\\
-	gzip & 27k & 494\\
-	lapack/blas & 954k & 16422\\
-	make & 34k & 993\\ 
-	sed & 23k & 292\\
-	tar & 73k & 1712\\
+	\multicolumn{1}{|c|}{Name} &
+        \multicolumn{1}{c|}{kLOC} &
+        \multicolumn{1}{c|}{$|P_R|$} \\ \hline
+	a2ps & 55 & 2012\\
+	gawk & 59 & 902\\ 
+	gnuchess & 38 & 1222\\ 
+	gnugo & 83 & 2801\\
+	grep & 35 & 820\\
+	gzip & 27 & 494\\
+	lapack/blas & 954 & 16422\\
+	make & 34 & 993\\ 
+	sed & 23 & 292\\
+	tar & 73 & 1712\\
 	\hline
@@
 \begin{center}
-\begin{tabular}{|l||r|r|r||r|r|r||r|r|r||r|r|r||r|r|r||} \hline
-\textbf{Benchmark} & \multicolumn{3}{c||}{\textbf{G/S}}
-& \multicolumn{3}{c||}{\textbf{PF/S}}
-& \multicolumn{3}{c||}{\textbf{PF/G}}
-& \multicolumn{3}{c||}{\textbf{G+PF/PF}}
-& \multicolumn{3}{c||}{\textbf{G+PF/G}} \\ \cline{2-16}
-& \textbf{lt} & \textbf{gt} & \textbf{un} 
-& \textbf{lt} & \textbf{gt} & \textbf{un} 
-& \textbf{lt} & \textbf{gt} & \textbf{un} 
-& \textbf{lt} & \textbf{gt} & \textbf{un} 
-& \textbf{lt} & \textbf{gt} & \textbf{un} \\
+\begin{tabular}{|l|rrr|rrr|rrr|rrr|rrr|} \hline
+\multicolumn{1}{|c|}{\textbf{Benchmark}}
+& \multicolumn{3}{c|}{\textbf{G/S}}
+& \multicolumn{3}{c|}{\textbf{PF/S}}
+& \multicolumn{3}{c|}{\textbf{PF/G}}
+& \multicolumn{3}{c|}{\textbf{G+PF/PF}}
+& \multicolumn{3}{c|}{\textbf{G+PF/G}} \\ %\cline{2-16}
+& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & inc.
+& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & inc.
+& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & inc.
+& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & inc.
+& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & inc. \\
  \hline
@@
 instance, \textbf{G/S} compares the benefits of \emph{Guided Static Analysis}
-over the classic Abstract interpretation algorithm. \textbf{lt} column gives the
-percentage of invariants more precise with the left-side technique,
-\textbf{gt} the percentage of invariants better with the right-side technique,
-and \textbf{un} gives the percentage of invariants that are uncomparable, i.e
+over the classic Abstract interpretation algorithm.
+The $\subseteq$ column gives the percentage of invariants more precise (smaller with respect to inclusion) with the left-side technique,
+$\subseteq$ the percentage of invariants better with the right-side technique,
+and ``inc.'' gives the percentage of invariants that are incomparable, i.e
 neither greater nor smaller.}
+% TODO: le reste c'est quand c'est egal?
 \label{tab:techniques}

### 2011-10-25 papers/PLDI2012/implicitization.tex
subseteq etc.

@@
 The $\subseteq$ column gives the percentage of invariants more precise (smaller with respect to inclusion) with the left-side technique,
-$\subseteq$ the percentage of invariants better with the right-side technique,
+$\supseteq$ the percentage of invariants better with the right-side technique,
 and ``inc.'' gives the percentage of invariants that are incomparable, i.e

### 2011-10-25 papers/PLDI2012/implicitization.tex
amelioration de la presentation du tableau

@@
 %\usepackage{color}
+
+\usepackage{siunitx}
 
@@
 \begin{center}
-\begin{tabular}{|l|rrr|rrr|rrr|rrr|rrr|} \hline
+\setlength{\tabcolsep}{1ex}
+\begin{tabular}{|l
+|S[tabnumalign=right,tabformat=2.2]S[tabnumalign=right,tabformat=1.2]S[tabnumalign=right,tabformat=1.2]%
+|S[tabnumalign=right,tabformat=2.2]S[tabnumalign=right,tabformat=1.2]S[tabnumalign=right,tabformat=1.2]%
+|S[tabnumalign=right,tabformat=2.2]S[tabnumalign=right,tabformat=1.2]S[tabnumalign=right,tabformat=1.2]%
+|S[tabnumalign=right,tabformat=2.2]S[tabnumalign=right,tabformat=1.2]S[tabnumalign=right,tabformat=1.2]%
+|S[tabnumalign=right,tabformat=2.2]S[tabnumalign=right,tabformat=1.2]S[tabnumalign=right,tabformat=1.2]|} \hline
 \multicolumn{1}{|c|}{\textbf{Benchmark}}
@@
 & \multicolumn{3}{c|}{\textbf{G+PF/G}} \\ %\cline{2-16}
-& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & inc.
-& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & inc.
-& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & inc.
-& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & inc.
-& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & inc. \\
+& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{inc.}
+& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{inc.}
+& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{inc.}
+& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{inc.}
+& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{inc.} \\
  \hline
- \input{table1.tex}
+ \input{table1}
 \end{tabular}
@@
 paper: classic Abstract Interpretation (S), \emph{Guided Static Analysis} (G),
-\emph{Path-focused} technique (PF), and our combined technique (G+PF). For
+\emph{Path-focusing} (PF), and our combined technique (G+PF). For
 instance, \textbf{G/S} compares the benefits of \emph{Guided Static Analysis}

### 2011-10-25 papers/PLDI2012/implicitization.tex
fix subsetneq

@@
 over the classic Abstract interpretation algorithm.
-The $\subseteq$ column gives the percentage of invariants more precise (smaller with respect to inclusion) with the left-side technique,
-$\supseteq$ the percentage of invariants better with the right-side technique,
+The $\subsetneq$ column gives the percentage of invariants more precise (smaller with respect to inclusion) with the left-side technique,
+$\supsetneq$ the percentage of invariants better with the right-side technique,
 and ``inc.'' gives the percentage of invariants that are incomparable, i.e

### 2011-10-25 papers/PLDI2012/implicitization.tex
uncomparable is correct

--- 2011-10-25 papers/PLDI2012/implicitization.tex
@@
 can easily be expressed as the
-cunjunction $b_{i_1}^s \wedge \bigwedge_{2 \leq k < n} b_{i_k} \wedge b_{i_n}^d$.
+conjunction $b_{i_1}^s \wedge \bigwedge_{2 \leq k < n} b_{i_k} \wedge b_{i_n}^d$.
 The Boolean $b_{i}^s$ is $true$ when the path starts at point $p_i$, whereas
@@
 Since we do not work on the entire transition graph but on a subset of it, we
-cunjunct the formula $f(p_i)$ with the actual set of working paths,
+conjoin the formula $f(p_i)$ with the actual set of working paths,
 noted $P$, expressed as a Boolean formula, where the Boolean variables are the
@@
 
-We distinguish two phases when analysing a subgraph $P$:
+We distinguish two phases when analyzing a subgraph $P$:
 \begin{enumerate}
@@
 		image is not in $-10 \leq \lstinline|x_old| \leq 10$.
-		Additionnally, the path that goes from $p_5$ to
+		Additionally, the path that goes from $p_5$ to
 		$p_11$, that goes through the \emph{else} branch of each 
@@
 & \multicolumn{3}{c|}{\textbf{G+PF/G}} \\ %\cline{2-16}
-& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{inc.}
-& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{inc.}
-& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{inc.}
-& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{inc.}
-& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{inc.} \\
+& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
+& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
+& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
+& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
+& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.} \\
  \hline
@@
 over the classic Abstract interpretation algorithm.
-The $\subseteq$ column gives the percentage of invariants more precise (smaller with respect to inclusion) with the left-side technique,
-$\supseteq$ the percentage of invariants better with the right-side technique,
-and ``inc.'' gives the percentage of invariants that are incomparable, i.e
+The $\subsetneq$ column gives the percentage of invariants more precise (smaller with respect to inclusion) with the left-side technique,
+$\supsetneq$ the percentage of invariants better with the right-side technique,
+and ``unc.'' gives the percentage of invariants that are uncomparable, i.e
 neither greater nor smaller.}
@@
 ``unknown'' result.
-This limitation occured very rarely in our experiments, except for the analysis
+This limitation occurred very rarely in our experiments, except for the analysis
 of \emph{Lapack/Blas}, where 798 over the 1602 functions have been skipped.

### 2011-10-26 papers/PLDI2012/implicitization.tex
ajout d'une option au package siunitx pour que ca compile avec la version 2 du package
on force siunitx a la version 1 avec une option speciale
siunitx
Utilisons dcolumn au lieu de siunitx!

@@
 
-\usepackage{siunitx}
+\usepackage{dcolumn}
 
@@
 \begin{tabular}{|l
-|S[tabnumalign=right,tabformat=2.2]S[tabnumalign=right,tabformat=1.2]S[tabnumalign=right,tabformat=1.2]%
-|S[tabnumalign=right,tabformat=2.2]S[tabnumalign=right,tabformat=1.2]S[tabnumalign=right,tabformat=1.2]%
-|S[tabnumalign=right,tabformat=2.2]S[tabnumalign=right,tabformat=1.2]S[tabnumalign=right,tabformat=1.2]%
-|S[tabnumalign=right,tabformat=2.2]S[tabnumalign=right,tabformat=1.2]S[tabnumalign=right,tabformat=1.2]%
-|S[tabnumalign=right,tabformat=2.2]S[tabnumalign=right,tabformat=1.2]S[tabnumalign=right,tabformat=1.2]|} \hline
+|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
+|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
+|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
+|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
+|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}|} \hline
 \multicolumn{1}{|c|}{\textbf{Benchmark}}

### 2011-10-26 papers/PLDI2012/implicitization.tex
temps d'exécution
reponse a un commentaire
ordre des mots + lisible

@@
 
-In such an analysis, information may flow forward (one computes the polyhedron after a program statement as the image, or a super-set thereof, of the polyhedron before by the semantics of the statement) or backward;
+In such an analysis, information may flow forward (one computes the polyhedron after a program statement as the image by the semantics of the statement, or a super-set thereof, of the polyhedron before) or backward;
 forward program analysis computes super-sets of the states reachable from the initialization of the program, backward program analysis computes super-sets of the states co-reachable from some property of interest (for instance, the violation of an assertion).

### 2011-10-26 papers/PLDI2012/implicitization.tex
phrase plus lisible

@@
 forward program analysis computes super-sets of the states reachable from the initialization of the program, backward program analysis computes super-sets of the states co-reachable from some property of interest (for instance, the violation of an assertion).
-In forward analysis, control-flow joins correspond to computation of convex hulls if using convex polyhedra (more generally, they correspond to least upper bounds in a lattice); in backward analysis, it is control-flow splits.
+In forward analysis, control-flow joins correspond to convex hulls if using convex polyhedra (more generally, they correspond to least upper bounds in a lattice); in backward analysis, it is control-flow splits that correspond to convex hulls.
 

### 2011-10-26 papers/PLDI2012/implicitization.tex
fix bst

@@
 Fortran.
-\begin{figure}[!h]
+\begin{table}[!h]
 	\centering
@@
 \label{fig:projects}
-\end{figure}
+\end{table}
 
@@
 and ``unc.'' gives the percentage of invariants that are uncomparable, i.e
-neither greater nor smaller.}
-% TODO: le reste c'est quand c'est egal?
+neither greater nor smaller;
+the code points where both invariants are equal make up the remaining percentage.}
 \label{tab:techniques}
@@
 
+\begin{table}[!h]
+	\centering
+\begin{tabular}{|l|r|r|r|r|r|} \hline
+	Benchmark & \textbf{S} & \textbf{G} & \textbf{PF} & \textbf{G+PF} &
+	\textbf{DIS} \\ \hline
+	\input{table_time.tex} \hline
+\end{tabular}
+\caption{Execution time for each technique, expressed in seconds}
+\label{tab:time}
+\end{table}
+
 \subsection{Precision of Abstract Domains}

### 2011-10-26 papers/PLDI2012/implicitization.tex
rephrase

@@
 
-\citet{Monniaux_Gonnord_SAS11}'s \emph{path-focused} technique
+\citet{Monniaux_Gonnord_SAS11}'s \emph{path-focusing} technique
 distinguishes the different paths in the program in order to avoid loss of
@@
 previous set $P$. Each starting points of paths in $P'$ is added into $A$.
-We apply the Path-focused technique\cite{Monniaux_Gonnord_SAS11} to
+We apply the path-focusing \cite{Monniaux_Gonnord_SAS11} to
 the elements of $A$ until $A$ is empty. 
@@
 
-The algorithm for computing an inductive invariant over a subgraph is the
-Path-focused algorithm from \citet{Monniaux_Gonnord_SAS11} with special
+For computing an inductive invariant over a subgraph, we use the
+Path-focusing algorithm from \citet{Monniaux_Gonnord_SAS11} with special
 treatment for self loops (line~\ref{alg=pf} in algorithm~\ref{algo:combined}).
@@
 	This is the phase explained in \ref{subsec:ascending}.
-\item  When the previous phase terminates, there is no more paths in
+\item  When the previous phase terminates, there are no more paths in
 $P$, starting at $p_i$, 
 that make the abstract values of the successors grow (otherwise, the SMT
-solver would not have answered ``\emph{unsat}''). Narrowing iterations preserves
+solver would not have answered ``\emph{unsat}''). Narrowing iterations preserve
 this property. 
@@
 we now simply check for the satisfiability of $f(p_i)$ instead of $f(p_i) \wedge
-P$. Since we already know that $f(p_i) \wedge P$ is \emph{unsatisfiable}, all
+P$. Since we already know that $f(p_i) \wedge P$ is unsatisfiable, all
 the paths given by the SMT solver won't be in $P$.
@@
 \item 
-the subset of paths $P$ is
-strictly increasing at each loop iteration, and bounded by the finite set of
+the subset of paths $P$ strictly increases at each loop iteration, and bounded by the finite set of
 paths in the entire graph. 
@@
 We revise the rate limiter described in \ref{subsec:rate_lim}. In this example,
-\emph{Path-focused} technique works well because all the paths starting at the
+\emph{Path-focusing} works well because all the paths starting at the
 loop header are actually self loops. In such a case, the technique performs a
 widening/narrowing sequence or accelerates the loop, thus leading to a precise
-invariant. However, in real cases, there also exist paths that are not
-self loops, where \emph{Path-focusing} only performs a widening. 
+invariant. However, in some cases, there also exist paths that are not
+self loops, in which case \emph{Path-focusing} applies widening. 
 This widening may induce unrecoverable loss of precision.
@@
 $\lstinline|x_old| \in~]-\infty,+\infty[$.
-Path-focused technique from \citet{Monniaux_Gonnord_SAS11} performs slightly
+Path-focusing \citep{Monniaux_Gonnord_SAS11} performs slightly
 better, and finds $\lstinline|x_old| \in~]-\infty,10000]$.
@@
 
-This technique is aimed at computing for a loop header $p_i$ a disjunctive invariant
+For a loop header $p_i$, this technique computes a disjunctive invariant
 $\bigvee_{1\leq j \leq m} X_{i,j}$. 
@@
 compute them only when needed.
-\item At each loop iteration of the algorithm explained in
-\cite{DBLP:conf/pldi/GulwaniZ10}, an image by a path and a union of abstract
-values are computed for each path and each disjunct. 
-Lots of these operations may be useless, in the sense that the resulting
-abstract value may not change.
-In our approach, we do this only for paths and disjuncts that makes the
-resulting abstract value grow.
+
+\item At each loop iteration of the original algorithm \citep{DBLP:conf/pldi/GulwaniZ10}, an image by each path inside the loop is computed for each disjunct of the invariant candidate.
+Yet, many of these images may be redundant: for instance, if our invariant candidate is $(0 \leq x \leq 10 \land 0 \leq y \leq 1000) \lor (x < -10 \land y < -10)$, then there is no point enumerating paths whose image is included in this invariant candidate.
+In our approach, we compute such an image only if it makes the resulting abstract value grow.
 \end{itemize}

### 2011-10-26 papers/PLDI2012/implicitization.tex
temps execution

--- 2011-10-26 papers/PLDI2012/implicitization.tex
@@
 In this section, our experimental results come from the analysis of various
-open-source projects listed in Figure \ref{fig:projects}, written in C, C++ and
+open-source projects listed in Table \ref{fig:projects}, written in C, C++ and
 Fortran.
-\begin{figure}[!h]
+\begin{table}[!h]
 	\centering
@@
 neither greater nor smaller.}
-% TODO: le reste c'est quand c'est egal?
+% TODO: le reste c'est quand c'est egal? => OUI
 \label{tab:techniques}

### 2011-10-26 papers/PLDI2012/implicitization.tex

--- 2011-10-26 papers/PLDI2012/implicitization.tex
@@
 In this section, our experimental results come from the analysis of various
-open-source projects listed in Figure \ref{fig:projects}, written in C, C++ and
+open-source projects listed in Table \ref{fig:projects}, written in C, C++ and
 Fortran.

### 2011-10-26 papers/PLDI2012/implicitization.tex
explications

@@
 We conducted extensive experiments on real-life programs in order to compare the
-different techniques. We also compared the precision of the analysis depending
-on the choice of the abstract domain.
+different techniques, mostly on open-source projects (Fig.~\ref{fig:projects}) written in C, C++ and Fortran.
 
-In order to compare the various techniques, we compare by inclusion the obtained
-abstract values at each control points in $P_R$.
-
-In this section, our experimental results come from the analysis of various
-open-source projects listed in Table \ref{fig:projects}, written in C, C++ and
-Fortran.
 \begin{table}[!h]
@@
 
-Table \ref{tab:techniques} gives the results of the comparison between the
-different analysis techniques, applied to the benchmarks described in Figure
-\ref{fig:projects}. These results are also summarized by Figure \ref{fig:techniques}.
+We compared the precision of the different techniques described in this article as follows:
+\begin{itemize}
+\item For each program, we distinguished a set $P_R = P_W$ of suitable widening points by a simple algorithm: for each procedure, we compute the strongly connected components of its control-flow graph using Tarjan's algorithm; the targets of the back-edges of the depth-first search are added to~$P_R$.
+Note that the resulting set is not necessarily minimal, but is sufficient to disconnect all cycles.
+(More sophisticated techniques are discussed in e.g. \citet{BourdonclePhd}).
 
+\item For each program and each pair $(T_1,T_2)$ of analysis techniques, we list the proportion of control points in $P_R$ where $T_1$ (resp.~$T_2$) gives a strictly stronger invariant, denoted by $\subsetneq$ (resp.
+$\supsetneq$), and the proportion of control points where the invariants given by $T_1$ and $T_2$ are uncomparable for the inclusion ordering (the remainder of the control points are thus those for which both techniques give the same invariant).
+\end{itemize}
+
+Our analysis tool currently only operates over scalar variable and cannot directly cope with arrays or memory accessed through pointers.
+We therefore run it after the ``memory to registers'' (\texttt{mem2reg}) optimization phase in LLVM, which lifts most memory accesses to scalar variables.
+The remaining memory accesses are treated as undefined values, as are the return values of function calls. For better precision, we also apply function inlining to the program.
+
+The analyses are performed using convex polyhedra as abstract domain.
+Variables defined by linear expressions over other variables (e.g.
+$y=x+1$) are not retained as dimensions in the polyhedra; therefore the only variables appearing in the polyhedra are those defined by $\phi$ operators (join points) in the SSA representation, and those introduced by constructs we cannot model as linear operations, such as loads from memory, return values from function calls, or bitwise arithmetic.
+
+% TODO: avec quel domaine abstrait?
+% TODO: discuter du big step SSA?
+% TODO: cite SSA
+
+Let us briefly comment the results given in more details in Table~\ref{tab:techniques} and Figure~\ref{fig:techniques}.
 \emph{Guided Static Analysis} from \citet{DBLP:conf/sas/GopanR07} improves the
@@
 in $P_R$.
-\emph{Path-focused} technique from \citet{Monniaux_Gonnord_SAS11} gives
+\emph{Path-focusing} from \citet{Monniaux_Gonnord_SAS11} gives
 statistically better results, and finds better invariants than \emph{Guided
-Static Analysis} in $7.28\%$ of the cases. However, it also looses precision in an
+Static Analysis} in $7.28\%$ of the cases. However, it also loses precision in an
 important number ($2.93\%$) of control points.
-Our combined technique gives the most promising results, since it is
+Finally, our combined technique gives the most promising results, since it is
 statistically more precise than the other techniques.
@@
 expressing the semantics of non-linear programs. 
-In this case, we skipped the function for which the SMT-solver returned the
+In this case, we skipped the functions for which the SMT-solver returned the
 ``unknown'' result.
@@
 of \emph{Lapack/Blas}, where 798 over the 1602 functions have been skipped.
-In such cases, such a technique should make the SMT-formula simpler to solve by using
-for instance \emph{linearization}. However, such transformations are not
-implemented in our analyzer.
+\emph{Lapack/Blas} implements matrix computations, which use floating-point multiplications.
+
+In cases where the formula is expressed in too rich a logic for the SMT-solver to deal with, a number of workarounds are possible:
+\begin{itemize}
+\item \emph{Linearization}, as per \citet{DBLP:conf/vmcai/Mine06}, which overapproximates nonlinear semantics by linear semantics.
+\item Replacing the results of nonlinear operations by ``unknown''.
+\end{itemize}
+However, such transformations are not yet implemented in our tool.
+% TODO: c'est si dur de remplacer * non lineaire par undef?
 

### 2011-10-26 papers/PLDI2012/implicitization.tex
experiments

--- 2011-10-26 papers/PLDI2012/implicitization.tex
@@
 
+For the computation of disjunctive invariants, we experimented with a heuristic
+of dynamic construction of the $\sigma$ function, adapted from
+\cite{DBLP:conf/pldi/GulwaniZ10}.  We start with one sigle
+disjunct per control point ($m=1$), and we start with an empty map $\sigma$. We also
+define a maximal bound on the number of disjunct per control point, noted $M$.
+$\sigma(j,k)$ is then constructed on the fly only when needed.
+When $\sigma(j,k)$ needs to be computed, we first compute the image of the
+abstract value $X_k$ by the path indexed by $j$, and try to find an existing
+disjunct of index $k'$ so that the convex hull of the two abstract values is
+exactly the union of them. If such an index exists, then $\sigma(j,k) = k'$.
+Otherwise:
+\begin{itemize}
+	\item if $m < M$, we increase $m$ by $1$ and define $\sigma(j,k) = m$
+	\item if $m = M$, we define $\sigma(j,k) = M$
+\end{itemize}
+
 \subsection{Precision of the various techniques}
@@
 ($11.64\%$). It probably means that our heuristic for the construction of
-$\sigma$ is not effective.
+$\sigma$ is not effective enough. Usually, disjunctive invariants computation
+finds precise loop invariants because they allow to distinguish for
+instance the first loop iterations. In our case, such benefits are not
+visible, since we have already unrolled every loops once.
 
@@
 implemented in our analyzer.
+
+Table \ref{tab:time} gives the execution time of the different analysis
+techniques. It is interesting to see that \emph{Guided Static Analysis} and
+\emph{Path-focusing} technique are sometimes faster than the classical
+algorithm. Our techniques are slower, but on the same order of magnitude than
+the previous techniques.
 

### 2011-10-26 papers/PLDI2012/implicitization.tex

--- 2011-10-26 papers/PLDI2012/implicitization.tex
@@
 % TODO: c'est si dur de remplacer * non lineaire par undef?
+
+Table \ref{tab:time} gives the execution time of the different analysis
+techniques. It is interesting to see that \emph{Guided Static Analysis} and
+\emph{Path-focusing} technique are sometimes faster than the classical
+algorithm. Our techniques are slower, but on the same order of magnitude than
+the previous techniques.
 

### 2011-10-26 papers/PLDI2012/implicitization.tex
typo
blah SSA

@@
 \end{table}
+
+\subsection{Analysis algorithm}
+Our tool operates over LLVM bitcode \citep{LLVM_langref,Lattner:2004:LCF:977395.977673}.
+LLVM bitcode is in \emph{single static assignment} (SSA) form: a given scalar variable is given a value at a single syntactic point in the program.
+In concrete terms, an assignment \lstinline|x=2*x+1;| gets translated into a definition $x_2 = 2x_1+1$, with distinct variables $x_1$ and $x_2$ corresponding to the same original variable \lstinline|x| at different points in the program.
+LLVM makes it easy to follow definition-use and use-definition chains: for a given variable (say, $x_2$) one can immediately obtain its definition (say, $2x_1+1$).
+One may see conversion to SSA form as a static precomputation of some of the symbolic propagations proposed by \citet{DBLP:conf/vmcai/Mine06} to enhance the precision of analyses.
+
+SSA introduces $\phi$-functions at the head of a control code to define variables whose value depends on which incoming edge to this control node was last taken.
+For instance, for \lstinline|if (...) { x = 2*x+1; } else { x= 0; }|, then $x_2$ is defined as $\phi(2x_1+1,0)$.
+In this framework, each variable is uniquely defined as an arithmetic ($+$, $-$, $\times$, $/$) function of other variables that themselves cannot be represented as affine linear functions, because they are defined using $\phi$-functions, bitwise arithmetic, loads from memory, or return values from function calls.
+
+This motivates a key implement decision of our tool: only those variables $v_1,\dots,v_n$ that are not defined by arithmetic operations are retained as coordinates in the abstract domain (e.g.
+as dimensions in polyhedra).
+A basic block of code therefore amounts to a \emph{parallel assignment} operation
+$(v_1,\dots,v_n) \allowbreak\mapsto\allowbreak
+(f_1(v_1,\dots,v_n), \allowbreak, \dots, \allowbreak
+ f_n(v_1,\dots,v_n))$;
+such operations are directly supported by APRON.
+A first benefit is to limit the number of dimensions in the abstract values, since polyhedra libraries typically perform worse with higher dimensions. TODO
+Note that this means that the abstract operation for a single path in path-focusing methods also is a (large) parallel assignment.
+
+Our analysis tool currently only operates over scalar variables from the SSA representation and thus cannot directly cope with arrays or memory accessed through pointers.
+We therefore run it after the ``memory to registers'' (\texttt{mem2reg}) optimization phase in LLVM, which lifts most memory accesses to scalar variables.
+The remaining memory accesses are treated as undefined values, as are the return values of function calls. For better precision, we also apply function inlining to the program.
 
@@
 
-Our analysis tool currently only operates over scalar variable and cannot directly cope with arrays or memory accessed through pointers.
-We therefore run it after the ``memory to registers'' (\texttt{mem2reg}) optimization phase in LLVM, which lifts most memory accesses to scalar variables.
-The remaining memory accesses are treated as undefined values, as are the return values of function calls. For better precision, we also apply function inlining to the program.
-
-The analyses are performed using convex polyhedra as abstract domain.
-Variables defined by linear expressions over other variables (e.g.
-$y=x+1$) are not retained as dimensions in the polyhedra; therefore the only variables appearing in the polyhedra are those defined by $\phi$ operators (join points) in the SSA representation, and those introduced by constructs we cannot model as linear operations, such as loads from memory, return values from function calls, or bitwise arithmetic.
-
 % TODO: avec quel domaine abstrait?

### 2011-10-26 papers/PLDI2012/implicitization.tex
petites modifs

--- 2011-10-26 papers/PLDI2012/implicitization.tex
@@
 We therefore run it after the ``memory to registers'' (\texttt{mem2reg}) optimization phase in LLVM, which lifts most memory accesses to scalar variables.
-The remaining memory accesses are treated as undefined values, as are the return values of function calls. For better precision, we also apply function inlining to the program.
+The remaining memory accesses are treated as undefined values, as are the return
+values of function calls. For better precision, we apply function inlining
+to the program, and we also unroll the loops once.
 
@@
 \emph{Path-focusing} technique are sometimes faster than the classical
-algorithm. Our techniques are slower, but on the same order of magnitude than
+algorithm. Our techniques are slower, but on the same order of magnitude as
 the previous techniques.

### 2011-10-26 papers/PLDI2012/implicitization.tex
on utilise des polyèdres

--- 2011-10-26 papers/PLDI2012/implicitization.tex
@@
 We therefore run it after the ``memory to registers'' (\texttt{mem2reg}) optimization phase in LLVM, which lifts most memory accesses to scalar variables.
-The remaining memory accesses are treated as undefined values, as are the return values of function calls. For better precision, we also apply function inlining to the program.
+The remaining memory accesses are treated as undefined values, as are the return
+values of function calls. For better precision, we also apply function inlining
+to the program, and we unroll every loop once.
 
@@
 
-\item For each program and each pair $(T_1,T_2)$ of analysis techniques, we list the proportion of control points in $P_R$ where $T_1$ (resp.~$T_2$) gives a strictly stronger invariant, denoted by $\subsetneq$ (resp.
-$\supsetneq$), and the proportion of control points where the invariants given by $T_1$ and $T_2$ are uncomparable for the inclusion ordering (the remainder of the control points are thus those for which both techniques give the same invariant).
+\item For each program and each pair $(T_1,T_2)$ of analysis techniques, we list
+	the proportion of control points in $P_R$ where $T_1$ (resp.~$T_2$) gives a
+	strictly stronger invariant, denoted by $\subsetneq$ (resp. $\supsetneq$),
+	and the proportion of control points where the invariants given by $T_1$ and
+	$T_2$ are uncomparable for the inclusion ordering (the remainder of the
+	control points are thus those for which both techniques give the same
+	invariant). We use convex polyhedra as the abstract domain.
 \end{itemize}
 
-% TODO: avec quel domaine abstrait?
 % TODO: discuter du big step SSA?

### 2011-10-26 papers/PLDI2012/implicitization.tex
and -> virgule
Anglais: actual -> current, et inside of -> inside
"inside of" existe aussi si inside est un nom (the inside of blabla),
mais sauf erreur de ma part, il était de trop là où je l'ai enlevé.

@@
 
-Most SMT solvers, including Z3%
+Most SMT solvers, including Z3,%
 \footnote{\url{http://research.microsoft.com/en-us/um/redmond/projects/z3/}}
-and Yices,%
+Yices,%
 \footnote{\url{http://yices.csl.sri.com/}}
@@
 The starting subset are the transitions reachable in one step from the loop initialization.
-If for a given subset $S$ in the sequence, no transitions outside of $S$ are reachable from the inductive invariant attached to $S$, then iterations stop;
+If for a given subset $S$ in the sequence, no transitions outside $S$ are reachable from the inductive invariant attached to $S$, then iterations stop;
 otherwise, add these transitions to $S$ and iterate more.
@@
 there exist a path starting at a point $p_i \in P_R$ in a state included in the
-actual invariant candidate $X_i$, and arriving at a point $p_j \in P_R$ in a
-state outside of $X_j$. In this case, we construct this path using the model and
+current invariant candidate $X_i$, and arriving at a point $p_j \in P_R$ in a
+state outside $X_j$. In this case, we construct this path using the model and
 update $X_j$. When $p_i = p_j$, meaning that the path is actually a self-loop,
@@
 \in P_R$ of $p_i$, such that the image of $X_{i}^s$ by the path transformation
-is not included in the actual $X_{j}^s$.
+is not included in the current $X_{j}^s$.
 Intuitively, such a path makes the abstract value $X_{j}^s$ grow, and thus is
@@
 		new subgraph. Again, there is no new path that goes to $p_11$ with an
-		image outside of $-10 \leq \lstinline|x_old| \leq 10$.
+		image outside $-10 \leq \lstinline|x_old| \leq 10$.
 	\item

### 2011-10-27 papers/PLDI2012/implicitization.tex
Déplacement du résumé de l'algo de « Adding new paths » vers l'intro de 3.
Au passage, l'ordre des étapes n'était plus bonne par rapport au nouvel algo.

@@
 
-\begin{algorithm}
-  \begin{algorithmic}
-    \STATE \textit{Subset} $\gets \emptyset$
-    \WHILE{true}
-    \STATE Add new paths to \textit{Subset} (lines~\ref{alg=start-add-paths}-\ref{alg=end-add-paths} in algorithm~\ref{algo:combined})
-    \STATE Ascending iteration on \textit{Subset} (lines~\ref{alg=start-ascending}-\ref{alg=end-ascending} in algorithm~\ref{algo:combined})
-    \STATE Narrow (line~\ref{alg=narrowing} in algorithm~\ref{algo:combined})
-    \ENDWHILE
-  \end{algorithmic}
-\end{algorithm}
+We distinguish three phases in the main loop of the analysis:
+\begin{enumerate}
+\item Whe start finding a relevant subset $P \cup P'$ of the graph.
+  Either the previous iteration or the initialization lead us to a
+  state where there are no more paths in the previous subset $P$,
+  starting at $p_i$, that make the abstract values of the successors
+  grow (otherwise, the SMT solver would not have answered
+  ``\emph{unsat}''). Narrowing iterations preserve this property.
+  However, there may exist such paths in the entire multigraph, that
+  are not in $P$. This phase computes these paths and adds them to
+  $P'$ (line~\ref{alg=computeNewPaths} in
+  algorithm~\ref{algo:combined}). We also use SMT solving to discover
+  these new paths, but we slightly change the SMT formula given to the
+  SMT solver: we now simply check for the satisfiability of $f(p_i)$
+  instead of $f(p_i) \wedge P$. Since we already know that $f(p_i)
+  \wedge P$ is unsatisfiable, all the paths given by the SMT solver
+  won't be in $P$. This phase is described in~\ref{subsec:addingpaths}
+  and corresponds to lines in~\ref{alg=start-add-paths} to
+  \ref{alg=end-add-paths} algorithm~\ref{algo:combined}.
+\item Given a new subset $P$, we search for paths starting at point
+  $p_i \in P_R$, such that these paths are in $P$, i.e are included in
+  the working subgraph. Each time we find a path, we update the
+  abstract value of the destination point of the path. This is the
+  phase explained in~\ref{subsec:ascending}, and corresponds to
+  lines~\ref{alg=start-ascending} to~\ref{alg=end-ascending} in
+  algorithm~\ref{algo:combined}.
+\item We perform narrowing iterations the usual way
+  (line~\ref{alg=narrowing} in algorithm~\ref{algo:combined}) and
+  reiterate from step 1 unless there are no more states to explore.
+\end{enumerate}
 
@@
 
-Since we do not work on the entire transition graph but on a subset of it, we
+The difference with~\citet{Monniaux_Gonnord_SAS11} is that we do not
+work on the entire transition graph but on a subset of it. Therefore we
 conjoin the formula $f(p_i)$ with the actual set of working paths,
@@
 
-We distinguish two phases when analyzing a subgraph $P$:
-\begin{enumerate}
-\item We search for paths starting at point $p_i \in P_R$, such that these paths
-	are in $P$, i.e are included in the working subgraph. Each time we find a
-	path, we update the abstract value of the destination point of the path.
-	This is the phase explained in \ref{subsec:ascending}.
-\item  When the previous phase terminates, there are no more paths in
-$P$, starting at $p_i$, 
-that make the abstract values of the successors grow (otherwise, the SMT
-solver would not have answered ``\emph{unsat}''). Narrowing iterations preserve
-this property. 
-However, there may exist such paths in the entire multigraph, that are not in
-$P$. The second phase
-computes these paths and adds them to $P'$ (line~\ref{alg=computeNewPaths} in
-algorithm~\ref{algo:combined}). We also use SMT solving to discover
-these new paths, but we slightly change the SMT formula given to the SMT solver:
-we now simply check for the satisfiability of $f(p_i)$ instead of $f(p_i) \wedge
-P$. Since we already know that $f(p_i) \wedge P$ is unsatisfiable, all
-the paths given by the SMT solver won't be in $P$.
-\end{enumerate}
+\MM{Cette partie est devenue toute petite, il y avait un résumé global
+  de l'algo ici. Ça me parait suffisant comme ça}
 

### 2011-10-27 papers/PLDI2012/implicitization.tex
comparaison des domaines abstraits

--- 2011-10-26 papers/PLDI2012/implicitization.tex
@@
 
+\begin{table*}
+\begin{center}
+\setlength{\tabcolsep}{1ex}
+\begin{tabular}{|l
+|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
+|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
+|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
+|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
+|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}|} \hline
+\multicolumn{1}{|c|}{\textbf{Benchmark}}
+& \multicolumn{3}{c|}{\textbf{PK/OCT}}
+& \multicolumn{3}{c|}{\textbf{PK/BOX}}
+& \multicolumn{3}{c|}{\textbf{OCT/BOX}}
+& \multicolumn{3}{c|}{\textbf{PK/PKEQ}}
+& \multicolumn{3}{c|}{\textbf{OCT/PKEQ}} \\ %\cline{2-16}
+& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
+& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
+& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
+& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
+& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.} \\
+ \hline
+ \input{table_domain.tex}
+\end{tabular}
+\end{center}
+\caption{Results of the comparison of the various abstract domains, when using
+the same technique (G+PF).}
+\label{tab:time}
+\end{table*}
+
 \section{Conclusion}

### 2011-10-27 papers/PLDI2012/implicitization.tex
Fusion de l'intro de la section 3. et de la section Algorithm

--- 2011-10-27 papers/PLDI2012/implicitization.tex
@@
 \newcommand{\DM}[1]{{\color{violet} TODO(DM): #1}}
+% Pour les commentaires
+\usepackage[utf8]{inputenc}
 
@@
 
+\subsection{Algorithm}
+
+The algorithm for Guided static analysis on the implicit multigraph is described
+in Algorithm \ref{algo:combined}.
+
+\MM{Il faudra peut-être une section ``détails d'implem'' où mettre
+  l'histoire des BDDs pour P. Je laisse ici faute de mieux (peut-être
+  que la section~\ref{sec:analysis-algorithm} ferait l'affaire ?),
+  mais ça distrait un peu le lecteur}
+
+The current working subset of paths, noted $P$ and initially empty, is
+stored using a compact representation, such as binary decision
+diagrams. We also maintain two sets of control points:
+\begin{itemize}
+	\item $A'$ contains the points in $P_R$ that may be the starting points of new
+		feasible paths.
+	\item $A$ contains the points in $P_R$ on which we apply the ascending iterations.
+	Each time the abstract value of a control point $p$ is updated, $p$ is
+	inserted in both $A$ and $A'$.
+\end{itemize}
+
+\begin{algorithm}
+	\caption{Guided static analysis on implicit multigraph}
+	\label{algo:combined}
+	\begin{algorithmic}[1] 
+	\input{algo/algo-combined.tex}
+	\end{algorithmic}
+\end{algorithm}
+
 We distinguish three phases in the main loop of the analysis:
 \begin{enumerate}
-\item Whe start finding a relevant subset $P \cup P'$ of the graph.
+\item \label{step:addingpaths} We start finding a new relevant subset
+  $P \cup P'$ of the graph.
   Either the previous iteration or the initialization lead us to a
@@
   \ref{alg=end-add-paths} algorithm~\ref{algo:combined}.
-\item Given a new subset $P$, we search for paths starting at point
+\item \label{step:ascending} Given a new subset $P$, we search for paths starting at point
   $p_i \in P_R$, such that these paths are in $P$, i.e are included in
@@
   algorithm~\ref{algo:combined}.
-\item We perform narrowing iterations the usual way
+\item  \label{step:narrowing} We perform narrowing iterations the usual way
   (line~\ref{alg=narrowing} in algorithm~\ref{algo:combined}) and
-  reiterate from step 1 unless there are no more states to explore.
+  reiterate from step 1 unless there are no more points to explore,
+  i.e. $A' = \emptyset$.
 \end{enumerate}
@@
 
-\subsection{Algorithm}
-
-The algorithm for Guided static analysis on the implicit multigraph is described
-in Algorithm \ref{algo:combined}.
-
-The current working subset of paths, noted $P$, is stored using a
-compact representation, such as binary decision diagrams.
-We also maintain two sets of control points:
-\begin{itemize}
-	\item $A'$ contains the points in $P_R$ that may be the starting points of new
-		feasible paths.
-	\item $A$ contains the points in $P_R$ on which we apply the ascending iterations.
-	Each time the abstract value of a control point $p$ is updated, $p$ is
-	inserted in both $A$ and $A'$.
-\end{itemize}
-
-We first check which paths become feasible, so that we can include them into the
-next subset of paths. The set $P'$ of new feasible paths is added to the
-previous set $P$. Each starting points of paths in $P'$ is added into $A$.
-We apply the path-focusing \cite{Monniaux_Gonnord_SAS11} to
+Step~\ref{step:ascending} corresponds to the application of
+path-focusing \cite{Monniaux_Gonnord_SAS11} to
 the elements of $A$ until $A$ is empty. 
@@
 These narrowing iterations allow to recover precision lost by
-widening, before computing and taking into account new feasible paths.
+widening, \emph{before} computing and taking into account new feasible paths.
 Thus, our technique combines both the advantages of \emph{Guided Static
 Analysis} and \emph{Path-focusing}.
-
-After narrowing, we restart the analysis, starting from the computation the new
-subgraph.
-
-\begin{algorithm}
-	\caption{Guided static analysis on implicit multigraph}
-	\label{algo:combined}
-	\begin{algorithmic}[1] 
-	\input{algo/algo-combined.tex}
-	\end{algorithmic}
-\end{algorithm}
 
@@
 \subsection{Analysis algorithm}
+\label{sec:analysis-algorithm}
 Our tool operates over LLVM bitcode \citep{LLVM_langref,Lattner:2004:LCF:977395.977673}.

### 2011-10-27 papers/PLDI2012/implicitization.tex
On ne distingue plus X_i^s et X_i^d
Le fait d'avoir deux valeurs abstraites pour chaque point est utile pour
le narrowing, et le fait de moralement couper les points en deux est
importante dans les formules du SMT-solving (b^s et b^d).

Mais pour la découverte de nouveaux chemins, on (Julien et moi) ne trouve
pas de cas où c'est intéressant de mettre à jour une valeur temporaire
X^d plutôt que de mettre à jour directement X^s (en faisant une mise à
jour sans élargissement).

Le code fait en réalité la distinction, donc pas exactement ce que fait
l'algo après ce commit, mais la distinction X^s/

@@
 unfeasible paths.
-In this section, we use the same notations as \ref{sec:path_focusing}, except
-that we call the abstract values $X_i^s$ instead of $X_i$ to leave room for
-another abstract value $X_i^d$ later in the algorithm.
+In this section, we use the same notations as \ref{sec:path_focusing}.
 
@@
 subsets of the paths between two points in $P_R$.
-We iterate the operations until we consider the whole program, i.e all the
-feasible paths. The result will then be an inductive invariant of the entire program.
+We iterate the operations until the whole program (i.e all the
+feasible paths) has been considered. The result will then be an
+inductive invariant of the entire program.
 
@@
   are not in $P$. This phase computes these paths and adds them to
-  $P'$ (line~\ref{alg=computeNewPaths} in
-  algorithm~\ref{algo:combined}). We also use SMT solving to discover
-  these new paths, but we slightly change the SMT formula given to the
-  SMT solver: we now simply check for the satisfiability of $f(p_i)$
-  instead of $f(p_i) \wedge P$. Since we already know that $f(p_i)
-  \wedge P$ is unsatisfiable, all the paths given by the SMT solver
-  won't be in $P$. This phase is described in~\ref{subsec:addingpaths}
+  $P'$. This phase is described in~\ref{subsec:addingpaths}
   and corresponds to lines in~\ref{alg=start-add-paths} to
@@
 
+\MM{Pas sûr que ça soit nécessaire, mais vu qu'on a eu une version
+  fausse avant, c'est peut-être bien de dire comment on est arrivés à
+  la meilleure version.}
+The order of steps is important: narrowing has to be performed before
+adding new paths, or some spurious new paths would be added to $P$,
+and starting with the addition of new paths avoids wasting time
+doing the ascending iterations on an empty graph.
+
 \subsection{Ascending iterations by Path-focusing}
 \label{subsec:ascending}
+
+\MM{Maintenant que l'ordre a changé dans l'algo, on peut peut-être
+  échanger cette section et la suivante ?}
 
@@
 model when satisfiable is a path that starts in $p_i$, goes to a successor $p_j
-\in P_R$ of $p_i$, such that the image of $X_{i}^s$ by the path transformation
-is not included in the current $X_{j}^s$.
-Intuitively, such a path makes the abstract value $X_{j}^s$ grow, and thus is
+\in P_R$ of $p_i$, such that the image of $X_{i}$ by the path transformation
+is not included in the current $X_{j}$.
+Intuitively, such a path makes the abstract value $X_{j}$ grow, and thus is
 an interesting path to focus on. We loop until the formula becomes unsatisfiable,
@@
 If we note $Succ(i)$ the set of indices $j$ such that $p_j \in P_R$ is a
-successor of $p_i$ in the expanded multigraph, and $X_i^s$ the abstract value
+successor of $p_i$ in the expanded multigraph, and $X_i$ the abstract value
 associated to $p_i$ :
@@
 \displaystyle\bigwedge_{j \in P_R \atop j \neq i} \neg
-b_j^s \wedge x \in X_i^s \wedge \displaystyle\bigvee_{j \in Succ(i)} (b_j^d \wedge
-x \notin X_j^s)$$
+b_j^s \wedge x \in X_i \wedge \displaystyle\bigvee_{j \in Succ(i)} (b_j^d \wedge
+x \notin X_j)$$
 
@@
 subgraphs, until the complete graph is reached.
-
 When the analysis of a subgraph is finished, meaning that the abstract values
@@
 $P'$ of new paths that become feasible regarding the current abstract values.
-
-\MM{Cette partie est devenue toute petite, il y avait un résumé global
-  de l'algo ici. Ça me parait suffisant comme ça}
-
-It is not necessary to add into $P'$ all the new feasible paths. We just have
-to add them one after another, until no more path can make the invariant grow,
-as proposed by Algorithm \ref{algo:computepaths}. For that, we use another
-abstract value associated to a control point $p_j$, noted $X_j^d$, which is
-distinct from $X_j^s$, and initialized to $X_j^s$.
+The paths in $P'$ are computed one after another, until no more path
+can make the invariant grow. This is line~\ref{alg=computeNewPaths} in
+Algorithm~\ref{algo:combined}, which corresponds to
+Algorithm~\ref{algo:computepaths}. We also use SMT solving to discover
+these new paths, but we subtly change the SMT formula given to the
+SMT solver: we now simply check for the satisfiability of $f(p_i)$
+instead of $f(p_i) \wedge P$. Since we already know that $f(p_i)
+\wedge P$ is unsatisfiable, none of the paths given by the SMT solver
+will be in $P$.
 
@@
 \begin{figure}[h]
+  \MM{Y a-t-il moyen d'avoir un truc plus lisible en impression noir
+    et blanc, genre des hachures ?}
   \begin{center}

### 2011-10-27 papers/PLDI2012/implicitization.tex
quelques TODOs dans le debut

@@
 \section{Introduction}
-% TODO expliquer 'implicite'
 Static analysis by abstract interpretation is a fully automatic program analysis method.
@@
 of the lattice, but the number of terms in the union may grow exponentially with the number of successive tests in the program to analyze, not to mention difficulties for designing suitable widening operators for enforcing the convergence of fixpoint iterations~\cite{PPL,BagnaraHZ08SCP,DBLP:journals/sttt/BagnaraHZ07}.
-The exponential growth of the number of terms in the union may be controlled by heuristics that judiciously apply least upper bound operations, as in the \emph{trace partitioning domain} \cite{Rival_Mauborgne_TOPLAS07} implemented in the Astr\'ee analyzer~\cite{ASTREE_TASE07,ASTREE_ESOP05,ASTREE_PLDI03}.
-% TODO ne garder que les papiers les plus significatifs par rapport au sujet évoqué
+The exponential growth of the number of terms in the union may be controlled by heuristics that judiciously apply least upper bound operations, as in the \emph{trace partitioning domain} \cite{Rival_Mauborgne_TOPLAS07} implemented in the Astr\'ee analyzer~\cite{ASTREE_PLDI03,ASTREE_ESOP05}.
 
@@
 Two recent proposals use SMT-solving \cite{Kroening_Strichman_08} as a decision procedure for the satisfiability of first-order arithmetic formulas in order to enumerate only paths that are needed for the progress of the analysis \cite{Gawlitza_Monniaux_ESOP11,Monniaux_Gonnord_SAS11}.
+They can equivalently be seen as analyses over a multigraph of transitions between some distinguished control nodes.
+This multigraph has an exponential number of edges, but is never explicitly represented in memory; instead, this graph is \emph{implicitly} or \emph{succinctly} represented: its edges are enumerated as needed as solutions to SMT problems.
 
@@
 \item We improve the technique for obtaining disjunctive invariants from \citet{DBLP:conf/pldi/GulwaniZ10} by replacing the explicit exhaustive enumeration of paths by a sequence of SMT queries.\label{contr:disjunctive} (See \S\ref{sec:disjunctive})
-% TODO: 'We implemented' pas une contribution; rephrase.
 \item We implemented these techniques, in addition to ``classical'' iterations and the original guided static analysis, inside a prototype static analyzer.
@@
 With respect to abstract domains on numerical variables, it uses the APRON library \cite{DBLP:conf/cav/JeannetM09}, which supports a variety of abstract domains from which we can choose with minimal changes to our analyzer.
+This article summarizes our implementation decisions for analysis of SSA-form programs.
 \item We conducted extensive experiments with this tool, on real-life programs, thus obtaining experimental comparisons between the various combinations of techniques. (\S\ref{sec:experiments})
@@
 
+Once an inductive invariant $\gamma((\abstr{x}_p)_{p \in P})$ has been obtained, one can attempt \emph{decreasing} or \emph{narrowing} iterations to reduce it.
+In their simplest form, this just means running the following operation until a fixpoint or a maximal number of iterations are reached: for any $p'$, replace $\abstr{x}_{p'}$ by $\abstr{x}_p \cap \left(\bigsqcup_{p \in P} \abstr{\tau}_{p,p'} (\abstr{x}_p)\right)$.
+The result also defines an inductive invariant.
+These decreasing iterations are indispensable to recover properties from guards (tests) in the program in most iteration settings; unfortunately, certain loops, particularly those involving identity (no-operation) transitions, may foil them: the iterations immediately reach a fixpoint and do not decrease further (see example in \S\ref{subsec:rate_lim}).
+Sections \ref{sec:guided} and \ref{sec:path_focusing} describe techniques that work around this problem.
+
 \subsection{SMT-solving}
@@
 
-The expected inductive invariant is $\verb|x_old| \in [-100000,100000]$, but classical abstract interpretation using intervals (or octagons or polyhedra) finds $\verb|x_old| \in (-\infty,+\infty)$.
-% TODO chercher ref Astree
+The expected inductive invariant is $\verb|x_old| \in [-100000,100000]$, but classical abstract interpretation using intervals (or octagons or polyhedra) finds $\verb|x_old| \in (-\infty,+\infty)$~\citep{ASTREE_ESOP05}.
 Let us briefly see why.
 
-Widening iterations converge to $\lstinline|x_old| \in (-\infty,+\infty)$; let us now see why decreasing iterations fail to recover the desired invariant. % TODO expliquer widening suivi de narrowing
+Widening iterations converge to $\lstinline|x_old| \in (-\infty,+\infty)$; let us now see why decreasing iterations fail to recover the desired invariant.
 The \lstinline|x >= x_old+10| test at line~6, if taken, yields $\lstinline|x_old| \in (-\infty,99990)$; followed by \lstinline|x = x_old+10|, we obtain $\lstinline|x| \in (-\infty,\allowbreak 100000]$, and the same after union with the no-operation ``else'' branch.

### 2011-10-27 papers/PLDI2012/implicitization.tex
TODOs, figure

@@
 Consider the following program:
-\lstinputlisting{gopan_reps.c}
+\lstinputlisting[label=lst:gopan_reps]{gopan_reps.c}
+
+\begin{figure}
+\begin{center}
+\includegraphics{figures/gopan_reps}
+\end{center}
+\caption{The invariant for Prog.~\ref{lst:gopan_reps}: the piecewise linear, solid line is the strongest invariant, the grayed polyhedron is its convex hull.}
+\label{fig:gopan_reps_invariant} 
+\end{figure}
 
@@
 From there, the ``else'' branch at line 5 may be taken; with further widening, $0 \leq y \leq x$ is obtained as a loop invariant, and thus the postcondition computed at line 9 is $x \geq 0 \land y = 0$.
-Yet the strongest invariant is $(0 \leq x \leq 51 \land y = x) \lor (51 \leq x \leq 102 \land x+y=102)$, and its convex hull, a convex polyhedron, is
+Yet the strongest invariant is $(0 \leq x \leq 51 \land y = x) \lor (51 \leq x \leq 102 \land x+y=102)$, and its convex hull, a convex polyhedron (Fig.~\ref{fig:gopan_reps_invariant}), is
 \begin{equation}
 y \leq x \land y \leq 102-x \land y \geq 0.\label{eqn:triangle}
-\end{equation} % TODO fig
+\end{equation}
 Intuitively, this disappointing result is obtained because widening extrapolates from the first iterations of the loop, but the loop has two different phases ($x \leq 50$ and $x > 50$) with different behaviors, thus the extrapolation from the first phase is not valid for the second.

### 2011-10-27 papers/PLDI2012/implicitization.tex
+description implem

@@
 such operations are directly supported by APRON.
-A first benefit is to limit the number of dimensions in the abstract values, since polyhedra libraries typically perform worse with higher dimensions. TODO
-Note that this means that the abstract operation for a single path in path-focusing methods also is a (large) parallel assignment.
+A first benefit is to limit the number of dimensions in the abstract values, since polyhedra libraries typically perform worse with higher dimensions.
+A second benefit is that the abstract operation for a single path in path-focusing methods also is a (large) parallel assignment.
+
+We further reduce the number of variables by projecting out those that are not live at the program point. Again, SSA use-def and def-use chains make it easy to compute liveness.
 
@@
 to the program, and we unroll every loop once.
+
+Our implementation of path-focusing currently does not use true acceleration techniques, as proposed by \citet{Monniaux_Gonnord_SAS11}.
+Instead, it simply runs widening and narrowing iterations on a single path.
 

### 2011-10-27 papers/PLDI2012/implicitization.tex
en bloc analysis

@@
 \usepackage{listings}
-\lstset{language=C,mathescape=true,numbers=left,numbersep=0.8ex,tabsize=2} % basicstyle={\rm\fontfamily{ppl}\selectfont} trop reconnaissable
+\lstset{language=C,mathescape=true,numbers=left,numbersep=0.8ex,tabsize=2,breaklines=true} % basicstyle={\rm\fontfamily{ppl}\selectfont} trop reconnaissable
 
@@
  f_n(v_1,\dots,v_n))$;
-such operations are directly supported by APRON.
-A first benefit is to limit the number of dimensions in the abstract values, since polyhedra libraries typically perform worse with higher dimensions.
-A second benefit is that the abstract operation for a single path in path-focusing methods also is a (large) parallel assignment.
+such operations are directly supported by APRON. This has three benefits:
+(i) it limits the number of dimensions in the abstract values, since polyhedra libraries typically perform worse with higher dimensions;
+(ii) the abstract operation for a single path in path-focusing methods also is a (large) parallel assignment;
+(iii) as suggested by \citet{DBLP:conf/vmcai/Mine06}, this approach is more precise than running abstract operations for each program line separately:
+for instance, for \lstinline|y=x; z=x-y;| with precondition $x \in [0,1]$, a line-by-line interval analysis obtains $y \in [0,1]$ and $z \in [-1,1]$ while our ``en bloc'' analysis symbolically simplifies $z = x - x = 0$ and thus $z \in [0,0]$.
 

### 2011-10-27 papers/PLDI2012/implicitization.tex
tassage
+ citation Laure

@@
 
+\usepackage{enumitem}
 \usepackage{dcolumn}
@@
 \begin{center}
-\includegraphics{figures/gopan_reps}
+\includegraphics[scale=0.75]{figures/gopan_reps}
 \end{center}
@@
 we can apply a widening/narrowing sequence, or even compute the transitive
-closure of the loop using abstract acceleration. % TODO citation laure gonnord
+closure of the loop (or an approximation thereof, or its application to $X_i$)
+using abstract acceleration~\cite{DBLP:conf/sas/GonnordH06}.
 
@@
 In cases where the formula is expressed in too rich a logic for the SMT-solver to deal with, a number of workarounds are possible:
-\begin{itemize}
+\begin{itemize}[itemsep=-\parsep,topsep=0pt,partopsep=0pt]
 \item \emph{Linearization}, as per \citet{DBLP:conf/vmcai/Mine06}, which overapproximates nonlinear semantics by linear semantics.

### 2011-10-27 ex/nested_loop4.c
bug fix in AIdis

@@ -2,8 +2,8 @@ void mat_product(unsigned n) {
 	n=100;
 if (n>0) {
   for(unsigned i=0; i<n; i++) {
-    for(unsigned j=0; j<n; j++) {
-      for(unsigned k=0; k<n; k++) {
+    for(unsigned j=0; j<i; j++) {
+      for(unsigned k=0; k<j; k++) {
       }
     }
   }

### 2011-10-27 src/AIdis.cc

@@ -215,7 +215,7 @@ void AIdis::computeNewPaths(Node * n) {
 	while (true) {
 		DEBUG(
 			Out->changeColor(raw_ostream::RED,true);
-			*Out << "-------------- NEW SMT SOLVE2 -------------------------\n";
+			*Out << "COMPUTEPATHS------ NEW SMT SOLVE -------------------------\n";
 			Out->resetColor();
 		);
 		// creating the SMTpass formula we want to check
@@ -245,8 +245,6 @@ void AIdis::computeNewPaths(Node * n) {
 
 		computeTransform(Xdisj->man_disj,n,path,*Xtemp);
 
-		Xdisj->change_environment(Xtemp->main->env,index);
-
 		AbstractDisj * SuccDisj = dynamic_cast<AbstractDisj*>(Succ->X_d[passID]);
 		int Sigma = sigma(path,index,Xtemp,false);
 		Join.clear();
@@ -258,6 +256,10 @@ void AIdis::computeNewPaths(Node * n) {
 
 		// there is a new path that has to be explored
 		pathtree[n->bb]->insert(path,true);
+		DEBUG(
+			*Out << "INSERTING INTO P THE PATH\n";
+			printPath(path);
+		);
 		A.push(n);
 	}
 }
@@ -350,7 +352,7 @@ void AIdis::computeNode(Node * n) {
 		is_computed[n] = true;
 		DEBUG(
 			Out->changeColor(raw_ostream::RED,true);
-			*Out << "-------------- NEW SMT SOLVE -------------------------\n";
+			*Out << "COMPUTENODE------- NEW SMT SOLVE -------------------------\n";
 			Out->resetColor();
 		);
 		LSMT->push_context();

### 2011-10-27 papers/PLDI2012/implicitization.tex
references dans la table des matieres
put title in pdf info
Revert "On ne distingue plus X_i^s et X_i^d"
This reverts commit 9dba15636aca443d1c5c5ffd5035f9f8d714ad91.
This reverts commit 3899366d305fbf8d450b66a06ff12bb04501f2f4.

--- 2011-10-27 papers/PLDI2012/implicitization.tex
@@
 \section{Introduction}
+% TODO expliquer 'implicite'
 Static analysis by abstract interpretation is a fully automatic program analysis method.
@@
 of the lattice, but the number of terms in the union may grow exponentially with the number of successive tests in the program to analyze, not to mention difficulties for designing suitable widening operators for enforcing the convergence of fixpoint iterations~\cite{PPL,BagnaraHZ08SCP,DBLP:journals/sttt/BagnaraHZ07}.
-The exponential growth of the number of terms in the union may be controlled by heuristics that judiciously apply least upper bound operations, as in the \emph{trace partitioning domain} \cite{Rival_Mauborgne_TOPLAS07} implemented in the Astr\'ee analyzer~\cite{ASTREE_PLDI03,ASTREE_ESOP05}.
+The exponential growth of the number of terms in the union may be controlled by heuristics that judiciously apply least upper bound operations, as in the \emph{trace partitioning domain} \cite{Rival_Mauborgne_TOPLAS07} implemented in the Astr\'ee analyzer~\cite{ASTREE_TASE07,ASTREE_ESOP05,ASTREE_PLDI03}.
+% TODO ne garder que les papiers les plus significatifs par rapport au sujet évoqué
 
@@
 Two recent proposals use SMT-solving \cite{Kroening_Strichman_08} as a decision procedure for the satisfiability of first-order arithmetic formulas in order to enumerate only paths that are needed for the progress of the analysis \cite{Gawlitza_Monniaux_ESOP11,Monniaux_Gonnord_SAS11}.
-They can equivalently be seen as analyses over a multigraph of transitions between some distinguished control nodes.
-This multigraph has an exponential number of edges, but is never explicitly represented in memory; instead, this graph is \emph{implicitly} or \emph{succinctly} represented: its edges are enumerated as needed as solutions to SMT problems.
 
@@
 With respect to abstract domains on numerical variables, it uses the APRON library \cite{DBLP:conf/cav/JeannetM09}, which supports a variety of abstract domains from which we can choose with minimal changes to our analyzer.
-This article summarizes our implementation decisions for analysis of SSA-form programs.
 \item We conducted extensive experiments with this tool, on real-life programs, thus obtaining experimental comparisons between the various combinations of techniques. (\S\ref{sec:experiments})
@@
 
-Once an inductive invariant $\gamma((\abstr{x}_p)_{p \in P})$ has been obtained, one can attempt \emph{decreasing} or \emph{narrowing} iterations to reduce it.
-In their simplest form, this just means running the following operation until a fixpoint or a maximal number of iterations are reached: for any $p'$, replace $\abstr{x}_{p'}$ by $\abstr{x}_p \cap \left(\bigsqcup_{p \in P} \abstr{\tau}_{p,p'} (\abstr{x}_p)\right)$.
-The result also defines an inductive invariant.
-These decreasing iterations are indispensable to recover properties from guards (tests) in the program in most iteration settings; unfortunately, certain loops, particularly those involving identity (no-operation) transitions, may foil them: the iterations immediately reach a fixpoint and do not decrease further (see example in \S\ref{subsec:rate_lim}).
-Sections \ref{sec:guided} and \ref{sec:path_focusing} describe techniques that work around this problem.
-
 \subsection{SMT-solving}
@@
 
-The expected inductive invariant is $\verb|x_old| \in [-100000,100000]$, but classical abstract interpretation using intervals (or octagons or polyhedra) finds $\verb|x_old| \in (-\infty,+\infty)$~\citep{ASTREE_ESOP05}.
+The expected inductive invariant is $\verb|x_old| \in [-100000,100000]$, but classical abstract interpretation using intervals (or octagons or polyhedra) finds $\verb|x_old| \in (-\infty,+\infty)$.
+% TODO chercher ref Astree
 Let us briefly see why.
 
-Widening iterations converge to $\lstinline|x_old| \in (-\infty,+\infty)$; let us now see why decreasing iterations fail to recover the desired invariant.
+Widening iterations converge to $\lstinline|x_old| \in (-\infty,+\infty)$; let us now see why decreasing iterations fail to recover the desired invariant. % TODO expliquer widening suivi de narrowing
 The \lstinline|x >= x_old+10| test at line~6, if taken, yields $\lstinline|x_old| \in (-\infty,99990)$; followed by \lstinline|x = x_old+10|, we obtain $\lstinline|x| \in (-\infty,\allowbreak 100000]$, and the same after union with the no-operation ``else'' branch.
@@
 Consider the following program:
-\lstinputlisting[label=lst:gopan_reps]{gopan_reps.c}
-
-\begin{figure}
-\begin{center}
-\includegraphics{figures/gopan_reps}
-\end{center}
-\caption{The invariant for Prog.~\ref{lst:gopan_reps}: the piecewise linear, solid line is the strongest invariant, the grayed polyhedron is its convex hull.}
-\label{fig:gopan_reps_invariant} 
-\end{figure}
+\lstinputlisting{gopan_reps.c}
 
@@
 From there, the ``else'' branch at line 5 may be taken; with further widening, $0 \leq y \leq x$ is obtained as a loop invariant, and thus the postcondition computed at line 9 is $x \geq 0 \land y = 0$.
-Yet the strongest invariant is $(0 \leq x \leq 51 \land y = x) \lor (51 \leq x \leq 102 \land x+y=102)$, and its convex hull, a convex polyhedron (Fig.~\ref{fig:gopan_reps_invariant}), is
+Yet the strongest invariant is $(0 \leq x \leq 51 \land y = x) \lor (51 \leq x \leq 102 \land x+y=102)$, and its convex hull, a convex polyhedron, is
 \begin{equation}
 y \leq x \land y \leq 102-x \land y \geq 0.\label{eqn:triangle}
-\end{equation}
+\end{equation} % TODO fig
 Intuitively, this disappointing result is obtained because widening extrapolates from the first iterations of the loop, but the loop has two different phases ($x \leq 50$ and $x > 50$) with different behaviors, thus the extrapolation from the first phase is not valid for the second.
@@
 unfeasible paths.
-In this section, we use the same notations as \ref{sec:path_focusing}.
+In this section, we use the same notations as \ref{sec:path_focusing}, except
+that we call the abstract values $X_i^s$ instead of $X_i$ to leave room for
+another abstract value $X_i^d$ later in the algorithm.
 
@@
 model when satisfiable is a path that starts in $p_i$, goes to a successor $p_j
-\in P_R$ of $p_i$, such that the image of $X_{i}$ by the path transformation
-is not included in the current $X_{j}$.
-Intuitively, such a path makes the abstract value $X_{j}$ grow, and thus is
+\in P_R$ of $p_i$, such that the image of $X_{i}^s$ by the path transformation
+is not included in the current $X_{j}^s$.
+Intuitively, such a path makes the abstract value $X_{j}^s$ grow, and thus is
 an interesting path to focus on. We loop until the formula becomes unsatisfiable,
@@
 If we note $Succ(i)$ the set of indices $j$ such that $p_j \in P_R$ is a
-successor of $p_i$ in the expanded multigraph, and $X_i$ the abstract value
+successor of $p_i$ in the expanded multigraph, and $X_i^s$ the abstract value
 associated to $p_i$ :
@@
 \displaystyle\bigwedge_{j \in P_R \atop j \neq i} \neg
-b_j^s \wedge x \in X_i \wedge \displaystyle\bigvee_{j \in Succ(i)} (b_j^d \wedge
-x \notin X_j)$$
+b_j^s \wedge x \in X_i^s \wedge \displaystyle\bigvee_{j \in Succ(i)} (b_j^d \wedge
+x \notin X_j^s)$$
 
@@
 will be in $P$.
+
+We use another
+abstract value associated to a control point $p_j$, noted $X_j^d$, which is
+distinct from $X_j^s$, and initialized to $X_j^s$.
 
@@
 such operations are directly supported by APRON.
-A first benefit is to limit the number of dimensions in the abstract values, since polyhedra libraries typically perform worse with higher dimensions.
-A second benefit is that the abstract operation for a single path in path-focusing methods also is a (large) parallel assignment.
-
-We further reduce the number of variables by projecting out those that are not live at the program point. Again, SSA use-def and def-use chains make it easy to compute liveness.
+A first benefit is to limit the number of dimensions in the abstract values, since polyhedra libraries typically perform worse with higher dimensions. TODO
+Note that this means that the abstract operation for a single path in path-focusing methods also is a (large) parallel assignment.
 
@@
 to the program, and we unroll every loop once.
-
-Our implementation of path-focusing currently does not use true acceleration techniques, as proposed by \citet{Monniaux_Gonnord_SAS11}.
-Instead, it simply runs widening and narrowing iterations on a single path.
 

### 2011-10-27 papers/PLDI2012/implicitization.tex
Clarification sur X^d

@@
 
-We use another
-abstract value associated to a control point $p_j$, noted $X_j^d$, which is
-distinct from $X_j^s$, and initialized to $X_j^s$.
+\MM{À changer si on décide de ne plus distinguer $X^s$ et $X^d$}
+
+To prevent the algorithm from adding too many path at a time, we
+use another abstract value associated to a control point $p_j$, noted
+$X_j^d$, which is distinct from $X_j^s$, and initialized to $X_j^s$.
+This value is updated when a point $p_j$ is the target of a new path,
+so that further SMT queries do not compute other paths with the same
+source and destination if it is not needed (because these new paths
+would not make $X_j^d$ grow, hence would not be returned by the SMT
+solver).
 

### 2011-10-27 papers/PLDI2012/implicitization.tex
typo

--- 2011-10-27 papers/PLDI2012/implicitization.tex
@@
 
-\usepackage{hyperref}
+\usepackage[pdfusetitle]{hyperref}
 
@@
 will be in $P$.
+
+\MM{À changer si on décide de ne plus distinguer $X^s$ et $X^d$}
+
+To prevent the algorithm from adding too many path at a time, we
+use another abstract value associated to a control point $p_j$, noted
+$X_j^d$, which is distinct from $X_j^s$, and initialized to $X_j^s$.
+This value is updated when a point $p_j$ is the target of a new path,
+so that further SMT queries do not compute other paths with the same
+source and destination if it is not needed (because these new paths
+would not make $X_j^d$ grow, hence would not be returned by the SMT
+solver).
 
@@
 of dynamic construction of the $\sigma$ function, adapted from
-\cite{DBLP:conf/pldi/GulwaniZ10}.  We start with one sigle
+\cite{DBLP:conf/pldi/GulwaniZ10}.  We start with one single
 disjunct per control point ($m=1$), and we start with an empty map $\sigma$. We also
@@
 
+\phantomsection\addcontentsline{toc}{section}{References} 
 \bibliographystyle{dmabbrvnat}

### 2011-10-28 papers/PLDI2012/implicitization.tex
Remarque sur widening pas monotone
nouveaux résultats de DIS

@@
 An alternative is to change the abstract domain, for instance by moving from intervals to octagons or convex polyhedra \cite{DBLP:journals/lisp/Mine06,Mine_PhD}, or the widening operator: for instance one may replace the classical widening on convex polyhedra \cite{CousotHalbwachs78,Halbwachs_PhD} by an improved version \cite{BagnaraHRZ05SCP}, or use widening ``up to'', that is, use linear inequalities syntactically present in the program source code as possible widening steps~\cite{Polka:FMSD:97}.
+
+\MM{Peut-être dire quelque part vers ici que le widening n'est en
+  général pas monotone, et que donc on ne peut jamais rien dire en
+  théorie sur la précision vu qu'être plus précis à une itération peut
+  faire perdre en précision la suivante ...}
 
@@
 |D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
+|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
 |D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}|} \hline
@@
 & \multicolumn{3}{c|}{\textbf{G+PF/PF}}
-& \multicolumn{3}{c|}{\textbf{G+PF/G}} \\ %\cline{2-16}
+& \multicolumn{3}{c|}{\textbf{G+PF/G}}
+& \multicolumn{3}{c|}{\textbf{DIS/G+PF}} \\ %\cline{2-16}
+& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
 & \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
@@
 \emph{Guided Static Analysis} from \citet{DBLP:conf/sas/GopanR07} improves the
-result of the classical Abstract Interpretation in $2.37\%$ of the control points
+result of the classical Abstract Interpretation in $2.33\%$ of the control points
 in $P_R$.
@@
 statistically better results, and finds better invariants than \emph{Guided
-Static Analysis} in $7.28\%$ of the cases. However, it also loses precision in an
-important number ($2.93\%$) of control points.
+Static Analysis} in $8.58\%$ of the cases. However, it also loses precision in an
+important number ($2.87\%$) of control points.
 Finally, our combined technique gives the most promising results, since it is
@@
 
+\JH{Paragraphe a reecrire:}
 The analysis using disjunctive invariants seems not to be as efficient as

### 2011-10-28 papers/PLDI2012/implicitization.tex
legendes des tables

@@
 \subsection{Precision of the various techniques}
+\label{sec:compare_techniques}
 
@@
 paper: classic Abstract Interpretation (S), \emph{Guided Static Analysis} (G),
-\emph{Path-focusing} (PF), and our combined technique (G+PF). For
+\emph{Path-focusing} (PF), our combined technique (G+PF), and its version using
+disjunctive invariants (DIS). For
 instance, \textbf{G/S} compares the benefits of \emph{Guided Static Analysis}
@@
 \caption{Results of the comparison of the various abstract domains, when using
-the same technique (G+PF).}
-\label{tab:time}
+the same technique (G+PF). We used as abstract domains Convex Polyhedra (PK),
+Octagons (OCT), intervals (BOX) and linear equalities (PKEQ).}
+\label{tab:domain}
 \end{table*}
+
+For each program and each pair $(D_1,D_2)$ of abstract domains, we compare by
+inclusion the invariants of the different control points in $P_R$ (where $P_R$
+is defined as in \ref{sec:compare_techniques}).
+The detailed results are given in Table \ref{tab:domain}.
 

### 2011-10-28 papers/PLDI2012/implicitization.tex
mise a jour resultats benchs
typos

@@
 
-There are many possible combinations of the above techniques, and it is not evident which ones perform more or less precisely or more or less efficiency on real-life examples.
+There are many possible combinations of the above techniques, and it is not
+evident which ones perform more or less precisely or more or less efficiently on real-life examples.
 One needs to experiment.
@@
 \begin{lstlisting}[numbers=none]
-if (wait) {
+if (wait()) {
    while (wait()) {}
@@
 abstract value $X_k$ by the path indexed by $j$, and try to find an existing
-disjunct of index $k'$ so that the convex hull of the two abstract values is
+disjunct of index $k'$ so that the least upper bound of the two abstract values is
 exactly the union of them. If such an index exists, then $\sigma(j,k) = k'$.
@@
 \begin{figure}[h]
-  \MM{Y a-t-il moyen d'avoir un truc plus lisible en impression noir
-    et blanc, genre des hachures ?}
   \begin{center}

### 2011-10-28 papers/PLDI2012/implicitization.tex
blah sur widening non monotone

--- 2011-10-28 papers/PLDI2012/implicitization.tex
@@
 An alternative is to change the abstract domain, for instance by moving from intervals to octagons or convex polyhedra \cite{DBLP:journals/lisp/Mine06,Mine_PhD}, or the widening operator: for instance one may replace the classical widening on convex polyhedra \cite{CousotHalbwachs78,Halbwachs_PhD} by an improved version \cite{BagnaraHRZ05SCP}, or use widening ``up to'', that is, use linear inequalities syntactically present in the program source code as possible widening steps~\cite{Polka:FMSD:97}.
-
-\MM{Peut-être dire quelque part vers ici que le widening n'est en
-  général pas monotone, et que donc on ne peut jamais rien dire en
-  théorie sur la précision vu qu'être plus précis à une itération peut
-  faire perdre en précision la suivante ...}
 
@@
 Sections \ref{sec:guided} and \ref{sec:path_focusing} describe techniques that work around this problem.
+
+Widening operations have a somewhat counterintuitive behavior: the result of an analysis (postcondition) is not necessarily monotonic with respect to the precondition.
+For instance, classical interval analysis with widening and decreasing iterations on the following loop yields a loop invariant $x \in [0,+\infty)$ if the precondition is $x \in [0,0]$ , but $x \in [0,10]$ (strictly stronger) if the precondition is $x \in [0,10]$ (strictly weaker):
+\lstinline|while (x != 10) x=x+1;|.
+This means that, when using widenings, being locally more precise (for instance, by using more expressive abstract domains, or widening operators that climb more slowly \cite{BagnaraHRZ05SCP}) does not necessarily translate into higher final precision.
+It is almost always possible to concoct examples where a supposedly more precise abstraction (e.g.
+polyhedra) yields less precise results than a less precise one (e.g.
+intervals), but are these examples representative of real programs?
+This justifies our recourse to evaluation on realistic examples~(\S~\ref{sec:compare_techniques}).
 

### 2011-10-28 papers/PLDI2012/implicitization.tex

@@
 
-There are many possible combinations of the above techniques, and it is not evident which ones perform more or less precisely or more or less efficiency on real-life examples.
-One needs to experiment.
-Unfortunately, the published literature on the subject lacks experimental comparative assessments.
-One purpose of this article is therefore to propose such experimental results.
-\medskip
+There are many possible combinations of the above techniques, and it is not evident which ones perform more or less precisely or more or less efficiency on real-life examples. One needs to experiment.
+Unfortunately, the published literature on the subject lacks experimental comparative assessments. One purpose of this article is therefore to propose such experimental results.
 

### 2011-10-28 papers/PLDI2012/implicitization.tex
contribution list

@@
 With respect to abstract domains on numerical variables, it uses the APRON library \cite{DBLP:conf/cav/JeannetM09}, which supports a variety of abstract domains from which we can choose with minimal changes to our analyzer.
-This article summarizes our implementation decisions for analysis of SSA-form programs.
-\item We conducted extensive experiments with this tool, on real-life programs, thus obtaining experimental comparisons between the various combinations of techniques. (\S\ref{sec:experiments})
+Section~\ref{sec:analysis-algorithm} summarizes our implementation decisions for analysis of SSA-form programs.
+
+\item We conducted extensive experiments with this tool, on real-life programs: (i) fixing the abstract domain, varying the iteration technique (\S~\ref{sec:compare_techniques}) (ii) fixing the iteration technique, varying the abstract domain (\S\ref{sec:compare_domains}).
 \end{enumerate}
@@
 \subsection{Precision of Abstract Domains}
-
+\label{sec:compare_domains}
 \begin{table*}

### 2011-10-28 papers/PLDI2012/implicitization.tex
petit fix
correction exemple

--- 2011-10-28 papers/PLDI2012/implicitization.tex
@@
 
-There are many possible combinations of the above techniques, and it is not evident which ones perform more or less precisely or more or less efficiency on real-life examples. One needs to experiment.
-Unfortunately, the published literature on the subject lacks experimental comparative assessments. One purpose of this article is therefore to propose such experimental results.
+\MM{Peut-être dire quelque part vers ici que le widening n'est en
+  général pas monotone, et que donc on ne peut jamais rien dire en
+  théorie sur la précision vu qu'être plus précis à une itération peut
+  faire perdre en précision la suivante ...}
+
+There are many possible combinations of the above techniques, and it is not
+evident which ones perform more or less precisely or more or less efficiently on real-life examples.
+One needs to experiment.
+Unfortunately, the published literature on the subject lacks experimental comparative assessments.
+One purpose of this article is therefore to propose such experimental results.
+\medskip
 
@@
 
-Widening operations have a somewhat counterintuitive behavior: the result of an analysis (postcondition) is not necessarily monotonic with respect to the precondition.
-For instance, classical interval analysis with widening and decreasing iterations on the following loop yields a loop invariant $x \in [0,+\infty)$ if the precondition is $x \in [0,0]$ , but $x \in [0,10]$ (strictly stronger) if the precondition is $x \in [0,10]$ (strictly weaker):
-\lstinline|while (x != 10) x=x+1;|.
-This means that, when using widenings, being locally more precise (for instance, by using more expressive abstract domains, or widening operators that climb more slowly \cite{BagnaraHRZ05SCP}) does not necessarily translate into higher final precision.
-It is almost always possible to concoct examples where a supposedly more precise abstraction (e.g.
-polyhedra) yields less precise results than a less precise one (e.g.
-intervals), but are these examples representative of real programs?
-This justifies our recourse to evaluation on realistic examples~(\S~\ref{sec:compare_techniques}).
-
 \subsection{SMT-solving}
@@
 
+\JH{$99990)$, ca designe bien une borne d'intervalle ouvert ? J'ai plus
+l'habitude a la notation $99990[$}
 Widening iterations converge to $\lstinline|x_old| \in (-\infty,+\infty)$; let us now see why decreasing iterations fail to recover the desired invariant.
-The \lstinline|x >= x_old+10| test at line~6, if taken, yields $\lstinline|x_old| \in (-\infty,99990)$; followed by \lstinline|x = x_old+10|, we obtain $\lstinline|x| \in (-\infty,\allowbreak 100000]$, and the same after union with the no-operation ``else'' branch.
-Line~7 yields $\lstinline|x| \in (-\infty,+\infty)$.
+The \lstinline|x > x_old+10| test at line~6, if taken, yields $\lstinline|x_old|
+\in (-\infty,99990)$; followed by \lstinline|x = x_old+10|, we obtain
+$\lstinline|x| \in (-\infty,\allowbreak 100000)$, and the same after union with the no-operation ``else'' branch. Line~7 yields $\lstinline|x| \in (-\infty,+\infty)$.
 
 We could use ``widening up to'' or ``widening with thresholds'', propagating the ``magic values'' $\pm 100000$ associated to \lstinline|x| into~\lstinline|x_old|, but these syntactic approaches cannot directly cope with programs for which $\lstinline|x|  \in [-100000,+100000]$ is itself obtained by analysis.
-The guided static analysis of \citet{DBLP:conf/sas/GopanR07} performs slightly better, and obtains $\verb|x_old| \in (-\infty,100000]$.
+The guided static analysis of \citet{DBLP:conf/sas/GopanR07} does not perform
+better, and also obtains $\verb|x_old| \in (-\infty,+\infty)$.
 

### 2011-10-28 papers/PLDI2012/implicitization.tex

--- 2011-10-28 papers/PLDI2012/implicitization.tex
@@
 
-% TODO: discuter du big step SSA?
-% TODO: cite SSA
-
 Let us briefly comment the results given in more details in Table~\ref{tab:techniques} and Figure~\ref{fig:techniques}.
@@
 However, such transformations are not yet implemented in our tool.
-% TODO: c'est si dur de remplacer * non lineaire par undef?
+\DM{c'est si dur de remplacer * non lineaire par undef?}
 

### 2011-10-28 papers/PLDI2012/implicitization.tex
vire citations de these en francais

--- 2011-10-28 papers/PLDI2012/implicitization.tex
@@
 All these methods improve the precision of the analysis by keeping the same abstract domain (say, convex polyhedra) but changing the operations applied and their ordering.
-An alternative is to change the abstract domain, for instance by moving from intervals to octagons or convex polyhedra \cite{DBLP:journals/lisp/Mine06,Mine_PhD}, or the widening operator: for instance one may replace the classical widening on convex polyhedra \cite{CousotHalbwachs78,Halbwachs_PhD} by an improved version \cite{BagnaraHRZ05SCP}, or use widening ``up to'', that is, use linear inequalities syntactically present in the program source code as possible widening steps~\cite{Polka:FMSD:97}.
+An alternative is to change the abstract domain, for instance by moving from intervals to octagons or convex polyhedra \cite{DBLP:journals/lisp/Mine06}, or the widening operator: for instance one may replace the classical widening on convex polyhedra \cite{CousotHalbwachs78} by an improved version \cite{BagnaraHRZ05SCP}, or use widening ``up to'', that is, use linear inequalities syntactically present in the program source code as possible widening steps~\cite{Polka:FMSD:97}.
 
@@
 
-Classical iterations on the domain of convex polyhedra \cite{Halbwachs_PhD,CousotHalbwachs78} or octagons \cite{Mine_PhD,DBLP:journals/lisp/Mine06} start with $x = 0 \land x = 0$, then continue with $x = y \land 0 \leq x \leq 1$.
+Classical iterations on the domain of convex polyhedra \cite{CousotHalbwachs78,BagnaraHRZ05SCP} or octagons \cite{DBLP:journals/lisp/Mine06} start with $x = 0 \land x = 0$, then continue with $x = y \land 0 \leq x \leq 1$.
 The widening operator extrapolates from these two iterations and yields $x = y \land x \geq 0$.

### 2011-10-28 papers/PLDI2012/implicitization.tex
hop
je remets phd de nicolas

@@
 When applied to imperative programs, it computes an inductive invariant mapping each program location (or a subset thereof) to a set of states represented symbolically~\cite{CousotCousot_JLC92}.
-For instance, if we are only interested in scalar numerical program variables, such a set may be a convex polyhedron (the set of solutions of a system of linear inequalities)~\cite{CousotHalbwachs78,PPL,BagnaraHZ08SCP}.
+For instance, if we are only interested in scalar numerical program variables, such a set may be a convex polyhedron (the set of solutions of a system of linear inequalities)~\cite{CousotHalbwachs78,Halbwachs_PhD,PPL,BagnaraHZ08SCP}.
 
@@
 All these methods improve the precision of the analysis by keeping the same abstract domain (say, convex polyhedra) but changing the operations applied and their ordering.
-An alternative is to change the abstract domain, for instance by moving from intervals to octagons or convex polyhedra \cite{DBLP:journals/lisp/Mine06}, or the widening operator: for instance one may replace the classical widening on convex polyhedra \cite{CousotHalbwachs78} by an improved version \cite{BagnaraHRZ05SCP}, or use widening ``up to'', that is, use linear inequalities syntactically present in the program source code as possible widening steps~\cite{Polka:FMSD:97}.
+An alternative is to change the abstract domain, for instance by moving from intervals to octagons or convex polyhedra \cite{DBLP:journals/lisp/Mine06}, or the widening operator: for instance one may replace the classical widening on convex polyhedra \cite{CousotHalbwachs78,Halbwachs_PhD} by an improved version \cite{BagnaraHRZ05SCP}, or use widening ``up to'', that is, use linear inequalities syntactically present in the program source code as possible widening steps~\cite{Polka:FMSD:97}.
 
@@
 
+\DM{Garde-t-on la these en francais de Nicolas?}
+
 \phantomsection\addcontentsline{toc}{section}{References} 

### 2011-10-28 papers/PLDI2012/implicitization.tex
reponse commentaire
retrait TODO
linearisation
fix typos

@@
 The guided static analysis of \citet{DBLP:conf/sas/GopanR07} does not perform
-better, and also obtains $\verb|x_old| \in (-\infty,+\infty)$.
+better, and also obtains $\lstinline|x_old| \in (-\infty,+\infty)$.
 
@@
 Guided static analysis from \citet{DBLP:conf/sas/GopanR07} yields at line 5
-$\lstinline|x_old| \in~]-\infty,+\infty[$.
+$\lstinline|x_old| \in (-\infty,+\infty)$.
 Path-focusing \citep{Monniaux_Gonnord_SAS11} performs slightly
-better, and finds $\lstinline|x_old| \in~]-\infty,10000]$.
+better, and finds $\lstinline|x_old| \in (-\infty,10000]$.
 

### 2011-10-28 papers/PLDI2012/implicitization.tex
Gulwani c'est disjunctive

@@
 Assuming we are  interested in a loop-free program fragment, the above approach of keeping symbolic unions gives the same results as performing the analysis separately over every path in the fragment.
-A recent method for finding inductive loop invariants \cite{DBLP:conf/pldi/GulwaniZ10} is based on this idea: each path inside the loop body is considered separately.
+A recent method for finding disjunctive loop invariants \cite{DBLP:conf/pldi/GulwaniZ10} is based on this idea: each path inside the loop body is considered separately.
 Two recent proposals use SMT-solving \cite{Kroening_Strichman_08} as a decision procedure for the satisfiability of first-order arithmetic formulas in order to enumerate only paths that are needed for the progress of the analysis \cite{Gawlitza_Monniaux_ESOP11,Monniaux_Gonnord_SAS11}.

### 2011-10-28 papers/PLDI2012/implicitization.tex
typo
uniformisation apparence du code

@@
 
-The expected inductive invariant is $\verb|x_old| \in [-100000,100000]$, but classical abstract interpretation using intervals (or octagons or polyhedra) finds $\verb|x_old| \in (-\infty,+\infty)$~\citep{ASTREE_ESOP05}.
+The expected inductive invariant is $\lstinline|x_old| \in [-100000,100000]$, but classical abstract interpretation using intervals (or octagons or polyhedra) finds $\lstinline|x_old| \in (-\infty,+\infty)$~\citep{ASTREE_ESOP05}.
 Let us briefly see why.
@@
 \begin{itemize}[itemsep=-\parsep,topsep=0pt,partopsep=0pt]
-\item \emph{Linearization}, as per \citet{DBLP:conf/vmcai/Mine06}, which overapproximates nonlinear semantics by linear semantics.
+\item \emph{Linearization}, as per \citet{DBLP:conf/vmcai/Mine06}, which overapproximates nonlinear semantics by linear semantics. However, such transformations are not implemented in our tool.
 \item Replacing the results of nonlinear operations by ``unknown''.
 \end{itemize}
-However, such transformations are not yet implemented in our tool.
-\DM{c'est si dur de remplacer * non lineaire par undef?}
 

### 2011-10-28 papers/PLDI2012/implicitization.tex
fix virgule

@@
 
-We could use ``widening up to'' or ``widening with thresholds'', propagating the ``magic values'' $\pm 100000$ associated to \lstinline|x| into~\lstinline|x_old|, but these syntactic approaches cannot directly cope with programs for which $\lstinline|x|  \in [-100000,+100000]$ is itself obtained by analysis.
+We could use ``widening up to'' or ``widening with thresholds'', propagating the ``magic values'' $\pm 100000$ associated to \lstinline|x| into \mbox{\lstinline|x_old|,} but these syntactic approaches cannot directly cope with programs for which $\lstinline|x|  \in [-100000,+100000]$ is itself obtained by analysis.
 The guided static analysis of \citet{DBLP:conf/sas/GopanR07} does not perform
@@
 
+\JH{Commenter les resultats}
+
 \section{Conclusion}

### 2011-10-28 papers/PLDI2012/implicitization.tex
(invar)

@@
 Analyzing this loop using widening and narrowing on convex polyhedra or octagons yields the loop invariant $0 \leq x \leq 51 \land y = x$.
-Now, the transition at line~5 becomes feasible; and we analyze the full loop, starting iterations from $0 \leq x \leq 51 \land y = x$, and obtain invariant~\ref{eqn:triangle}.
+Now, the transition at line~5 becomes feasible; and we analyze the full loop, starting iterations from $0 \leq x \leq 51 \land y = x$, and obtain invariant~(\ref{eqn:triangle}).
 

### 2011-10-28 papers/PLDI2012/implicitization.tex
resultats grid

--- 2011-10-28 papers/PLDI2012/implicitization.tex
@@
 
-The expected inductive invariant is $\lstinline|x_old| \in [-100000,100000]$, but classical abstract interpretation using intervals (or octagons or polyhedra) finds $\lstinline|x_old| \in (-\infty,+\infty)$~\citep{ASTREE_ESOP05}.
+The expected inductive invariant is $\verb|x_old| \in [-100000,100000]$, but classical abstract interpretation using intervals (or octagons or polyhedra) finds $\verb|x_old| \in (-\infty,+\infty)$~\citep{ASTREE_ESOP05}.
 Let us briefly see why.
@@
 & \multicolumn{3}{c|}{\textbf{PK/PKEQ}}
-& \multicolumn{3}{c|}{\textbf{OCT/PKEQ}} \\ %\cline{2-16}
+& \multicolumn{3}{c|}{\textbf{OCT/PKEQ}}
+& \multicolumn{3}{c|}{\textbf{PK/GRID}} \\ %\cline{2-16}
+& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
 & \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
@@
 the same technique (G+PF). We used as abstract domains Convex Polyhedra (PK),
-Octagons (OCT), intervals (BOX) and linear equalities (PKEQ).}
+Octagons (OCT), intervals (BOX), linear equalities (PKEQ) and linear congruences
+(GRID).}
 \label{tab:domain}

### 2011-10-28 papers/PLDI2012/implicitization.tex
explicite difference
phrase plus jolie

@@
 
-Guided static analysis \cite{DBLP:conf/sas/GopanR07} applies to the transition
+Guided static analysis, as proposed by \citet{DBLP:conf/sas/GopanR07}, applied to the transition
 graph of the program. We now present a new technique applying this analysis on the implicit
@@
 
-The algorithm for Guided static analysis on the implicit multigraph is described
-in Algorithm \ref{algo:combined}.
+Algorithm~ \ref{algo:combined} performs Guided static analysis on the implicitly represented multigraph.
 

### 2011-10-28 papers/PLDI2012/implicitization.tex
comparaison domaines

--- 2011-10-28 papers/PLDI2012/implicitization.tex
@@
 is defined as in \ref{sec:compare_techniques}).
-The detailed results are given in Table \ref{tab:domain}.
-
+The main results are given in Table \ref{tab:domain}.
+
+Statistically, the domain of convex polyhedra gives the better results, but
+regularly yields greater invariants than the domains of octagons/intervals.
+Experiences shows is it mostly due to the widening operator.
+The Octagon domain appears to be much better than intervals.
 \JH{Commenter les resultats}
+
+Analysis of linear equalities (PKEQ) performs very fast compared
+to other abstract domains, but yields very unprecise invariants.
+The comparison between the domain of convex polyhedra and linear congruences
+shows the percentage of points where the use of congruences refines the
+invariant. This is the case in $1.04\%$ of the control points in $P_R$.
 

### 2011-10-28 papers/PLDI2012/implicitization.tex
fix anglais

--- 2011-10-28 papers/PLDI2012/implicitization.tex
@@
 
-To prevent the algorithm from adding too many path at a time, we
+To prevent the algorithm from adding too many paths at a time, we
 use another abstract value associated to a control point $p_j$, noted
@@
 
-Guided static analysis from \citet{DBLP:conf/sas/GopanR07} yields at line 5
+Guided static analysis from \citet{DBLP:conf/sas/GopanR07} yields, at line 5,
 $\lstinline|x_old| \in (-\infty,+\infty)$.
@@
 better, and finds $\lstinline|x_old| \in (-\infty,10000]$.
-
-Now, see how our technique performs on this example.
+Now, let us see how our technique performs on this example.
 
@@
 $\lstinline|x_old| \in~[-10000,\allowbreak 10000]$. 
-Here, only 3 paths over the 25 have been computed during the analysis. In
+Here, only 3 paths out of the 25 have been computed during the analysis. In
 practice, depending on the order the SMT-solver returns the paths, other

### 2011-10-28 papers/PLDI2012/implicitization.tex

@@
 is defined as in \ref{sec:compare_techniques}).
-The detailed results are given in Table \ref{tab:domain}.
-
+The main results are given in Table \ref{tab:domain}.
+
+Statistically, the domain of convex polyhedra gives the better results, but
+regularly yields greater invariants than the domains of octagons/intervals.
+Experiences shows is it mostly due to the widening operator.
+The Octagon domain appears to be much better than intervals.
 \JH{Commenter les resultats}
+
+Analysis of linear equalities (PKEQ) performs very fast compared
+to other abstract domains, but yields very unprecise invariants.
+The comparison between the domain of convex polyhedra and linear congruences
+shows the percentage of points where the use of congruences refines the
+invariant. This is the case for $1.04\%$ of the control points in $P_R$.
 

### 2011-10-28 papers/PLDI2012/implicitization.tex
typo
fix anglais

@@
 
-\citet{DBLP:conf/pldi/GulwaniZ10} propose a technique for computing disjunctive invariants, by
+\citet{DBLP:conf/pldi/GulwaniZ10} proposed a technique for computing disjunctive invariants, by
 distinguishing all the paths inside a loop. In
@@
 $X_{i,\sigma(j,k)}$.
-Initially, the $\delta$-th abstract value is assigned to the initial states of
-$p_i$, and each other abstract value is assigned to bottom.
+Initially, the $\delta$-th abstract value contains the initial states of
+$p_i$, and all other abstract values contain~$\emptyset$.
 

### 2011-10-28 papers/PLDI2012/implicitization.tex
typo

--- 2011-10-28 papers/PLDI2012/implicitization.tex
@@
 
-To prevent the algorithm from adding too many paths at a time, we
+To prevent the algorithm from adding too many path at a time, we
 use another abstract value associated to a control point $p_j$, noted
@@
 
-Guided static analysis from \citet{DBLP:conf/sas/GopanR07} yields, at line 5,
+Guided static analysis from \citet{DBLP:conf/sas/GopanR07} yields at line 5
 $\lstinline|x_old| \in (-\infty,+\infty)$.
@@
 better, and finds $\lstinline|x_old| \in (-\infty,10000]$.
-Now, let us see how our technique performs on this example.
+
+Now, see how our technique performs on this example.
 
@@
 $\lstinline|x_old| \in~[-10000,\allowbreak 10000]$. 
-Here, only 3 paths out of the 25 have been computed during the analysis. In
+Here, only 3 paths over the 25 have been computed during the analysis. In
 practice, depending on the order the SMT-solver returns the paths, other
@@
 regularly yields greater invariants than the domains of octagons/intervals.
-Experiences shows is it mostly due to the widening operator.
+Experiences shows it is mostly due to the widening operator.
 The Octagon domain appears to be much better than intervals.

### 2011-10-28 papers/PLDI2012/implicitization.tex
fix anglais

--- 2011-10-28 papers/PLDI2012/implicitization.tex
@@
 value, and the path we focus on. If we dynamically construct this function
-during the analysis, we should store the results of the sigma function into a
-compact representation, such as Algebraic Decision Diagrams. Notice that the
-image of a couple $(j,k)$ by the function $\sigma$ will be computed only if
-needed.
+during the analysis, we should store the results of the $\sigma$ function into a
+compact representation, such as Algebraic Decision Diagrams.
+\DM{Pourquoi des ADD et pas des BDD?}
+Note that $\sigma(j,k)$ will be computed only if needed.
 

### 2011-10-28 papers/PLDI2012/implicitization.tex
llvm + apron

@@
 
-Most SMT solvers, including Z3,%
-\footnote{\url{http://research.microsoft.com/en-us/um/redmond/projects/z3/}}
+Most SMT solvers, including Z3,
+%\footnote{\url{http://research.microsoft.com/en-us/um/redmond/projects/z3/}}
 Yices,%
-\footnote{\url{http://yices.csl.sri.com/}}
+%\footnote{\url{http://yices.csl.sri.com/}}
 and all those supporting the full SMTLIB2 standard \cite{BarST-SMTLIB},
@@
 compact representation, such as Algebraic Decision Diagrams.
-\DM{Pourquoi des ADD et pas des BDD?}
 Note that $\sigma(j,k)$ will be computed only if needed.
@@
 Static Analysis} \cite{DBLP:conf/sas/GopanR07}.
-We also can choose among different abstract domains: convex polyhedra, octagons,
-intervals, grids (convex polyhedra and linear congruences).
+
+Our tool operates over LLVM bitcode \citep{LLVM_langref,Lattner:2004:LCF:977395.977673}, which is a target for several compilers, most notably Clang (supporting C and C++) and llvm-gcc (supporting C, C++, Fortran and Ada).
+Abstract domains are provided by the APRON library \citep{DBLP:conf/cav/JeannetM09}, and include convex polyhedra, octagons, intervals, and linear congruences.
 For SMT-solving, our analyzer uses Yices
-\cite{DBLP:conf/cav/DutertreM06} or Microsoft Z3\cite{DBLP:conf/tacas/MouraB08}
-as SMT-solver.
+\cite{DBLP:conf/cav/DutertreM06} or Microsoft Z3\cite{DBLP:conf/tacas/MouraB08}.
 
@@
 \label{sec:analysis-algorithm}
-Our tool operates over LLVM bitcode \citep{LLVM_langref,Lattner:2004:LCF:977395.977673}.
 LLVM bitcode is in \emph{single static assignment} (SSA) form: a given scalar variable is given a value at a single syntactic point in the program.

### 2011-10-28 papers/PLDI2012/implicitization.tex
change style des listings
position .
Ce coup-ci, on a peut-être une vraie explication pour les X^s/X^d

@@
 \usepackage{listings}
-\lstset{language=C,mathescape=true,numbers=left,numbersep=0.8ex,tabsize=2,breaklines=true} % basicstyle={\rm\fontfamily{ppl}\selectfont} trop reconnaissable
+\lstset{language=C,mathescape=true,numbers=left,numbersep=0.8ex,tabsize=2,breaklines=true,basicstyle={\rm\fontfamily{phv}\small\selectfont}}
 
@@
 
-To prevent the algorithm from adding too many paths at a time, we
-use another abstract value associated to a control point $p_j$, noted
-$X_j^d$, which is distinct from $X_j^s$, and initialized to $X_j^s$.
-This value is updated when a point $p_j$ is the target of a new path,
-so that further SMT queries do not compute other paths with the same
-source and destination if it is not needed (because these new paths
-would not make $X_j^d$ grow, hence would not be returned by the SMT
-solver).
+Also, to prevent the algorithm from adding too many paths at a time,
+and in particular to prevent the algorithm from trying to re-add the
+same path infinitely many times, we use another abstract value
+associated to a control point $p_j$, noted $X_j^d$, which is distinct
+from $X_j^s$, and initialized to $X_j^s$. This value is updated when a
+point $p_j$ is the target of a new path, and is used instead of
+$X_j^s$ in the SMT formula. This way, further SMT queries do not
+compute other paths with the same source and destination if it is not
+needed (because these new paths would not make $X_j^d$ grow, hence
+would not be returned by the SMT solver).
 
@@
 
+Our tool currently assumes that integer variables are unbounded mathematical integers ($\ZZ$) and floating-point variables are real (or rational) numbers.
+Techniques for sound analysis of bounded integers, including with wraparound, and of floating-point operations have been developed in e.g.
+the Astr\'ee system \citep{ASTREE_ESOP05,ASTREE_PLDI03}, but porting these techniques to our iteration schemes using SMT-solving requires supplemental work.
+
 Our implementation of path-focusing currently does not use true acceleration techniques, as proposed by \citet{Monniaux_Gonnord_SAS11}.
@@
   Analysis} (G), \emph{Path-focused} technique (PF), our combined technique
-  (G+PF), and its version with disjunctive invariants (DIS).}
+  (G+PF), and its version with disjunctive invariants (DIS).
+  The $\subsetneq$ bars gives the percentage of invariants stronger (more precise; smaller with respect to inclusion) with the left-side technique,
+$\supsetneq$ the percentage of invariants stronger with the right-side technique,
+and ``uncomparable'' gives the percentage of invariants that are uncomparable, i.e
+neither greater nor smaller;
+the code points where both invariants are equal make up the remaining percentage. See Tab.~\ref{tab:techniques} for details.}
   \label{fig:techniques}
@@
 \begin{center}
-\setlength{\tabcolsep}{1ex}
+\setlength{\tabcolsep}{0.75ex}
 \begin{tabular}{|l
@@
 over the classic Abstract interpretation algorithm.
-The $\subsetneq$ column gives the percentage of invariants more precise (smaller with respect to inclusion) with the left-side technique,
-$\supsetneq$ the percentage of invariants better with the right-side technique,
-and ``unc.'' gives the percentage of invariants that are uncomparable, i.e
-neither greater nor smaller;
-the code points where both invariants are equal make up the remaining percentage.}
+$\subsetneq$, $\supsetneq$ and ``unc.'' defined as in Fig.~\ref{fig:techniques}.}
 \label{tab:techniques}
@@
 Note that the resulting set is not necessarily minimal, but is sufficient to disconnect all cycles.
-(More sophisticated techniques are discussed in e.g. \citet{BourdonclePhd}).
+(More sophisticated techniques are discussed in e.g. \citet{BourdonclePhd}.)
 
@@
 
-In cases where the formula is expressed in too rich a logic for the SMT-solver to deal with, a number of workarounds are possible:
-\begin{itemize}[itemsep=-\parsep,topsep=0pt,partopsep=0pt]
-\item \emph{Linearization}, as per \citet{DBLP:conf/vmcai/Mine06}, which overapproximates nonlinear semantics by linear semantics. However, such transformations are not implemented in our tool.
-\item Replacing the results of nonlinear operations by ``unknown''.
-\end{itemize}
+In cases where the formula is expressed in too rich a logic for the SMT-solver to deal with, a number of workarounds are possible, including:
+(i) \emph{Linearization}, as per \citet{DBLP:conf/vmcai/Mine06}, which overapproximates nonlinear semantics by linear semantics.
+(ii) Replacing the results of nonlinear operations by ``unknown''.
+Neither is currently implemented in our tool.
 
@@
 techniques. It is interesting to see that \emph{Guided Static Analysis} and
-\emph{Path-focusing} technique are sometimes faster than the classical
-algorithm. Our techniques are slower, but on the same order of magnitude as
-the previous techniques.
+\emph{Path-focusing} are sometimes faster than the classical
+algorithm. The techniques we introduce in this article are slower, but speed is still on the same order of magnitude as the older ones.
 
@@
 Statistically, the domain of convex polyhedra gives the better results, but
-regularly yields greater invariants than the domains of octagons/intervals.
-Experiences shows is it mostly due to the widening operator.
+commonly yields greater invariants than the domains of octagons/intervals;
+this is a known weakness of its widening operator.
+% DM Ici on pourrait citer Monniaux+Le Guen si c'etait publie!
 The Octagon domain appears to be much better than intervals.
@@
 
-Analysis of linear equalities (PKEQ) performs very fast compared
-to other abstract domains, but yields very unprecise invariants.
+The Lapack/Blas benchmarks are unusual compared to the other programs.
+These libraries perform matrix computations, using nested loops over indices; such programs are the prime target for polyhedral loop optimization techniques and it is therefore unsurprising that polyhedra perform very well over them.
+
+The analysis of linear equalities (PKEQ) performs very fast compared
+to other abstract domains, but yields very imprecise invariants:
+it only detects relations of the form $\sum_i a_i x_i = C$ where $a_i$ and $C$ are constants.
 The comparison between the domain of convex polyhedra and linear congruences
@@
 invariant. This is the case for $1.04\%$ of the control points in $P_R$.
+\DM{Je croyais qu'on ne pouvait pas utiliser PPLGRID?}
 

### 2011-10-28 papers/PLDI2012/implicitization.tex
PPLgrid

@@
 
-Analysis of linear equalities (PKEQ) performs very fast compared
-to other abstract domains, but yields very unprecise invariants.
+The analysis of linear equalities (PKEQ) performs very fast compared
+to other abstract domains, but yields very imprecise invariants:
+it only detects relations of the form $\sum a_i x_i = C$ where $a_i$ and $C$ are constants.
 The comparison between the domain of convex polyhedra and linear congruences

### 2011-10-28 papers/PLDI2012/implicitization.tex
on analyse sur des entiers Z et des reels

@@
 to the program, and we unroll every loop once.
+
+Our tool currently assumes that integer variables are unbounded mathematical integers ($\ZZ$) and floating-point variables are real (or rational) numbers.
+Techniques for sound analysis of bounded integers, including with wraparound, and of floating-point operations have been developed in e.g.
+the Astr\'ee system \citep{ASTREE_ESOP05,ASTREE_PLDI03}, but porting these techniques to our iteration schemes using SMT-solving requires supplemental work.
 
@@
 
+The Lapack/Blas benchmarks are unusual compared to the other programs.
+These libraries perform matrix computations, using nested loops over indices; such programs are the prime target for polyhedral loop optimization techniques and it is therefore unsurprising that polyhedra perform very well over them.
+
 The analysis of linear equalities (PKEQ) performs very fast compared

### 2011-10-28 papers/PLDI2012/implicitization.tex
hop

@@
 to other abstract domains, but yields very imprecise invariants:
-it only detects relations of the form $\sum a_i x_i = C$ where $a_i$ and $C$ are constants.
+it only detects relations of the form $\sum_i a_i x_i = C$ where $a_i$ and $C$ are constants.
 The comparison between the domain of convex polyhedra and linear congruences

### 2011-10-28 papers/PLDI2012/implicitization.tex
tassage colonnes
Améliorations mineures et remarques

@@
 l'habitude a la notation $99990[$}
+\MM{Je crois que les anglophones utilisent plutôt $99990)$, mais
+  d'après Wikipedia ``Interval (mathematics)'', les deux sont acceptés
+  par le standard ISO}
 Widening iterations converge to $\lstinline|x_old| \in (-\infty,+\infty)$; let us now see why decreasing iterations fail to recover the desired invariant.
@@
 For a loop header $p_i$, this technique computes a disjunctive invariant
-$\bigvee_{1\leq j \leq m} X_{i,j}$. 
-To do so, one chooses an integer $\delta \in [1,m]$, and
-a mapping function $\sigma: [1,m] \times [1,n] \mapsto [1,m]$, assuming there
-are $n$ distinct paths in the loop.  For each
+$\bigvee_{1\leq j \leq m} X_{i,j}$. We denote by $n$ the number of
+distinct paths in the loop.
+To perform the analysis, one chooses an integer $\delta \in [1,m]$, and
+a mapping function $\sigma: [1,m] \times [1,n] \mapsto [1,m]$.
+\MM{C'est moi ou le $\tau_{i,k}$ ne correspond pas exactement à la
+  notation du début ? Au début, $\tau_{i,k}$ est la transition de $i$
+  à $k$, alors que là, on est dans le multigraphe donc il y a
+  plusieurs arrêtes entre deux points. Si je comprends bien,
+  $\tau_{i,k}$ est le $k$-ième chemin partant de $i$, c'est ça ?
+  Je mettrais bien ``... and for the $k$-th path starting from $i$,
+  denoted $\tau_{i,k}$, the image...'' si c'est bien ça. D'ailleurs,
+  faut-il forcément parler de boucle ici ? La technique ne s'applique
+  pas à des chemins quelconques ?}
+  For each
 abstract value of the disjunctive invariant, and for each path $\tau_{i,k}$ in the loop, the
@@
 We introduce in this formula Boolean variables $\{d_j, 1 \leq j \leq m\}$, so
-that we can easily find in the model which abstract value of the disjunction has
-to be chosen to make the invariant grow.
+that we can easily find in the model which abstract value of the
+disjunction of the source point has
+to be chosen to make the invariant of the destination grow.
 The resulting formula that is given to the SMT solver is defined
@@
 Note that $\sigma(j,k)$ will be computed only if needed.
+
+\MM{Je trouve ça assez frustrant d'arriver ici et de ne toujours pas
+  savoir quel $\sigma$ est choisi. A mon avis, on peut remonter la
+  description du $\sigma$ dynamique de la section suivante à ici. Ça
+  évite de le présenter comme un petit détail d'implémentation alors
+  que c'est quand même une vraie contrib par rapport au papier
+  original.}
 
@@
 \label{sec:analysis-algorithm}
-LLVM bitcode is in \emph{single static assignment} (SSA) form: a given scalar variable is given a value at a single syntactic point in the program.
+LLVM bitcode is in \emph{static single assignment} (SSA) form: a given scalar variable is given a value at a single syntactic point in the program.
 In concrete terms, an assignment \lstinline|x=2*x+1;| gets translated into a definition $x_2 = 2x_1+1$, with distinct variables $x_1$ and $x_2$ corresponding to the same original variable \lstinline|x| at different points in the program.
@@
 
+\MM{cf. ci-dessus, j'aurais mis ça dans la section invariants
+  disjonctifs}
 For the computation of disjunctive invariants, we experimented with a heuristic
@@
 disjunct of index $k'$ so that the least upper bound of the two abstract values is
-exactly the union of them. If such an index exists, then $\sigma(j,k) = k'$.
+exactly the union of them. If such an index exists, then we set $\sigma(j,k) = k'$.
 Otherwise:

### 2011-10-28 papers/PLDI2012/implicitization.tex
deplacement + check anglais du paragraphe sur le choix de sigma

@@
 
-\MM{Je trouve ça assez frustrant d'arriver ici et de ne toujours pas
-  savoir quel $\sigma$ est choisi. A mon avis, on peut remonter la
-  description du $\sigma$ dynamique de la section suivante à ici. Ça
-  évite de le présenter comme un petit détail d'implémentation alors
-  que c'est quand même une vraie contrib par rapport au papier
-  original.}
+We experimented with a heuristic
+of dynamic construction of the $\sigma$ function, adapted from
+\citet{DBLP:conf/pldi/GulwaniZ10}. 
+\DM{Cette heuristique, je croyais que c'était nous qui l'avions trouvée,
+et qu'elle n'était pas dans Gulwani?}
+We start with one single
+disjunct per control point ($m=1$) and an empty map $\sigma$.
+$M$ denotes an upper bound on the number of disjuncts per control point.
+$\sigma(j,k)$ is then constructed on the fly only when needed.
+When $\sigma(j,k)$ needs to be computed, we first compute the image of the
+abstract value $X_k$ by the path indexed by $j$, and try to find an existing
+disjunct of index $k'$ so that the least upper bound of the two abstract values is exactly their union (this can be tested using SMT-solving).
+If such an index exists, then we set $\sigma(j,k) = k'$.
+Otherwise:
+\begin{itemize}[itemsep=-1\parsep,topsep=0pt]
+	\item if $m < M$, we increase $m$ by $1$ and define $\sigma(j,k) = m$
+	\item if $m = M$, we define $\sigma(j,k) = M$
+\end{itemize}
 
@@
 Instead, it simply runs widening and narrowing iterations on a single path.
-
-\MM{cf. ci-dessus, j'aurais mis ça dans la section invariants
-  disjonctifs}
-For the computation of disjunctive invariants, we experimented with a heuristic
-of dynamic construction of the $\sigma$ function, adapted from
-\cite{DBLP:conf/pldi/GulwaniZ10}.  We start with one single
-disjunct per control point ($m=1$), and we start with an empty map $\sigma$. We also
-define a maximal bound on the number of disjunct per control point, noted $M$.
-$\sigma(j,k)$ is then constructed on the fly only when needed.
-When $\sigma(j,k)$ needs to be computed, we first compute the image of the
-abstract value $X_k$ by the path indexed by $j$, and try to find an existing
-disjunct of index $k'$ so that the least upper bound of the two abstract values is
-exactly the union of them. If such an index exists, then we set $\sigma(j,k) = k'$.
-Otherwise:
-\begin{itemize}
-	\item if $m < M$, we increase $m$ by $1$ and define $\sigma(j,k) = m$
-	\item if $m = M$, we define $\sigma(j,k) = M$
-\end{itemize}
 

### 2011-10-28 papers/PLDI2012/implicitization.tex
explain PK terminology

@@
 Our tool operates over LLVM bitcode \citep{LLVM_langref,Lattner:2004:LCF:977395.977673}, which is a target for several compilers, most notably Clang (supporting C and C++) and llvm-gcc (supporting C, C++, Fortran and Ada).
-Abstract domains are provided by the APRON library \citep{DBLP:conf/cav/JeannetM09}, and include convex polyhedra, octagons, intervals, and linear congruences.
+Abstract domains are provided by the APRON library \citep{DBLP:conf/cav/JeannetM09}, and include convex polyhedra (from the builtin Polka ``PK'' library), octagons, intervals, and linear congruences.
 For SMT-solving, our analyzer uses Yices
@@
 
-Our analysis tool currently only operates over scalar variables from the SSA representation and thus cannot directly cope with arrays or memory accessed through pointers.
+Our tool currently only operates over scalar variables from the SSA representation and thus cannot directly cope with arrays or memory accessed through pointers.
 We therefore run it after the ``memory to registers'' (\texttt{mem2reg}) optimization phase in LLVM, which lifts most memory accesses to scalar variables.

### 2011-10-28 papers/PLDI2012/implicitization.tex
centering column headers
parameters -> octagons

@@
 \begin{tabular}{|l|r|r|r|r|r|} \hline
-	Benchmark & \textbf{S} & \textbf{G} & \textbf{PF} & \textbf{G+PF} &
-	\textbf{DIS} \\ \hline
+	\multicolumn{1}{|c|}{Benchmark} &
+        \multicolumn{1}{c|}{\textbf{S}} &
+        \multicolumn{1}{c|}{\textbf{G}} &
+        \multicolumn{1}{c|}{\textbf{PF}} &
+        \multicolumn{1}{c|}{\textbf{G+PF}} &
+	\multicolumn{1}{c|}{\textbf{DIS}} \\ \hline
 	\input{table_time.tex} \hline
@@
 this is a known weakness of its widening operator.
-% DM Ici on pourrait citer Monniaux+Le Guen si c'etait publie!
-The Octagon domain appears to be much better than intervals.
+\DM{Ici on pourrait citer Monniaux+Le Guen si c'etait publie!}
+The Octagon domain appears to be much better than intervals; this is
+unsurprising since in most programs and libraries, bounds on loop indices are
+non constant: they depend on some parameters (array sizes etc.).
 \JH{Commenter les resultats}

### 2011-10-28 papers/PLDI2012/implicitization.tex
phi-variables

@@
 
-\titlebanner{banner above paper title}        % These are ignored unless
-\preprintfooter{short description of paper}   % 'preprint' option specified.
-
-\title{Implicit yaddayadda}
-\subtitle{Subtitle Text, if any}
+%\titlebanner{banner above paper title}        % These are ignored unless
+\preprintfooter{Succinct representations for abstract interpretation}   % 'preprint' option specified.
+
+\title{Succinct representations for abstract interpretation}
+%\subtitle{Subtitle Text, if any}
 
 \authorinfo{Anonymous}
-           {anonymous}
+           {The Internet}
            {anonymous.gal@gmail.com}
@@
 \begin{abstract}
-This is the text of the abstract.
+Abstract interpretation techniques can be made more precise by distinguishing paths inside loops, at the expensive of possibly exponential complexity.
+SMT-solving techniques and sparse representations of paths and sets of paths avoid this pitfall.
+
+We improve previously proposed techniques for guided static analysis and the generation of disjunctive invariants by combining them with techniques for succinct representations of paths and symbolic representations for transitions based on static single assignment.
+
+Because of the non-monotonicity of the results of abstract interpretation with widening operators, it is difficult to conclude that some abstraction is more precise than another based on theoretical local precision results.
+We thus conducted extensive comparisons between our new techniques and previous ones, on a variety of open-source packages.
+We also took the opportunity to compare the precision of various classical abstract domains (convex polyhedra, octagons, intervals).
 \end{abstract}
@@
 for instance, for \lstinline|y=x; z=x-y;| with precondition $x \in [0,1]$, a line-by-line interval analysis obtains $y \in [0,1]$ and $z \in [-1,1]$ while our ``en bloc'' analysis symbolically simplifies $z = x - x = 0$ and thus $z \in [0,0]$.
+
+\DM{Julien: pour la guided analysis, a priori ça arrive souvent cette histoire d'edge pas prise car non encore prenable?}
+In the event that a node is reachable only by a single control-flow edge (which may occur because of dead code, or during the first phases of guided static analysis), the $\phi$ operation reduces to a copy of the values flowing from that edge.
+In this case, our tool just propagates symbolic values through the predecessor node, without introducing $\phi$-variables.
 

### 2011-10-28 papers/PLDI2012/implicitization.tex
SSA in intro

@@
 With respect to abstract domains on numerical variables, it uses the APRON library \cite{DBLP:conf/cav/JeannetM09}, which supports a variety of abstract domains from which we can choose with minimal changes to our analyzer.
-Section~\ref{sec:analysis-algorithm} summarizes our implementation decisions for analysis of SSA-form programs.
+
+Our tool uses a \emph{single static assignment} (SSA) intermediate representation, which allows us to perform cheap symbolic propagations and simplifications,
+described in Section~\ref{sec:analysis-algorithm}.
 

### 2011-10-28 papers/PLDI2012/implicitization.tex
abstract + conclusion

@@
 Instead, it simply runs widening and narrowing iterations on a single path.
+The analysis is currently only forward, even though nothing in what we describe in this article is specific to forward analysis.
 
@@
 
-\section{Conclusion}
+\section{Conclusion and future prospects}
+Roughly, an analysis by abstract interpretation is defined by the choice of an iteration strategy and an abstract domain.
+In this article, we demonstrated that changes in the iteration algorithm can significantly improve precision, sometimes while improving analysis times.
+
+A common criticism of analysis techniques based on SMT-solving is that they
+do not scale up. Yet, our experiments show that, for numerical properties,
+they scale up to the size of typical functions and loops.
+It is however
+quite certain that, naively applied, they cannot scale to the kind of
+programs targeted by e.g. the Astr\'ee tool, that is, a dozens or hundreds
+of thousands of lines of code in a single loop operating over similar numbers
+of remanent variables.
+Actually, for such applications, only (quasi-)linear
+algorithms scale up, and ``cheap'' abstract domains such as octagons ($O(n^3)$ where $n$ is the number of variables) are not applied to the full variable set, but to restricted subsets thereof.
+It thus seems reasonable that techniques such as considering ``packs'' of related variables, slicing, etc. may similarly help SMT-based techniques to scale to global analyses.
+
+We compared the precision of different techniques and abstract domains by comparing the invariants for the inclusion ordering.
+A better metric is perhaps to take a client analysis --- such as the detection of overflows and array bound violations --- and compare the rates of alarms.
 
@@
 
-\DM{Garde-t-on la these en francais de Nicolas?}
-
 \phantomsection\addcontentsline{toc}{section}{References} 

### 2011-10-28 papers/PLDI2012/implicitization.tex
conclusion bof

@@
 
+We focused on numerical properties, because they are supported by easily available abstract libraries.
+Yet, in most programs, properties of data structures are important for proving interesting properties.
+Further investigations are needed not only on good abstractions for pointers (many are already known) but also on their conversion to SMT problems.
+
 % Cas a probleme:

### 2011-10-29 papers/PLDI2012/implicitization.tex

@@
 et qu'elle n'était pas dans Gulwani?}
+\JH{Non. Gulwani fait la même chose en un peu plus complique (voir commentaire
+suivant)}
 We start with one single
@@
 	\item if $m = M$, we define $\sigma(j,k) = M$ 
+		\JH{A la place, Gulwani cherche l'invariant qui a le plus de contraintes
+		d'egalite en commun pour definir son $sigma$. Nous, on le met dans M, pour simplifier}
 \end{itemize}
@@
 \DM{Julien: pour la guided analysis, a priori ça arrive souvent cette histoire d'edge pas prise car non encore prenable?}
+\JH{Oui, ca arrive regulierement}
 In the event that a node is reachable only by a single control-flow edge (which may occur because of dead code, or during the first phases of guided static analysis), the $\phi$ operation reduces to a copy of the values flowing from that edge.
@@
 non constant: they depend on some parameters (array sizes etc.).
-\JH{Commenter les resultats}
 
@@
 \DM{Je croyais qu'on ne pouvait pas utiliser PPLGRID?}
+\JH{On peut. C'est PKGRID qui ne fonctionne pas}
 

### 2011-10-31 papers/PLDI2012/implicitization.tex
macros pour desactiver commentaires
notations (correction des indices)

@@
 \newcommand{\DM}[1]{{\color{violet} TODO(DM): #1}}
+%\newcommand{\MM}[1]{}
+%\newcommand{\JH}[1]{}
+%\newcommand{\DM}[1]{}
+
 % Pour les commentaires
@@
 
-\JH{$99990)$, ca designe bien une borne d'intervalle ouvert ? J'ai plus
-l'habitude a la notation $99990[$}
-\MM{Je crois que les anglophones utilisent plutôt $99990)$, mais
-  d'après Wikipedia ``Interval (mathematics)'', les deux sont acceptés
-  par le standard ISO}
 Widening iterations converge to $\lstinline|x_old| \in (-\infty,+\infty)$; let us now see why decreasing iterations fail to recover the desired invariant.
@@
 
-For a loop header $p_i$, this technique computes a disjunctive invariant
-$\bigvee_{1\leq j \leq m} X_{i,j}$. We denote by $n$ the number of
-distinct paths in the loop.
-To perform the analysis, one chooses an integer $\delta \in [1,m]$, and
-a mapping function $\sigma: [1,m] \times [1,n] \mapsto [1,m]$.
+For each control point $p_i$, we compute a disjunctive invariant
+$\bigvee_{1\leq j \leq m_i} X_{i,j}$. We denote by $n_i$ the number of
+distinct paths starting at $p_i$.
+To perform the analysis, one chooses an integer $\delta_i \in [1,m_i]$, and
+a mapping function $\sigma_i: [1,m_i] \times [1,n_i] \mapsto [1,m_i]$.
 \MM{C'est moi ou le $\tau_{i,k}$ ne correspond pas exactement à la
@@
   pas à des chemins quelconques ?}
-  For each
-abstract value of the disjunctive invariant, and for each path $\tau_{i,k}$ in the loop, the
-image of $X_{i,j}$ by the path is joined with
+  The $k$-th path starting fom $p_i$ is denoted $\tau_{i,k}$.
+  The image of the $j$-th disjunct $X_{i,j}$ by the path $\tau_{i,k}$ is joined with 
   $X_{i,\sigma(j,k)}$.
-Initially, the $\delta$-th abstract value contains the initial states of
+Initially, the $\delta_i$-th abstract value contains the initial states of
 $p_i$, and all other abstract values contain~$\emptyset$.
 
-$m$, $\delta$ and $\sigma$ can be defined heuristically.
-For instance, one could define $\sigma$ so that $\sigma(j,k)$ only depends on the
+For each control point $p_i \in P_R$, $m_i$, $\delta_i$ and $\sigma_i$ can be defined heuristically.
+For instance, one could define $\sigma_i$ so that $\sigma_i(j,k)$ only depends on the
 last transition of the path, or else construct it dynamically during the
@@
 $d_j$ is \emph{true} in the model. Then, we can easily compute the value of 
-$\sigma(j,k)$, thus know the index of the disjunct to join with.
+$\sigma_i(j,k)$, thus know the index of the disjunct to join with.
 
@@
 
-The \emph{Update} function can classically assign to $X_{i,\sigma(j,k)}$ the
-value $X_{i,\sigma(j,k)} \widening (X_{i,\sigma(j,k)} \sqcup
+The \emph{Update} function can classically assign to $X_{i,\sigma_i(j,k)}$ the
+value $X_{i,\sigma_i(j,k)} \widening (X_{i,\sigma_i(j,k)} \sqcup
 \tau_{i,k}(X_{i,j}))$, or can integrate the special treatment for self loops
@@
 	\STATE Take $j \in \{ l | d_l = true\}$ 
-	\STATE Update($X_{i,\sigma(j,k)})$
+	\STATE Update($X_{i,\sigma_i(j,k)})$
 \ENDWHILE
@@
 
-The $\sigma$ function takes as parameters the index of the starting abstract
-value, and the path we focus on. If we dynamically construct this function
-during the analysis, we should store the results of the $\sigma$ function into a
+The $\sigma_i$ functions take as parameters the index of the starting abstract
+value, and the path we focus on. If we dynamically construct these functions
+during the analysis, we should store the results of the $\sigma_i$ functions into a
 compact representation, such as Algebraic Decision Diagrams.
-Note that $\sigma(j,k)$ will be computed only if needed.
+Note that $\sigma_i(j,k)$ will be computed only if needed.
 
 We experimented with a heuristic
-of dynamic construction of the $\sigma$ function, adapted from
+of dynamic construction of the $\sigma_i$ functions, adapted from
 \citet{DBLP:conf/pldi/GulwaniZ10}. 
@@
 We start with one single
-disjunct per control point ($m=1$) and an empty map $\sigma$.
+disjunct per control point ($m_i=1$ for each $i, p_i \in P_R$) 
+and empty maps $\sigma_i$.
 $M$ denotes an upper bound on the number of disjuncts per control point.
-$\sigma(j,k)$ is then constructed on the fly only when needed.
-When $\sigma(j,k)$ needs to be computed, we first compute the image of the
-abstract value $X_k$ by the path indexed by $j$, and try to find an existing
-disjunct of index $k'$ so that the least upper bound of the two abstract values is exactly their union (this can be tested using SMT-solving).
-If such an index exists, then we set $\sigma(j,k) = k'$.
+$\sigma_i(j,k)$ is then constructed on the fly only when needed.
+When $\sigma_i(j,k)$ needs to be computed, we first compute the image of the
+abstract value $X_{i,j}$ by the path indexed by $k$, and try to find an existing
+disjunct of index $j'$ so that the least upper bound of the two abstract values is exactly their union (this can be tested using SMT-solving).
+If such an index exists, then we set $\sigma_i(j,k) = j'$.
 Otherwise:
 \begin{itemize}[itemsep=-1\parsep,topsep=0pt]
-	\item if $m < M$, we increase $m$ by $1$ and define $\sigma(j,k) = m$
-	\item if $m = M$, we define $\sigma(j,k) = M$ 
+	\item if $m_i < M$, we increase $m_i$ by $1$ and define $\sigma_i(j,k) = m_i$
+	\item if $m_i = M$, we define $\sigma_i(j,k) = M$ 
 		\JH{A la place, Gulwani cherche l'invariant qui a le plus de contraintes

### 2011-10-31 papers/PLDI2012/implicitization.tex
modif remarque Matthieu

@@
   pas à des chemins quelconques ?}
+  \JH{J'ai corrige\dots}
   The $k$-th path starting fom $p_i$ is denoted $\tau_{i,k}$.
-  The image of the $j$-th disjunct $X_{i,j}$ by the path $\tau_{i,k}$ is joined with 
-  $X_{i,\sigma(j,k)}$.
+  The image of the $j$-th disjunct $X_{i,j}$ by the path $\tau_{i,k}$ is then
+  joined with $X_{i,\sigma(j,k)}$.
 Initially, the $\delta_i$-th abstract value contains the initial states of

### 2011-10-31 papers/PLDI2012/implicitization.tex
indices
ajout citation

--- 2011-10-29 papers/PLDI2012/implicitization.tex
@@
 \emph{Guided static analysis} was proposed by \citet{DBLP:conf/sas/GopanR07} as an improvement over classical upward Kleene iterations with widening.
-Consider the following program:
+Consider the following program, taken from \cite{DBLP:conf/sas/GopanR07}:
 \lstinputlisting[label=lst:gopan_reps]{gopan_reps.c}

### 2011-10-31 papers/PLDI2012/implicitization.tex
typos

--- 2011-10-31 papers/PLDI2012/implicitization.tex
@@
 
-Guided static analysis, as proposed by \citet{DBLP:conf/sas/GopanR07}, applied to the transition
+Guided static analysis, as proposed by \citet{DBLP:conf/sas/GopanR07}, applies to the transition
 graph of the program. We now present a new technique applying this analysis on the implicit
@@
   and corresponds to lines in~\ref{alg=start-add-paths} to
-  \ref{alg=end-add-paths} algorithm~\ref{algo:combined}.
+  \ref{alg=end-add-paths} in Algorithm~\ref{algo:combined}.
 \item \label{step:ascending} Given a new subset $P$, we search for paths starting at point
@@
   lines~\ref{alg=start-ascending} to~\ref{alg=end-ascending} in
-  algorithm~\ref{algo:combined}.
+  Algorithm~\ref{algo:combined}.
 \item  \label{step:narrowing} We perform narrowing iterations the usual way
@@
 
-The main loop in the expanded multigraph has then 24 distinct paths, half of
+The main loop in the expanded multigraph has then 8 distinct paths, half of
 them being self loops, and the other half going to the header of the nested loop.
@@
 		Additionally, the path that goes from $p_5$ to
-		$p_11$, that goes through the \emph{else} branch of each 
+		$p_{10}$, that goes through the \emph{else} branch of each 
 		\emph{if-then-else} at lines 7,8 and 9, is also 
 		$-10 \leq \lstinline|x_old| \leq 10$, and thus is also added into the
-		new subgraph. Again, there is no new path that goes to $p_11$ with an
+		new subgraph. Again, there is no new path that goes to $p_{10}$ with an
 		image outside $-10 \leq \lstinline|x_old| \leq 10$.
@@
 		Step 2). When we have reached an inductive invariant, we narrow, and the
-		result for both $p_5$ and $p_11$ is the polyhedron 
+		result for both $p_5$ and $p_{10}$ is the polyhedron 
 		$-10000 \leq \lstinline|x_old| \leq 10000$.
@@
 $\lstinline|x_old| \in~[-10000,\allowbreak 10000]$. 
-Here, only 3 paths out of the 25 have been computed during the analysis. In
+Here, only 3 paths out of the 8 have been computed during the analysis. In
 practice, depending on the order the SMT-solver returns the paths, other
@@
 	\node[PRstate] (n0) [below of=n00, yshift=0.8cm] {$p_5$};
-	\node[PRstate] (n1) [below of=n0] {$p_{11}$};
+	\node[PRstate] (n1) [below of=n0] {$p_{10}$};
 	\node (label) [below of=n1, yshift=1.2cm] {Step 1};
@@
 	\node[PRstate] (n0) [below of=n00, yshift=0.8cm] {$p_5$};
-	\node[PRstate] (n1) [below of=n0] {$p_{11}$};
+	\node[PRstate] (n1) [below of=n0] {$p_{10}$};
 	\node (label) [below of=n1, yshift=1.2cm] {Step 2};
@@
 \begin{algorithmic}[1] 
-\FORALL {$k \in \{1,..,m\} \setminus \{\delta\}$}
+\FORALL {$k \in \{1,..,m_i\} \setminus \{\delta_i\}$}
 	\STATE $X_{i,k} \gets \perp$
 \ENDFOR
-\STATE $X_{i,\delta} \gets I_{p_i}$
+\STATE $X_{i,\delta_i} \gets I_{p_i}$
 \end{algorithmic}

### 2011-11-01 papers/PLDI2012/implicitization.tex
X^d et X^s
typos + retrait TODO

@@
 \displaystyle\bigwedge_{j \in P_R \atop j \neq i} \neg
-b_j^s \wedge x \in X_i^s \wedge \displaystyle\bigvee_{j \in Succ(i)} (b_j^d \wedge
-x \notin X_j^s)$$
+b_j^s \wedge X_i^s \wedge \displaystyle\bigvee_{j \in Succ(i)} (b_j^d \wedge
+ \neg X_j^s)$$
 
@@
 point $p_j$ is the target of a new path, and is used instead of
-$X_j^s$ in the SMT formula. This way, further SMT queries do not
+$X_j^s$ in the SMT formula. We thus check the satisfiability of the formula
+$f'(p_i)$, where:
+$$f'(p_i) = \rho \wedge b_i^s \wedge 
+\displaystyle\bigwedge_{j \in P_R \atop j \neq i} \neg
+b_j^s \wedge X_i^s \wedge \displaystyle\bigvee_{j \in Succ(i)} (b_j^d \wedge
+\neg X_j^d)$$
+
+This way, further SMT queries do not
 compute other paths with the same source and destination if it is not
@@
 a mapping function $\sigma_i: [1,m_i] \times [1,n_i] \mapsto [1,m_i]$.
-\MM{C'est moi ou le $\tau_{i,k}$ ne correspond pas exactement à la
-  notation du début ? Au début, $\tau_{i,k}$ est la transition de $i$
-  à $k$, alors que là, on est dans le multigraphe donc il y a
-  plusieurs arrêtes entre deux points. Si je comprends bien,
-  $\tau_{i,k}$ est le $k$-ième chemin partant de $i$, c'est ça ?
-  Je mettrais bien ``... and for the $k$-th path starting from $i$,
-  denoted $\tau_{i,k}$, the image...'' si c'est bien ça. D'ailleurs,
-  faut-il forcément parler de boucle ici ? La technique ne s'applique
-  pas à des chemins quelconques ?}
-  \JH{J'ai corrige\dots}
   The $k$-th path starting fom $p_i$ is denoted $\tau_{i,k}$.
@@
  & \wedge 
-\displaystyle\bigvee_{1 \leq k \leq m} (d_k \wedge x \in X_{i,k} \wedge \bigwedge_{l \neq k}
+\displaystyle\bigvee_{1 \leq k \leq m_i} (d_k \wedge X_{i,k} \wedge \bigwedge_{l \neq k}
 \neg d_l) \\
@@
 \displaystyle\bigvee_{j \in Succ(i)} 
-(b_j^d \wedge \bigwedge_{1 \leq k \leq m} (x \notin X_{j,k}))
+(b_j^d \wedge \bigwedge_{1 \leq k \leq m_i} (\neg X_{j,k}))
 \end{eqnarray*}
@@
 invariant. This is the case for $1.04\%$ of the control points in $P_R$.
-\DM{Je croyais qu'on ne pouvait pas utiliser PPLGRID?}
-\JH{On peut. C'est PKGRID qui ne fonctionne pas}
 

### 2011-11-01 papers/PLDI2012/implicitization.tex
commentaires resultats

@@
 \emph{Guided Static Analysis} from \citet{DBLP:conf/sas/GopanR07} improves the
-result of the classical Abstract Interpretation in $2.33\%$ of the control points
+result of the classical Abstract Interpretation in $2.72\%$ of the control points
 in $P_R$.
@@
 statistically better results, and finds better invariants than \emph{Guided
-Static Analysis} in $8.58\%$ of the cases. However, it also loses precision in an
-important number ($2.87\%$) of control points.
+Static Analysis} in $9.89\%$ of the cases. However, it also loses precision in an
+important number ($2.68\%$) of control points.
 Finally, our combined technique gives the most promising results, since it is
@@
 
-\JH{Paragraphe a reecrire:}
-The analysis using disjunctive invariants seems not to be as efficient as
-expected, and often finds greater invariants than the combined technique
-($11.64\%$). It probably means that our heuristic for the construction of
-$\sigma$ is not effective enough. Usually, disjunctive invariants computation
-finds precise loop invariants because they allow to distinguish for
-instance the first loop iterations. In our case, such benefits are not
-visible, since we have already unrolled every loops once.
+The analysis using disjunctive invariants greatly improves the
+precision of the analysis (for $12.06\%$ of the control points in $P_R$), at an
+acceptable time cost (see Table \ref{tab:time}).
 

### 2011-11-01 papers/PLDI2012/implicitization.tex
Resultats d'experience Bagnara

@@
 & \multicolumn{3}{c|}{\textbf{PK/PKEQ}}
-& \multicolumn{3}{c|}{\textbf{OCT/PKEQ}}
-& \multicolumn{3}{c|}{\textbf{PK/GRID}} \\ %\cline{2-16}
+& \multicolumn{3}{c|}{\textbf{PK/GRID}}
+& \multicolumn{3}{c|}{\textbf{POLY*/POLY}} \\ %\cline{2-16}
 & \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
@@
 Octagons (OCT), intervals (BOX), linear equalities (PKEQ) and linear congruences
-(GRID).}
+(GRID). The last column compares the domain of Convex Polyhedra with the
+improved widening operator from \citet{BagnaraHRZ05SCP} (POLY*), and Convex Polyhedra
+using the classical widening operator (POLY).}
 \label{tab:domain}
@@
 invariant. This is the case for $1.04\%$ of the control points in $P_R$.
+
+Finally, we evaluated the benefits of the improved version of the widening
+operator for convex polyhedra from \citet{BagnaraHRZ05SCP}, compared to the
+classical widening. We found that the improved version from
+\citet{BagnaraHRZ05SCP} yields more precise invariants for $6.58\%$ of the
+control points in $P_R$.
 

### 2011-11-01 papers/PLDI2012/implicitization.tex
Remarques et typos sur les dernières pages

@@
 
+\MM{Et que prend-t-on pour $\delta$ ? C'est 1 tout le temps ?}
+
 \section{Experimental comparisons}
@@
 Static Analysis} \cite{DBLP:conf/sas/GopanR07}.
+
+\MM{On peut peut-être citer le nom de l'outil (on garde PAGAI ?) ici.
+  L'intérêt d'avoir un nom, c'est que les gens peuvent en parler plus
+  facilement, et c'est plus facile de retrouver le papier origine d'un
+  outil quand il cite son nom}
 
@@
 For instance, for \lstinline|if (...) { x = 2*x+1; } else { x= 0; }|, then $x_2$ is defined as $\phi(2x_1+1,0)$.
-In this framework, each variable is uniquely defined as an arithmetic ($+$, $-$, $\times$, $/$) function of other variables that themselves cannot be represented as affine linear functions, because they are defined using $\phi$-functions, bitwise arithmetic, loads from memory, or return values from function calls.
+In this framework, each variable is uniquely defined as an arithmetic ($+$, $-$, $\times$, $/$) function of other variables that themselves may not be representable as affine linear functions, because they are defined using $\phi$-functions, bitwise arithmetic, loads from memory, or return values from function calls.
 
@@
 \begin{figure}[h]
+  \MM{Je trouve que les zéro-virgule-quelque-chose comme .17 font
+    moches à côté des zéros, vu que les tableaux sont alignés sur le
+    ., les ``0'' sont tout à gauche et les ``.17'' tout à droite. Je
+    vote pour des 0.17 pour que l'alignement soit bon, ça rendrait la
+    lecture plus facile je trouve}
   \begin{center}
@@
 \item For each program, we distinguished a set $P_R = P_W$ of suitable widening points by a simple algorithm: for each procedure, we compute the strongly connected components of its control-flow graph using Tarjan's algorithm; the targets of the back-edges of the depth-first search are added to~$P_R$.
-Note that the resulting set is not necessarily minimal, but is sufficient to disconnect all cycles.
-(More sophisticated techniques are discussed in e.g. \citet{BourdonclePhd}.)
+Note that the resulting set is not necessarily minimal, but is sufficient to disconnect all cycles
+(More sophisticated techniques are discussed in e.g. \citet{BourdonclePhd}).
+\MM{Pourquoi cet item se trouve dans la section ``Precision of the
+  various techniques'' et pas dans la section précédente sur
+  l'implem ?}
 
@@
 \emph{Path-focusing} are sometimes faster than the classical
-algorithm. The techniques we introduce in this article are slower, but speed is still on the same order of magnitude as the older ones.
+algorithm.
+\MM{Si y'a la place, on peut dire un truc du style: ``This can be
+  explained by the fact that these algorithms start on a small subset
+  of the graph, and then may need fewer iterations on the complete
+  graph.''}
+The techniques we introduce in this article are slower, but speed is still on the same order of magnitude as the older ones.
 
@@
 Statistically, the domain of convex polyhedra gives the better results, but
-commonly yields greater invariants than the domains of octagons/intervals;
+commonly yields weaker invariants than the domains of octagons/intervals;
 this is a known weakness of its widening operator.
 \DM{Ici on pourrait citer Monniaux+Le Guen si c'etait publie!}
+\MM{Oui, ça fait un peu [citation needed] comme phrase ;-)}
 The Octagon domain appears to be much better than intervals; this is

### 2011-11-01 papers/PLDI2012/implicitization.tex
Encore des remarques.

@@
 
-\MM{À changer si on décide de ne plus distinguer $X^s$ et $X^d$}
-
+% À changer si on décide de ne plus distinguer $X^s$ et $X^d$
 Also, to prevent the algorithm from adding too many paths at a time,
@@
 during the analysis, we should store the results of the $\sigma_i$ functions into a
-compact representation, such as Algebraic Decision Diagrams.
+compact representation, such as Algebraic Decision Diagrams.\MM{Du
+  coup, j'aurais intégré cette phrase au paragraphe suivant. C'est des
+  ADD que tu utilises finalement ?}
 Note that $\sigma_i(j,k)$ will be computed only if needed.
@@
 suivant)}
+\MM{Du coup, la contribution~\ref{contr:disjunctive} devrait être
+  affaiblie un peu je suppose. Gulwani fait aussi du SMT-solving ?}
 We start with one single

### 2011-11-01 papers/PLDI2012/implicitization.tex
deplacement paragraphe calcul de P_R

@@
 \label{sec:analysis-algorithm}
+For each program, we distinguish a set $P_R = P_W$ of suitable widening points by a simple algorithm: for each procedure, compute the strongly connected components of its control-flow graph using Tarjan's algorithm; the targets of the back-edges of the depth-first search are added to~$P_R$.
+Note that the resulting set is not necessarily minimal, but is sufficient to disconnect all cycles --- more sophisticated techniques are discussed in e.g.
+\citet{BourdonclePhd}.
+
 LLVM bitcode is in \emph{static single assignment} (SSA) form: a given scalar variable is given a value at a single syntactic point in the program.
@@
 
-We compared the precision of the different techniques described in this article as follows:
-\begin{itemize}
-\item For each program, we distinguished a set $P_R = P_W$ of suitable widening points by a simple algorithm: for each procedure, we compute the strongly connected components of its control-flow graph using Tarjan's algorithm; the targets of the back-edges of the depth-first search are added to~$P_R$.
-Note that the resulting set is not necessarily minimal, but is sufficient to disconnect all cycles
-(More sophisticated techniques are discussed in e.g. \citet{BourdonclePhd}).
-\MM{Pourquoi cet item se trouve dans la section ``Precision of the
-  various techniques'' et pas dans la section précédente sur
-  l'implem ?}
-
-\item For each program and each pair $(T_1,T_2)$ of analysis techniques, we list
+For each program and each pair $(T_1,T_2)$ of analysis techniques, we list
 the proportion of control points in $P_R$ where $T_1$ (resp.~$T_2$) gives a
@@
 invariant). We use convex polyhedra as the abstract domain.
-\end{itemize}
 

### 2011-11-01 papers/PLDI2012/implicitization.tex
traitement de quelques remarques

@@
 
-\DM{Julien: pour la guided analysis, a priori ça arrive souvent cette histoire d'edge pas prise car non encore prenable?}
-\JH{Oui, ca arrive regulierement}
 In the event that a node is reachable only by a single control-flow edge (which may occur because of dead code, or during the first phases of guided static analysis), the $\phi$ operation reduces to a copy of the values flowing from that edge.
@@
 algorithm.
-\MM{Si y'a la place, on peut dire un truc du style: ``This can be
-  explained by the fact that these algorithms start on a small subset
-  of the graph, and then may need fewer iterations on the complete
-  graph.''}
+This seems due to the fact that these algorithms start on a small subset
+of the graph, and then may need fewer iterations on the complete graph.
 The techniques we introduce in this article are slower, but speed is still on the same order of magnitude as the older ones.
@@
 commonly yields weaker invariants than the domains of octagons/intervals;
-this is a known weakness of its widening operator.
-\DM{Ici on pourrait citer Monniaux+Le Guen si c'etait publie!}
-\MM{Oui, ça fait un peu [citation needed] comme phrase ;-)}
+this is a known weakness of its widening operator~\cite{Monniaux_LeGuen2011}.
 The Octagon domain appears to be much better than intervals; this is

### 2011-11-01 papers/PLDI2012/implicitization.tex
reponse a une question

@@
   affaiblie un peu je suppose. Gulwani fait aussi du SMT-solving ?}
+\DM{Non, Gulwani énumère tous les chemins un par un à tous les coups.}
 We start with one single

### 2011-11-02 papers/PLDI2012/implicitization.tex
+rq sur anonymat
subtitle
Oops, une remarque qui était mal placée
mise en forme de 'g' et fix commentaire
ajout de 0 devant les . dans les tableaux

@@
 \title{Succinct representations for abstract interpretation}
-%\subtitle{Subtitle Text, if any}
+\subtitle{Combined analysis algorithms and experimental evaluation}
 
@@
 \begin{eqnarray*}
-g(p_i) = & \rho \wedge b_i^s \wedge 
+g(p_i) & = & \rho \wedge b_i^s \wedge 
 \displaystyle\bigwedge_{j \in P_R \atop j \neq i} \neg b_j^s  \\
- & \wedge 
+ & & \wedge 
 \displaystyle\bigvee_{1 \leq k \leq m_i} (d_k \wedge X_{i,k} \wedge \bigwedge_{l \neq k}
 \neg d_l) \\
- & \wedge
+ & & \wedge
 \displaystyle\bigvee_{j \in Succ(i)} 
@@
 disjunct per control point ($m_i=1$ for each $i, p_i \in P_R$) 
-and empty maps $\sigma_i$.
+and empty maps $\sigma_i$. We also define $\delta_i = 1$ for each $i, p_i \in
+P_R$.
 $M$ denotes an upper bound on the number of disjuncts per control point.
@@
 
-\MM{Et que prend-t-on pour $\delta$ ? C'est 1 tout le temps ?}
-
 \section{Experimental comparisons}
@@
   outil quand il cite son nom}
+\DM{Dans la version finale, oui, mais pas dans une soumission anonymisée... suffit de Googlifier pagai sur imag.fr}
 
@@
 \begin{figure}[h]
-  \MM{Je trouve que les zéro-virgule-quelque-chose comme .17 font
-    moches à côté des zéros, vu que les tableaux sont alignés sur le
-    ., les ``0'' sont tout à gauche et les ``.17'' tout à droite. Je
-    vote pour des 0.17 pour que l'alignement soit bon, ça rendrait la
-    lecture plus facile je trouve}
   \begin{center}

### 2011-11-02 papers/PLDI2012/implicitization.tex
mauvaise citation

@@
 this is a known weakness of its widening operator~\cite{Monniaux_LeGuen2011}. 
+\JH{la citation n'est pas bonne. On tombe sur ``Automatic Modular Abstractions for Template Numerical Constraints''}
 The Octagon domain appears to be much better than intervals; this is

### 2011-11-02 papers/PLDI2012/implicitization.tex
Rappel de la contrib sur les invariants disjonctifs

@@
 \DM{Non, Gulwani énumère tous les chemins un par un à tous les coups.}
+\MM{OK, du coup j'ai ajouté une phrase à la fin de la section pour
+  rappeler notre contrib}
 We start with one single
@@
 \end{itemize}
+The main difference with the original
+algorithm~\cite{DBLP:conf/pldi/GulwaniZ10} is that we construct
+$\sigma_i(j,k)$ using SMT queries instead of enumerating a possibly
+exponential number of paths to find a solution.
 

### 2011-11-02 papers/PLDI2012/implicitization.tex
suppression commentaires + correction bibtex

--- 2011-11-02 papers/PLDI2012/implicitization.tex
@@
 Algorithm~ \ref{algo:combined} performs Guided static analysis on the implicitly represented multigraph.
-
-\MM{Il faudra peut-être une section ``détails d'implem'' où mettre
-  l'histoire des BDDs pour P. Je laisse ici faute de mieux (peut-être
-  que la section~\ref{sec:analysis-algorithm} ferait l'affaire ?),
-  mais ça distrait un peu le lecteur}
 
@@
 
-\MM{Pas sûr que ça soit nécessaire, mais vu qu'on a eu une version
-  fausse avant, c'est peut-être bien de dire comment on est arrivés à
-  la meilleure version.}
 The order of steps is important: narrowing has to be performed before
@@
 
-The $\sigma_i$ functions take as parameters the index of the starting abstract
-value, and the path we focus on. If we dynamically construct these functions
-during the analysis, we should store the results of the $\sigma_i$ functions into a
-compact representation, such as Algebraic Decision Diagrams.\MM{Du
-  coup, j'aurais intégré cette phrase au paragraphe suivant. C'est des
-  ADD que tu utilises finalement ?}
-Note that $\sigma_i(j,k)$ will be computed only if needed.
-
 We experimented with a heuristic
@@
 \citet{DBLP:conf/pldi/GulwaniZ10}. 
-\DM{Cette heuristique, je croyais que c'était nous qui l'avions trouvée,
-et qu'elle n'était pas dans Gulwani?}
-\JH{Non. Gulwani fait la même chose en un peu plus complique (voir commentaire
-suivant)}
-\MM{Du coup, la contribution~\ref{contr:disjunctive} devrait être
-  affaiblie un peu je suppose. Gulwani fait aussi du SMT-solving ?}
-\DM{Non, Gulwani énumère tous les chemins un par un à tous les coups.}
-We start with one single
-disjunct per control point ($m_i=1$ for each $i, p_i \in P_R$) 
-and empty maps $\sigma_i$. We also define $\delta_i = 1$ for each $i, p_i \in
-P_R$.
+%\DM{Cette heuristique, je croyais que c'était nous qui l'avions trouvée,
+%et qu'elle n'était pas dans Gulwani?}
+%\JH{Non. Gulwani fait la même chose en un peu plus complique (voir commentaire
+%suivant)}
+%\MM{Du coup, la contribution~\ref{contr:disjunctive} devrait être
+%  affaiblie un peu je suppose. Gulwani fait aussi du SMT-solving ?}
+%\DM{Non, Gulwani énumère tous les chemins un par un à tous les coups.}
+
+For each control point $p_i \in P_R$, we start with one single disjunct
+($m_i = 1$) and define $\delta_i = 1$.
 $M$ denotes an upper bound on the number of disjuncts per control point.
-$\sigma_i(j,k)$ is then constructed on the fly only when needed.
-When $\sigma_i(j,k)$ needs to be computed, we first compute the image of the
+
+The $\sigma_i$ functions take as parameters the index of the starting abstract
+value, and the path we focus on. 
+Since we dynamically construct these functions
+during the analysis, we store their already comupted image into a
+compact representation, such as Algebraic Decision Diagrams.
+$\sigma_i(j,k)$ is then constructed on the fly only when needed, and computed
+only once.
+When the value of $\sigma_i(j,k)$ is required but undefined, we first compute
+the image of the
 abstract value $X_{i,j}$ by the path indexed by $k$, and try to find an existing
@@
 	\item if $m_i = M$, we define $\sigma_i(j,k) = M$ 
-		\JH{A la place, Gulwani cherche l'invariant qui a le plus de contraintes
-		d'egalite en commun pour definir son $sigma$. Nous, on le met dans M, pour simplifier}
+		%\JH{A la place, Gulwani cherche l'invariant qui a le plus de contraintes
+		%d'egalite en commun pour definir son $sigma$. Nous, on le met dans M, pour simplifier}
 \end{itemize}
@@
 
-\MM{On peut peut-être citer le nom de l'outil (on garde PAGAI ?) ici.
-  L'intérêt d'avoir un nom, c'est que les gens peuvent en parler plus
-  facilement, et c'est plus facile de retrouver le papier origine d'un
-  outil quand il cite son nom}
-\DM{Dans la version finale, oui, mais pas dans une soumission anonymisée... suffit de Googlifier pagai sur imag.fr}
-
+%\MM{On peut peut-être citer le nom de l'outil (on garde PAGAI ?) ici.
+%  L'intérêt d'avoir un nom, c'est que les gens peuvent en parler plus
+%  facilement, et c'est plus facile de retrouver le papier origine d'un
+%  outil quand il cite son nom}
+%\DM{Dans la version finale, oui, mais pas dans une soumission anonymisée... suffit de Googlifier pagai sur imag.fr}
+%
 Our tool operates over LLVM bitcode \citep{LLVM_langref,Lattner:2004:LCF:977395.977673}, which is a target for several compilers, most notably Clang (supporting C and C++) and llvm-gcc (supporting C, C++, Fortran and Ada).

### 2011-11-02 papers/PLDI2012/implicitization.tex

@@
 \citet{DBLP:conf/pldi/GulwaniZ10}. 
-%\DM{Cette heuristique, je croyais que c'était nous qui l'avions trouvée,
-%et qu'elle n'était pas dans Gulwani?}
-%\JH{Non. Gulwani fait la même chose en un peu plus complique (voir commentaire
-%suivant)}
-%\MM{Du coup, la contribution~\ref{contr:disjunctive} devrait être
-%  affaiblie un peu je suppose. Gulwani fait aussi du SMT-solving ?}
-%\DM{Non, Gulwani énumère tous les chemins un par un à tous les coups.}
-
+\DM{Cette heuristique, je croyais que c'était nous qui l'avions trouvée,
+et qu'elle n'était pas dans Gulwani?}
+\JH{Non. Gulwani fait la même chose en un peu plus complique (voir commentaire
+suivant)}
+\MM{Du coup, la contribution~\ref{contr:disjunctive} devrait être
+  affaiblie un peu je suppose. Gulwani fait aussi du SMT-solving ?}
+\DM{Non, Gulwani énumère tous les chemins un par un à tous les coups.}
+\MM{OK, du coup j'ai ajouté une phrase à la fin de la section pour
+  rappeler notre contrib}
 For each control point $p_i \in P_R$, we start with one single disjunct

### 2011-11-02 papers/PLDI2012/implicitization.tex
cite PPL

@@
 improved widening operator from \citet{BagnaraHRZ05SCP} (POLY*), and Convex Polyhedra
-using the classical widening operator (POLY).}
+using the classical widening operator (POLY). POLY and POLY* use the Parma
+Polyhedral Library \cite{PPL}.}
 \label{tab:domain}

### 2011-11-02 papers/PLDI2012/implicitization.tex
explication plus claire du X^s / X^d

@@
 same path infinitely many times, we use another abstract value
-associated to a control point $p_j$, noted $X_j^d$, which is distinct
-from $X_j^s$, and initialized to $X_j^s$. This value is updated when a
-point $p_j$ is the target of a new path, and is used instead of
-$X_j^s$ in the SMT formula. We thus check the satisfiability of the formula
+associated to the control point $p_j$, noted $X_j^d$, which is distinct
+from $X_j^s$, and initialized to $X_j^s$ right before computing new paths. 
+In the SMT formula, we associate to $p_j$ the abstract value $X_j^s$ when $p_j$
+is the starting point of the path, and $X_j^d$ when it is its destination point.
+We thus check the satisfiability of the formula
 $f'(p_i)$, where:
@@
 
+$X_j^d$ is updated when the point $p_j$ is the target of a new path.
 This way, further SMT queries do not

### 2011-11-02 papers/PLDI2012/implicitization.tex
legende

@@
 over the classic Abstract interpretation algorithm.
-$\subsetneq$, $\supsetneq$ and ``unc.'' defined as in Fig.~\ref{fig:techniques}.}
+$\subsetneq$, $\supsetneq$ and ``unc.'' are defined as in Fig.~\ref{fig:techniques}.}
 \label{tab:techniques}
@@
 \caption{Results of the comparison of the various abstract domains, when using
-the same technique (G+PF). We used as abstract domains Convex Polyhedra (PK),
+the same technique (G+PF). We used as abstract domains Convex Polyhedra (PK and
+POLY),
 Octagons (OCT), intervals (BOX), linear equalities (PKEQ) and linear congruences
@@
 using the classical widening operator (POLY). POLY and POLY* use the Parma
-Polyhedral Library \cite{PPL}.}
+Polyhedral Library \cite{PPL}.
+$\subsetneq$, $\supsetneq$ and ``unc.'' are defined as in Fig.~\ref{fig:techniques}.}
 \label{tab:domain}

### 2011-11-02 papers/PLDI2012/implicitization.tex
typo

@@
 Since we dynamically construct these functions
-during the analysis, we store their already comupted image into a
+during the analysis, we store their already computed image into a
 compact representation, such as Algebraic Decision Diagrams.

### 2011-11-04 papers/PLDI2012/implicitization.tex
typos

@@
 \begin{abstract}
-Abstract interpretation techniques can be made more precise by distinguishing paths inside loops, at the expensive of possibly exponential complexity.
-SMT-solving techniques and sparse representations of paths and sets of paths avoid this pitfall.
+Abstract interpretation techniques can be made more precise by distinguishing
+paths inside loops, at the expense of possibly exponential complexity. SMT-solving techniques and sparse representations of paths and sets of paths avoid this pitfall.
 
@@
 
-There are many possible combinations of the above techniques, and it is not evident which ones perform more or less precisely or more or less efficiency on real-life examples. One needs to experiment.
+There are many possible combinations of the above techniques, and it is not
+evident which ones perform more or less precisely or more or less efficiently on real-life examples. One needs to experiment.
 Unfortunately, the published literature on the subject lacks experimental comparative assessments. One purpose of this article is therefore to propose such experimental results.
@@
 
-Once an inductive invariant $\gamma((\abstr{x}_p)_{p \in P})$ has been obtained, one can attempt \emph{decreasing} or \emph{narrowing} iterations to reduce it.
-In their simplest form, this just means running the following operation until a fixpoint or a maximal number of iterations are reached: for any $p'$, replace $\abstr{x}_{p'}$ by $\abstr{x}_p \cap \left(\bigsqcup_{p \in P} \abstr{\tau}_{p,p'} (\abstr{x}_p)\right)$.
-The result also defines an inductive invariant.
-These decreasing iterations are indispensable to recover properties from guards (tests) in the program in most iteration settings; unfortunately, certain loops, particularly those involving identity (no-operation) transitions, may foil them: the iterations immediately reach a fixpoint and do not decrease further (see example in \S\ref{subsec:rate_lim}).
-Sections \ref{sec:guided} and \ref{sec:path_focusing} describe techniques that work around this problem.
+Once an inductive invariant $\gamma((\abstr{x}_p)_{p \in P})$ has been obtained,
+one can attempt \emph{decreasing} or \emph{narrowing} iterations to reduce it.
+In their simplest form, this just means running the following operation until a
+fixpoint or a maximal number of iterations are reached: for any $p'$, replace
+$\abstr{x}_{p'}$ by $\abstr{x}_{p'} \cap \left(\bigsqcup_{p \in P}
+\abstr{\tau}_{p,p'} (\abstr{x}_p)\right)$. The result also defines an inductive
+invariant. These decreasing iterations are indispensable to recover properties
+from guards (tests) in the program in most iteration settings; unfortunately,
+certain loops, particularly those involving identity (no-operation) transitions,
+may foil them: the iterations immediately reach a fixpoint and do not decrease
+further (see example in \S\ref{subsec:rate_lim}). Sections \ref{sec:guided} and
+\ref{sec:path_focusing} describe techniques that work around this problem.
 
@@
 \begin{figure}
-\label{fig:multigraph}
 \centering
@@
 \caption{Expansion of the transition graph into a multigraph.}
+\label{fig:multigraph}
 \end{figure}

### 2011-11-04 papers/PLDI2012/implicitization.tex
capitalize
rephrase 2.3

--- 2011-11-02 papers/PLDI2012/implicitization.tex
@@
 
-\newcommand{\MM}[1]{{\color{blue} TODO(MM): #1}}
-\newcommand{\JH}[1]{{\color{teal} TODO(JH): #1}}
-\newcommand{\DM}[1]{{\color{violet} TODO(DM): #1}}
-%\newcommand{\MM}[1]{}
-%\newcommand{\JH}[1]{}
-%\newcommand{\DM}[1]{}
+%\newcommand{\MM}[1]{{\color{blue} TODO(MM): #1}}
+%\newcommand{\JH}[1]{{\color{teal} TODO(JH): #1}}
+%\newcommand{\DM}[1]{{\color{violet} TODO(DM): #1}}
+\newcommand{\MM}[1]{}
+\newcommand{\JH}[1]{}
+\newcommand{\DM}[1]{}
 
@@
 
-\title{Succinct representations for abstract interpretation}
+\title{Succinct Representations for Abstract Interpretation}
 \subtitle{Combined analysis algorithms and experimental evaluation}
@@
 We have transformed the first program into the second, manually pruning out infeasible paths; yet in general the resulting program could be exponentially larger than the first (as in Fig.~\ref{fig:multigraph}), even though not all feasible paths are needed to compute the invariant.
+
 Following recent suggestions \cite{Gawlitza_Monniaux_ESOP11,Monniaux_Gonnord_SAS11}, we avoid this space explosion by keeping the second program implicit while simulating its analysis.
-This means we work on an implicitly represented transition multigraph (Fig.~\ref{fig:multigraph}); its compact representation is given by the transition graph of the first program.
-
-Contribution \ref{contr:disjunctive} (\S\ref{sec:disjunctive}) replaces the exponential expansion of disjunctive invariant generation from \citet{DBLP:conf/pldi/GulwaniZ10} by an implicit representation;
-contribution \ref{contr:guided_multigraph} (\S\ref{sec:guided_multigraph}) recasts the ``guided analysis'' from \citet{DBLP:conf/sas/GopanR07} on an implicit representation of the paths in lieu of the individual transitions.
+This means we work on an implicitly represented transition multigraph (Fig.~\ref{fig:multigraph}); it is succinctly representated by the transition graph of the first program.
+Our first contribution (\S\ref{sec:guided_multigraph}) is to recast the ``guided analysis'' from \citet{DBLP:conf/sas/GopanR07} on such a succinct representation of the paths in lieu of the individual transitions.
+A similar explosion occurs in disjunctive invariant generation, following \citet{DBLP:conf/pldi/GulwaniZ10}; our second contribution (\S\ref{sec:disjunctive}) applies our implicit representation to their method.
 

### 2011-11-04 papers/PLDI2012/implicitization.tex
commentaires vraiment en commentaire, compatibilite xelatex

@@
 %\newcommand{\DM}[1]{{\color{violet} TODO(DM): #1}}
-\newcommand{\MM}[1]{}
-\newcommand{\JH}[1]{}
-\newcommand{\DM}[1]{}
-
-% Pour les commentaires
+%\newcommand{\MM}[1]{}
+%\newcommand{\JH}[1]{}
+%\newcommand{\DM}[1]{}
+
+\usepackage{ifxetex}
+\ifxetex
+\else
 \usepackage[utf8]{inputenc}
+\fi
 
@@
 
-\MM{Maintenant que l'ordre a changé dans l'algo, on peut peut-être
-  échanger cette section et la suivante ?}
+%\MM{Maintenant que l'ordre a changé dans l'algo, on peut peut-être
+%  échanger cette section et la suivante ?}
 
@@
 \citet{DBLP:conf/pldi/GulwaniZ10}. 
-\DM{Cette heuristique, je croyais que c'était nous qui l'avions trouvée,
-et qu'elle n'était pas dans Gulwani?}
-\JH{Non. Gulwani fait la même chose en un peu plus complique (voir commentaire
-suivant)}
-\MM{Du coup, la contribution~\ref{contr:disjunctive} devrait être
-  affaiblie un peu je suppose. Gulwani fait aussi du SMT-solving ?}
-\DM{Non, Gulwani énumère tous les chemins un par un à tous les coups.}
-\MM{OK, du coup j'ai ajouté une phrase à la fin de la section pour
-  rappeler notre contrib}
+%\DM{Cette heuristique, je croyais que c'était nous qui l'avions trouvée,
+%et qu'elle n'était pas dans Gulwani?}
+%\JH{Non. Gulwani fait la même chose en un peu plus complique (voir commentaire
+%suivant)}
+%\MM{Du coup, la contribution~\ref{contr:disjunctive} devrait être
+%  affaiblie un peu je suppose. Gulwani fait aussi du SMT-solving ?}
+%\DM{Non, Gulwani énumère tous les chemins un par un à tous les coups.}
+%\MM{OK, du coup j'ai ajouté une phrase à la fin de la section pour
+%  rappeler notre contrib}
 For each control point $p_i \in P_R$, we start with one single disjunct
@@
   \end{center} 
-  \vspace{-20px}
+  \vspace{-20pt}
   \caption{Comparison of the abstract values obtained by classical abstract

### 2011-11-05 papers/PLDI2012/implicitization.tex
typos

@@
 
-An additional claim of the methods that distinguish paths inside the loop body \cite{DBLP:conf/pldi/GulwaniZ10,Monniaux_Gonnord_SAS11} is that they tend to generate better invariants that methods that do not, by behaving better with respect to the \emph{widening operators} \cite{CousotCousot_JLC92}
+An additional claim of the methods that distinguish paths inside the loop body \cite{DBLP:conf/pldi/GulwaniZ10,Monniaux_Gonnord_SAS11} is that they tend to generate better invariants than methods that do not, by behaving better with respect to the \emph{widening operators} \cite{CousotCousot_JLC92}
 used for enforcing convergence when searching for loop invariants by Kleene iterations.
@@
 This program implements a construct commonly found in control programs (in e.g.
-automotive or avionics): a rate limiter.
-For the sake of simplicity, we chose it to be fed a nondeterministic input clamped between $[-100000,100000]$, but in a real system it would be integrated in a reactive control loop and its input connected to a complex system with unknown output range.
+automotive or avionics): a rate limiter. For the sake of simplicity, we chose it
+to be fed on a nondeterministic input clamped between $[-100000,100000]$, but in a real system it would be integrated in a reactive control loop and its input connected to a complex system with unknown output range.
 
@@
 
-The analysis starts by identifying the test taken and not taken during the first iteration of the loop, starting in the loop initialization.
+The analysis starts by identifying the tests taken and not taken during the first iteration of the loop, starting in the loop initialization.
 The branches not taken are pruned from the loop body, yielding:
@@
 
-Algorithm~ \ref{algo:combined} performs Guided static analysis on the implicitly represented multigraph.
+Algorithm~\ref{algo:combined} performs Guided static analysis on the implicitly represented multigraph.
 
@@
 
-We have implemented our proposed solutions inside a prototype intraprocedural
+We have implemented our proposed solutions inside a prototype of intraprocedural
 static analyzer,

### 2011-11-23 ex/compile_llvm.sh
presentation asopt

@@ -58,8 +58,8 @@ fi
 clang -DNDEBUG -fno-exceptions -emit-llvm -c $FILENAME -o $OUTPUT
 #opt -mem2reg -loopsimplify -lowerswitch $OUTPUT -o $OUTPUT
 #opt -mem2reg -lowerswitch -loops  -loop-simplify -loop-rotate -lcssa -loop-unroll -unroll-count=1 $OUTPUT -o $OUTPUT
-opt -mem2reg -lowerswitch -inline -loop-rotate $OUTPUT -o $OUTPUT
-#opt -mem2reg -lowerswitch -inline $OUTPUT -o $OUTPUT
+#opt -mem2reg -lowerswitch -inline -loop-rotate $OUTPUT -o $OUTPUT
+opt -mem2reg -lowerswitch -inline $OUTPUT -o $OUTPUT
 
 if [ $GRAPH -eq 1 ] ; then
 	opt -dot-cfg $OUTPUT -o $OUTPUT

### 2011-11-23 src/AIpf.cc

@@ -63,7 +63,7 @@ bool AIpf::runOnModule(Module &M) {
 				Out->changeColor(raw_ostream::MAGENTA,true);
 				*Out << "\n\nRESULT FOR BASICBLOCK: -------------------" << *b << "-----\n";
 				Out->resetColor();
-				n->X_d[passID]->print(true);
+				n->X_s[passID]->print(true);
 				N_Pr++;
 			}
 			//delete Nodes[b];
@@ -125,7 +125,17 @@ void AIpf::computeFunction(Function * F) {
 		*Out << "#######################################################\n";
 		Out->resetColor();
 	);
+
+	// we set X_d abstract values to bottom for narrowing
+	//for (Function::iterator i = F->begin(), e = F->end(); i != e; ++i) {
+	//	b = i;
+	//	if (Pr::getPr(*F)->count(i) && Nodes[b] != n) {
+	//		Nodes[b]->X_d[passID]->set_bottom(env);
+	//	}
+	//}
 	narrowingIter(n);
+	// then we move X_d abstract values to X_s abstract values
+	copy_Xd_to_Xs(F);
 }
 
 std::set<BasicBlock*> AIpf::getPredecessors(BasicBlock * b) const {
@@ -220,8 +230,8 @@ void AIpf::computeNode(Node * n) {
 		Xtemp->join_array(Xtemp->main->env,Join);
 
 		if (LI->isLoopHeader(Succ->bb) && ((Succ != n) || !only_join)) {
-				//Xtemp->widening(Succ->X_s[passID]);
-				Xtemp->widening_threshold(Succ->X_s[passID],&threshold);
+				Xtemp->widening(Succ->X_s[passID]);
+				//Xtemp->widening_threshold(Succ->X_s[passID],&threshold);
 				DEBUG(
 					*Out << "WIDENING! \n";
 				);

### 2011-12-22 papers/PLDI2012/response_PLDI2012.txt
review + brouillon rebuttal

"the title is fairly vague and non optimal -- perhaps
"succinct" should be replaced by "disjunctive"?"
Path Focusing technique and our combined technique (with or without disjunctive
invariants) do not work on the classical control flow graph, but on a multigraph
where all the paths inside loops are distinguished. The term "succint" refers to
the fact we actually do not compute all the paths of this multigraph, but work
on an succint representation thereof (using an SMT formula), thus avoiding the
enumeration of an exponential number of paths in a huge number of cases.

"The results are unsurprising"
The experimental results show that techniques that distinguish paths inside a
program give more precise results, but, more importantly, at a reasonable cost
(which was not evident).
It is also interesting to see that more complex technique perform sometimes
faster that more naive approaches.

"The problem with path focusing is that (1) it will process all
executable paths at once, including potentially many redundant paths
that don't contribute to the solution; and (2) its widening operation
lacks the precision conferred by guided static analysis."

This is not exactly the point. The problem with path focusing is that it will compute
the fixpoint iterations on the whole graph at once, using classical widening
operations in order to guarantee the convergence.
When the fixpoint has been reached, it finally apply some narrowing iterations
to recover precision.
The idea here is to apply some narrowing iterations during the analysis, in
order to recover some precision before continuing.
To clarify, widening is the same, but we perform regularly narrowing iterations,
and not only at the end of the ascending iterations.

"how much benefit this technique confers to other types of abstract
domains, including those that don't require widening"
Intuitively, Guided Static Analysis technique performs better because it
recovers precision lost by widening. It will probably not increase the precision
when using domains without widening/narrowing. However, the iteration strategy
will be changed, so the computation time may differ compared to a classical
abstract interpretation. We can do the same remarks on our combined technique
compared to Path Focusing.

"I'd like to know more about how applicable the technique
is to other kinds of abstract domains"
The general idea does not depend on the abstract domain. However, we have to
transform an abstract value into an SMT formula. Such transformation is easy in
our case, since it is a cunjunction of linear inequalities, but may still
require some research for other abstract domains that could require other
theories (for instance, arrays could be
encoded in SMT using uninterpreted functions, etc.)

"2 to 5% difference, does it make a difference at all?"
Our prototype of analyzer performs many over-approximations (no memory model,
etc.), thus leading to invariants that are often very unprecise.
Intuitively, I would say we could have obtained more difference between the
different techniques by reducing the number of over-approximation we do for the
time being.

### 2012-02-09 papers/SAS2012/gopan_reps.c
ca semble compiler

@@ -1,0 +1,8 @@
+int x = 0;
+int y = 0;
+while (1) {
+	if (x <= 50) y++;
+	else y--;
+	if (y < 0) break;
+	x++;
+}

### 2012-02-09 papers/SAS2012/rlim_int.c

@@ -1,0 +1,10 @@
+extern int input();
+void rate_limiter() {
+  int x_old = 0;
+  while (1) {
+    int x = input(-100000, 100000);
+    if (x > x_old+10) x = x_old+10;
+    if (x < x_old-10) x = x_old-10;
+    x_old = x;
+  }
+}

### 2012-02-09 papers/SAS2012/implicitization.tex
tassage

--- 2011-11-05 papers/PLDI2012/implicitization.tex
@@
-\documentclass[preprint]{sigplanconf}
-
-% The following \documentclass options may be useful:
-%
-% 10pt          To set in 10-point type instead of 9-point.
-% 11pt          To set in 11-point type instead of 9-point.
-% authoryear    To obtain author/year citation style instead of numeric.
-
-%\newcommand{\MM}[1]{{\color{blue} TODO(MM): #1}}
-%\newcommand{\JH}[1]{{\color{teal} TODO(JH): #1}}
-%\newcommand{\DM}[1]{{\color{violet} TODO(DM): #1}}
-%\newcommand{\MM}[1]{}
-%\newcommand{\JH}[1]{}
-%\newcommand{\DM}[1]{}
+\documentclass{llncs}
 
@@
 \usepackage{amsmath}
-\usepackage{mathcomp,amsfonts,amssymb,MnSymbol}
-\usepackage{dmnatbib}
+\usepackage{mathcomp,amsfonts,amssymb,mathabx}
 \usepackage[mathscr]{euscript}
@@
 
-\usepackage[pdfusetitle]{hyperref}
+\newcommand{\mytitle}{Succinct Representations for Abstract Interpretation}
+\usepackage[pdftitle={\mytitle},pdfauthor={Julien Henry, David Monniaux, Matthieu Moy}]{hyperref}
 
@@
 
+\graphicspath{{figures/}{gnuplot/}}
+
 % commentaires dans les algorithmes crees avec le package algorithmic
@@
 
+\title{\mytitle}
+\subtitle{Combined analysis algorithms and experimental evaluation}
+\author{Julien Henry\inst{1}\inst{2} \and David Monniaux\inst{1}\inst{3} \and Matthieu Moy\inst{1}\inst{4}}
+
+\institute{VERIMAG laboratory, Grenoble, France
+\and Université Joseph Fourier
+\and CNRS
+\and Grenoble-INP}
+
 \begin{document}
-
-\conferenceinfo{PLDI '12}{11-16 June 2012, Beijing} 
-\copyrightyear{2012} 
-\copyrightdata{[to be supplied]}
-
-%\titlebanner{banner above paper title}        % These are ignored unless
-\preprintfooter{Succinct representations for abstract interpretation}   % 'preprint' option specified.
-
-\title{Succinct Representations for Abstract Interpretation}
-\subtitle{Combined analysis algorithms and experimental evaluation}
-
-\authorinfo{Anonymous}
-           {The Internet}
-           {anonymous.gal@gmail.com}
-
 \maketitle
@@
 \end{abstract}
-
-% \category{CR-number}{subcategory}{third-level}
-\category{D.2.4}{Software/Program Verification}{Formal methods}
-\category{F.3.1}{Specifying and Verifying and Reasoning about Programs}{Assertions, Invariants, Mechanical verification, Pre- and post-conditions}
-\category{F.3.2}{Semantics of Programming Languages}{Program analysis}
-\category{F.4.1}{Mathematical Logic}{Logic and constraint programming, Mechanical theorem proving}
-\category{I.2.3}{Deduction and Theorem Proving}{Inference engines, Logic programming, Resolution}
-
-\terms
-theory, verification, algorithms
-
-\keywords
-SMT-solving, static analysis, abstract interpretation, disjunctions
 
@@
 \begin{enumerate}
-\item We recast the guided static analysis technique from \citet{DBLP:conf/sas/GopanR07} on the expanded multigraph from \citet{Monniaux_Gonnord_SAS11}, considering entire paths instead of individual transitions, using SMT queries and binary decision diagrams.\label{contr:guided_multigraph} (See \S\ref{sec:guided_multigraph})
-\item We improve the technique for obtaining disjunctive invariants from \citet{DBLP:conf/pldi/GulwaniZ10} by replacing the explicit exhaustive enumeration of paths by a sequence of SMT queries.\label{contr:disjunctive} (See \S\ref{sec:disjunctive})
+\item We recast the guided static analysis technique from \cite{DBLP:conf/sas/GopanR07} on the expanded multigraph from \cite{Monniaux_Gonnord_SAS11}, considering entire paths instead of individual transitions, using SMT queries and binary decision diagrams.\label{contr:guided_multigraph} (See \S\ref{sec:guided_multigraph})
+\item We improve the technique for obtaining disjunctive invariants from \cite{DBLP:conf/pldi/GulwaniZ10} by replacing the explicit exhaustive enumeration of paths by a sequence of SMT queries.\label{contr:disjunctive} (See \S\ref{sec:disjunctive})
 \item We implemented these techniques, in addition to ``classical'' iterations and the original guided static analysis, inside a prototype static analyzer.
@@
 
-The expected inductive invariant is $\lstinline|x_old| \in [-100000,100000]$, but classical abstract interpretation using intervals (or octagons or polyhedra) finds $\lstinline|x_old| \in (-\infty,+\infty)$~\citep{ASTREE_ESOP05}.
+The expected inductive invariant is $\lstinline|x_old| \in [-100000,100000]$, but classical abstract interpretation using intervals (or octagons or polyhedra) finds $\lstinline|x_old| \in (-\infty,+\infty)$~\cite{ASTREE_ESOP05}.
 Let us briefly see why.
@@
 We could use ``widening up to'' or ``widening with thresholds'', propagating the ``magic values'' $\pm 100000$ associated to \lstinline|x| into \mbox{\lstinline|x_old|,} but these syntactic approaches cannot directly cope with programs for which $\lstinline|x|  \in [-100000,+100000]$ is itself obtained by analysis.
-The guided static analysis of \citet{DBLP:conf/sas/GopanR07} does not perform
+The guided static analysis of \cite{DBLP:conf/sas/GopanR07} does not perform
 better, and also obtains $\lstinline|x_old| \in (-\infty,+\infty)$.
@@
 This means we work on an implicitly represented transition multigraph (Fig.~\ref{fig:multigraph}); it is succinctly representated by the transition graph of the first program.
-Our first contribution (\S\ref{sec:guided_multigraph}) is to recast the ``guided analysis'' from \citet{DBLP:conf/sas/GopanR07} on such a succinct representation of the paths in lieu of the individual transitions.
-A similar explosion occurs in disjunctive invariant generation, following \citet{DBLP:conf/pldi/GulwaniZ10}; our second contribution (\S\ref{sec:disjunctive}) applies our implicit representation to their method.
+Our first contribution (\S\ref{sec:guided_multigraph}) is to recast the ``guided analysis'' from \cite{DBLP:conf/sas/GopanR07} on such a succinct representation of the paths in lieu of the individual transitions.
+A similar explosion occurs in disjunctive invariant generation, following \cite{DBLP:conf/pldi/GulwaniZ10}; our second contribution (\S\ref{sec:disjunctive}) applies our implicit representation to their method.
 
@@
 \label{sec:guided}
-\emph{Guided static analysis} was proposed by \citet{DBLP:conf/sas/GopanR07} as an improvement over classical upward Kleene iterations with widening.
+\emph{Guided static analysis} was proposed by \cite{DBLP:conf/sas/GopanR07} as an improvement over classical upward Kleene iterations with widening.
 Consider the following program, taken from \cite{DBLP:conf/sas/GopanR07}:
@@
 \begin{center}
-\includegraphics[scale=0.75]{figures/gopan_reps}
+\includegraphics[scale=0.75]{gopan_reps}
 \end{center}
@@
 
-\citeauthor{DBLP:conf/sas/GopanR07}' idea is to analyze the first phase of the loop with a widening and narrowing sequence, and thus obtain $0 \leq x \leq 50 \land y = x$, and then analyze the second phase, finally obtaining invariant~\ref{eqn:triangle}; each phase is identified by the tests taken or not taken.
+Gopan and Reps' idea is to analyze the first phase of the loop with a widening and narrowing sequence, and thus obtain $0 \leq x \leq 50 \land y = x$, and then analyze the second phase, finally obtaining invariant~\ref{eqn:triangle}; each phase is identified by the tests taken or not taken.
 
@@
 
-\citet{Monniaux_Gonnord_SAS11}'s \emph{path-focusing} technique
+\cite{Monniaux_Gonnord_SAS11}'s \emph{path-focusing} technique
 distinguishes the different paths in the program in order to avoid loss of
@@
 
-Guided static analysis, as proposed by \citet{DBLP:conf/sas/GopanR07}, applies to the transition
+Guided static analysis, as proposed by \cite{DBLP:conf/sas/GopanR07}, applies to the transition
 graph of the program. We now present a new technique applying this analysis on the implicit
@@
 invariant computation of the subgraph is finished. As proposed by
-\citet{DBLP:conf/sas/GopanR07}, we can do some narrowing iterations.
+\cite{DBLP:conf/sas/GopanR07}, we can do some narrowing iterations.
 These narrowing iterations allow to recover precision lost by
@@
 For computing an inductive invariant over a subgraph, we use the
-Path-focusing algorithm from \citet{Monniaux_Gonnord_SAS11} with special
+Path-focusing algorithm from \cite{Monniaux_Gonnord_SAS11} with special
 treatment for self loops (line~\ref{alg=pf} in algorithm~\ref{algo:combined}).
@@
 
-The difference with~\citet{Monniaux_Gonnord_SAS11} is that we do not
+The difference with~\cite{Monniaux_Gonnord_SAS11} is that we do not
 work on the entire transition graph but on a subset of it. Therefore we
@@
 
-Guided static analysis from \citet{DBLP:conf/sas/GopanR07} yields, at line 5,
+Guided static analysis from \cite{DBLP:conf/sas/GopanR07} yields, at line 5,
 $\lstinline|x_old| \in (-\infty,+\infty)$.
-Path-focusing \citep{Monniaux_Gonnord_SAS11} performs slightly
+Path-focusing \cite{Monniaux_Gonnord_SAS11} performs slightly
 better, and finds $\lstinline|x_old| \in (-\infty,10000]$.
@@
 
-\citet{DBLP:conf/pldi/GulwaniZ10} proposed a technique for computing disjunctive invariants, by
+\cite{DBLP:conf/pldi/GulwaniZ10} proposed a technique for computing disjunctive invariants, by
 distinguishing all the paths inside a loop. In
@@
 
-\item At each loop iteration of the original algorithm \citep{DBLP:conf/pldi/GulwaniZ10}, an image by each path inside the loop is computed for each disjunct of the invariant candidate.
+\item At each loop iteration of the original algorithm \cite{DBLP:conf/pldi/GulwaniZ10}, an image by each path inside the loop is computed for each disjunct of the invariant candidate.
 Yet, many of these images may be redundant: for instance, if our invariant candidate is $(0 \leq x \leq 10 \land 0 \leq y \leq 1000) \lor (x < -10 \land y < -10)$, then there is no point enumerating paths whose image is included in this invariant candidate.
@@
 
-\begin{eqnarray*}
-g(p_i) & = & \rho \wedge b_i^s \wedge 
-\displaystyle\bigwedge_{j \in P_R \atop j \neq i} \neg b_j^s  \\
- & & \wedge 
+\begin{equation*}
+g(p_i) = \rho \wedge b_i^s \wedge 
+\displaystyle\bigwedge_{j \in P_R \atop j \neq i} \neg b_j^s 
+ \wedge 
 \displaystyle\bigvee_{1 \leq k \leq m_i} (d_k \wedge X_{i,k} \wedge \bigwedge_{l \neq k}
-\neg d_l) \\
- & & \wedge
+\neg d_l)
+\wedge
 \displaystyle\bigvee_{j \in Succ(i)} 
 (b_j^d \wedge \bigwedge_{1 \leq k \leq m_i} (\neg X_{j,k}))
-\end{eqnarray*}
+\end{equation*}
 
@@
 \tau_{i,k}(X_{i,j}))$, or can integrate the special treatment for self loops
-proposed by \citet{Monniaux_Gonnord_SAS11}, with widening/narrowing sequence or
+proposed by \cite{Monniaux_Gonnord_SAS11}, with widening/narrowing sequence or
 acceleration.
@@
 of dynamic construction of the $\sigma_i$ functions, adapted from
-\citet{DBLP:conf/pldi/GulwaniZ10}. 
+\cite{DBLP:conf/pldi/GulwaniZ10}. 
 %\DM{Cette heuristique, je croyais que c'était nous qui l'avions trouvée,
@@
 %
-Our tool operates over LLVM bitcode \citep{LLVM_langref,Lattner:2004:LCF:977395.977673}, which is a target for several compilers, most notably Clang (supporting C and C++) and llvm-gcc (supporting C, C++, Fortran and Ada).
-Abstract domains are provided by the APRON library \citep{DBLP:conf/cav/JeannetM09}, and include convex polyhedra (from the builtin Polka ``PK'' library), octagons, intervals, and linear congruences.
+Our tool operates over LLVM bitcode \cite{LLVM_langref,Lattner:2004:LCF:977395.977673}, which is a target for several compilers, most notably Clang (supporting C and C++) and llvm-gcc (supporting C, C++, Fortran and Ada).
+Abstract domains are provided by the APRON library \cite{DBLP:conf/cav/JeannetM09}, and include convex polyhedra (from the builtin Polka ``PK'' library), octagons, intervals, and linear congruences.
 For SMT-solving, our analyzer uses Yices
@@
 Note that the resulting set is not necessarily minimal, but is sufficient to disconnect all cycles --- more sophisticated techniques are discussed in e.g.
-\citet{BourdonclePhd}.
+\cite{BourdonclePhd}.
 
@@
 LLVM makes it easy to follow definition-use and use-definition chains: for a given variable (say, $x_2$) one can immediately obtain its definition (say, $2x_1+1$).
-One may see conversion to SSA form as a static precomputation of some of the symbolic propagations proposed by \citet{DBLP:conf/vmcai/Mine06} to enhance the precision of analyses.
+One may see conversion to SSA form as a static precomputation of some of the symbolic propagations proposed by \cite{DBLP:conf/vmcai/Mine06} to enhance the precision of analyses.
 
@@
 (ii) the abstract operation for a single path in path-focusing methods also is a (large) parallel assignment;
-(iii) as suggested by \citet{DBLP:conf/vmcai/Mine06}, this approach is more precise than running abstract operations for each program line separately:
+(iii) as suggested by \cite{DBLP:conf/vmcai/Mine06}, this approach is more precise than running abstract operations for each program line separately:
 for instance, for \lstinline|y=x; z=x-y;| with precondition $x \in [0,1]$, a line-by-line interval analysis obtains $y \in [0,1]$ and $z \in [-1,1]$ while our ``en bloc'' analysis symbolically simplifies $z = x - x = 0$ and thus $z \in [0,0]$.
@@
 Techniques for sound analysis of bounded integers, including with wraparound, and of floating-point operations have been developed in e.g.
-the Astr\'ee system \citep{ASTREE_ESOP05,ASTREE_PLDI03}, but porting these techniques to our iteration schemes using SMT-solving requires supplemental work.
-
-Our implementation of path-focusing currently does not use true acceleration techniques, as proposed by \citet{Monniaux_Gonnord_SAS11}.
+the Astr\'ee system \cite{ASTREE_ESOP05,ASTREE_PLDI03}, but porting these techniques to our iteration schemes using SMT-solving requires supplemental work.
+
+Our implementation of path-focusing currently does not use true acceleration techniques, as proposed by \cite{Monniaux_Gonnord_SAS11}.
 Instead, it simply runs widening and narrowing iterations on a single path.
@@
   \begin{center}
-    \input{./techniques.tex}
+    \input{gnuplot/techniques.tex}
   \end{center} 
@@
 Let us briefly comment the results given in more details in Table~\ref{tab:techniques} and Figure~\ref{fig:techniques}.
-\emph{Guided Static Analysis} from \citet{DBLP:conf/sas/GopanR07} improves the
+\emph{Guided Static Analysis} from \cite{DBLP:conf/sas/GopanR07} improves the
 result of the classical Abstract Interpretation in $2.72\%$ of the control points
 in $P_R$.
-\emph{Path-focusing} from \citet{Monniaux_Gonnord_SAS11} gives
+\emph{Path-focusing} from \cite{Monniaux_Gonnord_SAS11} gives
 statistically better results, and finds better invariants than \emph{Guided
@@
 In cases where the formula is expressed in too rich a logic for the SMT-solver to deal with, a number of workarounds are possible, including:
-(i) \emph{Linearization}, as per \citet{DBLP:conf/vmcai/Mine06}, which overapproximates nonlinear semantics by linear semantics.
+(i) \emph{Linearization}, as per \cite{DBLP:conf/vmcai/Mine06}, which overapproximates nonlinear semantics by linear semantics.
 (ii) Replacing the results of nonlinear operations by ``unknown''.
@@
 (GRID). The last column compares the domain of Convex Polyhedra with the
-improved widening operator from \citet{BagnaraHRZ05SCP} (POLY*), and Convex Polyhedra
+improved widening operator from \cite{BagnaraHRZ05SCP} (POLY*), and Convex Polyhedra
 using the classical widening operator (POLY). POLY and POLY* use the Parma
@@
 Finally, we evaluated the benefits of the improved version of the widening
-operator for convex polyhedra from \citet{BagnaraHRZ05SCP}, compared to the
+operator for convex polyhedra from \cite{BagnaraHRZ05SCP}, compared to the
 classical widening. We found that the improved version from
-\citet{BagnaraHRZ05SCP} yields more precise invariants for $6.58\%$ of the
+\cite{BagnaraHRZ05SCP} yields more precise invariants for $6.58\%$ of the
 control points in $P_R$.
@@
 
-% Cas a probleme:
-% Initialisation (x,y) à {(1,0), (0,1)}
-% puis boucle: si y = 0 alors x++
-% fait un polyèdre qui finit par perdre y <= 1 et x >= 0
-
-%\appendix
-%\section{Appendix Title}
-
-%This is the text of the appendix, if you need one.
-
-%\acks
-
-% Acknowledgments, if needed.
-
-% We recommend abbrvnat bibliography style.
-
 \phantomsection\addcontentsline{toc}{section}{References} 
-\bibliographystyle{dmabbrvnat}
+\bibliographystyle{splncs03}
 \bibliography{implicitization}
 \end{document}
-
-% Pour Emacs:
-% Local Variables:
-% reftex-cite-format: natbib
-% End:

### 2012-02-15 ex/pagai.sh
removed presentations and reports

@@ -85,9 +85,9 @@ DIR=`dirname $FILENAME`
 
 if [ -z $BITCODE ] ; then 
 	if [ $UNROLL -eq 0 ] ; then
-		BITCODE=/tmp/${NAME}_nounroll.bc
-	else
 		BITCODE=/tmp/${NAME}.bc
+	else
+		BITCODE=/tmp/${NAME}_unroll.bc
 	fi
 fi
 
@@ -107,8 +107,6 @@ fi
 
 NAME=`basename $BITCODE`
 RESULT=/tmp/${NAME%%.*}.result
-
-echo "running Pagai on $NAME"
 
 ulimit -t $TIME_LIMIT
 
@@ -135,14 +133,14 @@ else
 fi
 xs=$?
 case $xs in
- 0) echo "ok"
+ 0) echo "ok -- $NAME"
 	exit 0
 	 ;; # all fine
  *) if [ $xs -gt 127 ]; then
-       echo "killed"
+       echo "killed -- $NAME"
     else
-       echo "internal error"
+       echo "error -- $NAME"
     fi
-	exit 1
+	exit 0
 esac
 

### 2012-02-15 presentation/pagai.tex
presentation of pagai

\documentclass{beamer}

\usepackage{moy-slides}
\usepackage{tikz}
\usepackage{comment}
\usepackage{listings}

\title[Pagai]{The Pagai tool}

\subtitle{Path Analysis for the Generation of Arithmetic Invariants} % (optional)

\author[J. Henry, D. Monniaux, M. Moy]{\underline{Julien Henry}, David Monniaux, Matthieu
Moy}
% - Use the \inst{?} command only if the authors have different
%   affiliation.

\newcommand{\important}[1]{\color{trefle}{#1}}

\institute[Verimag]{
\includegraphics[height=1.1cm]{../report/images/logoverimag}}

\tikzstyle{arrow}=[->,line width=.05cm,draw=red!90!blue!60!black]

\usetikzlibrary{fit}					% fitting shapes to coordinates
\usetikzlibrary{snakes,arrows,shapes,backgrounds,shadows,automata,patterns}
%\date{}
\usepgflibrary{snakes}

\tikzstyle{state}=[circle,fill=black!25,minimum size=13pt,inner sep=0pt]
\tikzstyle{bb}=[rectangle,fill=black!15,minimum size=13pt,inner sep=0pt]
\tikzstyle{transition}=[->,rectangle,thick,draw=black!75,
  			  fill=black!20,minimum size=4mm]
\tikzstyle{transition2}=[->,rectangle,thick,draw=blue!75,
  			  fill=blue,minimum size=4mm,blue]
\tikzstyle{PRstate}=[circle,fill=red!90,minimum size=13pt,inner sep=0pt]
\tikzstyle{polyhedra}=[blue!25,opacity=0.5,pattern=north west lines,pattern
color=blue]
\tikzstyle{line}=[black,thick]
\tikzstyle{NewPath}=[rectangle,thick,draw=black!75,
  			  fill=black!20,minimum size=4mm,dash pattern=on 3pt off 3pt]
\tikzstyle{TakenPath}=[rectangle,thick,draw=blue!75,
  			  fill=blue,minimum size=4mm,blue]

\begin{document}

\begin{frame}
  \titlepage
\end{frame}

\begin{frame}
	\frametitle{Pagai}
	Compute numerical invariants from an LLVM bitcode file

	\vspace{1cm}

	LLVM bitcode is generated by clang or llvm-gcc :
	\begin{itemize}
		\item clang supports C and C++
		\item llvm-gcc supports C, C++, Fortran and Ada
	\end{itemize}
\end{frame}

\begin{frame}
	\frametitle{LLVM IR}
\begin{columns}
	\lstinputlisting{halbwachs-simple.c}
\begin{column}{1cm}
\end{column}
\begin{column}{10cm}
	\includegraphics[width=7cm]{f}
\end{column}
\end{columns}

\end{frame}

\begin{frame}
	\frametitle{Principle}

	Computes an invariant at the head of each BasicBlock

	\vspace{1cm}

	\begin{itemize}
		\item The invariant is only on live values.
		\item Computes the image of an abstract value by a path/transition in
			one ``big'' step.
	\end{itemize}
\end{frame}

\begin{frame}[containsverbatim]
	\frametitle{Example}
For each loop header, Pagai returns an invariant in the following form:
	\begin{verbatim}
RESULT FOR BASICBLOCK: -------------------
; <label>:1 
  %j.0 = phi i32 [ 0, %0 ], [ %j.1, %8 ]
  %i.0 = phi i32 [ 0, %0 ], [ %9, %8 ]
  %2 = icmp slt i32 %i.0, 100
  br i1 %2, label %3, label %10
-----
polyhedron of dim (2,0)
array of constraints of size 3
 0: -i.0 + j.0 >= 0
 1: -j.0 + 100 >= 0
 2: 100i.0 - j.0 >= 0
\end{verbatim}
\end{frame}

\begin{frame}
	\frametitle{SMT solvers}
	Pagai uses SMT solvers for Path-focusing techniques

	\begin{itemize}
		\item Yices
		\item Microsoft Z3
	\end{itemize}

	In practice, Z3 is better.
\end{frame}

\begin{frame}
	\frametitle{Apron}
B. Jeannet and A. Mine. CAV'2009.

\vspace{1cm}

Apron is a numerical abstract domain library.

\begin{itemize}
	\item BOX intervals library
	\item OCT octagon library
	\item NEWPOLKA Convex Polyhedra and Linear Equalities library 
\end{itemize}
Also provides an interface to the Parma Polyhedra Library
\end{frame}

\begin{frame}
	\frametitle{Techniques implemented}

	\begin{itemize}
		\item Classic Abstract interpretation
		\item Guided static analysis (Gopan \& Reps SAS'07)
		\item Path Focusing (Monniaux \& Gonnord SAS'11)
		\item Combined technique
		\item Combined technique with disjunctive invariants
	\end{itemize}

\end{frame}

\begin{frame}
	\frametitle{Limitations}

	\begin{itemize}
		\item no memory model
			\begin{itemize}
				\item LOAD $x \Rightarrow x \in (-\infty,+\infty)$   
				\item STORE $x$ does nothing 
			\end{itemize}
		\item intraprocedural
			\begin{itemize}
				\item a function call has the same effect than a LOAD
			\end{itemize}
	\end{itemize}

	\vspace{1cm}

	Remark : We use an LLVM pass (mem2reg) that 
	lifts most memory accesses to scalar variables.
\end{frame}

\begin{frame}
	\frametitle{Experiments}

	Various experiments have been done :

	\begin{itemize}
		\item benchmarks for WCET analysis or termination
		\item various GNU programs 
			\begin{itemize}
				\item grep, gnuchess, gnugo, gzip, etc.
			\end{itemize}
	\end{itemize}
\end{frame}

\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% ispell-local-dictionary: "american"
%%% End:

### 2012-02-15 presentation/pagai.tex
presentation de pagai : tableau de chiffres + slide assert

@@
 \usepackage{listings}
+\usepackage{array}
 
@@
 	\frametitle{Pagai}
-	Compute numerical invariants from an LLVM bitcode file
+
+	\begin{center}
+	Computes numerical invariants from an LLVM bitcode file
+	\end{center}
 
@@
 \begin{column}{10cm}
-	\includegraphics[width=7cm]{f}
+	\includegraphics[width=7cm]{f.pdf}
 \end{column}
@@
 \end{verbatim}
+\end{frame}
+
+\begin{frame}[containsverbatim]
+	\frametitle{Assert}
+Pagai can check that assert are always true.
+\begin{columns}
+	\lstinputlisting{gopan_assert.c}
+\begin{column}{1cm}
+\end{column}
+\begin{column}{10cm}
+\begin{verbatim}
+RESULT FOR BASICBLOCK: ------
+; <label>:15                                      
+  call void @assert_fail(...) 
+  unreachable
+-----
+empty polyhedron of dim (0,0)
+ASSERT OK
+\end{verbatim}
+\end{column}
+\end{columns}
 \end{frame}
@@
 
+\begin{frame}
+	\frametitle{Scalability}
+
+\begin{table}[!h]
+	\centering
+\begin{tabular}{|l|r|r||r|r|r|r|r|} \hline
+	 	 & & &
+        \multicolumn{5}{|c|}{Time (in seconds)} 
+		\\ \cline{4-8}
+		Name & kLOC & $|P_R|$ & S & G & PF & C & DIS \\ \hline
+		\input{table_time}
+	\hline
+\end{tabular}
+\label{fig:projects}
+\end{table}
+
+\end{frame}
+
 \end{document}

### 2012-02-20 papers/SAS2012/rlim_int_loop.c
Missing files to compile SAS

@@ -1,0 +1,12 @@
+extern int input();
+extern int wait();
+void rate_limiter() {
+  int x_old = 0;
+  while (1) {
+    int x = input(-100000, 100000);
+    if (x > x_old+10) x = x_old+10;
+    if (x < x_old-10) x = x_old-10;
+    x_old = x;
+    while (wait()) {}
+  }
+}

### 2012-02-21 papers/SAS2012/implicitization.tex
Rabottage, et correction de mise en forme suite à passage LLNCS

@@
 \usepackage{algorithmic,algorithm}
-%\usepackage{graphicx}
+\usepackage{paralist}
+\usepackage{graphicx}
 %\usepackage{color}
@@
 \usepackage{tikz}
-\usetikzlibrary{arrows,automata}
+\usetikzlibrary{arrows,automata,positioning}
 
@@
 
+\newcommand{\avirer}[1]{ {\color{red} #1} }
+\newcommand{\MM}[1]{{\color{blue} TODO(MM): #1}}
+%\newcommand{\JH}[1]{{\color{teal} TODO(JH): #1}}
+%\newcommand{\DM}[1]{{\color{violet} TODO(DM): #1}}
+%\newcommand{\MM}[1]{}
+%\newcommand{\JH}[1]{}
+%\newcommand{\DM}[1]{}
+
 \title{\mytitle}
@@
 \section{Bases}
-\subsection{Static analysis by abstract interpretation}
+\subsection{Static Analysis by Abstract Interpretation}
 \label{sec:static_analysis}
+\MM{On peut peut-être réduire un peu vu que tout le monde connait à
+  SAS ?}
+
 Let $X$ be the set of possible states of the program variables; for instance, if the program has 3 unbounded integer variables, then $X = \ZZ^3$.
@@
 
-Widening operations have a somewhat counterintuitive behavior: the result of an analysis (postcondition) is not necessarily monotonic with respect to the precondition.
-For instance, classical interval analysis with widening and decreasing iterations on the following loop yields a loop invariant $x \in [0,+\infty)$ if the precondition is $x \in [0,0]$ , but $x \in [0,10]$ (strictly stronger) if the precondition is $x \in [0,10]$ (strictly weaker):
-\lstinline|while (x != 10) x=x+1;|.
+Widening operations have a somewhat counterintuitive behavior: the
+result of an analysis (postcondition) is not necessarily monotonic
+with respect to the precondition. For instance, classical interval
+analysis with widening and decreasing iterations on the loop
+\lstinline|while (x != 10) x=x+1;| yields a loop invariant $x \in
+[0,+\infty)$ if the precondition is $x \in [0,0]$, but $x \in [0,10]$
+(strictly stronger) if the precondition is $x \in [0,10]$ (strictly
+weaker).
 This means that, when using widenings, being locally more precise (for instance, by using more expressive abstract domains, or widening operators that climb more slowly \cite{BagnaraHRZ05SCP}) does not necessarily translate into higher final precision.
@@
 intervals), but are these examples representative of real programs?
-This justifies our recourse to evaluation on realistic examples~(\S~\ref{sec:compare_techniques}).
+This justifies our recourse to evaluation on real, large programs~(\S~\ref{sec:compare_techniques}).
 
@@
 %\footnote{\url{http://research.microsoft.com/en-us/um/redmond/projects/z3/}}
-Yices,%
+Yices,
 %\footnote{\url{http://yices.csl.sri.com/}}
@@
 
-\subsection{A simple, motivating example}
+\subsection{A Simple, Motivating Example}
 \label{subsec:rate_lim}
@@
 This program implements a construct commonly found in control programs (in e.g.
-automotive or avionics): a rate limiter. For the sake of simplicity, we chose it
-to be fed on a nondeterministic input clamped between $[-100000,100000]$, but in a real system it would be integrated in a reactive control loop and its input connected to a complex system with unknown output range.
+automotive or avionics): a rate limiter.
+\avirer{For the sake of simplicity, we chose it
+to be fed on a nondeterministic input clamped between $[-100000,100000]$, but in a real system it would be integrated in a reactive control loop and its input connected to a complex system with unknown output range.}
 
@@
 \centering
-\begin{minipage}[c]{.19\textwidth}
-\begin{tikzpicture}[->,>=stealth',auto,node distance=1.1cm,
+\begin{minipage}[c]{.21\textwidth}
+\begin{tikzpicture}[->,>=stealth',auto,node distance=.2cm and .5cm,
                     semithick,font=\footnotesize]
@@
 	\node[PRstate] (n0) {$p$};
-	\node[state] (n1) [below left of=n0] {};
-	\node[state] (n2) [below right of=n0] {};
-	\node[state] (n3) [below right of=n1] {};
-	\node[state] (n4) [below left of=n3] {};
-	\node[state] (n5) [below right of=n3] {};
-	\node[state] (n6) [below right of=n4] {};
-	\node[state] (n7) [below left of=n6] {};
-	\node[state] (n8) [below right of=n6] {};
-	\node[PRstate] (n9) [below right of=n7] {$q$};
-
-  \path [transition] 
-		(n0) edge              node {} (n1);
-  \path [transition] 
-		(n0) edge              node {} (n2);
-  \path [transition] 
-		(n1) edge              node {} (n3);
-  \path [transition] 
-		(n2) edge              node {} (n3);
-  \path [transition] 
-		(n3) edge              node {} (n4);
-  \path [transition] 
-		(n3) edge              node {} (n5);
-  \path [transition] 
-		(n4) edge              node {} (n6);
-  \path [transition] 
-		(n5) edge              node {} (n6);
-  \path [transition] 
-		(n6) edge              node {} (n7);
-  \path [transition] 
-		(n6) edge              node {} (n8);
-  \path [transition] 
-		(n7) edge              node {} (n9);
-  \path [transition] 
-		(n8) edge              node {} (n9);
+	\node[state] (n1) [below left=of n0] {};
+	\node[state] (n2) [below right=of n0] {};
+	\node[state] (n3) [below right=of n1] {};
+	\node[state] (n4) [below left=of n3] {};
+	\node[state] (n5) [below right=of n3] {};
+	\node[state] (n6) [below right=of n4] {};
+	\node[state] (n7) [below left=of n6] {};
+	\node[state] (n8) [below right=of n6] {};
+	\node[PRstate] (n9) [below right=of n7] {$q$};
+
+  \path [transition] (n0) -- (n1);
+  \path [transition] (n0) -- (n2);
+  \path [transition] (n1) -- (n3);
+  \path [transition] (n2) -- (n3);
+  \path [transition] (n3) -- (n4);
+  \path [transition] (n3) -- (n5);
+  \path [transition] (n4) -- (n6);
+  \path [transition] (n5) -- (n6);
+  \path [transition] (n6) -- (n7);
+  \path [transition] (n6) -- (n8);
+  \path [transition] (n7) -- (n9);
+  \path [transition] (n8) -- (n9);
 \end{tikzpicture}
@@
 \begin{minipage}[c]{.19\textwidth}
-\begin{tikzpicture}[->,>=stealth',auto,node distance=1.1cm,
+\begin{tikzpicture}[->,>=stealth',auto,node distance=.2cm and .5cm,
                     semithick,font=\footnotesize]
@@
 	\node[PRstate] (n0) {$p$};
-	\node (n1) [below left of=n0] {};
-	\node (n2) [below right of=n0] {};
-	\node (n3) [below right of=n1] {};
-	\node (n4) [below left of=n3] {};
-	\node (n5) [below right of=n3] {};
-	\node (n6) [below right of=n4] {};
-	\node (n7) [below left of=n6] {};
-	\node (n8) [below right of=n6] {};
-	\node[PRstate] (n9) [below right of=n7] {$q$};
+	\node (n1) [below left=of n0] {};
+	\node (n2) [below right=of n0] {};
+	\node (n3) [below right=of n1] {};
+	\node (n4) [below left=of n3] {};
+	\node (n5) [below right=of n3] {};
+	\node (n6) [below right=of n4] {};
+	\node (n7) [below left=of n6] {};
+	\node (n8) [below right=of n6] {};
+	\node[PRstate] (n9) [below right=of n7] {$q$};
 
@@
 
-\subsection{Guided static analysis}
+\subsection{Guided Static Analysis}
 \label{sec:guided}
 \emph{Guided static analysis} was proposed by \cite{DBLP:conf/sas/GopanR07} as an improvement over classical upward Kleene iterations with widening.
-Consider the following program, taken from \cite{DBLP:conf/sas/GopanR07}:
+Consider the program in Fig.~\ref{fig:gopan_reps_invariant}, taken from \cite{DBLP:conf/sas/GopanR07}.
+
+\begin{figure}
+  \begin{minipage}{.3\linewidth}
     \lstinputlisting[label=lst:gopan_reps]{gopan_reps.c}
-
-\begin{figure}
-\begin{center}
+  \end{minipage}
+\hfill
+\begin{minipage}{.5\linewidth}
 \includegraphics[scale=0.75]{gopan_reps}
-\end{center}
-\caption{The invariant for Prog.~\ref{lst:gopan_reps}: the piecewise linear, solid line is the strongest invariant, the grayed polyhedron is its convex hull.}
+\vspace{-1ex}
+\begin{equation}
+y \leq x \;\land\; y \leq 102-x \;\land\; y \geq 0.\label{eqn:triangle}
+\end{equation}
+\end{minipage}
+\caption{Example program and its invariant: the piecewise linear, solid line is the strongest invariant, the grayed polyhedron is its convex hull.}
 \label{fig:gopan_reps_invariant} 
@@
 From there, the ``else'' branch at line 5 may be taken; with further widening, $0 \leq y \leq x$ is obtained as a loop invariant, and thus the postcondition computed at line 9 is $x \geq 0 \land y = 0$.
-Yet the strongest invariant is $(0 \leq x \leq 51 \land y = x) \lor (51 \leq x \leq 102 \land x+y=102)$, and its convex hull, a convex polyhedron (Fig.~\ref{fig:gopan_reps_invariant}), is
-\begin{equation}
-y \leq x \land y \leq 102-x \land y \geq 0.\label{eqn:triangle}
-\end{equation}
+Yet the strongest invariant is $(0 \leq x \leq 51 \land y = x) \lor (51 \leq x \leq 102 \land x+y=102)$, and its convex hull, a convex polyhedron (Fig.~\ref{fig:gopan_reps_invariant}).
+
 Intuitively, this disappointing result is obtained because widening extrapolates from the first iterations of the loop, but the loop has two different phases ($x \leq 50$ and $x > 50$) with different behaviors, thus the extrapolation from the first phase is not valid for the second.
@@
 variable is only assigned once in the transition graph. This is not a
-restriction, since there exist standard algorithms that transform a program into
+restriction, since there exists standard algorithms that transform a program into
 an SSA format.
@@
 In order to find focus paths, we solve an SMT formula which is satisfiable when
-there exist a path starting at a point $p_i \in P_R$ in a state included in the
+there exists a path starting at a point $p_i \in P_R$ in a state included in the
 current invariant candidate $X_i$, and arriving at a point $p_j \in P_R$ in a
@@
 
-\section{Guided analysis over the paths}
+\section{Guided Analysis over the Paths}
 \label{sec:guided_multigraph}
@@
 
-\subsection{Ascending iterations by Path-focusing}
+\subsection{Ascending Iterations by Path-focusing}
 \label{subsec:ascending}
@@
 associated to $p_i$ :
-$$f(p_i) = \rho \wedge b_i^s \wedge 
+\[
+f(p_i) = \rho \wedge b_i^s \wedge 
 \displaystyle\bigwedge_{j \in P_R \atop j \neq i} \neg
 b_j^s \wedge X_i^s \wedge \displaystyle\bigvee_{j \in Succ(i)} (b_j^d \wedge
- \neg X_j^s)$$
-
+ \neg X_j^s)
+\]
 The difference with~\cite{Monniaux_Gonnord_SAS11} is that we do not
@@
 
-\subsection{Adding new paths}
+\subsection{Adding New Paths}
 \label{subsec:addingpaths}
@@
 $f'(p_i)$, where:
-$$f'(p_i) = \rho \wedge b_i^s \wedge 
+\[
+f'(p_i) = \rho \wedge b_i^s \wedge 
 \displaystyle\bigwedge_{j \in P_R \atop j \neq i} \neg
 b_j^s \wedge X_i^s \wedge \displaystyle\bigvee_{j \in Succ(i)} (b_j^d \wedge
-\neg X_j^d)$$
-
+\neg X_j^d)
+\]
 $X_j^d$ is updated when the point $p_j$ is the target of a new path.
@@
 Termination of this algorithm is guaranteed, because:
-\begin{itemize}
+\begin{inparaenum}
 \item 
@@
 will eventually be empty after a finite number of loop iterations.
-\end{itemize}
+\end{inparaenum}
 
@@
 widening/narrowing sequence or accelerates the loop, thus leading to a precise
-invariant. However, in some cases, there also exist paths that are not
+invariant. However, in some cases, there also exists paths that are not
 self loops, in which case \emph{Path-focusing} applies widening. 
@@
 through the loop from the others:
-\begin{lstlisting}[numbers=none]
-if (wait()) {
-   while (wait()) {}
-}
-\end{lstlisting}
+\lstinline|if (wait()) {while (wait()) {}}|
 
@@
 		$p_{10}$, that goes through the \emph{else} branch of each 
-		\emph{if-then-else} at lines 7,8 and 9, is also 
+		\emph{if-then-else} at lines 7, 8 and 9, is also 
 		$-10 \leq \lstinline|x_old| \leq 10$, and thus is also added into the
@@
 \end{minipage} 
-\begin{minipage}[c]{.28\textwidth}
+\begin{minipage}[c]{.75\textwidth}
 \begin{tikzpicture}[->,>=stealth',auto,node distance=2.1cm,
@@
   \path [transition] 
-		(n0) edge node [left, yshift=-0.2cm, xshift=0.2cm] {
+		(n0) edge node [left, yshift=-0.2cm, xshift=0] {
 		$\begin{array}{r}
@@
 
-\section{Disjunctive invariants}
+\section{Disjunctive Invariants}
 \label{sec:disjunctive}
@@
 $\sigma_i(j,k)$, thus know the index of the disjunct to join with.
-
+%
 \begin{equation*}
@@
  \wedge 
-\displaystyle\bigvee_{1 \leq k \leq m_i} (d_k \wedge X_{i,k} \wedge \bigwedge_{l \neq k}
+\displaystyle\bigvee_{1 \leq k \leq m_i}\!\! (d_k \wedge X_{i,k} \wedge \bigwedge_{l \neq k}
 \neg d_l)
 \wedge
-\displaystyle\bigvee_{j \in Succ(i)} 
-(b_j^d \wedge \bigwedge_{1 \leq k \leq m_i} (\neg X_{j,k}))
+\displaystyle\bigvee_{j \in Succ(i)}\!\!
+(b_j^d \wedge \bigwedge_{1 \leq k \leq m_i}\!\! (\neg X_{j,k}))
 \end{equation*}
-
+%
 In our algorithm, the initialization of the abstract values slightly differs from
 algorithm \ref{algo:combined} line~\ref{alg=X-init}, since we now have to
-initialize each disjunct. Line~\ref{alg=X-init} is then replaced by:
-\begin{algorithm}[!h]
-\begin{algorithmic}[1] 
-\FORALL {$k \in \{1,..,m_i\} \setminus \{\delta_i\}$}
-	\STATE $X_{i,k} \gets \perp$
-\ENDFOR
-\STATE $X_{i,\delta_i} \gets I_{p_i}$
-\end{algorithmic}
-\end{algorithm}
+initialize each disjunct. Instead of Line~\ref{alg=X-init}, we initialize $X_{i,k}$ with $\perp$ for all
+$k \in \{1,..,m_i\} \setminus \{\delta_i\}$, and $X_{i,\delta_i}$ with
+$\gets I_{p_i}$.
+
+% MM: pas la place pour un vrai algo => inliné dans le texte.
+% Line~\ref{alg=X-init} is then replaced by:
+% %\begin{algorithm}[!h]
+% \begin{algorithmic}[1] 
+% \FORALL {$k \in \{1,..,m_i\} \setminus \{\delta_i\}$}
+% 	\STATE $X_{i,k} \gets \perp$
+% \ENDFOR
+% \STATE $X_{i,\delta_i} \gets I_{p_i}$
+% \end{algorithmic}
+% %\end{algorithm}
 
@@
 
-The \emph{Update} function can classically assign to $X_{i,\sigma_i(j,k)}$ the
-value $X_{i,\sigma_i(j,k)} \widening (X_{i,\sigma_i(j,k)} \sqcup
-\tau_{i,k}(X_{i,j}))$, or can integrate the special treatment for self loops
+The \emph{Update} function can classically assign  $X_{i,\sigma_i(j,k)} \widening (X_{i,\sigma_i(j,k)} \sqcup
+\tau_{i,k}(X_{i,j}))$ to $X_{i,\sigma_i(j,k)}$, or can integrate the special treatment for self loops
 proposed by \cite{Monniaux_Gonnord_SAS11}, with widening/narrowing sequence or
@@
 
-\section{Experimental comparisons}
+\section{Implementation and Experimental Comparisons}
 \label{sec:experiments}
@@
 For SMT-solving, our analyzer uses Yices
-\cite{DBLP:conf/cav/DutertreM06} or Microsoft Z3\cite{DBLP:conf/tacas/MouraB08}.
+\cite{DBLP:conf/cav/DutertreM06} or Microsoft Z3~\cite{DBLP:conf/tacas/MouraB08}.
 
@@
 
-\subsection{Analysis algorithm}
+\subsection{Analysis Algorithm}
 \label{sec:analysis-algorithm}
@@
 
-\subsection{Precision of the various techniques}
+\subsection{Precision of the Various Techniques}
 \label{sec:compare_techniques}
@@
 
-\section{Conclusion and future prospects}
+\section{Conclusion and Future Prospects}
 Roughly, an analysis by abstract interpretation is defined by the choice of an iteration strategy and an abstract domain.

### 2012-02-21 papers/SAS2012/implicitization.tex
On cite le nom de l'outil (PAGAI) maintenant que c'est plus anonyme

@@
 We have implemented our proposed solutions inside a prototype of intraprocedural
-static analyzer,
+static analyzer called PAGAI,
 as well as the classical abstract interpretation algorithm, and the state-of-the-art
@@
 %
-Our tool operates over LLVM bitcode \cite{LLVM_langref,Lattner:2004:LCF:977395.977673}, which is a target for several compilers, most notably Clang (supporting C and C++) and llvm-gcc (supporting C, C++, Fortran and Ada).
+PAGAI operates over LLVM bitcode \cite{LLVM_langref,Lattner:2004:LCF:977395.977673}, which is a target for several compilers, most notably Clang (supporting C and C++) and llvm-gcc (supporting C, C++, Fortran and Ada).
 Abstract domains are provided by the APRON library \cite{DBLP:conf/cav/JeannetM09}, and include convex polyhedra (from the builtin Polka ``PK'' library), octagons, intervals, and linear congruences.

### 2012-02-21 papers/SAS2012/implicitization.tex
Continuation du tassage

@@
 %\newcommand{\DM}[1]{{\color{violet} TODO(DM): #1}}
+%\newcommand{\avirer}[1]{}
 %\newcommand{\MM}[1]{}
@@
 
-In such an analysis, information may flow forward (one computes the polyhedron after a program statement as the image by the semantics of the statement, or a super-set thereof, of the polyhedron before) or backward;
+In such an analysis, information may flow forward \avirer{(one computes the polyhedron after a program statement as the image by the semantics of the statement, or a super-set thereof, of the polyhedron before)} or backward;
 forward program analysis computes super-sets of the states reachable from the initialization of the program, backward program analysis computes super-sets of the states co-reachable from some property of interest (for instance, the violation of an assertion).
@@
 
-All these methods improve the precision of the analysis by keeping the same abstract domain (say, convex polyhedra) but changing the operations applied and their ordering.
-An alternative is to change the abstract domain, for instance by moving from intervals to octagons or convex polyhedra \cite{DBLP:journals/lisp/Mine06}, or the widening operator: for instance one may replace the classical widening on convex polyhedra \cite{CousotHalbwachs78,Halbwachs_PhD} by an improved version \cite{BagnaraHRZ05SCP}, or use widening ``up to'', that is, use linear inequalities syntactically present in the program source code as possible widening steps~\cite{Polka:FMSD:97}.
-
-There are many possible combinations of the above techniques, and it is not
+All these methods improve the precision of the analysis by keeping the
+same abstract domain (say, convex polyhedra) but changing the
+operations applied and their ordering. An alternative is to change the
+abstract domain (e.g. octagons, convex polyhedra
+\cite{DBLP:journals/lisp/Mine06}), or the widening
+operator~\cite{BagnaraHRZ05SCP,Polka:FMSD:97}.
+
+\avirer{There are many possible combinations of the above techniques, and it is not
 evident which ones perform more or less precisely or more or less efficiently on real-life examples. One needs to experiment.
-Unfortunately, the published literature on the subject lacks experimental comparative assessments. One purpose of this article is therefore to propose such experimental results.
+Unfortunately, the published literature on the subject lacks experimental comparative assessments. One purpose of this article is therefore to propose such experimental results.}
 
@@
 \begin{enumerate}
-\item We recast the guided static analysis technique from \cite{DBLP:conf/sas/GopanR07} on the expanded multigraph from \cite{Monniaux_Gonnord_SAS11}, considering entire paths instead of individual transitions, using SMT queries and binary decision diagrams.\label{contr:guided_multigraph} (See \S\ref{sec:guided_multigraph})
-\item We improve the technique for obtaining disjunctive invariants from \cite{DBLP:conf/pldi/GulwaniZ10} by replacing the explicit exhaustive enumeration of paths by a sequence of SMT queries.\label{contr:disjunctive} (See \S\ref{sec:disjunctive})
+\item We recast the guided static analysis technique from \cite{DBLP:conf/sas/GopanR07} on the expanded multigraph from \cite{Monniaux_Gonnord_SAS11}, considering entire paths instead of individual transitions, using SMT queries and binary decision diagrams\label{contr:guided_multigraph} (See \S\ref{sec:guided_multigraph}).
+\item We improve the technique for obtaining disjunctive invariants from \cite{DBLP:conf/pldi/GulwaniZ10} by replacing the explicit exhaustive enumeration of paths by a sequence of SMT queries\label{contr:disjunctive} (See \S\ref{sec:disjunctive}).
 \item We implemented these techniques, in addition to ``classical'' iterations and the original guided static analysis, inside a prototype static analyzer.
 This tool uses the LLVM bitcode format \cite{Lattner:2004:LCF:977395.977673,LLVM_langref} as input, which can be produced by compilation from C, C++ and Fortran, enabling it to be run on many real-life programs.
-With respect to abstract domains on numerical variables, it uses the APRON library \cite{DBLP:conf/cav/JeannetM09}, which supports a variety of abstract domains from which we can choose with minimal changes to our analyzer.
-
+It uses the APRON library \cite{DBLP:conf/cav/JeannetM09}, which
+supports a variety of abstract domains for numerical variables, from
+which we can choose with minimal changes to our analyzer.
+
+\avirer{
 Our tool uses a \emph{single static assignment} (SSA) intermediate representation, which allows us to perform cheap symbolic propagations and simplifications,
 described in Section~\ref{sec:analysis-algorithm}.
+}
 
@@
 result of an analysis (postcondition) is not necessarily monotonic
-with respect to the precondition. For instance, classical interval
+with respect to the precondition.
+\avirer{
+For instance, classical interval
 analysis with widening and decreasing iterations on the loop
@@
 weaker).
-This means that, when using widenings, being locally more precise (for instance, by using more expressive abstract domains, or widening operators that climb more slowly \cite{BagnaraHRZ05SCP}) does not necessarily translate into higher final precision.
+}
+This means that, when using widenings, being locally more precise \avirer{(for instance, by using more expressive abstract domains, or widening operators that climb more slowly \cite{BagnaraHRZ05SCP})} does not necessarily translate into higher final precision.
+\avirer{
 It is almost always possible to concoct examples where a supposedly more precise abstraction (e.g.
@@
 intervals), but are these examples representative of real programs?
+}
 This justifies our recourse to evaluation on real, large programs~(\S~\ref{sec:compare_techniques}).
@@
 In concrete terms, an assignment \lstinline|x=2*x+1;| gets translated into a definition $x_2 = 2x_1+1$, with distinct variables $x_1$ and $x_2$ corresponding to the same original variable \lstinline|x| at different points in the program.
-LLVM makes it easy to follow definition-use and use-definition chains: for a given variable (say, $x_2$) one can immediately obtain its definition (say, $2x_1+1$).
-One may see conversion to SSA form as a static precomputation of some of the symbolic propagations proposed by \cite{DBLP:conf/vmcai/Mine06} to enhance the precision of analyses.
-
-SSA introduces $\phi$-functions at the head of a control code to define variables whose value depends on which incoming edge to this control node was last taken.
-For instance, for \lstinline|if (...) { x = 2*x+1; } else { x= 0; }|, then $x_2$ is defined as $\phi(2x_1+1,0)$.
-In this framework, each variable is uniquely defined as an arithmetic ($+$, $-$, $\times$, $/$) function of other variables that themselves may not be representable as affine linear functions, because they are defined using $\phi$-functions, bitwise arithmetic, loads from memory, or return values from function calls.
+\avirer{LLVM makes it easy to follow definition-use and use-definition chains: for a given variable (say, $x_2$) one can immediately obtain its definition (say, $2x_1+1$).
+One may see conversion to SSA form as a static precomputation of some of the symbolic propagations proposed by \cite{DBLP:conf/vmcai/Mine06} to enhance the precision of analyses.}
+
+SSA introduces $\phi$-functions at the head of a control code to
+define variables whose value depends on which incoming edge to this
+control node was last taken. For instance, for
+\lstinline|if (...) { x = 2*x+1; } else { x= 0; }|, then $x_2$ is defined as
+$\phi(2x_1+1,0)$. In this framework, variables are either defined as
+an arithmetic ($+$, $-$, $\times$, $/$) function of other variables,
+or not be representable as affine linear functions, because they are
+defined using $\phi$-functions, bitwise arithmetic, loads from memory,
+or return values from function calls.
 

### 2012-02-22 papers/SAS2012/implicitization.tex
un booléen pour activer/désactiver les commentaires
Encore un peu de retaillage

@@
 
-\newcommand{\avirer}[1]{ {\color{red} #1} }
+\usepackage{ifthen}
+\newboolean{draft}
+%\setboolean{draft}{true}
+\setboolean{draft}{false}
+
+\ifthenelse{\boolean{draft}}{
+  \newcommand{\avirer}[1]{{\color{red} #1}}
   \newcommand{\MM}[1]{{\color{blue} TODO(MM): #1}}
-%\newcommand{\JH}[1]{{\color{teal} TODO(JH): #1}}
-%\newcommand{\DM}[1]{{\color{violet} TODO(DM): #1}}
-%\newcommand{\avirer}[1]{}
-%\newcommand{\MM}[1]{}
-%\newcommand{\JH}[1]{}
-%\newcommand{\DM}[1]{}
+  \newcommand{\JH}[1]{{\color{teal} TODO(JH): #1}}
+  \newcommand{\DM}[1]{{\color{violet} TODO(DM): #1}}
+}{
+  \newcommand{\avirer}[1]{}
+  \newcommand{\MM}[1]{}
+  \newcommand{\JH}[1]{}
+  \newcommand{\DM}[1]{}
+}
 
@@
 
+\avirer{
 \begin{figure}
+\MM{J'aime bien cette figure, qui montre en un coup d'oeil l'idée,
+  mais elle prend vraiment beaucoup de place, je propose de
+  supprimer.}
+
 \centering
@@
 \end{figure}
+}
 
 Classical interval analysis on this program yields $\lstinline|x_old| \in [-100000,100000]$.
-We have transformed the first program into the second, manually pruning out infeasible paths; yet in general the resulting program could be exponentially larger than the first (as in Fig.~\ref{fig:multigraph}), even though not all feasible paths are needed to compute the invariant.
+We have transformed the first program into the second, manually pruning out infeasible paths; yet in general the resulting program could be exponentially larger than the first \avirer{(as in Fig.~\ref{fig:multigraph})}, even though not all feasible paths are needed to compute the invariant.
 
@@
 \begin{minipage}{.5\linewidth}
-\includegraphics[scale=0.75]{gopan_reps}
-\vspace{-1ex}
+\includegraphics[scale=0.7]{gopan_reps}
+\vspace{-1em}
 \begin{equation}

### 2012-02-22 papers/SAS2012/implicitization.tex
Fusion des deux tableaux benchmarks

@@
       \centering
-\begin{tabular}{|l|r|r|} \hline
+      \begin{tabular}{|l|r|r|r|r|r|r|r|} \hline
+        \multicolumn{1}{|c|}{} &
+        \multicolumn{2}{c|}{Size} &
+        \multicolumn{5}{c|}{Execution time (seconds)} \\ \hline
         \multicolumn{1}{|c|}{Name} & 
         \multicolumn{1}{c|}{kLOC}  & 
-        \multicolumn{1}{c|}{$|P_R|$} \\ \hline
-	a2ps & 55 & 2012\\
-	gawk & 59 & 902\\ 
-	gnuchess & 38 & 1222\\ 
-	gnugo & 83 & 2801\\
-	grep & 35 & 820\\
-	gzip & 27 & 494\\
-	lapack/blas & 954 & 16422\\
-	make & 34 & 993\\ 
-	sed & 23 & 292\\
-	tar & 73 & 1712\\
+        \multicolumn{1}{c|}{$|P_R|$}  &
+        \multicolumn{1}{c|}{\textbf{S}} &
+        \multicolumn{1}{c|}{\textbf{G}} &
+        \multicolumn{1}{c|}{\textbf{PF}} &
+        \multicolumn{1}{c|}{\textbf{G+PF}} &
+	\multicolumn{1}{c|}{\textbf{DIS}} \\ 
+        \hline
+        % M-x align RET powers ;-).
+a2ps-4.14      & 55  & 2012  & 64  & 17  & 66   & 175  & 512   \\
+gawk-4.0.0     & 59  & 902   & 59  & 8   & 18   & 43   & 129   \\
+gnuchess-6.0.0 & 38  & 1222  & 112 & 21  & 122  & 374  & 866   \\
+gnugo-3.8      & 83  & 2801  & 155 & 31  & 128  & 382  & 1403  \\
+grep-2.9       & 35  & 820   & 50  & 12  & 21   & 58   & 307   \\
+gzip-1.4       & 27  & 494   & 63  & 12  & 36   & 97   & 617   \\
+lapack-3.3.1   & 954 & 16422 & 721 & 246 & 2690 & 6052 & 22222 \\
+make-3.82      & 34  & 993   & 22  & 7   & 32   & 83   & 255   \\
+sed-4.2        & 23  & 292   & 35  & 7   & 23   & 69   & 521   \\
+tar-1.26       & 73  & 1712  & 270 & 39  & 110  & 304  & 1164  \\
+% TODO: mettre à jour les chiffres ?
 	\hline
@@
 precision of the analysis (for $12.06\%$ of the control points in $P_R$), at an
-acceptable time cost (see Table \ref{tab:time}).
+acceptable time cost (see Table \ref{tab:time}). \MM{TODO}
 
@@
 
-Table \ref{tab:time} gives the execution time of the different analysis
+Table \ref{tab:time} \MM{TODO} gives the execution time of the different analysis
 techniques. It is interesting to see that \emph{Guided Static Analysis} and
@@
 The techniques we introduce in this article are slower, but speed is still on the same order of magnitude as the older ones.
-
-\begin{table}[!h]
-	\centering
-\begin{tabular}{|l|r|r|r|r|r|} \hline
-	\multicolumn{1}{|c|}{Benchmark} &
-        \multicolumn{1}{c|}{\textbf{S}} &
-        \multicolumn{1}{c|}{\textbf{G}} &
-        \multicolumn{1}{c|}{\textbf{PF}} &
-        \multicolumn{1}{c|}{\textbf{G+PF}} &
-	\multicolumn{1}{c|}{\textbf{DIS}} \\ \hline
-	\input{table_time.tex} \hline
-\end{tabular}
-\caption{Execution time for each technique, expressed in seconds}
-\label{tab:time}
-\end{table}
 

### 2012-02-22 ex/pagai.sh
scripts

@@ -13,9 +13,12 @@ OPTIONS :
 	-u        : unroll loops once
 	-s        : silent mode
 	-c        : compare all techniques
+	-M        : compare narrowing
 	-G        : generate the .dot CFG
 	-y        : use Yices instead of Microsoft Z3
 	-b        : use the Bagnara Widening operator
+	-d        : domain 1
+	-D        : domain 2
 	-p        : use the trunk version of pagai
 	-t        : set a time limit (Pagai is killed after this time, default 800s)
 "
@@ -26,14 +29,17 @@ GRAPH=0
 YICES=0
 UNROLL=0
 COMPARE=0
+NCOMPARE=0
 TIME_LIMIT=800
 RESULT=
 BITCODE=
 PAGAI=Pagai
 REQUIRED=0
 SILENT=0
+DOMAIN1=pk
+DOMAIN2=pk
 
-while getopts "hpygrbuGi:o:ct:O:sp" opt ; do
+while getopts "hpygrbuGi:o:ct:O:spMd:D:" opt ; do
 	case $opt in
 		h)
 			usage
@@ -43,6 +49,12 @@ while getopts "hpygrbuGi:o:ct:O:sp" opt ; do
 			FILENAME=$OPTARG
 			REQUIRED=1
 			;;
+		d)
+			DOMAIN1=$OPTARG
+			;;
+		D)
+			DOMAIN2=$OPTARG
+			;;
 		u)
 			UNROLL=1
 			;;
@@ -51,6 +63,9 @@ while getopts "hpygrbuGi:o:ct:O:sp" opt ; do
 			;;
 		c)
 			COMPARE=1
+			;;
+		M)
+			NCOMPARE=1
 			;;
 		G)
 			GRAPH=1
@@ -119,25 +134,28 @@ RESULT=/tmp/${NAME%%.*}.result
 
 ulimit -t $TIME_LIMIT
 
+ARGS="--domain $DOMAIN1 --domain2 $DOMAIN2 "
+
 if [ -z $OUTPUT ] ; then 
 	if [ $COMPARE -eq 1 ] ; then
-			$PAGAI -c -i $BITCODE
+		$PAGAI $ARGS -c -i $BITCODE
+	elif [ $NCOMPARE -eq 1 ] ; then
[…]

### 2012-02-22 papers/SAS2012/implicitization.tex
Création d'une partie \appendix
Pour l'instant, le découpage est très arbitraire, mais au moins on n'est
pas loin des 15 pages.
Compactage

@@
 
-\item We conducted extensive experiments with this tool, on real-life programs: (i) fixing the abstract domain, varying the iteration technique (\S~\ref{sec:compare_techniques}) (ii) fixing the iteration technique, varying the abstract domain (\S\ref{sec:compare_domains}).
+\item We conducted extensive experiments with this tool, on real-life programs: (i) fixing the abstract domain, varying the iteration technique (\S\ref{sec:compare_techniques}) (ii) fixing the iteration technique, varying the abstract domain (\S\ref{sec:compare_domains}).
 \end{enumerate}
@@
 
-More generally, this analysis method considers an ascending sequence of subsets of the transitions in the loop body (left side of Fig.~\ref{fig:multigraph});
+More generally, this analysis method considers an ascending sequence of subsets of the transitions in the loop body \avirer{(left side of Fig.~\ref{fig:multigraph})};
 for each subset, an inductive invariant is computed for the program restricted to it.
@@
 Algorithm~\ref{algo:combined} performs Guided static analysis on the implicitly represented multigraph.
-
+%
 The current working subset of paths, noted $P$ and initially empty, is
@@
 diagrams. We also maintain two sets of control points:
-\begin{itemize}
-	\item $A'$ contains the points in $P_R$ that may be the starting points of new
+\begin{compactitem}
+	\item $A'$ : points in $P_R$ that may be the starting points of new
 		feasible paths.
-	\item $A$ contains the points in $P_R$ on which we apply the ascending iterations.
-	Each time the abstract value of a control point $p$ is updated, $p$ is
-	inserted in both $A$ and $A'$.
-\end{itemize}
+	\item $A$ : points in $P_R$ on which we apply the ascending iterations.
+	When the abstract value of a control point $p$ is updated, $p$ is
+	added to both $A$ and $A'$.
+\end{compactitem}
 
@@
 We distinguish three phases in the main loop of the analysis:
-\begin{enumerate}
+\begin{compactenum}
 \item \label{step:addingpaths} We start finding a new relevant subset
@@
   i.e. $A' = \emptyset$.
-\end{enumerate}
+\end{compactenum}
 
@@
 
-Suppose the main loop of the rate limiter contains a nested loop:
+Suppose the main loop of the rate limiter contains a nested loop like:
 \lstinputlisting{rlim_int_loop.c}
@@
 \caption{List of analyzed open-source projects, with their respective number of
-lines of code, and their number of control points in $P_R$}
+lines of code, number of control points in $P_R$, and execution time
+on various techniques\MM{TODO}}
 \label{fig:projects}
 \end{table}
+
+\section{Conclusion and Future Prospects}
+Roughly, an analysis by abstract interpretation is defined by the choice of an iteration strategy and an abstract domain.
+In this article, we demonstrated that changes in the iteration algorithm can significantly improve precision, sometimes while improving analysis times.
+
+A common criticism of analysis techniques based on SMT-solving is that they
+do not scale up. Yet, our experiments show that, for numerical properties,
+they scale up to the size of typical functions and loops.
+It is however
+quite certain that, naively applied, they cannot scale to the kind of
+programs targeted by e.g. the Astr\'ee tool, that is, a dozens or hundreds
+of thousands of lines of code in a single loop operating over similar numbers
+of remanent variables.
+Actually, for such applications, only (quasi-)linear
+algorithms scale up, and ``cheap'' abstract domains such as octagons ($O(n^3)$ where $n$ is the number of variables) are not applied to the full variable set, but to restricted subsets thereof.
+It thus seems reasonable that techniques such as considering ``packs'' of related variables, slicing, etc. may similarly help SMT-based techniques to scale to global analyses.
+
+We compared the precision of different techniques and abstract domains by comparing the invariants for the inclusion ordering.
+A better metric is perhaps to take a client analysis --- such as the detection of overflows and array bound violations --- and compare the rates of alarms.
+
+We focused on numerical properties, because they are supported by easily available abstract libraries.
+Yet, in most programs, properties of data structures are important for proving interesting properties.
+Further investigations are needed not only on good abstractions for pointers (many are already known) but also on their conversion to SMT problems.
+
+\phantomsection\addcontentsline{toc}{section}{References} 
+\bibliographystyle{splncs03}
+\bibliography{implicitization}
+
+\appendix
+
+\section{Experimental Results}
 
@@
 
-\section{Conclusion and Future Prospects}
-Roughly, an analysis by abstract interpretation is defined by the choice of an iteration strategy and an abstract domain.
-In this article, we demonstrated that changes in the iteration algorithm can significantly improve precision, sometimes while improving analysis times.
-
-A common criticism of analysis techniques based on SMT-solving is that they
-do not scale up. Yet, our experiments show that, for numerical properties,
-they scale up to the size of typical functions and loops.
-It is however
-quite certain that, naively applied, they cannot scale to the kind of
-programs targeted by e.g. the Astr\'ee tool, that is, a dozens or hundreds
-of thousands of lines of code in a single loop operating over similar numbers
-of remanent variables.
-Actually, for such applications, only (quasi-)linear
-algorithms scale up, and ``cheap'' abstract domains such as octagons ($O(n^3)$ where $n$ is the number of variables) are not applied to the full variable set, but to restricted subsets thereof.
-It thus seems reasonable that techniques such as considering ``packs'' of related variables, slicing, etc. may similarly help SMT-based techniques to scale to global analyses.
-
-We compared the precision of different techniques and abstract domains by comparing the invariants for the inclusion ordering.
-A better metric is perhaps to take a client analysis --- such as the detection of overflows and array bound violations --- and compare the rates of alarms.
-
-We focused on numerical properties, because they are supported by easily available abstract libraries.
-Yet, in most programs, properties of data structures are important for proving interesting properties.
-Further investigations are needed not only on good abstractions for pointers (many are already known) but also on their conversion to SMT problems.
-
-\phantomsection\addcontentsline{toc}{section}{References} 
-\bibliographystyle{splncs03}
-\bibliography{implicitization}
 \end{document}

### 2012-02-22 papers/SAS2012/gopan_reps.c

@@ -1,5 +1,4 @@
-int x = 0;
-int y = 0;
+int x = 0, y = 0;
 while (1) {
 	if (x <= 50) y++;
 	else y--;

### 2012-02-22 papers/SAS2012/rlim_int.c

@@ -1,4 +1,3 @@
-extern int input();
 void rate_limiter() {
   int x_old = 0;
   while (1) {

### 2012-02-22 papers/SAS2012/rlim_int_loop.c

@@ -1,5 +1,3 @@
-extern int input();
-extern int wait();
 void rate_limiter() {
   int x_old = 0;
   while (1) {

### 2012-02-22 papers/SAS2012/implicitization.tex
Encore du compactage

@@
 
+\renewcommand{\baselinestretch}{1}
+
 \begin{document}
@@
 This article makes the following contributions:
-\begin{enumerate}
+\begin{compactenum}
 \item We recast the guided static analysis technique from \cite{DBLP:conf/sas/GopanR07} on the expanded multigraph from \cite{Monniaux_Gonnord_SAS11}, considering entire paths instead of individual transitions, using SMT queries and binary decision diagrams\label{contr:guided_multigraph} (See \S\ref{sec:guided_multigraph}).
@@
 
-\item We conducted extensive experiments with this tool, on real-life programs: (i) fixing the abstract domain, varying the iteration technique (\S\ref{sec:compare_techniques}) (ii) fixing the iteration technique, varying the abstract domain (\S\ref{sec:compare_domains}).
-\end{enumerate}
+\item We conducted extensive experiments with this tool, on real-life programs\avirer{: (i) fixing the abstract domain, varying the iteration technique (\S\ref{sec:compare_techniques}) (ii) fixing the iteration technique, varying the abstract domain (\S\ref{sec:compare_domains})}.
+\end{compactenum}
 
@@
 An \emph{abstract domain} is a set $\abstr{X}$ equipped with a partial order $\sqsubseteq$; for instance, it can be the domain of convex polyhedra in $\QQ^3$ ordered by geometric inclusion.
-In this article, all abstract domains are supposed to contain machine-representable objects, and all $\sqsubseteq$ order relations are supposed to be decidable.
+\avirer{In this article, all abstract domains are supposed to contain machine-representable objects, and all $\sqsubseteq$ order relations are supposed to be decidable.}
 The concrete and abstract domains are connected by a monotone \emph{concretization} function $\gamma: \left(\abstr{X},\sqsubseteq\right) \rightarrow (\parts{X},\subseteq)$: an element $\abstr{x} \in \abstr{X}$ represents a set $\gamma(\abstr{x})$.%
 %
-\footnote{%
-Some presentations of abstract interpretation identify an element $\abstr{x}$ with the set of states $\gamma(\abstr{x})$ that it represents.
+% \footnote{%
+% Some presentations of abstract interpretation identify an element $\abstr{x}$ with the set of states $\gamma(\abstr{x})$ that it represents.
 This leads to simpler notations, but may also confuse if, as in some of our constructions, there exist several $\abstr{x}$ with identical $\gamma(\abstr{x})$, but distinct algorithmic behaviors.}
@@
 A concrete transition relation $\tau \subseteq X \times X$ is replaced by an abstract \emph{forward abstract transformer} $\abstr{\tau}: \abstr{X} \rightarrow \abstr{X}$, such that
-\begin{equation}
+\(
 \forall \abstr{x} \in \abstr{X}, x,x' \in X,~
@@
 x' \in \gamma \circ \abstr{\tau}(\abstr{x})
-\label{eqn:transformer_correct}
-\end{equation}
+\).
 It is easy to see that if to any control point $p \in P$ we attach an abstract element $\abstr{x}_p$ such that
@@
 associated to $p_i$ :
+\vspace{-.5em}
 \[
@@
  \neg X_j^s)
+\vspace{-.1em}
 \]
@@
 $f'(p_i)$, where:
+\vspace{-.5em}
 \[
@@
 \neg X_j^d)
+\vspace{-1em}
 \]
@@
 
-\begin{enumerate}
+\begin{compactenum}
 	\item The first subset of paths is depicted on Figure
@@
 		terminates.
-\end{enumerate}
+\end{compactenum}
 
@@
 Our method improves this technique in two ways :
-\begin{itemize}
+\begin{compactitem}
 \item Instead of enumerating the whole set of paths, we keep them implicit and
@@
 In our approach, we compute such an image only if it makes the resulting abstract value grow.
-\end{itemize}
+\end{compactitem}
 
@@
 Otherwise:
-\begin{itemize}[itemsep=-1\parsep,topsep=0pt]
+\begin{compactitem}
 	\item if $m_i < M$, we increase $m_i$ by $1$ and define $\sigma_i(j,k) = m_i$
@@
 		%d'egalite en commun pour definir son $sigma$. Nous, on le met dans M, pour simplifier}
-\end{itemize}
+\end{compactitem}
 The main difference with the original

### 2012-02-22 papers/SAS2012/implicitization.tex
Ça rentre presque dans les 15 pages

@@
 %\usepackage{color}
+\usepackage{placeins}
 
@@
 
+\avirer{
 Most SMT solvers, including Z3,
@@
 it may then backtrack some of the constraints and add other ones without restarting from scratch.
+}
 
@@
 Classical interval analysis on this program yields $\lstinline|x_old| \in [-100000,100000]$.
-We have transformed the first program into the second, manually pruning out infeasible paths; yet in general the resulting program could be exponentially larger than the first \avirer{(as in Fig.~\ref{fig:multigraph})}, even though not all feasible paths are needed to compute the invariant.
+We have transformed the program, manually pruning out infeasible paths; yet in general the resulting program could be exponentially larger than the first \avirer{(as in Fig.~\ref{fig:multigraph})}, even though not all feasible paths are needed to compute the invariant.
 
 Following recent suggestions \cite{Gawlitza_Monniaux_ESOP11,Monniaux_Gonnord_SAS11}, we avoid this space explosion by keeping the second program implicit while simulating its analysis.
-This means we work on an implicitly represented transition multigraph (Fig.~\ref{fig:multigraph}); it is succinctly representated by the transition graph of the first program.
+This means we work on an implicitly represented transition multigraph \avirer{(Fig.~\ref{fig:multigraph})}; it is succinctly representated by the transition graph of the first program.
 Our first contribution (\S\ref{sec:guided_multigraph}) is to recast the ``guided analysis'' from \cite{DBLP:conf/sas/GopanR07} on such a succinct representation of the paths in lieu of the individual transitions.
@@
 \begin{figure}
+  \vspace{-1em}
   \begin{minipage}{.3\linewidth}
@@
 
-Gopan and Reps' idea is to analyze the first phase of the loop with a widening and narrowing sequence, and thus obtain $0 \leq x \leq 50 \land y = x$, and then analyze the second phase, finally obtaining invariant~\ref{eqn:triangle}; each phase is identified by the tests taken or not taken.
+Gopan and Reps' idea is to analyze the first phase of the loop with a widening and narrowing sequence, and thus obtain $0 \leq x \leq 50 \land y = x$, and then analyze the second phase, finally obtaining invariant~(\ref{eqn:triangle}); each phase is identified by the tests taken or not taken.
 
@@
 
-Analyzing this loop using widening and narrowing on convex polyhedra or octagons yields the loop invariant $0 \leq x \leq 51 \land y = x$.
-Now, the transition at line~5 becomes feasible; and we analyze the full loop, starting iterations from $0 \leq x \leq 51 \land y = x$, and obtain invariant~(\ref{eqn:triangle}).
+Analyzing this loop using widening and narrowing on convex polyhedra
+or octagons yields the loop invariant $0 \leq x \leq 51 \land y = x$.
+Now, the transition at line~5 becomes feasible; and we analyze the
+full loop, starting iterations from $0 \leq x \leq 51 \land y = x$,
+and obtain invariant~(\ref{eqn:triangle}) in Fig~\ref{fig:gopan_reps_invariant}.
 
@@
 The order of steps is important: narrowing has to be performed before
-adding new paths, or some spurious new paths would be added to $P$,
-and starting with the addition of new paths avoids wasting time
+adding new paths, or spurious new paths would be added to $P$.
+Starting with the addition of new paths avoids% wasting time
 doing the ascending iterations on an empty graph.
@@
 %
+\vspace{-.5em}
 \begin{equation*}
@@
 (b_j^d \wedge \bigwedge_{1 \leq k \leq m_i}\!\! (\neg X_{j,k}))
+\vspace{-.5em}
 \end{equation*}
-%
+
 In our algorithm, the initialization of the abstract values slightly differs from
@@
 abstract value $X_{i,j}$ by the path indexed by $k$, and try to find an existing
-disjunct of index $j'$ so that the least upper bound of the two abstract values is exactly their union (this can be tested using SMT-solving).
+disjunct of index $j'$ so that the least upper bound of the two abstract values is exactly their union (using SMT-solving).
 If such an index exists, then we set $\sigma_i(j,k) = j'$.
@@
 different techniques, mostly on open-source projects (Fig.~\ref{fig:projects}) written in C, C++ and Fortran.
-
-\begin{table}[!h]
+\MM{TODO: résumer les conclusions sur les résultats}
+
+\begin{figure}[h!]
       \centering
@@
       \end{tabular}
-\caption{List of analyzed open-source projects, with their respective number of
-lines of code, number of control points in $P_R$, and execution time
-on various techniques\MM{TODO}}
       \label{fig:projects}
-\end{table}
+
+    \input{gnuplot/techniques.tex}
+  \vspace{-20pt}
+  \caption{Comparison of the abstract values obtained on several open-source projects. The table shows
+  their respective number of lines of code, number of control points in $P_R$, and execution time
+  on various techniques.
+  Techniques are classical abstract
+  interpretation (S), \emph{Guided Static
+  Analysis} (G), \emph{Path-focused} technique (PF), our combined technique
+  (G+PF), and its version with disjunctive invariants (DIS).
+  The $\subsetneq$ bars gives the percentage of invariants stronger (more precise; smaller with respect to inclusion) with the left-side technique,
+$\supsetneq$ the percentage of invariants stronger with the right-side technique,
+and ``uncomparable'' gives the percentage of invariants that are uncomparable, i.e
+neither greater nor smaller;
+the code points where both invariants are equal make up the remaining percentage. See Tab.~\ref{tab:techniques} for details.}
+  \label{fig:techniques}
+\end {figure}
 
@@
 
+\FloatBarrier
+
 \phantomsection\addcontentsline{toc}{section}{References} 
@@
 
-\begin{figure}[h]
-  \begin{center}
-    \input{gnuplot/techniques.tex}
-  \end{center} 
-  \vspace{-20pt}
-  \caption{Comparison of the abstract values obtained by classical abstract
-  interpretation (S), \emph{Guided Static
-  Analysis} (G), \emph{Path-focused} technique (PF), our combined technique
-  (G+PF), and its version with disjunctive invariants (DIS).
-  The $\subsetneq$ bars gives the percentage of invariants stronger (more precise; smaller with respect to inclusion) with the left-side technique,
-$\supsetneq$ the percentage of invariants stronger with the right-side technique,
-and ``uncomparable'' gives the percentage of invariants that are uncomparable, i.e
-neither greater nor smaller;
-the code points where both invariants are equal make up the remaining percentage. See Tab.~\ref{tab:techniques} for details.}
-  \label{fig:techniques}
-\end {figure}
+\MM{TODO: graphique supprimé ici}
 

### 2012-02-23 papers/SAS2012/implicitization.tex
URL de PAGAI + citation rapport M2

@@
 techniques \emph{Path Focusing} \cite{Monniaux_Gonnord_SAS11} and \emph{Guided
-Static Analysis} \cite{DBLP:conf/sas/GopanR07}.
+Static Analysis} \cite{DBLP:conf/sas/GopanR07}. It is
+available online at \url{https://forge.imag.fr/projects/pagai/}. The
+implementation is documented in~\cite{julien-henry-m2r}.
+\MM{pas idéal de citer un rapport de M2, mais ça a des chances de
+  faire taire les reviewers qui voudraient des détails sur comment
+  c'est fait, et on n'a pas la place de mettre plus. Le
+  paragraphe~\ref{sec:analysis-algorithm} a été supprimé depuis
+  PLDI ...}
 

### 2012-02-23 papers/SAS2012/implicitization.tex
Compactage

@@
 
-\renewcommand{\baselinestretch}{1}
+\renewcommand{\baselinestretch}{.97}
 
@@
 We thus conducted extensive comparisons between our new techniques and previous ones, on a variety of open-source packages.
-We also took the opportunity to compare the precision of various classical abstract domains (convex polyhedra, octagons, intervals).
 \end{abstract}
@@
 
+\MM{J'avais insisté pour qu'on écrive un truc comme ça, mais si on ne
+  met plus l'accent sur le côté expérimental, on peut sans doute
+  supprimer ce paragraphe completement ?}
+\avirer{
 Widening operations have a somewhat counterintuitive behavior: the
@@
 with respect to the precondition.
-\avirer{
-For instance, classical interval
-analysis with widening and decreasing iterations on the loop
-\lstinline|while (x != 10) x=x+1;| yields a loop invariant $x \in
-[0,+\infty)$ if the precondition is $x \in [0,0]$, but $x \in [0,10]$
-(strictly stronger) if the precondition is $x \in [0,10]$ (strictly
-weaker).
+% For instance, classical interval
+% analysis with widening and decreasing iterations on the loop
+% \lstinline|while (x != 10) x=x+1;| yields a loop invariant $x \in
+% [0,+\infty)$ if the precondition is $x \in [0,0]$, but $x \in [0,10]$
+% (strictly stronger) if the precondition is $x \in [0,10]$ (strictly
+% weaker).
+This means that, when using widenings, being locally more precise \avirer{(for instance, by using more expressive abstract domains, or widening operators that climb more slowly \cite{BagnaraHRZ05SCP})} does not necessarily translate into higher final precision.
+% It is almost always possible to concoct examples where a supposedly more precise abstraction (e.g.
+polyhedra) yields less precise results than a less precise one (e.g.
+intervals), but are these examples representative of real programs
+This justifies our recourse to evaluation on real, large programs~(\S~\ref{sec:compare_techniques}).
 }
-This means that, when using widenings, being locally more precise \avirer{(for instance, by using more expressive abstract domains, or widening operators that climb more slowly \cite{BagnaraHRZ05SCP})} does not necessarily translate into higher final precision.
-\avirer{
-It is almost always possible to concoct examples where a supposedly more precise abstraction (e.g.
-polyhedra) yields less precise results than a less precise one (e.g.
-intervals), but are these examples representative of real programs?
-}
-This justifies our recourse to evaluation on real, large programs~(\S~\ref{sec:compare_techniques}).
 
@@
 
-%\MM{Je trouve qu'il manque une vue de haut niveau quelque part par
-%  ici. Dire qu'en gros, l'algo, c'est (en pseudo-code ou en langue naturelle) :
-%
-%Sans ca, enchainer tout de suite sur ``ascending iterations'' est un peu brutal.
-%
-%Ensuite, on peut rafiner la définition de ``Subset'' ($A$ et $P$), et
-%enchainer sur l'algo 2.
-%}
-
 Step~\ref{step:ascending} corresponds to the application of
@@
 adding new paths, or spurious new paths would be added to $P$.
-Starting with the addition of new paths avoids% wasting time
+Starting with the addition of new paths avoids
 doing the ascending iterations on an empty graph.

### 2012-02-23 papers/SAS2012/implicitization.tex
blabla sur résultats expérimentaux

@@
 
+\MM{Bon, c'est vraiment frustrant de se limiter à ça, mais pour écrire
+  plus, faut trouver de la place}
+These results confirm that our combined technique improve the
+analysis in comparison with the two techniques taken individually, at
+a reasonable cost. The extension with disjunctive invariants increases
+precision in many cases, but with higher cost in terms of execution
+time.
+
 \section{Conclusion and Future Prospects}
@@
 \appendix
+
+\MM{TODO: la partie appendix est laissée en chantier. A priori, on ne
+  mettra que les tableaux de chiffres sans les commentaires, et les
+  tableaux risquent de changer avec les avancées de Julien.}
 

### 2012-02-23 papers/SAS2012/implicitization.tex
Ça rentre dans 15 pages !

@@
 
-\renewcommand{\baselinestretch}{.97}
+\renewcommand{\baselinestretch}{1}
 
@@
 The branches not taken are pruned from the loop body, yielding:
+\vspace{-.5em}
 \begin{lstlisting}[numbers=none]
@@
 \end{lstlisting}
+\vspace{-.5em}
 
@@
 that we call the abstract values $X_i^s$ instead of $X_i$ to leave room for
-another abstract value $X_i^d$ later in the algorithm.
+another abstract value $X_i^d$ later.
 
@@
 
-\subsection{Algorithm}
+The ascending iteration applies path-focusing
+\cite{Monniaux_Gonnord_SAS11} to a subset of the multigraph. As
+\cite{DBLP:conf/sas/GopanR07}, we do some narrowing, to recover
+precision lost by widening, \emph{before}
+computing and taking into account new feasible paths. Thus, our
+technique combines the advantages of \emph{Guided Static
+  Analysis} and \emph{Path-focusing}.
+
+% MM: ce titre bouffe pas mal de place pour une valeur ajoutée faible
+%\subsection{Algorithm}
 
@@
 
+\vspace{-1em}
 \begin{algorithm}
@@
 \end{algorithm}
+\vspace{-1em}
 
@@
 
-Step~\ref{step:ascending} corresponds to the application of
-path-focusing \cite{Monniaux_Gonnord_SAS11} to
-the elements of $A$ until $A$ is empty. 
-When $A$ becomes empty, it means the
-invariant computation of the subgraph is finished. As proposed by
-\cite{DBLP:conf/sas/GopanR07}, we can do some narrowing iterations.
-These narrowing iterations allow to recover precision lost by
-widening, \emph{before} computing and taking into account new feasible paths.
-Thus, our technique combines both the advantages of \emph{Guided Static
-Analysis} and \emph{Path-focusing}.
-
 The order of steps is important: narrowing has to be performed before
@@
 	\node[PRstate] (n1) [below of=n0] {$p_{10}$};
-	\node (label) [below of=n1, yshift=1.2cm] {Step 1};
+	\node (label) [left of=n00, node distance=1cm] {\bf Step 1};
 
@@
 	\node[PRstate] (n1) [below of=n0] {$p_{10}$};
-	\node (label) [below of=n1, yshift=1.2cm] {Step 2};
+	\node (label) [left of=n00, node distance=1cm] {\bf Step 2};
 
@@
 
+\vspace{-1em}
 \begin{algorithm}[!h]

### 2012-02-27 papers/SAS2012/gopan_reps.c
des details de presentation (p.ex. tables en annexe)

@@ -1,7 +1,7 @@
 int x = 0, y = 0;
 while (1) {
-	if (x <= 50) y++;
-	else y--;
-	if (y < 0) break;
-	x++;
+  if (x <= 50) y++;
+  else y--;
+  if (y < 0) break;
+  x++;
 }

### 2012-02-27 papers/SAS2012/rlim_int.c

@@ -5,5 +5,4 @@ void rate_limiter() {
     if (x > x_old+10) x = x_old+10;
     if (x < x_old-10) x = x_old-10;
     x_old = x;
-  }
-}
+} }

### 2012-02-27 papers/SAS2012/rlim_int2.c

@@ -5,5 +5,4 @@ void rate_limiter() {
     if (x > x_old+10) x = x_old+10;
     else if (x < x_old-10) x = x_old-10;
     else x_old = x;
-  }
-}
+} }

### 2012-02-27 papers/SAS2012/rlim_int_loop.c

@@ -6,5 +6,4 @@ void rate_limiter() {
     if (x < x_old-10) x = x_old-10;
     x_old = x;
     while (wait()) {}
-  }
-}
+} }

### 2012-03-02 papers/SAS2012/implicitization.tex
des details de presentation (p.ex. tables en annexe)
mathabx change tous les symboles maths; emulation

@@
 \usepackage{amsmath}
-\usepackage{mathcomp,amsfonts,amssymb,mathabx}
+\usepackage{mathcomp,amsfonts,amssymb}
 \usepackage[mathscr]{euscript}
@@
 \usepackage{dcolumn}
+\usepackage{rotating}
 
@@
 The set $\parts{X}$ of subsets of $X$, partially ordered by inclusion, is the \emph{concrete domain}.
-An \emph{abstract domain} is a set $\abstr{X}$ equipped with a partial order $\sqsubseteq$; for instance, it can be the domain of convex polyhedra in $\QQ^3$ ordered by geometric inclusion.
+An \emph{abstract domain} is a set $\abstr{X}$ equipped with a partial order $\sqsubseteq$ (the associated strict order being $\sqsubset$); for instance, it can be the domain of convex polyhedra in $\QQ^3$ ordered by geometric inclusion.
 \avirer{In this article, all abstract domains are supposed to contain machine-representable objects, and all $\sqsubseteq$ order relations are supposed to be decidable.}
@@
 \emph{Kleene iterations} compute such an inductive invariant as the stationary limit, if it exists, of the following system: for each $p$, initialize $\abstr{x}_p$ such that $\gamma(\abstr{x}_p)$ is a superset of the initial states at point $p$; then iterate the following:
-if $\abstr{\tau}_{p,p'} (\abstr{x}_p) \nsqsubseteq \abstr{x}_{p'}$, replace $\abstr{x}_{p'}$ by $\abstr{x}_{p'} \sqcup \abstr{\tau}_{p,p'} (\abstr{x}_p)$.
-Such a stationary limit is bound to exist if $\abstr{X}$ has no infinite ascending chain $a_1 \sqsubsetneq a_2 \sqsubsetneq \dots$; this condition is however not met by domains such as intervals or convex polyhedra.
+if $\abstr{\tau}_{p,p'} (\abstr{x}_p) \not\sqsubseteq \abstr{x}_{p'}$, replace $\abstr{x}_{p'}$ by $\abstr{x}_{p'} \sqcup \abstr{\tau}_{p,p'} (\abstr{x}_p)$.
+Such a stationary limit is bound to exist if $\abstr{X}$ has no infinite ascending chain $a_1 \sqsubset a_2 \sqsubset \dots$; this condition is however not met by domains such as intervals or convex polyhedra.
 
@@
 \label{fig:example-graph}
-\centering
 \begin{minipage}[c]{.15\textwidth}
-\begin{tikzpicture}[->,>=stealth',auto,node distance=2.1cm,
+\begin{tikzpicture}[->,>=stealth',auto,node distance=2.5cm,
                     semithick,font=\footnotesize]
@@
 \end{tikzpicture}
-\end{minipage} 
+\end{minipage}\hfill
 \begin{minipage}[c]{.75\textwidth}
@@
 
-\begin{table*}
+\begin{sidewaystable}
 \begin{center}
@@
 \label{tab:techniques}
-\end{table*}
+\end{sidewaystable}
 
@@
 \label{sec:compare_domains}
-\begin{table*}
+\begin{sidewaystable}
 \begin{center}
@@
 \label{tab:domain}
-\end{table*}
+\end{sidewaystable}
 

### 2012-03-02 papers/SAS2012/implicitization.tex
path

@@
 
-\cite{Monniaux_Gonnord_SAS11}'s \emph{path-focusing} technique
+Monniaux \& Gonnord's \emph{path-focusing} \cite{Monniaux_Gonnord_SAS11} technique
 distinguishes the different paths in the program in order to avoid loss of
@@
 
-\cite{DBLP:conf/pldi/GulwaniZ10} proposed a technique for computing disjunctive invariants, by
+Gulwani \& Zuleger \cite{DBLP:conf/pldi/GulwaniZ10} proposed a technique for computing disjunctive invariants, by
 distinguishing all the paths inside a loop. In

### 2012-03-05 papers/SAS2012/implicitization.tex
coquilles; reintegration de figures a droite des listings correspondants

--- 2012-02-23 papers/SAS2012/implicitization.tex
@@
 \label{subsec:rate_lim}
-Consider the following program, adapted from \cite{Monniaux_Gonnord_SAS11}, where \lstinline|input($a$, $b$)| stands for a nondeterministic input in $[a,b]$:
+Consider the following program, adapted from \cite{Monniaux_Gonnord_SAS11}, where \lstinline|input($a$, $b$)| stands for a nondeterministic input in $[a,b]$ (in the associated control-flow graph, $s$ is the start node and $e$ the end node):
+
+\begin{minipage}{9cm}
 \lstinputlisting{rlim_int.c}
-This program implements a construct commonly found in control programs (in e.g.
-automotive or avionics): a rate limiter.
-\avirer{For the sake of simplicity, we chose it
-to be fed on a nondeterministic input clamped between $[-100000,100000]$, but in a real system it would be integrated in a reactive control loop and its input connected to a complex system with unknown output range.}
-
-The expected inductive invariant is $\lstinline|x_old| \in [-100000,100000]$, but classical abstract interpretation using intervals (or octagons or polyhedra) finds $\lstinline|x_old| \in (-\infty,+\infty)$~\cite{ASTREE_ESOP05}.
-Let us briefly see why.
-
-Widening iterations converge to $\lstinline|x_old| \in (-\infty,+\infty)$; let us now see why decreasing iterations fail to recover the desired invariant.
-The \lstinline|x > x_old+10| test at line~6, if taken, yields $\lstinline|x_old|
-\in (-\infty,99990)$; followed by \lstinline|x = x_old+10|, we obtain
-$\lstinline|x| \in (-\infty,\allowbreak 100000)$, and the same after union with the no-operation ``else'' branch. Line~7 yields $\lstinline|x| \in (-\infty,+\infty)$.
-
-We could use ``widening up to'' or ``widening with thresholds'', propagating the ``magic values'' $\pm 100000$ associated to \lstinline|x| into \mbox{\lstinline|x_old|,} but these syntactic approaches cannot directly cope with programs for which $\lstinline|x|  \in [-100000,+100000]$ is itself obtained by analysis.
-The guided static analysis of \cite{DBLP:conf/sas/GopanR07} does not perform
-better, and also obtains $\lstinline|x_old| \in (-\infty,+\infty)$.
-
-In contrast, let us distinguish all four possible execution paths through the tests at lines 6 and~7. The path through both ``else'' branches is infeasible; the program is thus equivalent to:
-\lstinputlisting{rlim_int2.c}
-
-\avirer{
-\begin{figure}
-\MM{J'aime bien cette figure, qui montre en un coup d'oeil l'idée,
-  mais elle prend vraiment beaucoup de place, je propose de
-  supprimer.}
-
-\centering
-\begin{minipage}[c]{.21\textwidth}
+\end{minipage}%
+\begin{minipage}[c]{2cm}
 \begin{tikzpicture}[->,>=stealth',auto,node distance=.2cm and .5cm,
@@
 
-	\node[PRstate] (n0) {$p$};
+	\node[PRstate] (n0) {$s$};
 	\node[state] (n1) [below left=of n0] {};
@@
 	\node[state] (n6) [below right=of n4] {};
-	\node[state] (n7) [below left=of n6] {};
-	\node[state] (n8) [below right=of n6] {};
-	\node[PRstate] (n9) [below right=of n7] {$q$};
+	\node[PRstate] (n7) [below =of n6] {$e$};
 
@@
   \path [transition] (n6) -- (n7);
-  \path [transition] (n6) -- (n8);
-  \path [transition] (n7) -- (n9);
-  \path [transition] (n8) -- (n9);
 \end{tikzpicture}
 \end{minipage}
-$\Longrightarrow$
-\begin{minipage}[c]{.19\textwidth}
+
+This program implements a construct commonly found in control programs (in e.g.
+automotive or avionics): a rate or slope limiter.
+\avirer{For the sake of simplicity, we chose it
+to be fed on a nondeterministic input clamped between $[-100000,100000]$, but in a real system it would be integrated in a reactive control loop and its input connected to a complex system with unknown output range.}
+
+The expected inductive invariant is $\lstinline|x_old| \in [-100000,100000]$, but classical abstract interpretation using intervals (or octagons or polyhedra) finds $\lstinline|x_old| \in (-\infty,+\infty)$~\cite{ASTREE_ESOP05}.
+Let us briefly see why.
+
+Widening iterations converge to $\lstinline|x_old| \in (-\infty,+\infty)$; let us now see why decreasing iterations fail to recover the desired invariant.
+The \lstinline|x > x_old+10| test at line~6, if taken, yields $\lstinline|x_old|
+\in (-\infty,99990)$; followed by \lstinline|x = x_old+10|, we obtain
+$\lstinline|x| \in (-\infty,\allowbreak 100000)$, and the same after union with the no-operation ``else'' branch. Line~7 yields $\lstinline|x| \in (-\infty,+\infty)$.
+
+We could use ``widening up to'' or ``widening with thresholds'', propagating the ``magic values'' $\pm 100000$ associated to \lstinline|x| into \mbox{\lstinline|x_old|,} but these syntactic approaches cannot directly cope with programs for which $\lstinline|x|  \in [-100000,+100000]$ is itself obtained by analysis.
+The guided static analysis of \cite{DBLP:conf/sas/GopanR07} does not perform
+better, and also obtains $\lstinline|x_old| \in (-\infty,+\infty)$.
+
+In contrast, let us distinguish all four possible execution paths through the tests at lines 6 and~7.
+The path through both ``else'' branches is infeasible; the program is thus equivalent to a program with 3 paths:
+
+\begin{minipage}[c]{9cm}
+\lstinputlisting{rlim_int2.c}
+\end{minipage}
+\begin{minipage}[c]{2cm}
 \begin{tikzpicture}[->,>=stealth',auto,node distance=.2cm and .5cm,
@@
 
-	\node[PRstate] (n0) {$p$};
+	\node[PRstate] (n0) {$s$};
 	\node (n1) [below left=of n0] {};
@@
 	\node (n8) [below right=of n6] {};
-	\node[PRstate] (n9) [below right=of n7] {$q$};
-
-  \path [transition] 
-		(n0) edge  [out=0, in=0]            node {} (n9);
-  \path [transition] 
-		(n0) edge  [out=180, in=180]        node {} (n9);
-  \path [transition] 
-		(n0) edge  [out=205, in=155]            node {} (n9);
+	\node[PRstate] (n9) [below right=of n7] {$e$};
+
+%  \path [transition] 
+%		(n0) edge  [out=0, in=0]            node {} (n9);
+%  \path [transition] 
+%		(n0) edge  [out=180, in=180]        node {} (n9);
+%  \path [transition] 
+%		(n0) edge  [out=205, in=155]            node {} (n9);
   \path [transition] 
@@
 		(n0) edge  [out=255, in=105]            node {} (n9);
-  \path [transition] 
-		(n0) edge  [out=-25, in=25]            node {} (n9);
-  \path [transition] 
+%  \path [transition] 
+%		(n0) edge  [out=-25, in=25]            node {} (n9);
+  \path [transition,dotted] 
 		(n0) edge  [out=-50, in=50]            node {} (n9);
@@
 \end{minipage}
-\caption{Expansion of the transition graph into a multigraph.}
-\label{fig:multigraph}
-\end{figure}
-}
 
@@
 a \emph{destination} point, with only incoming transitions, so that the
-resulting graph is acyclic and there is no paths going through control
+resulting graph is acyclic and there are no paths going through control
 points in $P_R$.
@@
 Algorithm~\ref{algo:combined} performs Guided static analysis on the implicitly represented multigraph.
+$I_p$ denotes a set of initial states at program point~$p$ (thus $\emptyset$ for most $p$).
 %
@@
   $P \cup P'$ of the graph.
-  Either the previous iteration or the initialization lead us to a
+  Either the previous iteration or the initialization led us to a
   state where there are no more paths in the previous subset $P$,
@@
 
-Our technique computes the fixpoint iterations on a ascending sequence of
+Our technique computes the fixpoint iterations on an ascending sequence of
 subgraphs, until the complete graph is reached.
 When the analysis of a subgraph is finished, meaning that the abstract values
-for each control points has converged to an inductive invariant for this subgraph,
+for each control point has converged to an inductive invariant for this subgraph,
 the next subgraph to work on has to be computed.
@@
 \item 
-the subset of paths $P$ strictly increases at each loop iteration, and bounded by the finite set of
+the subset of paths $P$ strictly increases at each loop iteration,
+and is bounded by the finite set of
 paths in the entire graph. 
@@
 (ii) the abstract operation for a single path in path-focusing methods also is a (large) parallel assignment;
-(iii) as suggested by \cite{DBLP:conf/vmcai/Mine06}, this approach is more precise than running abstract operations for each program line separately:
+(iii) as suggested previously \cite{Colby_PhD,DBLP:conf/vmcai/Mine06}, this approach is more precise than running abstract operations for each program line separately:
 for instance, for \lstinline|y=x; z=x-y;| with precondition $x \in [0,1]$, a line-by-line interval analysis obtains $y \in [0,1]$ and $z \in [-1,1]$ while our ``en bloc'' analysis symbolically simplifies $z = x - x = 0$ and thus $z \in [0,0]$.

### 2012-03-08 papers/SAS2012/implicitization.tex
nouveaux chiffres

--- 2012-03-05 papers/SAS2012/implicitization.tex
@@
         % M-x align RET powers ;-).
-a2ps-4.14      & 55  & 2012  & 64  & 17  & 66   & 175  & 512   \\
-gawk-4.0.0     & 59  & 902   & 59  & 8   & 18   & 43   & 129   \\
-gnuchess-6.0.0 & 38  & 1222  & 112 & 21  & 122  & 374  & 866   \\
-gnugo-3.8      & 83  & 2801  & 155 & 31  & 128  & 382  & 1403  \\
-grep-2.9       & 35  & 820   & 50  & 12  & 21   & 58   & 307   \\
-gzip-1.4       & 27  & 494   & 63  & 12  & 36   & 97   & 617   \\
-lapack-3.3.1   & 954 & 16422 & 721 & 246 & 2690 & 6052 & 22222 \\
-make-3.82      & 34  & 993   & 22  & 7   & 32   & 83   & 255   \\
-sed-4.2        & 23  & 292   & 35  & 7   & 23   & 69   & 521   \\
-tar-1.26       & 73  & 1712  & 270 & 39  & 110  & 304  & 1164  \\
-% TODO: mettre à jour les chiffres ?
+		\input{table_time}
 	\hline

### 2012-03-12 papers/SAS2012/implicitization.tex
rqs Nicolas

--- 2012-03-07 papers/SAS2012/implicitization.tex
@@
 
-An additional claim of the methods that distinguish paths inside the loop body \cite{DBLP:conf/pldi/GulwaniZ10,Monniaux_Gonnord_SAS11} is that they tend to generate better invariants than methods that do not, by behaving better with respect to the \emph{widening operators} \cite{CousotCousot_JLC92}
+An additional claim in favor of the methods that distinguish paths inside the loop body \cite{DBLP:conf/pldi/GulwaniZ10,Monniaux_Gonnord_SAS11} is that they tend to generate better invariants than methods that do not, by behaving better with respect to the \emph{widening operators} \cite{CousotCousot_JLC92}
 used for enforcing convergence when searching for loop invariants by Kleene iterations.
@@
 
-\emph{Widening-accelerated Kleene iterations} proceed by replacing $\abstr{x}_{p'} \sqcup \abstr{\tau}_{p,p'} (\abstr{x}_p)$ by $\abstr{x}_{p'} \widening (\abstr{x}_{p'} \sqcup \abstr{\tau}_{p,p'} (\abstr{x}_p))$ where $\widening$ is a \emph{widening operator}: for all $\abstr{x},\abstr{y}$, $\gamma(\abstr{y}) \subseteq \gamma(\abstr{x} \widening \abstr{y})$, and there exists no sequence $\abstr{u}_1,\abstr{u}_2,\dots$ of the form $\abstr{u}_{n+1} = \abstr{u}_n \widening \abstr{v}_n$ where $\abstr{v}_n$ is another sequence.
-Such iterations necessarily lead to a stationary limit $(\abstr{x}_p)_{p \in P}$, which defines an inductive invariant $(\gamma(\abstr{x}_p))_{p \in P}$.
+\emph{Widening-accelerated Kleene iterations} proceed by replacing $\abstr{x}_{p'} \sqcup \abstr{\tau}_{p,p'} (\abstr{x}_p)$ by $\abstr{x}_{p'} \widening (\abstr{x}_{p'} \sqcup \abstr{\tau}_{p,p'} (\abstr{x}_p))$ where $\widening$ is a \emph{widening operator}: for all $\abstr{x},\abstr{y}$, $\gamma(\abstr{y}) \subseteq \gamma(\abstr{x} \widening \abstr{y})$, and any sequence $\abstr{u}_1,\abstr{u}_2,\dots$ of the form $\abstr{u}_{n+1} = \abstr{u}_n \widening \abstr{v}_n$, where $\abstr{v}_n$ is another sequence, become stationary.
+The stationary limit $(\abstr{x}_p)_{p \in P}$, defines an inductive invariant $(\gamma(\abstr{x}_p))_{p \in P}$.
 Note that this invariant is not, in general, the least one expressible in the abstract domain, and may depend on the iteration ordering (the successive choices~$p,p'$).

### 2012-03-12 papers/SAS2012/implicitization.tex
pbs vus par Nicolas

@@
 \label{subsec:rate_lim}
-Consider the following program, adapted from \cite{Monniaux_Gonnord_SAS11}, where \lstinline|input($a$, $b$)| stands for a nondeterministic input in $[a,b]$ (in the associated control-flow graph, $s$ is the start node and $e$ the end node):
+Consider the following program, adapted from \cite{Monniaux_Gonnord_SAS11}, where \lstinline|input($a$, $b$)| stands for a nondeterministic input in $[a,b]$ (the control-flow graph on the right depicts the loop body, $s$ is the start node and $e$ the end node):
 
@@
 Classical interval analysis on this program yields $\lstinline|x_old| \in [-100000,100000]$.
-We have transformed the program, manually pruning out infeasible paths; yet in general the resulting program could be exponentially larger than the first \avirer{(as in Fig.~\ref{fig:multigraph})}, even though not all feasible paths are needed to compute the invariant.
+We have transformed the program, manually pruning out infeasible paths; yet in general the resulting program could be exponentially larger than the first, even though not all feasible paths are needed to compute the invariant.
 
 Following recent suggestions \cite{Gawlitza_Monniaux_ESOP11,Monniaux_Gonnord_SAS11}, we avoid this space explosion by keeping the second program implicit while simulating its analysis.
-This means we work on an implicitly represented transition multigraph \avirer{(Fig.~\ref{fig:multigraph})}; it is succinctly representated by the transition graph of the first program.
+This means we work on an implicitly represented transition multigraph \avirer{(Fig.~\ref{fig:multigraph})}; it is succinctly represented by the transition graph of the first program.
 Our first contribution (\S\ref{sec:guided_multigraph}) is to recast the ``guided analysis'' from \cite{DBLP:conf/sas/GopanR07} on such a succinct representation of the paths in lieu of the individual transitions.
@@
 \vspace{-1em}
-\begin{equation}
+\begin{equation*}
 y \leq x \;\land\; y \leq 102-x \;\land\; y \geq 0.\label{eqn:triangle}
-\end{equation}
+\end{equation*}
 \end{minipage}
@@
 The widening operator extrapolates from these two iterations and yields $x = y \land x \geq 0$.
-From there, the ``else'' branch at line 5 may be taken; with further widening, $0 \leq y \leq x$ is obtained as a loop invariant, and thus the postcondition computed at line 9 is $x \geq 0 \land y = 0$.
+From there, the ``else'' branch at line~4 may be taken; with further widening, $0 \leq y \leq x$ is obtained as a loop invariant, and thus the computed loop postcondition is $x \geq 0 \land y = 0$.
 Yet the strongest invariant is $(0 \leq x \leq 51 \land y = x) \lor (51 \leq x \leq 102 \land x+y=102)$, and its convex hull, a convex polyhedron (Fig.~\ref{fig:gopan_reps_invariant}).

### 2012-03-12 papers/SAS2012/implicitization.tex
hop

@@
 or octagons yields the loop invariant $0 \leq x \leq 51 \land y = x$.
-Now, the transition at line~5 becomes feasible; and we analyze the
+Now, the transition at line~4 becomes feasible; and we analyze the
 full loop, starting iterations from $0 \leq x \leq 51 \land y = x$,
@@
 for each subset, an inductive invariant is computed for the program restricted to it.
-The starting subset are the transitions reachable in one step from the loop initialization.
+The starting subset consists in the transitions reachable in one step from the loop initialization.
 If for a given subset $S$ in the sequence, no transitions outside $S$ are reachable from the inductive invariant attached to $S$, then iterations stop;
@@
 unfeasible paths.
-In this section, we use the same notations as \ref{sec:path_focusing}, except
+In this section, we use the same notations as \S\ref{sec:path_focusing}, except
 that we call the abstract values $X_i^s$ instead of $X_i$ to leave room for

### 2012-03-13 papers/SAS2012/implicitization.tex
positionnement des floats

@@
 \begin{figure}
-\label{fig:example-graph}
 \begin{minipage}[c]{.15\textwidth}
@@
 \caption{Ascending sequence of subgraphs}
+\label{fig:example-graph}
 \end{figure}
@@
 different techniques, mostly on open-source projects (Fig.~\ref{fig:projects}) written in C, C++ and Fortran.
+These results confirm that our combined technique improve the
+analysis in comparison with the two techniques taken individually, at
+a reasonable cost. The extension with disjunctive invariants increases
+precision in many cases, but with higher cost in terms of execution
+time.
+
 \MM{TODO: résumer les conclusions sur les résultats}
 
-\begin{figure}[h!]
-      \centering
+\section{Conclusion and Future Prospects}
+Roughly, an analysis by abstract interpretation is defined by the choice of an iteration strategy and an abstract domain.
+In this article, we demonstrated that changes in the iteration algorithm can significantly improve precision, sometimes while improving analysis times.
+
+A common criticism of analysis techniques based on SMT-solving is that they
+do not scale up. Yet, our experiments show that, for numerical properties,
+they scale up to the size of typical functions and loops.
+It is however
+quite certain that, naively applied, they cannot scale to the kind of
+programs targeted by e.g. the Astr\'ee tool, that is, a dozens or hundreds
+of thousands of lines of code in a single loop operating over similar numbers
+of remanent variables.
+Actually, for such applications, only (quasi-)linear
+algorithms scale up, and ``cheap'' abstract domains such as octagons ($O(n^3)$ where $n$ is the number of variables) are not applied to the full variable set, but to restricted subsets thereof.
+It thus seems reasonable that techniques such as considering ``packs'' of related variables, slicing, etc. may similarly help SMT-based techniques to scale to global analyses.
+
+\begin{table}[h!]\centering
       \begin{tabular}{|l|r|r|r|r|r|r|r|} \hline
@@
       \end{tabular}
-      \label{fig:projects}
-
+\caption{Execution times for various techniques}
+\label{tab:time}
+\end{table}
+
+\begin{figure}[h!]%
+      \centering%
     \input{gnuplot/techniques.tex}
@@
   (G+PF), and its version with disjunctive invariants (DIS).
-  The $\subsetneq$ bars gives the percentage of invariants stronger (more precise; smaller with respect to inclusion) with the left-side technique,
-$\supsetneq$ the percentage of invariants stronger with the right-side technique,
+  The $\subsetneq$ bars (resp. $\supsetneq$) gives the percentage of invariants stronger (more precise; smaller with respect to inclusion) with the left-side (resp. right-side) technique,
 and ``uncomparable'' gives the percentage of invariants that are uncomparable, i.e
@@
   \label{fig:techniques}
+  \label{fig:projects}
 \end {figure}
@@
   plus, faut trouver de la place}
-These results confirm that our combined technique improve the
-analysis in comparison with the two techniques taken individually, at
-a reasonable cost. The extension with disjunctive invariants increases
-precision in many cases, but with higher cost in terms of execution
-time.
-
-\section{Conclusion and Future Prospects}
-Roughly, an analysis by abstract interpretation is defined by the choice of an iteration strategy and an abstract domain.
-In this article, we demonstrated that changes in the iteration algorithm can significantly improve precision, sometimes while improving analysis times.
-
-A common criticism of analysis techniques based on SMT-solving is that they
-do not scale up. Yet, our experiments show that, for numerical properties,
-they scale up to the size of typical functions and loops.
-It is however
-quite certain that, naively applied, they cannot scale to the kind of
-programs targeted by e.g. the Astr\'ee tool, that is, a dozens or hundreds
-of thousands of lines of code in a single loop operating over similar numbers
-of remanent variables.
-Actually, for such applications, only (quasi-)linear
-algorithms scale up, and ``cheap'' abstract domains such as octagons ($O(n^3)$ where $n$ is the number of variables) are not applied to the full variable set, but to restricted subsets thereof.
-It thus seems reasonable that techniques such as considering ``packs'' of related variables, slicing, etc. may similarly help SMT-based techniques to scale to global analyses.
 

### 2012-03-15 papers/SAS2012/implicitization.tex
on ne distingue plus X^s et X^d, et l'algo de calcul des nouveaux chemins a changé

@@
 unfeasible paths.
-In this section, we use the same notations as \S\ref{sec:path_focusing}, except
-that we call the abstract values $X_i^s$ instead of $X_i$ to leave room for
-another abstract value $X_i^d$ later.
+In this section, we use the same notations as \S\ref{sec:path_focusing}.
+%, except
+%that we call the abstract values $X_i^s$ instead of $X_i$ to leave room for
+%another abstract value $X_i^d$ later.
 
@@
 \item \label{step:addingpaths} We start finding a new relevant subset
-  $P \cup P'$ of the graph.
+  $P$ of the graph.
   Either the previous iteration or the initialization led us to a
@@
   are not in $P$. This phase computes these paths and adds them to
-  $P'$. This phase is described in~\ref{subsec:addingpaths}
+  the subset. This phase is described in~\ref{subsec:addingpaths}
   and corresponds to lines in~\ref{alg=start-add-paths} to
@@
 \displaystyle\bigwedge_{j \in P_R \atop j \neq i} \neg
-b_j^s \wedge X_i^s \wedge \displaystyle\bigvee_{j \in Succ(i)} (b_j^d \wedge
- \neg X_j^s)
+b_j^s \wedge X_i \wedge \displaystyle\bigvee_{j \in Succ(i)} (b_j^d \wedge
+ \neg X_j)
 \vspace{-.1em}
@@
 to give us a path included in $P$.
+Each time the invariant candidate of a point $p_j$ has been updated, $p_j$ is
+inserted into $A'$ since it may be the start of a new feasible paths.
 
@@
 these new paths, but we subtly change the SMT formula given to the
-SMT solver: we now simply check for the satisfiability of $f(p_i)$
-instead of $f(p_i) \wedge P$. Since we already know that $f(p_i)
-\wedge P$ is unsatisfiable, none of the paths given by the SMT solver
-will be in $P$.
-
+SMT solver: we now try to find a path that is not yet in $P$, but is feasible
+and makes the invariant candidate of its destination grow.
 % À changer si on décide de ne plus distinguer $X^s$ et $X^d$
-Also, to prevent the algorithm from adding too many paths at a time,
-and in particular to prevent the algorithm from trying to re-add the
-same path infinitely many times, we use another abstract value
-associated to the control point $p_j$, noted $X_j^d$, which is distinct
-from $X_j^s$, and initialized to $X_j^s$ right before computing new paths. 
-In the SMT formula, we associate to $p_j$ the abstract value $X_j^s$ when $p_j$
-is the starting point of the path, and $X_j^d$ when it is its destination point.
 We thus check the satisfiability of the formula
@@
 \[
-f'(p_i) = \rho \wedge b_i^s \wedge 
-\displaystyle\bigwedge_{j \in P_R \atop j \neq i} \neg
-b_j^s \wedge X_i^s \wedge \displaystyle\bigvee_{j \in Succ(i)} (b_j^d \wedge
-\neg X_j^d)
+f'(p_i) = f(p_i)\wedge \neg P
 \vspace{-1em}
 \]
-$X_j^d$ is updated when the point $p_j$ is the target of a new path.
-This way, further SMT queries do not
-compute other paths with the same source and destination if it is not
-needed (because these new paths would not make $X_j^d$ grow, hence
-would not be returned by the SMT solver).
+$X_j$ is updated using an abstract union when the point $p_j$ is the target of a new path. This way, further SMT queries do not
+compute other paths with the same source and destination if it is not needed
+(because these new paths would not make $X_j$ grow, hence would not be returned
+by the SMT solver).
 
@@
 
+When a new path has been found, it is immediately added into $P$. 
+We then have to add $p_i$ and $p_j$ into $A$ (since we do not apply widening in
+this section) and $p_j$ into $A'$, since $p_j$ may be the starting point of a
+new feasible path.
+
 \subsection{Termination}
@@
 paths in the entire graph. 
-\item the set $P'$ always verifies $P \cap P' = \emptyset$ by construction, 
-which guarantees that $P'$
-will eventually be empty after a finite number of loop iterations.
+\item
+	When computing new paths, we cunjunct our formula with $\neg P$, meaning
+	that we obtain each possible path only once. The number of path is finite,
+	so this computation always terminates.
+\item
+	The Path-focusing iterations terminate because of the properties of
+	widening.
 \end{inparaenum}

### 2012-03-15 papers/SAS2012/implicitization.tex
mise à jour de l'exemple

@@
 
-This new subgraph is the union of the paths of the previous one with a set
-$P'$ of new paths that become feasible regarding the current abstract values.
-The paths in $P'$ are computed one after another, until no more path
+This new subgraph contains all the paths from the previous one, and also new
+paths that become feasible regarding the current abstract values.
+The new paths in $P$ are computed one after another, until no more path
 can make the invariant grow. This is line~\ref{alg=computeNewPaths} in
@@
 \item
-	When computing new paths, we cunjunct our formula with $\neg P$, meaning
+	when computing new paths, we cunjunct our formula with $\neg P$, meaning
 	that we obtain each possible path only once. The number of path is finite,
@@
 \item
-	The Path-focusing iterations terminate because of the properties of
+	the Path-focusing iterations terminate because of the properties of
 	widening.
@@
 state. In this case, we have three elements in $P_R$.
-We also unroll the loop once, in order to distinguish the paths that do not go
-through the loop from the others:
-\lstinline|if (wait()) {while (wait()) {}}|
-
-The main loop in the expanded multigraph has then 8 distinct paths, half of
-them being self loops, and the other half going to the header of the nested loop.
-
-Guided static analysis from \cite{DBLP:conf/sas/GopanR07} yields, at line 5,
+
+The main loop in the expanded multigraph has then 4 distinct paths going to the header of the nested loop.
+
+Guided static analysis from \cite{DBLP:conf/sas/GopanR07} yields, at line 3,
 $\lstinline|x_old| \in (-\infty,+\infty)$.
-Path-focusing \cite{Monniaux_Gonnord_SAS11} performs slightly
-better, and finds $\lstinline|x_old| \in (-\infty,10000]$.
+Path-focusing \cite{Monniaux_Gonnord_SAS11} also
+finds $\lstinline|x_old| \in (-\infty,+\infty)$.
 Now, let us see how our technique performs on this example.
@@
 \begin{compactenum}
-	\item The first subset of paths is depicted on Figure
-		\ref{fig:example-graph} Step 1. We apply Path-Focusing on this graph,
-		and discover for $p_5$ the inductive invariant $\lstinline|x_old| = 0$.
+	\item The starting subgraph is depicted on Figure
+		\ref{fig:example-graph} Step 1. At the beginning, this graph has no
+		transitions.
 	\item
-		We then compute the set $P'$ of paths that have to be added into the
-		next subgraph. 
-		The image of $\lstinline|x_old| = 0$ by the path that goes from $p_5$ to
-		itself, and that goes through the \emph{else} branch of each 
+		We compute the new feasible paths that have to be added into the
+		subgraph. We first find the path from $p_1$ to $p_3$ and obtain at $p_3$
+		$\lstinline|x_old| = 0$. We then find the path
+
+		The image of $\lstinline|x_old| = 0$ by the path that goes from $p_3$ to
+		$p_8$, and that goes through the \emph{else} branch of each 
 		\emph{if-then-else}, is $-10 \leq \lstinline|x_old| \leq 10$. This path
-		is then added to our subgraph. Moreover, there is no other path whose
+		is then added to our subgraph.
+
+		Moreover, there is no other path starting at $p_3$ whose
 		image is not in $-10 \leq \lstinline|x_old| \leq 10$.
-		Additionally, the path that goes from $p_5$ to
-		$p_{10}$, that goes through the \emph{else} branch of each 
-		\emph{if-then-else} at lines 7, 8 and 9, is also 
-		$-10 \leq \lstinline|x_old| \leq 10$, and thus is also added into the
-		new subgraph. Again, there is no new path that goes to $p_{10}$ with an
-		image outside $-10 \leq \lstinline|x_old| \leq 10$.
+
+		Finally, since the abstract value associated to $p_8$ is  
+		$-10 \leq \lstinline|x_old| \leq 10$, the path from $p_8$ to $p_1$ is
+		feasible and is added into $P$. The final subgraph is depicted on Figure
+		\ref{fig:example-graph} Step 2.
+
 	\item
-		We apply Path Focusing on this new subgraph (\ref{fig:example-graph}
-		Step 2). When we have reached an inductive invariant, we narrow, and the
-		result for both $p_5$ and $p_{10}$ is the polyhedron 
-		$-10000 \leq \lstinline|x_old| \leq 10000$.
-	\item Finally, we compute the new subgraph. The SMT-solver does not find any
+		We then compute the ascending iterations by path-focusing. At the end of
+		these iterations, we obtain 
+$-\infty \leq \lstinline|x_old| \leq +\infty$ for both $p_3$ and $p_8$.
+	\item
+		We now can apply narrowing iterations, and recover the precision lost by
+		widening: we obtain 
+		$-10000 \leq \lstinline|x_old| \leq 10000$ at points $p_3$ and $p_8$.
+	\item Finally, we compute the next subgraph. The SMT-solver does not find any
 		new path that makes the abstract values grow, and the algorithm
 		terminates.
+
 \end{compactenum}
@@
 $\lstinline|x_old| \in~[-10000,\allowbreak 10000]$. 
-Here, only 3 paths out of the 8 have been computed during the analysis. In
+Here, only 3 paths out of the 6 have been computed during the analysis. In
 practice, depending on the order the SMT-solver returns the paths, other
@@
 \begin{figure}
-\begin{minipage}[c]{.15\textwidth}
-\begin{tikzpicture}[->,>=stealth',auto,node distance=2.5cm,
+\begin{minipage}[c]{.45\textwidth}
+\begin{tikzpicture}[->,>=stealth',auto,node distance=2.1cm,
                     semithick,font=\footnotesize]
 
-	\node[PRstate] (n00) {$p_4$};
-	\node[PRstate] (n0) [below of=n00, yshift=0.8cm] {$p_5$};
-	\node[PRstate] (n1) [below of=n0] {$p_{10}$};
+	\node[PRstate] (n00) {$p_1$};
+	\node[PRstate] (n0) [below of=n00, yshift=0.8cm] {$p_3$};
+	\node[PRstate] (n1) [below of=n0] {$p_{8}$};
 	\node (label) [left of=n00, node distance=1cm] {\bf Step 1};
 
-  \path [transition] 
-		(n00) edge  node {$\lstinline|x_old| \gets 0$} (n0);
 \end{tikzpicture}
@@
 
-	\node[PRstate] (n00) {$p_4$};
-	\node[PRstate] (n0) [below of=n00, yshift=0.8cm] {$p_5$};
-	\node[PRstate] (n1) [below of=n0] {$p_{10}$};
+	\node[PRstate] (n00) {$p_1$};
+	\node[PRstate] (n0) [below of=n00, yshift=0.8cm] {$p_3$};
+	\node[PRstate] (n1) [below of=n0] {$p_{8}$};
 	\node (label) [left of=n00, node distance=1cm] {\bf Step 2};
@@
   \path [transition] 
-		(n0) edge  [loop left] node [left, xshift=0.2cm] {
-		$\begin{array}{r}
-			-10000 \leq \lstinline|x| \leq 10000 \\
-			\lstinline|x_old|-10 \leq \lstinline|x| \leq \lstinline|x_old|+10 /\\
-			\lstinline|x_old| \gets \lstinline|x|
-		\end{array}$
-		} (n0);
-  \path [transition] 
-		(n0) edge node [left, yshift=-0.2cm, xshift=0] {
+		(n0) edge [bend left] node [left, yshift=-0.2cm, xshift=0] {
 		$\begin{array}{r}
@@
 		} (n1);
+  \path [transition] 
+		(n1) edge [bend left]  node {} (n0);
 \end{tikzpicture}
@@
 In this example, we see that our technique actually combines best of
-\emph{Guided Static Analysis} and \emph{Path Focusing}. Still, we need to unroll
-the loop once in order to get precise results. Section \ref{sec:disjunctive}
-proposes a new technique that gives precise invariants without needing such
-graph transformations.
+\emph{Guided Static Analysis} and \emph{Path Focusing}. 
+% JH: plus besoin de ca
+%Still, we need to unroll
+%the loop once in order to get precise results. Section \ref{sec:disjunctive}
+%proposes a new technique that gives precise invariants without needing such
+%graph transformations.
 

### 2012-03-23 papers/SAS2012/implicitization.tex
nouveaux chiffres, et commentaires mis à jour

@@
 \emph{Guided Static Analysis} from \cite{DBLP:conf/sas/GopanR07} improves the
-result of the classical Abstract Interpretation in $2.72\%$ of the control points
-in $P_R$.
+result of the classical Abstract Interpretation in $2.21\%$ of the control points in $P_R$.
 \emph{Path-focusing} from \cite{Monniaux_Gonnord_SAS11} gives
 statistically better results, and finds better invariants than \emph{Guided
-Static Analysis} in $9.89\%$ of the cases. However, it also loses precision in an
-important number ($2.68\%$) of control points.
+Static Analysis} in $5.00\%$ of the cases. However, it also loses
+precision compared to \emph{Guided Static Analysis}  in an
+important number ($4.00\%$) of control points.
+
+Unfortunately, these two techniques are less precise than the classical analysis
+in a important number of cases ($4.64\%$ for $G$, $5.14\%$ for $PF$).
+This could be explained by the non-monotonicity of the results with widening operators.
+
 Finally, our combined technique gives the most promising results, since it is
@@
 The analysis using disjunctive invariants greatly improves the
-precision of the analysis (for $12.06\%$ of the control points in $P_R$), at an
-acceptable time cost (see Table \ref{tab:time}). \MM{TODO}
+precision of the analysis (for $14.46\%$ of the control points in $P_R$), at an
+acceptable time cost compared to our combined algorithm (see Table \ref{tab:time}). \MM{TODO}
 
@@
 This limitation occurred very rarely in our experiments, except for the analysis
-of \emph{Lapack/Blas}, where 798 over the 1602 functions have been skipped.
+of \emph{Lapack/Blas}, where 668 over the 1498 functions have been skipped.
 \emph{Lapack/Blas} implements matrix computations, which use floating-point multiplications.
@@
 Table \ref{tab:time} \MM{TODO} gives the execution time of the different analysis
-techniques. It is interesting to see that \emph{Guided Static Analysis} and
-\emph{Path-focusing} are sometimes faster than the classical
-algorithm.
-This seems due to the fact that these algorithms start on a small subset
-of the graph, and then may need fewer iterations on the complete graph.
+techniques. It is interesting to see that 
+\emph{Path-focusing} technique is sometimes faster than the classical algorithm.
+This seems due to the fact that this technique avoids many convex hulls, which
+is costly, and also computes the image of an abstract value by an entire path
+instead of many single transitions.
 The techniques we introduce in this article are slower, but speed is still on the same order of magnitude as the older ones.

### 2012-03-23 papers/SAS2012/implicitization.tex
...

@@
 is costly, and also computes the image of an abstract value by an entire path
-instead of many single transitions.
+instead of many single transitions, thus leading to a large parallel assignment.
 The techniques we introduce in this article are slower, but speed is still on the same order of magnitude as the older ones.

### 2012-03-23 papers/SAS2012/implicitization.tex
Deux typos

@@
 		subgraph. We first find the path from $p_1$ to $p_3$ and obtain at $p_3$
-		$\lstinline|x_old| = 0$. We then find the path
+		$\lstinline|x_old| = 0$.
+                \MM{TODO: il y avait cette phrase pas finie `` We then
+                  find the path''. J'ai enlevé. C'est bon ?}
 
@@
 		Finally, since the abstract value associated to $p_8$ is  
-		$-10 \leq \lstinline|x_old| \leq 10$, the path from $p_8$ to $p_1$ is
+		$-10 \leq \lstinline|x_old| \leq 10$, the path from
+                $p_8$ to $p_3$\MM{TODO: y'avait $p_1$ ici, j'ai mis $p_3$, c'est bon ?} is
 		feasible and is added into $P$. The final subgraph is depicted on Figure

### 2012-03-23 papers/SAS2012/implicitization.tex
Suppression de quelques vspace négatifs
Réduction des annexes aux tableaux.

@@
 \begin{figure}
-  \vspace{-1em}
+%  \vspace{-1em}
   \begin{minipage}{.3\linewidth}
@@
 \includegraphics[scale=0.7]{gopan_reps}
-\vspace{-1em}
+%\vspace{-1em}
 \begin{equation*}
@@
 associated to $p_i$ :
-\vspace{-.5em}
+%\vspace{-.5em}
 \[
@@
  \neg X_j)
-\vspace{-.1em}
+%\vspace{-.1em}
 \]
@@
 $f'(p_i)$, where:
-\vspace{-.5em}
+%\vspace{-.5em}
 \[
 f'(p_i) = f(p_i)\wedge \neg P
-\vspace{-1em}
+%\vspace{-1em}
 \]
@@
 
-\subsection{Analysis Algorithm}
-\label{sec:analysis-algorithm}
-For each program, we distinguish a set $P_R = P_W$ of suitable widening points by a simple algorithm: for each procedure, compute the strongly connected components of its control-flow graph using Tarjan's algorithm; the targets of the back-edges of the depth-first search are added to~$P_R$.
-Note that the resulting set is not necessarily minimal, but is sufficient to disconnect all cycles --- more sophisticated techniques are discussed in e.g.
-\cite{BourdonclePhd}.
-
-LLVM bitcode is in \emph{static single assignment} (SSA) form: a given scalar variable is given a value at a single syntactic point in the program.
-In concrete terms, an assignment \lstinline|x=2*x+1;| gets translated into a definition $x_2 = 2x_1+1$, with distinct variables $x_1$ and $x_2$ corresponding to the same original variable \lstinline|x| at different points in the program.
-\avirer{LLVM makes it easy to follow definition-use and use-definition chains: for a given variable (say, $x_2$) one can immediately obtain its definition (say, $2x_1+1$).
-One may see conversion to SSA form as a static precomputation of some of the symbolic propagations proposed by \cite{DBLP:conf/vmcai/Mine06} to enhance the precision of analyses.}
-
-SSA introduces $\phi$-functions at the head of a control code to
-define variables whose value depends on which incoming edge to this
-control node was last taken. For instance, for
-\lstinline|if (...) { x = 2*x+1; } else { x= 0; }|, then $x_2$ is defined as
-$\phi(2x_1+1,0)$. In this framework, variables are either defined as
-an arithmetic ($+$, $-$, $\times$, $/$) function of other variables,
-or not be representable as affine linear functions, because they are
-defined using $\phi$-functions, bitwise arithmetic, loads from memory,
-or return values from function calls.
-
-This motivates a key implement decision of our tool: only those variables $v_1,\dots,v_n$ that are not defined by arithmetic operations are retained as coordinates in the abstract domain (e.g.
-as dimensions in polyhedra).
-A basic block of code therefore amounts to a \emph{parallel assignment} operation
-$(v_1,\dots,v_n) \allowbreak\mapsto\allowbreak
-(f_1(v_1,\dots,v_n), \allowbreak, \dots, \allowbreak
- f_n(v_1,\dots,v_n))$;
-such operations are directly supported by APRON. This has three benefits:
-(i) it limits the number of dimensions in the abstract values, since polyhedra libraries typically perform worse with higher dimensions;
-(ii) the abstract operation for a single path in path-focusing methods also is a (large) parallel assignment;
-(iii) as suggested previously \cite{Colby_PhD,DBLP:conf/vmcai/Mine06}, this approach is more precise than running abstract operations for each program line separately:
-for instance, for \lstinline|y=x; z=x-y;| with precondition $x \in [0,1]$, a line-by-line interval analysis obtains $y \in [0,1]$ and $z \in [-1,1]$ while our ``en bloc'' analysis symbolically simplifies $z = x - x = 0$ and thus $z \in [0,0]$.
-
-In the event that a node is reachable only by a single control-flow edge (which may occur because of dead code, or during the first phases of guided static analysis), the $\phi$ operation reduces to a copy of the values flowing from that edge.
-In this case, our tool just propagates symbolic values through the predecessor node, without introducing $\phi$-variables.
-
-We further reduce the number of variables by projecting out those that are not live at the program point. Again, SSA use-def and def-use chains make it easy to compute liveness.
-
-Our tool currently only operates over scalar variables from the SSA representation and thus cannot directly cope with arrays or memory accessed through pointers.
-We therefore run it after the ``memory to registers'' (\texttt{mem2reg}) optimization phase in LLVM, which lifts most memory accesses to scalar variables.
-The remaining memory accesses are treated as undefined values, as are the return
-values of function calls. For better precision, we also apply function inlining
-to the program, and we unroll every loop once.
-
-Our tool currently assumes that integer variables are unbounded mathematical integers ($\ZZ$) and floating-point variables are real (or rational) numbers.
-Techniques for sound analysis of bounded integers, including with wraparound, and of floating-point operations have been developed in e.g.
-the Astr\'ee system \cite{ASTREE_ESOP05,ASTREE_PLDI03}, but porting these techniques to our iteration schemes using SMT-solving requires supplemental work.
-
-Our implementation of path-focusing currently does not use true acceleration techniques, as proposed by \cite{Monniaux_Gonnord_SAS11}.
-Instead, it simply runs widening and narrowing iterations on a single path.
-The analysis is currently only forward, even though nothing in what we describe in this article is specific to forward analysis.
-
-\subsection{Precision of the Various Techniques}
-\label{sec:compare_techniques}
-
-\MM{TODO: graphique supprimé ici}
+Experimental results of our tool on the benchmark presented in
+Fig~\ref{fig:projects} are given below. The precision of different
+techniques is given in Table~\ref{tab:techniques}. The precision of
+abstract domains and widening operators in given in
+Table~\ref{tab:domain}.
 
@@
 
-For each program and each pair $(T_1,T_2)$ of analysis techniques, we list
-the proportion of control points in $P_R$ where $T_1$ (resp.~$T_2$) gives a
-strictly stronger invariant, denoted by $\subsetneq$ (resp. $\supsetneq$),
-and the proportion of control points where the invariants given by $T_1$ and
-$T_2$ are uncomparable for the inclusion ordering (the remainder of the
-control points are thus those for which both techniques give the same
-invariant). We use convex polyhedra as the abstract domain.
-
-Let us briefly comment the results given in more details in Table~\ref{tab:techniques} and Figure~\ref{fig:techniques}.
-\emph{Guided Static Analysis} from \cite{DBLP:conf/sas/GopanR07} improves the
-result of the classical Abstract Interpretation in $2.21\%$ of the control points in $P_R$.
-\emph{Path-focusing} from \cite{Monniaux_Gonnord_SAS11} gives
-statistically better results, and finds better invariants than \emph{Guided
-Static Analysis} in $5.00\%$ of the cases. However, it also loses
-precision compared to \emph{Guided Static Analysis}  in an
-important number ($4.00\%$) of control points.
-
-Unfortunately, these two techniques are less precise than the classical analysis
-in a important number of cases ($4.64\%$ for $G$, $5.14\%$ for $PF$).
-This could be explained by the non-monotonicity of the results with widening operators.
-
-Finally, our combined technique gives the most promising results, since it is
-statistically more precise than the other techniques.
-
-The analysis using disjunctive invariants greatly improves the
-precision of the analysis (for $14.46\%$ of the control points in $P_R$), at an
-acceptable time cost compared to our combined algorithm (see Table \ref{tab:time}). \MM{TODO}
-
-While experiencing with techniques that use SMT-solving, we encountered some
-limitations due to non-linear arithmetic in the analyzed programs. Indeed, 
-the SMT-solver is not able to decide the satisfiability of some SMT-formulae
-expressing the semantics of non-linear programs. 
-In this case, we skipped the functions for which the SMT-solver returned the
-``unknown'' result.
-This limitation occurred very rarely in our experiments, except for the analysis
-of \emph{Lapack/Blas}, where 668 over the 1498 functions have been skipped.
-\emph{Lapack/Blas} implements matrix computations, which use floating-point multiplications.
-
-In cases where the formula is expressed in too rich a logic for the SMT-solver to deal with, a number of workarounds are possible, including:
-(i) \emph{Linearization}, as per \cite{DBLP:conf/vmcai/Mine06}, which overapproximates nonlinear semantics by linear semantics.
-(ii) Replacing the results of nonlinear operations by ``unknown''.
-Neither is currently implemented in our tool.
-
-Table \ref{tab:time} \MM{TODO} gives the execution time of the different analysis
-techniques. It is interesting to see that 
-\emph{Path-focusing} technique is sometimes faster than the classical algorithm.
-This seems due to the fact that this technique avoids many convex hulls, which
-is costly, and also computes the image of an abstract value by an entire path
-instead of many single transitions, thus leading to a large parallel assignment.
-The techniques we introduce in this article are slower, but speed is still on the same order of magnitude as the older ones.
-
-\subsection{Precision of Abstract Domains}
 \label{sec:compare_domains}
@@
 
-For each program and each pair $(D_1,D_2)$ of abstract domains, we compare by
-inclusion the invariants of the different control points in $P_R$ (where $P_R$
-is defined as in \ref{sec:compare_techniques}).
-The main results are given in Table \ref{tab:domain}.
-
-Statistically, the domain of convex polyhedra gives the better results, but
-commonly yields weaker invariants than the domains of octagons/intervals;
-this is a known weakness of its widening operator~\cite{Monniaux_LeGuen2011}. 
-The Octagon domain appears to be much better than intervals; this is
-unsurprising since in most programs and libraries, bounds on loop indices are
-non constant: they depend on some parameters (array sizes etc.).
-
-The Lapack/Blas benchmarks are unusual compared to the other programs.
-These libraries perform matrix computations, using nested loops over indices; such programs are the prime target for polyhedral loop optimization techniques and it is therefore unsurprising that polyhedra perform very well over them.
-
-The analysis of linear equalities (PKEQ) performs very fast compared
-to other abstract domains, but yields very imprecise invariants:
-it only detects relations of the form $\sum_i a_i x_i = C$ where $a_i$ and $C$ are constants.
-The comparison between the domain of convex polyhedra and linear congruences
-shows the percentage of points where the use of congruences refines the
-invariant. This is the case for $1.04\%$ of the control points in $P_R$.
-
-Finally, we evaluated the benefits of the improved version of the widening
-operator for convex polyhedra from \cite{BagnaraHRZ05SCP}, compared to the
-classical widening. We found that the improved version from
-\cite{BagnaraHRZ05SCP} yields more precise invariants for $6.58\%$ of the
-control points in $P_R$.
-
 \end{document}

### 2012-03-23 papers/SAS2012/implicitization.tex
retrait de TODO

@@
 		$\lstinline|x_old| = 0$.
-                \MM{TODO: il y avait cette phrase pas finie `` We then
-                  find the path''. J'ai enlevé. C'est bon ?}
 
@@
 		$-10 \leq \lstinline|x_old| \leq 10$, the path from
-                $p_8$ to $p_3$\MM{TODO: y'avait $p_1$ ici, j'ai mis $p_3$, c'est bon ?} is
+                $p_8$ to $p_3$ is
 		feasible and is added into $P$. The final subgraph is depicted on Figure
@@
 & \multicolumn{3}{c|}{\textbf{PK/GRID}}
-& \multicolumn{3}{c|}{\textbf{POLY*/POLY}} \\ %\cline{2-16}
+& \multicolumn{3}{c|}{\textbf{POLY/POLY*}} \\ %\cline{2-16}
 & \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}

### 2012-03-26 papers/SAS2012/implicitization.tex
derniers résultats sur les domaines abstraits

@@
 techniques is given in Table~\ref{tab:techniques}. The precision of
-abstract domains and widening operators in given in
+abstract domains and widening operators is given in
 Table~\ref{tab:domain}.
@@
 POLY),
-Octagons (OCT), intervals (BOX), linear equalities (PKEQ) and linear congruences
+Octagons (OCT), intervals (BOX), linear equalities (PKEQ) and linear
+inequalities with linear congruences
 (GRID). The last column compares the domain of Convex Polyhedra with the

### 2012-05-05 papers/TAPAS2012/pagai.tex
TAPAS

--- 2011-11-05 papers/PLDI2012/implicitization.tex
@@
-\documentclass[preprint]{sigplanconf}
-
-% The following \documentclass options may be useful:
-%
-% 10pt          To set in 10-point type instead of 9-point.
-% 11pt          To set in 11-point type instead of 9-point.
-% authoryear    To obtain author/year citation style instead of numeric.
-
-%\newcommand{\MM}[1]{{\color{blue} TODO(MM): #1}}
-%\newcommand{\JH}[1]{{\color{teal} TODO(JH): #1}}
-%\newcommand{\DM}[1]{{\color{violet} TODO(DM): #1}}
-%\newcommand{\MM}[1]{}
-%\newcommand{\JH}[1]{}
-%\newcommand{\DM}[1]{}
-
+\documentclass{entcs} 
+\usepackage{entcsmacro}
 \usepackage{ifxetex}
@@
 \usepackage{mathcomp,amsfonts,amssymb,MnSymbol}
-\usepackage{dmnatbib}
+\usepackage{natbib}
 \usepackage[mathscr]{euscript}
@@
 \usepackage{listings}
-\lstset{language=C,mathescape=true,numbers=left,numbersep=0.8ex,tabsize=2,breaklines=true,basicstyle={\rm\fontfamily{phv}\small\selectfont}}
-
-\usepackage{tikz}
-\usetikzlibrary{arrows,automata}
-
-\tikzstyle{state}=[circle,fill=black!25,minimum size=13pt,inner sep=0pt]
-\tikzstyle{rstate}=[rectangle,fill=black!25,minimum size=13pt,inner sep=0pt]
-\tikzstyle{transition}=[rectangle,semithick,draw=black!75,
-  			  minimum size=4mm]
-\tikzstyle{transition2}=[transition,rectangle,thick,dashed,
-  			  minimum size=4mm]
-\tikzstyle{PRstate}=[circle,double,draw,fill=blue!15,minimum size=13pt,inner sep=0pt]
-\tikzstyle{polyhedra}=[blue!25,opacity=0.5,pattern=north west lines,pattern
-color=blue]
-\tikzstyle{line}=[black,thick]
-
-\usepackage[pdfusetitle]{hyperref}
 
@@
 
-% commentaires dans les algorithmes crees avec le package algorithmic
-\renewcommand{\algorithmiccomment}[1]{ {\color{gray} // #1}}
+\def\newblock{\hskip .11em plus .33em minus .07em}
 
+% The following is enclosed to allow easy detection of differences in
+% ascii coding.
+% Upper-case    A B C D E F G H I J K L M N O P Q R S T U V W X Y Z
+% Lower-case    a b c d e f g h i j k l m n o p q r s t u v w x y z
+% Digits        0 1 2 3 4 5 6 7 8 9
+% Exclamation   !           Double quote "          Hash (number) #
+% Dollar        $           Percent      %          Ampersand     &
+% Acute accent  '           Left paren   (          Right paren   )
+% Asterisk      *           Plus         +          Comma         ,
+% Minus         -           Point        .          Solidus       /
+% Colon         :           Semicolon    ;          Less than     <
+% Equals        =3D           Greater than >          Question mark ?
+% At            @           Left bracket [          Backslash     \
+% Right bracket ]           Circumflex   ^          Underscore    _
+% Grave accent  `           Left brace   {          Vertical bar  |
+% Right brace   }           Tilde        ~
+
+% A couple of exemplary definitions:
+
+\newcommand{\Nat}{{\mathbb N}}
+\newcommand{\Real}{{\mathbb R}}
+\def\lastname{Henry}
 \begin{document}
+\begin{frontmatter}
+  \title{PAGAI} 
+  \author{Julien Henry \thanksref{imag.fr}}
+  \address{VERIMAG\\ Universit\'e Joseph Fourier\\
+    Grenoble, France} 
+	\author{David Monniaux\thanksref{imag.fr}}
+  \address{VERIMAG\\CNRS\\
+    Grenoble, France} 
+	\author{Matthieu Moy\thanksref{grenoble-inp.fr}}
+  \address{VERIMAG\\Grenoble-INP\\
+    Grenoble, France}
 
-\conferenceinfo{PLDI '12}{11-16 June 2012, Beijing} 
-\copyrightyear{2012} 
-\copyrightdata{[to be supplied]}
+    \thanks[imag.fr]{Email:
+    \href{First Name.Last Name@imag.fr} {\texttt{\normalshape
+        First Name.Last Name@imag.fr}}} 
+    \thanks[grenoble-inp.fr]{Email:
+    \href{Matthieu.Moy@grenoble-inp.fr} {\texttt{\normalshape
+        Matthieu.Moy@grenoble-inp.fr}}} 
+\begin{abstract} 
+	We present design and implementation of PAGAI, a new static analyzer 
+	working over the LLVM compiler infrastructure, that computes inductive
+	invariants on the numerical variables of the program.
+	PAGAI implements various state-of-the-art techniques combining Abstract
+	Interpretation and Decision Procedures (SMT-solving), and offers
+	several abstract domains.
+	We thus can compare in time and precision the various combinations of
+	techniques and abstract domains. We conducted extensive experiments for
+	comparing these techniques on several benchmarks and GNU programs.
+\end{abstract}
+\begin{keyword}
+  Please list keywords from your paper here, separated by commas.
+\end{keyword}
+\end{frontmatter}
+\section{Introduction}\label{intro}
+Introduction
 
-%\titlebanner{banner above paper title}        % These are ignored unless
-\preprintfooter{Succinct representations for abstract interpretation}   % 'preprint' option specified.
-
-\title{Succinct Representations for Abstract Interpretation}
-\subtitle{Combined analysis algorithms and experimental evaluation}
-
-\authorinfo{Anonymous}
-           {The Internet}
-           {anonymous.gal@gmail.com}
-
-\maketitle
-
-\begin{abstract}
-Abstract interpretation techniques can be made more precise by distinguishing
-paths inside loops, at the expense of possibly exponential complexity. SMT-solving techniques and sparse representations of paths and sets of paths avoid this pitfall.
-
-We improve previously proposed techniques for guided static analysis and the generation of disjunctive invariants by combining them with techniques for succinct representations of paths and symbolic representations for transitions based on static single assignment.
-
-Because of the non-monotonicity of the results of abstract interpretation with widening operators, it is difficult to conclude that some abstraction is more precise than another based on theoretical local precision results.
-We thus conducted extensive comparisons between our new techniques and previous ones, on a variety of open-source packages.
-We also took the opportunity to compare the precision of various classical abstract domains (convex polyhedra, octagons, intervals).
-\end{abstract}
-
-% \category{CR-number}{subcategory}{third-level}
-\category{D.2.4}{Software/Program Verification}{Formal methods}
-\category{F.3.1}{Specifying and Verifying and Reasoning about Programs}{Assertions, Invariants, Mechanical verification, Pre- and post-conditions}
-\category{F.3.2}{Semantics of Programming Languages}{Program analysis}
-\category{F.4.1}{Mathematical Logic}{Logic and constraint programming, Mechanical theorem proving}
-\category{I.2.3}{Deduction and Theorem Proving}{Inference engines, Logic programming, Resolution}
-
-\terms
-theory, verification, algorithms
-
-\keywords
-SMT-solving, static analysis, abstract interpretation, disjunctions
-
-\section{Introduction}
-Static analysis by abstract interpretation is a fully automatic program analysis method.
-When applied to imperative programs, it computes an inductive invariant mapping each program location (or a subset thereof) to a set of states represented symbolically~\cite{CousotCousot_JLC92}.
-For instance, if we are only interested in scalar numerical program variables, such a set may be a convex polyhedron (the set of solutions of a system of linear inequalities)~\cite{CousotHalbwachs78,Halbwachs_PhD,PPL,BagnaraHZ08SCP}.
-
-In such an analysis, information may flow forward (one computes the polyhedron after a program statement as the image by the semantics of the statement, or a super-set thereof, of the polyhedron before) or backward;
-forward program analysis computes super-sets of the states reachable from the initialization of the program, backward program analysis computes super-sets of the states co-reachable from some property of interest (for instance, the violation of an assertion).
-In forward analysis, control-flow joins correspond to convex hulls if using convex polyhedra (more generally, they correspond to least upper bounds in a lattice); in backward analysis, it is control-flow splits that correspond to convex hulls.
-
-It is a known limitation of program analysis by abstract interpretation that this convex hull, or more generally, least upper bound operation, may introduce states that cannot occur in the real program: for instance, the convex hull of the intervals $[-2,-1]$ and $[1,2]$ is $[-2,2]$, strictly larger than the union of the two.
-Such introduction may prevent proving desired program properties, for instance $\neq 0$.
-The alternative is to keep the union symbolic (e.g.
-compute using $[-2,-1] \cup [1,2]$) and thus compute in the \emph{disjunctive completion}
-of the lattice, but the number of terms in the union may grow exponentially with the number of successive tests in the program to analyze, not to mention difficulties for designing suitable widening operators for enforcing the convergence of fixpoint iterations~\cite{PPL,BagnaraHZ08SCP,DBLP:journals/sttt/BagnaraHZ07}.
-The exponential growth of the number of terms in the union may be controlled by heuristics that judiciously apply least upper bound operations, as in the \emph{trace partitioning domain} \cite{Rival_Mauborgne_TOPLAS07} implemented in the Astr\'ee analyzer~\cite{ASTREE_PLDI03,ASTREE_ESOP05}.
-
-Assuming we are  interested in a loop-free program fragment, the above approach of keeping symbolic unions gives the same results as performing the analysis separately over every path in the fragment.
-A recent method for finding disjunctive loop invariants \cite{DBLP:conf/pldi/GulwaniZ10} is based on this idea: each path inside the loop body is considered separately.
-Two recent proposals use SMT-solving \cite{Kroening_Strichman_08} as a decision procedure for the satisfiability of first-order arithmetic formulas in order to enumerate only paths that are needed for the progress of the analysis \cite{Gawlitza_Monniaux_ESOP11,Monniaux_Gonnord_SAS11}.
-They can equivalently be seen as analyses over a multigraph of transitions between some distinguished control nodes.
-This multigraph has an exponential number of edges, but is never explicitly represented in memory; instead, this graph is \emph{implicitly} or \emph{succinctly} represented: its edges are enumerated as needed as solutions to SMT problems.
-
-An additional claim of the methods that distinguish paths inside the loop body \cite{DBLP:conf/pldi/GulwaniZ10,Monniaux_Gonnord_SAS11} is that they tend to generate better invariants than methods that do not, by behaving better with respect to the \emph{widening operators} \cite{CousotCousot_JLC92}
-used for enforcing convergence when searching for loop invariants by Kleene iterations.
-A related technique, \emph{guided static analysis} \cite{DBLP:conf/sas/GopanR07}, computes successive loop invariants for increasing subsets of the transitions taken into account, until all transitions are considered; again, the claim is that this approach avoids some gross over-approximation introduced by widenings.
-
-All these methods improve the precision of the analysis by keeping the same abstract domain (say, convex polyhedra) but changing the operations applied and their ordering.
-An alternative is to change the abstract domain, for instance by moving from intervals to octagons or convex polyhedra \cite{DBLP:journals/lisp/Mine06}, or the widening operator: for instance one may replace the classical widening on convex polyhedra \cite{CousotHalbwachs78,Halbwachs_PhD} by an improved version \cite{BagnaraHRZ05SCP}, or use widening ``up to'', that is, use linear inequalities syntactically present in the program source code as possible widening steps~\cite{Polka:FMSD:97}.
-
-There are many possible combinations of the above techniques, and it is not
-evident which ones perform more or less precisely or more or less efficiently on real-life examples. One needs to experiment.
-Unfortunately, the published literature on the subject lacks experimental comparative assessments. One purpose of this article is therefore to propose such experimental results.
-
-This article makes the following contributions:
-\begin{enumerate}
-\item We recast the guided static analysis technique from \citet{DBLP:conf/sas/GopanR07} on the expanded multigraph from \citet{Monniaux_Gonnord_SAS11}, considering entire paths instead of individual transitions, using SMT queries and binary decision diagrams.\label{contr:guided_multigraph} (See \S\ref{sec:guided_multigraph})
-\item We improve the technique for obtaining disjunctive invariants from \citet{DBLP:conf/pldi/GulwaniZ10} by replacing the explicit exhaustive enumeration of paths by a sequence of SMT queries.\label{contr:disjunctive} (See \S\ref{sec:disjunctive})
-\item We implemented these techniques, in addition to ``classical'' iterations and the original guided static analysis, inside a prototype static analyzer.
-This tool uses the LLVM bitcode format \cite{Lattner:2004:LCF:977395.977673,LLVM_langref} as input, which can be produced by compilation from C, C++ and Fortran, enabling it to be run on many real-life programs.
-With respect to abstract domains on numerical variables, it uses the APRON library \cite{DBLP:conf/cav/JeannetM09}, which supports a variety of abstract domains from which we can choose with minimal changes to our analyzer.
-
-Our tool uses a \emph{single static assignment} (SSA) intermediate representation, which allows us to perform cheap symbolic propagations and simplifications,
-described in Section~\ref{sec:analysis-algorithm}.
-
-\item We conducted extensive experiments with this tool, on real-life programs: (i) fixing the abstract domain, varying the iteration technique (\S~\ref{sec:compare_techniques}) (ii) fixing the iteration technique, varying the abstract domain (\S\ref{sec:compare_domains}).
-\end{enumerate}
-
-\section{Bases}
-\subsection{Static analysis by abstract interpretation}
-\label{sec:static_analysis}
-Let $X$ be the set of possible states of the program variables; for instance, if the program has 3 unbounded integer variables, then $X = \ZZ^3$.
-The set $\parts{X}$ of subsets of $X$, partially ordered by inclusion, is the \emph{concrete domain}.
-An \emph{abstract domain} is a set $\abstr{X}$ equipped with a partial order $\sqsubseteq$; for instance, it can be the domain of convex polyhedra in $\QQ^3$ ordered by geometric inclusion.
-In this article, all abstract domains are supposed to contain machine-representable objects, and all $\sqsubseteq$ order relations are supposed to be decidable.
-The concrete and abstract domains are connected by a monotone \emph{concretization} function $\gamma: \left(\abstr{X},\sqsubseteq\right) \rightarrow (\parts{X},\subseteq)$: an element $\abstr{x} \in \abstr{X}$ represents a set $\gamma(\abstr{x})$.%
-%
-\footnote{%
-Some presentations of abstract interpretation identify an element $\abstr{x}$ with the set of states $\gamma(\abstr{x})$ that it represents.
-This leads to simpler notations, but may also confuse if, as in some of our constructions, there exist several $\abstr{x}$ with identical $\gamma(\abstr{x})$, but distinct algorithmic behaviors.}
-
-We also assume a join operator $\sqcup: \abstr{X} \times \abstr{X} \rightarrow \abstr{X}$, with infix notation; in practice, it is generally a least upper bound operation, but we only need it to satisfy $\gamma(\abstr{x}) \cup \gamma(\abstr{y}) \subseteq \gamma(\abstr{x} \sqcup \abstr{y})$ for all $\abstr{x},\abstr{y}$.
-
-Classically, one considers the control-flow graph of the program, with edges labeled with concrete transition relations (e.g.
-$x' = x+1$ for an instruction \lstinline|x = x+1;|), and attaches an abstract element to each control point.
-A concrete transition relation $\tau \subseteq X \times X$ is replaced by an abstract \emph{forward abstract transformer} $\abstr{\tau}: \abstr{X} \rightarrow \abstr{X}$, such that
-\begin{equation}
-\forall \abstr{x} \in \abstr{X}, x,x' \in X,~
-x \in \gamma(\abstr{x}) \land (x,x') \in \tau \implies
-x' \in \gamma \circ \abstr{\tau}(\abstr{x})
-\label{eqn:transformer_correct}
-\end{equation}
-It is easy to see that if to any control point $p \in P$ we attach an abstract element $\abstr{x}_p$ such that
-(i) for any $p$, $\gamma(\abstr{x}_p)$ includes all initial states possible at control node $p$
-(ii) for any $p,p'$, $\abstr{\tau}_{p,p'} (\abstr{x}_p) \sqsubseteq \abstr{x}_{p'}$, noting $\tau_{p,p'}$ the transition from $p$ to~$p'$, then $(\gamma(\abstr{x}_p))_{p \in P}$ form an \emph{inductive invariant}: by induction, when the control point is $p$, the program state always lies in $\gamma(\abstr{x}_p)$.
-
-\emph{Kleene iterations} compute such an inductive invariant as the stationary limit, if it exists, of the following system: for each $p$, initialize $\abstr{x}_p$ such that $\gamma(\abstr{x}_p)$ is a superset of the initial states at point $p$; then iterate the following:
-if $\abstr{\tau}_{p,p'} (\abstr{x}_p) \nsqsubseteq \abstr{x}_{p'}$, replace $\abstr{x}_{p'}$ by $\abstr{x}_{p'} \sqcup \abstr{\tau}_{p,p'} (\abstr{x}_p)$.
-Such a stationary limit is bound to exist if $\abstr{X}$ has no infinite ascending chain $a_1 \sqsubsetneq a_2 \sqsubsetneq \dots$; this condition is however not met by domains such as intervals or convex polyhedra.
-
-\emph{Widening-accelerated Kleene iterations} proceed by replacing $\abstr{x}_{p'} \sqcup \abstr{\tau}_{p,p'} (\abstr{x}_p)$ by $\abstr{x}_{p'} \widening (\abstr{x}_{p'} \sqcup \abstr{\tau}_{p,p'} (\abstr{x}_p))$ where $\widening$ is a \emph{widening operator}: for all $\abstr{x},\abstr{y}$, $\gamma(\abstr{y}) \subseteq \gamma(\abstr{x} \widening \abstr{y})$, and there exists no sequence $\abstr{u}_1,\abstr{u}_2,\dots$ of the form $\abstr{u}_{n+1} = \abstr{u}_n \widening \abstr{v}_n$ where $\abstr{v}_n$ is another sequence.
-Such iterations necessarily lead to a stationary limit $(\abstr{x}_p)_{p \in P}$, which defines an inductive invariant $(\gamma(\abstr{x}_p))_{p \in P}$.
-Note that this invariant is not, in general, the least one expressible in the abstract domain, and may depend on the iteration ordering (the successive choices~$p,p'$).
-
-Once an inductive invariant $\gamma((\abstr{x}_p)_{p \in P})$ has been obtained,
-one can attempt \emph{decreasing} or \emph{narrowing} iterations to reduce it.
-In their simplest form, this just means running the following operation until a
-fixpoint or a maximal number of iterations are reached: for any $p'$, replace
-$\abstr{x}_{p'}$ by $\abstr{x}_{p'} \cap \left(\bigsqcup_{p \in P}
-\abstr{\tau}_{p,p'} (\abstr{x}_p)\right)$. The result also defines an inductive
-invariant. These decreasing iterations are indispensable to recover properties
-from guards (tests) in the program in most iteration settings; unfortunately,
-certain loops, particularly those involving identity (no-operation) transitions,
-may foil them: the iterations immediately reach a fixpoint and do not decrease
-further (see example in \S\ref{subsec:rate_lim}). Sections \ref{sec:guided} and
-\ref{sec:path_focusing} describe techniques that work around this problem.
-
-Widening operations have a somewhat counterintuitive behavior: the result of an analysis (postcondition) is not necessarily monotonic with respect to the precondition.
-For instance, classical interval analysis with widening and decreasing iterations on the following loop yields a loop invariant $x \in [0,+\infty)$ if the precondition is $x \in [0,0]$ , but $x \in [0,10]$ (strictly stronger) if the precondition is $x \in [0,10]$ (strictly weaker):
-\lstinline|while (x != 10) x=x+1;|.
-This means that, when using widenings, being locally more precise (for instance, by using more expressive abstract domains, or widening operators that climb more slowly \cite{BagnaraHRZ05SCP}) does not necessarily translate into higher final precision.
-It is almost always possible to concoct examples where a supposedly more precise abstraction (e.g.
-polyhedra) yields less precise results than a less precise one (e.g.
-intervals), but are these examples representative of real programs?
-This justifies our recourse to evaluation on realistic examples~(\S~\ref{sec:compare_techniques}).
-
-\subsection{SMT-solving}
-Boolean satisfiability (SAT) is the canonical NP-complete problem: given a propositional formula (e.g.
-$(a \lor \neg b) \land (\neg a \lor b \lor \neg c)$), decide whether it is satisfiable --- and, if so, output a satisfying assignment.
-Despite an exponential worst-case complexity, the DPLL algorithm \cite{Kroening_Strichman_08,Handbook_SAT} solves many useful SAT problems in practice.
-
-SAT was extended to \emph{satisfiability modulo theory} (SMT): in addition to propositional literals, SMT formulas admit atoms from a theory.
-For instance, the theories of linear integer arithmetic (LIA) and linear real arithmetic (LRA) have atoms of the form $a_1 x_1 + \dots + a_n x_n \bowtie C$ where $a_1,\dots,a_n,C$ are integer constants, $x_1,\dots,x_n$ are variables (interpreted over $\ZZ$ for LIA and $\RR$ or $\QQ$ for LRA), and $\bowtie$ is a comparison operator $=,\neq,<,\leq,>,\geq$.
-Satisfiability for LIA and LRA is NP-complete, yet tools based on DPLL(T) approach \cite{Kroening_Strichman_08,Handbook_SAT} solve many useful SMT problems in practice.
-All these tools provide a \emph{satisfying assignment} if the problem is satisfiable.
-
-Most SMT solvers, including Z3,
-%\footnote{\url{http://research.microsoft.com/en-us/um/redmond/projects/z3/}}
-Yices,%
-%\footnote{\url{http://yices.csl.sri.com/}}
-and all those supporting the full SMTLIB2 standard \cite{BarST-SMTLIB},
-offer an \emph{incremental} interface: the client program specifies the formula as an initially empty conjunction, to which additional constraints are added, and calls a ``check'' function answering whether it is satisfiable;
-it may then backtrack some of the constraints and add other ones without restarting from scratch.
-
-\subsection{A simple, motivating example}
-\label{subsec:rate_lim}
-Consider the following program, adapted from \cite{Monniaux_Gonnord_SAS11}, where \lstinline|input($a$, $b$)| stands for a nondeterministic input in $[a,b]$:
-\lstinputlisting{rlim_int.c}
-This program implements a construct commonly found in control programs (in e.g.
-automotive or avionics): a rate limiter. For the sake of simplicity, we chose it
-to be fed on a nondeterministic input clamped between $[-100000,100000]$, but in a real system it would be integrated in a reactive control loop and its input connected to a complex system with unknown output range.
-
-The expected inductive invariant is $\lstinline|x_old| \in [-100000,100000]$, but classical abstract interpretation using intervals (or octagons or polyhedra) finds $\lstinline|x_old| \in (-\infty,+\infty)$~\citep{ASTREE_ESOP05}.
-Let us briefly see why.
-
-Widening iterations converge to $\lstinline|x_old| \in (-\infty,+\infty)$; let us now see why decreasing iterations fail to recover the desired invariant.
-The \lstinline|x > x_old+10| test at line~6, if taken, yields $\lstinline|x_old|
-\in (-\infty,99990)$; followed by \lstinline|x = x_old+10|, we obtain
-$\lstinline|x| \in (-\infty,\allowbreak 100000)$, and the same after union with the no-operation ``else'' branch. Line~7 yields $\lstinline|x| \in (-\infty,+\infty)$.
-
-We could use ``widening up to'' or ``widening with thresholds'', propagating the ``magic values'' $\pm 100000$ associated to \lstinline|x| into \mbox{\lstinline|x_old|,} but these syntactic approaches cannot directly cope with programs for which $\lstinline|x|  \in [-100000,+100000]$ is itself obtained by analysis.
-The guided static analysis of \citet{DBLP:conf/sas/GopanR07} does not perform
-better, and also obtains $\lstinline|x_old| \in (-\infty,+\infty)$.
-
-In contrast, let us distinguish all four possible execution paths through the tests at lines 6 and~7. The path through both ``else'' branches is infeasible; the program is thus equivalent to:
-\lstinputlisting{rlim_int2.c}
-
-\begin{figure}
-\centering
-\begin{minipage}[c]{.19\textwidth}
-\begin{tikzpicture}[->,>=stealth',auto,node distance=1.1cm,
-                    semithick,font=\footnotesize]
-
-	\node[PRstate] (n0) {$p$};
-	\node[state] (n1) [below left of=n0] {};
-	\node[state] (n2) [below right of=n0] {};
-	\node[state] (n3) [below right of=n1] {};
-	\node[state] (n4) [below left of=n3] {};
-	\node[state] (n5) [below right of=n3] {};
-	\node[state] (n6) [below right of=n4] {};
-	\node[state] (n7) [below left of=n6] {};
-	\node[state] (n8) [below right of=n6] {};
-	\node[PRstate] (n9) [below right of=n7] {$q$};
-
-  \path [transition] 
-		(n0) edge              node {} (n1);
-  \path [transition] 
-		(n0) edge              node {} (n2);
-  \path [transition] 
-		(n1) edge              node {} (n3);
-  \path [transition] 
-		(n2) edge              node {} (n3);
-  \path [transition] 
-		(n3) edge              node {} (n4);
-  \path [transition] 
-		(n3) edge              node {} (n5);
-  \path [transition] 
-		(n4) edge              node {} (n6);
-  \path [transition] 
-		(n5) edge              node {} (n6);
-  \path [transition] 
-		(n6) edge              node {} (n7);
-  \path [transition] 
-		(n6) edge              node {} (n8);
-  \path [transition] 
-		(n7) edge              node {} (n9);
-  \path [transition] 
-		(n8) edge              node {} (n9);
-\end{tikzpicture}
-\end{minipage} 
-$\Longrightarrow$
-\begin{minipage}[c]{.19\textwidth}
-\begin{tikzpicture}[->,>=stealth',auto,node distance=1.1cm,
-                    semithick,font=\footnotesize]
-
-	\node[PRstate] (n0) {$p$};
-	\node (n1) [below left of=n0] {};
-	\node (n2) [below right of=n0] {};
-	\node (n3) [below right of=n1] {};
-	\node (n4) [below left of=n3] {};
-	\node (n5) [below right of=n3] {};
-	\node (n6) [below right of=n4] {};
-	\node (n7) [below left of=n6] {};
-	\node (n8) [below right of=n6] {};
-	\node[PRstate] (n9) [below right of=n7] {$q$};
-
-  \path [transition] 
-		(n0) edge  [out=0, in=0]            node {} (n9);
-  \path [transition] 
-		(n0) edge  [out=180, in=180]        node {} (n9);
-  \path [transition] 
-		(n0) edge  [out=205, in=155]            node {} (n9);
-  \path [transition] 
-		(n0) edge  [out=230, in=130]            node {} (n9);
-  \path [transition] 
-		(n0) edge  [out=255, in=105]            node {} (n9);
-  \path [transition] 
-		(n0) edge  [out=-25, in=25]            node {} (n9);
-  \path [transition] 
-		(n0) edge  [out=-50, in=50]            node {} (n9);
-  \path [transition] 
-		(n0) edge  [out=-75, in=75]            node {} (n9);
-\end{tikzpicture}
-\end{minipage}
-\caption{Expansion of the transition graph into a multigraph.}
-\label{fig:multigraph}
-\end{figure}
-
-Classical interval analysis on this program yields $\lstinline|x_old| \in [-100000,100000]$.
-We have transformed the first program into the second, manually pruning out infeasible paths; yet in general the resulting program could be exponentially larger than the first (as in Fig.~\ref{fig:multigraph}), even though not all feasible paths are needed to compute the invariant.
-
-Following recent suggestions \cite{Gawlitza_Monniaux_ESOP11,Monniaux_Gonnord_SAS11}, we avoid this space explosion by keeping the second program implicit while simulating its analysis.
-This means we work on an implicitly represented transition multigraph (Fig.~\ref{fig:multigraph}); it is succinctly representated by the transition graph of the first program.
-Our first contribution (\S\ref{sec:guided_multigraph}) is to recast the ``guided analysis'' from \citet{DBLP:conf/sas/GopanR07} on such a succinct representation of the paths in lieu of the individual transitions.
-A similar explosion occurs in disjunctive invariant generation, following \citet{DBLP:conf/pldi/GulwaniZ10}; our second contribution (\S\ref{sec:disjunctive}) applies our implicit representation to their method.
-
-\subsection{Guided static analysis}
-\label{sec:guided}
-\emph{Guided static analysis} was proposed by \citet{DBLP:conf/sas/GopanR07} as an improvement over classical upward Kleene iterations with widening.
-Consider the following program, taken from \cite{DBLP:conf/sas/GopanR07}:
-\lstinputlisting[label=lst:gopan_reps]{gopan_reps.c}
-
-\begin{figure}
-\begin{center}
-\includegraphics[scale=0.75]{figures/gopan_reps}
-\end{center}
-\caption{The invariant for Prog.~\ref{lst:gopan_reps}: the piecewise linear, solid line is the strongest invariant, the grayed polyhedron is its convex hull.}
-\label{fig:gopan_reps_invariant} 
-\end{figure}
-
-Classical iterations on the domain of convex polyhedra \cite{CousotHalbwachs78,BagnaraHRZ05SCP} or octagons \cite{DBLP:journals/lisp/Mine06} start with $x = 0 \land x = 0$, then continue with $x = y \land 0 \leq x \leq 1$.
-The widening operator extrapolates from these two iterations and yields $x = y \land x \geq 0$.
-From there, the ``else'' branch at line 5 may be taken; with further widening, $0 \leq y \leq x$ is obtained as a loop invariant, and thus the postcondition computed at line 9 is $x \geq 0 \land y = 0$.
-Yet the strongest invariant is $(0 \leq x \leq 51 \land y = x) \lor (51 \leq x \leq 102 \land x+y=102)$, and its convex hull, a convex polyhedron (Fig.~\ref{fig:gopan_reps_invariant}), is
-\begin{equation}
-y \leq x \land y \leq 102-x \land y \geq 0.\label{eqn:triangle}
-\end{equation}
-Intuitively, this disappointing result is obtained because widening extrapolates from the first iterations of the loop, but the loop has two different phases ($x \leq 50$ and $x > 50$) with different behaviors, thus the extrapolation from the first phase is not valid for the second.
-
-\citeauthor{DBLP:conf/sas/GopanR07}' idea is to analyze the first phase of the loop with a widening and narrowing sequence, and thus obtain $0 \leq x \leq 50 \land y = x$, and then analyze the second phase, finally obtaining invariant~\ref{eqn:triangle}; each phase is identified by the tests taken or not taken.
-
-The analysis starts by identifying the tests taken and not taken during the first iteration of the loop, starting in the loop initialization.
-The branches not taken are pruned from the loop body, yielding:
-\begin{lstlisting}[numbers=none]
-  while(1) {
-    if(x <= 50) y++;
-    else break; /* not taken in phase 1 */
-    if(y < 0) break;
-    x++;
-  }
-\end{lstlisting}
-
-Analyzing this loop using widening and narrowing on convex polyhedra or octagons yields the loop invariant $0 \leq x \leq 51 \land y = x$.
-Now, the transition at line~5 becomes feasible; and we analyze the full loop, starting iterations from $0 \leq x \leq 51 \land y = x$, and obtain invariant~(\ref{eqn:triangle}).
-
-More generally, this analysis method considers an ascending sequence of subsets of the transitions in the loop body (left side of Fig.~\ref{fig:multigraph});
-for each subset, an inductive invariant is computed for the program restricted to it.
-The starting subset are the transitions reachable in one step from the loop initialization.
-If for a given subset $S$ in the sequence, no transitions outside $S$ are reachable from the inductive invariant attached to $S$, then iterations stop;
-otherwise, add these transitions to $S$ and iterate more.
-Termination ensues from the finiteness of the control-flow graph.
-
-\subsection{Path-focusing}
-\label{sec:path_focusing}
-
-\citet{Monniaux_Gonnord_SAS11}'s \emph{path-focusing} technique
-distinguishes the different paths in the program in order to avoid loss of
-precision due to merge operations. Since the number of paths may be exponential,
-the technique keeps them implicit and computes them when needed using
-SMT-solving.
-The (accelerated) Kleene iterations (\S\ref{sec:static_analysis}) are computed over a reduced multigraph instead of the classical transition graph.
-
-Let $P$ be the set of control points in the transition graph, 
-$P_W \subseteq P$ the set of widening points
-such that removing the points in $P_W$ gives an acyclic graph.
-One can choose a set $P_R$ such that $P_W \subseteq P_R \subseteq P$.
-
-The set of paths is kept implicit by an SMT formula $\rho$ expressing
-the semantics of the program, assuming that the transition semantics can be
-expressed within a decidable theory. For an easy construction of $\rho$, 
-we also assume that the program is expressed in SSA form, meaning that each
-variable is only assigned once in the transition graph. This is not a
-restriction, since there exist standard algorithms that transform a program into
-an SSA format.
-
-This formula contains Boolean \emph{reachability predicates} $b_i$ for each
-control points $p_i \notin P_R$, $b_i^s$ and $b_i^d$ for each $p_i \in P_R$, so
-that a path 
-$p_{i_1} \rightarrow p_{i_2} \rightarrow \dots \rightarrow p_{i_n}$ 
-between two points $p_{i_1}, p_{i_n} \in P_R$ 
-can easily be expressed as the
-conjunction $b_{i_1}^s \wedge \bigwedge_{2 \leq k < n} b_{i_k} \wedge b_{i_n}^d$.
-The Boolean $b_{i}^s$ is $true$ when the path starts at point $p_i$, whereas
-$b_i^d$ is $true$ when the path arrives at $p_i$. In other words, we split the
-points in $P_R$ into a \emph{source} point, with only outgoing transitions, and
-a \emph{destination} point, with only incoming transitions, so that the
-resulting graph is acyclic and there is no paths going through control
-points in $P_R$.
-
-In order to find focus paths, we solve an SMT formula which is satisfiable when
-there exist a path starting at a point $p_i \in P_R$ in a state included in the
-current invariant candidate $X_i$, and arriving at a point $p_j \in P_R$ in a
-state outside $X_j$. In this case, we construct this path using the model and
-update $X_j$. When $p_i = p_j$, meaning that the path is actually a self-loop,
-we can apply a widening/narrowing sequence, or even compute the transitive
-closure of the loop (or an approximation thereof, or its application to $X_i$)
-using abstract acceleration~\cite{DBLP:conf/sas/GonnordH06}.
-
-\section{Guided analysis over the paths}
-\label{sec:guided_multigraph}
-
-Guided static analysis, as proposed by \citet{DBLP:conf/sas/GopanR07}, applies to the transition
-graph of the program. We now present a new technique applying this analysis on the implicit
-multigraph from \cite{Monniaux_Gonnord_SAS11}, thus avoiding control flow merges with
-unfeasible paths.
-In this section, we use the same notations as \ref{sec:path_focusing}, except
-that we call the abstract values $X_i^s$ instead of $X_i$ to leave room for
-another abstract value $X_i^d$ later in the algorithm.
-
-The combination of these two techniques aims at first discovering a precise
-inductive invariant for a subset of paths between two points in $P_R$, 
-by the mean of ascending and narrowing iterations. When an
-inductive invariant has been found, we add new feasible paths to the subset and
-compute an inductive invariant for this new subset, starting with the results
-from the previous analysis.
-In other words, our technique considers an ascending sequence of
-subsets of the paths between two points in $P_R$.
-We iterate the operations until the whole program (i.e all the
-feasible paths) has been considered. The result will then be an
-inductive invariant of the entire program.
-
-\subsection{Algorithm}
-
-Algorithm~\ref{algo:combined} performs Guided static analysis on the implicitly represented multigraph.
-
-The current working subset of paths, noted $P$ and initially empty, is
-stored using a compact representation, such as binary decision
-diagrams. We also maintain two sets of control points:
-\begin{itemize}
-	\item $A'$ contains the points in $P_R$ that may be the starting points of new
-		feasible paths.
-	\item $A$ contains the points in $P_R$ on which we apply the ascending iterations.
-	Each time the abstract value of a control point $p$ is updated, $p$ is
-	inserted in both $A$ and $A'$.
-\end{itemize}
-
-\begin{algorithm}
-	\caption{Guided static analysis on implicit multigraph}
-	\label{algo:combined}
-	\begin{algorithmic}[1] 
-	\input{algo/algo-combined.tex}
-	\end{algorithmic}
-\end{algorithm}
-
-We distinguish three phases in the main loop of the analysis:
-\begin{enumerate}
-\item \label{step:addingpaths} We start finding a new relevant subset
-  $P \cup P'$ of the graph.
-  Either the previous iteration or the initialization lead us to a
-  state where there are no more paths in the previous subset $P$,
-  starting at $p_i$, that make the abstract values of the successors
-  grow (otherwise, the SMT solver would not have answered
-  ``\emph{unsat}''). Narrowing iterations preserve this property.
-  However, there may exist such paths in the entire multigraph, that
-  are not in $P$. This phase computes these paths and adds them to
-  $P'$. This phase is described in~\ref{subsec:addingpaths}
-  and corresponds to lines in~\ref{alg=start-add-paths} to
-  \ref{alg=end-add-paths} in Algorithm~\ref{algo:combined}.
-\item \label{step:ascending} Given a new subset $P$, we search for paths starting at point
-  $p_i \in P_R$, such that these paths are in $P$, i.e are included in
-  the working subgraph. Each time we find a path, we update the
-  abstract value of the destination point of the path. This is the
-  phase explained in~\ref{subsec:ascending}, and corresponds to
-  lines~\ref{alg=start-ascending} to~\ref{alg=end-ascending} in
-  Algorithm~\ref{algo:combined}.
-\item  \label{step:narrowing} We perform narrowing iterations the usual way
-  (line~\ref{alg=narrowing} in algorithm~\ref{algo:combined}) and
-  reiterate from step 1 unless there are no more points to explore,
-  i.e. $A' = \emptyset$.
-\end{enumerate}
-
-%\MM{Je trouve qu'il manque une vue de haut niveau quelque part par
-%  ici. Dire qu'en gros, l'algo, c'est (en pseudo-code ou en langue naturelle) :
-%
-%Sans ca, enchainer tout de suite sur ``ascending iterations'' est un peu brutal.
-%
-%Ensuite, on peut rafiner la définition de ``Subset'' ($A$ et $P$), et
-%enchainer sur l'algo 2.
-%}
-
-Step~\ref{step:ascending} corresponds to the application of
-path-focusing \cite{Monniaux_Gonnord_SAS11} to
-the elements of $A$ until $A$ is empty. 
-When $A$ becomes empty, it means the
-invariant computation of the subgraph is finished. As proposed by
-\citet{DBLP:conf/sas/GopanR07}, we can do some narrowing iterations.
-These narrowing iterations allow to recover precision lost by
-widening, \emph{before} computing and taking into account new feasible paths.
-Thus, our technique combines both the advantages of \emph{Guided Static
-Analysis} and \emph{Path-focusing}.
-
-The order of steps is important: narrowing has to be performed before
-adding new paths, or some spurious new paths would be added to $P$,
-and starting with the addition of new paths avoids wasting time
-doing the ascending iterations on an empty graph.
-
-\subsection{Ascending iterations by Path-focusing}
-\label{subsec:ascending}
-
-%\MM{Maintenant que l'ordre a changé dans l'algo, on peut peut-être
-%  échanger cette section et la suivante ?}
-
-For computing an inductive invariant over a subgraph, we use the
-Path-focusing algorithm from \citet{Monniaux_Gonnord_SAS11} with special
-treatment for self loops (line~\ref{alg=pf} in algorithm~\ref{algo:combined}).
-
-In order to find which path to focus on, we construct an SMT formula $f(p_i)$, whose
-model when satisfiable is a path that starts in $p_i$, goes to a successor $p_j
-\in P_R$ of $p_i$, such that the image of $X_{i}^s$ by the path transformation
-is not included in the current $X_{j}^s$.
-Intuitively, such a path makes the abstract value $X_{j}^s$ grow, and thus is
-an interesting path to focus on. We loop until the formula becomes unsatisfiable,
-meaning that the analysis of $p_i$ is finished.
-
-If we note $Succ(i)$ the set of indices $j$ such that $p_j \in P_R$ is a
-successor of $p_i$ in the expanded multigraph, and $X_i^s$ the abstract value
-associated to $p_i$ :
-$$f(p_i) = \rho \wedge b_i^s \wedge 
-\displaystyle\bigwedge_{j \in P_R \atop j \neq i} \neg
-b_j^s \wedge X_i^s \wedge \displaystyle\bigvee_{j \in Succ(i)} (b_j^d \wedge
- \neg X_j^s)$$
-
-The difference with~\citet{Monniaux_Gonnord_SAS11} is that we do not
-work on the entire transition graph but on a subset of it. Therefore we
-conjoin the formula $f(p_i)$ with the actual set of working paths,
-noted $P$, expressed as a Boolean formula, where the Boolean variables are the
-\emph{reachability predicates} of the control points. We can easily construct
-this formula from the binary decision diagram using dynamic programming, and
-avoiding an exponentially sized formula. In other words, we force the SMT solver
-to give us a path included in $P$.
-
-\subsection{Adding new paths}
-\label{subsec:addingpaths}
-
-Our technique computes the fixpoint iterations on a ascending sequence of
-subgraphs, until the complete graph is reached.
-When the analysis of a subgraph is finished, meaning that the abstract values
-for each control points has converged to an inductive invariant for this subgraph,
-the next subgraph to work on has to be computed.
-
-This new subgraph is the union of the paths of the previous one with a set
-$P'$ of new paths that become feasible regarding the current abstract values.
-The paths in $P'$ are computed one after another, until no more path
-can make the invariant grow. This is line~\ref{alg=computeNewPaths} in
-Algorithm~\ref{algo:combined}, which corresponds to
-Algorithm~\ref{algo:computepaths}. We also use SMT solving to discover
-these new paths, but we subtly change the SMT formula given to the
-SMT solver: we now simply check for the satisfiability of $f(p_i)$
-instead of $f(p_i) \wedge P$. Since we already know that $f(p_i)
-\wedge P$ is unsatisfiable, none of the paths given by the SMT solver
-will be in $P$.
-
-% À changer si on décide de ne plus distinguer $X^s$ et $X^d$
-Also, to prevent the algorithm from adding too many paths at a time,
-and in particular to prevent the algorithm from trying to re-add the
-same path infinitely many times, we use another abstract value
-associated to the control point $p_j$, noted $X_j^d$, which is distinct
-from $X_j^s$, and initialized to $X_j^s$ right before computing new paths. 
-In the SMT formula, we associate to $p_j$ the abstract value $X_j^s$ when $p_j$
-is the starting point of the path, and $X_j^d$ when it is its destination point.
-We thus check the satisfiability of the formula
-$f'(p_i)$, where:
-$$f'(p_i) = \rho \wedge b_i^s \wedge 
-\displaystyle\bigwedge_{j \in P_R \atop j \neq i} \neg
-b_j^s \wedge X_i^s \wedge \displaystyle\bigvee_{j \in Succ(i)} (b_j^d \wedge
-\neg X_j^d)$$
-
-$X_j^d$ is updated when the point $p_j$ is the target of a new path.
-This way, further SMT queries do not
-compute other paths with the same source and destination if it is not
-needed (because these new paths would not make $X_j^d$ grow, hence
-would not be returned by the SMT solver).
-
-\begin{algorithm}
-	\caption{ComputeNewPaths}
-	\label{algo:computepaths}
-	\begin{algorithmic}[1] 
-	\input{algo/compute-paths.tex}
-	\end{algorithmic}
-\end{algorithm}
-
-\subsection{Termination}
-Termination of this algorithm is guaranteed, because:
-\begin{itemize}
-\item 
-the subset of paths $P$ strictly increases at each loop iteration, and bounded by the finite set of
-paths in the entire graph. 
-\item the set $P'$ always verifies $P \cap P' = \emptyset$ by construction, 
-which guarantees that $P'$
-will eventually be empty after a finite number of loop iterations.
-\end{itemize}
-
-\subsection{Example}
-
-We revise the rate limiter described in \ref{subsec:rate_lim}. In this example,
-\emph{Path-focusing} works well because all the paths starting at the
-loop header are actually self loops. In such a case, the technique performs a
-widening/narrowing sequence or accelerates the loop, thus leading to a precise
-invariant. However, in some cases, there also exist paths that are not
-self loops, in which case \emph{Path-focusing} applies widening. 
-This widening may induce unrecoverable loss of precision.
-
-Suppose the main loop of the rate limiter contains a nested loop:
-\lstinputlisting{rlim_int_loop.c}
-
-We choose $P_R$ as the set of loop headers of the function, plus the initial
-state. In this case, we have three elements in $P_R$.
-We also unroll the loop once, in order to distinguish the paths that do not go
-through the loop from the others:
-\begin{lstlisting}[numbers=none]
-if (wait()) {
-   while (wait()) {}
-}
-\end{lstlisting}
-
-The main loop in the expanded multigraph has then 8 distinct paths, half of
-them being self loops, and the other half going to the header of the nested loop.
-
-Guided static analysis from \citet{DBLP:conf/sas/GopanR07} yields, at line 5,
-$\lstinline|x_old| \in (-\infty,+\infty)$.
-Path-focusing \citep{Monniaux_Gonnord_SAS11} performs slightly
-better, and finds $\lstinline|x_old| \in (-\infty,10000]$.
-Now, let us see how our technique performs on this example.
-
-Figure \ref{fig:example-graph} shows the sequence of subset of paths during the
-analysis. The points in $P_R$ are noted $p_i$, where $i$ is the corresponding
-line in the code: for instance, $p_5$ corresponds to the header of the main
-loop.
-
-\begin{enumerate}
-	\item The first subset of paths is depicted on Figure
-		\ref{fig:example-graph} Step 1. We apply Path-Focusing on this graph,
-		and discover for $p_5$ the inductive invariant $\lstinline|x_old| = 0$.
-	\item
-		We then compute the set $P'$ of paths that have to be added into the
-		next subgraph. 
-		The image of $\lstinline|x_old| = 0$ by the path that goes from $p_5$ to
-		itself, and that goes through the \emph{else} branch of each 
-		\emph{if-then-else}, is $-10 \leq \lstinline|x_old| \leq 10$. This path
-		is then added to our subgraph. Moreover, there is no other path whose
-		image is not in $-10 \leq \lstinline|x_old| \leq 10$.
-		Additionally, the path that goes from $p_5$ to
-		$p_{10}$, that goes through the \emph{else} branch of each 
-		\emph{if-then-else} at lines 7,8 and 9, is also 
-		$-10 \leq \lstinline|x_old| \leq 10$, and thus is also added into the
-		new subgraph. Again, there is no new path that goes to $p_{10}$ with an
-		image outside $-10 \leq \lstinline|x_old| \leq 10$.
-	\item
-		We apply Path Focusing on this new subgraph (\ref{fig:example-graph}
-		Step 2). When we have reached an inductive invariant, we narrow, and the
-		result for both $p_5$ and $p_{10}$ is the polyhedron 
-		$-10000 \leq \lstinline|x_old| \leq 10000$.
-	\item Finally, we compute the new subgraph. The SMT-solver does not find any
-		new path that makes the abstract values grow, and the algorithm
-		terminates.
-\end{enumerate}
-
-Our technique gives us the expected invariant 
-$\lstinline|x_old| \in~[-10000,\allowbreak 10000]$. 
-Here, only 3 paths out of the 8 have been computed during the analysis. In
-practice, depending on the order the SMT-solver returns the paths, other
-feasible paths could have been added during the analysis.
-
-\begin{figure}
-\label{fig:example-graph}
-\centering
-\begin{minipage}[c]{.15\textwidth}
-\begin{tikzpicture}[->,>=stealth',auto,node distance=2.1cm,
-                    semithick,font=\footnotesize]
-
-	\node[PRstate] (n00) {$p_4$};
-	\node[PRstate] (n0) [below of=n00, yshift=0.8cm] {$p_5$};
-	\node[PRstate] (n1) [below of=n0] {$p_{10}$};
-	\node (label) [below of=n1, yshift=1.2cm] {Step 1};
-
-  \path [transition] 
-		(n00) edge  node {$\lstinline|x_old| \gets 0$} (n0);
-\end{tikzpicture}
-\end{minipage} 
-\begin{minipage}[c]{.28\textwidth}
-\begin{tikzpicture}[->,>=stealth',auto,node distance=2.1cm,
-                    semithick,font=\footnotesize]
-
-	\node[PRstate] (n00) {$p_4$};
-	\node[PRstate] (n0) [below of=n00, yshift=0.8cm] {$p_5$};
-	\node[PRstate] (n1) [below of=n0] {$p_{10}$};
-	\node (label) [below of=n1, yshift=1.2cm] {Step 2};
-
-  \path [transition] 
-		(n00) edge  node {$\lstinline|x_old| \gets 0$} (n0);
-  \path [transition] 
-		(n0) edge  [loop left] node [left, xshift=0.2cm] {
-		$\begin{array}{r}
-			-10000 \leq \lstinline|x| \leq 10000 \\
-			\lstinline|x_old|-10 \leq \lstinline|x| \leq \lstinline|x_old|+10 /\\
-			\lstinline|x_old| \gets \lstinline|x|
-		\end{array}$
-		} (n0);
-  \path [transition] 
-		(n0) edge node [left, yshift=-0.2cm, xshift=0.2cm] {
-		$\begin{array}{r}
-			-10000 \leq \lstinline|x| \leq 10000 \\
-			\lstinline|x_old|-10 \leq \lstinline|x| \leq \lstinline|x_old|+10 /\\
-			\lstinline|x_old| \gets \lstinline|x|
-		\end{array}$
-		} (n1);
-\end{tikzpicture}
-\end{minipage}
-\caption{Ascending sequence of subgraphs}
-\end{figure}
-
-In this example, we see that our technique actually combines best of
-\emph{Guided Static Analysis} and \emph{Path Focusing}. Still, we need to unroll
-the loop once in order to get precise results. Section \ref{sec:disjunctive}
-proposes a new technique that gives precise invariants without needing such
-graph transformations.
-
-\section{Disjunctive invariants}
-\label{sec:disjunctive}
-
-\citet{DBLP:conf/pldi/GulwaniZ10} proposed a technique for computing disjunctive invariants, by
-distinguishing all the paths inside a loop. In
-this section, we propose to improve this technique by using SMT queries to find
-interesting paths, the objective being to avoid an explicit exhaustive
-enumeration of an exponential number of paths.
-
-For each control point $p_i$, we compute a disjunctive invariant
-$\bigvee_{1\leq j \leq m_i} X_{i,j}$. We denote by $n_i$ the number of
-distinct paths starting at $p_i$.
-To perform the analysis, one chooses an integer $\delta_i \in [1,m_i]$, and
-a mapping function $\sigma_i: [1,m_i] \times [1,n_i] \mapsto [1,m_i]$.
-  The $k$-th path starting fom $p_i$ is denoted $\tau_{i,k}$.
-  The image of the $j$-th disjunct $X_{i,j}$ by the path $\tau_{i,k}$ is then
-  joined with $X_{i,\sigma(j,k)}$.
-Initially, the $\delta_i$-th abstract value contains the initial states of
-$p_i$, and all other abstract values contain~$\emptyset$.
-
-For each control point $p_i \in P_R$, $m_i$, $\delta_i$ and $\sigma_i$ can be defined heuristically.
-For instance, one could define $\sigma_i$ so that $\sigma_i(j,k)$ only depends on the
-last transition of the path, or else construct it dynamically during the
-analysis.
-
-Our method improves this technique in two ways :
-\begin{itemize}
-\item Instead of enumerating the whole set of paths, we keep them implicit and
-compute them only when needed.
-
-\item At each loop iteration of the original algorithm \citep{DBLP:conf/pldi/GulwaniZ10}, an image by each path inside the loop is computed for each disjunct of the invariant candidate.
-Yet, many of these images may be redundant: for instance, if our invariant candidate is $(0 \leq x \leq 10 \land 0 \leq y \leq 1000) \lor (x < -10 \land y < -10)$, then there is no point enumerating paths whose image is included in this invariant candidate.
-In our approach, we compute such an image only if it makes the resulting abstract value grow.
-\end{itemize}
-
-Our improvement consists in a modification of the SMT formula we solve in
-\ref{sec:guided_multigraph}.
-We introduce in this formula Boolean variables $\{d_j, 1 \leq j \leq m\}$, so
-that we can easily find in the model which abstract value of the
-disjunction of the source point has
-to be chosen to make the invariant of the destination grow.
-The resulting formula that is given to the SMT solver is defined
-by $g(p_i)$.
-When the formula is satisfiable, we know that the index $j$ of the starting
-disjunct that has to be chosen is the one for which the associate Boolean value
-$d_j$ is \emph{true} in the model. Then, we can easily compute the value of 
-$\sigma_i(j,k)$, thus know the index of the disjunct to join with.
-
-\begin{eqnarray*}
-g(p_i) & = & \rho \wedge b_i^s \wedge 
-\displaystyle\bigwedge_{j \in P_R \atop j \neq i} \neg b_j^s  \\
- & & \wedge 
-\displaystyle\bigvee_{1 \leq k \leq m_i} (d_k \wedge X_{i,k} \wedge \bigwedge_{l \neq k}
-\neg d_l) \\
- & & \wedge
-\displaystyle\bigvee_{j \in Succ(i)} 
-(b_j^d \wedge \bigwedge_{1 \leq k \leq m_i} (\neg X_{j,k}))
-\end{eqnarray*}
-
-In our algorithm, the initialization of the abstract values slightly differs from
-algorithm \ref{algo:combined} line~\ref{alg=X-init}, since we now have to
-initialize each disjunct. Line~\ref{alg=X-init} is then replaced by:
-\begin{algorithm}[!h]
-\begin{algorithmic}[1] 
-\FORALL {$k \in \{1,..,m_i\} \setminus \{\delta_i\}$}
-	\STATE $X_{i,k} \gets \perp$
-\ENDFOR
-\STATE $X_{i,\delta_i} \gets I_{p_i}$
-\end{algorithmic}
-\end{algorithm}
-
-Furthermore, the Path-focused algorithm (line~\ref{alg=pf} from algorithm
-\ref{algo:combined}) is enhanced to deal with disjunctive invariants, and
-is detailed in algorithm \ref{algo:disjunctive}.
-
-The \emph{Update} function can classically assign to $X_{i,\sigma_i(j,k)}$ the
-value $X_{i,\sigma_i(j,k)} \widening (X_{i,\sigma_i(j,k)} \sqcup
-\tau_{i,k}(X_{i,j}))$, or can integrate the special treatment for self loops
-proposed by \citet{Monniaux_Gonnord_SAS11}, with widening/narrowing sequence or
-acceleration.
-
-\begin{algorithm}[!h]
-\caption{Disjunctive invariant computation with implicit paths}\label{gulwani2}
-\label{algo:disjunctive}
-\begin{algorithmic}[1] 
-\WHILE {true}
-	\STATE $res \gets SmtSolve\left[g(p_i)\right]$
-	\IF {$res = unsat$}
-		\STATE \textbf{break}
-	\ENDIF
-	\STATE Compute the path $\tau_{i,k}$ from $res$ 
-	\STATE Take $j \in \{ l | d_l = true\}$ 
-	\STATE Update($X_{i,\sigma_i(j,k)})$
-\ENDWHILE
-\end{algorithmic}
-\end{algorithm}
-
-We experimented with a heuristic
-of dynamic construction of the $\sigma_i$ functions, adapted from
-\citet{DBLP:conf/pldi/GulwaniZ10}. 
-%\DM{Cette heuristique, je croyais que c'était nous qui l'avions trouvée,
-%et qu'elle n'était pas dans Gulwani?}
-%\JH{Non. Gulwani fait la même chose en un peu plus complique (voir commentaire
-%suivant)}
-%\MM{Du coup, la contribution~\ref{contr:disjunctive} devrait être
-%  affaiblie un peu je suppose. Gulwani fait aussi du SMT-solving ?}
-%\DM{Non, Gulwani énumère tous les chemins un par un à tous les coups.}
-%\MM{OK, du coup j'ai ajouté une phrase à la fin de la section pour
-%  rappeler notre contrib}
-For each control point $p_i \in P_R$, we start with one single disjunct
-($m_i = 1$) and define $\delta_i = 1$.
-$M$ denotes an upper bound on the number of disjuncts per control point.
-
-The $\sigma_i$ functions take as parameters the index of the starting abstract
-value, and the path we focus on. 
-Since we dynamically construct these functions
-during the analysis, we store their already computed image into a
-compact representation, such as Algebraic Decision Diagrams.
-$\sigma_i(j,k)$ is then constructed on the fly only when needed, and computed
-only once.
-When the value of $\sigma_i(j,k)$ is required but undefined, we first compute
-the image of the
-abstract value $X_{i,j}$ by the path indexed by $k$, and try to find an existing
-disjunct of index $j'$ so that the least upper bound of the two abstract values is exactly their union (this can be tested using SMT-solving).
-If such an index exists, then we set $\sigma_i(j,k) = j'$.
-Otherwise:
-\begin{itemize}[itemsep=-1\parsep,topsep=0pt]
-	\item if $m_i < M$, we increase $m_i$ by $1$ and define $\sigma_i(j,k) = m_i$
-	\item if $m_i = M$, we define $\sigma_i(j,k) = M$ 
-		%\JH{A la place, Gulwani cherche l'invariant qui a le plus de contraintes
-		%d'egalite en commun pour definir son $sigma$. Nous, on le met dans M, pour simplifier}
-\end{itemize}
-The main difference with the original
-algorithm~\cite{DBLP:conf/pldi/GulwaniZ10} is that we construct
-$\sigma_i(j,k)$ using SMT queries instead of enumerating a possibly
-exponential number of paths to find a solution.
-
-\section{Experimental comparisons}
-\label{sec:experiments}
+\section{Implementation}
 
@@
 
-Our implementation of path-focusing currently does not use true acceleration techniques, as proposed by \citet{Monniaux_Gonnord_SAS11}.
-Instead, it simply runs widening and narrowing iterations on a single path.
+Our implementation of path-focusing currently does not use true acceleration
+techniques, as proposed by \citet{Monniaux_Gonnord_SAS11}. Instead, it simply runs widening and narrowing iterations on a single path.
 The analysis is currently only forward, even though nothing in what we describe in this article is specific to forward analysis.
+
+\section{Experiments}
+\label{sec:experiments}
 
@@
   \begin{center}
-    \input{./techniques.tex}
+    %\input{./techniques.tex}
   \end{center} 
@@
  \hline
- \input{table1}
+ %\input{table1}
 \end{tabular}
@@
 	\multicolumn{1}{c|}{\textbf{DIS}} \\ \hline
-	\input{table_time.tex} \hline
+	%\input{table_time.tex} \hline
 \end{tabular}
@@
  \hline
- \input{table_domain.tex}
+ %\input{table_domain.tex}
 \end{tabular}
@@
 \phantomsection\addcontentsline{toc}{section}{References} 
-\bibliographystyle{dmabbrvnat}
-\bibliography{implicitization}
+\bibliographystyle{plain}
+\bibliography{pagai}
 \end{document}

### 2012-05-13 papers/TAPAS2012/pagai.tex
tapas

@@
 \usepackage{algorithmic,algorithm}
-%\usepackage{graphicx}
-%\usepackage{color}
+\usepackage{graphicx}
+\usepackage{color}
 
@@
 \section{Introduction}\label{intro}
-Introduction
+
+This paper introduces PAGAI, a new tool for static analysis, that
+computes numerical invariants at the different control points of a program.
+This tool uses the LLVM compiler infrastructure.
+
+Several other verification tools already use the LLVM framework, such as
+Calysto, KLEE, LAV, LLBMC.
+
+Our tool is dedicated to experiment various combination of analysis techniques
+by abstract interpretation.
+Indeed, depending on the iteration strategy, the choice of the abstract domain,
+etc, precision and cost of the analysis hugely vary. Our platform allows to
+compare various state-of-the-art techniques and abstract domains on real life
+programs, such as GNU projects.
 
@@
 For SMT-solving, our analyzer uses Yices
-\cite{DBLP:conf/cav/DutertreM06} or Microsoft Z3\cite{DBLP:conf/tacas/MouraB08}.
+\cite{DBLP:conf/cav/DutertreM06} or Microsoft Z3\cite{DBLP:conf/tacas/MouraB08}
+through their C API, and can also operate with any SMT-solver supporting the
+SMT-Lib 2 standard \cite{BarST-SMTLIB} through a pipe.
 
@@
   \begin{center}
-    %\input{./techniques.tex}
+    \input{./gnuplot/techniques.tex}
   \end{center} 
@@
 \begin{table*}
+	\tiny
 \begin{center}
@@
  \hline
- %\input{table1}
+ \input{table1}
 \end{tabular}
@@
 \begin{table}[!h]
-	\centering
+	\centering \tiny
 \begin{tabular}{|l|r|r|r|r|r|} \hline
@@
 	\multicolumn{1}{c|}{\textbf{DIS}} \\ \hline
-	%\input{table_time.tex} \hline
+	\input{table_time.tex} \hline
 \end{tabular}
@@
  \hline
- %\input{table_domain.tex}
+ \input{table_domain.tex}
 \end{tabular}

### 2012-05-13 papers/TAPAS2012/pagai.tex
problème de natbib corrigé
tapas

@@
 \usepackage{mathcomp,amsfonts,amssymb,MnSymbol}
-\usepackage{natbib}
+\usepackage[authoryear]{natbib}
 \usepackage[mathscr]{euscript}
@@
   \author{Julien Henry \thanksref{imag.fr}}
-  \address{VERIMAG\\ Universit\'e Joseph Fourier\\
+  \address{Universit\'e Joseph Fourier, VERIMAG\\
     Grenoble, France} 
 	\author{David Monniaux\thanksref{imag.fr}}
-  \address{VERIMAG\\CNRS\\
+  \address{CNRS, VERIMAG\\
     Grenoble, France} 
 	\author{Matthieu Moy\thanksref{grenoble-inp.fr}}
-  \address{VERIMAG\\Grenoble-INP\\
+  \address{Grenoble-INP, VERIMAG\\
     Grenoble, France}
@@
 \begin{keyword}
-  Please list keywords from your paper here, separated by commas.
+	Static Analysis, Program Verification, Abstract Interpretation, Decision
+	Procedure, Satisfiability Modulo Theories.
 \end{keyword}
@@
 computes numerical invariants at the different control points of a program.
-This tool uses the LLVM compiler infrastructure.
-
+This tool uses the internal representation of LLVM\citep{LLVM_langref,Lattner:2004:LCF:977395.977673},
+which is a target for several industrial-strengh compilers, most notably Clang (supporting C and
+C++) and llvm-gcc (supporting C, C++, Fortran and Ada).
 Several other verification tools already use the LLVM framework, such as
@@
 
-Our tool is dedicated to experiment various combination of analysis techniques
+Our tool is dedicated to experiment with various combination of analysis techniques
 by abstract interpretation.
@@
 programs, such as GNU projects.
+
+\section{Motivating Example}
+
+These techniques make a precise analysis of the paths inside the program
 
@@
 
-%\MM{On peut peut-être citer le nom de l'outil (on garde PAGAI ?) ici.
-%  L'intérêt d'avoir un nom, c'est que les gens peuvent en parler plus
-%  facilement, et c'est plus facile de retrouver le papier origine d'un
-%  outil quand il cite son nom}
-%\DM{Dans la version finale, oui, mais pas dans une soumission anonymisée... suffit de Googlifier pagai sur imag.fr}
-%
-Our tool operates over LLVM bitcode \citep{LLVM_langref,Lattner:2004:LCF:977395.977673}, which is a target for several compilers, most notably Clang (supporting C and C++) and llvm-gcc (supporting C, C++, Fortran and Ada).
 	Abstract domains are provided by the APRON library \citep{DBLP:conf/cav/JeannetM09}, and include convex polyhedra (from the builtin Polka ``PK'' library), octagons, intervals, and linear congruences.
@@
 
-% Cas a probleme:
-% Initialisation (x,y) à {(1,0), (0,1)}
-% puis boucle: si y = 0 alors x++
-% fait un polyèdre qui finit par perdre y <= 1 et x >= 0
-
-%\appendix
-%\section{Appendix Title}
-
-%This is the text of the appendix, if you need one.
-
-%\acks
-
-% Acknowledgments, if needed.
-
-% We recommend abbrvnat bibliography style.
-
 \phantomsection\addcontentsline{toc}{section}{References} 
-\bibliographystyle{plain}
+	\bibliographystyle{plainnat}
 \bibliography{pagai}

### 2012-05-16 papers/TAPAS2012/pagai.tex
TAPAS

@@
 \section{Introduction}\label{intro}
+Static analysis aims at automatically computing properties on programs, such as
+the possible values of their variables during execution. This allows to show for
+instance that a program has no runtime error (such as arithmetic overflow,
+division by zero), to compute loop invariants, etc. The main applications of
+static analysis are compile-time optimisations, and the proof of safety
+properties in critical systems.
+Abstract Interpretation is a general framework for fully automatic static
+analysis, that computes inductive invariants at various control points of an
+imperative program. Linear Relation Analysis (LRA) is a direct application of
+this framework, that computes an upper-approximation of the set of the reachable
+states of a numerical program.
 
@@
 This tool uses the internal representation of LLVM\citep{LLVM_langref,Lattner:2004:LCF:977395.977673},
-which is a target for several industrial-strengh compilers, most notably Clang (supporting C and
-C++) and llvm-gcc (supporting C, C++, Fortran and Ada).
+which is a target for several industrial-strengh compilers, most notably Clang
+(supporting C,
+C++ and Objective-C) and llvm-gcc (supporting C, C++, Fortran and Ada).
 Several other verification tools already use the LLVM framework, such as
@@
 Our tool is dedicated to experiment with various combination of analysis techniques
-by abstract interpretation.
+by Abstract Interpretation.
 Indeed, depending on the iteration strategy, the choice of the abstract domain,
-etc, precision and cost of the analysis hugely vary. Our platform allows to
-compare various state-of-the-art techniques and abstract domains on real life
+etc, precision and cost of the analysis may hugely vary. 
+We experimentally investigate techniques that take benefits of the improvements in SMT solving
+to choose the iteration strategy of the Abstract Interpretation framework.
+
+Our platform allows to compare such state-of-the-art techniques and abstract domains on real life
 programs, such as GNU projects.
@@
 
-We have implemented our proposed solutions inside a prototype of intraprocedural
-static analyzer,
+PAGAI is a prototype of intraprocedural static analyser, that implements our
+recent combined techniques \cite{Henry_Monniaux_Moy_SAS12}
 as well as the classical abstract interpretation algorithm, and the state-of-the-art
@@
 
-	Abstract domains are provided by the APRON library \citep{DBLP:conf/cav/JeannetM09}, and include convex polyhedra (from the builtin Polka ``PK'' library), octagons, intervals, and linear congruences.
+Abstract domains are provided by the APRON library
+\citep{DBLP:conf/cav/JeannetM09}, and include convex polyhedra (from the builtin
+Polka ``PK'' library), octagons, intervals, and linear congruences. It also has
+an interface with the Parma Polyhedra Library (CITATION!!), giving access to
+more abstract domains.
 For SMT-solving, our analyzer uses Yices
@@
 
-We conducted extensive experiments on real-life programs in order to compare the
-different techniques, mostly on open-source projects (Fig.~\ref{fig:projects}) written in C, C++ and Fortran.
-
 \begin{table}[!h]
+	\tiny
 	\centering
@@
 	make & 34 & 993\\ 
-	sed & 23 & 292\\
 	tar & 73 & 1712\\
@@
 
-This motivates a key implement decision of our tool: only those variables $v_1,\dots,v_n$ that are not defined by arithmetic operations are retained as coordinates in the abstract domain (e.g.
-as dimensions in polyhedra).
+This motivates a key implement decision of our tool: only those variables
+$v_1,\dots,v_n$ that are not defined by arithmetic operations are retained as
+coordinates in the abstract domain (e.g. as dimensions in polyhedra), assuming
+they are live at the associate control point.
+
+	For instance, assume that $x,y,z$ are numerical variables of a program,
+	$x$ is defined as $x = y+z$, and $x,y,z$ are live at point $p$. Instead of having
+	$x$ as a dimension for the abstract value at point $p$, we only have $y$ and $z$. All the properties
+	for $x$ can be directly extracted from the abstract value attached to $p$ and the information $x=y+z$.
+	This is an optimisation in the sense that there is redundant information in
+	the abstract value if both $x,y$ and $z$ are dimensions of $X_p$.
+
+	Then, classical definition of liveness can be adapted in our case:
+
+	\begin{definition}[Liveness by linearity]
+	A variable $v$ is \emph{live by linearity} at a control point $p$ if and
+	only if:
+	\begin{itemize}
+	\item 
+	its value is available at $p$, meaning that there is a path from the
+	point defining the variable to $p$,
+	\item one of these conditions holds:
+		\begin{itemize}
+		\item $v$ is live in $p$.
+		\item There is a variable $v'$, defined as a linear combination of other
+		variables $v_1, v_2, \dots, v_n$, so that $\exists i \in \{1,\dots,n\}, v = v_i$,
+		and $v'$ is live by linearity in $p$.
+		\end{itemize}
+	\end{itemize}
+	\end{definition}
+
+	Finally, a variable is a dimension in the abstract domain if and only if it
+	is live by linearity and it is not defined as a linear combination of
+	program variables.
+
 A basic block of code therefore amounts to a \emph{parallel assignment} operation
@@
 
-We further reduce the number of variables by projecting out those that are not live at the program point. Again, SSA use-def and def-use chains make it easy to compute liveness.
+\subsection{Limitations of the tool}
 
@@
 techniques, as proposed by \citet{Monniaux_Gonnord_SAS11}. Instead, it simply runs widening and narrowing iterations on a single path.
-The analysis is currently only forward, even though nothing in what we describe in this article is specific to forward analysis.
+The analysis is currently only forward, even though nothing in the techniques
+implemented is specific to forward analysis.
 
@@
 \label{sec:experiments}
+
+We conducted extensive experiments on real-life programs in order to compare the
+different techniques, mostly on open-source projects (Fig.~\ref{fig:projects}) written in C, C++ and Fortran.
 

### 2012-05-16 papers/TAPAS2012/pagai.tex
TAPAS

@@
 \begin{frontmatter}
-  \title{PAGAI} 
+  \title{PAGAI : a path sensitive static analyser} 
   \author{Julien Henry \thanksref{imag.fr}}
@@
 	PAGAI implements various state-of-the-art techniques combining Abstract
-	Interpretation and Decision Procedures (SMT-solving), and offers
-	several abstract domains.
+	Interpretation and Decision Procedures (SMT-solving), that allow to do a
+	precise analysis of the various paths in a program.
 	We thus can compare in time and precision the various combinations of
@@
 
+Exemple du boustrophedon ???
 These techniques make a precise analysis of the paths inside the program

### 2012-05-18 papers/TAPAS2012/pagai.tex
citation smtlib2

--- 2012-05-13 papers/TAPAS2012/pagai.tex
@@
 \usepackage{mathcomp,amsfonts,amssymb,MnSymbol}
-\usepackage[authoryear]{natbib}
+\usepackage{natbib}
 \usepackage[mathscr]{euscript}
@@
   \author{Julien Henry \thanksref{imag.fr}}
-  \address{Universit\'e Joseph Fourier, VERIMAG\\
+  \address{VERIMAG\\ Universit\'e Joseph Fourier\\
     Grenoble, France} 
 	\author{David Monniaux\thanksref{imag.fr}}
-  \address{CNRS, VERIMAG\\
+  \address{VERIMAG\\CNRS\\
     Grenoble, France} 
 	\author{Matthieu Moy\thanksref{grenoble-inp.fr}}
-  \address{Grenoble-INP, VERIMAG\\
+  \address{VERIMAG\\Grenoble-INP\\
     Grenoble, France}
@@
 \begin{keyword}
-	Static Analysis, Program Verification, Abstract Interpretation, Decision
-	Procedure, Satisfiability Modulo Theories.
+  Please list keywords from your paper here, separated by commas.
 \end{keyword}
@@
 \section{Introduction}\label{intro}
-
-This paper introduces PAGAI, a new tool for static analysis, that
-computes numerical invariants at the different control points of a program.
-This tool uses the internal representation of LLVM\citep{LLVM_langref,Lattner:2004:LCF:977395.977673},
-which is a target for several industrial-strengh compilers, most notably Clang (supporting C and
-C++) and llvm-gcc (supporting C, C++, Fortran and Ada).
-Several other verification tools already use the LLVM framework, such as
-Calysto, KLEE, LAV, LLBMC.
-
-Our tool is dedicated to experiment with various combination of analysis techniques
-by abstract interpretation.
-Indeed, depending on the iteration strategy, the choice of the abstract domain,
-etc, precision and cost of the analysis hugely vary. Our platform allows to
-compare various state-of-the-art techniques and abstract domains on real life
-programs, such as GNU projects.
-
-\section{Motivating Example}
-
-These techniques make a precise analysis of the paths inside the program
+Introduction
 
@@
 
+%\MM{On peut peut-être citer le nom de l'outil (on garde PAGAI ?) ici.
+%  L'intérêt d'avoir un nom, c'est que les gens peuvent en parler plus
+%  facilement, et c'est plus facile de retrouver le papier origine d'un
+%  outil quand il cite son nom}
+%\DM{Dans la version finale, oui, mais pas dans une soumission anonymisée... suffit de Googlifier pagai sur imag.fr}
+%
+Our tool operates over LLVM bitcode \citep{LLVM_langref,Lattner:2004:LCF:977395.977673}, which is a target for several compilers, most notably Clang (supporting C and C++) and llvm-gcc (supporting C, C++, Fortran and Ada).
 Abstract domains are provided by the APRON library \citep{DBLP:conf/cav/JeannetM09}, and include convex polyhedra (from the builtin Polka ``PK'' library), octagons, intervals, and linear congruences.
@@
 through their C API, and can also operate with any SMT-solver supporting the
-SMT-Lib 2 standard \cite{BarST-SMTLIB} through a pipe.
+SMT-Lib 2 standard \cite{BarST-SMTLIB,BarST-SMT-10} through a pipe.
 
@@
 
+% Cas a probleme:
+% Initialisation (x,y) à {(1,0), (0,1)}
+% puis boucle: si y = 0 alors x++
+% fait un polyèdre qui finit par perdre y <= 1 et x >= 0
+
+%\appendix
+%\section{Appendix Title}
+
+%This is the text of the appendix, if you need one.
+
+%\acks
+
+% Acknowledgments, if needed.
+
+% We recommend abbrvnat bibliography style.
+
 \phantomsection\addcontentsline{toc}{section}{References} 
-	\bibliographystyle{plainnat}
+\bibliographystyle{plain}
 \bibliography{pagai}

### 2012-05-18 papers/TAPAS2012/pagai.tex
TAPAS

@@
 \usepackage{color}
+\usepackage{tikz}
 
@@
 \def\newblock{\hskip .11em plus .33em minus .07em}
+
+\tikzstyle{arrow}=[->,line width=.05cm,draw=red!90!blue!60!black]
+
+\usetikzlibrary{snakes,arrows,shapes,backgrounds,shadows,automata,patterns}
+%\usepgflibrary{snakes}
+
+\tikzstyle{state}=[circle,fill=black!25,minimum size=13pt,inner sep=0pt]
+\tikzstyle{rstate}=[rectangle,fill=black!25,minimum size=13pt,inner sep=0pt]
+\tikzstyle{transition}=[rectangle,semithick,draw=black!75,
+  			  minimum size=4mm]
+\tikzstyle{transition2}=[transition,rectangle,thick,dashed,
+  			  minimum size=4mm]
+\tikzstyle{PRstate}=[circle,double,draw,fill=blue!15,minimum size=13pt,inner sep=0pt]
+\tikzstyle{polyhedra}=[blue!25,opacity=0.5,pattern=north west lines,pattern
+color=blue]
+\tikzstyle{line}=[black,thick]
+
+\lstnewenvironment{C}
+{\lstset{language=C,
+		basicstyle=\ttfamily,
+		commentstyle=\textit,
+		showstringspaces=false}}
+{}
 
@@
 \newcommand{\Real}{{\mathbb R}}
-\def\lastname{Henry}
+\def\lastname{Henry, Monniaux \& Moy}
 \begin{document}
@@
 
-Exemple du boustrophedon ???
-These techniques make a precise analysis of the paths inside the program
+It is well know fact in abstract interpretation that the least upper bound of two or
+more abstract values in the abstract domain 
+may induce unrecoverable loss of precision. This operation typically occurs with
+if-then-else statements and loops.
+
+An example of program where such a loss of precision occurs is depicted in
+Figure \ref{example}. In this program, the loop body has two feasible paths that
+are executed alternatively, depending on a variable ``phase''. Such kind of
+programs often occur in reactive systems. 
+Classical abstract interpretation does not prove that $\lstinline|x| < 100$,
+because of the abstract union at control point $n_5$. 
+To cope with this problem, an idea is compute disjunctive invariants, i.e a
+union of values of the abstract domain. However, the number of terms in the
+disjunction may grow exponentially with the number of tests in the program.
+A second technique is to distinguish all the paths inside the loop, so that we
+compute least upper bounds only at the widening points. In this example, if we
+distinguish all the 9 paths of the loop, we prove $\lstinline|x| < 100$. Again,
+this path enumeration grows exponentially with the number of tests. In order to
+avoid this blowup, we use SMT-solving to keep these paths implicit and to
+compute them only if needed.
+
+\begin{figure}[!h]
+\begin{minipage}[c]{.49\linewidth}
+\begin{C}
+int x = 0;
+int t = 0;
+int phase = 0;
+
+while (t < 100) {
+   if (phase == 0) 
+      x = x+2;
+   if (phase == 1) 
+      x = x-1;
+   phase = 1-phase;
+   t++;
+}
+assert(x <= 100);
+\end{C}
+\end{minipage}
+\begin{minipage}[c]{.49\linewidth}
+\begin{tikzpicture}[->,>=stealth',auto,node distance=1.5cm,
+                    semithick,font=\footnotesize]
+
+	\node[PRstate] (n0) {$n$};
+	\node[PRstate] (n00) [below of=n0] {$n_0$};
+	\node[PRstate] (nf) [left of=n00] {$n_f$};
+	\node[state] (n1) [below of=n00] {$n_1$};
+	\node[state] (n3) [below of=n1] {$n_3$};
+	\node[state] (n2) [left of=n3] {$n_2$};
+	\node[state] (n4) [right of=n3] {$n_4$};
+	\node[state] (n5) [below of=n3] {$n_5$};
+	\node[state] (n7) [below of=n5] {$n_7$};
+	\node[state] (n6) [left of=n7] {$n_6$};
+	\node[state] (n8) [right of=n7] {$n_8$};
+	\node[state] (n9) [below of=n7] {$n_9$};
+
+  \path [transition] 
+		(n0) edge              node  {$\begin{array}{l}
+		x \leftarrow 0 \\
+		phase \leftarrow 1 
+		\end{array}$} (n00);
+  \path [transition] 
+        (n00)  edge              node {$t>=100$} (nf);
+  \path [transition] 
+        (n00)  edge              node {$t<100$} (n1);
+  \path [transition] 
+        (n1) edge			   node [left] {$phase,0?$~ ~ ~ $<$} (n2);
+  \path [transition] 
+        (n1)  edge              node {$>$} (n3);
+  \path [transition] 
+        (n1)  edge              node [right] {$=$} (n4);
+  \path [transition] 
+        (n3) edge              node  {} (n5);
+  \path [transition] 
+        (n2) edge			   node {} (n5);
+  \path [transition] 
+        (n4) edge			   node [right] {$x \leftarrow x+2$} (n5);
+  \path [transition] 
+        (n5) edge			   node [left] {$phase,1?$ ~ ~ ~$<$} (n6);
+  \path [transition] 
+        (n5) edge			   node {$>$} (n7);
+  \path [transition] 
+        (n5) edge			   node [right] {$=$} (n8);
+  \path [transition] 
+        (n6) edge              node {} (n9);
+  \path [transition] 
+        (n7) edge              node {} (n9);
+  \path [transition] 
+        (n8) edge              node [right] {$x \leftarrow x-1$} (n9);
+  \path [transition] 
+        (n9) edge [out=0, in=0, distance=4.3cm] node [right] {
+		$\begin{array}{l}
+			phase \leftarrow \\
+			1-phase \\
+			t \leftarrow \\
+			t+1
+		\end{array}$} (n00);
+
+\end{tikzpicture}
+\end{minipage}
+\caption{Example of program, where the loop behaviour vary depending on a variable $phase$.}
+\label{example}
+\end{figure}
 
@@
 	A variable $v$ is \emph{live by linearity} at a control point $p$ if and
-	only if:
-	\begin{itemize}
-	\item 
-	its value is available at $p$, meaning that there is a path from the
-	point defining the variable to $p$,
-	\item one of these conditions holds:
+	only if one of these conditions holds:
 		\begin{itemize}
@@
 		and $v'$ is live by linearity in $p$.
-		\end{itemize}
 		\end{itemize}
@@
 
+\subsection{Use}
+
+PAGAI takes as input an LLVM bitcode file, and outputs an inductive invariant
+for each control point in $P_R$ (typically, the widening points).
+When a program contains \emph{assert} function call, PAGAI also outputs whether
+the statement has been proved.
+It is also possible to add some preconditions about the variables, etc, using a
+function \emph{assume}.
 \subsection{Limitations of the tool}
@@
 
-\section{Conclusion and future prospects}
-Roughly, an analysis by abstract interpretation is defined by the choice of an iteration strategy and an abstract domain.
-In this article, we demonstrated that changes in the iteration algorithm can significantly improve precision, sometimes while improving analysis times.
-
-A common criticism of analysis techniques based on SMT-solving is that they
-do not scale up. Yet, our experiments show that, for numerical properties,
-they scale up to the size of typical functions and loops.
-It is however
-quite certain that, naively applied, they cannot scale to the kind of
-programs targeted by e.g. the Astr\'ee tool, that is, a dozens or hundreds
-of thousands of lines of code in a single loop operating over similar numbers
-of remanent variables.
-Actually, for such applications, only (quasi-)linear
-algorithms scale up, and ``cheap'' abstract domains such as octagons ($O(n^3)$ where $n$ is the number of variables) are not applied to the full variable set, but to restricted subsets thereof.
-It thus seems reasonable that techniques such as considering ``packs'' of related variables, slicing, etc. may similarly help SMT-based techniques to scale to global analyses.
-
-We compared the precision of different techniques and abstract domains by comparing the invariants for the inclusion ordering.
-A better metric is perhaps to take a client analysis --- such as the detection of overflows and array bound violations --- and compare the rates of alarms.
-
-We focused on numerical properties, because they are supported by easily available abstract libraries.
-Yet, in most programs, properties of data structures are important for proving interesting properties.
-Further investigations are needed not only on good abstractions for pointers (many are already known) but also on their conversion to SMT problems.
+\section{Conclusion and future work}
 

### 2012-05-18 papers/TAPAS2012/pagai.tex
typo

@@
 
-It is well know fact in abstract interpretation that the least upper bound of two or
+It is well know in abstract interpretation that the least upper bound of two or
 more abstract values in the abstract domain 

### 2012-05-19 papers/TAPAS2012/pagai.tex
intro

@@
 \usepackage{amsmath}
-\usepackage{mathcomp,amsfonts,amssymb,MnSymbol}
+\usepackage{mathcomp,amsfonts,amssymb}
 \usepackage[authoryear]{natbib}
@@
 \begin{abstract} 
-	We present design and implementation of PAGAI, a new static analyzer 
-	working over the LLVM compiler infrastructure, that computes inductive
-	invariants on the numerical variables of the program.
-	PAGAI implements various state-of-the-art techniques combining Abstract
-	Interpretation and Decision Procedures (SMT-solving), that allow to do a
-	precise analysis of the various paths in a program.
-	We thus can compare in time and precision the various combinations of
-	techniques and abstract domains. We conducted extensive experiments for
-	comparing these techniques on several benchmarks and GNU programs.
+	We describe the design and the implementation of PAGAI,
+        a new static analyzer working over the LLVM compiler infrastructure,
+        which computes inductive invariants on the numerical variables of the
+        analyzed program.
+
+	PAGAI implements various state-of-the-art algorithms combining abstract
+	interpretation and decision procedures (SMT-solving),
+        focusing on distinction of paths inside the control flow graph while
+        avoiding systematic exponential enumerations.
+        It is parametric in the abstract domain in use, the algorithm,
+        and the decision procedure.
+
+	We compared the time and precision of various combinations of
+	analysis algorithms and abstract domains, with extensive
+        experiments both on personal benchmarks and widely available
+        GNU programs.
 \end{abstract}
@@
 \section{Introduction}\label{intro}
-Static analysis aims at automatically computing properties on programs, such as
-the possible values of their variables during execution. This allows to show for
-instance that a program has no runtime error (such as arithmetic overflow,
-division by zero), to compute loop invariants, etc. The main applications of
-static analysis are compile-time optimisations, and the proof of safety
-properties in critical systems.
+Sound static analysis automatically computes properties on programs, such as the possible values of their variables during execution.
+Applications include:
+showing that a program has no runtime error (such as arithmetic overflow, division by zero, array access out of bounds), as in e.g. the Astrée analyzer \cite{ASTREE_ESOP05};
+computing invariants for use with assisted proof systems (such as the B method, Frama-C\footnote{http://www.frama-c.com/}), thereby lessening the burden on the user;
+computing invariants for advanced optimization techniques in compilation (e.g. showing that two array cells are distinct, in order to allow instruction reordering between assignments to these cells).
+
 Abstract Interpretation is a general framework for fully automatic static

### 2012-05-19 papers/TAPAS2012/pagai.tex
intro, suite

@@
 
-Abstract Interpretation is a general framework for fully automatic static
-analysis, that computes inductive invariants at various control points of an
-imperative program. Linear Relation Analysis (LRA) is a direct application of
-this framework, that computes an upper-approximation of the set of the reachable
-states of a numerical program.
+This article introduces PAGAI, a new tool for fully automatic static analysis.
+PAGAI takes as input a program in the ``bitcode'' intermediate representation of LLVM\citep{LLVM_langref,Lattner:2004:LCF:977395.977673}, a modern compilation framework.
+LLVM bitcode is a target for several industrial-strengh compilers, most notably Clang (supporting C, C++, Objective-C and Objective-C++) and llvm-gcc (supporting C, C++, Objective-C, Objective-C++, Fortran and Ada);
+furthermore, a growing number of analysis tools, testing tools, etc. are currently built around this platform (Calysto, KLEE, LAV, LLBMC).
+The output of PAGAI is a list of inductive invariants for a selected subset of the control nodes of the original program:
+for structured source programs, PAGAI will provide an inductive invariant for the head of every loop.
 
-This paper introduces PAGAI, a new tool for static analysis, that
-computes numerical invariants at the different control points of a program.
-This tool uses the internal representation of LLVM\citep{LLVM_langref,Lattner:2004:LCF:977395.977673},
-which is a target for several industrial-strengh compilers, most notably Clang
-(supporting C,
-C++ and Objective-C) and llvm-gcc (supporting C, C++, Fortran and Ada).
-Several other verification tools already use the LLVM framework, such as
-Calysto, KLEE, LAV, LLBMC.
+PAGAI is based on abstract interpretation, a general framework for fully automatic static analysis.
+PAGAI infers invariants of a selected form; by default it performs \emph{linear relation analysis}, which obtains invariants as conjunctions of linear inequalities (or, equivalently, convex polyhedra), but it also supports most of the abstract domains provided by the APRON library (products of intervals, octagons, linear congruences, and combinations thereof) through a runtime option.
+% TODO: citer des réfs sur APRON, les octogones etc.
+
+Abstract interpretation is a , which computes inductive invariants at various control points of an
+imperative program. Linear relation analysis (LRA) is an application of
+this framework, where the invariants are conjunctions of linear inequalities (or, equivalently, convex polyhedra), or, possibly, disjunctions of such conjunctions (or, equivalently, unions of convex polyhedra).
 

### 2012-05-19 papers/TAPAS2012/pagai.tex
intro

@@
 PAGAI infers invariants of a selected form; by default it performs \emph{linear relation analysis}, which obtains invariants as conjunctions of linear inequalities (or, equivalently, convex polyhedra), but it also supports most of the abstract domains provided by the APRON library (products of intervals, octagons, linear congruences, and combinations thereof) through a runtime option.
+Depending on the algorithm (\emph{iteration strategy}) selected, PAGAI may also infer invariants as \emph{disjunctions} of elements of the abstract domain (e.g. unions of convex polyhedra).
 % TODO: citer des réfs sur APRON, les octogones etc.
 
-Abstract interpretation is a , which computes inductive invariants at various control points of an
-imperative program. Linear relation analysis (LRA) is an application of
-this framework, where the invariants are conjunctions of linear inequalities (or, equivalently, convex polyhedra), or, possibly, disjunctions of such conjunctions (or, equivalently, unions of convex polyhedra).
+Textbook descriptions of abstract interpretation-based static analysis state that an inductive invariant is computed at every control point of the program.
+In contrast, PAGAI abstracts straight-line sequences of statements en bloc, computing invariants only at points where control flow branches or merges.
+Furthermore, several algorithms implemented in PAGAI compute invariants only at the heads of loops (or, in general control-flow graph, at nodes forming a \emph{feedback vertex set}, whose removal breaks all cycles in the graph), expanding the rest of the control flow to a possibly exponential number of straight-line sequences of statements between the selected nodes.
+In order to avoid explicit enumerations of exponential sets, PAGAI uses decision procedures for arithmetic theories, based on the \emph{satisfiability modulo theory} (SMT) approach.
+Depending on the abstract domain in use and the arithmetic constructs found in the program statements, PAGAI may use a variety of SMT-solvers, including Z3 and Yices; work is underway for a generic SMTLIB2 interface allowing the connection of more solvers, including MathSAT, SMTinterpol and CVC4.
+% TODO refs
 
-Our tool is dedicated to experiment with various combination of analysis techniques
-by Abstract Interpretation.
-Indeed, depending on the iteration strategy, the choice of the abstract domain,
-etc, precision and cost of the analysis may hugely vary. 
-We experimentally investigate techniques that take benefits of the improvements in SMT solving
-to choose the iteration strategy of the Abstract Interpretation framework.
-
-Our platform allows to compare such state-of-the-art techniques and abstract domains on real life
-programs, such as GNU projects.
+The PAGAI tool is dedicated to experimenting new analysis algorithms.
+It allows independent selection of abstract domain and iteration strategy, and partially independent selection of decision procedure, and thus is well-suited for comparisons.
+We thus conducted extensive experiments both on examples we produced ourselves (sometimes inspired by industrial code) and on GNU programs.
+The ability to run on any C or C++ code, through the LLVM system, is especially useful for the latter: front-ends for many analysis tools put restrictions (e.g.
+no backward \lstinline|goto| instructions, no pointer arithmetic...), often satisfied by safety-critical embedded programs, but not by generic programs.
 

### 2012-05-19 papers/TAPAS2012/pagai.tex
bla bla

@@
 furthermore, a growing number of analysis tools, testing tools, etc. are currently built around this platform (Calysto, KLEE, LAV, LLBMC).
+
 The output of PAGAI is a list of inductive invariants for a selected subset of the control nodes of the original program:
@@
 \section{Motivating Example}
-
-It is well know in abstract interpretation that the least upper bound of two or
-more abstract values in the abstract domain 
-may induce unrecoverable loss of precision. This operation typically occurs with
-if-then-else statements and loops.
+In most forward abstract interpretation-based analyzes, when control flows from several nodes into a single node, the abstract value at that node is obtained by computing the least upper bound of the incoming abstract values in the abstract domain (in backward analysis, this occurs when control flows from a single node to several nodes).
+If the abstract domain is convex polyhedra, then this means computing the convex hull of the incoming polyhedra.
+Such an operation may induce unrecoverable loss of precision by introducing spurious states that cannot occur in concrete program runs.
 
 An example of program where such a loss of precision occurs is depicted in
-Figure \ref{example}. In this program, the loop body has two feasible paths that
-are executed alternatively, depending on a variable ``phase''. Such kind of
-programs often occur in reactive systems. 
-Classical abstract interpretation does not prove that $\lstinline|x| < 100$,
-because of the abstract union at control point $n_5$. 
-To cope with this problem, an idea is compute disjunctive invariants, i.e a
-union of values of the abstract domain. However, the number of terms in the
-disjunction may grow exponentially with the number of tests in the program.
-A second technique is to distinguish all the paths inside the loop, so that we
-compute least upper bounds only at the widening points. In this example, if we
-distinguish all the 9 paths of the loop, we prove $\lstinline|x| < 100$. Again,
-this path enumeration grows exponentially with the number of tests. In order to
-avoid this blowup, we use SMT-solving to keep these paths implicit and to
-compute them only if needed.
+Figure \ref{example}.
+In this program, the loop body has two feasible paths that
+are executed alternatively, depending on a variable ``phase''.
+Such programs, with active code paths depending on global ``mode'' or ``phase'' variables, often occur in reactive systems.
+
+Removing program point $n_0$ breaks all cycles; we are thus primarily concerned with obtaining an inductive invariant at that point.
+We consider the domain of convex polyhedra and thus wish to obtain this invariant as a polyhedron.
+Because convex polyhedra form a lattice of infinite height, we use Kleene iterations (pushing abstract values through control-flow edges) with a widening scheme, which ensures convergence in finite time to an inductive invariant, followed by decreasing (narrowing) iterations.
+
+At program point $n_5$, classical forward abstract interpretation with convex polyhedra computes the convex hull of three incoming polyhedra over variables $(\lstinline|phase|,\lstinline|x|,\lstinline|t|)$.
+This convex hull introduces extra states, unreachable in the concrete programs, for the analysis of the fragment from $n_5$ to~$n_9$.
+When analyzing the whole loop, these extra states prevent proving $x < 100$.
+
+To cope with this problem, an solution is compute disjunctive invariants at all intermediate nodes: at $n_5$, keep an explicit list of three polyhedra, and thus obtain a list of nine polyhedra at $n_9$.
+We pass the convex hull of these polyhedra to the widening operator at point $n_0$ (which operates on polyhedra, not on lists of polyhedra).
+The drawback is that the number of elements in the lists may grow exponentially with the number of successive tests.
+
+A second solution, equivalent to the preceding with respect to final results but different in its operation, is to distinguish all nine paths inside the loop (from $n_0$ to $n_0$), compute the final outcome of each path, and compute the convex hull of these outcomes.
+Again, the number of such paths is exponential in the number of tests.
+We therefore consider paths only if they will make the final convex hull grow.
+The next path to consider is obtained from a model of an arithmetic formula expressing this condition \cite{Monniaux_Gonnord_SAS11}; if this formula is unsatisfiable, this means there is no such path and thus the convex hull obtained truly is the one for all paths.
+The advantages of this method over the preceding one are twofold: there is no exponentially large list of abstract elements, and the satisfiability query for the formula is handed over to a \emph{satisfiability modulo theory} (SMT) solver.
+Modern SMT-solvers are very efficient; their caching mechanisms may, for instance, remember that taking a certain branch in the code is incompatible with taking another one (if a Boolean is associated with passing through each branch, then this is just a \emph{blocking clause} inside the underlying SAT-solver).
+The algorithms implemented in PAGAI are variants of this second approach.
 

### 2012-05-19 papers/TAPAS2012/pagai.tex
bla bla

@@
 In order to avoid explicit enumerations of exponential sets, PAGAI uses decision procedures for arithmetic theories, based on the \emph{satisfiability modulo theory} (SMT) approach.
-Depending on the abstract domain in use and the arithmetic constructs found in the program statements, PAGAI may use a variety of SMT-solvers, including Z3 and Yices; work is underway for a generic SMTLIB2 interface allowing the connection of more solvers, including MathSAT, SMTinterpol and CVC4.
+
 % TODO refs
@@
 
-PAGAI is a prototype of intraprocedural static analyser, that implements our
+PAGAI is a prototype interprocedural static analyser, that implements our
 recent combined techniques \cite{Henry_Monniaux_Moy_SAS12}
-as well as the classical abstract interpretation algorithm, and the state-of-the-art
-techniques \emph{Path Focusing} \cite{Monniaux_Gonnord_SAS11} and \emph{Guided
-Static Analysis} \cite{DBLP:conf/sas/GopanR07}.
+as well as the classical abstract interpretation algorithm, and the state-of-the-art techniques \emph{Path Focusing} \cite{Monniaux_Gonnord_SAS11} and \emph{Guided Static Analysis} \cite{DBLP:conf/sas/GopanR07}.
 
@@
 \citep{DBLP:conf/cav/JeannetM09}, and include convex polyhedra (from the builtin
-Polka ``PK'' library), octagons, intervals, and linear congruences. It also has
-an interface with the Parma Polyhedra Library (CITATION!!), giving access to
-more abstract domains.
+Polka ``PK'' library), octagons, and products of intervals. It also has
+an interface with the Parma Polyhedra Library (TODO CITATION!!), giving access to
+more abstract domains (for instance, it can use a reduced product of polyhedra and linear congruences, producing invariants such as $0 \leq x \leq 1001 \land x \equiv 0 \pmod 7$).
+
 For SMT-solving, our analyzer uses Yices
 \cite{DBLP:conf/cav/DutertreM06} or Microsoft Z3\cite{DBLP:conf/tacas/MouraB08}
-through their C API, and can also operate with any SMT-solver supporting the
-SMT-Lib 2 standard \cite{BarST-SMTLIB,BarST-SMT-10} through a pipe.
+through their C API.
+An implementation of communications with the SMT-solver by textual messages sent through a pipe following the SMT-Lib~2 standard \cite{BarST-SMTLIB,BarST-SMT-10} is underway, and now partially supports Z3 and SMTinterpol (plans include support for CVC4 and OpenSMT).
+It is unfortunately impossible to ignore differences between solvers behind the supposedly standard interface, since different solvers may support slightly different sets of operators and settings and may return models in different formats.
 

### 2012-05-19 papers/TAPAS2012/pagai.tex
bla bla sur feedback vertex set

@@
 through their C API.
-An implementation of communications with the SMT-solver by textual messages sent through a pipe following the SMT-Lib~2 standard \cite{BarST-SMTLIB,BarST-SMT-10} is underway, and now partially supports Z3 and SMTinterpol (plans include support for CVC4 and OpenSMT).
+An implementation of communications with the SMT-solver by textual messages sent through a pipe following the SMT-Lib~2 standard \cite{BarST-SMTLIB,BarST-SMT-10} is underway, and now partially supports Z3 and SMTinterpol (plans include support for e.g.
+CVC4 and OpenSMT).
 It is unfortunately impossible to ignore differences between solvers behind the supposedly standard interface, since different solvers may support slightly different sets of operators and settings and may return models in different formats.
@@
 For each program, we distinguish a set $P_R = P_W$ of suitable widening points by a simple algorithm: for each procedure, compute the strongly connected components of its control-flow graph using Tarjan's algorithm; the targets of the back-edges of the depth-first search are added to~$P_R$.
-Note that the resulting set is not necessarily minimal, but is sufficient to disconnect all cycles --- more sophisticated techniques are discussed in e.g.
-\citet{BourdonclePhd}.
+The resulting \emph{feedback vertex set} is not necessarily minimal, but is sufficient to disconnect all cycles --- more sophisticated techniques are discussed in e.g. \citet{BourdonclePhd}.
+It would be possible to obtain a feedback vertex set minimal with respect to inclusion by successive removal of nodes; however obtaining one of minimal cardinality is a NP-complete problem.
+It is however unclear whether more advanced selection techniques would finally yield stronger invariants; the current simple scheme has the advantage that, when run over a control-flow graph obtained from a structured program, it marks heads of loops, which is a ``natural'' choice.
 

### 2012-05-19 papers/TAPAS2012/pagai.tex
bla bla sur les variables

@@
 In concrete terms, an assignment \lstinline|x=2*x+1;| gets translated into a definition $x_2 = 2x_1+1$, with distinct variables $x_1$ and $x_2$ corresponding to the same original variable \lstinline|x| at different points in the program.
+Because LLVM generally assigns rather straightforward names (e.g.
+\lstinline|x.0| for the first renaming of variable \lstinline|x|), the user can map the invariants back to the original source code; an automatic back-to-source mapping, based on debugging information, is being developed.
+
 LLVM makes it easy to follow definition-use and use-definition chains: for a given variable (say, $x_2$) one can immediately obtain its definition (say, $2x_1+1$).
@@
 
-SSA introduces $\phi$-functions at the head of a control code to define variables whose value depends on which incoming edge to this control node was last taken.
+SSA introduces $\phi$-functions at the head of a control code to define variables whose value depends on which incoming edge was last taken to reach this control node.
 For instance, for \lstinline|if (...) { x = 2*x+1; } else { x= 0; }|, then $x_2$ is defined as $\phi(2x_1+1,0)$.
-In this framework, each variable is uniquely defined as an arithmetic ($+$, $-$, $\times$, $/$) function of other variables that themselves may not be representable as affine linear functions, because they are defined using $\phi$-functions, bitwise arithmetic, loads from memory, or return values from function calls.
+
+In this framework, each variable is uniquely defined as an arithmetic ($+$, $-$, $\times$, $/$) function of other variables that themselves may not be representable as arithmetic functions, because they are defined using $\phi$-functions, loads from memory, return values from function calls, or other numerical operations (e.g.
+bitwise operators) that are not representable with our class of basic arithmetic operations.
+We may even vary the class of arithmetic operations, for instance, by restricting ourselves to linear ones.
 
@@
 coordinates in the abstract domain (e.g. as dimensions in polyhedra), assuming
-they are live at the associate control point.
+they are live at the associated control point.
 
@@
 	$x$ as a dimension for the abstract value at point $p$, we only have $y$ and $z$. All the properties
-	for $x$ can be directly extracted from the abstract value attached to $p$ and the information $x=y+z$.
+	for $x$ can be directly extracted from the abstract value attached to $p$ and the relation $x=y+z$.
 	This is an optimisation in the sense that there is redundant information in
@@
 such operations are directly supported by APRON. This has three benefits:
-(i) it limits the number of dimensions in the abstract values, since polyhedra libraries typically perform worse with higher dimensions;
+(i) it limits the number of dimensions in the abstract values, since polyhedra libraries typically perform worse with higher dimensions;%
+\footnote{The additional dimensions express linear equalities between variables, which could be dealt with efficiently within polyhedra libraries, which anyway keep a basis of the linear affine variety containing the polyhedra.
+They should therefore cost little assuming some sparse representation of the constraints.
+Alas, several libraries, including APRON, compute with \emph{dense} vectors and matrices, which means that any increase in dimensions slows computations.}
 (ii) the abstract operation for a single path in path-focusing methods also is a (large) parallel assignment;

### 2012-05-19 papers/TAPAS2012/pagai.tex
hop

@@
 function \emph{assume}.
-\subsection{Limitations of the tool}
+
+\emph{Assert} and \emph{assume} are implemented as C macros.
+\lstinline|assert(x)| is roughly defined as \lstinline|if (!
+x) __assert_fail();|, and the tool just tests for the reachability of \lstinline|__assert_fail();|: if it is unreachable, then the assertion is true.
+
+\subsection{Current limitations of the tool}
 
@@
 We therefore run it after the ``memory to registers'' (\texttt{mem2reg}) optimization phase in LLVM, which lifts most memory accesses to scalar variables.
-The remaining memory accesses are treated as undefined values, as are the return
-values of function calls. For better precision, we also apply function inlining
-to the program, and we unroll every loop once.
+The remaining memory accesses are treated as nondeterministic choices.
+
+The analysis is currently intraprocedural: function calls are ignored in a sound way (the return value is a nondeterministic choice, the value of all variables escaping from the local scope is discarded...).
+In order to increase precision, we apply function inlining as an LLVM optimization phase.
+Plans for interprocedural analysis include computing input/output summaries for functions as elements of the abstract domain (e.g.
+if the function operates over variables $x$ and $y$, then one could compute a polyhedron over $(x,y,x',y')$ encompassing all input-output pairs) or as more general formulas.
+
+Since it is often advantageous to distinguish whether a loop has been executed at least once,%
+\footnote{Consider the very simple loop \lstinline|for(int i=0; i<n; i++) {}|.
+The obvious loop invariant is $0 \leq \lstinline|i| \leq \lstinline|n|$, but it is valid only if $\lstinline|n| > 0$.
+One would thus need to use disjunctive loop invariants to obtain
+$0 \leq \lstinline|i| \leq \lstinline|n| \lor
+ (\lstinline|i|=0 \land \lstinline|n| \leq 0)$.
+It is much simpler to unroll the loop once.}
+we unroll every loop once, again with a LLVM optimization phase.
 
@@
 the Astr\'ee system \citep{ASTREE_ESOP05,ASTREE_PLDI03}, but porting these techniques to our iteration schemes using SMT-solving requires supplemental work.
+It is unclear whether one should use bitvector arithmetic inside the SMT formula, or use mathematical integer with explicit control splits for wraparound (e.g.
+an operation $z = x+y$ over $n$-bit signed integers would appear as the disjunction of three statements $z = x+y \land -2^{n-1} \leq x+y < 2^{n-1}$,
+$z = x+y+2^n \land -x+y < -2^{n-1}$,
+$z = x+y-2^n \land x+y \geq 2^{n-1}$).
 
@@
 techniques, as proposed by \citet{Monniaux_Gonnord_SAS11}. Instead, it simply runs widening and narrowing iterations on a single path.
+We currently analyze each strongly connected component of the control-flow graph in topological order; thus each loop nest gets analyzed as a single fixed point.
+An alternative method would be to recursively decompose the strongly connected components (for structured programs, this amounts to reconstructing the nested loop structure) and summarize the inner loops before analyzing the outer loop.
+
 The analysis is currently only forward, even though nothing in the techniques
-implemented is specific to forward analysis.
+implemented is specific to forward analysis. A possible extension would therefore be backward analysis from the \lstinline|__assert_fail()| statements.
 

### 2012-05-22 papers/TAPAS2012/pagai.tex
tassage en 12 pages

@@
 \usepackage{mathcomp,amsfonts,amssymb}
-\usepackage[authoryear]{natbib}
+\usepackage[sort,numbers]{natbib}
 \usepackage[mathscr]{euscript}
@@
 
+\lstset{language=C,
+  basicstyle={\fontfamily{ppl}\selectfont}, % Palatino
+  commentstyle=\textit,
+  showstringspaces=false}
+
 \lstnewenvironment{C}
-{\lstset{language=C,
-		basicstyle=\ttfamily,
-		commentstyle=\textit,
-		showstringspaces=false}}
+{\lstset{language=C}}
 {}
@@
     Grenoble, France}
+
+  \begin{minipage}{0.45\textwidth}\begin{center}
   \author{David Monniaux\thanksref{imag.fr}}
@@
     Grenoble, France}
+  \end{center}\end{minipage}
+  \begin{minipage}{0.45\textwidth}\begin{center}
   \author{Matthieu Moy\thanksref{grenoble-inp.fr}}
@@
     Grenoble, France} 
+  \end{center}\end{minipage}
 
     \thanks[imag.fr]{Email:
-    \href{First Name.Last Name@imag.fr} {\texttt{\normalshape
+    \href{FirstName.LastName@imag.fr} {\texttt{\normalshape
         First Name.Last Name@imag.fr}}} 
@@
 
-	PAGAI implements various state-of-the-art algorithms combining abstract
-	interpretation and decision procedures (SMT-solving),
+	PAGAI implements various state-of-the-art algorithms combining
+        abstract interpretation and decision procedures (SMT-solving),
         focusing on distinction of paths inside the control flow graph while
         avoiding systematic exponential enumerations.
-        It is parametric in the abstract domain in use, the algorithm,
-        and the decision procedure.
+        It is parametric in the abstract domain in use, the iteration
+        algorithm, and the decision procedure.
 
@@
 Applications include:
-showing that a program has no runtime error (such as arithmetic overflow, division by zero, array access out of bounds), as in e.g. the Astrée analyzer \cite{ASTREE_ESOP05};
-computing invariants for use with assisted proof systems (such as the B method, Frama-C\footnote{http://www.frama-c.com/}), thereby lessening the burden on the user;
+showing that a program cannot encounter a runtime error (such as arithmetic overflow, division by zero, array access out of bounds), as in e.g. the Astrée analyzer \cite{ASTREE_ESOP05};
+computing invariants for use with assisted proof systems (such as the B method), thereby lessening the burden on the user;
 computing invariants for advanced optimization techniques in compilation (e.g. showing that two array cells are distinct, in order to allow instruction reordering between assignments to these cells).
+All these applications need invariants on numerical quantities.
 
@@
 PAGAI takes as input a program in the ``bitcode'' intermediate representation of LLVM\citep{LLVM_langref,Lattner:2004:LCF:977395.977673}, a modern compilation framework.
-LLVM bitcode is a target for several industrial-strengh compilers, most notably Clang (supporting C, C++, Objective-C and Objective-C++) and llvm-gcc (supporting C, C++, Objective-C, Objective-C++, Fortran and Ada);
+LLVM bitcode is a target for several industrial-strengh compilers, most notably Clang (supporting C, C++, Objective-C and Objective-C++) and llvm-gcc (supporting, in addition to these, Fortran and Ada);
 furthermore, a growing number of analysis tools, testing tools, etc. are currently built around this platform (Calysto, KLEE, LAV, LLBMC).
@@
 PAGAI is based on abstract interpretation, a general framework for fully automatic static analysis.
-PAGAI infers invariants of a selected form; by default it performs \emph{linear relation analysis}, which obtains invariants as conjunctions of linear inequalities (or, equivalently, convex polyhedra), but it also supports most of the abstract domains provided by the APRON library (products of intervals, octagons, linear congruences, and combinations thereof) through a runtime option.
-Depending on the algorithm (\emph{iteration strategy}) selected, PAGAI may also infer invariants as \emph{disjunctions} of elements of the abstract domain (e.g. unions of convex polyhedra).
-% TODO: citer des réfs sur APRON, les octogones etc.
+PAGAI infers invariants of a selected form; by default it performs \emph{linear relation analysis}, which obtains invariants as conjunctions of linear inequalities (or, equivalently, convex polyhedra), but it also supports other abstract domains through a runtime option.
+Depending on the \emph{iteration algorithm} selected, PAGAI may also infer invariants as \emph{disjunctions} of elements of the abstract domain (e.g. unions of convex polyhedra).
 
@@
 
-% TODO refs
-
 The PAGAI tool is dedicated to experimenting new analysis algorithms.
@@
 We thus conducted extensive experiments both on examples we produced ourselves (sometimes inspired by industrial code) and on GNU programs.
+
 The ability to run on any C or C++ code, through the LLVM system, is especially useful for the latter: front-ends for many analysis tools put restrictions (e.g.
 no backward \lstinline|goto| instructions, no pointer arithmetic...), often satisfied by safety-critical embedded programs, but not by generic programs.
+Our tool suffers no such restrictions, though it may in some cases apply coarse abstractions which may possibly yield weak invariants.
 
@@
 An example of program where such a loss of precision occurs is depicted in
-Figure \ref{example}.
+Fig.~\ref{example}.
 In this program, the loop body has two feasible paths that
@@
 Again, the number of such paths is exponential in the number of tests.
-We therefore consider paths only if they will make the final convex hull grow.
-The next path to consider is obtained from a model of an arithmetic formula expressing this condition \cite{Monniaux_Gonnord_SAS11}; if this formula is unsatisfiable, this means there is no such path and thus the convex hull obtained truly is the one for all paths.
+but we consider paths only if they will make the final convex hull grow.
+The next path to consider is obtained from a model of an arithmetic formula expressing this growth condition \cite{Monniaux_Gonnord_SAS11}; if this formula is unsatisfiable, this means there is no such path and thus the convex hull encompasses the outcome of all paths.
 The advantages of this method over the preceding one are twofold: there is no exponentially large list of abstract elements, and the satisfiability query for the formula is handed over to a \emph{satisfiability modulo theory} (SMT) solver.
@@
 Polka ``PK'' library), octagons, and products of intervals. It also has
-an interface with the Parma Polyhedra Library (TODO CITATION!!), giving access to
-more abstract domains (for instance, it can use a reduced product of polyhedra and linear congruences, producing invariants such as $0 \leq x \leq 1001 \land x \equiv 0 \pmod 7$).
+an interface with the Parma Polyhedra Library \cite{BagnaraHZ08SCP}, giving access to
+more abstract domains (e.g. a reduced product of polyhedra and linear congruences, producing invariants such as $0 \leq x \leq 1001 \land x \equiv 0 \pmod 7$).
 
-For SMT-solving, our analyzer uses Yices
-\cite{DBLP:conf/cav/DutertreM06} or Microsoft Z3\cite{DBLP:conf/tacas/MouraB08}
+For SMT-solving, our analyzer uses
+Yices \cite{DBLP:conf/cav/DutertreM06}
+or Microsoft Z3\cite{DBLP:conf/tacas/MouraB08}
 through their C API.
 An implementation of communications with the SMT-solver by textual messages sent through a pipe following the SMT-Lib~2 standard \cite{BarST-SMTLIB,BarST-SMT-10} is underway, and now partially supports Z3 and SMTinterpol (plans include support for e.g.
-CVC4 and OpenSMT).
-It is unfortunately impossible to ignore differences between solvers behind the supposedly standard interface, since different solvers may support slightly different sets of operators and settings and may return models in different formats.
+CVC4 and OpenSMT).%
+\footnote{%
+It is unfortunately impossible to ignore differences between solvers behind the supposedly standard interface, since different solvers may support slightly different sets of operators and settings and may return models in different formats.}
 
@@
 For each program, we distinguish a set $P_R = P_W$ of suitable widening points by a simple algorithm: for each procedure, compute the strongly connected components of its control-flow graph using Tarjan's algorithm; the targets of the back-edges of the depth-first search are added to~$P_R$.
-The resulting \emph{feedback vertex set} is not necessarily minimal, but is sufficient to disconnect all cycles --- more sophisticated techniques are discussed in e.g. \citet{BourdonclePhd}.
-It would be possible to obtain a feedback vertex set minimal with respect to inclusion by successive removal of nodes; however obtaining one of minimal cardinality is a NP-complete problem.
+The resulting \emph{feedback vertex set} is not necessarily minimal, but is sufficient to disconnect all cycles --- more sophisticated techniques are discussed in e.g. \citet{BourdonclePhd}.%
+\footnote{
+It would be possible to obtain a feedback vertex set minimal with respect to inclusion by successive removal of nodes; however obtaining one of minimal cardinality is a NP-complete problem.}
 It is however unclear whether more advanced selection techniques would finally yield stronger invariants; the current simple scheme has the advantage that, when run over a control-flow graph obtained from a structured program, it marks heads of loops, which is a ``natural'' choice.
@@
 Because LLVM generally assigns rather straightforward names (e.g.
-\lstinline|x.0| for the first renaming of variable \lstinline|x|), the user can map the invariants back to the original source code; an automatic back-to-source mapping, based on debugging information, is being developed.
+\lstinline|x.0| for the first renaming of variable \lstinline|x|), the user can map the invariants back to the original source code; an automatic and more robust back-to-source mapping, based on debugging information, is being developed.
 
@@
 bitwise operators) that are not representable with our class of basic arithmetic operations.
-We may even vary the class of arithmetic operations, for instance, by restricting ourselves to linear ones.
+We may vary the class of arithmetic operations, for instance, by restricting ourselves to linear ones.
 
@@
 for each control point in $P_R$ (typically, the widening points).
-When a program contains \emph{assert} function call, PAGAI also outputs whether
+When a program contains an \lstinline|assert| function call, PAGAI also outputs whether
 the statement has been proved.
 It is also possible to add some preconditions about the variables, etc, using a
-function \emph{assume}.
+function \lstinline|assume|.
 
-\emph{Assert} and \emph{assume} are implemented as C macros.
+\lstinline|assert| and \lstinline|assume| are implemented as C macros.
 \lstinline|assert(x)| is roughly defined as \lstinline|if (!
@@
 
-\subsection{Current limitations of the tool}
+\subsection{Current limitations of the tool, possible future works}
 
@@
 techniques, as proposed by \citet{Monniaux_Gonnord_SAS11}. Instead, it simply runs widening and narrowing iterations on a single path.
+
 We currently analyze each strongly connected component of the control-flow graph in topological order; thus each loop nest gets analyzed as a single fixed point.
 An alternative method would be to recursively decompose the strongly connected components (for structured programs, this amounts to reconstructing the nested loop structure) and summarize the inner loops before analyzing the outer loop.
-
 The analysis is currently only forward, even though nothing in the techniques
@@
 \subsection{Precision of the various techniques}
-\label{sec:compare_techniques}
+% \label{sec:compare_techniques}
 
-\begin{figure}[h]
-  \begin{center}
-    \input{./gnuplot/techniques.tex}
-  \end{center} 
-  \vspace{-20pt}
-  \caption{Comparison of the abstract values obtained by classical abstract
-  interpretation (S), \emph{Guided Static
-  Analysis} (G), \emph{Path-focused} technique (PF), our combined technique
-  (G+PF), and its version with disjunctive invariants (DIS).
-  The $\subsetneq$ bars gives the percentage of invariants stronger (more precise; smaller with respect to inclusion) with the left-side technique,
-$\supsetneq$ the percentage of invariants stronger with the right-side technique,
-and ``uncomparable'' gives the percentage of invariants that are uncomparable, i.e
-neither greater nor smaller;
-the code points where both invariants are equal make up the remaining percentage. See Tab.~\ref{tab:techniques} for details.}
-  \label{fig:techniques}
-\end {figure}
+% \begin{figure}[h]
+%   \begin{center}
+%     \input{./gnuplot/techniques.tex}
+%   \end{center} 
+%   \vspace{-20pt}
+%   \caption{Comparison of the abstract values obtained by classical abstract
+%   interpretation (S), \emph{Guided Static
+%   Analysis} (G), \emph{Path-focused} technique (PF), our combined technique
+%   (G+PF), and its version with disjunctive invariants (DIS).
+%   The $\subsetneq$ bars gives the percentage of invariants stronger (more precise; smaller with respect to inclusion) with the left-side technique,
+% $\supsetneq$ the percentage of invariants stronger with the right-side technique,
+% and ``uncomparable'' gives the percentage of invariants that are uncomparable, i.e
+% neither greater nor smaller;
+% the code points where both invariants are equal make up the remaining percentage. See Tab.~\ref{tab:techniques} for details.}
+%   \label{fig:techniques}
+% \end {figure}
 
@@
 
-Let us briefly comment the results given in more details in Table~\ref{tab:techniques} and Figure~\ref{fig:techniques}.
+Let us briefly comment the results given in more details in Table~\ref{tab:techniques}.
 \emph{Guided Static Analysis} from \citet{DBLP:conf/sas/GopanR07} improves the
@@
 
-While experiencing with techniques that use SMT-solving, we encountered some
+While experimenting with techniques that use SMT-solving, we encountered some
 limitations due to non-linear arithmetic in the analyzed programs. Indeed, 
@@
 \emph{Lapack/Blas} implements matrix computations, which use floating-point multiplications.
-
 In cases where the formula is expressed in too rich a logic for the SMT-solver to deal with, a number of workarounds are possible, including:
@@
 using the classical widening operator (POLY). POLY and POLY* use the Parma
-Polyhedral Library \cite{PPL}.
+Polyhedral Library \cite{BagnaraHZ08SCP}.
 $\subsetneq$, $\supsetneq$ and ``unc.'' are defined as in Fig.~\ref{fig:techniques}.}
@@
 For each program and each pair $(D_1,D_2)$ of abstract domains, we compare by
-inclusion the invariants of the different control points in $P_R$ (where $P_R$
-is defined as in \ref{sec:compare_techniques}).
-The main results are given in Table \ref{tab:domain}.
+inclusion the invariants of the different control points in $P_R=P_W$ (Tab.~\ref{tab:domain}).
 
@@
 
-\section{Conclusion and future work}
+\phantomsection\addcontentsline{toc}{section}{References} 
+{\newcommand{\issn}[1]{{\footnotesize ISSN \href{http://www.worldcat.org/issn/#1}{#1}}}
+\newcommand{\isbn}[1]{{\footnotesize ISBN \href{http://www.worldcat.org/isbn/#1}{#1}}}
+\newcommand{\doix}[1]{\href{http://dx.doi.org/#1}{#1}\endgroup}
+\newcommand{\doi}{\begingroup\footnotesize doi: \catcode`\_=13\def\_{\textunderscore}\doix}
 
-\phantomsection\addcontentsline{toc}{section}{References} 
-	\bibliographystyle{plainnat}
+\bibliographystyle{dmplainnat}
 \bibliography{pagai}
+}
 \end{document}

### 2012-05-22 papers/TAPAS2012/pagai.tex
12 pages

@@
 \usepackage{dcolumn}
+\usepackage{listings}
 
-\usepackage{listings}
+\hypersetup{pageanchor=false}
 
@@
 
+\def\arraystretch{1.2}
 \def\newblock{\hskip .11em plus .33em minus .07em}
@@
 
-\newcommand{\Nat}{{\mathbb N}}
-\newcommand{\Real}{{\mathbb R}}
 \def\lastname{Henry, Monniaux \& Moy}
+
 \begin{document}
@@
 
-\begin{table}[!h]
-	\tiny
-	\centering
+\begin{table}[!htb]
+\begin{center}
 \begin{tabular}{|l|r|r|} \hline
@@
 \end{tabular}
+\end{center}
+
 \caption{List of analyzed open-source projects, with their respective number of
 lines of code, and their number of control points in $P_R$}
-\label{fig:projects}
+\label{tab:projects}
 \end{table}
@@
 	only if one of these conditions holds:
-		\begin{itemize}
-		\item $v$ is live in $p$.
-		\item There is a variable $v'$, defined as a linear combination of other
+		(i) $v$ is live in $p$.
+        (ii) There is a variable $v'$, defined as a linear combination of other
 		variables $v_1, v_2, \dots, v_n$, so that $\exists i \in \{1,\dots,n\}, v = v_i$,
 		and $v'$ is live by linearity in $p$.
-		\end{itemize}
 	\end{definition}
@@
 We conducted extensive experiments on real-life programs in order to compare the
-different techniques, mostly on open-source projects (Fig.~\ref{fig:projects}) written in C, C++ and Fortran.
+different techniques, mostly on open-source projects (Tab.~\ref{tab:projects}) written in C, C++ and Fortran.
 
@@
  \input{table1}
+\hline
 \end{tabular}
@@
 over the classic Abstract interpretation algorithm.
-$\subsetneq$, $\supsetneq$ and ``unc.'' are defined as in Fig.~\ref{fig:techniques}.}
+$\subsetneq$ (resp.~$\supsetneq$) gives the percentage of invariants stronger (more precise; smaller with respect to inclusion) with the left-side (resp. right-side) technique,
+and ``uncomparable'' gives the percentage of invariants that are uncomparable, i.e neither greater nor smaller;
+the code points where both invariants are equal make up the remaining percentage}
 \label{tab:techniques}
@@
 
-\begin{table}[!h]
-	\centering \tiny
+\begin{table}[!tb]
+\begin{center}\small
 \begin{tabular}{|l|r|r|r|r|r|} \hline
@@
 \end{tabular}
+\end{center}
 \caption{Execution time for each technique, expressed in seconds}
@@
 \label{sec:compare_domains}
-\begin{table*}
+\begin{table}[!tb]
 	\tiny
@@
 Polyhedral Library \cite{BagnaraHZ08SCP}.
-$\subsetneq$, $\supsetneq$ and ``unc.'' are defined as in Fig.~\ref{fig:techniques}.}
+$\subsetneq$, $\supsetneq$ and ``unc.'' are defined as in Tab.~\ref{tab:techniques}.}
 \label{tab:domain}
-\end{table*}
+\end{table}
 

### 2012-05-22 papers/TAPAS2012/pagai.tex
tassage
un peu de future work ?

--- 2012-05-19 papers/TAPAS2012/pagai.tex
@@
 Table \ref{tab:time} gives the execution time of the different analysis
-techniques. It is interesting to see that \emph{Guided Static Analysis} and
-\emph{Path-focusing} are sometimes faster than the classical
+techniques. It is interesting to see that
+\emph{Path-focusing} is sometimes faster than the classical
 algorithm.
@@
 
-\section{Conclusion and future work}
+It is not totally relevant to compare by inclusion the abstract values obtained
+by the various analysis techniques. Indeed, a slightly smaller invariant may not
+always be useful to prove the desired properties.
+A future work is thus to experiment with
+better comparison metrics, such as:
+\begin{itemize}
+	\item the number of \emph{assert} that have been proved in the code.
+		Unfortunately, it is difficult to find good benchmarks or real life
+		programs with many \emph{assert} statements.
+	\item the number of false alarms in a client analysis that detects array
+		bound violations, arithmetic overflows, etc.
+\end{itemize}
 

### 2012-05-22 papers/TAPAS2012/pagai.tex

--- 2012-05-22 papers/TAPAS2012/pagai.tex
@@
 \usepackage{entcsmacro}
+
 \usepackage{ifxetex}
@@
 \begin{figure}[!h]
-\begin{minipage}[c]{.49\linewidth}
+\begin{minipage}[c]{.35\linewidth}
 \begin{C}
@@
 \end{minipage}
-\begin{minipage}[c]{.49\linewidth}
+\begin{minipage}[c]{.64\linewidth}
 \begin{tikzpicture}[->,>=stealth',auto,node distance=1.5cm,
@@
 
+It is not totally relevant to compare by inclusion the abstract values obtained
+by the various analysis techniques. Indeed, a slightly smaller invariant may not
+always be useful to prove the desired properties.
+A future work is thus to experiment with
+better comparison metrics, such as:
+\begin{itemize}
+	\item the number of \emph{assert} that have been proved in the code.
+		Unfortunately, it is difficult to find good benchmarks or real life
+		programs with many \emph{assert} statements.
+	\item the number of false alarms in a client analysis that detects array
+		bound violations, arithmetic overflows, etc.
+\end{itemize}
+
 \phantomsection\addcontentsline{toc}{section}{References} 

### 2012-05-23 papers/SAS2012/implicitization.tex
separation entre version format LMCS et version format article ou amsart

--- 2012-03-23 papers/SAS2012/implicitization.tex
@@
-\documentclass{llncs}
-
-\usepackage{ifxetex}
-\ifxetex
-\else
-\usepackage[utf8]{inputenc}
-\fi
-
-\usepackage{amsmath}
-\usepackage{mathcomp,amsfonts,amssymb}
-\usepackage[mathscr]{euscript}
-\usepackage{algorithmic,algorithm}
-\usepackage{paralist}
-\usepackage{graphicx}
-%\usepackage{color}
-\usepackage{placeins}
-
-\usepackage{enumitem}
-\usepackage{dcolumn}
-\usepackage{rotating}
-
-\usepackage{listings}
-\lstset{language=C,mathescape=true,numbers=left,numbersep=0.8ex,tabsize=2,breaklines=true,basicstyle={\rm\fontfamily{phv}\small\selectfont}}
-
-\usepackage{tikz}
-\usetikzlibrary{arrows,automata,positioning}
-
-\tikzstyle{state}=[circle,fill=black!25,minimum size=13pt,inner sep=0pt]
-\tikzstyle{rstate}=[rectangle,fill=black!25,minimum size=13pt,inner sep=0pt]
-\tikzstyle{transition}=[rectangle,semithick,draw=black!75,
-  			  minimum size=4mm]
-\tikzstyle{transition2}=[transition,rectangle,thick,dashed,
-  			  minimum size=4mm]
-\tikzstyle{PRstate}=[circle,double,draw,fill=blue!15,minimum size=13pt,inner sep=0pt]
-\tikzstyle{polyhedra}=[blue!25,opacity=0.5,pattern=north west lines,pattern
-color=blue]
-\tikzstyle{line}=[black,thick]
-
-\newcommand{\mytitle}{Succinct Representations for Abstract Interpretation}
-\usepackage[pdftitle={\mytitle},pdfauthor={Julien Henry, David Monniaux, Matthieu Moy}]{hyperref}
-
-\newcommand{\abstr}[1]{#1^\sharp}
-\newcommand{\parts}[1]{\mathscr{P}(#1)}
-\newcommand{\ZZ}{\mathbb{Z}}
-\newcommand{\QQ}{\mathbb{Q}}
-\newcommand{\RR}{\mathbb{R}}
-\newcommand{\widening}{\mathop{\triangledown}}
-
-\graphicspath{{figures/}{gnuplot/}}
-
-% commentaires dans les algorithmes crees avec le package algorithmic
-\renewcommand{\algorithmiccomment}[1]{ {\color{gray} // #1}}
-
-\usepackage{ifthen}
-\newboolean{draft}
-%\setboolean{draft}{true}
-\setboolean{draft}{false}
-
-\ifthenelse{\boolean{draft}}{
-  \newcommand{\avirer}[1]{{\color{red} #1}}
-  \newcommand{\MM}[1]{{\color{blue} TODO(MM): #1}}
-  \newcommand{\JH}[1]{{\color{teal} TODO(JH): #1}}
-  \newcommand{\DM}[1]{{\color{violet} TODO(DM): #1}}
-}{
-  \newcommand{\avirer}[1]{}
-  \newcommand{\MM}[1]{}
-  \newcommand{\JH}[1]{}
-  \newcommand{\DM}[1]{}
-}
-
-\title{\mytitle}
-\subtitle{Combined analysis algorithms and experimental evaluation}
-\author{Julien Henry\inst{1}\inst{2} \and David Monniaux\inst{1}\inst{3} \and Matthieu Moy\inst{1}\inst{4}}
-
-\institute{VERIMAG laboratory, Grenoble, France
-\and Université Joseph Fourier
-\and CNRS
-\and Grenoble-INP}
-
-\renewcommand{\baselinestretch}{1}
-
-\begin{document}
 \maketitle
@@
 Further investigations are needed not only on good abstractions for pointers (many are already known) but also on their conversion to SMT problems.
-
-\FloatBarrier
-
-\phantomsection\addcontentsline{toc}{section}{References} 
-\bibliographystyle{splncs03}
-\bibliography{implicitization}
-
-\appendix
-
-\MM{TODO: la partie appendix est laissée en chantier. A priori, on ne
-  mettra que les tableaux de chiffres sans les commentaires, et les
-  tableaux risquent de changer avec les avancées de Julien.}
-
-\section{Experimental Results}
-
-Experimental results of our tool on the benchmark presented in
-Fig~\ref{fig:projects} are given below. The precision of different
-techniques is given in Table~\ref{tab:techniques}. The precision of
-abstract domains and widening operators in given in
-Table~\ref{tab:domain}.
-
-\begin{sidewaystable}
-\begin{center}
-\setlength{\tabcolsep}{0.75ex}
-\begin{tabular}{|l
-|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
-|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
-|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
-|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
-|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
-|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}|} \hline
-\multicolumn{1}{|c|}{\textbf{Benchmark}}
-& \multicolumn{3}{c|}{\textbf{G/S}}
-& \multicolumn{3}{c|}{\textbf{PF/S}}
-& \multicolumn{3}{c|}{\textbf{PF/G}}
-& \multicolumn{3}{c|}{\textbf{G+PF/PF}}
-& \multicolumn{3}{c|}{\textbf{G+PF/G}}
-& \multicolumn{3}{c|}{\textbf{DIS/G+PF}} \\ %\cline{2-16}
-& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
-& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
-& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
-& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
-& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
-& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.} \\
- \hline
- \input{table1}
-\end{tabular}
-\end{center}
-\caption{Result of the comparison of the various techniques described in this
-paper: classic Abstract Interpretation (S), \emph{Guided Static Analysis} (G),
-\emph{Path-focusing} (PF), our combined technique (G+PF), and its version using
-disjunctive invariants (DIS). For
-instance, \textbf{G/S} compares the benefits of \emph{Guided Static Analysis}
-over the classic Abstract interpretation algorithm.
-$\subsetneq$, $\supsetneq$ and ``unc.'' are defined as in Fig.~\ref{fig:techniques}.}
-\label{tab:techniques}
-\end{sidewaystable}
-
-\label{sec:compare_domains}
-\begin{sidewaystable}
-\begin{center}
-\setlength{\tabcolsep}{0.75ex}
-\begin{tabular}{|l
-|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
-|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
-|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
-|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
-|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
-|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}|} \hline
-\multicolumn{1}{|c|}{\textbf{Benchmark}}
-& \multicolumn{3}{c|}{\textbf{PK/OCT}}
-& \multicolumn{3}{c|}{\textbf{PK/BOX}}
-& \multicolumn{3}{c|}{\textbf{OCT/BOX}}
-& \multicolumn{3}{c|}{\textbf{PK/PKEQ}}
-& \multicolumn{3}{c|}{\textbf{PK/GRID}}
-& \multicolumn{3}{c|}{\textbf{POLY/POLY*}} \\ %\cline{2-16}
-& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
-& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
-& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
-& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
-& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
-& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.} \\
- \hline
- \input{table_domain.tex}
-\end{tabular}
-\end{center}
-\caption{Results of the comparison of the various abstract domains, when using
-the same technique (G+PF). We used as abstract domains Convex Polyhedra (PK and
-POLY),
-Octagons (OCT), intervals (BOX), linear equalities (PKEQ) and linear congruences
-(GRID). The last column compares the domain of Convex Polyhedra with the
-improved widening operator from \cite{BagnaraHRZ05SCP} (POLY*), and Convex Polyhedra
-using the classical widening operator (POLY). POLY and POLY* use the Parma
-Polyhedral Library \cite{PPL}.
-$\subsetneq$, $\supsetneq$ and ``unc.'' are defined as in Fig.~\ref{fig:techniques}.}
-\label{tab:domain}
-\end{sidewaystable}
-
-\end{document}

### 2012-05-23 papers/SAS2012/implicitization_lmcs.tex
separation entre version format LMCS et version format article ou amsart

--- 2011-11-04 papers/PLDI2012/implicitization.tex
@@
-\documentclass[preprint]{sigplanconf}
+\documentclass{llncs}
 
-% The following \documentclass options may be useful:
-%
-% 10pt          To set in 10-point type instead of 9-point.
-% 11pt          To set in 11-point type instead of 9-point.
-% authoryear    To obtain author/year citation style instead of numeric.
-
-%\newcommand{\MM}[1]{{\color{blue} TODO(MM): #1}}
-%\newcommand{\JH}[1]{{\color{teal} TODO(JH): #1}}
-%\newcommand{\DM}[1]{{\color{violet} TODO(DM): #1}}
-\newcommand{\MM}[1]{}
-\newcommand{\JH}[1]{}
-\newcommand{\DM}[1]{}
-
-% Pour les commentaires
+\usepackage{ifxetex}
+\ifxetex
+\else
 \usepackage[utf8]{inputenc}
+\fi
 
 \usepackage{amsmath}
-\usepackage{mathcomp,amsfonts,amssymb,MnSymbol}
-\usepackage{dmnatbib}
+\usepackage{mathcomp,amsfonts,amssymb}
 \usepackage[mathscr]{euscript}
 \usepackage{algorithmic,algorithm}
-%\usepackage{graphicx}
+\usepackage{paralist}
+\usepackage{graphicx}
 %\usepackage{color}
+\usepackage{placeins}
 
@@
 
+\usepackage{ifthen}
+\newboolean{draft}
+%\setboolean{draft}{true}
+\setboolean{draft}{false}
+
+\ifthenelse{\boolean{draft}}{
+  \newcommand{\avirer}[1]{{\color{red} #1}}
+  \newcommand{\MM}[1]{{\color{blue} TODO(MM): #1}}
+  \newcommand{\JH}[1]{{\color{teal} TODO(JH): #1}}
+  \newcommand{\DM}[1]{{\color{violet} TODO(DM): #1}}
+}{
+  \newcommand{\avirer}[1]{}
+  \newcommand{\MM}[1]{}
+  \newcommand{\JH}[1]{}
+  \newcommand{\DM}[1]{}
+}
+
+\title{\mytitle}
+\subtitle{Combined analysis algorithms and experimental evaluation}
+\author{Julien Henry\inst{1}\inst{2} \and David Monniaux\inst{1}\inst{3} \and Matthieu Moy\inst{1}\inst{4}}
+
+\institute{VERIMAG laboratory, Grenoble, France
+\and Université Joseph Fourier
+\and CNRS
+\and Grenoble-INP}
+
+\renewcommand{\baselinestretch}{1}
+
 \begin{document}
+\input{implicitization}
 
-\conferenceinfo{PLDI '12}{11-16 June 2012, Beijing} 
-\copyrightyear{2012} 
-\copyrightdata{[to be supplied]}
+\FloatBarrier
 
-%\titlebanner{banner above paper title}        % These are ignored unless
-\preprintfooter{Succinct representations for abstract interpretation}   % 'preprint' option specified.
+\phantomsection\addcontentsline{toc}{section}{References} 
+\bibliographystyle{splncs03}
+\bibliography{implicitization}
 
-\title{Succinct Representations for Abstract Interpretation}
-\subtitle{Combined analysis algorithms and experimental evaluation}
+\appendix
 
-\authorinfo{Anonymous}
-           {The Internet}
-           {anonymous.gal@gmail.com}
+\MM{TODO: la partie appendix est laissée en chantier. A priori, on ne
+  mettra que les tableaux de chiffres sans les commentaires, et les
+  tableaux risquent de changer avec les avancées de Julien.}
 
-\maketitle
+\section{Experimental Results}
 
-\begin{abstract}
-Abstract interpretation techniques can be made more precise by distinguishing
-paths inside loops, at the expense of possibly exponential complexity. SMT-solving techniques and sparse representations of paths and sets of paths avoid this pitfall.
+Experimental results of our tool on the benchmark presented in
+Fig~\ref{fig:projects} are given below. The precision of different
+techniques is given in Table~\ref{tab:techniques}. The precision of
+abstract domains and widening operators is given in
+Table~\ref{tab:domain}.
 
-We improve previously proposed techniques for guided static analysis and the generation of disjunctive invariants by combining them with techniques for succinct representations of paths and symbolic representations for transitions based on static single assignment.
-
-Because of the non-monotonicity of the results of abstract interpretation with widening operators, it is difficult to conclude that some abstraction is more precise than another based on theoretical local precision results.
-We thus conducted extensive comparisons between our new techniques and previous ones, on a variety of open-source packages.
-We also took the opportunity to compare the precision of various classical abstract domains (convex polyhedra, octagons, intervals).
-\end{abstract}
-
-% \category{CR-number}{subcategory}{third-level}
-\category{D.2.4}{Software/Program Verification}{Formal methods}
-\category{F.3.1}{Specifying and Verifying and Reasoning about Programs}{Assertions, Invariants, Mechanical verification, Pre- and post-conditions}
-\category{F.3.2}{Semantics of Programming Languages}{Program analysis}
-\category{F.4.1}{Mathematical Logic}{Logic and constraint programming, Mechanical theorem proving}
-\category{I.2.3}{Deduction and Theorem Proving}{Inference engines, Logic programming, Resolution}
-
-\terms
-theory, verification, algorithms
-
-\keywords
-SMT-solving, static analysis, abstract interpretation, disjunctions
-
-\section{Introduction}
-Static analysis by abstract interpretation is a fully automatic program analysis method.
-When applied to imperative programs, it computes an inductive invariant mapping each program location (or a subset thereof) to a set of states represented symbolically~\cite{CousotCousot_JLC92}.
-For instance, if we are only interested in scalar numerical program variables, such a set may be a convex polyhedron (the set of solutions of a system of linear inequalities)~\cite{CousotHalbwachs78,Halbwachs_PhD,PPL,BagnaraHZ08SCP}.
-
-In such an analysis, information may flow forward (one computes the polyhedron after a program statement as the image by the semantics of the statement, or a super-set thereof, of the polyhedron before) or backward;
-forward program analysis computes super-sets of the states reachable from the initialization of the program, backward program analysis computes super-sets of the states co-reachable from some property of interest (for instance, the violation of an assertion).
-In forward analysis, control-flow joins correspond to convex hulls if using convex polyhedra (more generally, they correspond to least upper bounds in a lattice); in backward analysis, it is control-flow splits that correspond to convex hulls.
-
-It is a known limitation of program analysis by abstract interpretation that this convex hull, or more generally, least upper bound operation, may introduce states that cannot occur in the real program: for instance, the convex hull of the intervals $[-2,-1]$ and $[1,2]$ is $[-2,2]$, strictly larger than the union of the two.
-Such introduction may prevent proving desired program properties, for instance $\neq 0$.
-The alternative is to keep the union symbolic (e.g.
-compute using $[-2,-1] \cup [1,2]$) and thus compute in the \emph{disjunctive completion}
-of the lattice, but the number of terms in the union may grow exponentially with the number of successive tests in the program to analyze, not to mention difficulties for designing suitable widening operators for enforcing the convergence of fixpoint iterations~\cite{PPL,BagnaraHZ08SCP,DBLP:journals/sttt/BagnaraHZ07}.
-The exponential growth of the number of terms in the union may be controlled by heuristics that judiciously apply least upper bound operations, as in the \emph{trace partitioning domain} \cite{Rival_Mauborgne_TOPLAS07} implemented in the Astr\'ee analyzer~\cite{ASTREE_PLDI03,ASTREE_ESOP05}.
-
-Assuming we are  interested in a loop-free program fragment, the above approach of keeping symbolic unions gives the same results as performing the analysis separately over every path in the fragment.
-A recent method for finding disjunctive loop invariants \cite{DBLP:conf/pldi/GulwaniZ10} is based on this idea: each path inside the loop body is considered separately.
-Two recent proposals use SMT-solving \cite{Kroening_Strichman_08} as a decision procedure for the satisfiability of first-order arithmetic formulas in order to enumerate only paths that are needed for the progress of the analysis \cite{Gawlitza_Monniaux_ESOP11,Monniaux_Gonnord_SAS11}.
-They can equivalently be seen as analyses over a multigraph of transitions between some distinguished control nodes.
-This multigraph has an exponential number of edges, but is never explicitly represented in memory; instead, this graph is \emph{implicitly} or \emph{succinctly} represented: its edges are enumerated as needed as solutions to SMT problems.
-
-An additional claim of the methods that distinguish paths inside the loop body \cite{DBLP:conf/pldi/GulwaniZ10,Monniaux_Gonnord_SAS11} is that they tend to generate better invariants that methods that do not, by behaving better with respect to the \emph{widening operators} \cite{CousotCousot_JLC92}
-used for enforcing convergence when searching for loop invariants by Kleene iterations.
-A related technique, \emph{guided static analysis} \cite{DBLP:conf/sas/GopanR07}, computes successive loop invariants for increasing subsets of the transitions taken into account, until all transitions are considered; again, the claim is that this approach avoids some gross over-approximation introduced by widenings.
-
-All these methods improve the precision of the analysis by keeping the same abstract domain (say, convex polyhedra) but changing the operations applied and their ordering.
-An alternative is to change the abstract domain, for instance by moving from intervals to octagons or convex polyhedra \cite{DBLP:journals/lisp/Mine06}, or the widening operator: for instance one may replace the classical widening on convex polyhedra \cite{CousotHalbwachs78,Halbwachs_PhD} by an improved version \cite{BagnaraHRZ05SCP}, or use widening ``up to'', that is, use linear inequalities syntactically present in the program source code as possible widening steps~\cite{Polka:FMSD:97}.
-
-There are many possible combinations of the above techniques, and it is not
-evident which ones perform more or less precisely or more or less efficiently on real-life examples. One needs to experiment.
-Unfortunately, the published literature on the subject lacks experimental comparative assessments. One purpose of this article is therefore to propose such experimental results.
-
-This article makes the following contributions:
-\begin{enumerate}
-\item We recast the guided static analysis technique from \citet{DBLP:conf/sas/GopanR07} on the expanded multigraph from \citet{Monniaux_Gonnord_SAS11}, considering entire paths instead of individual transitions, using SMT queries and binary decision diagrams.\label{contr:guided_multigraph} (See \S\ref{sec:guided_multigraph})
-\item We improve the technique for obtaining disjunctive invariants from \citet{DBLP:conf/pldi/GulwaniZ10} by replacing the explicit exhaustive enumeration of paths by a sequence of SMT queries.\label{contr:disjunctive} (See \S\ref{sec:disjunctive})
-\item We implemented these techniques, in addition to ``classical'' iterations and the original guided static analysis, inside a prototype static analyzer.
-This tool uses the LLVM bitcode format \cite{Lattner:2004:LCF:977395.977673,LLVM_langref} as input, which can be produced by compilation from C, C++ and Fortran, enabling it to be run on many real-life programs.
-With respect to abstract domains on numerical variables, it uses the APRON library \cite{DBLP:conf/cav/JeannetM09}, which supports a variety of abstract domains from which we can choose with minimal changes to our analyzer.
-
-Our tool uses a \emph{single static assignment} (SSA) intermediate representation, which allows us to perform cheap symbolic propagations and simplifications,
-described in Section~\ref{sec:analysis-algorithm}.
-
-\item We conducted extensive experiments with this tool, on real-life programs: (i) fixing the abstract domain, varying the iteration technique (\S~\ref{sec:compare_techniques}) (ii) fixing the iteration technique, varying the abstract domain (\S\ref{sec:compare_domains}).
-\end{enumerate}
-
-\section{Bases}
-\subsection{Static analysis by abstract interpretation}
-\label{sec:static_analysis}
-Let $X$ be the set of possible states of the program variables; for instance, if the program has 3 unbounded integer variables, then $X = \ZZ^3$.
-The set $\parts{X}$ of subsets of $X$, partially ordered by inclusion, is the \emph{concrete domain}.
-An \emph{abstract domain} is a set $\abstr{X}$ equipped with a partial order $\sqsubseteq$; for instance, it can be the domain of convex polyhedra in $\QQ^3$ ordered by geometric inclusion.
-In this article, all abstract domains are supposed to contain machine-representable objects, and all $\sqsubseteq$ order relations are supposed to be decidable.
-The concrete and abstract domains are connected by a monotone \emph{concretization} function $\gamma: \left(\abstr{X},\sqsubseteq\right) \rightarrow (\parts{X},\subseteq)$: an element $\abstr{x} \in \abstr{X}$ represents a set $\gamma(\abstr{x})$.%
-%
-\footnote{%
-Some presentations of abstract interpretation identify an element $\abstr{x}$ with the set of states $\gamma(\abstr{x})$ that it represents.
-This leads to simpler notations, but may also confuse if, as in some of our constructions, there exist several $\abstr{x}$ with identical $\gamma(\abstr{x})$, but distinct algorithmic behaviors.}
-
-We also assume a join operator $\sqcup: \abstr{X} \times \abstr{X} \rightarrow \abstr{X}$, with infix notation; in practice, it is generally a least upper bound operation, but we only need it to satisfy $\gamma(\abstr{x}) \cup \gamma(\abstr{y}) \subseteq \gamma(\abstr{x} \sqcup \abstr{y})$ for all $\abstr{x},\abstr{y}$.
-
-Classically, one considers the control-flow graph of the program, with edges labeled with concrete transition relations (e.g.
-$x' = x+1$ for an instruction \lstinline|x = x+1;|), and attaches an abstract element to each control point.
-A concrete transition relation $\tau \subseteq X \times X$ is replaced by an abstract \emph{forward abstract transformer} $\abstr{\tau}: \abstr{X} \rightarrow \abstr{X}$, such that
-\begin{equation}
-\forall \abstr{x} \in \abstr{X}, x,x' \in X,~
-x \in \gamma(\abstr{x}) \land (x,x') \in \tau \implies
-x' \in \gamma \circ \abstr{\tau}(\abstr{x})
-\label{eqn:transformer_correct}
-\end{equation}
-It is easy to see that if to any control point $p \in P$ we attach an abstract element $\abstr{x}_p$ such that
-(i) for any $p$, $\gamma(\abstr{x}_p)$ includes all initial states possible at control node $p$
-(ii) for any $p,p'$, $\abstr{\tau}_{p,p'} (\abstr{x}_p) \sqsubseteq \abstr{x}_{p'}$, noting $\tau_{p,p'}$ the transition from $p$ to~$p'$, then $(\gamma(\abstr{x}_p))_{p \in P}$ form an \emph{inductive invariant}: by induction, when the control point is $p$, the program state always lies in $\gamma(\abstr{x}_p)$.
-
-\emph{Kleene iterations} compute such an inductive invariant as the stationary limit, if it exists, of the following system: for each $p$, initialize $\abstr{x}_p$ such that $\gamma(\abstr{x}_p)$ is a superset of the initial states at point $p$; then iterate the following:
-if $\abstr{\tau}_{p,p'} (\abstr{x}_p) \nsqsubseteq \abstr{x}_{p'}$, replace $\abstr{x}_{p'}$ by $\abstr{x}_{p'} \sqcup \abstr{\tau}_{p,p'} (\abstr{x}_p)$.
-Such a stationary limit is bound to exist if $\abstr{X}$ has no infinite ascending chain $a_1 \sqsubsetneq a_2 \sqsubsetneq \dots$; this condition is however not met by domains such as intervals or convex polyhedra.
-
-\emph{Widening-accelerated Kleene iterations} proceed by replacing $\abstr{x}_{p'} \sqcup \abstr{\tau}_{p,p'} (\abstr{x}_p)$ by $\abstr{x}_{p'} \widening (\abstr{x}_{p'} \sqcup \abstr{\tau}_{p,p'} (\abstr{x}_p))$ where $\widening$ is a \emph{widening operator}: for all $\abstr{x},\abstr{y}$, $\gamma(\abstr{y}) \subseteq \gamma(\abstr{x} \widening \abstr{y})$, and there exists no sequence $\abstr{u}_1,\abstr{u}_2,\dots$ of the form $\abstr{u}_{n+1} = \abstr{u}_n \widening \abstr{v}_n$ where $\abstr{v}_n$ is another sequence.
-Such iterations necessarily lead to a stationary limit $(\abstr{x}_p)_{p \in P}$, which defines an inductive invariant $(\gamma(\abstr{x}_p))_{p \in P}$.
-Note that this invariant is not, in general, the least one expressible in the abstract domain, and may depend on the iteration ordering (the successive choices~$p,p'$).
-
-Once an inductive invariant $\gamma((\abstr{x}_p)_{p \in P})$ has been obtained,
-one can attempt \emph{decreasing} or \emph{narrowing} iterations to reduce it.
-In their simplest form, this just means running the following operation until a
-fixpoint or a maximal number of iterations are reached: for any $p'$, replace
-$\abstr{x}_{p'}$ by $\abstr{x}_{p'} \cap \left(\bigsqcup_{p \in P}
-\abstr{\tau}_{p,p'} (\abstr{x}_p)\right)$. The result also defines an inductive
-invariant. These decreasing iterations are indispensable to recover properties
-from guards (tests) in the program in most iteration settings; unfortunately,
-certain loops, particularly those involving identity (no-operation) transitions,
-may foil them: the iterations immediately reach a fixpoint and do not decrease
-further (see example in \S\ref{subsec:rate_lim}). Sections \ref{sec:guided} and
-\ref{sec:path_focusing} describe techniques that work around this problem.
-
-Widening operations have a somewhat counterintuitive behavior: the result of an analysis (postcondition) is not necessarily monotonic with respect to the precondition.
-For instance, classical interval analysis with widening and decreasing iterations on the following loop yields a loop invariant $x \in [0,+\infty)$ if the precondition is $x \in [0,0]$ , but $x \in [0,10]$ (strictly stronger) if the precondition is $x \in [0,10]$ (strictly weaker):
-\lstinline|while (x != 10) x=x+1;|.
-This means that, when using widenings, being locally more precise (for instance, by using more expressive abstract domains, or widening operators that climb more slowly \cite{BagnaraHRZ05SCP}) does not necessarily translate into higher final precision.
-It is almost always possible to concoct examples where a supposedly more precise abstraction (e.g.
-polyhedra) yields less precise results than a less precise one (e.g.
-intervals), but are these examples representative of real programs?
-This justifies our recourse to evaluation on realistic examples~(\S~\ref{sec:compare_techniques}).
-
-\subsection{SMT-solving}
-Boolean satisfiability (SAT) is the canonical NP-complete problem: given a propositional formula (e.g.
-$(a \lor \neg b) \land (\neg a \lor b \lor \neg c)$), decide whether it is satisfiable --- and, if so, output a satisfying assignment.
-Despite an exponential worst-case complexity, the DPLL algorithm \cite{Kroening_Strichman_08,Handbook_SAT} solves many useful SAT problems in practice.
-
-SAT was extended to \emph{satisfiability modulo theory} (SMT): in addition to propositional literals, SMT formulas admit atoms from a theory.
-For instance, the theories of linear integer arithmetic (LIA) and linear real arithmetic (LRA) have atoms of the form $a_1 x_1 + \dots + a_n x_n \bowtie C$ where $a_1,\dots,a_n,C$ are integer constants, $x_1,\dots,x_n$ are variables (interpreted over $\ZZ$ for LIA and $\RR$ or $\QQ$ for LRA), and $\bowtie$ is a comparison operator $=,\neq,<,\leq,>,\geq$.
-Satisfiability for LIA and LRA is NP-complete, yet tools based on DPLL(T) approach \cite{Kroening_Strichman_08,Handbook_SAT} solve many useful SMT problems in practice.
-All these tools provide a \emph{satisfying assignment} if the problem is satisfiable.
-
-Most SMT solvers, including Z3,
-%\footnote{\url{http://research.microsoft.com/en-us/um/redmond/projects/z3/}}
-Yices,%
-%\footnote{\url{http://yices.csl.sri.com/}}
-and all those supporting the full SMTLIB2 standard \cite{BarST-SMTLIB},
-offer an \emph{incremental} interface: the client program specifies the formula as an initially empty conjunction, to which additional constraints are added, and calls a ``check'' function answering whether it is satisfiable;
-it may then backtrack some of the constraints and add other ones without restarting from scratch.
-
-\subsection{A simple, motivating example}
-\label{subsec:rate_lim}
-Consider the following program, adapted from \cite{Monniaux_Gonnord_SAS11}, where \lstinline|input($a$, $b$)| stands for a nondeterministic input in $[a,b]$:
-\lstinputlisting{rlim_int.c}
-This program implements a construct commonly found in control programs (in e.g.
-automotive or avionics): a rate limiter.
-For the sake of simplicity, we chose it to be fed a nondeterministic input clamped between $[-100000,100000]$, but in a real system it would be integrated in a reactive control loop and its input connected to a complex system with unknown output range.
-
-The expected inductive invariant is $\lstinline|x_old| \in [-100000,100000]$, but classical abstract interpretation using intervals (or octagons or polyhedra) finds $\lstinline|x_old| \in (-\infty,+\infty)$~\citep{ASTREE_ESOP05}.
-Let us briefly see why.
-
-Widening iterations converge to $\lstinline|x_old| \in (-\infty,+\infty)$; let us now see why decreasing iterations fail to recover the desired invariant.
-The \lstinline|x > x_old+10| test at line~6, if taken, yields $\lstinline|x_old|
-\in (-\infty,99990)$; followed by \lstinline|x = x_old+10|, we obtain
-$\lstinline|x| \in (-\infty,\allowbreak 100000)$, and the same after union with the no-operation ``else'' branch. Line~7 yields $\lstinline|x| \in (-\infty,+\infty)$.
-
-We could use ``widening up to'' or ``widening with thresholds'', propagating the ``magic values'' $\pm 100000$ associated to \lstinline|x| into \mbox{\lstinline|x_old|,} but these syntactic approaches cannot directly cope with programs for which $\lstinline|x|  \in [-100000,+100000]$ is itself obtained by analysis.
-The guided static analysis of \citet{DBLP:conf/sas/GopanR07} does not perform
-better, and also obtains $\lstinline|x_old| \in (-\infty,+\infty)$.
-
-In contrast, let us distinguish all four possible execution paths through the tests at lines 6 and~7. The path through both ``else'' branches is infeasible; the program is thus equivalent to:
-\lstinputlisting{rlim_int2.c}
-
-\begin{figure}
-\centering
-\begin{minipage}[c]{.19\textwidth}
-\begin{tikzpicture}[->,>=stealth',auto,node distance=1.1cm,
-                    semithick,font=\footnotesize]
-
-	\node[PRstate] (n0) {$p$};
-	\node[state] (n1) [below left of=n0] {};
-	\node[state] (n2) [below right of=n0] {};
-	\node[state] (n3) [below right of=n1] {};
-	\node[state] (n4) [below left of=n3] {};
-	\node[state] (n5) [below right of=n3] {};
-	\node[state] (n6) [below right of=n4] {};
-	\node[state] (n7) [below left of=n6] {};
-	\node[state] (n8) [below right of=n6] {};
-	\node[PRstate] (n9) [below right of=n7] {$q$};
-
-  \path [transition] 
-		(n0) edge              node {} (n1);
-  \path [transition] 
-		(n0) edge              node {} (n2);
-  \path [transition] 
-		(n1) edge              node {} (n3);
-  \path [transition] 
-		(n2) edge              node {} (n3);
-  \path [transition] 
-		(n3) edge              node {} (n4);
-  \path [transition] 
-		(n3) edge              node {} (n5);
-  \path [transition] 
-		(n4) edge              node {} (n6);
-  \path [transition] 
-		(n5) edge              node {} (n6);
-  \path [transition] 
-		(n6) edge              node {} (n7);
-  \path [transition] 
-		(n6) edge              node {} (n8);
-  \path [transition] 
-		(n7) edge              node {} (n9);
-  \path [transition] 
-		(n8) edge              node {} (n9);
-\end{tikzpicture}
-\end{minipage} 
-$\Longrightarrow$
-\begin{minipage}[c]{.19\textwidth}
-\begin{tikzpicture}[->,>=stealth',auto,node distance=1.1cm,
-                    semithick,font=\footnotesize]
-
-	\node[PRstate] (n0) {$p$};
-	\node (n1) [below left of=n0] {};
-	\node (n2) [below right of=n0] {};
-	\node (n3) [below right of=n1] {};
-	\node (n4) [below left of=n3] {};
-	\node (n5) [below right of=n3] {};
-	\node (n6) [below right of=n4] {};
-	\node (n7) [below left of=n6] {};
-	\node (n8) [below right of=n6] {};
-	\node[PRstate] (n9) [below right of=n7] {$q$};
-
-  \path [transition] 
-		(n0) edge  [out=0, in=0]            node {} (n9);
-  \path [transition] 
-		(n0) edge  [out=180, in=180]        node {} (n9);
-  \path [transition] 
-		(n0) edge  [out=205, in=155]            node {} (n9);
-  \path [transition] 
-		(n0) edge  [out=230, in=130]            node {} (n9);
-  \path [transition] 
-		(n0) edge  [out=255, in=105]            node {} (n9);
-  \path [transition] 
-		(n0) edge  [out=-25, in=25]            node {} (n9);
-  \path [transition] 
-		(n0) edge  [out=-50, in=50]            node {} (n9);
-  \path [transition] 
-		(n0) edge  [out=-75, in=75]            node {} (n9);
-\end{tikzpicture}
-\end{minipage}
-\caption{Expansion of the transition graph into a multigraph.}
-\label{fig:multigraph}
-\end{figure}
-
-Classical interval analysis on this program yields $\lstinline|x_old| \in [-100000,100000]$.
-We have transformed the first program into the second, manually pruning out infeasible paths; yet in general the resulting program could be exponentially larger than the first (as in Fig.~\ref{fig:multigraph}), even though not all feasible paths are needed to compute the invariant.
-
-Following recent suggestions \cite{Gawlitza_Monniaux_ESOP11,Monniaux_Gonnord_SAS11}, we avoid this space explosion by keeping the second program implicit while simulating its analysis.
-This means we work on an implicitly represented transition multigraph (Fig.~\ref{fig:multigraph}); it is succinctly representated by the transition graph of the first program.
-Our first contribution (\S\ref{sec:guided_multigraph}) is to recast the ``guided analysis'' from \citet{DBLP:conf/sas/GopanR07} on such a succinct representation of the paths in lieu of the individual transitions.
-A similar explosion occurs in disjunctive invariant generation, following \citet{DBLP:conf/pldi/GulwaniZ10}; our second contribution (\S\ref{sec:disjunctive}) applies our implicit representation to their method.
-
-\subsection{Guided static analysis}
-\label{sec:guided}
-\emph{Guided static analysis} was proposed by \citet{DBLP:conf/sas/GopanR07} as an improvement over classical upward Kleene iterations with widening.
-Consider the following program, taken from \cite{DBLP:conf/sas/GopanR07}:
-\lstinputlisting[label=lst:gopan_reps]{gopan_reps.c}
-
-\begin{figure}
-\begin{center}
-\includegraphics[scale=0.75]{figures/gopan_reps}
-\end{center}
-\caption{The invariant for Prog.~\ref{lst:gopan_reps}: the piecewise linear, solid line is the strongest invariant, the grayed polyhedron is its convex hull.}
-\label{fig:gopan_reps_invariant} 
-\end{figure}
-
-Classical iterations on the domain of convex polyhedra \cite{CousotHalbwachs78,BagnaraHRZ05SCP} or octagons \cite{DBLP:journals/lisp/Mine06} start with $x = 0 \land x = 0$, then continue with $x = y \land 0 \leq x \leq 1$.
-The widening operator extrapolates from these two iterations and yields $x = y \land x \geq 0$.
-From there, the ``else'' branch at line 5 may be taken; with further widening, $0 \leq y \leq x$ is obtained as a loop invariant, and thus the postcondition computed at line 9 is $x \geq 0 \land y = 0$.
-Yet the strongest invariant is $(0 \leq x \leq 51 \land y = x) \lor (51 \leq x \leq 102 \land x+y=102)$, and its convex hull, a convex polyhedron (Fig.~\ref{fig:gopan_reps_invariant}), is
-\begin{equation}
-y \leq x \land y \leq 102-x \land y \geq 0.\label{eqn:triangle}
-\end{equation}
-Intuitively, this disappointing result is obtained because widening extrapolates from the first iterations of the loop, but the loop has two different phases ($x \leq 50$ and $x > 50$) with different behaviors, thus the extrapolation from the first phase is not valid for the second.
-
-\citeauthor{DBLP:conf/sas/GopanR07}' idea is to analyze the first phase of the loop with a widening and narrowing sequence, and thus obtain $0 \leq x \leq 50 \land y = x$, and then analyze the second phase, finally obtaining invariant~\ref{eqn:triangle}; each phase is identified by the tests taken or not taken.
-
-The analysis starts by identifying the test taken and not taken during the first iteration of the loop, starting in the loop initialization.
-The branches not taken are pruned from the loop body, yielding:
-\begin{lstlisting}[numbers=none]
-  while(1) {
-    if(x <= 50) y++;
-    else break; /* not taken in phase 1 */
-    if(y < 0) break;
-    x++;
-  }
-\end{lstlisting}
-
-Analyzing this loop using widening and narrowing on convex polyhedra or octagons yields the loop invariant $0 \leq x \leq 51 \land y = x$.
-Now, the transition at line~5 becomes feasible; and we analyze the full loop, starting iterations from $0 \leq x \leq 51 \land y = x$, and obtain invariant~(\ref{eqn:triangle}).
-
-More generally, this analysis method considers an ascending sequence of subsets of the transitions in the loop body (left side of Fig.~\ref{fig:multigraph});
-for each subset, an inductive invariant is computed for the program restricted to it.
-The starting subset are the transitions reachable in one step from the loop initialization.
-If for a given subset $S$ in the sequence, no transitions outside $S$ are reachable from the inductive invariant attached to $S$, then iterations stop;
-otherwise, add these transitions to $S$ and iterate more.
-Termination ensues from the finiteness of the control-flow graph.
-
-\subsection{Path-focusing}
-\label{sec:path_focusing}
-
-\citet{Monniaux_Gonnord_SAS11}'s \emph{path-focusing} technique
-distinguishes the different paths in the program in order to avoid loss of
-precision due to merge operations. Since the number of paths may be exponential,
-the technique keeps them implicit and computes them when needed using
-SMT-solving.
-The (accelerated) Kleene iterations (\S\ref{sec:static_analysis}) are computed over a reduced multigraph instead of the classical transition graph.
-
-Let $P$ be the set of control points in the transition graph, 
-$P_W \subseteq P$ the set of widening points
-such that removing the points in $P_W$ gives an acyclic graph.
-One can choose a set $P_R$ such that $P_W \subseteq P_R \subseteq P$.
-
-The set of paths is kept implicit by an SMT formula $\rho$ expressing
-the semantics of the program, assuming that the transition semantics can be
-expressed within a decidable theory. For an easy construction of $\rho$, 
-we also assume that the program is expressed in SSA form, meaning that each
-variable is only assigned once in the transition graph. This is not a
-restriction, since there exist standard algorithms that transform a program into
-an SSA format.
-
-This formula contains Boolean \emph{reachability predicates} $b_i$ for each
-control points $p_i \notin P_R$, $b_i^s$ and $b_i^d$ for each $p_i \in P_R$, so
-that a path 
-$p_{i_1} \rightarrow p_{i_2} \rightarrow \dots \rightarrow p_{i_n}$ 
-between two points $p_{i_1}, p_{i_n} \in P_R$ 
-can easily be expressed as the
-conjunction $b_{i_1}^s \wedge \bigwedge_{2 \leq k < n} b_{i_k} \wedge b_{i_n}^d$.
-The Boolean $b_{i}^s$ is $true$ when the path starts at point $p_i$, whereas
-$b_i^d$ is $true$ when the path arrives at $p_i$. In other words, we split the
-points in $P_R$ into a \emph{source} point, with only outgoing transitions, and
-a \emph{destination} point, with only incoming transitions, so that the
-resulting graph is acyclic and there is no paths going through control
-points in $P_R$.
-
-In order to find focus paths, we solve an SMT formula which is satisfiable when
-there exist a path starting at a point $p_i \in P_R$ in a state included in the
-current invariant candidate $X_i$, and arriving at a point $p_j \in P_R$ in a
-state outside $X_j$. In this case, we construct this path using the model and
-update $X_j$. When $p_i = p_j$, meaning that the path is actually a self-loop,
-we can apply a widening/narrowing sequence, or even compute the transitive
-closure of the loop (or an approximation thereof, or its application to $X_i$)
-using abstract acceleration~\cite{DBLP:conf/sas/GonnordH06}.
-
-\section{Guided analysis over the paths}
-\label{sec:guided_multigraph}
-
-Guided static analysis, as proposed by \citet{DBLP:conf/sas/GopanR07}, applies to the transition
-graph of the program. We now present a new technique applying this analysis on the implicit
-multigraph from \cite{Monniaux_Gonnord_SAS11}, thus avoiding control flow merges with
-unfeasible paths.
-In this section, we use the same notations as \ref{sec:path_focusing}, except
-that we call the abstract values $X_i^s$ instead of $X_i$ to leave room for
-another abstract value $X_i^d$ later in the algorithm.
-
-The combination of these two techniques aims at first discovering a precise
-inductive invariant for a subset of paths between two points in $P_R$, 
-by the mean of ascending and narrowing iterations. When an
-inductive invariant has been found, we add new feasible paths to the subset and
-compute an inductive invariant for this new subset, starting with the results
-from the previous analysis.
-In other words, our technique considers an ascending sequence of
-subsets of the paths between two points in $P_R$.
-We iterate the operations until the whole program (i.e all the
-feasible paths) has been considered. The result will then be an
-inductive invariant of the entire program.
-
-\subsection{Algorithm}
-
-Algorithm~ \ref{algo:combined} performs Guided static analysis on the implicitly represented multigraph.
-
-The current working subset of paths, noted $P$ and initially empty, is
-stored using a compact representation, such as binary decision
-diagrams. We also maintain two sets of control points:
-\begin{itemize}
-	\item $A'$ contains the points in $P_R$ that may be the starting points of new
-		feasible paths.
-	\item $A$ contains the points in $P_R$ on which we apply the ascending iterations.
-	Each time the abstract value of a control point $p$ is updated, $p$ is
-	inserted in both $A$ and $A'$.
-\end{itemize}
-
-\begin{algorithm}
-	\caption{Guided static analysis on implicit multigraph}
-	\label{algo:combined}
-	\begin{algorithmic}[1] 
-	\input{algo/algo-combined.tex}
-	\end{algorithmic}
-\end{algorithm}
-
-We distinguish three phases in the main loop of the analysis:
-\begin{enumerate}
-\item \label{step:addingpaths} We start finding a new relevant subset
-  $P \cup P'$ of the graph.
-  Either the previous iteration or the initialization lead us to a
-  state where there are no more paths in the previous subset $P$,
-  starting at $p_i$, that make the abstract values of the successors
-  grow (otherwise, the SMT solver would not have answered
-  ``\emph{unsat}''). Narrowing iterations preserve this property.
-  However, there may exist such paths in the entire multigraph, that
-  are not in $P$. This phase computes these paths and adds them to
-  $P'$. This phase is described in~\ref{subsec:addingpaths}
-  and corresponds to lines in~\ref{alg=start-add-paths} to
-  \ref{alg=end-add-paths} in Algorithm~\ref{algo:combined}.
-\item \label{step:ascending} Given a new subset $P$, we search for paths starting at point
-  $p_i \in P_R$, such that these paths are in $P$, i.e are included in
-  the working subgraph. Each time we find a path, we update the
-  abstract value of the destination point of the path. This is the
-  phase explained in~\ref{subsec:ascending}, and corresponds to
-  lines~\ref{alg=start-ascending} to~\ref{alg=end-ascending} in
-  Algorithm~\ref{algo:combined}.
-\item  \label{step:narrowing} We perform narrowing iterations the usual way
-  (line~\ref{alg=narrowing} in algorithm~\ref{algo:combined}) and
-  reiterate from step 1 unless there are no more points to explore,
-  i.e. $A' = \emptyset$.
-\end{enumerate}
-
-%\MM{Je trouve qu'il manque une vue de haut niveau quelque part par
-%  ici. Dire qu'en gros, l'algo, c'est (en pseudo-code ou en langue naturelle) :
-%
-%Sans ca, enchainer tout de suite sur ``ascending iterations'' est un peu brutal.
-%
-%Ensuite, on peut rafiner la définition de ``Subset'' ($A$ et $P$), et
-%enchainer sur l'algo 2.
-%}
-
-Step~\ref{step:ascending} corresponds to the application of
-path-focusing \cite{Monniaux_Gonnord_SAS11} to
-the elements of $A$ until $A$ is empty. 
-When $A$ becomes empty, it means the
-invariant computation of the subgraph is finished. As proposed by
-\citet{DBLP:conf/sas/GopanR07}, we can do some narrowing iterations.
-These narrowing iterations allow to recover precision lost by
-widening, \emph{before} computing and taking into account new feasible paths.
-Thus, our technique combines both the advantages of \emph{Guided Static
-Analysis} and \emph{Path-focusing}.
-
-The order of steps is important: narrowing has to be performed before
-adding new paths, or some spurious new paths would be added to $P$,
-and starting with the addition of new paths avoids wasting time
-doing the ascending iterations on an empty graph.
-
-\subsection{Ascending iterations by Path-focusing}
-\label{subsec:ascending}
-
-\MM{Maintenant que l'ordre a changé dans l'algo, on peut peut-être
-  échanger cette section et la suivante ?}
-
-For computing an inductive invariant over a subgraph, we use the
-Path-focusing algorithm from \citet{Monniaux_Gonnord_SAS11} with special
-treatment for self loops (line~\ref{alg=pf} in algorithm~\ref{algo:combined}).
-
-In order to find which path to focus on, we construct an SMT formula $f(p_i)$, whose
-model when satisfiable is a path that starts in $p_i$, goes to a successor $p_j
-\in P_R$ of $p_i$, such that the image of $X_{i}^s$ by the path transformation
-is not included in the current $X_{j}^s$.
-Intuitively, such a path makes the abstract value $X_{j}^s$ grow, and thus is
-an interesting path to focus on. We loop until the formula becomes unsatisfiable,
-meaning that the analysis of $p_i$ is finished.
-
-If we note $Succ(i)$ the set of indices $j$ such that $p_j \in P_R$ is a
-successor of $p_i$ in the expanded multigraph, and $X_i^s$ the abstract value
-associated to $p_i$ :
-$$f(p_i) = \rho \wedge b_i^s \wedge 
-\displaystyle\bigwedge_{j \in P_R \atop j \neq i} \neg
-b_j^s \wedge X_i^s \wedge \displaystyle\bigvee_{j \in Succ(i)} (b_j^d \wedge
- \neg X_j^s)$$
-
-The difference with~\citet{Monniaux_Gonnord_SAS11} is that we do not
-work on the entire transition graph but on a subset of it. Therefore we
-conjoin the formula $f(p_i)$ with the actual set of working paths,
-noted $P$, expressed as a Boolean formula, where the Boolean variables are the
-\emph{reachability predicates} of the control points. We can easily construct
-this formula from the binary decision diagram using dynamic programming, and
-avoiding an exponentially sized formula. In other words, we force the SMT solver
-to give us a path included in $P$.
-
-\subsection{Adding new paths}
-\label{subsec:addingpaths}
-
-Our technique computes the fixpoint iterations on a ascending sequence of
-subgraphs, until the complete graph is reached.
-When the analysis of a subgraph is finished, meaning that the abstract values
-for each control points has converged to an inductive invariant for this subgraph,
-the next subgraph to work on has to be computed.
-
-This new subgraph is the union of the paths of the previous one with a set
-$P'$ of new paths that become feasible regarding the current abstract values.
-The paths in $P'$ are computed one after another, until no more path
-can make the invariant grow. This is line~\ref{alg=computeNewPaths} in
-Algorithm~\ref{algo:combined}, which corresponds to
-Algorithm~\ref{algo:computepaths}. We also use SMT solving to discover
-these new paths, but we subtly change the SMT formula given to the
-SMT solver: we now simply check for the satisfiability of $f(p_i)$
-instead of $f(p_i) \wedge P$. Since we already know that $f(p_i)
-\wedge P$ is unsatisfiable, none of the paths given by the SMT solver
-will be in $P$.
-
-% À changer si on décide de ne plus distinguer $X^s$ et $X^d$
-Also, to prevent the algorithm from adding too many paths at a time,
-and in particular to prevent the algorithm from trying to re-add the
-same path infinitely many times, we use another abstract value
-associated to the control point $p_j$, noted $X_j^d$, which is distinct
-from $X_j^s$, and initialized to $X_j^s$ right before computing new paths. 
-In the SMT formula, we associate to $p_j$ the abstract value $X_j^s$ when $p_j$
-is the starting point of the path, and $X_j^d$ when it is its destination point.
-We thus check the satisfiability of the formula
-$f'(p_i)$, where:
-$$f'(p_i) = \rho \wedge b_i^s \wedge 
-\displaystyle\bigwedge_{j \in P_R \atop j \neq i} \neg
-b_j^s \wedge X_i^s \wedge \displaystyle\bigvee_{j \in Succ(i)} (b_j^d \wedge
-\neg X_j^d)$$
-
-$X_j^d$ is updated when the point $p_j$ is the target of a new path.
-This way, further SMT queries do not
-compute other paths with the same source and destination if it is not
-needed (because these new paths would not make $X_j^d$ grow, hence
-would not be returned by the SMT solver).
-
-\begin{algorithm}
-	\caption{ComputeNewPaths}
-	\label{algo:computepaths}
-	\begin{algorithmic}[1] 
-	\input{algo/compute-paths.tex}
-	\end{algorithmic}
-\end{algorithm}
-
-\subsection{Termination}
-Termination of this algorithm is guaranteed, because:
-\begin{itemize}
-\item 
-the subset of paths $P$ strictly increases at each loop iteration, and bounded by the finite set of
-paths in the entire graph. 
-\item the set $P'$ always verifies $P \cap P' = \emptyset$ by construction, 
-which guarantees that $P'$
-will eventually be empty after a finite number of loop iterations.
-\end{itemize}
-
-\subsection{Example}
-
-We revise the rate limiter described in \ref{subsec:rate_lim}. In this example,
-\emph{Path-focusing} works well because all the paths starting at the
-loop header are actually self loops. In such a case, the technique performs a
-widening/narrowing sequence or accelerates the loop, thus leading to a precise
-invariant. However, in some cases, there also exist paths that are not
-self loops, in which case \emph{Path-focusing} applies widening. 
-This widening may induce unrecoverable loss of precision.
-
-Suppose the main loop of the rate limiter contains a nested loop:
-\lstinputlisting{rlim_int_loop.c}
-
-We choose $P_R$ as the set of loop headers of the function, plus the initial
-state. In this case, we have three elements in $P_R$.
-We also unroll the loop once, in order to distinguish the paths that do not go
-through the loop from the others:
-\begin{lstlisting}[numbers=none]
-if (wait()) {
-   while (wait()) {}
-}
-\end{lstlisting}
-
-The main loop in the expanded multigraph has then 8 distinct paths, half of
-them being self loops, and the other half going to the header of the nested loop.
-
-Guided static analysis from \citet{DBLP:conf/sas/GopanR07} yields, at line 5,
-$\lstinline|x_old| \in (-\infty,+\infty)$.
-Path-focusing \citep{Monniaux_Gonnord_SAS11} performs slightly
-better, and finds $\lstinline|x_old| \in (-\infty,10000]$.
-Now, let us see how our technique performs on this example.
-
-Figure \ref{fig:example-graph} shows the sequence of subset of paths during the
-analysis. The points in $P_R$ are noted $p_i$, where $i$ is the corresponding
-line in the code: for instance, $p_5$ corresponds to the header of the main
-loop.
-
-\begin{enumerate}
-	\item The first subset of paths is depicted on Figure
-		\ref{fig:example-graph} Step 1. We apply Path-Focusing on this graph,
-		and discover for $p_5$ the inductive invariant $\lstinline|x_old| = 0$.
-	\item
-		We then compute the set $P'$ of paths that have to be added into the
-		next subgraph. 
-		The image of $\lstinline|x_old| = 0$ by the path that goes from $p_5$ to
-		itself, and that goes through the \emph{else} branch of each 
-		\emph{if-then-else}, is $-10 \leq \lstinline|x_old| \leq 10$. This path
-		is then added to our subgraph. Moreover, there is no other path whose
-		image is not in $-10 \leq \lstinline|x_old| \leq 10$.
-		Additionally, the path that goes from $p_5$ to
-		$p_{10}$, that goes through the \emph{else} branch of each 
-		\emph{if-then-else} at lines 7,8 and 9, is also 
-		$-10 \leq \lstinline|x_old| \leq 10$, and thus is also added into the
-		new subgraph. Again, there is no new path that goes to $p_{10}$ with an
-		image outside $-10 \leq \lstinline|x_old| \leq 10$.
-	\item
-		We apply Path Focusing on this new subgraph (\ref{fig:example-graph}
-		Step 2). When we have reached an inductive invariant, we narrow, and the
-		result for both $p_5$ and $p_{10}$ is the polyhedron 
-		$-10000 \leq \lstinline|x_old| \leq 10000$.
-	\item Finally, we compute the new subgraph. The SMT-solver does not find any
-		new path that makes the abstract values grow, and the algorithm
-		terminates.
-\end{enumerate}
-
-Our technique gives us the expected invariant 
-$\lstinline|x_old| \in~[-10000,\allowbreak 10000]$. 
-Here, only 3 paths out of the 8 have been computed during the analysis. In
-practice, depending on the order the SMT-solver returns the paths, other
-feasible paths could have been added during the analysis.
-
-\begin{figure}
-\label{fig:example-graph}
-\centering
-\begin{minipage}[c]{.15\textwidth}
-\begin{tikzpicture}[->,>=stealth',auto,node distance=2.1cm,
-                    semithick,font=\footnotesize]
-
-	\node[PRstate] (n00) {$p_4$};
-	\node[PRstate] (n0) [below of=n00, yshift=0.8cm] {$p_5$};
-	\node[PRstate] (n1) [below of=n0] {$p_{10}$};
-	\node (label) [below of=n1, yshift=1.2cm] {Step 1};
-
-  \path [transition] 
-		(n00) edge  node {$\lstinline|x_old| \gets 0$} (n0);
-\end{tikzpicture}
-\end{minipage} 
-\begin{minipage}[c]{.28\textwidth}
-\begin{tikzpicture}[->,>=stealth',auto,node distance=2.1cm,
-                    semithick,font=\footnotesize]
-
-	\node[PRstate] (n00) {$p_4$};
-	\node[PRstate] (n0) [below of=n00, yshift=0.8cm] {$p_5$};
-	\node[PRstate] (n1) [below of=n0] {$p_{10}$};
-	\node (label) [below of=n1, yshift=1.2cm] {Step 2};
-
-  \path [transition] 
-		(n00) edge  node {$\lstinline|x_old| \gets 0$} (n0);
-  \path [transition] 
-		(n0) edge  [loop left] node [left, xshift=0.2cm] {
-		$\begin{array}{r}
-			-10000 \leq \lstinline|x| \leq 10000 \\
-			\lstinline|x_old|-10 \leq \lstinline|x| \leq \lstinline|x_old|+10 /\\
-			\lstinline|x_old| \gets \lstinline|x|
-		\end{array}$
-		} (n0);
-  \path [transition] 
-		(n0) edge node [left, yshift=-0.2cm, xshift=0.2cm] {
-		$\begin{array}{r}
-			-10000 \leq \lstinline|x| \leq 10000 \\
-			\lstinline|x_old|-10 \leq \lstinline|x| \leq \lstinline|x_old|+10 /\\
-			\lstinline|x_old| \gets \lstinline|x|
-		\end{array}$
-		} (n1);
-\end{tikzpicture}
-\end{minipage}
-\caption{Ascending sequence of subgraphs}
-\end{figure}
-
-In this example, we see that our technique actually combines best of
-\emph{Guided Static Analysis} and \emph{Path Focusing}. Still, we need to unroll
-the loop once in order to get precise results. Section \ref{sec:disjunctive}
-proposes a new technique that gives precise invariants without needing such
-graph transformations.
-
-\section{Disjunctive invariants}
-\label{sec:disjunctive}
-
-\citet{DBLP:conf/pldi/GulwaniZ10} proposed a technique for computing disjunctive invariants, by
-distinguishing all the paths inside a loop. In
-this section, we propose to improve this technique by using SMT queries to find
-interesting paths, the objective being to avoid an explicit exhaustive
-enumeration of an exponential number of paths.
-
-For each control point $p_i$, we compute a disjunctive invariant
-$\bigvee_{1\leq j \leq m_i} X_{i,j}$. We denote by $n_i$ the number of
-distinct paths starting at $p_i$.
-To perform the analysis, one chooses an integer $\delta_i \in [1,m_i]$, and
-a mapping function $\sigma_i: [1,m_i] \times [1,n_i] \mapsto [1,m_i]$.
-  The $k$-th path starting fom $p_i$ is denoted $\tau_{i,k}$.
-  The image of the $j$-th disjunct $X_{i,j}$ by the path $\tau_{i,k}$ is then
-  joined with $X_{i,\sigma(j,k)}$.
-Initially, the $\delta_i$-th abstract value contains the initial states of
-$p_i$, and all other abstract values contain~$\emptyset$.
-
-For each control point $p_i \in P_R$, $m_i$, $\delta_i$ and $\sigma_i$ can be defined heuristically.
-For instance, one could define $\sigma_i$ so that $\sigma_i(j,k)$ only depends on the
-last transition of the path, or else construct it dynamically during the
-analysis.
-
-Our method improves this technique in two ways :
-\begin{itemize}
-\item Instead of enumerating the whole set of paths, we keep them implicit and
-compute them only when needed.
-
-\item At each loop iteration of the original algorithm \citep{DBLP:conf/pldi/GulwaniZ10}, an image by each path inside the loop is computed for each disjunct of the invariant candidate.
-Yet, many of these images may be redundant: for instance, if our invariant candidate is $(0 \leq x \leq 10 \land 0 \leq y \leq 1000) \lor (x < -10 \land y < -10)$, then there is no point enumerating paths whose image is included in this invariant candidate.
-In our approach, we compute such an image only if it makes the resulting abstract value grow.
-\end{itemize}
-
-Our improvement consists in a modification of the SMT formula we solve in
-\ref{sec:guided_multigraph}.
-We introduce in this formula Boolean variables $\{d_j, 1 \leq j \leq m\}$, so
-that we can easily find in the model which abstract value of the
-disjunction of the source point has
-to be chosen to make the invariant of the destination grow.
-The resulting formula that is given to the SMT solver is defined
-by $g(p_i)$.
-When the formula is satisfiable, we know that the index $j$ of the starting
-disjunct that has to be chosen is the one for which the associate Boolean value
-$d_j$ is \emph{true} in the model. Then, we can easily compute the value of 
-$\sigma_i(j,k)$, thus know the index of the disjunct to join with.
-
-\begin{eqnarray*}
-g(p_i) & = & \rho \wedge b_i^s \wedge 
-\displaystyle\bigwedge_{j \in P_R \atop j \neq i} \neg b_j^s  \\
- & & \wedge 
-\displaystyle\bigvee_{1 \leq k \leq m_i} (d_k \wedge X_{i,k} \wedge \bigwedge_{l \neq k}
-\neg d_l) \\
- & & \wedge
-\displaystyle\bigvee_{j \in Succ(i)} 
-(b_j^d \wedge \bigwedge_{1 \leq k \leq m_i} (\neg X_{j,k}))
-\end{eqnarray*}
-
-In our algorithm, the initialization of the abstract values slightly differs from
-algorithm \ref{algo:combined} line~\ref{alg=X-init}, since we now have to
-initialize each disjunct. Line~\ref{alg=X-init} is then replaced by:
-\begin{algorithm}[!h]
-\begin{algorithmic}[1] 
-\FORALL {$k \in \{1,..,m_i\} \setminus \{\delta_i\}$}
-	\STATE $X_{i,k} \gets \perp$
-\ENDFOR
-\STATE $X_{i,\delta_i} \gets I_{p_i}$
-\end{algorithmic}
-\end{algorithm}
-
-Furthermore, the Path-focused algorithm (line~\ref{alg=pf} from algorithm
-\ref{algo:combined}) is enhanced to deal with disjunctive invariants, and
-is detailed in algorithm \ref{algo:disjunctive}.
-
-The \emph{Update} function can classically assign to $X_{i,\sigma_i(j,k)}$ the
-value $X_{i,\sigma_i(j,k)} \widening (X_{i,\sigma_i(j,k)} \sqcup
-\tau_{i,k}(X_{i,j}))$, or can integrate the special treatment for self loops
-proposed by \citet{Monniaux_Gonnord_SAS11}, with widening/narrowing sequence or
-acceleration.
-
-\begin{algorithm}[!h]
-\caption{Disjunctive invariant computation with implicit paths}\label{gulwani2}
-\label{algo:disjunctive}
-\begin{algorithmic}[1] 
-\WHILE {true}
-	\STATE $res \gets SmtSolve\left[g(p_i)\right]$
-	\IF {$res = unsat$}
-		\STATE \textbf{break}
-	\ENDIF
-	\STATE Compute the path $\tau_{i,k}$ from $res$ 
-	\STATE Take $j \in \{ l | d_l = true\}$ 
-	\STATE Update($X_{i,\sigma_i(j,k)})$
-\ENDWHILE
-\end{algorithmic}
-\end{algorithm}
-
-We experimented with a heuristic
-of dynamic construction of the $\sigma_i$ functions, adapted from
-\citet{DBLP:conf/pldi/GulwaniZ10}. 
-\DM{Cette heuristique, je croyais que c'était nous qui l'avions trouvée,
-et qu'elle n'était pas dans Gulwani?}
-\JH{Non. Gulwani fait la même chose en un peu plus complique (voir commentaire
-suivant)}
-\MM{Du coup, la contribution~\ref{contr:disjunctive} devrait être
-  affaiblie un peu je suppose. Gulwani fait aussi du SMT-solving ?}
-\DM{Non, Gulwani énumère tous les chemins un par un à tous les coups.}
-\MM{OK, du coup j'ai ajouté une phrase à la fin de la section pour
-  rappeler notre contrib}
-For each control point $p_i \in P_R$, we start with one single disjunct
-($m_i = 1$) and define $\delta_i = 1$.
-$M$ denotes an upper bound on the number of disjuncts per control point.
-
-The $\sigma_i$ functions take as parameters the index of the starting abstract
-value, and the path we focus on. 
-Since we dynamically construct these functions
-during the analysis, we store their already computed image into a
-compact representation, such as Algebraic Decision Diagrams.
-$\sigma_i(j,k)$ is then constructed on the fly only when needed, and computed
-only once.
-When the value of $\sigma_i(j,k)$ is required but undefined, we first compute
-the image of the
-abstract value $X_{i,j}$ by the path indexed by $k$, and try to find an existing
-disjunct of index $j'$ so that the least upper bound of the two abstract values is exactly their union (this can be tested using SMT-solving).
-If such an index exists, then we set $\sigma_i(j,k) = j'$.
-Otherwise:
-\begin{itemize}[itemsep=-1\parsep,topsep=0pt]
-	\item if $m_i < M$, we increase $m_i$ by $1$ and define $\sigma_i(j,k) = m_i$
-	\item if $m_i = M$, we define $\sigma_i(j,k) = M$ 
-		%\JH{A la place, Gulwani cherche l'invariant qui a le plus de contraintes
-		%d'egalite en commun pour definir son $sigma$. Nous, on le met dans M, pour simplifier}
-\end{itemize}
-The main difference with the original
-algorithm~\cite{DBLP:conf/pldi/GulwaniZ10} is that we construct
-$\sigma_i(j,k)$ using SMT queries instead of enumerating a possibly
-exponential number of paths to find a solution.
-
-\section{Experimental comparisons}
-\label{sec:experiments}
-
-We have implemented our proposed solutions inside a prototype intraprocedural
-static analyzer,
-as well as the classical abstract interpretation algorithm, and the state-of-the-art
-techniques \emph{Path Focusing} \cite{Monniaux_Gonnord_SAS11} and \emph{Guided
-Static Analysis} \cite{DBLP:conf/sas/GopanR07}.
-
-%\MM{On peut peut-être citer le nom de l'outil (on garde PAGAI ?) ici.
-%  L'intérêt d'avoir un nom, c'est que les gens peuvent en parler plus
-%  facilement, et c'est plus facile de retrouver le papier origine d'un
-%  outil quand il cite son nom}
-%\DM{Dans la version finale, oui, mais pas dans une soumission anonymisée... suffit de Googlifier pagai sur imag.fr}
-%
-Our tool operates over LLVM bitcode \citep{LLVM_langref,Lattner:2004:LCF:977395.977673}, which is a target for several compilers, most notably Clang (supporting C and C++) and llvm-gcc (supporting C, C++, Fortran and Ada).
-Abstract domains are provided by the APRON library \citep{DBLP:conf/cav/JeannetM09}, and include convex polyhedra (from the builtin Polka ``PK'' library), octagons, intervals, and linear congruences.
-For SMT-solving, our analyzer uses Yices
-\cite{DBLP:conf/cav/DutertreM06} or Microsoft Z3\cite{DBLP:conf/tacas/MouraB08}.
-
-We conducted extensive experiments on real-life programs in order to compare the
-different techniques, mostly on open-source projects (Fig.~\ref{fig:projects}) written in C, C++ and Fortran.
-
-\begin{table}[!h]
-	\centering
-\begin{tabular}{|l|r|r|} \hline
-	\multicolumn{1}{|c|}{Name} &
-        \multicolumn{1}{c|}{kLOC} &
-        \multicolumn{1}{c|}{$|P_R|$} \\ \hline
-	a2ps & 55 & 2012\\
-	gawk & 59 & 902\\ 
-	gnuchess & 38 & 1222\\ 
-	gnugo & 83 & 2801\\
-	grep & 35 & 820\\
-	gzip & 27 & 494\\
-	lapack/blas & 954 & 16422\\
-	make & 34 & 993\\ 
-	sed & 23 & 292\\
-	tar & 73 & 1712\\
-	\hline
-\end{tabular}
-\caption{List of analyzed open-source projects, with their respective number of
-lines of code, and their number of control points in $P_R$}
-\label{fig:projects}
-\end{table}
-
-\subsection{Analysis algorithm}
-\label{sec:analysis-algorithm}
-For each program, we distinguish a set $P_R = P_W$ of suitable widening points by a simple algorithm: for each procedure, compute the strongly connected components of its control-flow graph using Tarjan's algorithm; the targets of the back-edges of the depth-first search are added to~$P_R$.
-Note that the resulting set is not necessarily minimal, but is sufficient to disconnect all cycles --- more sophisticated techniques are discussed in e.g.
-\citet{BourdonclePhd}.
-
-LLVM bitcode is in \emph{static single assignment} (SSA) form: a given scalar variable is given a value at a single syntactic point in the program.
-In concrete terms, an assignment \lstinline|x=2*x+1;| gets translated into a definition $x_2 = 2x_1+1$, with distinct variables $x_1$ and $x_2$ corresponding to the same original variable \lstinline|x| at different points in the program.
-LLVM makes it easy to follow definition-use and use-definition chains: for a given variable (say, $x_2$) one can immediately obtain its definition (say, $2x_1+1$).
-One may see conversion to SSA form as a static precomputation of some of the symbolic propagations proposed by \citet{DBLP:conf/vmcai/Mine06} to enhance the precision of analyses.
-
-SSA introduces $\phi$-functions at the head of a control code to define variables whose value depends on which incoming edge to this control node was last taken.
-For instance, for \lstinline|if (...) { x = 2*x+1; } else { x= 0; }|, then $x_2$ is defined as $\phi(2x_1+1,0)$.
-In this framework, each variable is uniquely defined as an arithmetic ($+$, $-$, $\times$, $/$) function of other variables that themselves may not be representable as affine linear functions, because they are defined using $\phi$-functions, bitwise arithmetic, loads from memory, or return values from function calls.
-
-This motivates a key implement decision of our tool: only those variables $v_1,\dots,v_n$ that are not defined by arithmetic operations are retained as coordinates in the abstract domain (e.g.
-as dimensions in polyhedra).
-A basic block of code therefore amounts to a \emph{parallel assignment} operation
-$(v_1,\dots,v_n) \allowbreak\mapsto\allowbreak
-(f_1(v_1,\dots,v_n), \allowbreak, \dots, \allowbreak
- f_n(v_1,\dots,v_n))$;
-such operations are directly supported by APRON. This has three benefits:
-(i) it limits the number of dimensions in the abstract values, since polyhedra libraries typically perform worse with higher dimensions;
-(ii) the abstract operation for a single path in path-focusing methods also is a (large) parallel assignment;
-(iii) as suggested by \citet{DBLP:conf/vmcai/Mine06}, this approach is more precise than running abstract operations for each program line separately:
-for instance, for \lstinline|y=x; z=x-y;| with precondition $x \in [0,1]$, a line-by-line interval analysis obtains $y \in [0,1]$ and $z \in [-1,1]$ while our ``en bloc'' analysis symbolically simplifies $z = x - x = 0$ and thus $z \in [0,0]$.
-
-In the event that a node is reachable only by a single control-flow edge (which may occur because of dead code, or during the first phases of guided static analysis), the $\phi$ operation reduces to a copy of the values flowing from that edge.
-In this case, our tool just propagates symbolic values through the predecessor node, without introducing $\phi$-variables.
-
-We further reduce the number of variables by projecting out those that are not live at the program point. Again, SSA use-def and def-use chains make it easy to compute liveness.
-
-Our tool currently only operates over scalar variables from the SSA representation and thus cannot directly cope with arrays or memory accessed through pointers.
-We therefore run it after the ``memory to registers'' (\texttt{mem2reg}) optimization phase in LLVM, which lifts most memory accesses to scalar variables.
-The remaining memory accesses are treated as undefined values, as are the return
-values of function calls. For better precision, we also apply function inlining
-to the program, and we unroll every loop once.
-
-Our tool currently assumes that integer variables are unbounded mathematical integers ($\ZZ$) and floating-point variables are real (or rational) numbers.
-Techniques for sound analysis of bounded integers, including with wraparound, and of floating-point operations have been developed in e.g.
-the Astr\'ee system \citep{ASTREE_ESOP05,ASTREE_PLDI03}, but porting these techniques to our iteration schemes using SMT-solving requires supplemental work.
-
-Our implementation of path-focusing currently does not use true acceleration techniques, as proposed by \citet{Monniaux_Gonnord_SAS11}.
-Instead, it simply runs widening and narrowing iterations on a single path.
-The analysis is currently only forward, even though nothing in what we describe in this article is specific to forward analysis.
-
-\subsection{Precision of the various techniques}
-\label{sec:compare_techniques}
-
-\begin{figure}[h]
-  \begin{center}
-    \input{./techniques.tex}
-  \end{center} 
-  \vspace{-20px}
-  \caption{Comparison of the abstract values obtained by classical abstract
-  interpretation (S), \emph{Guided Static
-  Analysis} (G), \emph{Path-focused} technique (PF), our combined technique
-  (G+PF), and its version with disjunctive invariants (DIS).
-  The $\subsetneq$ bars gives the percentage of invariants stronger (more precise; smaller with respect to inclusion) with the left-side technique,
-$\supsetneq$ the percentage of invariants stronger with the right-side technique,
-and ``uncomparable'' gives the percentage of invariants that are uncomparable, i.e
-neither greater nor smaller;
-the code points where both invariants are equal make up the remaining percentage. See Tab.~\ref{tab:techniques} for details.}
-  \label{fig:techniques}
-\end {figure}
-
-\begin{table*}
+\begin{sidewaystable}
 \begin{center}
@@
 \label{tab:techniques}
-\end{table*}
+\end{sidewaystable}
 
-For each program and each pair $(T_1,T_2)$ of analysis techniques, we list
-the proportion of control points in $P_R$ where $T_1$ (resp.~$T_2$) gives a
-strictly stronger invariant, denoted by $\subsetneq$ (resp. $\supsetneq$),
-and the proportion of control points where the invariants given by $T_1$ and
-$T_2$ are uncomparable for the inclusion ordering (the remainder of the
-control points are thus those for which both techniques give the same
-invariant). We use convex polyhedra as the abstract domain.
-
-Let us briefly comment the results given in more details in Table~\ref{tab:techniques} and Figure~\ref{fig:techniques}.
-\emph{Guided Static Analysis} from \citet{DBLP:conf/sas/GopanR07} improves the
-result of the classical Abstract Interpretation in $2.72\%$ of the control points
-in $P_R$.
-\emph{Path-focusing} from \citet{Monniaux_Gonnord_SAS11} gives
-statistically better results, and finds better invariants than \emph{Guided
-Static Analysis} in $9.89\%$ of the cases. However, it also loses precision in an
-important number ($2.68\%$) of control points.
-Finally, our combined technique gives the most promising results, since it is
-statistically more precise than the other techniques.
-
-The analysis using disjunctive invariants greatly improves the
-precision of the analysis (for $12.06\%$ of the control points in $P_R$), at an
-acceptable time cost (see Table \ref{tab:time}).
-
-While experiencing with techniques that use SMT-solving, we encountered some
-limitations due to non-linear arithmetic in the analyzed programs. Indeed, 
-the SMT-solver is not able to decide the satisfiability of some SMT-formulae
-expressing the semantics of non-linear programs. 
-In this case, we skipped the functions for which the SMT-solver returned the
-``unknown'' result.
-This limitation occurred very rarely in our experiments, except for the analysis
-of \emph{Lapack/Blas}, where 798 over the 1602 functions have been skipped.
-\emph{Lapack/Blas} implements matrix computations, which use floating-point multiplications.
-
-In cases where the formula is expressed in too rich a logic for the SMT-solver to deal with, a number of workarounds are possible, including:
-(i) \emph{Linearization}, as per \citet{DBLP:conf/vmcai/Mine06}, which overapproximates nonlinear semantics by linear semantics.
-(ii) Replacing the results of nonlinear operations by ``unknown''.
-Neither is currently implemented in our tool.
-
-Table \ref{tab:time} gives the execution time of the different analysis
-techniques. It is interesting to see that \emph{Guided Static Analysis} and
-\emph{Path-focusing} are sometimes faster than the classical
-algorithm.
-This seems due to the fact that these algorithms start on a small subset
-of the graph, and then may need fewer iterations on the complete graph.
-The techniques we introduce in this article are slower, but speed is still on the same order of magnitude as the older ones.
-
-\begin{table}[!h]
-	\centering
-\begin{tabular}{|l|r|r|r|r|r|} \hline
-	\multicolumn{1}{|c|}{Benchmark} &
-        \multicolumn{1}{c|}{\textbf{S}} &
-        \multicolumn{1}{c|}{\textbf{G}} &
-        \multicolumn{1}{c|}{\textbf{PF}} &
-        \multicolumn{1}{c|}{\textbf{G+PF}} &
-	\multicolumn{1}{c|}{\textbf{DIS}} \\ \hline
-	\input{table_time.tex} \hline
-\end{tabular}
-\caption{Execution time for each technique, expressed in seconds}
-\label{tab:time}
-\end{table}
-
-\subsection{Precision of Abstract Domains}
 \label{sec:compare_domains}
-\begin{table*}
+\begin{sidewaystable}
 \begin{center}
@@
 POLY),
-Octagons (OCT), intervals (BOX), linear equalities (PKEQ) and linear congruences
+Octagons (OCT), intervals (BOX), linear equalities (PKEQ) and linear
+inequalities with linear congruences
 (GRID). The last column compares the domain of Convex Polyhedra with the
-improved widening operator from \citet{BagnaraHRZ05SCP} (POLY*), and Convex Polyhedra
+improved widening operator from \cite{BagnaraHRZ05SCP} (POLY*), and Convex Polyhedra
 using the classical widening operator (POLY). POLY and POLY* use the Parma
@@
 \label{tab:domain}
-\end{table*}
+\end{sidewaystable}
 
-For each program and each pair $(D_1,D_2)$ of abstract domains, we compare by
-inclusion the invariants of the different control points in $P_R$ (where $P_R$
-is defined as in \ref{sec:compare_techniques}).
-The main results are given in Table \ref{tab:domain}.
-
-Statistically, the domain of convex polyhedra gives the better results, but
-commonly yields weaker invariants than the domains of octagons/intervals;
-this is a known weakness of its widening operator~\cite{Monniaux_LeGuen2011}. 
-The Octagon domain appears to be much better than intervals; this is
-unsurprising since in most programs and libraries, bounds on loop indices are
-non constant: they depend on some parameters (array sizes etc.).
-
-The Lapack/Blas benchmarks are unusual compared to the other programs.
-These libraries perform matrix computations, using nested loops over indices; such programs are the prime target for polyhedral loop optimization techniques and it is therefore unsurprising that polyhedra perform very well over them.
-
-The analysis of linear equalities (PKEQ) performs very fast compared
-to other abstract domains, but yields very imprecise invariants:
-it only detects relations of the form $\sum_i a_i x_i = C$ where $a_i$ and $C$ are constants.
-The comparison between the domain of convex polyhedra and linear congruences
-shows the percentage of points where the use of congruences refines the
-invariant. This is the case for $1.04\%$ of the control points in $P_R$.
-
-Finally, we evaluated the benefits of the improved version of the widening
-operator for convex polyhedra from \citet{BagnaraHRZ05SCP}, compared to the
-classical widening. We found that the improved version from
-\citet{BagnaraHRZ05SCP} yields more precise invariants for $6.58\%$ of the
-control points in $P_R$.
-
-\section{Conclusion and future prospects}
-Roughly, an analysis by abstract interpretation is defined by the choice of an iteration strategy and an abstract domain.
-In this article, we demonstrated that changes in the iteration algorithm can significantly improve precision, sometimes while improving analysis times.
-
-A common criticism of analysis techniques based on SMT-solving is that they
-do not scale up. Yet, our experiments show that, for numerical properties,
-they scale up to the size of typical functions and loops.
-It is however
-quite certain that, naively applied, they cannot scale to the kind of
-programs targeted by e.g. the Astr\'ee tool, that is, a dozens or hundreds
-of thousands of lines of code in a single loop operating over similar numbers
-of remanent variables.
-Actually, for such applications, only (quasi-)linear
-algorithms scale up, and ``cheap'' abstract domains such as octagons ($O(n^3)$ where $n$ is the number of variables) are not applied to the full variable set, but to restricted subsets thereof.
-It thus seems reasonable that techniques such as considering ``packs'' of related variables, slicing, etc. may similarly help SMT-based techniques to scale to global analyses.
-
-We compared the precision of different techniques and abstract domains by comparing the invariants for the inclusion ordering.
-A better metric is perhaps to take a client analysis --- such as the detection of overflows and array bound violations --- and compare the rates of alarms.
-
-We focused on numerical properties, because they are supported by easily available abstract libraries.
-Yet, in most programs, properties of data structures are important for proving interesting properties.
-Further investigations are needed not only on good abstractions for pointers (many are already known) but also on their conversion to SMT problems.
-
-% Cas a probleme:
-% Initialisation (x,y) à {(1,0), (0,1)}
-% puis boucle: si y = 0 alors x++
-% fait un polyèdre qui finit par perdre y <= 1 et x >= 0
-
-%\appendix
-%\section{Appendix Title}
-
-%This is the text of the appendix, if you need one.
-
-%\acks
-
-% Acknowledgments, if needed.
-
-% We recommend abbrvnat bibliography style.
-
-\phantomsection\addcontentsline{toc}{section}{References} 
-\bibliographystyle{dmabbrvnat}
-\bibliography{implicitization}
 \end{document}
-
-% Pour Emacs:
-% Local Variables:
-% reftex-cite-format: natbib
-% End:

### 2012-05-23 papers/SAS2012/implicitization.tex
mise en page version finale, debut

@@
-\maketitle
-
-\begin{abstract}
-Abstract interpretation techniques can be made more precise by distinguishing
-paths inside loops, at the expense of possibly exponential complexity. SMT-solving techniques and sparse representations of paths and sets of paths avoid this pitfall.
-
-We improve previously proposed techniques for guided static analysis and the generation of disjunctive invariants by combining them with techniques for succinct representations of paths and symbolic representations for transitions based on static single assignment.
-
-Because of the non-monotonicity of the results of abstract interpretation with widening operators, it is difficult to conclude that some abstraction is more precise than another based on theoretical local precision results.
-We thus conducted extensive comparisons between our new techniques and previous ones, on a variety of open-source packages.
-\end{abstract}
-
 \section{Introduction}

### 2012-05-23 papers/SAS2012/implicitization_article.tex
mise en page version finale, debut

--- 2012-05-23 papers/SAS2012/implicitization_lmcs.tex
@@
-\documentclass{llncs}
+\documentclass[a4paper]{amsart}
 
@@
 \usepackage{dcolumn}
-\usepackage{rotating}
+
+\usepackage[numbers]{dmnatbib}
 
@@
 
-\title{\mytitle}
-\subtitle{Combined analysis algorithms and experimental evaluation}
-\author{Julien Henry\inst{1}\inst{2} \and David Monniaux\inst{1}\inst{3} \and Matthieu Moy\inst{1}\inst{4}}
-
-\institute{VERIMAG laboratory, Grenoble, France
-\and Université Joseph Fourier
-\and CNRS
-\and Grenoble-INP}
-
 \renewcommand{\baselinestretch}{1}
@@
 \begin{document}
+\title[Succinct representations for abstract interpretation]{Succinct representations for abstract interpretation \\Combined analysis algorithms and experimental evaluation}
+%\thanks{\raisebox{-0.5em}{\includegraphics[height=1.5em]{ASOPT_officiel_nb}} This work was partially funded by ANR project ``ASOPT''}
+
+\author{Julien Henry}
+\address{J. Henry\\VERIMAG\\2 av de Vignate\\38610 Gières\\France}
+\email{Julien.Henry@imag.fr}
+\urladdr{http://www-verimag.imag.fr/~jhenry/}
+\thanks{Julien Henry is a graduate student at Université Joseph Fourier, VERIMAG laboratory. VERIMAG is a joint laboratory of Université Joseph Fourier, CNRS and Grenoble-INP}
+
+\author{David Monniaux}
+\address{D. Monniaux\\VERIMAG\\2 av de Vignate\\38610 Gières\\France}
+\email{David.Monniaux@imag.fr}
+\urladdr{http://www-verimag.imag.fr/~monniaux/}
+\thanks{David Monniaux is researcher at CNRS, VERIMAG laboratory}
+
+\author{Matthieu Moy}
+\address{M. Moy\\VERIMAG\\2 av de Vignate\\38610 Gières\\France}
+\email{Matthieu.Moy@grenoble-inp.fr}
+\urladdr{http://www-verimag.imag.fr/~moy/}
+\thanks{Matthieu Moy is assistant professor at Grenoble-INP, VERIMAG laboratory}
+
+\begin{abstract}
+\input{abstract}
+\end{abstract}
+
+\maketitle
+
 \input{implicitization}
@@
 
-\phantomsection\addcontentsline{toc}{section}{References} 
-\bibliographystyle{splncs03}
+\bibliographystyle{dmplainnat}
 \bibliography{implicitization}
-
-\appendix
-
-\MM{TODO: la partie appendix est laissée en chantier. A priori, on ne
-  mettra que les tableaux de chiffres sans les commentaires, et les
-  tableaux risquent de changer avec les avancées de Julien.}
-
-\section{Experimental Results}
-
-Experimental results of our tool on the benchmark presented in
-Fig~\ref{fig:projects} are given below. The precision of different
-techniques is given in Table~\ref{tab:techniques}. The precision of
-abstract domains and widening operators is given in
-Table~\ref{tab:domain}.
-
-\begin{sidewaystable}
-\begin{center}
-\setlength{\tabcolsep}{0.75ex}
-\begin{tabular}{|l
-|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
-|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
-|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
-|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
-|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
-|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}|} \hline
-\multicolumn{1}{|c|}{\textbf{Benchmark}}
-& \multicolumn{3}{c|}{\textbf{G/S}}
-& \multicolumn{3}{c|}{\textbf{PF/S}}
-& \multicolumn{3}{c|}{\textbf{PF/G}}
-& \multicolumn{3}{c|}{\textbf{G+PF/PF}}
-& \multicolumn{3}{c|}{\textbf{G+PF/G}}
-& \multicolumn{3}{c|}{\textbf{DIS/G+PF}} \\ %\cline{2-16}
-& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
-& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
-& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
-& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
-& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
-& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.} \\
- \hline
- \input{table1}
-\end{tabular}
-\end{center}
-\caption{Result of the comparison of the various techniques described in this
-paper: classic Abstract Interpretation (S), \emph{Guided Static Analysis} (G),
-\emph{Path-focusing} (PF), our combined technique (G+PF), and its version using
-disjunctive invariants (DIS). For
-instance, \textbf{G/S} compares the benefits of \emph{Guided Static Analysis}
-over the classic Abstract interpretation algorithm.
-$\subsetneq$, $\supsetneq$ and ``unc.'' are defined as in Fig.~\ref{fig:techniques}.}
-\label{tab:techniques}
-\end{sidewaystable}
-
-\label{sec:compare_domains}
-\begin{sidewaystable}
-\begin{center}
-\setlength{\tabcolsep}{0.75ex}
-\begin{tabular}{|l
-|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
-|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
-|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
-|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
-|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
-|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}|} \hline
-\multicolumn{1}{|c|}{\textbf{Benchmark}}
-& \multicolumn{3}{c|}{\textbf{PK/OCT}}
-& \multicolumn{3}{c|}{\textbf{PK/BOX}}
-& \multicolumn{3}{c|}{\textbf{OCT/BOX}}
-& \multicolumn{3}{c|}{\textbf{PK/PKEQ}}
-& \multicolumn{3}{c|}{\textbf{PK/GRID}}
-& \multicolumn{3}{c|}{\textbf{POLY/POLY*}} \\ %\cline{2-16}
-& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
-& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
-& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
-& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
-& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
-& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.} \\
- \hline
- \input{table_domain.tex}
-\end{tabular}
-\end{center}
-\caption{Results of the comparison of the various abstract domains, when using
-the same technique (G+PF). We used as abstract domains Convex Polyhedra (PK and
-POLY),
-Octagons (OCT), intervals (BOX), linear equalities (PKEQ) and linear
-inequalities with linear congruences
-(GRID). The last column compares the domain of Convex Polyhedra with the
-improved widening operator from \cite{BagnaraHRZ05SCP} (POLY*), and Convex Polyhedra
-using the classical widening operator (POLY). POLY and POLY* use the Parma
-Polyhedral Library \cite{PPL}.
-$\subsetneq$, $\supsetneq$ and ``unc.'' are defined as in Fig.~\ref{fig:techniques}.}
-\label{tab:domain}
-\end{sidewaystable}
-
 \end{document}

### 2012-05-23 papers/SAS2012/implicitization_lmcs.tex
mise en page version finale, debut
en 17 pages

@@
 \usepackage{dcolumn}
-\usepackage{rotating}
 
@@
 \begin{document}
+\maketitle
+
+\begin{abstract}
+\input{abstract}
+\end{abstract}
+
 \input{implicitization}
@@
 
-\appendix
-
-\MM{TODO: la partie appendix est laissée en chantier. A priori, on ne
-  mettra que les tableaux de chiffres sans les commentaires, et les
-  tableaux risquent de changer avec les avancées de Julien.}
-
-\section{Experimental Results}
-
-Experimental results of our tool on the benchmark presented in
-Fig~\ref{fig:projects} are given below. The precision of different
-techniques is given in Table~\ref{tab:techniques}. The precision of
-abstract domains and widening operators is given in
-Table~\ref{tab:domain}.
-
-\begin{sidewaystable}
-\begin{center}
-\setlength{\tabcolsep}{0.75ex}
-\begin{tabular}{|l
-|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
-|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
-|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
-|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
-|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
-|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}|} \hline
-\multicolumn{1}{|c|}{\textbf{Benchmark}}
-& \multicolumn{3}{c|}{\textbf{G/S}}
-& \multicolumn{3}{c|}{\textbf{PF/S}}
-& \multicolumn{3}{c|}{\textbf{PF/G}}
-& \multicolumn{3}{c|}{\textbf{G+PF/PF}}
-& \multicolumn{3}{c|}{\textbf{G+PF/G}}
-& \multicolumn{3}{c|}{\textbf{DIS/G+PF}} \\ %\cline{2-16}
-& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
-& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
-& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
-& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
-& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
-& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.} \\
- \hline
- \input{table1}
-\end{tabular}
-\end{center}
-\caption{Result of the comparison of the various techniques described in this
-paper: classic Abstract Interpretation (S), \emph{Guided Static Analysis} (G),
-\emph{Path-focusing} (PF), our combined technique (G+PF), and its version using
-disjunctive invariants (DIS). For
-instance, \textbf{G/S} compares the benefits of \emph{Guided Static Analysis}
-over the classic Abstract interpretation algorithm.
-$\subsetneq$, $\supsetneq$ and ``unc.'' are defined as in Fig.~\ref{fig:techniques}.}
-\label{tab:techniques}
-\end{sidewaystable}
-
-\label{sec:compare_domains}
-\begin{sidewaystable}
-\begin{center}
-\setlength{\tabcolsep}{0.75ex}
-\begin{tabular}{|l
-|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
-|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
-|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
-|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
-|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
-|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}|} \hline
-\multicolumn{1}{|c|}{\textbf{Benchmark}}
-& \multicolumn{3}{c|}{\textbf{PK/OCT}}
-& \multicolumn{3}{c|}{\textbf{PK/BOX}}
-& \multicolumn{3}{c|}{\textbf{OCT/BOX}}
-& \multicolumn{3}{c|}{\textbf{PK/PKEQ}}
-& \multicolumn{3}{c|}{\textbf{PK/GRID}}
-& \multicolumn{3}{c|}{\textbf{POLY/POLY*}} \\ %\cline{2-16}
-& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
-& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
-& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
-& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
-& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
-& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.} \\
- \hline
- \input{table_domain.tex}
-\end{tabular}
-\end{center}
-\caption{Results of the comparison of the various abstract domains, when using
-the same technique (G+PF). We used as abstract domains Convex Polyhedra (PK and
-POLY),
-Octagons (OCT), intervals (BOX), linear equalities (PKEQ) and linear
-inequalities with linear congruences
-(GRID). The last column compares the domain of Convex Polyhedra with the
-improved widening operator from \cite{BagnaraHRZ05SCP} (POLY*), and Convex Polyhedra
-using the classical widening operator (POLY). POLY and POLY* use the Parma
-Polyhedral Library \cite{PPL}.
-$\subsetneq$, $\supsetneq$ and ``unc.'' are defined as in Fig.~\ref{fig:techniques}.}
-\label{tab:domain}
-\end{sidewaystable}
-
 \end{document}

### 2012-05-23 papers/SAS2012/implicitization.tex
remarques de Sriram, #1

@@
 \label{sec:disjunctive}
+While many (most?) useful program invariants on numerical variables can be expressed as conjunctions of inequalities and congruences, it is sometimes necessary to introduce disjunctions.
+For instance, the loop \lstinline|for (int i=0; i<n; i++) {...}| has head invariant $0 \leq i \leq n \lor (i=0 \land n<0)$.
+For this very simple example, a simple syntactic transformation of the control structure (into \lstinline|i=0; if (i<n) do {...} while (i<n)|) is sufficient, but in more complex cases more advanced analyses are necessary
+\cite{DBLP:conf/emsoft/BalakrishnanSIG09,DBLP:journals/fmsd/Jeannet03,DBLP:conf/cav/SharmaDDA11,Monniaux_Bodin_APLAS11};
+in intuitive terms, they discover \emph{phases} or \emph{modes} in loops.
 
@@
 Further investigations are needed not only on good abstractions for pointers (many are already known) but also on their conversion to SMT problems.
+
+%%% Local Variables: 
+%%% mode: latex
+%%% TeX-master: "implicitization_lmcs.tex"
+%%% reftex-cite-format: natbib
+%%% End:

### 2012-05-23 papers/SAS2012/implicitization.tex
Harris et al. Pas con.

@@
 using abstract acceleration~\cite{DBLP:conf/sas/GonnordH06}.
+
+We assume that we can encode the concrete semantics of the program into the SMT formula, or at least an abstraction thereof at least as precise as the one applied by the abstract interpreter (in simple terms: we want to avoid the case where the SMT solver exhibits a possible path, but the static analyzer realizes that this path is infeasible; this would lead to nontermination, because the SMT solver would exhibit the same path on the next iteration).
+A workaround would be to apply \emph{satisfiability modulo path programs} \cite{DBLP:conf/popl/HarrisSIG10}: from each path ruled infeasible by abstract interpretation, extract a blocking clause for the SAT solver underlying the SMT-solver.
 

### 2012-05-23 papers/SAS2012/implicitization.tex
fixes review 2

@@
 
+PAGAI currently neither models the memory heap nor performs interprocedural analysis.
+Instead, LLVM optimization phases are applied prior to analysis, in order to inline non-recursive function calls and lift certain memory accesses to operations on explicit numerical variables
+(e.g. \lstinline|y=t[0]*t[0];| preceded by \lstinline|t[0]=x;| without any aliased write in between is replaced by \lstinline|y=x*x;|).
+The remaining memory reads are considered as indeterminates, and memory writes are ignored; this is a sound abstraction.
+
 We conducted extensive experiments on real-life programs in order to compare the

### 2012-05-23 papers/PLDI2012/response_PLDI2012.txt

@@
+Reviewer #1
 "the title is fairly vague and non optimal -- perhaps
 "succinct" should be replaced by "disjunctive"?"
-Path Focusing technique and our combined technique (with or without disjunctive
-invariants) do not work on the classical control flow graph, but on a multigraph
-where all the paths inside loops are distinguished. The term "succint" refers to
-the fact we actually do not compute all the paths of this multigraph, but work
-on an succint representation thereof (using an SMT formula), thus avoiding the
-enumeration of an exponential number of paths in a huge number of cases.
+Path focusing and our combined technique (with or without disjunctive invariants) do not work on the classical control flow graph, but on a multigraph where all the paths inside loops are distinguished.
+"Succinct" refers to the fact we actually do not represent explicitly all these paths, but instead list them on demand as solutions to an SMT formula; see Papadimitriou "succinct representations of graphs" using SAT formulas.
 
@@
 program give more precise results, but, more importantly, at a reasonable cost
-(which was not evident).
-It is also interesting to see that more complex technique perform sometimes
-faster that more naive approaches.
+(non-obvious).
+Interestingly, more precise techniques (e.g. guided static analysis) perform sometimes faster that more naive approaches.
 
-"The problem with path focusing is that (1) it will process all
-executable paths at once, including potentially many redundant paths
-that don't contribute to the solution; and (2) its widening operation
-lacks the precision conferred by guided static analysis."
+Reviewer #2
+"The problem with guided static analysis is [...] thus losing precision"
+Precisely, the precision loss is due to "least upper bound" operators at intermediate control joins, which may add spurious states.
 
-This is not exactly the point. The problem with path focusing is that it will compute
-the fixpoint iterations on the whole graph at once, using classical widening
-operations in order to guarantee the convergence.
-When the fixpoint has been reached, it finally apply some narrowing iterations
-to recover precision.
-The idea here is to apply some narrowing iterations during the analysis, in
-order to recover some precision before continuing.
-To clarify, widening is the same, but we perform regularly narrowing iterations,
-and not only at the end of the ascending iterations.
+"The problem with path focusing is that [...] analysis."
+This is not exactly the point.
+The problem with path focusing is that it will compute the fixpoint iterations on the whole graph at once, using classical widening operations in order to guarantee the convergence.
+When a fixpoint has been reached, it finally applies some narrowing iterations to recover precision.
+Guided static analysis instead applies narrowing iterations at intermediate point during the analysis, in order to recover some precision, not only at the end of the ascending iterations.
 
@@
 domains, including those that don't require widening"
-Intuitively, Guided Static Analysis technique performs better because it
+Intuitively, guided static analysis performs better because it
 recovers precision lost by widening. It will probably not increase the precision
-when using domains without widening/narrowing. However, the iteration strategy
-will be changed, so the computation time may differ compared to a classical
-abstract interpretation. We can do the same remarks on our combined technique
-compared to Path Focusing.
+when using domains without widening/narrowing.
+However, since the iteration strategy is changed, the computation time may differ compared to a classical abstract interpretation. The same applies to our combined technique compared to Path Focusing.
 
-"I'd like to know more about how applicable the technique
-is to other kinds of abstract domains"
+"I'd like to know more about how applicable the technique is to other kinds of abstract domains"
 The general idea does not depend on the abstract domain. However, we have to
 transform an abstract value into an SMT formula. Such transformation is easy in
-our case, since it is a cunjunction of linear inequalities, but may still
-require some research for other abstract domains that could require other
-theories (for instance, arrays could be
-encoded in SMT using uninterpreted functions, etc.)
+our case, since a polyhedron in constraint form is a conjunction of linear inequalities, but further work may be needed for efficient encoding of e.g.
+alias and pointer information (these can be encoded as uninterpreted functions with functional updates, aka arrays, but solving the resulting formulas may be costly).
 
+Reviewer #3
 "2 to 5% difference, does it make a difference at all?"
-Our prototype of analyzer performs many over-approximations (no memory model,
-etc.), thus leading to invariants that are often very unprecise.
-Intuitively, I would say we could have obtained more difference between the
-different techniques by reducing the number of over-approximation we do for the
-time being.
+Our prototype of analyzer performs many over-approximations, especially since it has no memory model: the program is preprocessed by a LLVM memory-to-scalar pass, and all memory references that this pass cannot turn into SSA variables are considered undefined.
+This often results in very imprecise invariants regardless of the iteration method and domain used.
+
+We thus expect that a less rough treatment of memory will yield more marked differences between iteration techniques.

### 2012-05-23 papers/SAS2012/abstract.tex

--- 2011-11-04 papers/PLDI2012/implicitization.tex
@@
-\documentclass[preprint]{sigplanconf}
-
-% The following \documentclass options may be useful:
-%
-% 10pt          To set in 10-point type instead of 9-point.
-% 11pt          To set in 11-point type instead of 9-point.
-% authoryear    To obtain author/year citation style instead of numeric.
-
-%\newcommand{\MM}[1]{{\color{blue} TODO(MM): #1}}
-%\newcommand{\JH}[1]{{\color{teal} TODO(JH): #1}}
-%\newcommand{\DM}[1]{{\color{violet} TODO(DM): #1}}
-\newcommand{\MM}[1]{}
-\newcommand{\JH}[1]{}
-\newcommand{\DM}[1]{}
-
-% Pour les commentaires
-\usepackage[utf8]{inputenc}
-
-\usepackage{amsmath}
-\usepackage{mathcomp,amsfonts,amssymb,MnSymbol}
-\usepackage{dmnatbib}
-\usepackage[mathscr]{euscript}
-\usepackage{algorithmic,algorithm}
-%\usepackage{graphicx}
-%\usepackage{color}
-
-\usepackage{enumitem}
-\usepackage{dcolumn}
-
-\usepackage{listings}
-\lstset{language=C,mathescape=true,numbers=left,numbersep=0.8ex,tabsize=2,breaklines=true,basicstyle={\rm\fontfamily{phv}\small\selectfont}}
-
-\usepackage{tikz}
-\usetikzlibrary{arrows,automata}
-
-\tikzstyle{state}=[circle,fill=black!25,minimum size=13pt,inner sep=0pt]
-\tikzstyle{rstate}=[rectangle,fill=black!25,minimum size=13pt,inner sep=0pt]
-\tikzstyle{transition}=[rectangle,semithick,draw=black!75,
-  			  minimum size=4mm]
-\tikzstyle{transition2}=[transition,rectangle,thick,dashed,
-  			  minimum size=4mm]
-\tikzstyle{PRstate}=[circle,double,draw,fill=blue!15,minimum size=13pt,inner sep=0pt]
-\tikzstyle{polyhedra}=[blue!25,opacity=0.5,pattern=north west lines,pattern
-color=blue]
-\tikzstyle{line}=[black,thick]
-
-\usepackage[pdfusetitle]{hyperref}
-
-\newcommand{\abstr}[1]{#1^\sharp}
-\newcommand{\parts}[1]{\mathscr{P}(#1)}
-\newcommand{\ZZ}{\mathbb{Z}}
-\newcommand{\QQ}{\mathbb{Q}}
-\newcommand{\RR}{\mathbb{R}}
-\newcommand{\widening}{\mathop{\triangledown}}
-
-% commentaires dans les algorithmes crees avec le package algorithmic
-\renewcommand{\algorithmiccomment}[1]{ {\color{gray} // #1}}
-
-\begin{document}
-
-\conferenceinfo{PLDI '12}{11-16 June 2012, Beijing} 
-\copyrightyear{2012} 
-\copyrightdata{[to be supplied]}
-
-%\titlebanner{banner above paper title}        % These are ignored unless
-\preprintfooter{Succinct representations for abstract interpretation}   % 'preprint' option specified.
-
-\title{Succinct Representations for Abstract Interpretation}
-\subtitle{Combined analysis algorithms and experimental evaluation}
-
-\authorinfo{Anonymous}
-           {The Internet}
-           {anonymous.gal@gmail.com}
-
-\maketitle
-
-\begin{abstract}
 Abstract interpretation techniques can be made more precise by distinguishing
@@
 We thus conducted extensive comparisons between our new techniques and previous ones, on a variety of open-source packages.
-We also took the opportunity to compare the precision of various classical abstract domains (convex polyhedra, octagons, intervals).
-\end{abstract}
-
-% \category{CR-number}{subcategory}{third-level}
-\category{D.2.4}{Software/Program Verification}{Formal methods}
-\category{F.3.1}{Specifying and Verifying and Reasoning about Programs}{Assertions, Invariants, Mechanical verification, Pre- and post-conditions}
-\category{F.3.2}{Semantics of Programming Languages}{Program analysis}
-\category{F.4.1}{Mathematical Logic}{Logic and constraint programming, Mechanical theorem proving}
-\category{I.2.3}{Deduction and Theorem Proving}{Inference engines, Logic programming, Resolution}
-
-\terms
-theory, verification, algorithms
-
-\keywords
-SMT-solving, static analysis, abstract interpretation, disjunctions
-
-\section{Introduction}
-Static analysis by abstract interpretation is a fully automatic program analysis method.
-When applied to imperative programs, it computes an inductive invariant mapping each program location (or a subset thereof) to a set of states represented symbolically~\cite{CousotCousot_JLC92}.
-For instance, if we are only interested in scalar numerical program variables, such a set may be a convex polyhedron (the set of solutions of a system of linear inequalities)~\cite{CousotHalbwachs78,Halbwachs_PhD,PPL,BagnaraHZ08SCP}.
-
-In such an analysis, information may flow forward (one computes the polyhedron after a program statement as the image by the semantics of the statement, or a super-set thereof, of the polyhedron before) or backward;
-forward program analysis computes super-sets of the states reachable from the initialization of the program, backward program analysis computes super-sets of the states co-reachable from some property of interest (for instance, the violation of an assertion).
-In forward analysis, control-flow joins correspond to convex hulls if using convex polyhedra (more generally, they correspond to least upper bounds in a lattice); in backward analysis, it is control-flow splits that correspond to convex hulls.
-
-It is a known limitation of program analysis by abstract interpretation that this convex hull, or more generally, least upper bound operation, may introduce states that cannot occur in the real program: for instance, the convex hull of the intervals $[-2,-1]$ and $[1,2]$ is $[-2,2]$, strictly larger than the union of the two.
-Such introduction may prevent proving desired program properties, for instance $\neq 0$.
-The alternative is to keep the union symbolic (e.g.
-compute using $[-2,-1] \cup [1,2]$) and thus compute in the \emph{disjunctive completion}
-of the lattice, but the number of terms in the union may grow exponentially with the number of successive tests in the program to analyze, not to mention difficulties for designing suitable widening operators for enforcing the convergence of fixpoint iterations~\cite{PPL,BagnaraHZ08SCP,DBLP:journals/sttt/BagnaraHZ07}.
-The exponential growth of the number of terms in the union may be controlled by heuristics that judiciously apply least upper bound operations, as in the \emph{trace partitioning domain} \cite{Rival_Mauborgne_TOPLAS07} implemented in the Astr\'ee analyzer~\cite{ASTREE_PLDI03,ASTREE_ESOP05}.
-
-Assuming we are  interested in a loop-free program fragment, the above approach of keeping symbolic unions gives the same results as performing the analysis separately over every path in the fragment.
-A recent method for finding disjunctive loop invariants \cite{DBLP:conf/pldi/GulwaniZ10} is based on this idea: each path inside the loop body is considered separately.
-Two recent proposals use SMT-solving \cite{Kroening_Strichman_08} as a decision procedure for the satisfiability of first-order arithmetic formulas in order to enumerate only paths that are needed for the progress of the analysis \cite{Gawlitza_Monniaux_ESOP11,Monniaux_Gonnord_SAS11}.
-They can equivalently be seen as analyses over a multigraph of transitions between some distinguished control nodes.
-This multigraph has an exponential number of edges, but is never explicitly represented in memory; instead, this graph is \emph{implicitly} or \emph{succinctly} represented: its edges are enumerated as needed as solutions to SMT problems.
-
-An additional claim of the methods that distinguish paths inside the loop body \cite{DBLP:conf/pldi/GulwaniZ10,Monniaux_Gonnord_SAS11} is that they tend to generate better invariants that methods that do not, by behaving better with respect to the \emph{widening operators} \cite{CousotCousot_JLC92}
-used for enforcing convergence when searching for loop invariants by Kleene iterations.
-A related technique, \emph{guided static analysis} \cite{DBLP:conf/sas/GopanR07}, computes successive loop invariants for increasing subsets of the transitions taken into account, until all transitions are considered; again, the claim is that this approach avoids some gross over-approximation introduced by widenings.
-
-All these methods improve the precision of the analysis by keeping the same abstract domain (say, convex polyhedra) but changing the operations applied and their ordering.
-An alternative is to change the abstract domain, for instance by moving from intervals to octagons or convex polyhedra \cite{DBLP:journals/lisp/Mine06}, or the widening operator: for instance one may replace the classical widening on convex polyhedra \cite{CousotHalbwachs78,Halbwachs_PhD} by an improved version \cite{BagnaraHRZ05SCP}, or use widening ``up to'', that is, use linear inequalities syntactically present in the program source code as possible widening steps~\cite{Polka:FMSD:97}.
-
-There are many possible combinations of the above techniques, and it is not
-evident which ones perform more or less precisely or more or less efficiently on real-life examples. One needs to experiment.
-Unfortunately, the published literature on the subject lacks experimental comparative assessments. One purpose of this article is therefore to propose such experimental results.
-
-This article makes the following contributions:
-\begin{enumerate}
-\item We recast the guided static analysis technique from \citet{DBLP:conf/sas/GopanR07} on the expanded multigraph from \citet{Monniaux_Gonnord_SAS11}, considering entire paths instead of individual transitions, using SMT queries and binary decision diagrams.\label{contr:guided_multigraph} (See \S\ref{sec:guided_multigraph})
-\item We improve the technique for obtaining disjunctive invariants from \citet{DBLP:conf/pldi/GulwaniZ10} by replacing the explicit exhaustive enumeration of paths by a sequence of SMT queries.\label{contr:disjunctive} (See \S\ref{sec:disjunctive})
-\item We implemented these techniques, in addition to ``classical'' iterations and the original guided static analysis, inside a prototype static analyzer.
-This tool uses the LLVM bitcode format \cite{Lattner:2004:LCF:977395.977673,LLVM_langref} as input, which can be produced by compilation from C, C++ and Fortran, enabling it to be run on many real-life programs.
-With respect to abstract domains on numerical variables, it uses the APRON library \cite{DBLP:conf/cav/JeannetM09}, which supports a variety of abstract domains from which we can choose with minimal changes to our analyzer.
-
-Our tool uses a \emph{single static assignment} (SSA) intermediate representation, which allows us to perform cheap symbolic propagations and simplifications,
-described in Section~\ref{sec:analysis-algorithm}.
-
-\item We conducted extensive experiments with this tool, on real-life programs: (i) fixing the abstract domain, varying the iteration technique (\S~\ref{sec:compare_techniques}) (ii) fixing the iteration technique, varying the abstract domain (\S\ref{sec:compare_domains}).
-\end{enumerate}
-
-\section{Bases}
-\subsection{Static analysis by abstract interpretation}
-\label{sec:static_analysis}
-Let $X$ be the set of possible states of the program variables; for instance, if the program has 3 unbounded integer variables, then $X = \ZZ^3$.
-The set $\parts{X}$ of subsets of $X$, partially ordered by inclusion, is the \emph{concrete domain}.
-An \emph{abstract domain} is a set $\abstr{X}$ equipped with a partial order $\sqsubseteq$; for instance, it can be the domain of convex polyhedra in $\QQ^3$ ordered by geometric inclusion.
-In this article, all abstract domains are supposed to contain machine-representable objects, and all $\sqsubseteq$ order relations are supposed to be decidable.
-The concrete and abstract domains are connected by a monotone \emph{concretization} function $\gamma: \left(\abstr{X},\sqsubseteq\right) \rightarrow (\parts{X},\subseteq)$: an element $\abstr{x} \in \abstr{X}$ represents a set $\gamma(\abstr{x})$.%
-%
-\footnote{%
-Some presentations of abstract interpretation identify an element $\abstr{x}$ with the set of states $\gamma(\abstr{x})$ that it represents.
-This leads to simpler notations, but may also confuse if, as in some of our constructions, there exist several $\abstr{x}$ with identical $\gamma(\abstr{x})$, but distinct algorithmic behaviors.}
-
-We also assume a join operator $\sqcup: \abstr{X} \times \abstr{X} \rightarrow \abstr{X}$, with infix notation; in practice, it is generally a least upper bound operation, but we only need it to satisfy $\gamma(\abstr{x}) \cup \gamma(\abstr{y}) \subseteq \gamma(\abstr{x} \sqcup \abstr{y})$ for all $\abstr{x},\abstr{y}$.
-
-Classically, one considers the control-flow graph of the program, with edges labeled with concrete transition relations (e.g.
-$x' = x+1$ for an instruction \lstinline|x = x+1;|), and attaches an abstract element to each control point.
-A concrete transition relation $\tau \subseteq X \times X$ is replaced by an abstract \emph{forward abstract transformer} $\abstr{\tau}: \abstr{X} \rightarrow \abstr{X}$, such that
-\begin{equation}
-\forall \abstr{x} \in \abstr{X}, x,x' \in X,~
-x \in \gamma(\abstr{x}) \land (x,x') \in \tau \implies
-x' \in \gamma \circ \abstr{\tau}(\abstr{x})
-\label{eqn:transformer_correct}
-\end{equation}
-It is easy to see that if to any control point $p \in P$ we attach an abstract element $\abstr{x}_p$ such that
-(i) for any $p$, $\gamma(\abstr{x}_p)$ includes all initial states possible at control node $p$
-(ii) for any $p,p'$, $\abstr{\tau}_{p,p'} (\abstr{x}_p) \sqsubseteq \abstr{x}_{p'}$, noting $\tau_{p,p'}$ the transition from $p$ to~$p'$, then $(\gamma(\abstr{x}_p))_{p \in P}$ form an \emph{inductive invariant}: by induction, when the control point is $p$, the program state always lies in $\gamma(\abstr{x}_p)$.
-
-\emph{Kleene iterations} compute such an inductive invariant as the stationary limit, if it exists, of the following system: for each $p$, initialize $\abstr{x}_p$ such that $\gamma(\abstr{x}_p)$ is a superset of the initial states at point $p$; then iterate the following:
-if $\abstr{\tau}_{p,p'} (\abstr{x}_p) \nsqsubseteq \abstr{x}_{p'}$, replace $\abstr{x}_{p'}$ by $\abstr{x}_{p'} \sqcup \abstr{\tau}_{p,p'} (\abstr{x}_p)$.
-Such a stationary limit is bound to exist if $\abstr{X}$ has no infinite ascending chain $a_1 \sqsubsetneq a_2 \sqsubsetneq \dots$; this condition is however not met by domains such as intervals or convex polyhedra.
-
-\emph{Widening-accelerated Kleene iterations} proceed by replacing $\abstr{x}_{p'} \sqcup \abstr{\tau}_{p,p'} (\abstr{x}_p)$ by $\abstr{x}_{p'} \widening (\abstr{x}_{p'} \sqcup \abstr{\tau}_{p,p'} (\abstr{x}_p))$ where $\widening$ is a \emph{widening operator}: for all $\abstr{x},\abstr{y}$, $\gamma(\abstr{y}) \subseteq \gamma(\abstr{x} \widening \abstr{y})$, and there exists no sequence $\abstr{u}_1,\abstr{u}_2,\dots$ of the form $\abstr{u}_{n+1} = \abstr{u}_n \widening \abstr{v}_n$ where $\abstr{v}_n$ is another sequence.
-Such iterations necessarily lead to a stationary limit $(\abstr{x}_p)_{p \in P}$, which defines an inductive invariant $(\gamma(\abstr{x}_p))_{p \in P}$.
-Note that this invariant is not, in general, the least one expressible in the abstract domain, and may depend on the iteration ordering (the successive choices~$p,p'$).
-
-Once an inductive invariant $\gamma((\abstr{x}_p)_{p \in P})$ has been obtained,
-one can attempt \emph{decreasing} or \emph{narrowing} iterations to reduce it.
-In their simplest form, this just means running the following operation until a
-fixpoint or a maximal number of iterations are reached: for any $p'$, replace
-$\abstr{x}_{p'}$ by $\abstr{x}_{p'} \cap \left(\bigsqcup_{p \in P}
-\abstr{\tau}_{p,p'} (\abstr{x}_p)\right)$. The result also defines an inductive
-invariant. These decreasing iterations are indispensable to recover properties
-from guards (tests) in the program in most iteration settings; unfortunately,
-certain loops, particularly those involving identity (no-operation) transitions,
-may foil them: the iterations immediately reach a fixpoint and do not decrease
-further (see example in \S\ref{subsec:rate_lim}). Sections \ref{sec:guided} and
-\ref{sec:path_focusing} describe techniques that work around this problem.
-
-Widening operations have a somewhat counterintuitive behavior: the result of an analysis (postcondition) is not necessarily monotonic with respect to the precondition.
-For instance, classical interval analysis with widening and decreasing iterations on the following loop yields a loop invariant $x \in [0,+\infty)$ if the precondition is $x \in [0,0]$ , but $x \in [0,10]$ (strictly stronger) if the precondition is $x \in [0,10]$ (strictly weaker):
-\lstinline|while (x != 10) x=x+1;|.
-This means that, when using widenings, being locally more precise (for instance, by using more expressive abstract domains, or widening operators that climb more slowly \cite{BagnaraHRZ05SCP}) does not necessarily translate into higher final precision.
-It is almost always possible to concoct examples where a supposedly more precise abstraction (e.g.
-polyhedra) yields less precise results than a less precise one (e.g.
-intervals), but are these examples representative of real programs?
-This justifies our recourse to evaluation on realistic examples~(\S~\ref{sec:compare_techniques}).
-
-\subsection{SMT-solving}
-Boolean satisfiability (SAT) is the canonical NP-complete problem: given a propositional formula (e.g.
-$(a \lor \neg b) \land (\neg a \lor b \lor \neg c)$), decide whether it is satisfiable --- and, if so, output a satisfying assignment.
-Despite an exponential worst-case complexity, the DPLL algorithm \cite{Kroening_Strichman_08,Handbook_SAT} solves many useful SAT problems in practice.
-
-SAT was extended to \emph{satisfiability modulo theory} (SMT): in addition to propositional literals, SMT formulas admit atoms from a theory.
-For instance, the theories of linear integer arithmetic (LIA) and linear real arithmetic (LRA) have atoms of the form $a_1 x_1 + \dots + a_n x_n \bowtie C$ where $a_1,\dots,a_n,C$ are integer constants, $x_1,\dots,x_n$ are variables (interpreted over $\ZZ$ for LIA and $\RR$ or $\QQ$ for LRA), and $\bowtie$ is a comparison operator $=,\neq,<,\leq,>,\geq$.
-Satisfiability for LIA and LRA is NP-complete, yet tools based on DPLL(T) approach \cite{Kroening_Strichman_08,Handbook_SAT} solve many useful SMT problems in practice.
-All these tools provide a \emph{satisfying assignment} if the problem is satisfiable.
-
-Most SMT solvers, including Z3,
-%\footnote{\url{http://research.microsoft.com/en-us/um/redmond/projects/z3/}}
-Yices,%
-%\footnote{\url{http://yices.csl.sri.com/}}
-and all those supporting the full SMTLIB2 standard \cite{BarST-SMTLIB},
-offer an \emph{incremental} interface: the client program specifies the formula as an initially empty conjunction, to which additional constraints are added, and calls a ``check'' function answering whether it is satisfiable;
-it may then backtrack some of the constraints and add other ones without restarting from scratch.
-
-\subsection{A simple, motivating example}
-\label{subsec:rate_lim}
-Consider the following program, adapted from \cite{Monniaux_Gonnord_SAS11}, where \lstinline|input($a$, $b$)| stands for a nondeterministic input in $[a,b]$:
-\lstinputlisting{rlim_int.c}
-This program implements a construct commonly found in control programs (in e.g.
-automotive or avionics): a rate limiter.
-For the sake of simplicity, we chose it to be fed a nondeterministic input clamped between $[-100000,100000]$, but in a real system it would be integrated in a reactive control loop and its input connected to a complex system with unknown output range.
-
-The expected inductive invariant is $\lstinline|x_old| \in [-100000,100000]$, but classical abstract interpretation using intervals (or octagons or polyhedra) finds $\lstinline|x_old| \in (-\infty,+\infty)$~\citep{ASTREE_ESOP05}.
-Let us briefly see why.
-
-Widening iterations converge to $\lstinline|x_old| \in (-\infty,+\infty)$; let us now see why decreasing iterations fail to recover the desired invariant.
-The \lstinline|x > x_old+10| test at line~6, if taken, yields $\lstinline|x_old|
-\in (-\infty,99990)$; followed by \lstinline|x = x_old+10|, we obtain
-$\lstinline|x| \in (-\infty,\allowbreak 100000)$, and the same after union with the no-operation ``else'' branch. Line~7 yields $\lstinline|x| \in (-\infty,+\infty)$.
-
-We could use ``widening up to'' or ``widening with thresholds'', propagating the ``magic values'' $\pm 100000$ associated to \lstinline|x| into \mbox{\lstinline|x_old|,} but these syntactic approaches cannot directly cope with programs for which $\lstinline|x|  \in [-100000,+100000]$ is itself obtained by analysis.
-The guided static analysis of \citet{DBLP:conf/sas/GopanR07} does not perform
-better, and also obtains $\lstinline|x_old| \in (-\infty,+\infty)$.
-
-In contrast, let us distinguish all four possible execution paths through the tests at lines 6 and~7. The path through both ``else'' branches is infeasible; the program is thus equivalent to:
-\lstinputlisting{rlim_int2.c}
-
-\begin{figure}
-\centering
-\begin{minipage}[c]{.19\textwidth}
-\begin{tikzpicture}[->,>=stealth',auto,node distance=1.1cm,
-                    semithick,font=\footnotesize]
-
-	\node[PRstate] (n0) {$p$};
-	\node[state] (n1) [below left of=n0] {};
-	\node[state] (n2) [below right of=n0] {};
-	\node[state] (n3) [below right of=n1] {};
-	\node[state] (n4) [below left of=n3] {};
-	\node[state] (n5) [below right of=n3] {};
-	\node[state] (n6) [below right of=n4] {};
-	\node[state] (n7) [below left of=n6] {};
-	\node[state] (n8) [below right of=n6] {};
-	\node[PRstate] (n9) [below right of=n7] {$q$};
-
-  \path [transition] 
-		(n0) edge              node {} (n1);
-  \path [transition] 
-		(n0) edge              node {} (n2);
-  \path [transition] 
-		(n1) edge              node {} (n3);
-  \path [transition] 
-		(n2) edge              node {} (n3);
-  \path [transition] 
-		(n3) edge              node {} (n4);
-  \path [transition] 
-		(n3) edge              node {} (n5);
-  \path [transition] 
-		(n4) edge              node {} (n6);
-  \path [transition] 
-		(n5) edge              node {} (n6);
-  \path [transition] 
-		(n6) edge              node {} (n7);
-  \path [transition] 
-		(n6) edge              node {} (n8);
-  \path [transition] 
-		(n7) edge              node {} (n9);
-  \path [transition] 
-		(n8) edge              node {} (n9);
-\end{tikzpicture}
-\end{minipage} 
-$\Longrightarrow$
-\begin{minipage}[c]{.19\textwidth}
-\begin{tikzpicture}[->,>=stealth',auto,node distance=1.1cm,
-                    semithick,font=\footnotesize]
-
-	\node[PRstate] (n0) {$p$};
-	\node (n1) [below left of=n0] {};
-	\node (n2) [below right of=n0] {};
-	\node (n3) [below right of=n1] {};
-	\node (n4) [below left of=n3] {};
-	\node (n5) [below right of=n3] {};
-	\node (n6) [below right of=n4] {};
-	\node (n7) [below left of=n6] {};
-	\node (n8) [below right of=n6] {};
-	\node[PRstate] (n9) [below right of=n7] {$q$};
-
-  \path [transition] 
-		(n0) edge  [out=0, in=0]            node {} (n9);
-  \path [transition] 
-		(n0) edge  [out=180, in=180]        node {} (n9);
-  \path [transition] 
-		(n0) edge  [out=205, in=155]            node {} (n9);
-  \path [transition] 
-		(n0) edge  [out=230, in=130]            node {} (n9);
-  \path [transition] 
-		(n0) edge  [out=255, in=105]            node {} (n9);
-  \path [transition] 
-		(n0) edge  [out=-25, in=25]            node {} (n9);
-  \path [transition] 
-		(n0) edge  [out=-50, in=50]            node {} (n9);
-  \path [transition] 
-		(n0) edge  [out=-75, in=75]            node {} (n9);
-\end{tikzpicture}
-\end{minipage}
-\caption{Expansion of the transition graph into a multigraph.}
-\label{fig:multigraph}
-\end{figure}
-
-Classical interval analysis on this program yields $\lstinline|x_old| \in [-100000,100000]$.
-We have transformed the first program into the second, manually pruning out infeasible paths; yet in general the resulting program could be exponentially larger than the first (as in Fig.~\ref{fig:multigraph}), even though not all feasible paths are needed to compute the invariant.
-
-Following recent suggestions \cite{Gawlitza_Monniaux_ESOP11,Monniaux_Gonnord_SAS11}, we avoid this space explosion by keeping the second program implicit while simulating its analysis.
-This means we work on an implicitly represented transition multigraph (Fig.~\ref{fig:multigraph}); it is succinctly representated by the transition graph of the first program.
-Our first contribution (\S\ref{sec:guided_multigraph}) is to recast the ``guided analysis'' from \citet{DBLP:conf/sas/GopanR07} on such a succinct representation of the paths in lieu of the individual transitions.
-A similar explosion occurs in disjunctive invariant generation, following \citet{DBLP:conf/pldi/GulwaniZ10}; our second contribution (\S\ref{sec:disjunctive}) applies our implicit representation to their method.
-
-\subsection{Guided static analysis}
-\label{sec:guided}
-\emph{Guided static analysis} was proposed by \citet{DBLP:conf/sas/GopanR07} as an improvement over classical upward Kleene iterations with widening.
-Consider the following program, taken from \cite{DBLP:conf/sas/GopanR07}:
-\lstinputlisting[label=lst:gopan_reps]{gopan_reps.c}
-
-\begin{figure}
-\begin{center}
-\includegraphics[scale=0.75]{figures/gopan_reps}
-\end{center}
-\caption{The invariant for Prog.~\ref{lst:gopan_reps}: the piecewise linear, solid line is the strongest invariant, the grayed polyhedron is its convex hull.}
-\label{fig:gopan_reps_invariant} 
-\end{figure}
-
-Classical iterations on the domain of convex polyhedra \cite{CousotHalbwachs78,BagnaraHRZ05SCP} or octagons \cite{DBLP:journals/lisp/Mine06} start with $x = 0 \land x = 0$, then continue with $x = y \land 0 \leq x \leq 1$.
-The widening operator extrapolates from these two iterations and yields $x = y \land x \geq 0$.
-From there, the ``else'' branch at line 5 may be taken; with further widening, $0 \leq y \leq x$ is obtained as a loop invariant, and thus the postcondition computed at line 9 is $x \geq 0 \land y = 0$.
-Yet the strongest invariant is $(0 \leq x \leq 51 \land y = x) \lor (51 \leq x \leq 102 \land x+y=102)$, and its convex hull, a convex polyhedron (Fig.~\ref{fig:gopan_reps_invariant}), is
-\begin{equation}
-y \leq x \land y \leq 102-x \land y \geq 0.\label{eqn:triangle}
-\end{equation}
-Intuitively, this disappointing result is obtained because widening extrapolates from the first iterations of the loop, but the loop has two different phases ($x \leq 50$ and $x > 50$) with different behaviors, thus the extrapolation from the first phase is not valid for the second.
-
-\citeauthor{DBLP:conf/sas/GopanR07}' idea is to analyze the first phase of the loop with a widening and narrowing sequence, and thus obtain $0 \leq x \leq 50 \land y = x$, and then analyze the second phase, finally obtaining invariant~\ref{eqn:triangle}; each phase is identified by the tests taken or not taken.
-
-The analysis starts by identifying the test taken and not taken during the first iteration of the loop, starting in the loop initialization.
-The branches not taken are pruned from the loop body, yielding:
-\begin{lstlisting}[numbers=none]
-  while(1) {
-    if(x <= 50) y++;
-    else break; /* not taken in phase 1 */
-    if(y < 0) break;
-    x++;
-  }
-\end{lstlisting}
-
-Analyzing this loop using widening and narrowing on convex polyhedra or octagons yields the loop invariant $0 \leq x \leq 51 \land y = x$.
-Now, the transition at line~5 becomes feasible; and we analyze the full loop, starting iterations from $0 \leq x \leq 51 \land y = x$, and obtain invariant~(\ref{eqn:triangle}).
-
-More generally, this analysis method considers an ascending sequence of subsets of the transitions in the loop body (left side of Fig.~\ref{fig:multigraph});
-for each subset, an inductive invariant is computed for the program restricted to it.
-The starting subset are the transitions reachable in one step from the loop initialization.
-If for a given subset $S$ in the sequence, no transitions outside $S$ are reachable from the inductive invariant attached to $S$, then iterations stop;
-otherwise, add these transitions to $S$ and iterate more.
-Termination ensues from the finiteness of the control-flow graph.
-
-\subsection{Path-focusing}
-\label{sec:path_focusing}
-
-\citet{Monniaux_Gonnord_SAS11}'s \emph{path-focusing} technique
-distinguishes the different paths in the program in order to avoid loss of
-precision due to merge operations. Since the number of paths may be exponential,
-the technique keeps them implicit and computes them when needed using
-SMT-solving.
-The (accelerated) Kleene iterations (\S\ref{sec:static_analysis}) are computed over a reduced multigraph instead of the classical transition graph.
-
-Let $P$ be the set of control points in the transition graph, 
-$P_W \subseteq P$ the set of widening points
-such that removing the points in $P_W$ gives an acyclic graph.
-One can choose a set $P_R$ such that $P_W \subseteq P_R \subseteq P$.
-
-The set of paths is kept implicit by an SMT formula $\rho$ expressing
-the semantics of the program, assuming that the transition semantics can be
-expressed within a decidable theory. For an easy construction of $\rho$, 
-we also assume that the program is expressed in SSA form, meaning that each
-variable is only assigned once in the transition graph. This is not a
-restriction, since there exist standard algorithms that transform a program into
-an SSA format.
-
-This formula contains Boolean \emph{reachability predicates} $b_i$ for each
-control points $p_i \notin P_R$, $b_i^s$ and $b_i^d$ for each $p_i \in P_R$, so
-that a path 
-$p_{i_1} \rightarrow p_{i_2} \rightarrow \dots \rightarrow p_{i_n}$ 
-between two points $p_{i_1}, p_{i_n} \in P_R$ 
-can easily be expressed as the
-conjunction $b_{i_1}^s \wedge \bigwedge_{2 \leq k < n} b_{i_k} \wedge b_{i_n}^d$.
-The Boolean $b_{i}^s$ is $true$ when the path starts at point $p_i$, whereas
-$b_i^d$ is $true$ when the path arrives at $p_i$. In other words, we split the
-points in $P_R$ into a \emph{source} point, with only outgoing transitions, and
-a \emph{destination} point, with only incoming transitions, so that the
-resulting graph is acyclic and there is no paths going through control
-points in $P_R$.
-
-In order to find focus paths, we solve an SMT formula which is satisfiable when
-there exist a path starting at a point $p_i \in P_R$ in a state included in the
-current invariant candidate $X_i$, and arriving at a point $p_j \in P_R$ in a
-state outside $X_j$. In this case, we construct this path using the model and
-update $X_j$. When $p_i = p_j$, meaning that the path is actually a self-loop,
-we can apply a widening/narrowing sequence, or even compute the transitive
-closure of the loop (or an approximation thereof, or its application to $X_i$)
-using abstract acceleration~\cite{DBLP:conf/sas/GonnordH06}.
-
-\section{Guided analysis over the paths}
-\label{sec:guided_multigraph}
-
-Guided static analysis, as proposed by \citet{DBLP:conf/sas/GopanR07}, applies to the transition
-graph of the program. We now present a new technique applying this analysis on the implicit
-multigraph from \cite{Monniaux_Gonnord_SAS11}, thus avoiding control flow merges with
-unfeasible paths.
-In this section, we use the same notations as \ref{sec:path_focusing}, except
-that we call the abstract values $X_i^s$ instead of $X_i$ to leave room for
-another abstract value $X_i^d$ later in the algorithm.
-
-The combination of these two techniques aims at first discovering a precise
-inductive invariant for a subset of paths between two points in $P_R$, 
-by the mean of ascending and narrowing iterations. When an
-inductive invariant has been found, we add new feasible paths to the subset and
-compute an inductive invariant for this new subset, starting with the results
-from the previous analysis.
-In other words, our technique considers an ascending sequence of
-subsets of the paths between two points in $P_R$.
-We iterate the operations until the whole program (i.e all the
-feasible paths) has been considered. The result will then be an
-inductive invariant of the entire program.
-
-\subsection{Algorithm}
-
-Algorithm~ \ref{algo:combined} performs Guided static analysis on the implicitly represented multigraph.
-
-The current working subset of paths, noted $P$ and initially empty, is
-stored using a compact representation, such as binary decision
-diagrams. We also maintain two sets of control points:
-\begin{itemize}
-	\item $A'$ contains the points in $P_R$ that may be the starting points of new
-		feasible paths.
-	\item $A$ contains the points in $P_R$ on which we apply the ascending iterations.
-	Each time the abstract value of a control point $p$ is updated, $p$ is
-	inserted in both $A$ and $A'$.
-\end{itemize}
-
-\begin{algorithm}
-	\caption{Guided static analysis on implicit multigraph}
-	\label{algo:combined}
-	\begin{algorithmic}[1] 
-	\input{algo/algo-combined.tex}
-	\end{algorithmic}
-\end{algorithm}
-
-We distinguish three phases in the main loop of the analysis:
-\begin{enumerate}
-\item \label{step:addingpaths} We start finding a new relevant subset
-  $P \cup P'$ of the graph.
-  Either the previous iteration or the initialization lead us to a
-  state where there are no more paths in the previous subset $P$,
-  starting at $p_i$, that make the abstract values of the successors
-  grow (otherwise, the SMT solver would not have answered
-  ``\emph{unsat}''). Narrowing iterations preserve this property.
-  However, there may exist such paths in the entire multigraph, that
-  are not in $P$. This phase computes these paths and adds them to
-  $P'$. This phase is described in~\ref{subsec:addingpaths}
-  and corresponds to lines in~\ref{alg=start-add-paths} to
-  \ref{alg=end-add-paths} in Algorithm~\ref{algo:combined}.
-\item \label{step:ascending} Given a new subset $P$, we search for paths starting at point
-  $p_i \in P_R$, such that these paths are in $P$, i.e are included in
-  the working subgraph. Each time we find a path, we update the
-  abstract value of the destination point of the path. This is the
-  phase explained in~\ref{subsec:ascending}, and corresponds to
-  lines~\ref{alg=start-ascending} to~\ref{alg=end-ascending} in
-  Algorithm~\ref{algo:combined}.
-\item  \label{step:narrowing} We perform narrowing iterations the usual way
-  (line~\ref{alg=narrowing} in algorithm~\ref{algo:combined}) and
-  reiterate from step 1 unless there are no more points to explore,
-  i.e. $A' = \emptyset$.
-\end{enumerate}
-
-%\MM{Je trouve qu'il manque une vue de haut niveau quelque part par
-%  ici. Dire qu'en gros, l'algo, c'est (en pseudo-code ou en langue naturelle) :
-%
-%Sans ca, enchainer tout de suite sur ``ascending iterations'' est un peu brutal.
-%
-%Ensuite, on peut rafiner la définition de ``Subset'' ($A$ et $P$), et
-%enchainer sur l'algo 2.
-%}
-
-Step~\ref{step:ascending} corresponds to the application of
-path-focusing \cite{Monniaux_Gonnord_SAS11} to
-the elements of $A$ until $A$ is empty. 
-When $A$ becomes empty, it means the
-invariant computation of the subgraph is finished. As proposed by
-\citet{DBLP:conf/sas/GopanR07}, we can do some narrowing iterations.
-These narrowing iterations allow to recover precision lost by
-widening, \emph{before} computing and taking into account new feasible paths.
-Thus, our technique combines both the advantages of \emph{Guided Static
-Analysis} and \emph{Path-focusing}.
-
-The order of steps is important: narrowing has to be performed before
-adding new paths, or some spurious new paths would be added to $P$,
-and starting with the addition of new paths avoids wasting time
-doing the ascending iterations on an empty graph.
-
-\subsection{Ascending iterations by Path-focusing}
-\label{subsec:ascending}
-
-\MM{Maintenant que l'ordre a changé dans l'algo, on peut peut-être
-  échanger cette section et la suivante ?}
-
-For computing an inductive invariant over a subgraph, we use the
-Path-focusing algorithm from \citet{Monniaux_Gonnord_SAS11} with special
-treatment for self loops (line~\ref{alg=pf} in algorithm~\ref{algo:combined}).
-
-In order to find which path to focus on, we construct an SMT formula $f(p_i)$, whose
-model when satisfiable is a path that starts in $p_i$, goes to a successor $p_j
-\in P_R$ of $p_i$, such that the image of $X_{i}^s$ by the path transformation
-is not included in the current $X_{j}^s$.
-Intuitively, such a path makes the abstract value $X_{j}^s$ grow, and thus is
-an interesting path to focus on. We loop until the formula becomes unsatisfiable,
-meaning that the analysis of $p_i$ is finished.
-
-If we note $Succ(i)$ the set of indices $j$ such that $p_j \in P_R$ is a
-successor of $p_i$ in the expanded multigraph, and $X_i^s$ the abstract value
-associated to $p_i$ :
-$$f(p_i) = \rho \wedge b_i^s \wedge 
-\displaystyle\bigwedge_{j \in P_R \atop j \neq i} \neg
-b_j^s \wedge X_i^s \wedge \displaystyle\bigvee_{j \in Succ(i)} (b_j^d \wedge
- \neg X_j^s)$$
-
-The difference with~\citet{Monniaux_Gonnord_SAS11} is that we do not
-work on the entire transition graph but on a subset of it. Therefore we
-conjoin the formula $f(p_i)$ with the actual set of working paths,
-noted $P$, expressed as a Boolean formula, where the Boolean variables are the
-\emph{reachability predicates} of the control points. We can easily construct
-this formula from the binary decision diagram using dynamic programming, and
-avoiding an exponentially sized formula. In other words, we force the SMT solver
-to give us a path included in $P$.
-
-\subsection{Adding new paths}
-\label{subsec:addingpaths}
-
-Our technique computes the fixpoint iterations on a ascending sequence of
-subgraphs, until the complete graph is reached.
-When the analysis of a subgraph is finished, meaning that the abstract values
-for each control points has converged to an inductive invariant for this subgraph,
-the next subgraph to work on has to be computed.
-
-This new subgraph is the union of the paths of the previous one with a set
-$P'$ of new paths that become feasible regarding the current abstract values.
-The paths in $P'$ are computed one after another, until no more path
-can make the invariant grow. This is line~\ref{alg=computeNewPaths} in
-Algorithm~\ref{algo:combined}, which corresponds to
-Algorithm~\ref{algo:computepaths}. We also use SMT solving to discover
-these new paths, but we subtly change the SMT formula given to the
-SMT solver: we now simply check for the satisfiability of $f(p_i)$
-instead of $f(p_i) \wedge P$. Since we already know that $f(p_i)
-\wedge P$ is unsatisfiable, none of the paths given by the SMT solver
-will be in $P$.
-
-% À changer si on décide de ne plus distinguer $X^s$ et $X^d$
-Also, to prevent the algorithm from adding too many paths at a time,
-and in particular to prevent the algorithm from trying to re-add the
-same path infinitely many times, we use another abstract value
-associated to the control point $p_j$, noted $X_j^d$, which is distinct
-from $X_j^s$, and initialized to $X_j^s$ right before computing new paths. 
-In the SMT formula, we associate to $p_j$ the abstract value $X_j^s$ when $p_j$
-is the starting point of the path, and $X_j^d$ when it is its destination point.
-We thus check the satisfiability of the formula
-$f'(p_i)$, where:
-$$f'(p_i) = \rho \wedge b_i^s \wedge 
-\displaystyle\bigwedge_{j \in P_R \atop j \neq i} \neg
-b_j^s \wedge X_i^s \wedge \displaystyle\bigvee_{j \in Succ(i)} (b_j^d \wedge
-\neg X_j^d)$$
-
-$X_j^d$ is updated when the point $p_j$ is the target of a new path.
-This way, further SMT queries do not
-compute other paths with the same source and destination if it is not
-needed (because these new paths would not make $X_j^d$ grow, hence
-would not be returned by the SMT solver).
-
-\begin{algorithm}
-	\caption{ComputeNewPaths}
-	\label{algo:computepaths}
-	\begin{algorithmic}[1] 
-	\input{algo/compute-paths.tex}
-	\end{algorithmic}
-\end{algorithm}
-
-\subsection{Termination}
-Termination of this algorithm is guaranteed, because:
-\begin{itemize}
-\item 
-the subset of paths $P$ strictly increases at each loop iteration, and bounded by the finite set of
-paths in the entire graph. 
-\item the set $P'$ always verifies $P \cap P' = \emptyset$ by construction, 
-which guarantees that $P'$
-will eventually be empty after a finite number of loop iterations.
-\end{itemize}
-
-\subsection{Example}
-
-We revise the rate limiter described in \ref{subsec:rate_lim}. In this example,
-\emph{Path-focusing} works well because all the paths starting at the
-loop header are actually self loops. In such a case, the technique performs a
-widening/narrowing sequence or accelerates the loop, thus leading to a precise
-invariant. However, in some cases, there also exist paths that are not
-self loops, in which case \emph{Path-focusing} applies widening. 
-This widening may induce unrecoverable loss of precision.
-
-Suppose the main loop of the rate limiter contains a nested loop:
-\lstinputlisting{rlim_int_loop.c}
-
-We choose $P_R$ as the set of loop headers of the function, plus the initial
-state. In this case, we have three elements in $P_R$.
-We also unroll the loop once, in order to distinguish the paths that do not go
-through the loop from the others:
-\begin{lstlisting}[numbers=none]
-if (wait()) {
-   while (wait()) {}
-}
-\end{lstlisting}
-
-The main loop in the expanded multigraph has then 8 distinct paths, half of
-them being self loops, and the other half going to the header of the nested loop.
-
-Guided static analysis from \citet{DBLP:conf/sas/GopanR07} yields, at line 5,
-$\lstinline|x_old| \in (-\infty,+\infty)$.
-Path-focusing \citep{Monniaux_Gonnord_SAS11} performs slightly
-better, and finds $\lstinline|x_old| \in (-\infty,10000]$.
-Now, let us see how our technique performs on this example.
-
-Figure \ref{fig:example-graph} shows the sequence of subset of paths during the
-analysis. The points in $P_R$ are noted $p_i$, where $i$ is the corresponding
-line in the code: for instance, $p_5$ corresponds to the header of the main
-loop.
-
-\begin{enumerate}
-	\item The first subset of paths is depicted on Figure
-		\ref{fig:example-graph} Step 1. We apply Path-Focusing on this graph,
-		and discover for $p_5$ the inductive invariant $\lstinline|x_old| = 0$.
-	\item
-		We then compute the set $P'$ of paths that have to be added into the
-		next subgraph. 
-		The image of $\lstinline|x_old| = 0$ by the path that goes from $p_5$ to
-		itself, and that goes through the \emph{else} branch of each 
-		\emph{if-then-else}, is $-10 \leq \lstinline|x_old| \leq 10$. This path
-		is then added to our subgraph. Moreover, there is no other path whose
-		image is not in $-10 \leq \lstinline|x_old| \leq 10$.
-		Additionally, the path that goes from $p_5$ to
-		$p_{10}$, that goes through the \emph{else} branch of each 
-		\emph{if-then-else} at lines 7,8 and 9, is also 
-		$-10 \leq \lstinline|x_old| \leq 10$, and thus is also added into the
-		new subgraph. Again, there is no new path that goes to $p_{10}$ with an
-		image outside $-10 \leq \lstinline|x_old| \leq 10$.
-	\item
-		We apply Path Focusing on this new subgraph (\ref{fig:example-graph}
-		Step 2). When we have reached an inductive invariant, we narrow, and the
-		result for both $p_5$ and $p_{10}$ is the polyhedron 
-		$-10000 \leq \lstinline|x_old| \leq 10000$.
-	\item Finally, we compute the new subgraph. The SMT-solver does not find any
-		new path that makes the abstract values grow, and the algorithm
-		terminates.
-\end{enumerate}
-
-Our technique gives us the expected invariant 
-$\lstinline|x_old| \in~[-10000,\allowbreak 10000]$. 
-Here, only 3 paths out of the 8 have been computed during the analysis. In
-practice, depending on the order the SMT-solver returns the paths, other
-feasible paths could have been added during the analysis.
-
-\begin{figure}
-\label{fig:example-graph}
-\centering
-\begin{minipage}[c]{.15\textwidth}
-\begin{tikzpicture}[->,>=stealth',auto,node distance=2.1cm,
-                    semithick,font=\footnotesize]
-
-	\node[PRstate] (n00) {$p_4$};
-	\node[PRstate] (n0) [below of=n00, yshift=0.8cm] {$p_5$};
-	\node[PRstate] (n1) [below of=n0] {$p_{10}$};
-	\node (label) [below of=n1, yshift=1.2cm] {Step 1};
-
-  \path [transition] 
-		(n00) edge  node {$\lstinline|x_old| \gets 0$} (n0);
-\end{tikzpicture}
-\end{minipage} 
-\begin{minipage}[c]{.28\textwidth}
-\begin{tikzpicture}[->,>=stealth',auto,node distance=2.1cm,
-                    semithick,font=\footnotesize]
-
-	\node[PRstate] (n00) {$p_4$};
-	\node[PRstate] (n0) [below of=n00, yshift=0.8cm] {$p_5$};
-	\node[PRstate] (n1) [below of=n0] {$p_{10}$};
-	\node (label) [below of=n1, yshift=1.2cm] {Step 2};
-
-  \path [transition] 
-		(n00) edge  node {$\lstinline|x_old| \gets 0$} (n0);
-  \path [transition] 
-		(n0) edge  [loop left] node [left, xshift=0.2cm] {
-		$\begin{array}{r}
-			-10000 \leq \lstinline|x| \leq 10000 \\
-			\lstinline|x_old|-10 \leq \lstinline|x| \leq \lstinline|x_old|+10 /\\
-			\lstinline|x_old| \gets \lstinline|x|
-		\end{array}$
-		} (n0);
-  \path [transition] 
-		(n0) edge node [left, yshift=-0.2cm, xshift=0.2cm] {
-		$\begin{array}{r}
-			-10000 \leq \lstinline|x| \leq 10000 \\
-			\lstinline|x_old|-10 \leq \lstinline|x| \leq \lstinline|x_old|+10 /\\
-			\lstinline|x_old| \gets \lstinline|x|
-		\end{array}$
-		} (n1);
-\end{tikzpicture}
-\end{minipage}
-\caption{Ascending sequence of subgraphs}
-\end{figure}
-
-In this example, we see that our technique actually combines best of
-\emph{Guided Static Analysis} and \emph{Path Focusing}. Still, we need to unroll
-the loop once in order to get precise results. Section \ref{sec:disjunctive}
-proposes a new technique that gives precise invariants without needing such
-graph transformations.
-
-\section{Disjunctive invariants}
-\label{sec:disjunctive}
-
-\citet{DBLP:conf/pldi/GulwaniZ10} proposed a technique for computing disjunctive invariants, by
-distinguishing all the paths inside a loop. In
-this section, we propose to improve this technique by using SMT queries to find
-interesting paths, the objective being to avoid an explicit exhaustive
-enumeration of an exponential number of paths.
-
-For each control point $p_i$, we compute a disjunctive invariant
-$\bigvee_{1\leq j \leq m_i} X_{i,j}$. We denote by $n_i$ the number of
-distinct paths starting at $p_i$.
-To perform the analysis, one chooses an integer $\delta_i \in [1,m_i]$, and
-a mapping function $\sigma_i: [1,m_i] \times [1,n_i] \mapsto [1,m_i]$.
-  The $k$-th path starting fom $p_i$ is denoted $\tau_{i,k}$.
-  The image of the $j$-th disjunct $X_{i,j}$ by the path $\tau_{i,k}$ is then
-  joined with $X_{i,\sigma(j,k)}$.
-Initially, the $\delta_i$-th abstract value contains the initial states of
-$p_i$, and all other abstract values contain~$\emptyset$.
-
-For each control point $p_i \in P_R$, $m_i$, $\delta_i$ and $\sigma_i$ can be defined heuristically.
-For instance, one could define $\sigma_i$ so that $\sigma_i(j,k)$ only depends on the
-last transition of the path, or else construct it dynamically during the
-analysis.
-
-Our method improves this technique in two ways :
-\begin{itemize}
-\item Instead of enumerating the whole set of paths, we keep them implicit and
-compute them only when needed.
-
-\item At each loop iteration of the original algorithm \citep{DBLP:conf/pldi/GulwaniZ10}, an image by each path inside the loop is computed for each disjunct of the invariant candidate.
-Yet, many of these images may be redundant: for instance, if our invariant candidate is $(0 \leq x \leq 10 \land 0 \leq y \leq 1000) \lor (x < -10 \land y < -10)$, then there is no point enumerating paths whose image is included in this invariant candidate.
-In our approach, we compute such an image only if it makes the resulting abstract value grow.
-\end{itemize}
-
-Our improvement consists in a modification of the SMT formula we solve in
-\ref{sec:guided_multigraph}.
-We introduce in this formula Boolean variables $\{d_j, 1 \leq j \leq m\}$, so
-that we can easily find in the model which abstract value of the
-disjunction of the source point has
-to be chosen to make the invariant of the destination grow.
-The resulting formula that is given to the SMT solver is defined
-by $g(p_i)$.
-When the formula is satisfiable, we know that the index $j$ of the starting
-disjunct that has to be chosen is the one for which the associate Boolean value
-$d_j$ is \emph{true} in the model. Then, we can easily compute the value of 
-$\sigma_i(j,k)$, thus know the index of the disjunct to join with.
-
-\begin{eqnarray*}
-g(p_i) & = & \rho \wedge b_i^s \wedge 
-\displaystyle\bigwedge_{j \in P_R \atop j \neq i} \neg b_j^s  \\
- & & \wedge 
-\displaystyle\bigvee_{1 \leq k \leq m_i} (d_k \wedge X_{i,k} \wedge \bigwedge_{l \neq k}
-\neg d_l) \\
- & & \wedge
-\displaystyle\bigvee_{j \in Succ(i)} 
-(b_j^d \wedge \bigwedge_{1 \leq k \leq m_i} (\neg X_{j,k}))
-\end{eqnarray*}
-
-In our algorithm, the initialization of the abstract values slightly differs from
-algorithm \ref{algo:combined} line~\ref{alg=X-init}, since we now have to
-initialize each disjunct. Line~\ref{alg=X-init} is then replaced by:
-\begin{algorithm}[!h]
-\begin{algorithmic}[1] 
-\FORALL {$k \in \{1,..,m_i\} \setminus \{\delta_i\}$}
-	\STATE $X_{i,k} \gets \perp$
-\ENDFOR
-\STATE $X_{i,\delta_i} \gets I_{p_i}$
-\end{algorithmic}
-\end{algorithm}
-
-Furthermore, the Path-focused algorithm (line~\ref{alg=pf} from algorithm
-\ref{algo:combined}) is enhanced to deal with disjunctive invariants, and
-is detailed in algorithm \ref{algo:disjunctive}.
-
-The \emph{Update} function can classically assign to $X_{i,\sigma_i(j,k)}$ the
-value $X_{i,\sigma_i(j,k)} \widening (X_{i,\sigma_i(j,k)} \sqcup
-\tau_{i,k}(X_{i,j}))$, or can integrate the special treatment for self loops
-proposed by \citet{Monniaux_Gonnord_SAS11}, with widening/narrowing sequence or
-acceleration.
-
-\begin{algorithm}[!h]
-\caption{Disjunctive invariant computation with implicit paths}\label{gulwani2}
-\label{algo:disjunctive}
-\begin{algorithmic}[1] 
-\WHILE {true}
-	\STATE $res \gets SmtSolve\left[g(p_i)\right]$
-	\IF {$res = unsat$}
-		\STATE \textbf{break}
-	\ENDIF
-	\STATE Compute the path $\tau_{i,k}$ from $res$ 
-	\STATE Take $j \in \{ l | d_l = true\}$ 
-	\STATE Update($X_{i,\sigma_i(j,k)})$
-\ENDWHILE
-\end{algorithmic}
-\end{algorithm}
-
-We experimented with a heuristic
-of dynamic construction of the $\sigma_i$ functions, adapted from
-\citet{DBLP:conf/pldi/GulwaniZ10}. 
-\DM{Cette heuristique, je croyais que c'était nous qui l'avions trouvée,
-et qu'elle n'était pas dans Gulwani?}
-\JH{Non. Gulwani fait la même chose en un peu plus complique (voir commentaire
-suivant)}
-\MM{Du coup, la contribution~\ref{contr:disjunctive} devrait être
-  affaiblie un peu je suppose. Gulwani fait aussi du SMT-solving ?}
-\DM{Non, Gulwani énumère tous les chemins un par un à tous les coups.}
-\MM{OK, du coup j'ai ajouté une phrase à la fin de la section pour
-  rappeler notre contrib}
-For each control point $p_i \in P_R$, we start with one single disjunct
-($m_i = 1$) and define $\delta_i = 1$.
-$M$ denotes an upper bound on the number of disjuncts per control point.
-
-The $\sigma_i$ functions take as parameters the index of the starting abstract
-value, and the path we focus on. 
-Since we dynamically construct these functions
-during the analysis, we store their already computed image into a
-compact representation, such as Algebraic Decision Diagrams.
-$\sigma_i(j,k)$ is then constructed on the fly only when needed, and computed
-only once.
-When the value of $\sigma_i(j,k)$ is required but undefined, we first compute
-the image of the
-abstract value $X_{i,j}$ by the path indexed by $k$, and try to find an existing
-disjunct of index $j'$ so that the least upper bound of the two abstract values is exactly their union (this can be tested using SMT-solving).
-If such an index exists, then we set $\sigma_i(j,k) = j'$.
-Otherwise:
-\begin{itemize}[itemsep=-1\parsep,topsep=0pt]
-	\item if $m_i < M$, we increase $m_i$ by $1$ and define $\sigma_i(j,k) = m_i$
-	\item if $m_i = M$, we define $\sigma_i(j,k) = M$ 
-		%\JH{A la place, Gulwani cherche l'invariant qui a le plus de contraintes
-		%d'egalite en commun pour definir son $sigma$. Nous, on le met dans M, pour simplifier}
-\end{itemize}
-The main difference with the original
-algorithm~\cite{DBLP:conf/pldi/GulwaniZ10} is that we construct
-$\sigma_i(j,k)$ using SMT queries instead of enumerating a possibly
-exponential number of paths to find a solution.
-
-\section{Experimental comparisons}
-\label{sec:experiments}
-
-We have implemented our proposed solutions inside a prototype intraprocedural
-static analyzer,
-as well as the classical abstract interpretation algorithm, and the state-of-the-art
-techniques \emph{Path Focusing} \cite{Monniaux_Gonnord_SAS11} and \emph{Guided
-Static Analysis} \cite{DBLP:conf/sas/GopanR07}.
-
-%\MM{On peut peut-être citer le nom de l'outil (on garde PAGAI ?) ici.
-%  L'intérêt d'avoir un nom, c'est que les gens peuvent en parler plus
-%  facilement, et c'est plus facile de retrouver le papier origine d'un
-%  outil quand il cite son nom}
-%\DM{Dans la version finale, oui, mais pas dans une soumission anonymisée... suffit de Googlifier pagai sur imag.fr}
-%
-Our tool operates over LLVM bitcode \citep{LLVM_langref,Lattner:2004:LCF:977395.977673}, which is a target for several compilers, most notably Clang (supporting C and C++) and llvm-gcc (supporting C, C++, Fortran and Ada).
-Abstract domains are provided by the APRON library \citep{DBLP:conf/cav/JeannetM09}, and include convex polyhedra (from the builtin Polka ``PK'' library), octagons, intervals, and linear congruences.
-For SMT-solving, our analyzer uses Yices
-\cite{DBLP:conf/cav/DutertreM06} or Microsoft Z3\cite{DBLP:conf/tacas/MouraB08}.
-
-We conducted extensive experiments on real-life programs in order to compare the
-different techniques, mostly on open-source projects (Fig.~\ref{fig:projects}) written in C, C++ and Fortran.
-
-\begin{table}[!h]
-	\centering
-\begin{tabular}{|l|r|r|} \hline
-	\multicolumn{1}{|c|}{Name} &
-        \multicolumn{1}{c|}{kLOC} &
-        \multicolumn{1}{c|}{$|P_R|$} \\ \hline
-	a2ps & 55 & 2012\\
-	gawk & 59 & 902\\ 
-	gnuchess & 38 & 1222\\ 
-	gnugo & 83 & 2801\\
-	grep & 35 & 820\\
-	gzip & 27 & 494\\
-	lapack/blas & 954 & 16422\\
-	make & 34 & 993\\ 
-	sed & 23 & 292\\
-	tar & 73 & 1712\\
-	\hline
-\end{tabular}
-\caption{List of analyzed open-source projects, with their respective number of
-lines of code, and their number of control points in $P_R$}
-\label{fig:projects}
-\end{table}
-
-\subsection{Analysis algorithm}
-\label{sec:analysis-algorithm}
-For each program, we distinguish a set $P_R = P_W$ of suitable widening points by a simple algorithm: for each procedure, compute the strongly connected components of its control-flow graph using Tarjan's algorithm; the targets of the back-edges of the depth-first search are added to~$P_R$.
-Note that the resulting set is not necessarily minimal, but is sufficient to disconnect all cycles --- more sophisticated techniques are discussed in e.g.
-\citet{BourdonclePhd}.
-
-LLVM bitcode is in \emph{static single assignment} (SSA) form: a given scalar variable is given a value at a single syntactic point in the program.
-In concrete terms, an assignment \lstinline|x=2*x+1;| gets translated into a definition $x_2 = 2x_1+1$, with distinct variables $x_1$ and $x_2$ corresponding to the same original variable \lstinline|x| at different points in the program.
-LLVM makes it easy to follow definition-use and use-definition chains: for a given variable (say, $x_2$) one can immediately obtain its definition (say, $2x_1+1$).
-One may see conversion to SSA form as a static precomputation of some of the symbolic propagations proposed by \citet{DBLP:conf/vmcai/Mine06} to enhance the precision of analyses.
-
-SSA introduces $\phi$-functions at the head of a control code to define variables whose value depends on which incoming edge to this control node was last taken.
-For instance, for \lstinline|if (...) { x = 2*x+1; } else { x= 0; }|, then $x_2$ is defined as $\phi(2x_1+1,0)$.
-In this framework, each variable is uniquely defined as an arithmetic ($+$, $-$, $\times$, $/$) function of other variables that themselves may not be representable as affine linear functions, because they are defined using $\phi$-functions, bitwise arithmetic, loads from memory, or return values from function calls.
-
-This motivates a key implement decision of our tool: only those variables $v_1,\dots,v_n$ that are not defined by arithmetic operations are retained as coordinates in the abstract domain (e.g.
-as dimensions in polyhedra).
-A basic block of code therefore amounts to a \emph{parallel assignment} operation
-$(v_1,\dots,v_n) \allowbreak\mapsto\allowbreak
-(f_1(v_1,\dots,v_n), \allowbreak, \dots, \allowbreak
- f_n(v_1,\dots,v_n))$;
-such operations are directly supported by APRON. This has three benefits:
-(i) it limits the number of dimensions in the abstract values, since polyhedra libraries typically perform worse with higher dimensions;
-(ii) the abstract operation for a single path in path-focusing methods also is a (large) parallel assignment;
-(iii) as suggested by \citet{DBLP:conf/vmcai/Mine06}, this approach is more precise than running abstract operations for each program line separately:
-for instance, for \lstinline|y=x; z=x-y;| with precondition $x \in [0,1]$, a line-by-line interval analysis obtains $y \in [0,1]$ and $z \in [-1,1]$ while our ``en bloc'' analysis symbolically simplifies $z = x - x = 0$ and thus $z \in [0,0]$.
-
-In the event that a node is reachable only by a single control-flow edge (which may occur because of dead code, or during the first phases of guided static analysis), the $\phi$ operation reduces to a copy of the values flowing from that edge.
-In this case, our tool just propagates symbolic values through the predecessor node, without introducing $\phi$-variables.
-
-We further reduce the number of variables by projecting out those that are not live at the program point. Again, SSA use-def and def-use chains make it easy to compute liveness.
-
-Our tool currently only operates over scalar variables from the SSA representation and thus cannot directly cope with arrays or memory accessed through pointers.
-We therefore run it after the ``memory to registers'' (\texttt{mem2reg}) optimization phase in LLVM, which lifts most memory accesses to scalar variables.
-The remaining memory accesses are treated as undefined values, as are the return
-values of function calls. For better precision, we also apply function inlining
-to the program, and we unroll every loop once.
-
-Our tool currently assumes that integer variables are unbounded mathematical integers ($\ZZ$) and floating-point variables are real (or rational) numbers.
-Techniques for sound analysis of bounded integers, including with wraparound, and of floating-point operations have been developed in e.g.
-the Astr\'ee system \citep{ASTREE_ESOP05,ASTREE_PLDI03}, but porting these techniques to our iteration schemes using SMT-solving requires supplemental work.
-
-Our implementation of path-focusing currently does not use true acceleration techniques, as proposed by \citet{Monniaux_Gonnord_SAS11}.
-Instead, it simply runs widening and narrowing iterations on a single path.
-The analysis is currently only forward, even though nothing in what we describe in this article is specific to forward analysis.
-
-\subsection{Precision of the various techniques}
-\label{sec:compare_techniques}
-
-\begin{figure}[h]
-  \begin{center}
-    \input{./techniques.tex}
-  \end{center} 
-  \vspace{-20px}
-  \caption{Comparison of the abstract values obtained by classical abstract
-  interpretation (S), \emph{Guided Static
-  Analysis} (G), \emph{Path-focused} technique (PF), our combined technique
-  (G+PF), and its version with disjunctive invariants (DIS).
-  The $\subsetneq$ bars gives the percentage of invariants stronger (more precise; smaller with respect to inclusion) with the left-side technique,
-$\supsetneq$ the percentage of invariants stronger with the right-side technique,
-and ``uncomparable'' gives the percentage of invariants that are uncomparable, i.e
-neither greater nor smaller;
-the code points where both invariants are equal make up the remaining percentage. See Tab.~\ref{tab:techniques} for details.}
-  \label{fig:techniques}
-\end {figure}
-
-\begin{table*}
-\begin{center}
-\setlength{\tabcolsep}{0.75ex}
-\begin{tabular}{|l
-|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
-|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
-|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
-|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
-|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
-|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}|} \hline
-\multicolumn{1}{|c|}{\textbf{Benchmark}}
-& \multicolumn{3}{c|}{\textbf{G/S}}
-& \multicolumn{3}{c|}{\textbf{PF/S}}
-& \multicolumn{3}{c|}{\textbf{PF/G}}
-& \multicolumn{3}{c|}{\textbf{G+PF/PF}}
-& \multicolumn{3}{c|}{\textbf{G+PF/G}}
-& \multicolumn{3}{c|}{\textbf{DIS/G+PF}} \\ %\cline{2-16}
-& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
-& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
-& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
-& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
-& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
-& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.} \\
- \hline
- \input{table1}
-\end{tabular}
-\end{center}
-\caption{Result of the comparison of the various techniques described in this
-paper: classic Abstract Interpretation (S), \emph{Guided Static Analysis} (G),
-\emph{Path-focusing} (PF), our combined technique (G+PF), and its version using
-disjunctive invariants (DIS). For
-instance, \textbf{G/S} compares the benefits of \emph{Guided Static Analysis}
-over the classic Abstract interpretation algorithm.
-$\subsetneq$, $\supsetneq$ and ``unc.'' are defined as in Fig.~\ref{fig:techniques}.}
-\label{tab:techniques}
-\end{table*}
-
-For each program and each pair $(T_1,T_2)$ of analysis techniques, we list
-the proportion of control points in $P_R$ where $T_1$ (resp.~$T_2$) gives a
-strictly stronger invariant, denoted by $\subsetneq$ (resp. $\supsetneq$),
-and the proportion of control points where the invariants given by $T_1$ and
-$T_2$ are uncomparable for the inclusion ordering (the remainder of the
-control points are thus those for which both techniques give the same
-invariant). We use convex polyhedra as the abstract domain.
-
-Let us briefly comment the results given in more details in Table~\ref{tab:techniques} and Figure~\ref{fig:techniques}.
-\emph{Guided Static Analysis} from \citet{DBLP:conf/sas/GopanR07} improves the
-result of the classical Abstract Interpretation in $2.72\%$ of the control points
-in $P_R$.
-\emph{Path-focusing} from \citet{Monniaux_Gonnord_SAS11} gives
-statistically better results, and finds better invariants than \emph{Guided
-Static Analysis} in $9.89\%$ of the cases. However, it also loses precision in an
-important number ($2.68\%$) of control points.
-Finally, our combined technique gives the most promising results, since it is
-statistically more precise than the other techniques.
-
-The analysis using disjunctive invariants greatly improves the
-precision of the analysis (for $12.06\%$ of the control points in $P_R$), at an
-acceptable time cost (see Table \ref{tab:time}).
-
-While experiencing with techniques that use SMT-solving, we encountered some
-limitations due to non-linear arithmetic in the analyzed programs. Indeed, 
-the SMT-solver is not able to decide the satisfiability of some SMT-formulae
-expressing the semantics of non-linear programs. 
-In this case, we skipped the functions for which the SMT-solver returned the
-``unknown'' result.
-This limitation occurred very rarely in our experiments, except for the analysis
-of \emph{Lapack/Blas}, where 798 over the 1602 functions have been skipped.
-\emph{Lapack/Blas} implements matrix computations, which use floating-point multiplications.
-
-In cases where the formula is expressed in too rich a logic for the SMT-solver to deal with, a number of workarounds are possible, including:
-(i) \emph{Linearization}, as per \citet{DBLP:conf/vmcai/Mine06}, which overapproximates nonlinear semantics by linear semantics.
-(ii) Replacing the results of nonlinear operations by ``unknown''.
-Neither is currently implemented in our tool.
-
-Table \ref{tab:time} gives the execution time of the different analysis
-techniques. It is interesting to see that \emph{Guided Static Analysis} and
-\emph{Path-focusing} are sometimes faster than the classical
-algorithm.
-This seems due to the fact that these algorithms start on a small subset
-of the graph, and then may need fewer iterations on the complete graph.
-The techniques we introduce in this article are slower, but speed is still on the same order of magnitude as the older ones.
-
-\begin{table}[!h]
-	\centering
-\begin{tabular}{|l|r|r|r|r|r|} \hline
-	\multicolumn{1}{|c|}{Benchmark} &
-        \multicolumn{1}{c|}{\textbf{S}} &
-        \multicolumn{1}{c|}{\textbf{G}} &
-        \multicolumn{1}{c|}{\textbf{PF}} &
-        \multicolumn{1}{c|}{\textbf{G+PF}} &
-	\multicolumn{1}{c|}{\textbf{DIS}} \\ \hline
-	\input{table_time.tex} \hline
-\end{tabular}
-\caption{Execution time for each technique, expressed in seconds}
-\label{tab:time}
-\end{table}
-
-\subsection{Precision of Abstract Domains}
-\label{sec:compare_domains}
-\begin{table*}
-\begin{center}
-\setlength{\tabcolsep}{0.75ex}
-\begin{tabular}{|l
-|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
-|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
-|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
-|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
-|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
-|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}|} \hline
-\multicolumn{1}{|c|}{\textbf{Benchmark}}
-& \multicolumn{3}{c|}{\textbf{PK/OCT}}
-& \multicolumn{3}{c|}{\textbf{PK/BOX}}
-& \multicolumn{3}{c|}{\textbf{OCT/BOX}}
-& \multicolumn{3}{c|}{\textbf{PK/PKEQ}}
-& \multicolumn{3}{c|}{\textbf{PK/GRID}}
-& \multicolumn{3}{c|}{\textbf{POLY*/POLY}} \\ %\cline{2-16}
-& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
-& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
-& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
-& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
-& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
-& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.} \\
- \hline
- \input{table_domain.tex}
-\end{tabular}
-\end{center}
-\caption{Results of the comparison of the various abstract domains, when using
-the same technique (G+PF). We used as abstract domains Convex Polyhedra (PK and
-POLY),
-Octagons (OCT), intervals (BOX), linear equalities (PKEQ) and linear congruences
-(GRID). The last column compares the domain of Convex Polyhedra with the
-improved widening operator from \citet{BagnaraHRZ05SCP} (POLY*), and Convex Polyhedra
-using the classical widening operator (POLY). POLY and POLY* use the Parma
-Polyhedral Library \cite{PPL}.
-$\subsetneq$, $\supsetneq$ and ``unc.'' are defined as in Fig.~\ref{fig:techniques}.}
-\label{tab:domain}
-\end{table*}
-
-For each program and each pair $(D_1,D_2)$ of abstract domains, we compare by
-inclusion the invariants of the different control points in $P_R$ (where $P_R$
-is defined as in \ref{sec:compare_techniques}).
-The main results are given in Table \ref{tab:domain}.
-
-Statistically, the domain of convex polyhedra gives the better results, but
-commonly yields weaker invariants than the domains of octagons/intervals;
-this is a known weakness of its widening operator~\cite{Monniaux_LeGuen2011}. 
-The Octagon domain appears to be much better than intervals; this is
-unsurprising since in most programs and libraries, bounds on loop indices are
-non constant: they depend on some parameters (array sizes etc.).
-
-The Lapack/Blas benchmarks are unusual compared to the other programs.
-These libraries perform matrix computations, using nested loops over indices; such programs are the prime target for polyhedral loop optimization techniques and it is therefore unsurprising that polyhedra perform very well over them.
-
-The analysis of linear equalities (PKEQ) performs very fast compared
-to other abstract domains, but yields very imprecise invariants:
-it only detects relations of the form $\sum_i a_i x_i = C$ where $a_i$ and $C$ are constants.
-The comparison between the domain of convex polyhedra and linear congruences
-shows the percentage of points where the use of congruences refines the
-invariant. This is the case for $1.04\%$ of the control points in $P_R$.
-
-Finally, we evaluated the benefits of the improved version of the widening
-operator for convex polyhedra from \citet{BagnaraHRZ05SCP}, compared to the
-classical widening. We found that the improved version from
-\citet{BagnaraHRZ05SCP} yields more precise invariants for $6.58\%$ of the
-control points in $P_R$.
-
-\section{Conclusion and future prospects}
-Roughly, an analysis by abstract interpretation is defined by the choice of an iteration strategy and an abstract domain.
-In this article, we demonstrated that changes in the iteration algorithm can significantly improve precision, sometimes while improving analysis times.
-
-A common criticism of analysis techniques based on SMT-solving is that they
-do not scale up. Yet, our experiments show that, for numerical properties,
-they scale up to the size of typical functions and loops.
-It is however
-quite certain that, naively applied, they cannot scale to the kind of
-programs targeted by e.g. the Astr\'ee tool, that is, a dozens or hundreds
-of thousands of lines of code in a single loop operating over similar numbers
-of remanent variables.
-Actually, for such applications, only (quasi-)linear
-algorithms scale up, and ``cheap'' abstract domains such as octagons ($O(n^3)$ where $n$ is the number of variables) are not applied to the full variable set, but to restricted subsets thereof.
-It thus seems reasonable that techniques such as considering ``packs'' of related variables, slicing, etc. may similarly help SMT-based techniques to scale to global analyses.
-
-We compared the precision of different techniques and abstract domains by comparing the invariants for the inclusion ordering.
-A better metric is perhaps to take a client analysis --- such as the detection of overflows and array bound violations --- and compare the rates of alarms.
-
-We focused on numerical properties, because they are supported by easily available abstract libraries.
-Yet, in most programs, properties of data structures are important for proving interesting properties.
-Further investigations are needed not only on good abstractions for pointers (many are already known) but also on their conversion to SMT problems.
-
-% Cas a probleme:
-% Initialisation (x,y) à {(1,0), (0,1)}
-% puis boucle: si y = 0 alors x++
-% fait un polyèdre qui finit par perdre y <= 1 et x >= 0
-
-%\appendix
-%\section{Appendix Title}
-
-%This is the text of the appendix, if you need one.
-
-%\acks
-
-% Acknowledgments, if needed.
-
-% We recommend abbrvnat bibliography style.
-
-\phantomsection\addcontentsline{toc}{section}{References} 
-\bibliographystyle{dmabbrvnat}
-\bibliography{implicitization}
-\end{document}
-
-% Pour Emacs:
-% Local Variables:
-% reftex-cite-format: natbib
-% End:

### 2012-05-23 papers/TAPAS2012/pagai.tex

@@
 x) __assert_fail();|, and the tool just tests for the reachability of \lstinline|__assert_fail();|: if it is unreachable, then the assertion is true.
+\lstinline|assume| works with the same principle, and is defined as \lstinline|if (! x) __assumption_declared()|. Since \lstinline|__assert_fail| and
+\lstinline|__assumption_declared| are \emph{noreturn} functions, the code after
+the test verifies \lstinline|x|.
 
@@
 \emph{Guided Static Analysis} from \citet{DBLP:conf/sas/GopanR07} improves the
-result of the classical Abstract Interpretation in $2.72\%$ of the control points
+result of the classical Abstract Interpretation in $2.21\%$ of the control points
 in $P_R$.
-\emph{Path-focusing} from \citet{Monniaux_Gonnord_SAS11} gives
-statistically better results, and finds better invariants than \emph{Guided
-Static Analysis} in $9.89\%$ of the cases. However, it also loses precision in an
-important number ($2.68\%$) of control points.
+\emph{Path-focusing} from \citet{Monniaux_Gonnord_SAS11} 
+finds better invariants in $4.13\%$ of the control points.
+
+However, these two techniques also lose precision in an important number
+($4.64\%$ for G, $5.14\%$ for PF) of control points, and obtain worse results
+than the classical many times. This result is unexpected, and could be partially
+explained by bad behaviour of the widening operator.
+
 Finally, our combined technique gives the most promising results, since it is
-statistically more precise than the other techniques.
+statistically more precise than the other techniques. It improves the precision
+of the inductive invariant in $8.29\%$ to $9.86\%$ of the control points
+compared to the three previous techniques. Still, we obtain worse result in a
+non-negligible number of cases ($2.02\%$).
 
 The analysis using disjunctive invariants greatly improves the
-precision of the analysis (for $12.06\%$ of the control points in $P_R$), at an
-acceptable time cost (see Table \ref{tab:time}).
+precision of the analysis (for $14.46\%$ of the control points in $P_R$ compared
+to G+PF), at the expense of a higher time cost (see Table \ref{tab:time}).
+It also gives worse results in $6.85\%$ of the points, most probably because of
+a non optimal choice of the $\sigma$ function, detailed in
+\cite{Henry_Monniaux_Moy_SAS12}.
 
@@
 algorithm.
-This seems due to the fact that these algorithms start on a small subset
-of the graph, and then may need fewer iterations on the complete graph.
-The techniques we introduce in this article are slower, but speed is still on the same order of magnitude as the older ones.
+This seems due to the fact that this algorithm
+computes inductive invariant on a small number of control points compared to
+classical approaches, thus leading to fewer operations over abstract values.
 
@@
 & \multicolumn{3}{c|}{\textbf{PK/PKEQ}}
-& \multicolumn{3}{c|}{\textbf{PK/GRID}}
-& \multicolumn{3}{c|}{\textbf{POLY*/POLY}} \\ %\cline{2-16}
+& \multicolumn{3}{c|}{\textbf{PK/PKGRID}}
+& \multicolumn{3}{c|}{\textbf{POLY/POLY*}} \\ %\cline{2-16}
 & \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
@@
 POLY),
-Octagons (OCT), intervals (BOX), linear equalities (PKEQ) and linear congruences
-(GRID). The last column compares the domain of Convex Polyhedra with the
+Octagons (OCT), intervals (BOX), linear equalities (PKEQ) and the reduced
+product of NewPolka convex polyhedra with linear congruences
+from the Parma Polyhedra Library \cite{BagnaraHZ08SCP}.
+(PKGRID). The last column compares the domain of Convex Polyhedra with the
 improved widening operator from \citet{BagnaraHRZ05SCP} (POLY*), and Convex Polyhedra
-using the classical widening operator (POLY). POLY and POLY* use the Parma
-Polyhedral Library \cite{BagnaraHZ08SCP}.
+using the classical widening operator (POLY). POLY and POLY* use the PPL\cite{BagnaraHZ08SCP}.
 $\subsetneq$, $\supsetneq$ and ``unc.'' are defined as in Tab.~\ref{tab:techniques}.}
@@
 
-The Lapack/Blas benchmarks are unusual compared to the other programs.
-These libraries perform matrix computations, using nested loops over indices; such programs are the prime target for polyhedral loop optimization techniques and it is therefore unsurprising that polyhedra perform very well over them.
+The Lapack/Blas benchmarks are unusual compared to the other programs. These
+libraries perform matrix computations, using nested loops over indices; such
+programs are the prime target for polyhedral loop optimization techniques and it
+is therefore unsurprising that polyhedra and octagons perform very well over them.
 
@@
 it only detects relations of the form $\sum_i a_i x_i = C$ where $a_i$ and $C$ are constants.
-The comparison between the domain of convex polyhedra and linear congruences
-shows the percentage of points where the use of congruences refines the
-invariant. This is the case for $1.04\%$ of the control points in $P_R$.
+
+Using the reduced product of convex prolyhedra with linear congruences
+(PKGRID) improves the analysis by $2.52\%$.
 
@@
 classical widening. We found that the improved version from
-\citet{BagnaraHRZ05SCP} yields more precise invariants for $6.58\%$ of the
+\citet{BagnaraHRZ05SCP} yields more precise invariants for $3.70\%$ of the
 control points in $P_R$.
 
+\subsection{Future Work}
 It is not totally relevant to compare by inclusion the abstract values obtained

### 2014-09-18 WCET/generateSMTwcet.py
WCET tools

@@ -1,0 +1,540 @@
+import sys
+import commands
+import re
+import argparse
+import json
+from heapq import heappush, heappop
+
+number_articulations = 0
+labels = {}
+vars = {}
+debug = False
+cut_id = 0
+first_label = ''
+last_label = ''
+first_var = ''
+first_var_bd = ''
+last_var = ''
+heap_cuts = []
+my_graph = {}
+
+def match_names(blocks):
+    global labels
+    global vars
+    global first_label
+    global last_label
+    global first_var
+    global first_var_bd
+    global last_var
+
+    first_var_ok = False
+    for i in range(1, len(blocks)):
+        id, instructions = blocks[i].split(':', 1)
+
+        var = id[1:]
+        if instructions.find('<label>:') != -1:
+            #print 'NOT IMPLEMENTED'
+            label = instructions.split('<label>:')[1].split(' ')[0]
+        else:
+            label = instructions.split('\n', 1)[1].split(':')[0]
+
+        if "bd_" in var:
+            if first_var_ok:
+                last_label = label
+                last_var = var
+            else:
+                first_var_ok = True
+                first_label = label
+                first_var_bd = var
+                var = var.replace("bd","bs")
+                first_var = var
+
+        labels[var] = label
+        vars[label] = var
+
+label_costs = {}
+transition_costs = {}
+def readmatchingfile(matchingfile):
+    global costs
+    #for line in matchingfile:
+    #    name,arm,cost = line.split(',')
+    #    name = name.replace("(%", "")
+    #    name = name.replace(" ", "")
+    #    cost = cost.replace(")\n", "")
+    #    cost = int(cost.replace(" ", ""))
+    #    label_costs[name] = cost
+    for line in matchingfile:
+        source,dest,cost = line.split(',')
+        source = source.replace("(", "")
+        source = source.replace("%", "")
+        source = source.replace(" ", "")
+        dest = dest.replace("%", "")
+        dest = dest.replace(" ", "")
+        cost = cost.replace(")\n", "")
+        cost = int(cost.replace(" ", ""))
+        if dest == "":
+            label_costs[source] = cost
+        else:
+            #print source
+            #print dest
+            if source not in transition_costs:
+                transition_costs[source] = {}
+            transition_costs[source][dest] = cost
+
+heads = []
+tails = []
+semanticcut = []
+semanticcut_maxcost = []
+
+def addautomaticcuts():
+    global my_graph
+    #for mnode in my_graph[last_var][2]:
[…]

### 2014-09-18 WCET/go.sh

@@ -1,0 +1,273 @@
+#!/bin/bash
+
+SCRIPTDIR=`dirname $0`
+
+export LD_LIBRARY_PATH=$SCRIPTDIR/../external/z3/lib/:$LD_LIBRARY_PATH
+export PAGAI_PATH=$SCRIPTDIR/../src/
+
+function usage () {
+echo "
+Usage:
+./go.sh <OPTION> -c [FILE]
+
+OPTION :
+	-h        : help
+	-c <FILE> : C input file
+	-M <FILE> : matching file
+	-S <FILE> : longest syntactic path file
+	-b        : FILE given by option -c is already a bitcode FILE
+	-s        : do not add summaries in the SMT formula
+	-l        : avoid non-linear arithmetic in the SMT formula
+	-d        : create .dot file
+	-n        : skip bitcode optimisations
+	-I        : with input functions when running charcute_bitcode
+	-m        : print model of the longest feasible path
+	-T <int>  : set the ulimit -t time (default:120)
+	-R        : Recursive cuts
+"
+}
+
+SUMMARIES=1
+NONLINEAR=1
+DOT=0
+SKIPCLANG=0
+SKIPOPTIM=0
+PRINTMODEL=""
+DOMATCHING=0
+PRINTSYNTACTIC=0
+RECURSIVECUTS=0
+WITHINPUTFUNCTION=""
+MATCHINGFILE=""
+ULIMIT=120
+
+while getopts “hc:lsdbmnM:S:IT:R” opt ; do
+	case $opt in
+		h)
+			usage
+			exit 1
+            ;;
+		c)
+			FILE=$OPTARG
+			;;
+		M)
+            DOMATCHING=1
+			MATCHINGFILE=$OPTARG
+			;;
+		S)
+            PRINTSYNTACTIC=1
+            LONGESTSYNTACTICFILE=$OPTARG
+			;;
+		T)
+            ULIMIT=$OPTARG
+			;;
+		s)
+			SUMMARIES=0
+			;;
+		l)
+			NONLINEAR=0
+			;;
+		d)
+			DOT=1
+			;;
+		b)
+			SKIPCLANG=1
+			;;
+		I)
+			# not required anymore 
+			WITHINPUTFUNCTION=""
+			;;
+		n)
+			SKIPOPTIM=1
[…]

### 2014-09-18 src/Analyzer.cc

@@ -34,6 +34,7 @@ bool skipnonlinear;
 bool has_timeout;
 bool optimize;
 bool withinlining;
+bool withbrutalunrolling;
 bool onlydumpll;
 bool wcet_settings;
 bool invasmetadata;
@@ -166,6 +167,10 @@ bool check_overflow() {
 
 bool inline_functions() {
 	return withinlining;
+}
+
+bool brutal_unrolling() {
+	return withbrutalunrolling;
 }
 
 bool dumpll() {
@@ -488,6 +493,7 @@ int main(int argc, char* argv[]) {
 	has_timeout = false;
 	optimize = false;
 	withinlining = true;
+	withbrutalunrolling = false;
 	onlydumpll = false;
 	wcet_settings = false;
 	annotatedFilename = NULL;
@@ -521,6 +527,7 @@ int main(int argc, char* argv[]) {
 			{"svcomp",  no_argument, 0, 'V'},
 			{"optimize",  no_argument, 0, 'O'},
 			{"noinline",  no_argument, 0, 'I'},
+			{"loop-unroll",  no_argument, 0, 'U'},
 			{"dump-ll",  no_argument, 0, 'z'},
 			{"wcet",  no_argument, 0, 'w'},
 			{NULL, 0, 0, 0}
@@ -528,7 +535,7 @@ int main(int argc, char* argv[]) {
 	/* getopt_long stores the option index here. */
 	int option_index = 0;
 
-	 while ((o = getopt_long(argc, argv, "qa:ShDi:o:s:c:Cft:d:e:nNMTm:k:pvb:VOIzwB:",long_options,&option_index)) != -1) {
+	 while ((o = getopt_long(argc, argv, "qa:ShDi:o:s:c:Cft:d:e:nNMTm:k:pvb:VOIzwB:U",long_options,&option_index)) != -1) {
         switch (o) {
 			case 0:
 				assert(false);
@@ -560,6 +567,9 @@ int main(int argc, char* argv[]) {
         	    break;
         	case 'I':
         	    withinlining = false;
+        	    break;
+        	case 'U':
+        	    withbrutalunrolling = true;
         	    break;
         	case 'z':
 				onlydumpll = true;

### 2014-09-18 src/Analyzer.h

@@ -130,6 +130,7 @@ bool optimizeBC();
 
 bool WCETSettings();
 bool inline_functions();
+bool brutal_unrolling();
 bool dumpll();
 
 extern llvm::raw_ostream *Out;

### 2014-09-18 src/Execute.cc

@@ -217,12 +217,18 @@ void execute::exec(std::string InputFilename, std::string OutputFilename) {
 	
 	PassManager OptPasses;
 	OptPasses.add(new GlobalToLocal());
-	//OptPasses.add(createLoopSimplifyPass());
 	OptPasses.add(createLoopRotatePass());
-	//OptPasses.add(createLCSSAPass());
-	//OptPasses.add(createLoopUnrollPass(-1,1,1,0));
+
 	if (!WCETSettings())
 		OptPasses.add(new RemoveUndet());
+
+	if (brutal_unrolling()) {
+		OptPasses.add(createLoopSimplifyPass());
+		OptPasses.add(createLCSSAPass());
+		OptPasses.add(createIndVarSimplifyPass());
+		OptPasses.add(createScalarReplAggregatesPass());
+		OptPasses.add(createLoopUnrollPass(INT_MAX,INT_MAX,1,0));
+	}
 	OptPasses.add(createPromoteMemoryToRegisterPass());
 	OptPasses.run(*M);
 

### 2014-09-18 src/globaltolocal.cc

@@ -59,10 +59,16 @@ bool GlobalToLocal::runOnModule(Module &M) {
 				Builder.SetInsertPoint(entry->begin());
 				std::string name(global->getName());
 				AllocaInst * A = Builder.CreateAlloca(ty,NULL,Twine(name));
-				if (init) Builder.CreateStore(init,global,false);
+				if (init) {
+					Builder.CreateStore(init,global,false);
+				}
 				//we replace only inside the function
 				replaceAllUsesInFunction(i,global,A);
 				//global->replaceAllUsesWith(A);
+				if (!init) {
+					LoadInst * load = Builder.CreateLoad(global,false);
+					Builder.CreateStore(load,A);
+				}
 			}
 		}
 	}

### 2014-09-18 WCET/go.sh
wcet

@@ -5,6 +5,27 @@ SCRIPTDIR=`dirname $0`
 export LD_LIBRARY_PATH=$SCRIPTDIR/../external/z3/lib/:$LD_LIBRARY_PATH
 export PAGAI_PATH=$SCRIPTDIR/../src/
 
+function check_path () {
+	if [ -z "`which pagai`" ] ; then
+		echo "pagai not found in $PAGAI_PATH"
+		exit
+	fi
+	if [ -z "`which z3`" ] ; then
+		echo "z3 not found in \$PATH. Please update your \$PATH variable"
+		exit
+	fi
+	if [ -z "`which clang`" ] ; then
+		echo "clang not found in \$PATH. Please update your \$PATH variable"
+		exit
+	fi
+	if [ -z "`which opt`" ] ; then
+		echo "opt not found in \$PATH. Please update your \$PATH variable"
+		exit
+	fi
+}
+
+check_path
+
 function usage () {
 echo "
 Usage:
@@ -20,7 +41,6 @@ OPTION :
 	-l        : avoid non-linear arithmetic in the SMT formula
 	-d        : create .dot file
 	-n        : skip bitcode optimisations
-	-I        : with input functions when running charcute_bitcode
 	-m        : print model of the longest feasible path
 	-T <int>  : set the ulimit -t time (default:120)
 	-R        : Recursive cuts
@@ -40,7 +60,7 @@ WITHINPUTFUNCTION=""
 MATCHINGFILE=""
 ULIMIT=120
 
-while getopts “hc:lsdbmnM:S:IT:R” opt ; do
+while getopts “hc:lsdbmnM:S:T:R” opt ; do
 	case $opt in
 		h)
 			usage
@@ -71,10 +91,6 @@ while getopts “hc:lsdbmnM:S:IT:R” opt ; do
 			;;
 		b)
 			SKIPCLANG=1
-			;;
-		I)
-			# not required anymore 
-			WITHINPUTFUNCTION=""
 			;;
 		n)
 			SKIPOPTIM=1
@@ -118,13 +134,16 @@ function ccompile {
 }
 
 function bcoptimize {
-	$PAGAI_PATH/pagai -i $dir/$name.bc --dump-ll --wcet --loop-unroll > $dir/$name.ll
[…]

### 2014-09-18 src/Execute.cc

@@ -227,7 +227,7 @@ void execute::exec(std::string InputFilename, std::string OutputFilename) {
 		OptPasses.add(createLCSSAPass());
 		OptPasses.add(createIndVarSimplifyPass());
 		OptPasses.add(createScalarReplAggregatesPass());
-		OptPasses.add(createLoopUnrollPass(INT_MAX,INT_MAX,1,0));
+		//OptPasses.add(createLoopUnrollPass(INT_MAX,INT_MAX,1,0));
 	}
 	OptPasses.add(createPromoteMemoryToRegisterPass());
 	OptPasses.run(*M);

### 2014-09-18 src/SMTpass.cc

@@ -668,6 +668,7 @@ void SMTpass::computeRhoRec(Function &F,
 		ToBeComputed.pop();
 		bool first = (visited->count(b) > 0);
 		visited->insert(b);
+		if (first) continue;
 
 		if (!FPr->inPr(b) || b == dest) {
 			// we recursively construct Rho, starting from the predecessors

### 2014-09-23 WCET/generateSMTwcet.py
WCET

@@ -54,6 +54,9 @@ def match_names(blocks):
 
 label_costs = {}
 transition_costs = {}
+# read the file given by Mihail's code, that consists in tuples
+# of the form (source,destination,cost) if cost is attached to transitions
+# or form (block,cost) if cost is attached to blocks
 def readmatchingfile(matchingfile):
     global costs
     #for line in matchingfile:
@@ -533,7 +536,6 @@ def main():
             c = heappop(heap_cuts)
             cuts_file.write(c[1]+' '+c[2]+'\n')
         cuts_file.close()
-
 
     #print json.dumps(my_graph[last_var])
 sys.setrecursionlimit(10000)

### 2014-09-23 WCET/go.sh

@@ -35,13 +35,10 @@ OPTION :
 	-h        : help
 	-c <FILE> : C input file
 	-M <FILE> : matching file
-	-S <FILE> : longest syntactic path file
 	-b        : FILE given by option -c is already a bitcode FILE
 	-s        : do not add summaries in the SMT formula
 	-l        : avoid non-linear arithmetic in the SMT formula
-	-d        : create .dot file
 	-n        : skip bitcode optimisations
-	-m        : print model of the longest feasible path
 	-T <int>  : set the ulimit -t time (default:120)
 	-R        : Recursive cuts
 "
@@ -49,7 +46,6 @@ OPTION :
 
 SUMMARIES=1
 NONLINEAR=1
-DOT=0
 SKIPCLANG=0
 SKIPOPTIM=0
 PRINTMODEL=""
@@ -60,7 +56,7 @@ WITHINPUTFUNCTION=""
 MATCHINGFILE=""
 ULIMIT=120
 
-while getopts “hc:lsdbmnM:S:T:R” opt ; do
+while getopts “hc:lsbnM:T:R” opt ; do
 	case $opt in
 		h)
 			usage
@@ -73,10 +69,6 @@ while getopts “hc:lsdbmnM:S:T:R” opt ; do
             DOMATCHING=1
 			MATCHINGFILE=$OPTARG
 			;;
-		S)
-            PRINTSYNTACTIC=1
-            LONGESTSYNTACTICFILE=$OPTARG
-			;;
 		T)
             ULIMIT=$OPTARG
 			;;
@@ -86,17 +78,11 @@ while getopts “hc:lsdbmnM:S:T:R” opt ; do
 		l)
 			NONLINEAR=0
 			;;
-		d)
-			DOT=1
-			;;
 		b)
 			SKIPCLANG=1
 			;;
 		n)
 			SKIPOPTIM=1
-			;;
-		m)
-			PRINTMODEL="-m"
 			;;
 		R) 
 			RECURSIVECUTS=1
@@ -158,23 +144,24 @@ if [ $SKIPOPTIM -eq 0 ] ; then
 fi
 echo -e "IR file is " $GREEN "$dir/$name.bc" $NORMAL
 
-if [ $DOT -eq 1 ] ; then
-	opt -dot-cfg $dir/$name.bc -o $dir/$name.bc
-fi
 
 gensmtformula
 
 if [ $DOMATCHING -eq 1 ] ; then
 	TEXFILE=benchmarksedges.tex
 	MATCHING=".edges"
[…]

### 2014-09-23 WCET/go.sh
wcet

@@ -37,7 +37,7 @@ OPTION :
 	-M <FILE> : matching file
 	-b        : FILE given by option -c is already a bitcode FILE
 	-s        : do not add summaries in the SMT formula
-	-l        : avoid non-linear arithmetic in the SMT formula
+	-l        : allow non-linear arithmetic in the SMT formula
 	-n        : skip bitcode optimisations
 	-T <int>  : set the ulimit -t time (default:120)
 	-R        : Recursive cuts
@@ -45,7 +45,7 @@ OPTION :
 }
 
 SUMMARIES=1
-NONLINEAR=1
+NONLINEAR=0
 SKIPCLANG=0
 SKIPOPTIM=0
 PRINTMODEL=""
@@ -76,7 +76,7 @@ while getopts “hc:lsbnM:T:R” opt ; do
 			SUMMARIES=0
 			;;
 		l)
-			NONLINEAR=0
+			NONLINEAR=1
 			;;
 		b)
 			SKIPCLANG=1

### 2014-09-26 WCET/go.sh
work on carlit

@@ -4,6 +4,7 @@ SCRIPTDIR=`dirname $0`
 
 export LD_LIBRARY_PATH=$SCRIPTDIR/../external/z3/lib/:$LD_LIBRARY_PATH
 export PAGAI_PATH=$SCRIPTDIR/../src/
+export PATH=$SCRIPTDIR/../external/llvm/bin/:$PATH
 
 function check_path () {
 	if [ -z "`which pagai`" ] ; then

### 2014-10-08 ex/wcet_bicycle2.c
examples

@@ -1,7 +1,13 @@
-void toto() {
+#include "../pagai_assert.h"
+
+void bicycle() {
   int count=0, phase=0;
   for(int i=0; i<10000; i++) {
-    if (phase == 0) { count += 2; phase = 1; }
-    else if (phase == 1) { count += 1; phase = 0; }
+    if (phase == 0) { 
+		count += 2; phase = 1; 
+	} else if (phase == 1) { 
+		count += 1; phase = 0; 
+	}
   }
+  assert(count <= 15000);
 }
